### 2013-08-22 CompExam.aviolov.tex
Initial commit

%%This is a very basic article template.
%%There is just one section and two subsections.
\documentclass{article}

\usepackage{amsmath}
\usepackage{amscd}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsthm}

\usepackage{circuitikz}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows,snakes,backgrounds}
% \usetikz
\usepackage{subfig}

\usepackage{algpseudocode}
\usepackage{algorithm}

\usepackage[super]{nth}
% \usepackage{appendix}
% \usepackage{listings}
% \usepackage{color}

\usepackage{hyperref}
%\usepackage{url}

\usepackage{cleveref}
\usepackage{cancel}

\usepackage{aviolov_style}
\usepackage{local_style}

\newtheorem{thm}{Theorem}[section]
\newtheorem{lemma}{Theorem}[thm]
% \theoremstyle{definition}
\newtheorem{ex}{Example}[thm]
\newtheorem{defn}{Definition}[thm]

\begin{document}

\title{Stochastic Optimal Control - with a view towards spike control in neurons
\\
{\small Advanced Comprehensive Exam in Mathematics}
\\
{\small Dept. of Math - University of Ottawa}
}
\author{Alexandre Iolov 
$<$\href{mailto:aiolo040@uottawa.ca}
		{aiolo040 at uottawa dot ca}$>$}

\date{January 10, 2013}

\maketitle

\abstract{We review main concepts from stochastic optimal control theory,
including stochastic differential equations, dynamic programing and Pontryagin's
Maximum Principle. We then discuss an optimal control problem in a simple
mathematical model of a neuron where the feedback is minimal and we have a
quasi-open-loop control which necessitates the use of a Maximum Principle for
the system's transition density.}

\tableofcontents

\section{Overview}
Stochastic Optimal Control in continuous time deals with finding
the input which optimizes the performance of a continuous-time dynamical systems
that contain a significant random component in their evolutions.

Here we present a brief overview of the subject starting from the basic notions
of Stochastic Differential Equations (SDEs), 
and then introducing the main tools of Optimal Control, namely the
Hamilton-Jacobi-Bellman equation of Dynamic Programming and the state-adjoint
equations arising from Pontryagin's Maximum Principle. We will not state
proofs, although we sometimes provide heuristic derivations of the main
results.

If I had to learn the material all over again, I would proceed my readings as
follows: First I would read the online notes by Evans on both
SDEs,\cite{Evansa}, and Optimal Control, \cite{Evansb}. I would then read
Jacobs' book, \cite{Jacobs}, especially for the chapter on the Fokker-Planck
equation (ch 7) and then Oksendal \cite{Oksendal2007} for a
more thorough mathematical treatment of SDEs. Having familiarized myself with SDE's I would
return to Optimal Control in the Deterministic Case, starting with \cite{Kirk2004} for the
finite dimensional deterministic case (ODEs) and then \cite{Fleming1975} for the
extension to the stochastic case. I would keep the following for reference \cite{Krylov2008},
\cite{Hanson2007} and finally \cite{Fattorini1999} for the
deterministic infinite-dimensional (PDEs) case.

I have not used any particularly sophisticated numerical
methods, but, for example, Hanson \cite{Hanson2007} discusses both the
Euler-Maruyama scheme for simulating SDEs and the Crank-Nicholson Method for
solving PDEs which is all I have used in my codes.

All of the theoretical results presented in the sequel will be from
the above references.

\section{Stochastic Differential Equations}
In view of our ultimate goals, we will restrict ourselves to  stochastic
processes whose sample paths have continuous paths, i.e.\ to SDEs driven by
Brownian motion.

Since the reader is more likely well familiar with the material in this section,
we will not provide proofs, but only state the results with a view towards
establishing the notation for the sequel.

We will assume that the reader is familiar with the following concepts:
\begin{itemize} 
  \item a probability space, $\{\O, \sAlg, P\}$ consisting of a
  probability space, a sigma-algebra and a probability measure
  \item a continuous-time stochastic process, $X_t$
  \item a filtration $\Fil(t)$ and in particular the filtration generated by a
  stochastic process.
\end{itemize}

\subsection{The Wiener Process and the Ito Integral}
The Wiener Process is the fundamental building block of the stochastic calculus,
it is often called Brownian Motion and we will denote it $\{W_t\}_{t\geq 0}$. It
satisfies:
\begin{defn}Wiener Process, $W_t$:
\begin{enumerate}
  \item $W_0 = 0$
  \item $W_t - W_s = N(0, |t-s|)$ , i.e. normally distributed with mean 0,
  variance $|t-s|$)
  \item $\forall \{t_i\}_1^N, \quad \{W_{t_i} - W_{t_{i-1}} \}_2^N \sim$
  independent, i.e $W_t$ has independent increments
\end{enumerate}
\end{defn}
The fact that the finite incremental distributions suffice to specify a unique
continuous-time stochastic process is known as Kolmogorov's extension theorem.
\begin{ex}[from \cite{Evansa} \#21.] Inverted Wiener Process:
\\
Let $W_t$ be a 1-d Wiener Process and define:
\def \Wbar {{\bar{W} }}
\begin{equation*}
\Wbar(t) := \begin{cases}
t W(1/t) & t >0
\\
0 & t = 0
\end{cases}
\end{equation*}
Show $\Wbar(t)- \Wbar(s) \sim N(0, t-s)$.

\begin{proof}[Soln]
\begin{align*}
\Wbar(t) - \Wbar(s) =& t W(1/t) - s W(1/s), \, t>s
\\
=& (t-s)W(1/t) + s W(1/t) - sW(1/s) & \text{recall:} 1/s > 1/t
\\
=& (t-s)N(0, 1/t) - sN(0, 1/s -1/t) & \text{by indnt increments}
\\
=& N\left(0, \frac{(t-s)^2}t + \frac{s^2 }{ 1/s -1/t} \right) & \text{
addition of indnt normal RVs.}
\end{align*}
With some algebra, the variance in the last expression reduces to $t-s$
\end{proof}
\end{ex}

Now, we collect a few more relevant properties of $W_t$:
\begin{enumerate}
  \item The sample paths $W_{[0,\infty)}(\o)$
are almost surely (a.s.) continuous and are in fact Holder continuous for any
exponent $\g < 1/2$
\item The sample paths $W_{[0,\infty)}(\o)$ are nowhere differentiable
\item $W_t$ is a Markov process: $\Prob[W_t \in B | \s(W_{s' \leq s })] =
\Prob[W_t \in B \,| \,W_s]$, for any Borel set, $B \subset \R$, where $\s(W_{s'
\leq s })$ is the filtration generated by the $W_t$
\end{enumerate}

We now turn to defining stochastic integrals based on the Wiener Process.

\begin{defn} Progressively Measurable Functions:

Let $\Fil_t$ be the filtration generated by the Wiener Process.

Let $X_t$ be a stochastic process which is $\Fil_t$-measurable $\forall t$ and
which is jointly measurable in $(t,\o)$. We call such an $X_t$
\emph{progressively measurable}
\end{defn}

\begin{defn} $\Ltwopm, \Lonepm$

We define $\Ltwopm[0,T]$ as the space of all progressively measurable
$X$ such that
\begin{equation*}
\Exp \left[ \int_{[0,T]} X_t^2 \intd{t} \right] < \infty
\end{equation*}

Similarly, we define $\Lonepm[0,T]$ as the space of all progressively measurable
$X$ such that
\begin{equation*}
\Exp \left[ \int_{[0,T]} X_t \intd{t} \right] < \infty
\end{equation*} 
\end{defn}

$\Ltwopm$ will be the class of functions for which the Ito integral is
well-defined. 
\begin{defn} Ito Integral:
\label{defn:ito_integral}

Let $P^n := {a = t^n_1 \ldots t^n_{m_n} = b}$ be a partition of the interval
$[a,b] \subset [0, \infty)$. Let $|P_n| = \sup_i|t_i - t_{i-1}|$. 
Take $|P_n| \rightarrow_n 0 $ and consider an $ X_t \in \Ltwopm[a,b]$,

then
\begin{equation}
\int_{[a,b]} X_t \intd{W} := \lim_{n \rightarrow \infty}  
\sum_{i=1}^{m_n} X_{t_i}\left( W(t_{i+1}) - W(t_{i})\right)
\end{equation}

\end{defn}

To be precise, our definition is actually a theorem, and the real
definition is one that uses step functions and passes to the limit.  Also we
will write the limits of integration $\int_0^T \cdot  \intd{W}$ or $\int_{[0,T]}
\cdot  \intd{W}$ interchangeably.
% To justify def'n \ref{defn:ito_integral}, we recall:
% \begin{lemma}[Quadratic Variation]
% \begin{equation}
% \lim_{n \rightarrow \infty} \sum_{i=1}^{m_n} \left( W(t_i) - W(t_{i-1})\right)^2
% = b-a \quad \text{in }  L^2(\O)
% \end{equation}
% \end{lemma}
% The above can be read, heuristically as $\lim \sum (\Delta W)^2 \rightarrow \sum
% \Delta t$, which is the meaning behind the colloquial $\intd{W} =
% \sqrt{dt}$.

Again, we state without proof a few interesting properties of $\int X \intd{W}$:
\begin{thm} Ito Integral Properties

\begin{enumerate}
  \item $\Exp[\int_0^T X \intd{W} ] = 0$ 
  \item $\Exp[\left(\int_0^T X \intd{W}\right)^2 ] = \Exp[\int_0^T X^2
  \intd{t}]$
  \item $I(t) = \int_0^t X \intd{W} $ is a martingale 
  \item $I(t) = \int_0^t X \intd{W} $ has continuous sample paths
\end{enumerate}
\end{thm}

\subsection{Ito SDEs and Ito's Lemma}
\begin{defn} an Ito SDE:

We write
\begin{equation}
dX_t = Fdt + G dW
\end{equation}
on $0 \leq t \leq T$, if $X$ is a real-valued stochastic process satisfying:
\begin{equation*}
X(r) = X(s) + \int_{[s,r]} F \intd{t} + \int_{[s,r]} G \intd{W}
\quad 0\leq s \leq r \leq T
\end{equation*}
for some $F \in \Lonepm[0,T]$ and $G \in \Ltwopm[0,T]$
\end{defn}

We are now ready to present the celebrated Ito Lemma which is the chain-rule of
stochastic calculus:

\begin{thm}[Ito Lemma] Suppose $X$ satisfies the stochastic differential $dX_t =
Fdt + G dW$ as above and take $u(x,t) \in C^{2,1}[ \R \times [0,T]]$.

Set $Y(t) = u(X,t)$
\\
then
$$
dY =  \left( \di_t u + \di_x u \cdot F + \di^2_x u \cdot \frac{G^2}2 \right)
\intd{t} + \left(   \di_x u\cdot G  \right)\intd{W}
$$
\end{thm}

% We now present two examples, one fundamental and one more involved:
% \begin{ex}[pg 67 in \cite{Evansa}] The 'Stochastic' Exponential:
% 
% Let $Y = \exp(\l W - \l^2 t / 2)$. In the notation of the above Thm, we have:
% $dX = dW, F=0, G=1$ and $u(x,t) = \exp(\l x - \l^2 t / 2)$.
% 
% then
% \begin{align*}
% dY =& 
% \left( \di_t u + \di_x u \cdot F + \di^2_x u \cdot \frac{G^2}2 \right)\intd{t} +
% \left(   \di_x u \cdot G  \right)\intd{W}
% \\
% =&
% \left( -\frac{\l^2 t}{2}\exp(\l x - \l^2 t / 2) + \l^2\exp(\l x - \l^2 t / 2)  \right)\intd{t}
% \\ 
% &+ \left(   \l \exp(\l x - \l^2 t / 2) \right)\intd{W}  
% \\
% =&  \l \left( \exp(\l x - \l^2 t / 2) \right)\intd{W}  
% \\
% =& \l Y dW
% \end{align*}
% \end{ex}
% 
% \begin{ex}[pg. 67 in \cite{Evansa}] Hermite Polynomials:
% 
% Let $h_n(x,t) = \frac{(-t)^n}{n!} e^{x^2/2t} \di^n_x \left( e^{x^2/2t} 
% \right)$ be the $n$th Hermite polynomial
% \\
% then
% $$
% \int_0^t h_n(W, t) \intd{W} = h_{n+1}(W(t), t)) 
% $$
% i.e. in Ito calculus, the Hermite polynomials are like the power monomials
% $\frac{t^n}{n!}$ in the ordinary calculus.
% \begin{proof} Since
% $$
% d_\l^n [\exp(- \frac{(x-\l t)^2}{2t} ] |_{\l=0} =
%  (-t)^n d_x^n [ \exp(-\frac{x^2}{2t})] $$
% We will have:
% \begin{align*}
% d_\l^n [\exp(\l x - \l^2 t / 2) ] |_{\l=0} &= (-t)^n  e^{x^2/2t} d_x^n [
% \exp(-\frac{x^2}{2t})]
% \\
% &= n! h_n(x,t)
% \end{align*}
% Thus we can expand $\exp(\l x - \l^2 t / 2) $ in a Taylor series at $\l=0$ as:
% $$
% \exp(\l x - \l^2 t / 2) = \sum_n \l^n h_n(x,t)
% $$
% Now recall $Y$ from the previous example:
% \begin{equation*}
% Y(t) = \exp(\l W(t) - \frac{\l^2 t }2) = \sum_n \l^n h_n(W(t),t)
% \end{equation*}
% which satisfies:
% \begin{equation*}
% Y(t) = 1 + \l \int Y \intd{W}
% \end{equation*}
% Naturally we will plug the series expansion in the SDE to obtain:
% \begin{align*}
% \sum_{n=0}^{\infty} \l^n h_n(x,t) =&
% 1 + \l \int_0^t  \sum_{n=0}^{\infty} \l^n
% h_n(x,t)\intd{W}
% \\
% =& 1 + \sum_{n=1}^{\infty} \l^n\int_0^t 
% h_{n-1}(x,t)\intd{W} \quad {\textrm{bump $n$ up by one}}
% \end{align*}
% Equating the coefficients for $\l^n$ on both sides leads to the result
% \end{proof}
% \end{ex}

Finally, we are in a position to define an Ito SDE:

\begin{defn}[Ito SDE] Let
$F(x,t):\R\times [0,T]\ra \R,  G(x,t):\R\times [0,T]\ra \R,$ be given functions.
We say that a stochastic process $X$ satisfies:
\begin{equation}
dX =F dt + G dW
\end{equation} 
over $[0,T]$ if
\begin{enumerate}
  \item $X$ is progressively measurable wrt. $\Fil_t$
  \item $F(X,t) \in \Lonepm$
  \item $G(X,t) \in \Ltwopm$ 
  \item $X_t = X_0 + \int_0^t F(X_s, s) \intd{s} + \int_0^t G(X_s, s) \intd{W_s}$
\end{enumerate}
\end{defn}

% \begin{ex} The solution to 
% $$
% \begin{cases}
% dX = F(t)X dt + G(t) X dW
% \\
% X_0 = 1
% \end{cases}
% $$
% is 
% $$
% X_t = \exp\left( \int_0^t F(s) - \frac{G^2(s)}{2} \intd{s} + \int_0^t G
% \intd{W} \right) $$
% \begin{proof}
% Let $Y = \int_0^t f(s) - \frac{g^2(s)}{2} \intd{s} + \int_0^t g
% \intd{W} $
% and guess that $X = e^Y$
% then, using Ito's lemma:
% \begin{align*}
% dX =& (\di_y [e^y] dY + \frac{1}{2}\di^2_y[e^y]  g^2 dt
% \\ =&
% (e^y  f dt - e^y g^2 / 2) dt + e^y g dW + \frac{1}{2} e^y \cdot  g^2 dt
% \\ =& e^y \left( f dt +g dW \right)
% \\ =& X \left( f dt +g dW \right)
% \end{align*}
% \end{proof}
% \end{ex}

% \begin{ex}[Brownian Bridge, pg 80 in \cite{Evansa}] The solution of the SDE
% $$
% \begin{cases}
% dB = -\frac{B}{1-t} dt + dW
% \\
% B_0= 0
% \end{cases}
% $$
% is
% $$
% B(t) = (1-t) \int_0^t \frac{1}{1-s} dW
% $$
% \begin{proof}
% It is not clear how to apply Ito's lemma to this problem, so instead we just
% directly calculate the differential:
% \begin{align*}
% B(t^+) - B(t) =& (1-t^+) \int_0^{t^+} \frac{1}{1-s} dW - (1-t) \int_0^t
% \frac{1}{1-s} dW
% \\ =&
% (1-t^+) \int_t^{t^+}\frac{1}{1-s} dW  - (t^+-t) \int_0^t
% \frac{1}{1-s} dW
% \\ \approx&
% \int_t^{t^+} dW - \Delta t \frac{B(t)}{1-t}  
% \end{align*}
% \end{proof}
% \end{ex}
% Actually we can say a bit more about the Brownian bridge, in particular we
% will calculate $\Exp[B^2]$
%  \begin{ex}[Evans .37]
% \begin{align*}
% \Exp[B^2] =& \Exp\left[\left( (1-t) \int_0^t \frac{1}{1-s} dW \right)^2 \right]
% \\
% =&(1-t)^2 \Exp\left[ \int_0^t\left(  \frac{1}{1-s}\right)^2 \intd{s} \right]
% \\
% =&(1-t)^2 [ \frac{1}{1-t} - 1 ]
% \\
% =&(1-t) - (1-t)^2
% \end{align*}
% 
% Further note that  $\lim_{t\ra 1^-}\Exp[B^2] = 0$ and that $B$ is a martingale
% and so its first moment is always equal to its initial value, $0$. So indeed the 2nd
% moment is the variance and with the variance converging to zero, we can use
% Chebyshev's inequality to conclude that $\lim_{t\ra 1^-}B(t) = 0 $(a.s). This
% justifies the name of the Brownian bridge - it is a normally distributed RV.
% which is clamped to equal 0 at both $t= \{0,1\}$.
% \end{ex}

As an example we will discuss the Ornstein-Uhlenbeck process, which is the
basis for the models we will face later on.
\begin{ex}[O-U Process] Let $X$ follow:
\begin{equation*}
dX = \th (\a - X_t) dt + \b dW
\end{equation*}
with an initial condition, $X_0$, which may be an arbitrary distribution
independent of the Wiener Process,$W$.
We can solve this as follows:
\begin{align*}
dX =& \th (\a - X_t) dt + \b dW
\\
dX + \th X_t) dt=& \th \a dt + \b dW
\\
e^{\th t} dX + e^{\th t}\th X_t) dt=& \th e^{\th t}\a dt + \b e^{\th t} dW
\\
Xe^{\th t} - X_0 = & \int \th e^{\th t}\a dt +  \int \b e^{\th t} dW
\\
X_t =& e^{-\th t} X_0 + \a(1-e^{-\th t}) +  \frac{\b e^{-\th t}}{\sqrt{2\th}}
W(e^{2\th t}-1)
\end{align*}
which means that $X$ forgets its initial conditions exponentially fast and
converges to a normal random variable with mean $\a$ and variance
$\tfrac{\b^2}{2\th}$
\end{ex}

At the end of this sub-section, we state Ito's Lemma with multiple Wiener
processes and then for a multidimensional state.

\begin{thm}[Ito Lemma for $dW \in  \R^{m}$] Suppose $X$ satisfies the stochastic
differential $dX_t = F dt + G dW$, where $ F \in \R, G \in \R^{1\times m}$ and
$dW = \left(dW^{(i)}\right) \in  \R^{m}$ is a vector of independent Wiener
Processes. Take $u(x,t) \in C^{2,1}[ \R \times [0,T]]$. Let $D :=\left[{\sum_k
G_{k} G_{k} } \right] \in \R$

Set $Y(t) = u(X,t)$
\\
then
$$
dY =  \left( \di_t u +  \di_{x} u \cdot F + 
  \di^2_{x} u \cdot D \right)
\intd{t} +
 \left(  \di_{x} u  \cdot  G \right) dW  
$$
\end{thm}

\begin{thm}[Ito Lemma for $X \in R^n$]
Suppose $X$ satisfies the stochastic differential $dX_t = F dt + G dW$, where $
F \in \R^n, G \in \R^{n\times m}$ and $dW = \left(dW^{(i)}\right) \in  \R^{ 
m}$ is a vector of independent Wiener Processes. Take $u(x,t) \in C^{2,1}[ \R^n
\times [0,T]]$. Let $D_{ij} := \tfrac{1}{2} \left[{\sum_k G_{ik} G_{jk} }
\right]_{ij} \in \R^{n\times n} $

Set $Y(t) = u(X,t)$
\\
then
$$
dY =  \left( \di_t u + \sum_i \di_{x_i} u \cdot f_i + 
\sum_{i,j} \di^2_{x_i x_j} u \cdot D_{ij} \right)
\intd{t} +
 \left(  \sum_{ij} \di_{x_i} u  \cdot G_{ij} dW^{(j)} 
\right)$$
\end{thm}

\subsection{Fokker-Planck and Kolmogorov's Backward Equations}
The Fokker-Planck equation and Kolmogorov's Backward equation describe the
forward (resp. backward)  evolution of the probability density of $X_t$

For the rest of this section we will work in $ \R^n$, i.e. $X$ will be an $n-$
dimensional vector that satisfies the stochastic differential 
\begin{equation}
dX_t = F(X,t) dt + G(X,t) dW,
\label{eq:generic_Ito_SDE_Rn}
\end{equation}
$F \in \R^n, G \in \R^{n \times m}$,
$dW = (dW^{(i)}) \in  \R^{  m}$ is an $m$-dimensional Brownian motion and we
write $D_{ij} :=
\tfrac{1}{2} 
\left[{\sum_k G_{ik} G_{jk} } \right]_{ij} \in \R^{n\times n} $.

% Let $$\F(x,t| y,s) = \Prob[X_t \preceq x| X_s = y] $$ be the forward transition
% distribution and

Let $$ \f(x,t| y,s) \intd{x} =  \Prob[X_t \in \intd{x} | X_s = y] $$ be the
transition probability density. Then $\f$ satisfies:
\begin{equation}
\di_t \f= -\sum_i \di_{x_i} \left[ F_i(x,t) \f(x,t) \right] 
+ 
\sum_{i,j}  \di^2_{x_i x_j} \left[ D_{ij}(x,t) \f(x,t) \right]
\label{eq:FP_pde}
\end{equation}
This is called the \emph{Fokker-Planck} or \emph{Forward Kolmogorov} equation.
It can also be seen as a continuity euquation or a cosevation of probability
equation. To this end define the probability current, $\p \in \R^n$ as:
$$
\phi_i =   F_i(x,t) \f(x,t) 
+ 
\sum_{j}  \di_{x_j} \left[ D_{ij}(x,t) \f(x,t)\right]
$$
then the Fokker-Planck equation, \cref{eq:FP_pde}, can be written as:
$$
\di_t \f= -\grad \cdot \phi
$$
which just says that the change in probability is the
difference between the flow in and the flow out.

Initial conditions for $\f$ are given by the distribution of $X_s$, if $s=0$
then this is the initial distribution of $X$. Boundary conditions (BCs) depend
on the domain of $X$ and what happens to $X$ once it hits its boundary. If the domain
is all of $\R^n$ then we only insist that $\lim_{|x| \ra \infty} \f = 0$. If the
domain has boundaries, then there are two common scenarios which we will also
encounter in the sequel:
\begin{enumerate}  
  \item absorbing BCs
  \item reflecting BCs
\end{enumerate}
At an absorbing boundary, the particle $X$ is removed and $\f=0$ there.
In this situation, we will not have conservation of probability and the integral
of $\f$ over $X$'s domain will monotonically decrease.

At a reflecting boundary, the particle $X$ bounces back into its domain and we
will have that the probability current $\p\cdot n = 0$ where $n$ is the outward
normal at the reflecting boundary.

We now consider $\f$ as functions of $y,s$, holding $x,t$ fixed, then
\begin{equation}
-\di_s \f= \sum_i  \left[F_i(y,s) \cdot \di_{y_i}\f(|y,s) \right] 
+ 
\sum_{i,j}   \left[ D_{ij}(x,t) \di^2_{y_i y_j}\f(|y,s) \right]
\label{eq:FP_backward_pde}
\end{equation} 
This is the \emph{Backward Kolmogorov} equation for $\f$. Note the minus sign
in front $\di_s \f$. A mnemonic for the signs of the Backward vs. Forward
equation is that as $t$ increases $\f$ diffuses and so $\di_t$ and $\di^2_{x}$
have the same sign, but as $s$ increases $\f$ anti-diffuses and they have
opposite signs.

The differential operator on the right-hand side of
\cref{eq:FP_backward_pde} occurs often in the study of SDEs and has its own
name.
\begin{defn}[Generator of an SDE] the Generator $A$ of $X$ is defined by:
$$
A[\psi(x)] = \lim_{t \searrow 0^+} \frac{\Exp[\psi(X_t)]  - \psi(x)}{t} ;  \quad X_0 =
x \in \R^n
$$
\end{defn} 
\begin{lemma} For an Ito SDE as in \cref{eq:generic_Ito_SDE_Rn} 
$$
A[\psi(x)] = \sum_i F_i(x,t) \cdot \di_{x_i} \psi + \sum_{i,j} D_{ij}(x,t)
\di^2_{x_i x_j}\psi $$
\end{lemma}

\subsection{Stopping Times}
\begin{defn}
Let $\Fil(t)$ be some filtration, a random variable $\t$ is called a
\emph{stopping time} if 
$$
\{\o : \t(\o) \leq t\} \in \Fil(t) \, \forall t  
$$
\end{defn}
The colloquial way of describing stopping times is at any time we know
whether $\t$ has occurred or not. For a counterexample, the time that a Wiener
Process achieves its maximum over some interval is not a stopping time, since at any
given time, we do not know if the maximum has occurred or not. The most common
example of a stopping time is the first hitting-time, which is the first time
$X_t$ leaves a set. 
\begin{thm}[First-Hitting time] Let $E \subset \R^n, E \neq
\phi$ be open or closed

then $$\t := \inf\{ t \geq 0 | X_t \in E\}$$ is a stopping time.
\end{thm}

The reason stopping times are very useful is that all the facts so far quoted
for Ito calculus using integrals $\int_0^T \intd{W}$ remain true if $T$ is
replaced by a stopping time $\t$.

Also it allows us to link SDE's and PDEs using the generator, $A$, of the
diffusion: 
\begin{thm}[Dynkin's formula] Given $\t$ a stopping time, $\Exp[\t] < \infty$.
Then:
$$
\Exp[u(X_\t, \t)] =
u(x, 0) + \Exp\left[\int_0^\t \di_t u + A[u] \intd{s}
\right]
 $$
\end{thm}
This provides a link between PDEs and stochastic processes and allows us to go
back and forth in that we can find probabilistic results by
solving a PDE or we can approximate a PDE by simulating a stochastic process
and averaging.

\begin{ex}[\cite{Evansb} pg 99 - Expected hitting time to a boundary]
\label{ex:mean_hitting_time}
 Let $\O
\subset \R^n$ be a bounded open set with smooth boundary $\di \O$ then it is a
basic fact from PDEs theory that
\begin{equation}
\begin{cases}
-\frac{1}{2} \grad^2 u = 1  & \text{over } \O
\\
u =  0 &\text{on } \di \O
\end{cases}
\end{equation}
has a unique $C^\infty(\O)$ solution.

Let $X = W_t + x$ for any $x \in \O$ and define
 $$\t_x := \text{first time } X \text{ hits } \di \O$$
then the generator of $X$ is $A[\psi] = -\grad^2(\psi)/2$ and we will
have:

\begin{align*}
\Exp[u(X_\t)] - \Exp[ u(x_0)] =& \Exp \left[ -\frac{1}{2}\int_0^\t \grad^2
u\right]
\\
=& \Exp \left[ - \int_0^\t 1
\right]
\\
=-&\Exp [\t]
\end{align*}
Finally, invoke $u$'s BCs, $u|_{\di E} = 0$, and $X$'s ICs, $X_0=0 = x$ to
conclude that $$ u(x) = \Exp [\t]$$
The solution to the PDE evaluated at $x$ is the expected exit time from
$E$ for an $X_t$  starting at $x$.
\end{ex}
% Similarly we can calculate more moments of the hitting time.
% \begin{ex}[Mean squared hitting time to a boundary] Let $E
% \label{ex:mean_squared_hitting_time}
%  \subset \R$ be a bounded open set and let $u$ solve:
%  \begin{equation}
%  \begin{cases}
%  (\a - \tfrac{x}{\tc}) \di_x u + \frac{\b^2}{2} \di_x^2 u = -\frac t 2  &
%  \text{over } E
%  \\
%  u =  0 &\text{on } \di E
%  \end{cases}
%  \end{equation}
% 
% Then $- t / 2 = A[u]$ for $A$, the generator, corresponding to $dX = (\a - x /
% \tc) dt + \b dW$.
% Define $$\t_x := \text{first time } X \text{ leaves }  E$$
% then
%  \begin{align*}
%  \Exp[u(X_\t)] - \Exp[ u(X_0)] =& 
%  \Exp \left[  \int_0^\t   (\a - \tfrac{x}{\tc}) \di_x u + \frac{\b^2}{2} \di_x^2
%  u \right]
%  \\
%  =& \Exp \left[ - \int_0^\t -\frac t 2 
%  \right]
%  \\
%  =-&\Exp [\t^2]
%  \end{align*}
% And we can conclude that 
% $$ u(x) = \Exp [\t^2]$$
% The solution to the PDE evaluated at $x$ is the expected squared exit time from
% $E$ for an $X_t$  starting at $x$.
% \end{ex}

\subsection{Feynman-Kac formula}
The Feynman-Kac formula is a generalization of the above idea. We state it in
1-D. 
\begin{thm}[Feynman-Kac Formula] Consider the (linear) PDE, with terminal
conditions:
\begin{equation}
\begin{cases}
\di_t u(x,t) + f \di_x u + \frac {g^2}2 \di_x^2 u - V u + q 
&= 0
\\
u(x,\tf) &= \psi(x)
\end{cases}
\label{eq:PDE_feynman_kac}
\end{equation}
The dynamics can also be written:
$$
\di_t u(x,t) + A[u] - V u + q = 0
$$
for the generator $A = f \cdot \di_x + g^2/2 \cdot \di_x^2$ corresponding to the
SDE: $$
dX_t = f(X,t) dt + g(X,t) dW
$$
Then for $t< \tf$
\begin{align*}
u(x,t) =& \Exp\bigg[
\int_t^\tf q(x,s)\cdot \exp \left(-\int_t^s V(X_\t) \intd{\t} \right)\intd{s} \\
&\quad + \, \psi(X_t) \cdot \exp\left(-\int_t^\tf V(X_\t,\t) \intd{\t} \right)  
\quad \Big| \,X_t = x \bigg]
\end{align*}
\begin{proof}[Proof]
This can be seen as a more elaborate extension of example
\ref{ex:mean_hitting_time}. We set: 
$$
Y_r = u(x,t)+ 
\int_t^r q(x,s)\cdot\exp \left(-\int_t^s V(X_\t) \intd{\t} \right)\intd{s} +
u(X_r,r)\cdot\exp\left(-\int_t^r V(X_\t,\t) \intd{\t} \right)   
$$
then 
\begin{align*}
dY =& q(x,r)\cdot\exp \left(-\int_t^r V(X_\t) \intd{\t} \right)
\\
& - u(X_r,r) \cdot V(X_r, r)  \exp\left(-\int_t^r V(X_\t,\t)
\\
& + du \cdot \exp\left(-\int_t^r V(X_\t,\t) \intd{\t} \right) 
\intd{\t}\right)
\end{align*}
Now we recall Ito's lemma for $du(X,t)$, and replace it in the last line:
\begin{align*}
dY_r =& q(x,r)\cdot\exp \left(-\int_t^r V(X_\t) \intd{\t} \right) \intd{r}
\\
& - u(X_r,r) \cdot V(X_r, r)  \exp\left(-\int_t^r V(X_\t,\t)
\intd{\t}\right)s
\\
& + \Big( \di_t u + A[u] \Big) \cdot \exp\left(-\int_t^r V(X_\t,\t) \intd{\t}
\right) \intd{r}
\\
& + \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot \exp\left(-\int_t^r V(X_\t,\t)
\intd{\t} \right) \intd{W}
\end{align*}
Now we can group the terms in front of the exponential term on the first
three lines to form:
\begin{align*}
dY_r =& \Big(\di_t u + A[u] - u(X_r,r) V(X_r, r) + q(X_r,r) \Big) \cdot\exp
\left(-\int_t^r V(X_\t) \intd{\t} \right)
\intd{r}
\\
& + \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot \exp\left(-\int_t^r V(X_\t,\t)
\intd{\t} \right) \intd{W}
\end{align*}
Since, $u$ satisfies the PDE, \cref{eq:PDE_feynman_kac}, the term on the first
line is zero so $dY$ is just
$$
dY_r = \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot \exp\left(-\int_t^r V(X_\t,\t)
\intd{\t} \right) \intd{W}
$$
integrating we have: 
$$
Y_\tf - Y_t = \int^\tf_t \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot
\exp\left(-\int_t^r V(X_\t,\t)
\intd{\t} \right) \intd{W}
$$
Now we need conditions such that the expectation of the stochastic integral is
zero. For example, it suffices that $V$ is bounded, and often in applications
$V$ is a positive rate.
Then, we   take conditional expectations, conditional on $X_t=x$:
\begin{equation*}
\begin{gathered}
\Exp\Big[Y_\tf | X_t = x\Big] - \Exp\Big[Y_t | X_t = x\Big]  
= 
\\
\Exp\Bigg[\int^\tf_t \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot
\exp\left(-\int_t^r V(X_\t,\t)
\intd{\t} \right) \intd{W} \,\mid\, X_t = x\Bigg] 
\\ = 0
\end{gathered}
\end{equation*}
Unravelling the definition of $Y_t$, we get
\begin{align*}
\Exp\bigg[
\int_t^\tf q(x,s)\cdot \exp \left(-\int_t^s V(X_\t) \intd{\t} \right)\intd{s} 
 \quad + &
\\
 \, \psi(X_t) \cdot \exp\left(-\int_t^\tf V(X_\t,\t) \intd{\t} \right)  
\, \Big| \,X_t = x \bigg] &=\, u(x,t)  
\end{align*}

\end{proof}
A more detailed proof can be found in \cite{Oksendal2007} Thm. 8.2.1
\end{thm}

We will not use the Feynman-Kac formula in what follows, but
we have included it here, because there are some very interesting developments in stochastic optimal
control, which rely on solving a many-dimensional PDE using stochastic
simulation, the link being made via the Feynman-Kac formula - see
\cite{Kappen2005},\cite{Palmer2011}.

% \subsection{Numerical Simulation of SDEs}

\section{Deterministic Optimal Control in Finite Dimension}
Optimal Control theory has three main components - a state, $x$, a control, $u$
and an objective $J$ which is a functional of $x,u$ and which we try to either
minimize or maximize. Here, we will be minimizing.

We introduce the theory for a finite dimensional state, $x(t) \in \R^{n_x}$, but
we will not address state constraints. Our control will also be
finite-dimensional $u(t) \in \R^{n_u}$ and it will be allowed to take values in
some closed domain, $\Udomain$, of allowed controls, e.g.\ $\Udomain = \{ u_t
\in \R^{n_u} \, \st |u| \leq \Umax \}$ for a control constrained to a sphere.
The objective $J$ will be given by a time-integral-plus-terminal cost:
\begin{equation}
J[u] = \int_0^\tf L(x_s, u_s) \intd{s} + M(x_\tf)
\label{eq:generic_objective_function_deterministic}
\end{equation}
Where $\tf$ could be either a variable or fixed.

The dynamics of $x$ are given by some controlled ordinary differential
equation (ODE) given some initial conditions:
\begin{equation}
\dot{x} = f(x,t); \quad x(0) = x_0
\label{eq:generic_dynamics_deterministic}
\end{equation}

Then we seek the optimal control $u^*$, such that:
\begin{equation}
u^* = \argmin_{u \in \Udomain} J[u]  
\label{eq:generic_objective_argmin}
\end{equation}

Note that $x$ may also have final conditions, which we will deal with later,
and/or path constraints. If there are path-constraints, we talk about a
state-constrained problem. However, we will not face path constraints in the
sequel and so will not say more about how to handle them.

There are essentially three methods for solving equation
\cref{eq:generic_objective_function_deterministic}. Two, Pontryagin's Minimum
Principle and Dynamic Programing, are analytic in nature providing a set of
equations that characterize the minimum and then solving these equations,
numerically if necessary. The third, the direct method, simply discretizes the
problem and solves the resulting nonlinear programing (NLP) problem via standard
NLP techniques. We will describe both Pontryagin's Minimum Principle and Dynamic
Programing below and indeed we will use both later on.

\subsection{Pontryagin's Maximum Principle}
Pontryagin's Principle is derived from on a variational argument, similar to the
Euler-Lagrange equations from the Calculus of Variations, and it characterizes
the optimal trajectory/control pair. It can be thought of as a generalization of
the zero-tangent rule (Fermat's Rule) for finding optima in single-variable
calculus and like both the Euler-Lagrange equations and Fermat's Rule it
provides necessary, but not sufficient conditions for a minimum.

Pontryagin's Principle also holds a similarity to Hamiltonian Mechanics, which
should not be surprising, since Hamiltonian Mechanics is founded on
the minimization of action.

There are several versions of Pontryagin's Principle, depending on whether $x$ has
terminal conditions or not, whether the final time, $\tf$ is specified or a
variable in itself and most importantly whether the state has constraints or
not. We will state the versions that are relevant for our needs, namely, $x$ has
no constraints and $\tf$ is specified.

We first introduce the Pontryagin Hamiltonian, $\H$, also called the control
theory Hamiltonian and an adjoint state, $p \in \R^{n_x}$:
\begin{defn} [$\H$] The control theory Hamiltonian is the function:
$$
\H(x,p, u) := f(x,u) \cdot p + L(x,u) \quad (x,p \in R^{n_x}, u \in \Udomain)
$$ 
\end{defn}

\begin{thm}[Pontryagin's Fixed-Time, Free-End-Point] Assume $u^*(t)$ is
optimal for
\cref{eq:generic_objective_function_deterministic,eq:generic_dynamics_deterministic}
and that $x^*$ is the corresponding trajectory.

Then there exists a function $p^*:[0,\tf]:\ra \R^{n_x}$ st.
\begin{equation}
\begin{cases}
\dot{x}^*(t) &=  \grad_p \H(x^*,p^*, u^*) = f(x^*, u^*)
\\
x(0) &= x_0
\end{cases}
\label{eq:pontryagin_state_ode}
\end{equation}
\begin{equation}
\begin{cases}
\dot{p}^*(t) &= -\grad_x \H(x^*,p^*, u^*) = -\grad_xL - p \grad_x f
\\
p^*(\tf) &= \grad_x M(x)
\end{cases}
\label{eq:pontryagin_adjoint_ode} 
\end{equation}
and
\begin{equation}
\H(x^*(t), p^*(t), u^*(t)) = \min_{u(t) \in \Udomain(t)}  \H(x^*(t), p^*(t),
u(t))
\label{eq:pontryaginH_optimality_condition} 
\end{equation}
\end{thm}
We will not give a proof of this theorem (see A.2 in \cite{Evansb}),
but we do provide a heuristic derivation in
\cref{sec:Pontryagin_heuristic_derivation}, which is very helpful in deriving
the necessary conditions for PDE dynamics and it explains the otherwise
mysterious origins of $\H$ and $p$.

At each time, $t$, we obtain the optimal control $u(t)$, by solving
\cref{eq:pontryaginH_optimality_condition}. This can be easy, hard or useless.
In the easy case, the minimum is analytically obvious, e.g.\ if the Hamiltonian
is quadratic in $u$. If it is linear in $u$ then the minimum will occur on the
boundaries of the control and we will talk about a bang-bang control.
Conversely, solving the minimum of $\H$ with respect to $u$ may be impossible,
since $x,p$ are unknown. In this case we need to have an iterative scheme which
alternately solves
\cref{eq:pontryaginH_optimality_condition} 
and then \cref{eq:pontryagin_state_ode,eq:pontryagin_adjoint_ode}. Basic
numerical techniques using this idea are discussed in the appendix in
\cite{Kirk2004}. However, if we are going to be solving the problem numerically,
then the direct method might be superior, see \cite{Ross2005} for a discussion.

Finally, it is possible that \cref{eq:pontryaginH_optimality_condition} does not
depend on $u$ explicitly (it will still depend  on $u$ implicitly through $x,p$)
and so we cannot find $u$ at all by minimizing $\H$ wrt.\ $u$. This happens, for
example, if the cost, $L$, does not depend on $u$, the dynamics, $f$, are linear
in $u$ and $p=0$ over an interval of positive length. Such an optimal control
problem is called singular and it is discussed briefly in \cite{Kirk2004}, and
the theory can still be useful in finding a solution, but many complications can
and do arise. We face a singular control problem in our own work later, however
instead of tackling it head on, we merely regularize the problem - thus we will
not say much more on Singular Control Problems.

The version of Pontryagin's Principle for Fixed End Point, meaning $x$ has some
terminal conditions, is very similar to the one with free end-point. Suppose 
\begin{equation}
\dot{x} = f(x,t); 
\quad
\begin{cases}
x(0) = x_0
\\
x(\tf) = x_f
\end{cases}
\label{eq:generic_dynamics_deterministic_fixed_end_point}
\end{equation}

\begin{thm}[Pontryagin's Fixed-Time, Fixed-End-Point] Assume $u^*(t)$ is
optimal for
\cref{eq:generic_objective_function_deterministic,eq:generic_dynamics_deterministic_fixed_end_point}
and that $x^*$ is the corresponding trajectory.

Then there exists a function $p^*:[0,\tf]:\ra \R^{n_x}$ st.
\begin{equation}
\begin{cases}
\dot{x}^*(t) &=  \grad_p \H(x^*,p^*, u^*) = f(x^*, u^*)
\\
x(0) &= x_0
\\
x(\tf) &= x_f
\end{cases}
\label{eq:pontryagin_state_ode}
\end{equation}
\begin{equation}
\dot{p}^*(t) &= -\grad_x \H(x^*,p^*, u^*) = -\grad_xL - p \grad_x f
\label{eq:pontryagin_adjoint_ode} 
\end{equation}
and
\begin{equation}
\H(x^*(t), p^*(t), u^*(t)) = \min_{u(t) \in \Udomain(t)}  \H(x^*(t), p^*(t),
u(t))
\label{eq:pontryaginH_optimality_condition} 
\end{equation}
\end{thm}
I.e. $p$ loses its final conditions because $x$ has them. This relates to the
variational nature of Pontryagin's Principle and the idea that $p$ represents
the sensitivity of the objective wrt. small changes in $x$. Thus, if $x$ is
fixed, there are no variations and $p$ can be anything.
% To illustrate Pontryagin's Principle we will solve the moon
% lander problem (Ex. 4.4.4 \cite{Evansb}), with the twist of holding the final
% time fixed. 
% \begin{ex}
% 
% \end{ex}

\subsection{Dynamic Programing - the Hamilton-Jacobi-Bellman equation} 
Dynamic Programing uses backwards recursion to tabulate the optimal control
starting from the terminal time. The basic object in dynamic programing is the value function, $v$. In order to
introduce it, we first extend our definition for the objective, $J$, to include later
times, with different initial conditions.

The running cost-to-go corresponding to
\cref{eq:generic_objective_function_deterministic} is:
\begin{equation}
J[u; x, t] = \int_t^\tf L(x_s, u_s) \intd{s} + M(x_\tf), \quad x_t = x
\label{eq:generic_cost_to_go_deterministic} 
\end{equation}
So that $J[u; x_0, 0] = J[u]$.

The dynamic programming value function, $v$ is then defined by:
$$
v(x,t) = \inf_{u \in \Udomain} J[u; x,t] 
$$
it is also called the optimal cost-to-go.

We immediately note the terminal conditions on $v$:
\begin{equation}
v(x,\tf ) = M(x)
\end{equation}

The Dynamic Programing approach is then to solve for $v$ backwards and in the
process infer the control $u$.
\begin{thm}[Hamilton-Jacobi-Bellman Equation] Assume that $v$ is $C^{1,1}$ in
$x,t$. Then $v$ solves:
\begin{equation}
\begin{cases}
-\di_t v(x,t) &= \max_{u(t) \in \Udomain(t)} \left\{ f(x,t) \cdot \di_x v +
L(x,u) \right\}
\\
v(x,\tf) &= M(x)
\end{cases}
\end{equation}
\end{thm}
We will not derive the HJB equation here, but we will derive a similar version
later for the stochastic case, when the dynamics are given by SDEs.

% \subsection{Model Predictive Control}
% Solving the HJB equations and obtaining a feedback control is usually
% impossible. Thus a practical alternative is to solve a simplified problem on a
% limited horizon and use the simplified solution now and then after stepping
% forward in time and observing the system in a new position, to solve a revised
% limited horizon problem and use the new solution.
% Let us make this more precise.

\section{Stochastic Optimal Control}
Here we extend the Dynamic Programing theory to the case where the dynamics are
governed by an SDE:
\begin{equation}
dX = F(X,t, u) dt + G(X,t) dW
\end{equation} 
i.e.\ we assume control affects only the drift term. This 
assumption is not necessary, but it corresponds to our problems.

For stochastic dynamics one can only optimize probabilistically and so
$J$ becomes:
\begin{equation}
J[u] = \Exp \left[ \int_0^\tf L(x_s, u_s) \intd{s} + M(X_\tf)
\right] 
\label{eq:generic_objective_function_stochastic}
\end{equation}
while the running cost-to-go is:
\begin{equation}
J[u; x, t] = \Exp \left[ \int_t^\tf L(X_s, u_s) \intd{s} + M(X_\tf) \right], 
\quad X_t = x
\label{eq:generic_cost_to_go_stochastic}
\end{equation}
the value function is:
\begin{equation}
v(x,t) := \inf_{u \in \Udomain} J[u; x,t]
\end{equation}

Finally, we again have that $v$ satisfies a certain PDE:
\begin{thm}[Stochastic HJB] 
\label{thm:stochastic_hjb}
\begin{equation}
\begin{cases}
-\di_t v(x,t) &=  \max_{u(t) \in \Udomain(t)} \big\{ L(x,u)  +
F(x,t,u) \cdot \di_x v
\big\} + \frac{G^2(x,t)}{2} \di_x^2v
\\
v(x,\tf) &= M(x)
\end{cases}
\end{equation}
\begin{proof}[Heuristic Derivation (from \cite{Evansb})] Suppose we are at time
$t$ and $X_t = x$.

Take a time increment $[t, t+h]$ and assume that during that time we apply a
control $u$ and subsequently we apply the the optimal control $u^*$. The
running-cost will then break down as: 
$$
J[u; x,t] = \Exp \left[ \int_t^{t+h} L(X_s, u_s) \intd{s}  + v(X_{t+h},
t+h) \right] $$

Since $v(x,t) = \inf J(u; x,t)$, we must have that:
$$
v(x,t) \leq  \Exp \left[ \int_t^{t+h} L(X_s, u_s) \intd{s}  + v(X_{t+h},
t+h) \right] $$
or, rearranging, that
\begin{align*}
0 \leq^&  \Exp \left[ \int_t^{t+h} L(X_s, u_s) \intd{s} \right]  
+ \underbrace{\Exp \left[ v(X_{t+h}, t+h) - v(x,t) \right]}_{\Exp\left[ dv
\right]}
\end{align*}
Now, $\Exp\left[ dv \right]$ can be expressed using Dynkin's Formula, as:
$$
\Exp[dv] = \int_t^{t+h} \di_t v +  F\cdot \di_x v + \frac{G^2 }{2}\cdot \di_x^2
v \intd{s} $$
Plugging that back, we get
$$
0 \leq \int_t^{t+h} L(X,u) +  \di_tv +  F\cdot \di_x v + \frac{G^2 }{2}\cdot
\di_x^2 v \intd{s} $$
Taking $h \ra 0$ we get 
$$
0 \leq  L(x,u) +  \di_tv(x,t) +  F\cdot \di_x v(x,t) + \frac{G^2 }{2}\cdot
\di_x^2 v(x,t) $$
and we conjecture that for the actual optimal control, the inequality becomes an
equality:
\begin{equation}
\begin{cases}
0 &=   L(x,u^*)+ \di_t v(x,t) +
 F(x,t,u^*)\cdot \di_x v(x,t) + \frac{G^2(x,t)}{2}\cdot \di_x^2 v(x,t)
\\
u^*(t) &= \argmax_{u \in \Udomain(t)}  
\big\{L(x,u) + F(x,t,u)\cdot \di_x v \big\}
\end{cases}
\end{equation}
\end{proof}
\end{thm}
% A rigorous proof can be found in \cite{Krylov2008}, sec???ch???

\section{Deterministic Optimal Control in Infinite Dimensions}
Now we discuss Optimal Control when the dynamics are deterministic but of
infinite-dimension, i.e. the state evolution is a PDE instead of an ODE.  The
reason we discuss this, is that optimization of ordinary SDEs in an open-loop
control is actually a deterministic optimal control problem in infinite
dimensions.

Why?

Since the objective is probabilistic and the probability density evolution is
given by the (deterministic) Fokker-Planck PDE, we need to optimize a
deterministic system whose dynamics are given by the Fokker-Planck equation.

There is no all-encompassing generalization of Pontryagin's Principle to
infinite dimensions, but a Maximum Principle can be derived for certain systems.
To illustrate the idea we will discuss a simple case as in \cite{Palmer2011}.
Our controlled SDE will be given by: 
$$ dX_t = F(X,t, u) dt + G(X,t) dW, $$
The corresponding probability density is given by: $$ \di_t \f=
-\sum_i \di_{x_i} \left[ F_i(x,t,u) \f(x,t) \right] + \sum_{i,j}  \di^2_{x_i
x_j} \left[ D_{ij}(x,t) \f(x,t) \right] $$ which we will write as $$ \dot{\f} =
\L_{u} [\f] $$ where $\L_{u}$ is the differential operator corresponding to the
RHS of the Fokker-Planck equation parametrized by the control $u$. For now we
will assume that there are no BCs, and the domain of $X$ is all of $\R^n$.

A natural inner-product, $ \langle.,. \rangle$ on functions in $L^2[\R^n]$ is
given by: $$
 \langle\psi,\phi  \rangle = \int_{\R^n} \psi(x) \phi(x) \intd{x}
$$
Our objective will be to minimize:
\begin{equation}
J[u] = \int_0^\tf  \langle L(\cdot,s,u), \f_u(s) \rangle \intd{s} +  
\langle M(\cdot),\f(\tf) \rangle
\label{eq:generic_objective_function_probabilistic}  
\end{equation}.

\Cref{eq:generic_objective_function_probabilistic} is a very natural cost
function in the stochastic dynamics case, because it can also be written as: 
$$
J[u] = \Exp \left[ \int_0^\tf  L(X_s, s,u) \intd{s} + M(X_\tf) \right] 
$$ which
is just \cref{eq:generic_objective_function_stochastic}. We use
\cref{eq:generic_objective_function_stochastic} if we have some feedback and
want to use dynamic programming. We use
\cref{eq:generic_objective_function_probabilistic} if we have no observations
and need to run an open-loop control.

Since, \cref{eq:generic_objective_function_probabilistic} is the objective
function for deterministic dynamics, we can use Pontryagin's Principle adapted
to the case of infinite dimensional dynamics (PDEs).

Introduce an adjoint co-state, $\l \in \R^n$ and let the Hamiltonian, $\H$ be:
$$
\H(\f, \l, u)= \langle \phi, L + \Lstar[\l] \rangle$$
where $\Lstar$ is the adjoint operator to $\L$ defined via:
$$
\langle \L [\phi], \l \rangle =\langle \phi, \Lstar[\l] \rangle
$$

Let $\l$ evolve (backwards) according to:
\begin{equation*}
\begin{cases}
-\di_t \l =& \Lstar[\l] + L
\\
\l(\tf) =& M(x)  
\end{cases}
\end{equation*}
then the optimal control, $u$, is given via
$$
u = \argmin_u {\H(\f, \l, u)}
$$

An important fact, useful for computations, is that the total derivative of the
objective with respect to the control is given by the partial derivative of the
Hamiltonian with respect to the control:
\begin{equation}
\grad_u \big[J[u] \big] = \di_u \H(\f, \l, u)
\label{eq:objective_gradient_wrt_control}
\end{equation}

So for a given $u$, we can compute $\f,\l$, calculate the objective gradient
using \cref{eq:objective_gradient_wrt_control} and then iterate in the direction
of descent (if we are minimizing).

All that remains is to compute $\Lstar$. It turns out that if we do not have any
BCs, then
$$
\Lstar[\l] = \sum_i F_i(x,t) \cdot \di_{x_i} \l + \sum_{i,j} D_{ij}(x,t)
\di^2_{x_i x_j}\l 
$$ 
which is just the generator of $X$!

We will see later, that things become more complicated when we need to
consider a PDE's BCs.

\section{Closed-loop vs.\ Open-Loop control}
At this point we stop to briefly discuss the difference between
open-loop control and closed-loop control.

By closed-loop control, we mean that at any time we have observations of the
system state, $x$ and its value can influence the control. In open loop control,
we decide on the value of $u(t)$ a priori and apply it independently of the
value of $x(t)$, based only on the knowledge of the system dynamics and the
initial conditions.

If the system dynamics are deterministic there is no added value in closed-loop
control, since the state evolution is perfectly predictable and one can infer
where $x$ will be based only on the dynamics and the applied control. For
stochastic dynamics this is no longer true and there is something to be gained
by observing the state, since different controls can be applied depending on the
different realizations of the stochastic process driving the dynamics.
Naturally, if one has observations of the system, one would prefer to use this
and apply a closed-loop control. Thus, an open-loop control would be used
only if one has no reliable knowledge of the system state.

\section{A Stochastic Optimal Control Problem from Mathematical Neuroscience}
Neurons relay information by means of voltage spikes - sudden sharp increases in
voltage. Although many details remain unclear, the information content is
thought to be contained in the length of the time-interval between these spikes.
In the simplest case, this can be thought of as a rate  -the average number of
spikes per time interval, but more complicated coding schemes are hypothesized to
exist.

Many experiments allow for manipulating an individual neural cell. A natural
goal then is to make a cell produce a given spike train. This may arise, for
example, in brain-machine interfaces or in artificial prosthetics.

We now consider this problem in the framework of Stochastic Optimal Control and
give both a feedback solution, when the cell voltage is explicitly observable as
well as an open-loop solution when only the occurrence of a spike is
robustly observable.

\subsection{Problem Formulation}
The most basic representative model for a neuron is the noisy
leaky-integrate-and-fire model:
\begin{equation}
\begin{gathered}
dX_s = (\a(t) - \frac{X_s}{\tc} ) \intd{s} + \b \intd{W_s},
\\
X(0) = .0,
\\
X(\ts) = \xth \implies  
\begin{cases}
X(\ts^+) = .0 &  
\end{cases}
\end{gathered}
\label{eq:X_evolution_uo}
\end{equation}
In this $X$ represents the membrane electric potential which tries to revert to
$0$ with a time-const of $\tc$, $dW$ is a Brownian motion increment with
intensity $\b$ and $\a(t)$ is our external control. Having last spiked at
time $0$, the potential hits $\xth$ at time $\ts$, we record a spike
of duration $\ts$, the potential resets to $0$ and we start all over again.

Suppose we want to ensure that spikes occur exactly at the predetermined
times $\tns$. Then given the renewal nature of our spiking model, we can work
sequentially, focusing only on the first of these intervals, of duration $t_1 =
\T$, and then ask: how do we choose $\a(t)$ st. $\ts = \T$. There are several
ways to specify what we mean by 'make $\ts = \T$', all of which result in optimal
control problems. A natural one is a least squares solution:
\begin{equation}
\a(t) = \argmin_{\a(t)} \{\, \Exp[(\ts- \T)^2] \,\}
\label{eq:OC_LS_variance}   
\end{equation}

It is also natural to assume that our control has certain limitations. The
simplest and most common are simple bounded constraints:
\begin{equation}
\a(t) \in [\amin, \amax]
\label{eq:bound_constraints_alpha}
\end{equation}

% \subsection{Optimal Control Formulation}
In addition to \cref{eq:OC_LS_variance}, our objective will also contain a
running cost based on the control. This regularizes the problem eliminating the
subtleties of singular-control situations and it might be argued appropritate
to avoid excessive control.

So we will seek an optimal control, $\a$ that solves:
\begin{equation}
\a(t) = \argmin_{\a(t)} \Big\{ \,
\Exp\left[
\e \int_0^\ts  \a^2 \intd{s}
+  
(\ts - \T \big)^2 \right]  \, \Big\}
\label{eq:OC_LS_variance_energy}   
\end{equation}
where $\e$ measure how much weight we put on minimizing the energy expenditure
and if $\e \rightarrow 0$, then we do not care at all.

Experimentally it is often difficult to observe precisely the exact voltage
trajectory, but the occurrence of a spike is much more easily observable. As
such a closed-loop approach is not implementable since the state is unknown. In
the sequel we will perform three calculations. First we will derive the HJB
equation for a closed-loop control if the system was observable, and use this as
a benchmark for the best possible performance of an open-loop control. Then we
will calculate the most simple-minded open-loop control, which is just the
deterministic control of the dynamics when assuming there is no noise, $\b=0$.
THen we will calculate the optimal open-loop control by optimizing the
transition density. And finally we will compare the performance of all three
controls.

\section{Closed-Loop Solution - Dynamic Programing}
Given $t, X_t = x$, let $\t$ be the (unknown) remaining time to spike. For
example, given $t=0, X=.0$, $\t = \ts$, the entire duration of the inter-spike interval.
Starting from a general $t, x$, our remaining-cost objective, $J[\a; x,t]$,
will be:
\begin{equation}
J[\a, x,t]  = \Big\{ \,
\Exp\left[
\e \int_t^\ts  \a^2 \intd{s}
+
(\t - (\T-t) \big)^2 |x,t
\right] \, \Big\}
\label{eq:OC_LS_mean_objective}   
\end{equation}
i.e.\ if time $t$ has elapsed without a spike, we now want to minimize
the difference between $\t$ and ($T-t$), given the current state $x$. For now we
assume that we know $x$, later we will treat the problem when that is not so.

We are now in a position to set up the HJB equation for this problem. First
introduce the value function, $v(x,t)$ as:
\begin{multline}
v(x,t) = \argmin_{\a(s \geq t)} \{J[\a, x,t] \} =
\\ 
\argmin_{\a(s \geq t)} 
\Big\{ \,
\Exp\left[
\e \int_t^\ts  \a^2 \intd{s}
+
(\t - (\T-t) \big)^2 |x,t
\right] \, \Big\}
\label{eq:OC_LS_mean_value}   
\end{multline}

The corresponding HJB PDE for $v$ is:
\begin{equation}
\di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + \min_{\a \in [\amin, \amax]}\big\{
\e \a^2 + (\a-\tfrac{x}{\tc}) \di_x v \big\} = 0
\label{eq:OC_LS_mean_HJB}
\end{equation}

We have been canny enough to set up a quadratic optimization problem in the
control, which means that the optimal control is given analytically via:
\begin{equation}
\a(x,t) = \frac{-\di_x v}{2\e} 
\label{eq:OC_LS_variance_energy_quadratic_control}
\end{equation}
However we need to recall the bounds on the control, $\amin < \a < \amax$, and
thus
\begin{equation}
\astar(x,t) = \min \left(\amax, \max\left(\amin, -\frac{\di_x v}{2\e}\right)
\right)
\label{eq:OC_LS_variance_energy_quadratic_control_bounds}
\end{equation}
With this the HJB becomes
\begin{equation}
\di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + 
\e (\astar)^2 + (\astar-\tfrac{x}{\tc}) \di_x v 
= 0
\label{eq:OC_LS_mean_HJB_bounded_control}
\end{equation}

In the sequel we will drop the superscript star on the optimal control and
just write $\a(t)$ for the solution of
\cref{eq:OC_LS_variance_energy_quadratic_control_bounds}. 
% Thus the control can be solved out of the HJB equation, which now becomes:
% \begin{equation}
% \di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + \tfrac{(\di_x v)^2}{\e}
% -\tfrac{x}{\tc}\di_x v= 0
% \label{eq:OC_LS_mean_HJB}
% \end{equation}

Now, we need to consider boundary conditions (BCs). If $X_t =
\xth$ then we have a spike now and we know that $\ts = t$. Thus:
$$
v(\xth,t) = (t-\T)^2
$$
Now for $x \ll .0$, we can assume that $v$ doesn't really depend on $x$ so say 
$$
\di_x v(\xmin, t) = .0
$$
This is legitimate if we impose artificial reflecting boundaries for the
diffusion at $\xmin$. So we need to determine for which $\xmin$ is a reflecting
boundary consistent with the dynamics. For example, we can take $\xmin$ to be
two standard deviations of the stationary distribution of the non-controlled, $\a=0$, \cref{eq:X_evolution_uo}.
% Ooops,
% we need $\di_x v \neq 0$ and here we are setting $\di_x v = 0$ in one of the BCs\ldots oh well.

Finally we come to the Terminal Conditions (TCs) for $v$. The idea is simple: if
we reach $\T$ without having spiked we will just slam on the accelerator: i.e.\
$$t>T \implies \a(t) = \amax$$ Thus: $$ v(x,T) = \Exp[\t^2|x, \amax] $$.

Thus we restate the HJB equation in its fully specified form:
\begin{equation}
\begin{gathered}
\di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + 
\e \a^2 + (\a -\tfrac{x}{\tc}) \di_x v 
= 0
\\
\a (x,t) = \min \left(\amax, \max\left(\amin, -\frac{\di_x v}{2\e}\right)
\right)
\\
\begin{cases}
v(\xth,t) = (t-\T)^2  \quad & \textrm{upper BC}
\\
\di_x v(\xmin, t)  = .0  \quad &\textrm{lower BC}
\\
v(x,T)  = \Exp[\t^2|x, \amax]  \quad& \textrm{TC}
\end{cases}
\end{gathered}
\label{eq:OC_LS_HJB_full}
\end{equation} 
We are solving $v(x,t)$ over $[\xmin, \xth] \times [0,T]$.

\subsection{Calculating the TCs} 
\label{sec:calculating_TCs}
At this point we are in a position to solve the HJB \cref{eq:OC_LS_HJB_full}.
But first we need a more concrete expression for the TCs. Let
\begin{equation}
\begin{array}{lcl}
\Tone(x_0) &=& \Exp[\t|x_0, \amax] 
\\
\Ttwo(x_0) &=& \Exp[\t^2|x_0, \amax]  
\end{array}
\end{equation}
where as before, $\t$ is the remaining time-to-spike, given $X_0 = x_0, \a(t) =
\amax$. Naturally, $\Ti$ depend on $x_0$ and are affected by the value of the
various parameters, $\a,\b,\t$.

Actually, we only  need $\Ttwo$ for the HJB equation above, but we'll see that
$\Tone$ will be needed as an auxiliary.

Indeed, analytical expressions exist for $\Ti$, \cite{Inoue1995}. However in my
experience, they are very finicky numerically, especially as they involve
infinite alternating sums which can easily overflow in numeric calculations.
Moreover, they are not particulary convenient if you need $\Ti (x_0)$ for a wide
range of $x$'s as we do here. Instead we will turn to:

\subsubsection{the Backward Kolmogorov Equation}
Here we follow Jacobs, \cite{Jacobs}, in deriving and solving a differential
equation for $\Ti(x_0)$.

Basically we are calculating the mean-time-squared to exit an interval $[\xmin,
\xth]$. Theoretically, $\xmin = -\infty$, but for our purposes we will set it to
some finite value and impose reflecting boundaries there - this is a very good
approximation for $\xmin \ll 0$, as long as $\a>0$, which is what we have.

Let $B(y,t|x) = \Prob[X_0 = y| X_t = x]$ for a fixed $x$. Then $B$ solves the
following (backward Kolmogorov) PDE:
\begin{equation}
-\di_t B = (\a - \frac y\tc)\di_yB - \frac {\b^2} 2 \di_y^2 B
\label{eq:backward_Xdensity}
\end{equation}

Let $\G(t,y)$ be the survival function for $T$ for $X_0 = y$. 
$$\G(t,y) = \Prob[T>t | X_0 = y] = \Prob[X_{s\leq t} \in [\xmin,
\xth] | X_0 = y]$$.

Because the dynamics do not depend explicitly on time, we have:
$$
\G(t,y) = \int_\xmin^\xth B(y,t|x) \intd{x}
$$
and so $\G$, being a definite integral of $B$ satisfies the same PDE:
\begin{equation}
-\di_t \G = (\a - \frac y\tc) \di_y\G - \frac {\b^2} 2 \di_y^2 \G
\label{eq:backward_SDF}
\end{equation}

With the BCs, ICs:
\begin{equation}
\begin{cases}
\G(t,\xth) = .0 & \quad (T = t \implies \Prob[T>t] = 0)
\\
\di_x \G(t,\xmin) = .0  & \quad (\textrm{due to reflecting boundary for X})
\\
\G(0, y) = 1. & \quad (T \textrm{ is definitely} > 0 \textrm{ for } x_0 \in
(\xmin, \xth) )
\end{cases}
\end{equation}

\subsubsection{A BVP for $\Ti(x_0)$}
If we solve the PDE for $\G$ over $t \in [0,\infty)$, then we can calculate
$\Ttwo(x_0)$ as:
\begin{eqnarray}
\Ttwo(x_0) &=& \int_0^\infty t^2 \cdot (-\di_t\G(t,x_0)) \intd{t}
\\
		   &=& \int_0^\infty 2t \cdot \G(t,x_0) \intd{t}
\end{eqnarray}
assuming it exists.

Since $\Ttwo$ is an integral over time, we can reduce the PDE in
\cref{eq:backward_SDF} for the distribution to a BVP for the moment(s) by
integrating over time. Here is how this is done: multiply both sides  of
\cref{eq:backward_SDF} by $2t$, and integrate with respect to time:
\begin{equation}
\int_0^\infty 2t (-\di_t \G) \intd{t}
= 
\int_0^\infty  2t \left[ (\a - \frac y\tc) \di_y\G - \frac {\b^2} 2
\di_y^2 \G\right]
\intd{t}
\end{equation}
The LHS turns out to be $2\Tone$, so 
\begin{equation}
2\Tone
=(\a - \frac y\tc)   \di_y \Ttwo 
- \frac {\b^2} 2
\di_y^2 \Ttwo
\label{eq:BVP_Ttwo}
\end{equation}
Very similarly we can derive an equation for $\Tone$, namely:
\begin{equation}
1
=(\a - \frac y\tc)   \di_y \Tone 
- \frac {\b^2} 2
\di_y^2 \Tone
\label{eq:BVP_Tone}
\end{equation}
In both cases, the BCs will be:
\begin{equation}
\begin{cases}
\Ti(\xth) = .0 & \quad (\textrm{we are already spking})
\\
\di_x \Ti(\xmin) = .0  & \quad (\textrm{due to reflecting boundary for }X)
\end{cases}
\label{eq:BVP_Ti_BCs}
\end{equation}

\Cref{eq:BVP_Tone,eq:BVP_Ttwo} are seemingly second order ODEs, but are
actually first order and can be written as:
\begin{equation}
f(y)
= (\a - \frac y\tc) S(y) 
- \frac {\b^2} 2 S'(y) ; \quad  S(\xmin) = .0
\label{eq:BVP_T_generic}
\end{equation}
and we can recover $\Ti$ from $S$ via:
$$\Ti(x_0) =   -\int_{x_0}^\xth S(y)\intd{y}$$.
It is clear that thus defined, $\Ti$ will satisfy the BCs of
\cref{eq:BVP_Ti_BCs}.

% Solving \cref{eq:BVP_T_generic} for $S$ gives: \begin{eqnarray} S(y) &=& -
% \exp\left(\smallint \tfrac UD \right) \cdot \int_\xmin^{y} \exp\left(-\int
% \tfrac UD\right) \frac{f(\xi)}D \intd{\xi}
% \\
% &=& - \exp\left( \frac{\a y - \frac{y^2}{2\tc}}D \right) \cdot \int_\xmin^{y} 
%  \exp\left(- \frac{\a \xi - \tfrac{\xi^2}{2\tc}}D \right) \frac{f(\xi)}D
% \intd{\xi} \label{eq:BVP_T_generic_soln} \end{eqnarray}
We can solve \cref{eq:BVP_T_generic} analytically, but the analytical solution
turns out to be awkward for numerical calculations, so we
are better off just calculating the numerical solutions to \cref{eq:BVP_Tone,eq:BVP_Ttwo} directly
via an ODE solver, starting from $S(\xmin) = .0$ and integrating until $\xth$.

With this we can compare $\Ttwo$ between simulations and the numerical
solutions see \cref{fig:Ti_empirical_vs_analytic}.
\begin{figure}[htp]
\begin{center}
\subfloat[]
{
\includegraphics[width=0.48\textwidth]
{/home/alex/Workspaces/Latex/OptSpike/Figs/Moments_a=10_b=1_N=128.png}
}
\subfloat[ ]
{
\includegraphics[width=0.48\textwidth]
{/home/alex/Workspaces/Latex/OptSpike/Figs/Moments_a=10_b=10_N=128.png}
}
\\
\subfloat[ ]
{
\includegraphics[width=0.48\textwidth]
{/home/alex/Workspaces/Latex/OptSpike/Figs/Moments_a=15_b=1_N=128.png}
}
\subfloat[ ]
{
\includegraphics[width=0.48\textwidth]
{/home/alex/Workspaces/Latex/OptSpike/Figs/Moments_a=15_b=10_N=128.png}
}
\caption[]{Comparison between solutions of equations
\cref{eq:BVP_Tone,eq:BVP_Ttwo} and empirical statistics from simulating the
hitting times of \cref{eq:X_evolution_uo} for four different pairs of $\a,\b$.
In the top panel we show the first moment of the remaining time-to-spike
starting from $x$, $\Tone(x)$, and in the bottom panel, we show the second
moment, $\Ttwo(x)$. We have used only $N=128$ samples to form the moment
statistics.}
\label{fig:Ti_empirical_vs_analytic}
\end{center}
\end{figure}

\subsection{Solving the HJB - the numerical method for
\cref{eq:OC_LS_mean_HJB_bounded_control}}
We now have a PDE for $v$ and an algorithm for computing all the BCs and TCs of
this PDE. It is time to discuss the numerical method for solving
\cref{eq:OC_LS_HJB_full}.
Since we are in 1-d and we have the simplest geometry, the simplest thing to do
is try a centred Finite Difference scheme, upwind wherever necessary and hope
for the best.

Things will break down if $\e$ is very small in which case the quadratic term,
$(\di_x v)^2 / \e$, will dominate and cause 'havoc' or if $\b$ is very small in
which case the smoothing effect of the second derivative will be negligible and
we will be in the zone of nonlinear conservation equations. Let us hope for the
best now and assume these two problems do not arise, for the parameters we
select.

Recall that we are solving backwards in time. Then for variables at the next
time-step (already computed) we will write $(\cdot)^+$; while variables at the
previous time step (to-be computed), will be denoted as$(\cdot)^-$.

Similarly given some node in space, we will write
$(\cdot)_-,(\cdot)_0,(\cdot)_+$ as the left node, the node itself and the right
node.

Then our finite-difference scheme in time looks like:
\begin{equation}
\frac{v^- - v^+}{\Delta t} = \tfrac{\b^2}2  .5*(\di_x^2 v^- + \di_x^2 v^+) +
(\a^+ - \tfrac{x}{\tc})  .5*(\di_x v^- + \di_x v^+) + \e (\a ^+)^2
\end{equation}

I.e.\ we have removed the non-linearity in $v$ by computing the control, $\a$ at
the next, already computed, time-step and used central-differencing
otherwise.

For the spatial derivatives, we will not upwind for now and also use central
differencing:
$$\di_x v = \frac{v_+ - v_-}{2\Delta x}$$
and as is standard:
$$\di^2_x v = \frac{v_+ - 2 v_0 + v_-}{(\Delta x)^2}$$

We will not discuss a detailed study of the convergence properties of the
numerical method, but just demonstrate its performance on a representative
parameter set in \cref{fig:HJB_attempt} we see that the method is stable for a
particular parameter set.

% \begin{figure}[ht]
% \begin{center}
%   \includegraphics[width=.7\textwidth]{Figs/KolmogorovBVPMoments_a=10_t=9_b=8.png}
%   \caption[]{TCs for alpha = 1.0;  tauchar = .9; beta = .75;}
%   \label{fig:TCs_extended_x}
% \end{center}
% \end{figure}

Let's discuss the most salient features of \cref{fig:HJB_attempt}. At the end,
when $t = \T$, the value function, $v$ is monotonely decreasing. This makes
sense, becaue the lower $x$ is the more time on average will it take us to spike
and the more delay will we have from our desired spike time. However, as we go
back in time $v$ inverts, in the sense that for $t \ll \T$, we would like to
be somewhat far from the boundary to avoid the risk of spiking early.

Now focus on the controls, \cref{fig:HJB_attempt_control}: Naturally, at the
end we have the control assume its maximum value, $\a(t>=\T) = \amax$. As we
start to go back in time however, two things start to happen, the control $\a$
decreases for $x \ll \xth$, and it becomes negative for $x \lesssim \xth$. 
That is fully consistent with the problem objective - for
$t<\T$, we want to distance $X_t$ from the threshold to avoid an early spike.
\begin{figure}[h!]
\begin{center}
\subfloat[Solution, $v$]
{
\label{fig:HJB_attempt_soln}
\includegraphics[width=0.48\textwidth]
{Figs/HJB/soln_al=-20_au=20__t=5_b=8.png}
}
\subfloat[Controls, $\a$]
{
\label{fig:HJB_attempt_control}
\includegraphics[width=0.48\textwidth]
{Figs/HJB/control_al=-20_au=20__t=5_b=8.png}
}
\caption[]{The solution and controls for the HJB PDE parametrized by:
$\tc = 0.50,  \beta = 0.75,
\amin = -2.00, \amax = 2.00, \eps=0.10,
\T = 1.80$. The $k$ index indicates the time-step, and recall that we are
running backwards in time, so the computation actually happens from right-to-left, bottom-to-top}
\label{fig:HJB_attempt}
\end{center}
\end{figure}

\section{Deterministic Control}
\label{sec:deterministic_ode_control}
Now we will compute the optimal control in the case that $\b = 0$. The reason we
do this is that since we have assumed that we have no feedback, we attempt to
devise an open-controller by controlling the non-random dynamics. The
calculation of this control turns out to be very simple:

First let us assume that we actually can get to $\xth$ from $0$ in $\T$ time
given $\amax$ and no noise.

Then 
\begin{equation}
\begin{gathered}
J = \int_0^\T \e \a^2 \intd{t}
\\
\st
\begin{cases}
\dot{x} = (\a(t) - \frac{x}{\tc})
\\
x_\T = \xth
\\
x_0 = .0
\end{cases} 
\end{gathered}
\end{equation}
The (Pontryagin) Hamiltonian becomes:
\begin{equation}
\H(x,p, \a) = \e \a^2 + p \cdot (\a - x/\tc )
\end{equation}
the adjoint is given by:
\begin{equation}
-\dot{p} = \di_x \H = - p / \tc 
\end{equation} 
and the (unconstrained) control by:
\begin{equation}
\a(t) = \frac p {2 \e}
\end{equation}
which looks very similar to the stochastic case, once you put $p = \di_x v$.

We can actually solve for the adjoint analytically: 
$$
p = p_0 \exp(t / \tc)
$$
where we can freely choose $p_0$ to later recover the TCs for $x$.
$$
\a(t) = \frac{p_0 \exp(t / \tc)}{2\e}
$$
Finally the solution for $x$ is:
\begin{align*}
\dot{x} & = \a(t) -  \frac{x}{\tc}
\\ & = \frac{p_0 \exp(t / \tc)}{2\e} - \frac{x}{\tc}
\\
e^{t/\tc}x &= \int^t e^{s/\tc}\frac{p_0 \exp(s / \tc)}{2\e} \intd{s}
\\
x &= e^{-t/\tc} \cdot \frac{p_0 }{2\e} \tc \left[  \exp(2 t / \tc) -1 \right] 
\\
&= \frac{p_0 }{\e} \tc \cdot  \sinh (t/ \tc)
\end{align*}
And so: 
$$
x(\T) = \xth \implies p_0 = \e \frac{\xth  }{ \tc \sinh (\T/ \tc) }
$$
And we have a full solution.

If we have saturation, i.e. if $\a$ hits its upper bound, then the situation is
mildly more complicated. Since $p$ is monotonically increasing, as long as $p_0
> \amin$ then $\a(t) > \amin \forall t$. So the saturation is simplified to:
\begin{equation}
\a(t) = \min \left(\frac{p_0 \exp(t / \tc)}{2\e} , \amax \right)
\label{eq:deterministic_control_bounds}
\end{equation}
In particular, $\a$ will start by following $p(t)$ and then eventually hit its
upper bound and stay there until $\T$.

The only complication then is to calculate when $\frac{p_0 \exp(t / \tc)}{2\e} =
\amax$ and then redo the calculation for $x$ with the new control in order to
obtain the new value of $p_0$. Or we can just brute-force solve for $p_0$, by a
simple shooting method. Using the latter approach, a solution-control pair for
the parameter set we will use later is shown in \cref{fig:deterministic_soln_controls}. 
%\usepackage{graphics} is needed for \includegraphics
\begin{figure}[htp]
\begin{center}
  \includegraphics[width=.7\textwidth]{Figs/ControlSimulator/deterministic_example.png}
  \caption[]{An example trajectory for the deterministic dynamics. Above is the
  state evolution, $x(t)$ given $\b=0$ and below is the applied control,
  $\a(t)$. }
  \label{fig:deterministic_soln_controls}
\end{center}
\end{figure}

\section{Open-loop Stochastic Control}
As we have mentioned before, when the value of $X_t$ is unknown, the best one
can do is to use the transition density to perform the optimization. Since the
transition density follows a (deterministic) PDE, we will thus apply a Maximum
Principle for PDEs as a method of obtaining the optimal control.

\subsection{Fokker-Planck equation for the state density evolution}

We will need the transition density of $X$, conditional on no
spikes having occurred. We will write this as:
\begin{align*}
f(x,t) \intd{x} &= \Prob[X_t\in \intd{x} | X_0 = .0, X_{s<t} < \xth]  								  
\end{align*}
Then $f$ satisfies a Fokker-Planck equation with absorbing boundaries
\begin{equation}
\begin{gathered}
\di_t f(x,t) =
				\frac{\b^2 }{2}\cdot \di^2_x f -  
				\di_x \Big[ (\a(t)- \frac{x}{\tc})  \cdot f \Big]
\\
\\
\begin{array}{lll}
	&\textrm{st.\ }& 
	\left\{ \begin{array}{lcl}
	 f(x,0) &=& \delta(x) \quad (\textrm{delta function}
	\\
	(\a(t)- \tfrac x \tc)f - \di_x \tfrac {\b^2}2 f] \big|_{x=\xmin} &\equiv& 0
	\quad (\textrm{reflecting BCs at some } \xmin )
	\\
	f(x,t) |_{x=\xth} &\equiv& 0 \quad (\textrm{absorbing BCs at } \xth)
\end{array} \right. 
\end{array}
\label{eq:FP_pde_OU_PDF}
\end{gathered}
\end{equation}
In theory, $\xmin = -\infty$, but in the numerics below we will need to
truncate it to some finite value.

The probability flow, $\phi$, here is
$$
\phi(x,t) = (\a- \tfrac x \tc)f - \di_x [\tfrac {\b^2}2 f]
$$
as such the $f$ dynamics can also be written as:
$$
\di_t f(x,t) = - \di_x \phi(x,t)
$$
and the lower BC is:
$$
\phi(x,t) |_{x=\xmin} \equiv 0
$$

We will also need a short hand notation for the differential operator on the
right side of \cref{eq:FP_pde_OU_PDF}. Let
$$
\L_\a[\cdot] := \di^2_x [ \frac{\b^2 }{2} \cdot] -  
				\di_x  [ (\a(t)- \frac{x}{\tc}) \cdot]
$$
This is a linear and bounded differential operator. . We will usually omit the
subscript $\a$, but have written it now, since the operator is parametrized by
$\a$, that is by the control.

\subsection{Restating the objective in terms of $f$}
Let us recall our objective, \cref{eq:OC_LS_variance}:
$$
J(\a) = \Exp\left[
(\ts - \T \big)^2 
+  
\e \int_0^\ts  \a^2(s) \intd{s}
\right]
$$
Rewriting this in terms of $f$ reads:
\begin{multline}
J(\a) = 
\int_\xmin^\xth \Ttwo(x) f(x,\T) \intd{x}
\\
+ \int_0^\ts \phi(\xth, s) (s-\T)^2 \intd{s}
\\
+  \e \int_0^\ts  \a^2(s)  \int_\xmin^\xth f(x,s) \intd{x} \intd{s}
\label{eq:OC_LS_variance_density}
\end{multline}

Let us explain in more detail each term on the RHS in
\cref{eq:OC_LS_variance_density}:

The first term
$$
\int_\xmin^\xth \Ttwo(x) f(x,\T) \intd{x}
$$
counts the cost of trajectories which spike too late:
this cost is the expected squared time-to-hit starting at $x$, with  $\a
= \amax$, weighted by the probability of $X_\T = x$ , $f(x,\T)$. Recall
that we calculated $\Ttwo(x)$ in \cref{sec:calculating_TCs}.

The second term
$$
\int_0^\ts \phi(s,t) (s-\T)^2 \intd{s}
$$
counts the cost of trajectories which spike too soon, that
is the squared difference between realized spike time, $t$ and desired spike
time $\T$, weighted by the probability of spike at $t$ which is just the outflow of
probability, $\phi(\xth,t)$. Note further that due to the homogeneous
Dirichlet BC on $f$ at $\xth$, the outflow is simply:
$$
\phi(\xth, t) = -\frac{\b^2 }{2} \di_x f(\xth, t) 
$$

Finally, the third term
$$
\e \int_0^\ts  \a^2(s)  \left[  \int_\xmin^\xth f(x,s) \intd{x} \right] \intd{s}
$$
is the energy cost: the inner integral, $\int_\xmin^\xth f(x,s) \intd{x}$
takes into account that we incur an energy cost only for those trajectories
which have not yet spiked.

With that our control $\a$ will naturally be found via:
$$
\a = \argmin J[\a]
$$

\subsection{Optimizing with $f$ using a Pontryagin-type Maximum Principle}
\label{sec:PDE_max_principle_for_pdf}
By now our ostensibly stochastic optimal control problem modelled by SDEs
has become a deterministic optimal control problem modelled by PDEs. Our control
$\a$ influences the evolution density $f$ and via $f$, the integrals in the
objective, $J$.

As in the ODE theory, we will introduce an adjoint function, $p$, which will be
crucial in setting up necessary conditions for the optimal control.

The basic formulation for the adjoint PDEs and the abstract Hamiltonian can be
found in \cite{Fattorini1999,Palmer2011}. However our cost functional is a
little strange as it involves values of $\di_x f$ at the boundary. As such we
will derive the optimality condition heuristically, appealing to the
references for rigorous proofs.

Before proceeding, one might wish to read
\cref{sec:Pontryagin_heuristic_derivation} for a simpler version of the
following derivation with exactly the same ideas.

We start by assuming that $f,\a$ are the optimum trajectory/control pair and we
ask what necessary conditions must be true for $f,\a$. In particular if we take
a small variation around $\a$, we will have:
\begin{align*}
\a_\e = \a + \e \da
\\
f_\e = f + \e \df
\end{align*}

With that the variation of the cost with respect to the control at the optimal
$\a$ can be calculated as:
\begin{equation}
\frac {dJ}{d\e} \Big|_{\e = 0} =
- \int_0^\T (s-\T)^2 \frac {\b^2}{2} \di_x  \df \intd{s}  
+ \int_\xmin^\xth \Ttwo\df\intd{x}
+ \tilde{\e} \int 2 \a \da \intd{s}  
\label{eq:J_variation_PDE}
\end{equation}
This must be non-negative for a true minimum, but before we can say more, we
need to eliminate $\df$ from $\frac {dJ}{d\e}$! In order to eliminate $\df$
from \cref{eq:J_variation_PDE} we will introduce an adjoint variable, $p$. What
we would want from this adjoint variable $p$ would be that:
\begin{equation}
p(x,\T) = \Ttwo(x)
\label{eq:adjoint_TC_desideratum}
\end{equation}  
and 
\begin{equation}
\di_t \left[ \int_\xmin^\xth  p \cdot\df \intd{x} \right] = (s-\T)^2 \frac
{\b^2}{2}
\di_x
\df
 + \text{stuff}
\label{eq:adjoint_dynamics_desideratum}
\end{equation}
where 'stuff' does not depend on $\df$.

If we had such a $p$ satisfying
\cref{eq:adjoint_TC_desideratum,eq:adjoint_dynamics_desideratum} then
\cref{eq:J_variation_PDE} would read:
\begin{equation}
\frac {dJ}{d\e} \Big|_{\e = 0} =
\tilde{\e} \int 2 \a\da  \intd{s} +   \int_0^\T  \text{'stuff'}\intd{s}
\label{eq:J_variation_PDE_nodf}
\end{equation}
which does not depend on $\df$! and we would be on our way. In order to find
$p(x,t)$ we will need to
\begin{enumerate}
  \item calculate $\di_t \df$
  \item specify $\di_t p $ st. \cref{eq:adjoint_dynamics_desideratum} holds.
\end{enumerate}

\subsection{The PDE for $\df$}
First, calculate the PDE that $\df$ satisfies:
\begin{multline}
\di_t f_\e =  \L_{\a_\e} [f_\e]
\\
= -\di_x [ (\a_\e - \tfrac{x}\tc) f_\e  - D \di_x f_\e]
\\
=  -\di_x [ (\a + \e\da - \tfrac{x}\tc) f_\e  - D \di_x f_\e]
\\
=  -\di_x [ (\a + \e\da - \tfrac{x}\tc) (f + \e \df)  - D \di_x (f + \e \df)]
\\
=  -\di_x [ (\a + \e\da - \tfrac{x}\tc) (f)  - D \di_x (f)]
- \di_x [ (\a + \e\da - \tfrac{x}\tc) (\e \df)  - D \di_x (\e \df)]
\\
= \L_\a f  -\di_x [ \e\da\cdot  f]
  +\e \L_\a \df + o(\e) 
\\
= \di_t f + \e \di_t \df
\end{multline}
Of course, we have $\di_t f = \L_\a f$, so we conclude that:
\begin{equation}
\di_t \df = \L_\a \df -\di_x [\da \cdot f] 
\end{equation}
Now what are the ICs? Well, both $f_\e$ and $f$ must satisfy the system ICs in
\cref{eq:FP_pde_OU_PDF}, so $$\df (x,0) = 0$$.
Finally, what are the BCs?
On the upper side:
$$
f_\e |_\xth = 0 = [ f  + \e \df |_\xth = 0 + \e \df|_\xth \implies \df |_\xth =
0 $$
While on the lower side:
\begin{align*}
0 =& [ (\a_\e - x/\tc)f_\e - D \di_x f_\e |_\xmin
\\
=& (\a +\e \da - x/\tc)(f + \e \df ) - D \di_x (f + \e \df) |_\xmin
\\
=& [(\a +\e \da - x/\tc)(f) - D \di_x (f) |_\xmin
\\
& + \e [ (\a +\e \da - x/\tc)(\df) - D \di_x (\df) |_\xmin
\\
=& \cancel{[(\a - x/\tc)f - D \di_x (f) |_\xmin}
\\
&+ \e \da f|_\xmin
\\
&+ \e [ (\a - x/\tc)(\df) - D \di_x (\df) |_\xmin   + o(\e)
\end{align*}

so $\df$ satisfies the following system of equations:
\begin{equation}
\begin{gathered}
\begin{aligned}
\di_t \df 
&= \L_\a \df -\di_x [\da \cdot f]
\\
&= -\di_x \left[ (\a - \tfrac x\tc)\df - \tfrac{\b^2}2 \di_x \df \right]
-\di_xf \cdot\da
\end{aligned}
\\
\begin{cases}
\df(x,0) &= 0
\\
[ (\a - \frac x \tc)(\df) - D \di_x (\df) |_\xmin &= -\da \cdot f|_\xmin
\\
\df|_\xth &= 0
\end{cases}
\end{gathered}
\label{eq:FP_variation_near_optimum}
\end{equation} 
With \cref{eq:FP_variation_near_optimum} we are ready to design the dynamics of
$p$ such that $\df$ is eliminated from the optimality equation,
\cref{eq:J_variation_PDE}.

\subsection{Calculating the adjoint }
Recall our desideratum, we want
$$
p(x,\T) = \Ttwo(x) \quad (\text{required TCs})
$$
and
$$
\int_0^\T \di_t (p \cdot \df) = \int_0^\T(s-\T)^2 \frac {\b^2}{2} \di_x 
\df \intd{s} + \text{stuff}
$$
This is equivalent to having
$$
\di_t (p \cdot \df) =  (s-\T)^2 \frac {\b^2}{2} \di_x 
\df  + \text{stuff}
$$
Let us calculate $\di_t (p \cdot \df)$
\begin{align*}
\di_t (p \cdot \df) =& \df \cdot \di_t p + p \cdot   \di_t \df
 \\
 =& \df \cdot \di_t p + p \cdot \left( \L [\df] -\di_xf \cdot\da \right)  
 \\
 =& \df \cdot \di_t p + p \cdot \L  [\df] - \underbrace{ p \cdot \di_xf \cdot\da}_{\text{stuff}}
\end{align*}
Now our job is clear - we want 
$$
\di_t p = -\Lstar[p]
$$
Where $\Lstar$ is defined by:
\begin{equation}
\begin{gathered}
\int_\xmin^\xth  \Lstar [p] \df \intd{x} 
=
\int_\xmin^\xth p \L [\df] \intd{x} - (s-\T)^2 \frac {\b^2}{2}\di_x \df +
\text{possibly more stuff}
\\
\\
\forall \df \quad \st \begin{cases}
\df(x,t) |_{x=\xth} & = 0 
\\
(\a - \tfrac x \tc)\df - \di_x \tfrac {\b^2}2 \df \big|_{x=\xmin} &=  -\da
\cdot f
\end{cases}
\end{gathered}
\end{equation}
where 'possibly more stuff' should be independent of $\df$.

Let us take a deep breath and calculate $\Lstar$:
\begin{align*}
\int p \L [\df] \intd{x} =&
\int p \underbrace{\left(\frac{\b^2 }{2}\cdot \di^2_x \df -  
			 \di_x \big[ (\a- \frac{x}{\tc})\cdot \df \big]\right)}_{\L [\df]} \intd{x}
\\
=& [ -p ((\a- \frac{x}{\tc})  \cdot \df)\quad  \vert^\xth_\xmin
+ \int \di_xp ((\a- \frac{x}{\tc})  \cdot \df) \intd{x}
\\
& + [p (\frac{\b^2 }{2}\cdot \di_x \df \quad \vert ^\xth_\xmin
- [ \di_x p \cdot \frac{\b^2 }{2}  \df\quad \vert^\xth_\xmin 
+ \int \frac{\b^2 }{2} \di^2_xp    \cdot \df \intd{x} 
\\
=& \int \Lstar [p] \df \intd{x} 
+\underbrace{[-p(\a- \frac x \tc) \cdot \df 
+ p \frac{\b^2}{2} \di_x \df 
-\di_x p \frac{\b^2}{2} \df \,\vert^\xth_\xmin }_{\text{BC terms}}
\end{align*}

Now we want to make the 'BC terms' equal to $(s-T)^2 \frac{\b^2}{2} \di_x \df$
+ 'possibly more stuff'.
In doing so, we will discover the correct BCs associated with the adjoint
operator, $\Lstar$:
\begin{align*}
\textrm{BC terms} =& 
\Big[ -p (\a- \frac{x}{\tc})  \cdot \df 
+ p \frac{\b^2 }{2}\cdot \di_x \df   
- \di_x p \cdot \frac{\b^2 }{2}  \df   \quad ^\xth_\xmin\vert 
\\
=& -\cancel{p (\a- \frac{x}{\tc})  \cdot \df \Big|_\xth}
 + p (\a- \frac{x}{\tc}) \cdot \df \Big|_\xmin
\\ 
& + [p \frac{\b^2 }{2}\cdot \di_x \df \Big|_\xth
 - [p \frac{\b^2 }{2}\cdot \di_x \df \Big|_\xmin
\\
&- \cancel{[ \di_x p \cdot \frac{\b^2 }{2}   \df\Big|_\xth}
 + [ \di_x p \cdot \frac{\b^2 }{2}   \df\Big|_\xmin
\\
=& \cancelto{-p (-\da f)}{-p ((\a- \tfrac x \tc)\df - \di_x \tfrac {\b^2}2
\df) \Big|_\xmin} 
+ [p \frac{\b^2 }{2}\cdot \di_x f \Big|_\xth
+ [ \di_x p \cdot \frac{\b^2 }{2}   f\Big|_\xmin
\\
=& (s-T)^2 \frac{\b^2 }{2}\cdot \di_x \df\Big|_\xth 
+  p (\da f)\Big|_\xmin
\end{align*}

Since we want this to be true for arbitrary $\df$, we need to enforce the
following (adjoint) BCs: 
$$
\begin{cases}
p |_\xth &= (s-T)^2
\\
\di_xp |_\xmin &= 0
\end{cases}
$$

In summary, the adjoint PDE equation is given by: 
\begin{equation}
\begin{gathered}
\begin{aligned}
\di_t p =& - \Lstar[p]
\\
		=&  
			- \Big[ \frac{\b^2 }{2}\cdot \di^2_x p +  
			(\a(t)- \frac{x}{\tc})  \cdot \di_x p \Big] 
\end{aligned}
\\
\st  
\begin{cases}
	p(x,\T) &= \Ttwo(x)
	\\ 
	\di_x p  \big|_{x=\xmin} &\equiv 0
	\\
	p \big|_{x=\xth} &\equiv (s-T)^2
\end{cases}
\label{eq:adjoint_pde_OU}
\end{gathered}
\end{equation}

And most importantly:
\begin{equation}
\int \Lstar [p] \df \intd{x}
=  
\int p \L [\df] \intd{x} 
  - (s-T)^2 \frac{\b^2 }{2}\cdot \di_x \df\Big|_\xth 
  -  p  f \cdot \da \Big|_\xmin
\label{eq:L_to_Lstar_relation}
\end{equation}

\subsection{A Pontryagin-type Minimum Principle for the optimal control $\a$}
Using \cref{eq:L_to_Lstar_relation} and the TCs in \cref{eq:adjoint_pde_OU}, we
will have:
\begin{align*}
\int \Ttwo \df \intd{x} =& 
\int_\xmin^\xth p \df \intd{x}
\\ =& \int_0^\T \di_t \int_\xmin^\xth p \cdot \df \intd{x} \intd{s}
\\
=& \int_0^\T \int _\xmin^\xth - \df \cdot \Lstar[p]
 + p \cdot \L_\a [\df] 
 -  p \cdot \di_x [\da \cdot f]  \intd{x}  \intd{s}
\\
=& 
\int_0^\T \Big\{  (s-T)^2 \frac{\b^2 }{2}\cdot \di_x \df\Big|_\xth 
+  p (\da f)\Big|_\xmin
\\
& \quad
- \int _\xmin^\xth p \cdot \di_x [\da \cdot f] \intd{x} \Big\} \intd{s}
\end{align*}

Now put this back into the equation of optimality, \cref{eq:J_variation_PDE}.
\begin{align*}
\frac {dJ}{d\e} \Big|_{\e = 0} =&
- \int_0^\T (s-\T)^2 \frac {\b^2}{2} \di_x  \df \intd{s}  
+ \int_\xmin^\xth \Ttwo\df\intd{x}
+ \tilde{\e} \int 2 \a \da \intd{s}
\\
=& 
\int_0^\T \Big\{
 p (\da f)\Big|_\xmin
+ \tilde{\e}  2 \a \da
- \int _\xmin^\xth p \cdot \di_x [\da \cdot f] \intd{x} 
 \Big\} \intd{s}
 \\
=& 
\int_0^\T \da \Big\{
 p f \Big|_\xmin
+ \tilde{\e}  2 \a
- \int _\xmin^\xth p \cdot \di_x f \intd{x} 
 \Big\} \intd{s}
\end{align*}

Now we pull the famous fundamental lemma of the calculus of variations out and
conclude that
\begin{equation}
\Big\{
 \tilde{\e}  2 \a(t)
+ p f \Big|_\xmin
- \int _\xmin^\xth p \cdot \di_x f \intd{x} 
\Big\} = 0
\quad \forall t \in [0,T]
\label{eq:J_necessary_condition}
\end{equation}

Working in reverse we can recognize a Hamiltonian, $\H$ as:
$$
\H(f,p,\a) =  \int \cref{eq:J_necessary_condition} \intd{\a} 
$$

Naturally, $f,p$ must satisfy their respective
\cref{eq:FP_pde_OU_PDF,eq:adjoint_pde_OU}.

And our optimal control, $\a$,  can be found via:
\begin{equation}
\a (t) = \frac { - p(\xmin,t) f(\xmin,t)
+ \int _\xmin^\xth p(x,t) \cdot \di_x f(x,t) \intd{x} 
}{2\e}
\label{eq:optimal_alpha_4_PDF_optimization}
\end{equation}
As always, subject to the appropriate bounds, \cref{eq:bound_constraints_alpha}.

\subsection{An Iterative Numerical Solution}
We have an entangled situation: $f,p$ depend on $\a$ through
\cref{eq:adjoint_pde_OU,eq:FP_pde_OU_PDF}, while $\a$, the optimal control, is
given by \cref{eq:optimal_alpha_4_PDF_optimization} which requires  the
solution to $f,p$ . So we need some kind of iteration to find $\a$ and the corresponding $f,p$.

The key to such an iteration is that the quantity 
$$\di_\a \H =  \Big\{
 \tilde{\e}  2 \a(t)
+ p(x,t) f(x,t) \Big|_\xmin
- \int _\xmin^\xth p \cdot \di_x f(x,t) \intd{x} 
\Big\}
$$
gives us the direction of increase
and we can use it as a gradient in a descent algorithm. 
This we can also write as:
\begin{align*}
\di_\a \H =&  \Big\{
 \tilde{\e}  2 \a(t)
+ p(x,t) f(x,t) \Big|_\xmin
- [p(x,t) f(x,t) \, _\xmin^\xth\Big|
+ \int _\xmin^\xth \di_x p \cdot f(x,t) \intd{x} 
\Big\}
\\
=&
 \Big\{
 \tilde{\e}  2 \a(t)
- p(x,t) f(x,t)\Big|_\xth
+ 2p(x,t) f(x,t) \Big|_\xmin
+ \int _\xmin^\xth \di_x p \cdot f(x,t) \intd{x} 
\Big\}
\end{align*}
which has certain numerical advantages.

So the plan is: we choose an initial $\a_1(t)$ - the deterministic solution
from \cref{sec:deterministic_ode_control} is a natural choice. Then solve for $f,p$.
Then evaluate $J$ and iterate until convergence. For our convergence
criterion we will take that the incremental difference in $J$ is smaller than
some difference small tolerance. The algorithm is specified more
precisely in Algorithm \ref{alg:hamiltonian_descent_4_PDF_OC}.
\begin{algorithm}
\begin{algorithmic}
\State $J_0 = \infty $ 
\State $\a_1 = $ deterministic $\a$, the optimal control given $\b=0$ from
\cref{sec:deterministic_ode_control}
\For { $k= 1\dots k_{\max}$}
	\State $\di_t f_{k} = \L_{a_{k}} [ f_{k}]$
	\State $-\di_t p_{k} = \Lstar_{a_{k}} [ p_{k}]$
	\State $J_{k} \gets J[\a_k, f_k]$
	\If {$|J_{k} - J_{k-1}| < tol $}
		\State BREAK
	\EndIf
	\State $\da = \di_a \H(f_{k} ,p_{k},  \a_{k})$

	\If {$\sup_t |\da(t)| < tol$}
		\State BREAK
	\EndIf

	\State $\a_{k+1} \gets \a_{k} - s \cdot \da  $
\EndFor
\State \Return $J_k, \a_k$
\end{algorithmic}
\caption{Optimal Control Descent Algorithm}
\label{alg:hamiltonian_descent_4_PDF_OC}
\end{algorithm}

We show one realization of $\di_a \H$ in \cref{fig:grad_aH_deterministic_alpha}.
\begin{figure}[htp]
\begin{center}
\subfloat[$\a_0$]
{\includegraphics[width=.49\textwidth]{Figs/FP_Adjoint/Deterministic_alpha_di_aH.png}
 }
  \subfloat[$\a_1$]
{  \includegraphics[width=.49\textwidth]{Figs/FP_Adjoint/First_control_iteration.png}
}
\caption[labelInTOC]{Iteratively calculating $\a$ for the
open-loop optimal control. Plotted is an example of
$-\di_a \H$ calculated at $\a(t) = \a_1(t)$, where $a_1$ is the deterministic control. The green curve, $-\di_a \H$
indicates the direction that we want to change the control, if it is negative
we decrease and if it is positive we increase it. On the right panel, b), we
show the result of applying the incremental change by also plotting the
resulting $\a_1(t)$. }
\label{fig:grad_aH_deterministic_alpha}
\end{center}
\end{figure}

Finally, we are ready to calculate the open-loop stochastic optimal control for
this parameter set. The results are in
\cref{fig:FP_adjoint_objective_control_convergence}.

\begin{figure}[h]
\begin{center}
\subfloat[$\a_k(t)$ ]
{
\includegraphics[width=0.48\textwidth]
{Figs/FP_Adjoint/ExampleControlConvergence_control.png}
}
\subfloat[$J_k$ ]
{
\includegraphics[width=0.48\textwidth]
{Figs/FP_Adjoint/ExampleControlConvergence_objective.png}
}
\caption[ ]{The result of an entire optimization iteration, on the left the
iterates of the control, $\a_k(t)$, on the right the progress of the
objective with each iteration. The final control, in purple, is inhibitory at
the beginning of the time interval, $\a < 0$, unlike the the deterministic
control solution. 
We see a significant
reduction in the objective value, $J$, and thus an improvement, on the
order of 50\%.}
\label{fig:FP_adjoint_objective_control_convergence}
\end{center}
\end{figure}

% \begin{table}[h] 
% \centering
% \begin{tabular}{lc}
% Control Law & Squared Error \\
% \hline
% Deterministic &  0.647 \\
% Stochastic &  0.336\\
% Theory (Value function) & 0.285
% \end{tabular}
% \caption{Realized performance of the different control laws and the theoretical
% expected performance of the stochastic law (last row)}
% \label{tab:realized_avg_errors_det_vs_stoch}
% \end{table}
% 
% \begin{figure}[h]
% \begin{center}
% \subfloat[A]
% {
% \label{fig:controlled_traj_ex1}
% \includegraphics[width=0.33\textwidth]
% {Figs/ControlSimulator/example_controlled_trajectories_id1.png}
% }
% \subfloat[B]
% {
% \label{fig:controlled_traj_ex2}
% \includegraphics[width=0.33\textwidth]
% {Figs/ControlSimulator/example_controlled_trajectories_id4.png}
% }
% \subfloat[C]
% {
% \label{fig:controlled_traj_ex3}
% \includegraphics[width=0.33\textwidth]
% {Figs/ControlSimulator/example_controlled_trajectories_id3.png}
% }
% \caption[]{Examples for the controlled trajectories using both the deterministic
% and the stochastic control approaches. The red vertical line in the plots
% indicates the desired spike-time, $\T$. In A, B the performance of both
% control laws is essentially the same, but in C we see the advantage of the
% stochastic approach. $\tc, \b = [.75, 1.25]$.}
% \label{fig:control_trajectories_examples}
% \end{center}
% \end{figure}
% \begin{figure}[htp]
% \begin{center}
%   \includegraphics[width=.9\textwidth]{Figs/ControlSimulator/example_controlled_trajectories_hists.png}
%   \caption[labelInTOC]{Histogram of the spike timing error for the
%   deterministic (left) vs. the stochastic (right) control laws.}
%   \label{fig:error_histograms_det_vs_stoch}
% \end{center}
% \end{figure}

\section{Test Comparison of Different Controls}
\label{sec:probabilistic_numerical_test}
We will now compare the performance of the three obtained controls: the
closed-loop stochastic obtained via the HJB equation, the open-loop
deterministic control obtained when setting $\b=0$ and the open-loop stochastic
control obtained by optimizing using the transition density.

For our comparison we will sample $128$ realizations of the controlled system
and apply in turn each of the three controls. Naturally, for each path, we will
apply exactly the same random disturbances to each controller.

We will set $\e = .001$ and $\amax = 2.0$. As such the energy cost is of
secondary importance and the paramount effect on the objective is the difference
$(\ts - \T)$, which we are trying to minimize. We set $\tc, \b = [.75, 1.25]$.

The performance of the three controllers is given in
\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch}. We see that while the
open-loop stochastic control is not as good as the feedback stochastic
control, it is much closer in performance to it than to the blind
deterministic control.

For illustration sake, we also show the error
histograms in \cref{fig:error_histograms_det_vs_openloop_vs_stoch} and visualize
some trajectories in \cref{fig:control_trajectories_examples}.

\begin{table}[h] 
\centering
\begin{tabular}{lcc}
Control Law & Empirical Squared Error & Theoretical Squared Error \\
\hline
Deterministic &  0.621 & - \\
Open-Loop Stochastic & 0.356 & 0.382\\
Feedback Stochastic &  0.288& 0.285\\
\hline
\end{tabular}
\caption{Realized and theoretical performance of the different control laws. The empirical
performance is obtained using 128 sample paths. The theoretical performance is
found using the $J$ in for the open-loop stochastic control and the value
function $v(x=0, t =0)$ for the closed-loop stochastic control.}
\label{tab:realized_avg_errors_det_vs_openloop_vs_stoch}
\end{table}
\begin{figure}[h]
\begin{center}
\subfloat[]
{
\label{fig:controlled_traj_ex1}
\includegraphics[width=0.33\textwidth]
{Figs/ControlSimulator/3controls_example_trajectories_id3.png}
}
\subfloat[ ]
{
\label{fig:controlled_traj_ex2}
\includegraphics[width=0.33\textwidth]
{Figs/ControlSimulator/3controls_example_trajectories_id5.png}
}
\subfloat[ ]
{
\label{fig:controlled_traj_ex3}
\includegraphics[width=0.33\textwidth]
{Figs/ControlSimulator/3controls_example_trajectories_id6.png}
}
\caption[]{Examples for the controlled trajectories using the deterministic,
open-loop stochastic and feedback stochastic control approaches. The
black vertical line in the plots indicates the desired spike-time, $\T$.
For the parameter values are $\tc, \b = [.75, 1.25]$. The three panels are
three different realizations of the model dynamics.}
\label{fig:control_trajectories_examples}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
  \includegraphics[width=.95\textwidth]
  {Figs/ControlSimulator/3controls_example_trajectories_hists.png}
  \caption[labelInTOC]{Histogram of the spike timing error for the
  deterministic (left) vs. open-loop stochastic(centre) vs. feedback stochastic
  (right) control laws. This is for the same problem as in
  \cref{fig:control_trajectories_examples}. We have used $N=128$ sample paths
  to form the statistics}
  \label{fig:error_histograms_det_vs_openloop_vs_stoch} 
\end{center}
\end{figure}

\section{Research Goals}
I hope to be able to use the techniques of open-loop stochastic control
discussed herein to address problems in stochastic control where for whatever
reasons, the dynamic programing formalism is not applicable.

One such reason might be that the state is unobservable. An example of such a
system is the control of blood glucose for diabetes or intensive care
patients, where the patients' blood glucose is modelled to follow a stochastic
differential equation and actual observations are expensive and invasive and
thus rare, on the order of a few hours. In this scenario, the applied external
glucose or insulin to the patient in the time in-between measurements can be
thought of as the solution to an optimal control problem where the objective is
to minimize deviations of the patient's blood glucose from some baseline.

An alternative reason to use the open-loop control techniques is when optimizing
with multi-dimensional systems at which point the dynamic programing approach
suffers from 'the curse of dimensionality' which is the colloquial name for the
exponential increase in necessary computing resources. If the optimality
conditions for the open-loop control lend themselves to a Feynman-Kac-type
representation, then one might replace solving the PDEs by an appropriate
Monte-Carlo process and thus make the computations tractable.

\clearpage
\appendix
\section{Heuristic Derivation of Pontryagin's Minimum Principle}
\label{sec:Pontryagin_heuristic_derivation}
Here we will show a heuristic derivation for the Pontryagin Minimum Principle
for linear ODE dynamics. It should motivate the manipulations in
\cref{sec:PDE_max_principle_for_pdf}.

The following is almost a direct transcription of A.1 in
\cite{Evansb}, with the added simplification of linear dynamics
and linear cost and with the notation adapted to what we have used
hereto.

The dynamics are given by:
$$
\dot{x} = A(t, \a) x
$$
where $A$ is a matrix in $\R^{n_x \times n_x}$ whose values depend on the
time, $t$ and the control $\a$. The cost, $J$, is given by:
$$
J[\a] = \int_0^\tf L(t,\a) \cdot x \intd{s} + M \cdot x(\tf)
$$
where $L(t,\a), M$
are vectors in $\R^{n_x}$. Both the dynamics and the cost are linear in the state $x$.
Suppose $x,\a$ are the optimal trajectory - then small variations in $\a$ can
only worsen $J$. Let such an acceptable variation be given by:
$$
\a_\e = \a + \e \b
$$
for some small $\e$. We will have a corresponding dynamic, $x_\e$ st:
$$
x_\e = x + \e \dx+ o(\e)
$$
Now obtain a differential equation for the variation $\dx$:
\begin{align*}
\dot{x_\e} &=  \dot{x} + \e \dot{\dx} 
\\
&= A(\a_\e) x_\e = A(\a) x + \e \dot{\dx}
\end{align*}
Now divide through by $\e$ and let it go to zero:
$$
\implies \dot{\dx} = A \dx + [\grad_\a [ A(t,\a)x]] \cdot\b
$$
Note that $$\dx(0) = 0$$ since $x(0)$ is specified and so there is no
variation there.

With that the derivative of $J$ at $\a$ will read:
\begin{multline}
\frac {dJ}{d\e} \Big|_{\a} = \lim \frac{ J[\a_\e] - J[\a]}{\e} 
= \int L(\a,t) \dx +  \grad_\a L \cdot \b \intd{s} + M \dx(\tf)
\geq 0 
\label{eq:J_variation_ODE}
\end{multline}
The problem is that the state variation $\dx$ also appears in this derivative,
whereas we just want the control variation $\b$ to appear there.

The trick then is to introduce the adjoint variable $p$ with appropriate
dynamics such that all terms involving $\dx$ in \cref{eq:J_variation_ODE} are
eliminated.

'Clearly', we want to replace the term $ M \cdot \dx $ first. Thus we set 
$$
p(\tf) = M
$$

Now notice the following:
$$p(\tf)\dx(\tf) = \cancelto{0}{p(0)\dx(0)} + \int_0^\tf d_t (p \cdot \dx)$$

So if we make $\int d_t (p \cdot \dx) = - \int r(\a,t) \dx$ we would be
eliminate the $\dx$ terms in the expresion for $dJ/d\eps$ entirely!

Now the way is clear:

Let 
$$ 
\dot{p} = -A^T p - L(\a,t)
$$
Then indeed:
\begin{multline}
d_t (p \cdot \dx) = \dot{p} \dx + p \dot{\dx}
\\
= (-A^T p - r)\dx + p (A\dx+ [\grad_\a [ A(t,\a)x]] \cdot\b)
\\
= -r\dx + p ([\grad_\a [ A(t,\a)x]] \cdot\b)
\end{multline}
That is almost what we wanted and it will work, since it only contains
variations wrt the control $\b$, not the state, $\dx$.

Thus
\cref{eq:J_variation_ODE} is transformed to read:
\begin{multline}
\frac {dJ}{d\e} \Big|_{\a} = 
\int r(\a,t) \dx +  \grad_\a r \cdot \b \intd{s} + \psi \dx(\tf)
\\
=
\int r(\a,t) \dx +  \grad_\a r \cdot \b \intd{s} + p(\tf) \dx(\tf)
\\
= \int r(\a,t) \dx +  \grad_\a r \cdot \b \intd{s} - \int r \dx - p ([\grad_\a [
A(t,\a)x]] \cdot\b)
\\
=
\int_0^\tf  
[\underbrace{p ([\grad_\a [A(t,\a)x]] ) + \grad_a r)}_{\di_\a \H} ] 
\b \intd{s}
\end{multline}
As always in these variational arguments, the final point is that since $\b$ is
arbitrary, we must have that the rest of the integrand be identically zero -
thus we see where the control theory Hamiltonian, $\H$, comes from.

Except for the nitty-gritty manipulations of the BCs in
\cref{sec:PDE_max_principle_for_pdf}, what is done there is exactly what is done
here - a heuristic derivation for a necessary condition for a minimum.

\bibliographystyle{plain}
\bibliography{library,local}

\end{document}

### 2013-08-22 IOPLaTeXGuidelines.tex
Initial commit

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}  
\begin{document}

\title[Author guidelines for IOP journals in  \LaTeXe]{How to prepare and submit an article for 
publication in an IOP journal using \LaTeXe}

\author{author of this}

\address{IOP 
Publishing, Dirac
House, Temple Back, Bristol BS1 6BE, UK}
\ead{custserv@iop.org}
\begin{abstract}
This document describes the  preparation of an article using \LaTeXe\ and 
\verb"iopart.cls" (the IOP \LaTeXe\ preprint class file).
This class file is designed to help 
authors produce preprints in a form suitable for submission to any of the
journals published by IOP Publishing.
Authors submitting to any IOP journal, i.e.\ 
both single- and double-column ones, should follow the guidelines set out here. 
On acceptance, their TeX code will be converted to 
the appropriate format for the journal concerned.

\end{abstract}

%Uncomment for PACS numbers title message
%\pacs{00.00, 20.00, 42.10}
% Keywords required only for MST, PB, PMB, PM, JOA, JOB? 
%\vspace{2pc}
%\noindent{\it Keywords}: Article preparation, IOP journals
% Uncomment for Submitted to journal title message
%\submitto{\JPA}
% Comment out if separate title page not required
\maketitle

\section{Introduction: file preparation and submission}
The \verb"iopart" \LaTeXe\ article class file is provided to help authors prepare articles for submission to the IOP journals. Submission is not restricted to those using of this class file and articles prepared using any other class or style files can also be submitted. Use of the \verb"iopart" class file does, however, help to speed the publication process and it is hoped it will make it easy for authors to prepare their articles. This document describes and demonstrates the use of \verb"iopart.cls" to produce a preprint for submission and refereeing. The detailed formatting for print and web use will be undertaken by IOP as part of the production process after the article has been accepted for publication, but use of \verb"iopart" does make this simpler.

\subsection{What you will need to supply}
Full details on how to submit files to a particular IOP journal are contained in the document {\it Guidelines for authors} which is discussed in section \ref{agide}. Here we mention the key points you need to consider when preparing your files for submission.

\subsubsection{{\bf Text}} The \LaTeX\ source code of your paper with all figures included in 
the source code (see section \ref{figinc}). Although we recommend that authors use the \verb"iopart" class file, articles prepared using almost any  version of \TeX\ or \LaTeX\ 
can be handled.  
\begin{itemize}
\item {\bf Using \LaTeX\ packages} Most \LaTeXe\ packages can be used if they are 
available in the normal distribution of \LaTeXe; however, if it is essential to use 
a non-standard package then any extra files needed to process the article must 
also be sent in. Authors should be aware that the final version will be printed on 
a different page size and using different fonts to the preprint version so that any 
special effects used should not contain material that is not easily scalable.
Alterations to the source code will be made during the production process in order to 
conform to IOP house style and journal format.
\end{itemize}
\subsubsection{\bf Figures} Figures must be included in an article as encapsulated PostScript files
(see section \ref{figinc}) or created using standard \LaTeX\ drawing commands. 
 Please name all figure files using the guidelines in section \ref{fname}. For more detail
on preparation of EPS figure files, please refer to IOP's graphics guidelines (\gguide) 
which can be downloaded from authors.iop.org.

\begin{itemize}
\item {\bf Including figures\label{fig1}} All figures can be included within the body of the text 
at an appropriate point or grouped together with their captions at the end of the article. A standard graphics inclusion package such as \verb"graphicx" should be used for figure inclusion.
Authors should avoid using special effects generated by including verbatim
PostScript code. Wherever possible, please try to use standard \LaTeX\ tools 
and packages.
\end{itemize}

\subsubsection{\bf BiBTeX \label{bibby}}  If you use BiBTeX to prepare your
references, our preferred  \verb".bst" styles are:

\begin{itemize}
\item for the numerical (Vancouver) reference style we recommend that authors use 
 \verb"unsrt.bst";
\item for alphabetic (Harvard) style reference we recommend that authors use
the \verb"jphysicsB.bst" BiBTeX style file together with the harvard \LaTeX\ package. 
\verb"jphysicsB.bst" is supplied as part of the \verb"harvard.sty" package which can be 
downloaded from www.ctan.org.
\end{itemize}

Please make sure that you include your .bib bibliographic database file(s) and any 
.bst style file(s) you have used (other than unsrt.bst or jphysicsB.bst).

\subsubsection{\bf Copyrighted material} If you wish to illustrate your article using material for which you do not own the copyright then you must seek permission from the copyright holder, usually both the author and the publisher. It is the author's responsibility to obtain copyright permissions and this should be done prior to submitting your article. If you have obtained permission, please provide full details of the permission granted---for example, copies of the text of any e-mails or a copy of any letters you may have received. Figure captions must include an acknowledgment of the original source of the material even when permission to use has been obtained. Full details on copyright can be found in IOP's {\it Guidelines for authors} which can be accessed online at authors.iop.org.

\subsection{Naming your files}
Please name all your files, both figures and text as follows:
\begin{itemize}
\item use only characters from the set a to z, A to Z, 0 to 9 and underscore (\_); 
\item do not use spaces in file names;
\item include an extension to indicate the file type (e.g., .tex, .eps, .txt, etc);
\item do not use any accented characters in file names; for example, \'a, \^e, \~n, \"o, etc because 
they can cause difficulties when processing your files.
\end{itemize}
\subsubsection{\label{fname}Naming your figure files} In addition to the above points, please give each figure file a name which indicates the number of the figure it contains; for example, figure1.eps, figure2a.eps etc. If the figure file contains a figure with multiple parts, for example figure 2(a) to 2(e), give it a name such as figure2a\_2e.eps, and so forth.
\subsection{Submitting your article files to an IOP journal}
For most IOP journals you send your files directly to us by web, e-mail or FTP; however, there are a number of exceptions to this. For certain IOP journals you need to send your files to a different institution or to an external editor. How to find the submission address for a particular journal is discussed in section 1.4, below.
\subsection{Where to send your files \label{agide}}
Full details of how to submit files to a particular IOP journal are contained in IOP's {\it Guidelines for authors} which can be accessed online by going to authors.iop.org. From this document's table of contents, select the `How to submit' link to access full details of the submission address for a particular journal.

\subsection{Obtaining the class file and documentation}
The  \LaTeXe\ class file and guidelines can be downloaded from \verb"authors.iop.org".
Files are available in zipped format for PCs, tar compressed format files for
Unix or as a stuffit archive for Macintosh.

\section{Preparing your article}

Please follow these guidelines as closely as possible, particularly with regard to the preparation of the reference list.

At the start of the \LaTeX\ source please include 
commented material to identify the journal, author and reference number if 
known. The first non-commented line should be 
\verb"\documentclass[12pt]{iopart}"  to load the preprint class 
file. Omitting \verb"[12pt]" produces an article with the normal journal
page and type sizes. The start of the article text is signalled by 
\verb"\begin{document}".
Authors of very long articles may find it convenient to separate 
their article into a series of files each containing a section, each of which is called 
in turn by the primary file.

Authors may add their own macros 
at the start of an article provided they do not overwrite existing definitions and 
that they send copies of their new macros with their text file. Macros for the individual paper not included 
in a style file should be inserted in the preamble to the paper with comments to
describe any complex or non-obvious ones.
\verb"iopart" can be used with other package files such
as those loading the AMS extension fonts 
\verb"msam" and \verb"msbm" (these fonts provide the 
blackboard bold alphabet and various extra maths symbols as well as symbols useful in figure 
captions); an extra style file \verb"iopams.sty" is provided to load these
packages and provide extra definitions for bold Greek letters.

\subsection{Double-column journals}
Authors writing for double-column journals should use the \verb"iopart" class file. Conversion from the single-column format to the double-column output required for printing will be done during the production process. 
However, authors should bear in mind that all mathematical formulae will need to be fitted
into the width of a single column, so individual lines of equations should not occupy more than two thirds of the line width in this preprint form.

\section{The title and abstract page} 
The code for setting the title page information is slightly different from
the normal default in \LaTeX.

\subsection{Titles and article types}
The title is set using the command
\verb"\title{#1}", where \verb"#1" is the title of the article. The
first letter 
of the title should be capitalized with the rest in lower case. 
Mathematical expressions within the title may be left in light-face type.

If the title is too long to use as a running head at the top of each page (apart from the
first) a short
form can be provided as an optional argument (in square brackets)
before the full title, i.e.\ \verb"\title[Short title]{Full title}".

For article types other than papers, \verb"iopart.cls"
has a generic heading \verb"\article[Short title]{TYPE}{Full title}" 
and the specific definitions given in table~\ref{arttype}. In each case (apart from Letters
to the Editor) an 
optional argument can be used immediately after the control sequence name
to specify the short title; where no short title is given the full title
will be used as the running head. 
For Letters use \verb"\letter{Full title}", no short title is required as 
the running head is automatically defined to be {\it Letter to the Editor}.
The generic heading could be used for 
articles such as those presented at a conference or workshop, e.g.
\small\begin{verbatim}
\article[Short title]{WORKSHOP ON HIGH-ENERGY PHYSICS}{Title}
\end{verbatim}\normalsize
Footnotes to titles may be given, though acknowledgment of 
funding should be included in the acknowledgments section rather than here. A footnote can be included by using \verb"\footnote{Text of footnote.}" immediately after the title.

\begin{table}
\caption{\label{arttype}Types of article defined in the {\tt iopart.cls} 
class file.}
\footnotesize\rm
\begin{tabular*}{\textwidth}{@{}l*{15}{@{\extracolsep{0pt plus12pt}}l}}
\br
Command&Type&Heading on first page\\
\mr
\verb"\title{#1}"&Paper&---\\
\verb"\review{#1}"&Review&REVIEW\\
\verb"\topical{#1}"&Topical review&TOPICAL REVIEW\\
\verb"\comment{#1}"&Comment&COMMENT\\
\verb"\note{#1}"&Note&NOTE\\
\verb"\paper{#1}"&Paper&---\\
\verb"\prelim{#1}"&Preliminary communication&PRELIMINARY COMMUNICATION\\
\verb"\rapid{#1}"&Rapid communication&RAPID COMMUNICATION\\
\verb"\letter{#1}"&Letter&LETTER TO THE EDITOR\\
\verb"\article{#1}{#2}"&Other articles&Whatever is entered as {\tt 
\#1}\\
\br
\end{tabular*}
\end{table}

\subsection{Authors' names and addresses}
The next information required is the list of authors' names and 
their affiliations. For the authors' names type \verb"\author{#1}", 
where \verb"#1" is the 
list of all authors' names. The style for the names is initials then
family name, with a comma after all but the last 
two names, which are separated by `and'. Initials should {\it not} be followed by full stops. First (given) names may be used if 
desired and Chinese-style names included in the form they should be printed in. If the authors are at different addresses a superscripted number, e.g. $^1$, \verb"$^1$", should be used after each 
name to reference the author to his/her address. 
If an author has additional information to appear as a footnote, such as 
a permanent address, a normal \LaTeX\ footnote command
should be given after the family name and address marker 
with this extra information.

The addresses of the authors' affiliations follow the list of authors. 
Each address is set by using
\verb"\address{#1}" with the address as the single parameter in braces. 
If there is more 
than one address then the appropriate superscripted number, followed by a space, should come at the start of
the address.

Please also add the e-mail addresses for at least one of the authors. This is done by inserting the 
command \verb"\ead{#1}" after the postal address(es) where \verb"#1" is the e-mail address.  
See section~\ref{startsample} for sample coding. For more than one e-mail address, please use the command 
\verb"\eads{\mailto{#1}, \mailto{#2}}" with \verb"\mailto" surrounding each e-mail address.

\subsection{The abstract}
The abstract follows the addresses and
should give readers concise information about the content 
of the article and indicate the main results obtained and conclusions 
drawn. It should be self-contained with no reference to 
figures, tables, equations, bibliographic references etc
included and should not normally exceed 200 
words.
To indicate the start 
of the abstract type \verb"\begin{abstract}" followed by the text of the 
abstract (not in braces).  The abstract should normally be restricted 
to a single paragraph and is terminated by the command
\verb"\end{abstract}"

\subsection{Subject classification numbers}
Any Physics and Astronomy Classification System (PACS) 
codes or Mathematics Subject Classification (MSC) scheme 
numbers should come immediately after the abstract. 
Classification codes can greatly help in the choice of suitable 
referees and allocation of articles to subject areas. 
For {\it Inverse Problems} and {\it Nonlinearity} authors may 
use either PACS or MSC codes.

PACS or MSC numbers are included after the abstract 
using \verb"\pacs{#1}" and \verb"\ams{#1}" respectively.

After any classification numbers the command
\verb"\submitto{#1}" can be inserted, where \verb"#1" is the journal name written in full or the appropriate control sequence as
given in table~\ref{jlab1}. This command is not essential to the running of the file.

\subsubsection{Information on PACS and MSC}
For more information on PACS and MSC see
\begin{itemize}
\item MSC: http://www.ams.org/msc
\item PACS: http://www.aip.org/pacs
\end{itemize}

\subsection{Keywords}
Keywords should be provided for submissions to {\it Measurement Science and Technology, Physical Biology, Physiological Measurement}, and both parts of {\it Journal of Optics\/}. Add these as a new paragraph starting \verb"\noindent{\it Keywords\/}:" after the end of the abstract.

\subsection{Making a separate title page}
The command \verb"\maketitle" forces a page break after the point where it
is inserted and so to keep the header material on a separate page from the
body of the text insert \verb"\maketitle" or \verb"\newpage" before the start of the text. 
If \verb"\maketitle" is not included the text of the
article will start immediately after the abstract.

\subsection{Sample coding for the start of an article}
\label{startsample}
The code for the start of a title page of a typical paper might read:
\small\begin{verbatim}
\documentclass[12pt]{iopart}
\begin{document}
\title[The anomalous magnetic moment of the 
neutrino]{The anomalous magnetic moment of the 
neutrino and its relation to the solar neutrino problem}

\author{P J Smith$^1$, T M Collins$^2$, 
R J Jones$^3$\footnote{Present address:
Department of Physics, University of Bristol, Tyndalls Park Road, 
Bristol BS8 1TS, UK.} and Janet Williams$^3$}

\address{$^1$ Mathematics Faculty, Open University, 
Milton Keynes MK7~6AA, UK}
\address{$^2$ Department of Mathematics, 
Imperial College, Prince Consort Road, London SW7~2BZ, UK}
\address{$^3$ Department of Computer Science, 
University College London, Gower Street, London WC1E~6BT, UK}
\ead{williams@ucl.ac.uk}

\begin{abstract}
...
\end{abstract}
\pacs{1315, 9440T}
\submitto{\JPG}
\maketitle
\end{verbatim}
\normalsize

\section{The text}
\subsection{Sections, subsections and subsubsections}
The text of articles may be divided into sections, subsections and, where necessary, 
subsubsections. To start a new section, end the previous paragraph and 
then include \verb"\section" followed by the section heading within braces. 
Numbering of sections is done {\it automatically} in the headings: 
sections will be numbered 1, 2, 3, etc, subsections will be numbered 
2.1, 2.2,  3.1, etc, and subsubsections will be numbered 2.3.1, 2.3.2, 
etc.  Cross references to other sections in the text should, where
possible, be made using 
labels (see section~\ref{xrefs}) but can also
be made manually. See section~\ref{eqnum} for information on the numbering of displayed equations. Subsections and subsubsections are 
similar to sections but 
the commands are \verb"\subsection" and \verb"\subsubsection" respectively. 
Sections have a bold heading, subsections an italic heading and 
subsubsections an italic heading with the text following on directly.
\small\begin{verbatim}
\section{This is the section title}
\subsection{This is the subsection title}
\end{verbatim}\normalsize

The first section is normally an introduction,  which should state clearly 
the object of the work, its scope and the main advances reported, with 
brief references to relevant results by other workers. In long papers it is 
helpful to indicate the way in which the paper is arranged and the results 
presented.

Footnotes should be avoided whenever possible and can often be included in the text as phrases or sentences in parentheses. If required, they should be used only for brief notes that do not fit conveniently into the text. The use of 
displayed mathematics in footnotes should be avoided wherever possible and no equations within a footnote should be numbered. 
The standard \LaTeX\ macro \verb"\footnote" should be used.

\subsection{Acknowledgments}
Authors wishing to acknowledge assistance or encouragement from 
colleagues, special work by technical staff or financial support from 
organizations should do so in an unnumbered `Acknowledgments' section 
immediately following the last numbered section of the paper. The 
command \verb"\ack" sets the acknowledgments heading as an unnumbered
section. For Letters 
\verb"\ack" does not set a heading but leaves a line space and does not 
indent the next paragraph.

\subsection{Appendices}
Technical detail that it is necessary to include, but that interrupts 
the flow of the article, may be consigned to an appendix. 
Any appendices should be included at the end of the main text of the paper, after the acknowledgments section (if any) but before the reference list.
If there are 
two or more appendices they should be called Appendix A, Appendix B, etc. 
Numbered equations will be in the form (A.1), (A.2), etc,
figures will appear as figure A1, figure B1, etc and tables as table A1,
table B1, etc.

The command \verb"\appendix" is used to signify the start of the
appendices. Thereafter \verb"\section", \verb"\subsection", etc, will 
give headings appropriate for an appendix. To obtain a simple heading of 
`Appendix' use the code \verb"\section*{Appendix}". If it contains
numbered equations, figures or tables the command \verb"\appendix" should
precede it and \verb"\setcounter{section}{1}" must follow it.

\subsection{Some matters of style}
It will help the readers if your article is written in a clear,
consistent and concise manner. The Production Department at IOP
Publishing will try to make sure that your work is presented to its
readers in the best possible way without sacrificing the individuality of
your writing. Some recommended 
points to note, however, are the following.
\begin{enumerate}
\item Authors are often inconsistent in the use of `ize' and `ise' endings.
We recommend using `-ize' spellings (diagonalize, 
renormalization, minimization, etc) but there are some common 
exceptions to this, for example: devise, 
promise and advise.

\item English spellings are mandatory (colour, flavour, behaviour, tunnelling, artefact, focused, focusing, fibre, etc) except in the following journals where either American or English spellings are acceptable: {\it Physical Biology}, {\it Smart Materials and Structures} and {\it Journal of Micromechanics and Microengineering}.

\item The words table and figure should be written 
in full and {\bf not} abbreviaged to tab. and fig. Do not include `eq.'\, `equation' etc before an equation number or `ref.'\, `reference' etc before a reference number.
\end{enumerate}

Please check your article carefully for accuracy, consistency and clarity before
submission. Remember that your article will probably be read by many
people whose native language is not English and who may not therefore 
be aware of many of the subtle meanings of words or idiomatic phases
present in the English language. It therefore helps if you try to keep
sentences as short and simple as possible.

\section{Mathematics}
\subsection{Two-line constructions}
The great advantage of \LaTeX\ 
over other text processing systems is its 
ability to handle mathematics to almost any degree of complexity. However, 
in order to produce an article suitable for publication both within a print journal and online, 
authors should exercise some restraint on the constructions used. Some equations using very small characters which are clear in a preprint style article may be difficult read in a smaller format or online at a relatively low resolution.
For simple fractions in the text the solidus \verb"/", as in 
$\lambda/2\pi$, should be used instead of \verb"\frac" or \verb"\over", 
care being taken to use parentheses where necessary to avoid ambiguity, 
for example to distinguish between $1/(n-1)$ and $1/n-1$. Exceptions to 
this are the proper fractions $\frac12$, $\frac13$, $\frac34$, 
etc, which are better left in this form. In displayed equations 
horizontal lines are preferable to solidi provided the equation is 
kept within a height of two lines. A two-line solidus should be 
avoided where possible; the construction $(\ldots)^{-1}$ should be 
used instead; for example use:
\begin{equation*}
\frac{1}{M_{\rm a}}\left(\int^\infty_0{\rm d}
\omega\;\frac{|S_o|^2}{N}\right)^{-1}\qquad\mbox{instead of}\qquad
\frac{1}{M_{\rm a}}\biggl/\int^\infty_0{\rm d}
\omega\;\frac{|S_o|^2}{N}.
\end{equation*}

\subsection{Roman and italic in mathematics}
In mathematics mode \LaTeX\ automatically sets variables in an italic 
font. In most cases authors should accept this italicization. However, 
there are some cases where it is better to use a Roman font; for 
instance, IOP journals use a Roman d for a differential d, a Roman e 
for an exponential e and a Roman i for the square root of $-1$. To 
accommodate this and to simplify the  typing of equations we have 
provided some extra definitions. \verb"\rmd", \verb"\rme" and \verb"\rmi" 
now give Roman d, e and i respectively for use in equations, 
e.g.\ $\rmi x\rme^{2x}\rmd x/\rmd y$ 
is obtained by typing \verb"$\rmi x\rme^{2x}\rmd x/\rmd y$".

Certain other common mathematical functions, such as cos, sin, det and 
ker, should appear in Roman type. \LaTeX\ provides macros for 
most of these functions 
(in the cases above, \verb"\cos", \verb"\sin", \verb"\det" and \verb"\ker" 
respectively); we have also provided 
additional definitions for $\Tr$, $\tr$ and 
$\Or$ (\verb"\Tr", \verb"\tr" and \verb"\Or", respectively).

Subscripts and superscripts should be in Roman type if they are labels 
rather than variables or characters that take values. For example in the 
equation
\[
\epsilon_m=-g\mu_{\rm n}Bm
\]
$m$, the $z$ component of the nuclear spin, is italic because it can have 
different values whereas n is Roman because it 
is a label meaning nuclear ($\mu_{\rm n}$ 
is the nuclear magneton).

\subsection{Displayed equations in double-column journals}
Authors should bear in mind that all mathematical formulae in double-column journals will need to be fitted
into the width of a single column, so individual lines of equations should not occupy more than two thirds of the line width in this preprint form.

\subsection{Special characters for mathematics}
Bold italic characters can be used in our journals to signify vectors (rather
than using an upright bold or an over arrow). To obtain this effect
use \verb"\bi{#1}" within maths mode, e.g. $\bi{ABCdef}$. If upright 
bold characters are required in maths use \verb"\mathbf{#1}" within maths
mode, e.g. $\mathbf{XYZabc}$. The calligraphic (script) uppercase alphabet
is obtained with \verb"\mathcal{AB}" or \verb"\cal{CD}" 
($\mathcal{AB}\cal{CD}$).

The American Mathematical Society provides a series of extra symbol fonts
to use with \LaTeX\ and packages containing the character definitions to
use these fonts. Authors wishing to use Fraktur 
\ifiopams$\mathfrak{ABC}$ \fi
or Blackboard Bold \ifiopams$\mathbb{XYZ}$ \fi can include the appropriate
AMS package (e.g. amsgen, amsfonts, amsbsy, amssymb) with a 
\verb"\usepackage" command or add the command \verb"\usepackage{iopams}"
which loads the four AMS packages mentioned above and also provides
definitions for extra bold characters (all Greek letters and some
additional other symbols).

The package iopams uses the definition \verb"\boldsymbol" in amsbsy
which allows individual non-alphabetical symbols and Greek letters to be 
made bold within equations.
The bold Greek lowercase letters \ifiopams$\balpha \ldots\bomega$,\fi 
are obtained with the commands 
\verb"\balpha" \dots\ \verb"\bomega" (but note that
bold eta\ifiopams, $\bfeta$,\fi\ is \verb"\bfeta" rather than \verb"\beta")
and the capitals\ifiopams, $\bGamma\ldots\bOmega$,\fi\ with commands 
\verb"\bGamma" \dots\
\verb"\bOmega". Bold versions of the following symbols are
predefined in iopams: 
bold partial\ifiopams, $\bpartial$,\fi\ \verb"\bpartial",
bold `ell'\ifiopams, $\bell$,\fi\  \verb"\bell", 
bold imath\ifiopams, $\bimath$,\fi\  \verb"\bimath", 
bold jmath\ifiopams, $\bjmath$,\fi\  \verb"\bjmath", 
bold infinity\ifiopams, $\binfty$,\fi\ \verb"\binfty", 
bold nabla\ifiopams, $\bnabla$,\fi\ \verb"\bnabla", 
bold centred dot\ifiopams, $\bdot$,\fi\  \verb"\bdot", other 
characters are made bold using 
\verb"\boldsymbol{\symbolname}".

Please do not use the style file \verb"amsmath.sty" (part of the AMSTeX package) in conjunction with \verb"iopart.cls". This will result in several errors. To make use of the macros defined in \verb"amsmath.sty", we have provided the file \verb"setstack.sty" which reproduces the following useful macros from \verb"amsmath.sty":
\small\begin{verbatim}
\overset \underset \sideset \substack \boxed   \leftroot
\uproot  \dddot    \ddddot  \varrow   \harrow
\end{verbatim}\normalsize

Table~\ref{math-tab2} lists some other macros for use in 
mathematics with a brief description of their purpose.

\begin{table}
\caption{\label{math-tab2}Other macros defined in IOP macros for use in maths.}
\begin{tabular*}{\textwidth}{@{}l*{15}{@{\extracolsep{0pt plus
12pt}}l}}
\br
Macro&Result&Description\\
\mr
\verb"\fl"&&Start line of equation full left\\
\verb"\case{#1}{#2}"&$\case{\#1}{\#2}$&Text style fraction in display\\
\verb"\Tr"&$\Tr$&Roman Tr (Trace)\\
\verb"\tr"&$\tr$&Roman tr (trace)\\
\verb"\Or"&$\Or$&Roman O (of order of)\\
\verb"\tdot{#1}"&$\tdot{x}$&Triple dot over character\\
\verb"\lshad"&$\lshad$&Text size left shadow bracket\\
\verb"\rshad"&$\rshad$&Text size right shadow bracket\\
\br
\end{tabular*}
\end{table}

\subsection{Alignment of displayed equations}
The normal style for aligning displayed equations is to align them left rather than centre them. The class file automatically does this and indents each line of a display. To make any line start at the left margin of the page, add \verb"\fl" at its start (to indicate full left).

Using the \verb"eqnarray" environment equations will naturally be aligned left without the use of any ampersands for alignment, see equations (\ref{eq1}) and (\ref{eq2})
\begin{eqnarray}
\alpha + \beta =\gamma^2 \label{eq1}\\
\alpha^2 + 2\gamma + \cos\theta = \delta \label{eq2} 
\end{eqnarray}
This is the normal equation style for our journals.

Where some secondary alignment is needed, for instance a second part of an equation on a second line, a single ampersand is added at the point of alignment in each line  (see  (\ref{eq3}) and (\ref{eq4})).
\begin{eqnarray}
\alpha &=2\gamma^2 + \cos\theta + \frac{XY \sin\theta}{X+ Y\cos\theta} \label{eq3}\\
 & = \delta\theta PQ \cos\gamma. \label{eq4} 
\end{eqnarray}

Two points of alignment are possible using two ampersands for alignment (see  (\ref{eq5}) and (\ref{eq6})).  Note in this case extra space \verb"\qquad" is added before the second ampersand in the longest line (the top one) to separate the condition from the equation. 
\begin{eqnarray}
\alpha &=2\gamma^2 + \cos\theta + \frac{XY \sin\theta}{X+ Y\cos\theta}\qquad& \theta > 1 \label{eq5}\\
 & = \delta\theta PQ \cos\gamma &\theta \leq 1\label{eq6} 
\end{eqnarray}

For a long equation which has to be split over more than one line the first line should start at the left margin, this is achieved by inserting \verb"\fl" (full left) at the start of the line, the use of the alignment parameter \verb"&" is not necessary unless some secondary alignment is needed.
\begin{eqnarray}
\fl \alpha + 2\gamma^2 = \cos\theta + \frac{XY \sin\theta}{X+ Y\cos\theta} +  \frac{XY \sin\theta}{X- Y\cos\theta} +
+ \left(\frac{XY \sin\theta}{X+ Y\cos\theta}\right)^2 \nonumber\\
+  \left(\frac{XY \sin\theta}{X- Y\cos\theta}\right)^2\label{eq7} 
\end{eqnarray}

The Plain \TeX\ command \verb"\eqalign" can be used within an \verb"equation" environment to obtain a multiline equation with a single centred number, for example
\begin{equation}
\eqalign{\alpha + \beta =\gamma^2 \cr
\alpha^2 + 2\gamma + \cos\theta = \delta.}\label{eq2} 
\end{equation}

\subsection{Miscellaneous points}
Exponential expressions, especially those containing subscripts or 
superscripts, are clearer if the notation $\exp(\ldots)$ is used except for 
simple examples. For instance $\exp[\rmi(kx-\omega t)]$ and $\exp(z^2)$ are 
preferred to $\e^{\rmi(kx-\omega t)}$ and $\e^{z^2}$, but 
$\e^x$ 
is acceptable. Similarly the square root sign $\sqrt{\phantom{b}}$ should 
only be used with relatively
simple expressions, e.g.\ $\sqrt2$ and $\sqrt{a^2+b^2}$;
in other cases the 
power $1/2$ should be used; for example, $[(x^2+y^2)/xy(x-y)]^{1/2}$.

It is important to distinguish between $\ln = \log_\e$ and $\lg 
=\log_{10}$. Braces, brackets and parentheses should be used in the 
following order: $\{[(\;)]\}$. The same ordering of brackets should be 
used within each size. However, this ordering can be ignored if the
brackets have a 
special meaning (e.g.\ if they denote an average or a function).  Decimal 
fractions should 
always be preceded by a zero: for example 0.123 {\bf not} .123. For long 
numbers commas are not inserted but instead a thin space is added after 
every third character away from the position of the decimal point unless 
this leaves a single separated character: e.g.\ $60\,000$, $0.123\,456\,78$ 
but 4321 and 0.7325.

Equations should be followed by a full stop (periods) when at the end
of a sentence.

\subsection{Equation numbering}
\label{eqnum}
\LaTeX\ provides facilities for automatically numbering equations 
and these should be used where possible. Sequential numbering (1), (2), 
etc, is the default numbering system although, if the command
\verb"\eqnobysec" is included in the preamble, equation numbering
by section is obtained, e.g.\ 
(2.1), (2.2), etc. In articles with several appendices equation numbering
by section is useful in the appendices even when sequential numbering has
been used throughout the main body of the text and is switched on by the 
\verb"\appendix" command. Equation numbering by section {\it must} 
be used for {\it Reports on Progress in 
Physics}. When referring to an equation in the text it is not normally necessary to include the word equation before the number, 
which should be parentheses. Do not 
use abbreviations such as eqn or eq.
When cross-referencing is used, \verb"(\ref{<label>})"
 will produce `(\verb"<eqnum>")',
\verb"\eref{<label>}" produces `(\verb"<eqnum>")' and
\verb"\Eref{<label>}" produces `Equation (\verb"<eqnum>")', 
where \verb"<label>" is the label to produce equation number \verb"<eqnum>".

Sometimes it is useful to number equations as parts of the same
basic equation. This can be accomplished by inserting the 
commands \verb"\numparts" before the equations concerned and 
\verb"\endnumparts" when reverting to the normal sequential numbering.
The equations below show the previous equations numbered as separate parts
using \verb"\numparts ... \endnumparts" and the \verb"eqnarray"
environment
\numparts
\begin{eqnarray}
T_{11}&=(1+P_\e)I_{\uparrow\uparrow}-(1-P_\e)
I_{\uparrow\downarrow}\\
T_{-1-1}&=(1+P_\e)I_{\downarrow\downarrow}-(1-P_\e)I_{\uparrow\downarrow}\\
S_{11}&=(3+P_\e)I_{\downarrow\uparrow}-(3-P_e)I_{\uparrow\uparrow}\\
S_{-1-1}&=(3+P_\e)I_{\uparrow\downarrow}-(3-P_\e)
I_{\downarrow\downarrow}
\end{eqnarray}
\endnumparts

\subsection{Miscellaneous extra commands for displayed equations}
The \verb"\cases" command of Plain \TeX\ is available
for use with \LaTeX\ but has been amended slightly to 
increase the space between the equation and the condition. 
\Eref{cases} 
demonstrates simply the output from the \verb"\cases" command
\begin{equation}
\label{cases}
X=\cases{1&for $x \ge 0$\\
-1&for $x<0$\\}
\end{equation}
%The code used was:
%\small\begin{verbatim}
%\begin{equation}
%\label{cases}
%X=\cases{1&for $x \ge 0$\\
%-1&for $x<0$\\}
%\end{equation}
%\end{verbatim}
%\normalsize

To obtain text style fractions within displayed maths the command 
\verb"\case{#1}{#2}" can be used (see equations (2) and (5)) instead
of the usual \verb"\frac{#1}{#2}" command or \verb"{#1 \over #2}".

When two or more short equations are on the same line they should be 
separated by a `qquad space' (\verb"\qquad"), rather than
\verb"\quad" or any combination of \verb"\,", \verb"\>", \verb"\;" 
and \verb"\ ".

\section{Referencing\label{except}}
Two different styles of referencing are in common use: 
the Harvard alphabetical system and the Vancouver numerical system. 
All IOP journals allow the use of the Harvard or Vancouver system 
(authors should use the style they are comfortable with), 
apart from the following journals

\begin{itemize}
\item {\it Physics in Medicine and Biology} 
\item {\it Physiological Measurement}
\end{itemize}

\noindent for which authors must use the Harvard referencing style.

\subsection{Alphabetical (Harvard) system}
In the Harvard system the name of the author appears in the text together 
with the year of publication. As appropriate, either the date or the name 
and date are included within parentheses. Where there are only two authors 
both names should be given in the text; if there are more than two 
authors only the first name should appear followed by `{\it et al}' 
(which can be obtained by typing \verb"\etal"). When two or 
more references to work by one author or group of authors occur for the 
same year they should be identified by including a, b, etc after the date 
(e.g.\ 1986a). If several references to different pages of the same article 
occur the appropriate page number may be given in the text, e.g.\ Kitchen 
(1982, p 39).

The reference list at the end of an article consists of an 
unnumbered `References' section containing an
alphabetical listing by authors' names and in date order for each 
author or group of identical authors. The reference list in the 
preprint style is started by including the command \verb"\section*{References}" and then
\verb"\begin{harvard}".
Individual references start with \verb"\item[]" and the reference list is completed with \verb"\end{harvard}".
There is also a shortened form of the coding; \verb"\section*{References}"
and \verb"\begin{harvard}" can be replaced by the single command
\verb"\References" and \verb"\end{harvard}" can be shortened to
\verb"\endrefs".

\subsection{Numerical (Vancouver) system}
In the numerical system references are numbered sequentially 
throughout the text. The numbers occur within square brackets and one 
number can be used to designate several references. A numerical 
reference list in the \verb"iopart" style is started by including the 
command \verb"\section*{References}" and then
\verb"\begin{thebibliography}{<num>}", where \verb"<num>" is the largest
number in the reference list (or any other number with the same number
of digits).  The 
reference list gives the references in 
numerical, not alphabetical, order and is completed by
\verb"\end{thebibliography}". Short forms of the commands are again
available: \verb"\Bibliography{<num>}" can be used at the start of the
references section and \verb"\endbib" at the end.

A variant of this system is to use labels instead of numbers within 
square brackets, this method is allowed although not recommended.

\subsection{BiBTeX\label{bibtex}}
Note that for the journals lised at the start of section {\ref{except}} authors should use the
appropriate \verb".bst" file to produce references in the correct style---see section \ref{bibby} for
a list of reccommended \verb".bst" files.

\subsection{References, general}
A complete reference should provide the reader with enough information to locate
the article concerned. Titles of journal articles can also be included and are
required for {\it Inverse problems, Journal of Neural Engineering, Measurement
Science and Technology, Physical Biology, Physics in Medicine and Biology\/} and
{\it Physiological Measurement}. Final page numbers of references are required
for {\it Reports on Progress in Physics\/} and {\it Physiological Measurement}.

Up to ten authors may be given in a particular reference; where 
there are more than ten only the first should be given followed by 
`{\it et al}'. If an author is unsure of an abbreviation and the 
journal is not given in appendix B, it is best to leave the title in full.

The terms {\it loc.\ cit.}\ and {\it ibid}.\ should not be used. 
Unpublished conferences and reports should generally not be included 
in the reference list if a published version of the work exists. Articles in the course of publication should 
include the article title and the journal of publication, if known. 
A reference to a thesis submitted for a higher degree may be included 
if it has not been superseded by a published 
paper but the institution where the work was done should be included.

The basic structure of a reference in the reference list is the same in both the alphabetical and numerical systems, the only difference the code at the start of the reference. Alphabetic references are preceded by \verb"\item[]", numeric by \verb"\bibitem{label}" or just \verb"\item" to generate a number or \verb"\nonum" where a reference is not the first in a group of references under the same number.

Cross referencing between the text and the 
reference list is most useful in the numeric system but is not necessary for alphabetic referencing 
in the Harvard system as adding or deleting a reference 
does not normally change any of the other references. 
Care should, however, be taken to make sure that the text citations match the reference list and vice versa.

Note that footnotes to the text should not be 
included in reference list, but should appear at the bottom of the relevant page by using the \verb"\footnote" command.

\subsection{References to journal articles}
A normal reference to a journal article contains three changes of 
font:
the authors and date appear in Roman type, the journal title in 
italic, the volume number in bold and the page numbers in Roman again. 
A typical journal entry would be:

\smallskip
\begin{harvard}
\item[] Cisneros A 1971 {\it Astrophys.\ Space Sci.} {\bf 10} 87
\end{harvard}
\smallskip

\noindent which would be obtained by typing, within the references
environment 
\small\begin{verbatim}
\item[] Cisneros A 1971 {\it Astrophys. Space Sci.} {\bf 10} 87
\end{verbatim}\normalsize

Features to note are the following.

\begin{enumerate}
\item The authors should be in the form surname (with only the first 
letter capitalized) {\bf followed} by the initials with {\bf no} 
periods after the initials. Authors should be separated by a comma 
except for the last two which should be separated by `and' with no 
comma preceding it. For journals that accept titles of articles in the 
reference list,  the title should be in Roman (upright)
lower case letters, except for an initial 
capital, and should follow the date.

\item The year of publication follows the authors and is not in parentheses. An article title (in Roman) would follow the year.

\item The journal is in italic and is abbreviated. \ref{jlabs} 
gives a list of macros that will give the correct abbreviation for 
many of the common journals. If a journal has several parts denoted by 
different letters the part letter should be inserted after the journal in Roman type, e.g.\ 
{\it Phys.\ Rev.\ \rm A}.

\item The volume number is bold; the page number is Roman.
 Both the initial and final page numbers should be given where possible. The final page number should be in 
the shortest possible form and separated from the initial page number by an en rule (\verb"--"), e.g.\ 1203--14.

\item Where there are two or more references with identical authors, 
the authors' names should be repeated for the second and subsequent references. Each individual publication should be presented as a separate reference, although in the numerical system one number can be used for several references. This facilitates linking in the electronic version of the journal. 
\end{enumerate}
\subsection{Electronic journal references}
These often do not always follow the conventional {\it year-journal-volume-page numbers} pattern. Some examples are:

\numrefs{1}
\item Carlip S and Vera R 1998 {\it Phys. Rev.} D {\bf 58} 011345 
\item Davies K and Brown G 1997 {\it J. High Energy Phys.} JHEP12(1997)002
\item Hannestad S 2005 {\it J. Cosmol. Astropart. Phys.} JCAP02(2005)011
\item Hilhorst H J 2005 {\it J. Stat. Mech.} L02003
\item Gundlach C 1999 {\it Liv. Rev. Rel.} 1994-4
\endnumrefs

\subsection{Preprint references}
Preprint may be referenced but if the article concerned has been published in a peer-reviewed journal, that reference should take precedence. If only a preprint reference can be given, it is helpful to include the article title. Examples are:
\vskip6pt
\numrefs{1}
\item Neilson D and Choptuik M 2000 {\it Class. Quantum Grav.} {\bf 17} 761 ({\it Preprint} gr-qc/9812053)
\item Harrison M 1999 Dipheomorphism-invariant manifolds {\it Preprint} hep-th/9909196
\endnumrefs

\subsection{References to books, conference proceedings and reports}

References to books, proceedings and reports are similar, but have only two
changes of font. The authors and date of publication are in Roman, the 
title of the book is in italic, and the editors, publisher, 
town of publication 
and page number are in Roman. A typical reference to a book and a
conference paper might be

\smallskip
\begin{harvard}
\item[] Dorman L I 1975 {\it Variations of Galactic Cosmic Rays} 
(Moscow: Moscow State University Press) p~103
\item[] Caplar R and Kulisic P 1973 {\it Proc.\
Int.\ Conf.\ on Nuclear Physics (Munich)} vol~1 (Amsterdam:  
North-Holland/American Elsevier) p~517
\end{harvard}
\smallskip

\noindent which would be obtained by typing
\small\begin{verbatim}
\item[] Dorman L I 1975 {\it Variations of Galactic Cosmic Rays} 
(Moscow: Moscow State University Press) p~103
\item[] Caplar R and Kulisic P 1973 {\it Proc. Int. Conf. on Nuclear 
Physics (Munich)} vol~1 (Amsterdam: North-Holland/American 
Elsevier) p~517
\end{verbatim}\normalsize
\noindent respectively.

Features to note are the following.
\begin{enumerate}
\item Book titles are in italic and should be spelt out in full with 
initial capital letters for all except minor words. Words such as 
Proceedings, Symposium, International, Conference, Second, etc should 
be abbreviated to Proc., Symp., Int., Conf., 2nd, 
respectively, but the rest of the title should be given in full, 
followed by the date of the conference and the 
town or city where the conference was held. For 
Laboratory Reports the Laboratory should be spelt out wherever 
possible, e.g.\ {\it Argonne National Laboratory Report}.

\item The volume number as, for example, vol~2, should be followed by 
the editors, if any, in a form such as ed~A~J~Smith and P~R~Jones. Use 
\etal if there are more than two editors. Next comes the town of 
publication and publisher, within brackets and separated by a colon, 
and finally the page numbers preceded by p if only one number is given 
or pp if both the initial and final numbers are given.

\item If a book is part of a series (for examples, {\it Springer Tracts in Modern Physics\/}), the series title and volume number is given in parentheses after the book title. Whereas for an individual volume in a multivolume set, the set title is given first, then the volume title. 
\end{enumerate}
\smallskip
\begin{harvard}
\item[]Morse M 1996 Supersonic beam sources {\it Atomic Molecular and Optical Physics\/} ({\it Experimental Methods in the Physical Sciences\/} vol 29) ed F B Dunning and R Hulet (San Diego: Academic)  
\item[]Fulco C E, Liverman C T and Sox H C (eds) 2000 {\it Gulf War and Health\/} vol 1 {\it Depleted Uranium, Pyridostigmine Bromide, Sarin, and Vaccines\/} (Washington, DC: The National Academies Press)
\end{harvard}

\section{Cross referencing\label{xrefs}}
The facility to cross reference items in the text is very useful when 
composing articles as the precise form of the article may be uncertain at the start 
and  revisions and amendments may subsequently be made. 
\LaTeX\ provides excellent facilities for doing cross-referencing
and these can be very useful in preparing articles.

\subsection{References}
\label{refs}
Cross referencing is useful for numeric reference lists because, if it 
is used, adding 
another reference to the list does not then involve renumbering all 
subsequent references. It is not necessary for referencing 
in the Harvard system where the final reference list is alphabetical 
and normally no other changes are necessary when a reference is added or
deleted.
Two passes are necessary initially to get the cross references right 
but once they are correct a single run is usually sufficient provided an 
\verb".aux" file is available and the file 
is run to the end each time.
\verb"\cite" and \verb"\bibitem" are used to link citations in the text
with their entry in the reference list;
if the 
reference list contains an entry \verb"\bibitem{label}", 
then \verb"\bibitem{label}" 
will produce the correct number in the reference list and 
\verb"\cite{label}" will produce the number within square brackets in the 
text. \verb"label" may contain alphabetic letters, 
or punctuation characters but must not contain spaces or commas. It is also
recommended that the underscore character \_{} is not used in cross
referencing. 
Thus labels for the form 
\verb"eq:partial", \verb"fig:run1", \verb"eq:dy'", 
etc, may be used. When several 
references occur together in the text \verb"\cite" may be used with 
multiple labels with commas but no spaces separating them; 
the output will be the 
numbers within a single pair of square brackets with a comma and a 
thin space separating the numbers. Thus \verb"\cite{label1,label2,label4}"
would give [1,\,2,\,4]. Note that no attempt is made by the style file to sort the 
labels and no shortening of groups of consecutive numbers is done.
Authors should therefore try to use multiple labels in the correct 
order.

The numbers for the cross referencing are generated in the order the 
references appear in the reference list, so that if the entries in the 
list are not in the order in which the references appear in the text 
then the 
numbering within the text will not be sequential. To correct this 
change the ordering of the entries in the reference list and then 
rerun {\it twice}.

\subsection{Equation numbers, sections, subsections, figures and 
tables}
Labels for equation numbers, sections, subsections, figures and tables 
are all defined with the \verb"\label{label}" command and cross references 
to them are made with the \verb"\ref{label}" command.

Any section, subsection, subsubsection, appendix or subappendix 
command defines a section type label, e.g. 1, 2.2, A2, A1.2 depending 
on context. A typical article might have in the code of its introduction 
`The results are discussed in section\verb"~\ref{disc}".' and
the heading for the discussion section would be:
\small\begin{verbatim}
\section{Results}\label{disc}
\end{verbatim}\normalsize
Labels to sections, etc, may occur anywhere within that section except
within another numbered environment. 
Within a maths environment labels can be used to tag equations which are 
referred to within the text.

In addition to the standard \verb"\ref{<label>}" the abbreviated
forms given in the \tref{abrefs}
are available for reference to standard parts of the text

\Table{\label{abrefs}Alternatives to the normal references $\backslash${\tt ref} 
and the text generated by
them. Note it is not normally necessary to include the word equation
before an equation number except where the number starts a sentence. The
versions producing an initial capital should only be used at the start of
sentences.} 
\br
Reference&Text produced\\
\mr
\verb"\eref{<label>}"&(\verb"<num>")\\
\verb"\Eref{<label>}"&Equation (\verb"<num>")\\
\verb"\fref{<label>}"&figure \verb"<num>"\\
\verb"\Fref{<label>}"&Figure \verb"<num>"\\
\verb"\sref{<label>}"&section \verb"<num>"\\
\verb"\Sref{<label>}"&Section \verb"<num>"\\
\verb"\tref{<label>}"&table \verb"<num>"\\
\verb"\Tref{<label>}"&Table \verb"<num>"\\
\br
\endTable

\section{Tables and table captions}
Tables are numbered serially and referred to in the text 
by number (table 1, etc, {\bf not} tab. 1). Each table should have an 
explanatory caption which should be as concise as possible. If a table 
is divided into parts these should be labelled \pt(a), \pt(b), 
\pt(c), etc but there should be only one caption for the whole 
table, not separate ones for each part.

In the preprint style the tables may be included in the text 
or listed separately after the reference list 
starting on a new page.

\subsection{The basic table format}
The standard form for a table is:
\small\begin{verbatim}
\begin{table}
\caption{\label{label}Table caption.}
\begin{indented}
\item[]\begin{tabular}{@{}llll}
\br
Head 1&Head 2&Head 3&Head 4\\
\mr
1.1&1.2&1.3&1.4\\
2.1&2.2&2.3&2.4\\
\br
\end{tabular}
\end{indented}
\end{table}
\end{verbatim}\normalsize

Points to note are:
\begin{enumerate}
\item The caption comes before the table. It should have a period at
the end.

\item Tables are normally set in a smaller type than the text.
The normal style is for tables to be indented. This is accomplished
by using \verb"\begin{indented}" \dots\ \verb"\end{indented}"
and putting \verb"\item[]" before the start of the tabular environment.
Omit these
commands for any tables which will not fit on the page when indented.

\item The default is for columns to be aligned left and 
adding \verb"@{}" omits the extra space before the first column.

\item Tables have only horizontal rules and no vertical ones. The rules at
the top and bottom are thicker than internal rules and are set with
\verb"\br" (bold rule). 
The rule separating the headings from the entries is set with
\verb"\mr" (medium rule).

\item Numbers in columns should be aligned on the decimal point;
to help do this a control sequence \verb"\lineup" has been defined 
which sets \verb"\0" equal to a space the size of a digit, \verb"\m"
to be a space the width of a minus sign, and \verb"\-" to be a left
overlapping minus sign. \verb"\-" is for use in text mode while the other
two commands may be used in maths or text.
(\verb"\lineup" should only be used within a table
environment after the caption so that \verb"\-" has its normal meaning
elsewhere.) See table~\ref{tabone} for an example of a table where
\verb"\lineup" has been used.
\end{enumerate}

\begin{table}
\caption{\label{tabone}A simple example produced using the standard table commands 
and $\backslash${\tt lineup} to assist in aligning columns on the 
decimal point. The width of the 
table and rules is set automatically by the 
preamble.}

\begin{indented}
\lineup
\item[]\begin{tabular}{@{}*{7}{l}}
\br                              
$\0\0A$&$B$&$C$&\m$D$&\m$E$&$F$&$\0G$\cr 
\mr
\0\023.5&60  &0.53&$-20.2$&$-0.22$ &\01.7&\014.5\cr
\0\039.7&\-60&0.74&$-51.9$&$-0.208$&47.2 &146\cr 
\0123.7 &\00 &0.75&$-57.2$&\m---   &---  &---\cr 
3241.56 &60  &0.60&$-48.1$&$-0.29$ &41   &\015\cr 
\br
\end{tabular}
\end{indented}
\end{table}

\subsection{Simplified coding and extra features for tables}
The basic coding format can be simplified using extra commands provided in
the \verb"iopart" class file. The commands up to and including 
the start of the tabular environment
can be replaced by
\small\begin{verbatim}
\Table{\label{label}Table caption}
\end{verbatim}\normalsize
this also activates the definitions within \verb"\lineup".
The final three lines can also be reduced to \verb"\endTable" or
\verb"\endtab". Similarly for a table which does not fit in when indented
\verb"\fulltable{\label{label}caption}" \dots\ \verb"\endfulltable" or \verb"\endtab"
can be used. \LaTeX\ optional positional parameters can, if desired, be added after 
\verb"\Table{\label{label}caption}" and \verb"\fulltable{\label{label}caption}".

\verb"\centre{#1}{#2}" can be used to centre a heading 
\verb"#2" over \verb"#1" 
columns and \verb"\crule{#1}" puts a rule across 
\verb"#1" columns. A negative 
space \verb"\ns" is usually useful to reduce the space between a centred 
heading and a centred rule. \verb"\ns" should occur immediately after the 
\verb"\\" of the row containing the centred heading (see code for
\tref{tabl3}). A small space can be 
inserted between rows of the table 
with \verb"\ms" and a half line space with \verb"\bs" 
(both must follow a \verb"\\" but should not have a 
\verb"\\" following them).

\Table{\label{tabl3}A table with headings spanning two columns and containing notes. 
To improve the 
visual effect a negative skip ($\backslash${\tt ns})
has been put in between the lines of the 
headings. Commands set-up by $\backslash${\tt lineup} are used to aid 
alignment in columns. $\backslash${\tt lineup} is defined within
the $\backslash${\tt Table} definition.}
\br
&&&\centre{2}{Separation energies}\\
\ns
&Thickness&&\crule{2}\\
Nucleus&(mg\,cm$^{-2}$)&Composition&$\gamma$, n (MeV)&$\gamma$, 2n (MeV)\\
\mr
$^{181}$Ta&$19.3\0\pm 0.1^{\rm a}$&Natural&7.6&14.2\\
$^{208}$Pb&$\03.8\0\pm 0.8^{\rm b}$&99\%\ enriched&7.4&14.1\\
$^{209}$Bi&$\02.86\pm 0.01^{\rm b}$&Natural&7.5&14.4\\
\br
\end{tabular}
\item[] $^{\rm a}$ Self-supporting.
\item[] $^{\rm b}$ Deposited over Al backing.
\end{indented}
\end{table}

Units should not normally be given within the body of a table but 
given in brackets following the column heading; however, they can be 
included in the caption for long column headings or complicated units. 
Where possible tables should not be broken over pages. 
If a table has related notes these should appear directly below the table
rather than at the bottom of the page. Notes can be designated with
footnote symbols (preferable when there are only a few notes) or
superscripted small roman letters. The notes are set to the same width as
the table and in normal tables follow after \verb"\end{tabular}", each
note preceded by \verb"\item[]". For a full width table \verb"\noindent"
should precede the note rather than \verb"\item[]". To simplify the coding 
\verb"\tabnotes" can, if desired, replace \verb"\end{tabular}" and 
\verb"\endtabnotes" replaces
\verb"\end{indented}\end{table}".

If all the tables are grouped at the end of a document
the command \verb"\Tables" is used to start a new page and 
set a heading `Tables and table captions'. If the tables follow an appendix then add the command \verb"\noappendix" to revert to normal style numbering.

\section{Figures and figure captions}

Figures (with their captions) can be incorporated into the text at the appropriate position or grouped together
at the end of the article. If the figures are at the end of the article and follow an appendix then add the command \verb"\noappendix" to revert to normal style numbering.

\subsection{Using copyright material}
If you wish to illustrate your paper using material for which you 
do not own the copyright then you must seek permission from 
the copyright holder, usually both the author and the publisher of the previous work. It is the author's responsibility to obtain 
copyright permissions and this should be done
prior to submitting your paper. 
If you have obtained permission, please provide full details of 
the permission granted---for example, copies of the text of any e-mails 
or a copy of any letters you may have received. Figure captions must include an acknowledgment of the original source of the material even when permission to use has been obtained.

\subsection{Inclusion of graphics files\label{figinc}}
Using the \verb"graphicx" package graphics files can 
be included within figure and center environments at an 
appropriate point within the text using code such as:
\small\begin{verbatim}
\includegraphics{file.eps}
\end{verbatim}\normalsize
The \verb"graphicx" package supports various optional arguments
to control the appearance of the figure. Other similar 
packages can also be used (e.g. \verb"graphics", \verb"epsf"). 
For more detail about graphics inclusion see the documentation 
of the \verb"graphicx" package, refer to one of the books on \LaTeX\ {\cite{book1}}
or download some of the excellent free documentation available via the Comprehensive
TeX Archive Network (CTAN) http://www.ctan.org---{in particular see \cite{eps}}.
IOP's graphics guidelines, \gguide, provide further information on preparing EPS files---a copy
should have accompanied this document but it may be download from authors.iop.org.

\subsection{Captions}
Each figure should have a brief caption describing it and, if 
necessary, interpreting the various lines and symbols on the figure. 
As much lettering as possible should be removed from the figure itself 
and included in the caption. If a figure has parts, these should be 
labelled ($a$), ($b$), ($c$), etc and all parts should be described 
within a single caption. \Tref{blobs} gives the definitions for describing 
symbols and lines often used within figure captions (more symbols are 
available when using the optional packages loading the AMS extension fonts).

\subsection{Supplementary Data}
All of our journals encourage authors to submit supplementary data attachments to 
enhance the online versions of published research articles. Supplementary data 
enhancements typically consist of video clips, animations or
data files, tables of extra information or extra figures. They can 
add to the reader's understanding and present results in attractive ways that go 
beyond what can be presented in the print version of the journal. 
The printed journal remains the archival version, and supplementary data items are 
supplements which enhance a reader's understanding of the paper but are not 
essential to that understanding. For electronic-only journals, supplementary data attachments 
may be used to convey essential information. Guidelines on supplementary data file formats 
are contained in the document {\it Guidelines for authors} which can be accessed 
online by going to authors.iop.org.

\begin{table}[t]
\caption{\label{blobs}Control sequences to describe lines and symbols in figure 
captions.}
\begin{indented}
\item[]\begin{tabular}{@{}lllll}
\br
Control sequence&Output&&Control sequence&Output\\
\mr
\verb"\dotted"&\dotted        &&\verb"\opencircle"&\opencircle\\
\verb"\dashed"&\dashed        &&\verb"\opentriangle"&\opentriangle\\
\verb"\broken"&\broken&&\verb"\opentriangledown"&\opentriangledown\\
\verb"\longbroken"&\longbroken&&\verb"\fullsquare"&\fullsquare\\
\verb"\chain"&\chain          &&\verb"\opensquare"&\opensquare\\
\verb"\dashddot"&\dashddot    &&\verb"\fullcircle"&\fullcircle\\
\verb"\full"&\full            &&\verb"\opendiamond"&\opendiamond\\
\br
\end{tabular}
\end{indented}
\end{table}

\clearpage

\appendix
\section{List of macros for formatting text, figures and tables}

\begin{table}[hb]
\caption{Macros available for use in text. Parameters in square brackets are optional.}
\footnotesize\rm
\begin{tabular}{@{}*{7}{l}}
\br
Macro name&Purpose\\
\mr
\verb"\title[#1]{#2}"&Title of article and short title (optional)\\
\verb"\paper[#1]{#2}"&Title of paper and short title (optional)\\
\verb"\letter{#1}"&Title of Letter to the Editor\\
\verb"\comment[#1]{#2}"&Title of Comment and short title (optional)\\
\verb"\topical[#1]{#2}"&Title of Topical Review and short title 
(optional)\\
\verb"\review[#1]{#2}"&Title of review article and short title (optional)\\
\verb"\note[#1]{#2}"&Title of Note and short title (optional)\\
\verb"\prelim[#1]{#2}"&Title of Preliminary Communication \& short title\\
\verb"\author{#1}"&List of all authors\\
\verb"\article[#1]{#2}{#3}"&Type and title of other articles and 
short title (optional)\\
\verb"\address{#1}"&Address of author\\
\verb"\pacs{#1}"&PACS classification codes\\
\verb"\pacno{#1}"&Single PACS classification code\\
\verb"\ams{#1}"&Mathematics Classification Scheme\\
\verb"\submitto{#1}"&`Submitted to' message\\
\verb"\maketitle"&Creates title page\\
\verb"\begin{abstract}"&Start of abstract\\
\verb"\end{abstract}"&End of abstract\\
\verb"\nosections"&Inserts space before text when no sections\\
\verb"\section{#1}"&Section heading\\
\verb"\subsection{#1}"&Subsection heading\\
\verb"\subsubsection{#1}"&Subsubsection heading\\
\verb"\appendix"&Start of appendixes\\
\verb"\ack"&Acknowledgments heading\\
\verb"\References"&Heading for reference list\\
\verb"\begin{harvard}"&Start of alphabetic reference list\\
\verb"\end{harvard}"&End of alphabetic reference list\\
\verb"\begin{thebibliography}{#1}"&Start of numeric reference list\\
\verb"\end{thebibliography}"&End of numeric reference list\\
\verb"\etal"&\etal for text and reference lists\\
\verb"\nonum"&Unnumbered entry in numerical reference list\\
\br
\end{tabular}
\end{table}

\begin{table}
\caption{Macros defined within {\tt iopart.cls}
for use with figures and tables.}
\begin{indented}
\item[]\begin{tabular}{@{}l*{15}{l}}
\br
Macro name&Purpose\\
\mr
\verb"\Figures"&Heading for list of figure captions\\
\verb"\Figure{#1}"&Figure caption\\
\verb"\Tables"&Heading for tables and table captions\\
\verb"\Table{#1}"&Table caption\\
\verb"\fulltable{#1}"&Table caption for full width table\\
\verb"\endTable"&End of table created with \verb"\Table"\\
\verb"\endfulltab"&End of table created with \verb"\fulltable"\\
\verb"\endtab"&End of table\\
\verb"\br"&Bold rule for tables\\
\verb"\mr"&Medium rule for tables\\
\verb"\ns"&Small negative space for use in table\\
\verb"\centre{#1}{#2}"&Centre heading over columns\\
\verb"\crule{#1}"&Centre rule over columns\\
\verb"\lineup"&Set macros for alignment in columns\\
\verb"\m"&Space equal to width of minus sign\\
\verb"\-"&Left overhanging minus sign\\
\verb"\0"&Space equal to width of a digit\\
\br
\end{tabular}
\end{indented}
\end{table}

\clearpage
\section{Control sequences for journal
abbreviations}
\label{jlabs}
\begin{table}[hb]
\caption{\label{jlab1}Abbreviations for the IOP journals.}
\begin{indented}
\item[]
\begin{tabular}{@{}lll}
\br
Macro name&{\rm Short form of journal title}&Years relevant\\
\mr
{\it Current journals}\\
\mr
\verb"\CQG"&Class. Quantum Grav.\\
\verb"\EJP"&Eur. J. Phys.\\
\verb"\IP"&Inverse Problems\\
\verb"\JHEP"&J. High Energy Phys.&1999 and onwards\\
\verb"\JMM"&J. of Michromech. and Microeng.\\
\verb"\JNE"&J. Neural Eng.\\
\verb"\JOA"&J. Opt. A: Pure and Applied Opt.&1998 and onwards\\
\verb"\JOB"&J. Opt. B: Quantum and Semiclass. Opt.&1999 and onwards\\
\verb"\JPA"&J. Phys. A: Math. Gen.\\
\verb"\jpb"&J. Phys. B: At. Mol. Opt. Phys.&1988 and onwards\\
\verb"\JPCM"&J. Phys: Condens. Matter&1989 and onwards\\
\verb"\JPD"&J. Phys. D: Appl. Phys.\\
\verb"\jpg"&J. Phys. G: Nucl. Part. Phys.&1989 and onwards\\
\verb"\MSMSE"&Modelling Simul. Mater. Sci. Eng.\\
\verb"\MST"&Meas. Sci. Technol.&1990 and onwards\\
\verb"\NJP"&New J. Phys.&1999 and onwards\\
\verb"\NL"&Nonlinearity\\
\verb"\NT"&Nanotechnology\\
\verb"\PB"&Phys. Biol.\\
\verb"\PM"&Physiol. Meas.\\
\verb"\PMB"&Phys. Med. Biol.\\
\verb"\PPCF"&Plasma Physics and Controlled Fusion\\
\verb"\PSST"&Plasma Sources Sci. Technol.\\
\verb"\RPP"&Rep. Prog. Phys.\\
\verb"\SST"&Semicond. Sci. Technol.\\
\verb"\SMS"&Smart Mater. Struct.\\
\verb"\SUST"&Supercond. Sci. Technol.\\
\mr
{\it No Longer Published}\\
\mr
\verb"\JPB"&J. Phys. B: At. Mol. Phys.&1968--1987\\
\verb"\JPC"&J. Phys. C: Solid State Phys.&1968--1988\\
\verb"\JPE"&J. Phys. E: Sci. Instrum.&1968--1989\\
\verb"\JPF"&J. Phys. F: Met. Phys.&1971--1988\\
\verb"\JPG"&J. Phys. G: Nucl. Phys.&1975--1988\\
\verb"\PAO"&Pure and Applied Opt.&1992--1998\\
\verb"\QO"&Quantum Opt.&1989--1994\\
\verb"\QSO"&Quantum and Semiclass. Opt.&1995--1998\\
\br
\end{tabular}
\end{indented}
\end{table}

\begin{table}

\caption{\label{jlab2}Abbreviations for some more 
common non-IOP journals.}
\begin{indented}
\item[]\begin{tabular}{@{}ll}
\br
Macro name&{\rm Short form of journal}\\
\mr
\verb"\AC"&Acta Crystallogr.\\
\verb"\AM"&Acta Metall.\\
\verb"\AP"&Ann. Phys., Lpz\\
\verb"\APNY"&Ann. Phys., NY\\
\verb"\APP"&Ann. Phys., Paris\\
\verb"\CJP"&Can. J. Phys.\\
\verb"\GRG"&Gen. Rel. Grav.\\
\verb"\JAP"&J. Appl. Phys.\\
\verb"\JCP"&J. Chem. Phys.\\
\verb"\JJAP"&Japan. J. Appl. Phys.\\
\verb"\JMMM"&J. Magn. Magn. Mater.\\
\verb"\JMP"&J. Math. Phys.\\
\verb"\JOSA"&J. Opt. Soc. Am.\\
\verb"\JP"&J. Physique\\
\verb"\JPhCh"&J. Phys. Chem.\\
\verb"\JPSJ"&J. Phys. Soc. Japan\\
\verb"\JQSRT"&J. Quant. Spectrosc. Radiat. Transfer\\
\verb"\NC"&Nuovo Cimento\\
\verb"\NIM"&Nucl. Instrum. Methods\\
\verb"\NP"&Nucl. Phys.\\
\verb"\PF"&Phys. Fluids\\
\verb"\PL"&Phys. Lett.\\
\verb"\PR"&Phys. Rev.\\
\verb"\PRL"&Phys. Rev. Lett.\\
\verb"\PRS"&Proc. R. Soc.\\
\verb"\PS"&Phys. Scr.\\
\verb"\PSS"&Phys. Status Solidi\\
\verb"\PTRS"&Phil. Trans. R. Soc.\\
\verb"\RMP"&Rev. Mod. Phys.\\
\verb"\RSI"&Rev. Sci. Instrum.\\
\verb"\SSC"&Solid State Commun.\\
\verb"\SPJ"&Sov. Phys.--JETP\\
\verb"\ZP"&Z. Phys.\\
\br
\end{tabular}
\end{indented}
\end{table}

\section*{References}
\begin{thebibliography}{10}
\bibitem{book1} Goosens M, Rahtz S and Mittelbach F 1997 {\it The \LaTeX\ Graphics Companion\/} 
(Reading, MA: Addison-Wesley)
\bibitem{eps} Reckdahl K 1997 {\it Using Imported Graphics in \LaTeX\ } (search CTAN for the file `epslatex.pdf')
\end{thebibliography}

\end{document}

### 2013-08-22 OU_HittingTimeVariance.tex
Initial commit

--- 2013-08-22 CompExam.aviolov.tex
@@
-%%This is a very basic article template.
-%%There is just one section and two subsections.
 \documentclass{article}
@@
 
-\usepackage{algpseudocode}
-\usepackage{algorithm}
-
 \usepackage[super]{nth}
@@
 \usepackage{cleveref}
-\usepackage{cancel}
 
@@
 
-\newtheorem{thm}{Theorem}[section]
-\newtheorem{lemma}{Theorem}[thm]
-% \theoremstyle{definition}
-\newtheorem{ex}{Example}[thm]
-\newtheorem{defn}{Definition}[thm]
-
 \begin{document}
 
-\title{Stochastic Optimal Control - with a view towards spike control in neurons
-\\
-{\small Advanced Comprehensive Exam in Mathematics}
-\\
-{\small Dept. of Math - University of Ottawa}
-}
+\title{The 2nd moment of the Hitting time of an OU Process} 
 \author{Alexandre Iolov 
@@
 
-\date{January 10, 2013}
+\date{\today}
 
@@
 
-\abstract{We review main concepts from stochastic optimal control theory,
-including stochastic differential equations, dynamic programing and Pontryagin's
-Maximum Principle. We then discuss an optimal control problem in a simple
-mathematical model of a neuron where the feedback is minimal and we have a
-quasi-open-loop control which necessitates the use of a Maximum Principle for
-the system's transition density.}
+\abstract{Calculate the 2nd moment of the hitting time of an OU process given
+an arbitrary initial position}
 
-\tableofcontents
+\def \t {{ T}}
 
-\section{Overview}
-Stochastic Optimal Control in continuous time deals with finding
-the input which optimizes the performance of a continuous-time dynamical systems
-that contain a significant random component in their evolutions.
-
-Here we present a brief overview of the subject starting from the basic notions
-of Stochastic Differential Equations (SDEs), 
-and then introducing the main tools of Optimal Control, namely the
-Hamilton-Jacobi-Bellman equation of Dynamic Programming and the state-adjoint
-equations arising from Pontryagin's Maximum Principle. We will not state
-proofs, although we sometimes provide heuristic derivations of the main
-results.
-
-If I had to learn the material all over again, I would proceed my readings as
-follows: First I would read the online notes by Evans on both
-SDEs,\cite{Evansa}, and Optimal Control, \cite{Evansb}. I would then read
-Jacobs' book, \cite{Jacobs}, especially for the chapter on the Fokker-Planck
-equation (ch 7) and then Oksendal \cite{Oksendal2007} for a
-more thorough mathematical treatment of SDEs. Having familiarized myself with SDE's I would
-return to Optimal Control in the Deterministic Case, starting with \cite{Kirk2004} for the
-finite dimensional deterministic case (ODEs) and then \cite{Fleming1975} for the
-extension to the stochastic case. I would keep the following for reference \cite{Krylov2008},
-\cite{Hanson2007} and finally \cite{Fattorini1999} for the
-deterministic infinite-dimensional (PDEs) case.
-
-I have not used any particularly sophisticated numerical
-methods, but, for example, Hanson \cite{Hanson2007} discusses both the
-Euler-Maruyama scheme for simulating SDEs and the Crank-Nicholson Method for
-solving PDEs which is all I have used in my codes.
-
-All of the theoretical results presented in the sequel will be from
-the above references.
-
-\section{Stochastic Differential Equations}
-In view of our ultimate goals, we will restrict ourselves to  stochastic
-processes whose sample paths have continuous paths, i.e.\ to SDEs driven by
-Brownian motion.
-
-Since the reader is more likely well familiar with the material in this section,
-we will not provide proofs, but only state the results with a view towards
-establishing the notation for the sequel.
-
-We will assume that the reader is familiar with the following concepts:
-\begin{itemize} 
-  \item a probability space, $\{\O, \sAlg, P\}$ consisting of a
-  probability space, a sigma-algebra and a probability measure
-  \item a continuous-time stochastic process, $X_t$
-  \item a filtration $\Fil(t)$ and in particular the filtration generated by a
-  stochastic process.
-\end{itemize}
-
-\subsection{The Wiener Process and the Ito Integral}
-The Wiener Process is the fundamental building block of the stochastic calculus,
-it is often called Brownian Motion and we will denote it $\{W_t\}_{t\geq 0}$. It
-satisfies:
-\begin{defn}Wiener Process, $W_t$:
-\begin{enumerate}
-  \item $W_0 = 0$
-  \item $W_t - W_s = N(0, |t-s|)$ , i.e. normally distributed with mean 0,
-  variance $|t-s|$)
-  \item $\forall \{t_i\}_1^N, \quad \{W_{t_i} - W_{t_{i-1}} \}_2^N \sim$
-  independent, i.e $W_t$ has independent increments
-\end{enumerate}
-\end{defn}
-The fact that the finite incremental distributions suffice to specify a unique
-continuous-time stochastic process is known as Kolmogorov's extension theorem.
-\begin{ex}[from \cite{Evansa} \#21.] Inverted Wiener Process:
-\\
-Let $W_t$ be a 1-d Wiener Process and define:
-\def \Wbar {{\bar{W} }}
-\begin{equation*}
-\Wbar(t) := \begin{cases}
-t W(1/t) & t >0
-\\
-0 & t = 0
-\end{cases}
-\end{equation*}
-Show $\Wbar(t)- \Wbar(s) \sim N(0, t-s)$.
-
-\begin{proof}[Soln]
-\begin{align*}
-\Wbar(t) - \Wbar(s) =& t W(1/t) - s W(1/s), \, t>s
-\\
-=& (t-s)W(1/t) + s W(1/t) - sW(1/s) & \text{recall:} 1/s > 1/t
-\\
-=& (t-s)N(0, 1/t) - sN(0, 1/s -1/t) & \text{by indnt increments}
-\\
-=& N\left(0, \frac{(t-s)^2}t + \frac{s^2 }{ 1/s -1/t} \right) & \text{
-addition of indnt normal RVs.}
-\end{align*}
-With some algebra, the variance in the last expression reduces to $t-s$
-\end{proof}
-\end{ex}
-
-Now, we collect a few more relevant properties of $W_t$:
-\begin{enumerate}
-  \item The sample paths $W_{[0,\infty)}(\o)$
-are almost surely (a.s.) continuous and are in fact Holder continuous for any
-exponent $\g < 1/2$
-\item The sample paths $W_{[0,\infty)}(\o)$ are nowhere differentiable
-\item $W_t$ is a Markov process: $\Prob[W_t \in B | \s(W_{s' \leq s })] =
-\Prob[W_t \in B \,| \,W_s]$, for any Borel set, $B \subset \R$, where $\s(W_{s'
-\leq s })$ is the filtration generated by the $W_t$
-\end{enumerate}
-
-We now turn to defining stochastic integrals based on the Wiener Process.
-
-\begin{defn} Progressively Measurable Functions:
-
-Let $\Fil_t$ be the filtration generated by the Wiener Process.
-
-Let $X_t$ be a stochastic process which is $\Fil_t$-measurable $\forall t$ and
-which is jointly measurable in $(t,\o)$. We call such an $X_t$
-\emph{progressively measurable}
-\end{defn}
-
-\begin{defn} $\Ltwopm, \Lonepm$
-
-We define $\Ltwopm[0,T]$ as the space of all progressively measurable
-$X$ such that
-\begin{equation*}
-\Exp \left[ \int_{[0,T]} X_t^2 \intd{t} \right] < \infty
-\end{equation*}
-
-Similarly, we define $\Lonepm[0,T]$ as the space of all progressively measurable
-$X$ such that
-\begin{equation*}
-\Exp \left[ \int_{[0,T]} X_t \intd{t} \right] < \infty
-\end{equation*} 
-\end{defn}
-
-$\Ltwopm$ will be the class of functions for which the Ito integral is
-well-defined. 
-\begin{defn} Ito Integral:
-\label{defn:ito_integral}
-
-Let $P^n := {a = t^n_1 \ldots t^n_{m_n} = b}$ be a partition of the interval
-$[a,b] \subset [0, \infty)$. Let $|P_n| = \sup_i|t_i - t_{i-1}|$. 
-Take $|P_n| \rightarrow_n 0 $ and consider an $ X_t \in \Ltwopm[a,b]$,
-
-then
-\begin{equation}
-\int_{[a,b]} X_t \intd{W} := \lim_{n \rightarrow \infty}  
-\sum_{i=1}^{m_n} X_{t_i}\left( W(t_{i+1}) - W(t_{i})\right)
-\end{equation}
-
-\end{defn}
-
-To be precise, our definition is actually a theorem, and the real
-definition is one that uses step functions and passes to the limit.  Also we
-will write the limits of integration $\int_0^T \cdot  \intd{W}$ or $\int_{[0,T]}
-\cdot  \intd{W}$ interchangeably.
-% To justify def'n \ref{defn:ito_integral}, we recall:
-% \begin{lemma}[Quadratic Variation]
-% \begin{equation}
-% \lim_{n \rightarrow \infty} \sum_{i=1}^{m_n} \left( W(t_i) - W(t_{i-1})\right)^2
-% = b-a \quad \text{in }  L^2(\O)
-% \end{equation}
-% \end{lemma}
-% The above can be read, heuristically as $\lim \sum (\Delta W)^2 \rightarrow \sum
-% \Delta t$, which is the meaning behind the colloquial $\intd{W} =
-% \sqrt{dt}$.
-
-Again, we state without proof a few interesting properties of $\int X \intd{W}$:
-\begin{thm} Ito Integral Properties
-
-\begin{enumerate}
-  \item $\Exp[\int_0^T X \intd{W} ] = 0$ 
-  \item $\Exp[\left(\int_0^T X \intd{W}\right)^2 ] = \Exp[\int_0^T X^2
-  \intd{t}]$
-  \item $I(t) = \int_0^t X \intd{W} $ is a martingale 
-  \item $I(t) = \int_0^t X \intd{W} $ has continuous sample paths
-\end{enumerate}
-\end{thm}
-
-\subsection{Ito SDEs and Ito's Lemma}
-\begin{defn} an Ito SDE:
-
-We write
-\begin{equation}
-dX_t = Fdt + G dW
-\end{equation}
-on $0 \leq t \leq T$, if $X$ is a real-valued stochastic process satisfying:
-\begin{equation*}
-X(r) = X(s) + \int_{[s,r]} F \intd{t} + \int_{[s,r]} G \intd{W}
-\quad 0\leq s \leq r \leq T
-\end{equation*}
-for some $F \in \Lonepm[0,T]$ and $G \in \Ltwopm[0,T]$
-\end{defn}
-
-We are now ready to present the celebrated Ito Lemma which is the chain-rule of
-stochastic calculus:
-
-\begin{thm}[Ito Lemma] Suppose $X$ satisfies the stochastic differential $dX_t =
-Fdt + G dW$ as above and take $u(x,t) \in C^{2,1}[ \R \times [0,T]]$.
-
-Set $Y(t) = u(X,t)$
-\\
-then
-$$
-dY =  \left( \di_t u + \di_x u \cdot F + \di^2_x u \cdot \frac{G^2}2 \right)
-\intd{t} + \left(   \di_x u\cdot G  \right)\intd{W}
-$$
-\end{thm}
-
-% We now present two examples, one fundamental and one more involved:
-% \begin{ex}[pg 67 in \cite{Evansa}] The 'Stochastic' Exponential:
-% 
-% Let $Y = \exp(\l W - \l^2 t / 2)$. In the notation of the above Thm, we have:
-% $dX = dW, F=0, G=1$ and $u(x,t) = \exp(\l x - \l^2 t / 2)$.
-% 
-% then
-% \begin{align*}
-% dY =& 
-% \left( \di_t u + \di_x u \cdot F + \di^2_x u \cdot \frac{G^2}2 \right)\intd{t} +
-% \left(   \di_x u \cdot G  \right)\intd{W}
-% \\
-% =&
-% \left( -\frac{\l^2 t}{2}\exp(\l x - \l^2 t / 2) + \l^2\exp(\l x - \l^2 t / 2)  \right)\intd{t}
-% \\ 
-% &+ \left(   \l \exp(\l x - \l^2 t / 2) \right)\intd{W}  
-% \\
-% =&  \l \left( \exp(\l x - \l^2 t / 2) \right)\intd{W}  
-% \\
-% =& \l Y dW
-% \end{align*}
-% \end{ex}
-% 
-% \begin{ex}[pg. 67 in \cite{Evansa}] Hermite Polynomials:
-% 
-% Let $h_n(x,t) = \frac{(-t)^n}{n!} e^{x^2/2t} \di^n_x \left( e^{x^2/2t} 
-% \right)$ be the $n$th Hermite polynomial
-% \\
-% then
-% $$
-% \int_0^t h_n(W, t) \intd{W} = h_{n+1}(W(t), t)) 
-% $$
-% i.e. in Ito calculus, the Hermite polynomials are like the power monomials
-% $\frac{t^n}{n!}$ in the ordinary calculus.
-% \begin{proof} Since
-% $$
-% d_\l^n [\exp(- \frac{(x-\l t)^2}{2t} ] |_{\l=0} =
-%  (-t)^n d_x^n [ \exp(-\frac{x^2}{2t})] $$
-% We will have:
-% \begin{align*}
-% d_\l^n [\exp(\l x - \l^2 t / 2) ] |_{\l=0} &= (-t)^n  e^{x^2/2t} d_x^n [
-% \exp(-\frac{x^2}{2t})]
-% \\
-% &= n! h_n(x,t)
-% \end{align*}
-% Thus we can expand $\exp(\l x - \l^2 t / 2) $ in a Taylor series at $\l=0$ as:
-% $$
-% \exp(\l x - \l^2 t / 2) = \sum_n \l^n h_n(x,t)
-% $$
-% Now recall $Y$ from the previous example:
-% \begin{equation*}
-% Y(t) = \exp(\l W(t) - \frac{\l^2 t }2) = \sum_n \l^n h_n(W(t),t)
-% \end{equation*}
-% which satisfies:
-% \begin{equation*}
-% Y(t) = 1 + \l \int Y \intd{W}
-% \end{equation*}
-% Naturally we will plug the series expansion in the SDE to obtain:
-% \begin{align*}
-% \sum_{n=0}^{\infty} \l^n h_n(x,t) =&
-% 1 + \l \int_0^t  \sum_{n=0}^{\infty} \l^n
-% h_n(x,t)\intd{W}
-% \\
-% =& 1 + \sum_{n=1}^{\infty} \l^n\int_0^t 
-% h_{n-1}(x,t)\intd{W} \quad {\textrm{bump $n$ up by one}}
-% \end{align*}
-% Equating the coefficients for $\l^n$ on both sides leads to the result
-% \end{proof}
-% \end{ex}
-
-Finally, we are in a position to define an Ito SDE:
-
-\begin{defn}[Ito SDE] Let
-$F(x,t):\R\times [0,T]\ra \R,  G(x,t):\R\times [0,T]\ra \R,$ be given functions.
-We say that a stochastic process $X$ satisfies:
-\begin{equation}
-dX =F dt + G dW
-\end{equation} 
-over $[0,T]$ if
-\begin{enumerate}
-  \item $X$ is progressively measurable wrt. $\Fil_t$
-  \item $F(X,t) \in \Lonepm$
-  \item $G(X,t) \in \Ltwopm$ 
-  \item $X_t = X_0 + \int_0^t F(X_s, s) \intd{s} + \int_0^t G(X_s, s) \intd{W_s}$
-\end{enumerate}
-\end{defn}
-
-% \begin{ex} The solution to 
-% $$
-% \begin{cases}
-% dX = F(t)X dt + G(t) X dW
-% \\
-% X_0 = 1
-% \end{cases}
-% $$
-% is 
-% $$
-% X_t = \exp\left( \int_0^t F(s) - \frac{G^2(s)}{2} \intd{s} + \int_0^t G
-% \intd{W} \right) $$
-% \begin{proof}
-% Let $Y = \int_0^t f(s) - \frac{g^2(s)}{2} \intd{s} + \int_0^t g
-% \intd{W} $
-% and guess that $X = e^Y$
-% then, using Ito's lemma:
-% \begin{align*}
-% dX =& (\di_y [e^y] dY + \frac{1}{2}\di^2_y[e^y]  g^2 dt
-% \\ =&
-% (e^y  f dt - e^y g^2 / 2) dt + e^y g dW + \frac{1}{2} e^y \cdot  g^2 dt
-% \\ =& e^y \left( f dt +g dW \right)
-% \\ =& X \left( f dt +g dW \right)
-% \end{align*}
-% \end{proof}
-% \end{ex}
-
-% \begin{ex}[Brownian Bridge, pg 80 in \cite{Evansa}] The solution of the SDE
-% $$
-% \begin{cases}
-% dB = -\frac{B}{1-t} dt + dW
-% \\
-% B_0= 0
-% \end{cases}
-% $$
-% is
-% $$
-% B(t) = (1-t) \int_0^t \frac{1}{1-s} dW
-% $$
-% \begin{proof}
-% It is not clear how to apply Ito's lemma to this problem, so instead we just
-% directly calculate the differential:
-% \begin{align*}
-% B(t^+) - B(t) =& (1-t^+) \int_0^{t^+} \frac{1}{1-s} dW - (1-t) \int_0^t
-% \frac{1}{1-s} dW
-% \\ =&
-% (1-t^+) \int_t^{t^+}\frac{1}{1-s} dW  - (t^+-t) \int_0^t
-% \frac{1}{1-s} dW
-% \\ \approx&
-% \int_t^{t^+} dW - \Delta t \frac{B(t)}{1-t}  
-% \end{align*}
-% \end{proof}
-% \end{ex}
-% Actually we can say a bit more about the Brownian bridge, in particular we
-% will calculate $\Exp[B^2]$
-%  \begin{ex}[Evans .37]
-% \begin{align*}
-% \Exp[B^2] =& \Exp\left[\left( (1-t) \int_0^t \frac{1}{1-s} dW \right)^2 \right]
-% \\
-% =&(1-t)^2 \Exp\left[ \int_0^t\left(  \frac{1}{1-s}\right)^2 \intd{s} \right]
-% \\
-% =&(1-t)^2 [ \frac{1}{1-t} - 1 ]
-% \\
-% =&(1-t) - (1-t)^2
-% \end{align*}
-% 
-% Further note that  $\lim_{t\ra 1^-}\Exp[B^2] = 0$ and that $B$ is a martingale
-% and so its first moment is always equal to its initial value, $0$. So indeed the 2nd
-% moment is the variance and with the variance converging to zero, we can use
-% Chebyshev's inequality to conclude that $\lim_{t\ra 1^-}B(t) = 0 $(a.s). This
-% justifies the name of the Brownian bridge - it is a normally distributed RV.
-% which is clamped to equal 0 at both $t= \{0,1\}$.
-% \end{ex}
-
-As an example we will discuss the Ornstein-Uhlenbeck process, which is the
-basis for the models we will face later on.
-\begin{ex}[O-U Process] Let $X$ follow:
-\begin{equation*}
-dX = \th (\a - X_t) dt + \b dW
-\end{equation*}
-with an initial condition, $X_0$, which may be an arbitrary distribution
-independent of the Wiener Process,$W$.
-We can solve this as follows:
-\begin{align*}
-dX =& \th (\a - X_t) dt + \b dW
-\\
-dX + \th X_t) dt=& \th \a dt + \b dW
-\\
-e^{\th t} dX + e^{\th t}\th X_t) dt=& \th e^{\th t}\a dt + \b e^{\th t} dW
-\\
-Xe^{\th t} - X_0 = & \int \th e^{\th t}\a dt +  \int \b e^{\th t} dW
-\\
-X_t =& e^{-\th t} X_0 + \a(1-e^{-\th t}) +  \frac{\b e^{-\th t}}{\sqrt{2\th}}
-W(e^{2\th t}-1)
-\end{align*}
-which means that $X$ forgets its initial conditions exponentially fast and
-converges to a normal random variable with mean $\a$ and variance
-$\tfrac{\b^2}{2\th}$
-\end{ex}
-
-At the end of this sub-section, we state Ito's Lemma with multiple Wiener
-processes and then for a multidimensional state.
-
-\begin{thm}[Ito Lemma for $dW \in  \R^{m}$] Suppose $X$ satisfies the stochastic
-differential $dX_t = F dt + G dW$, where $ F \in \R, G \in \R^{1\times m}$ and
-$dW = \left(dW^{(i)}\right) \in  \R^{m}$ is a vector of independent Wiener
-Processes. Take $u(x,t) \in C^{2,1}[ \R \times [0,T]]$. Let $D :=\left[{\sum_k
-G_{k} G_{k} } \right] \in \R$
-
-Set $Y(t) = u(X,t)$
-\\
-then
-$$
-dY =  \left( \di_t u +  \di_{x} u \cdot F + 
-  \di^2_{x} u \cdot D \right)
-\intd{t} +
- \left(  \di_{x} u  \cdot  G \right) dW  
-$$
-\end{thm}
-
-\begin{thm}[Ito Lemma for $X \in R^n$]
-Suppose $X$ satisfies the stochastic differential $dX_t = F dt + G dW$, where $
-F \in \R^n, G \in \R^{n\times m}$ and $dW = \left(dW^{(i)}\right) \in  \R^{ 
-m}$ is a vector of independent Wiener Processes. Take $u(x,t) \in C^{2,1}[ \R^n
-\times [0,T]]$. Let $D_{ij} := \tfrac{1}{2} \left[{\sum_k G_{ik} G_{jk} }
-\right]_{ij} \in \R^{n\times n} $
-
-Set $Y(t) = u(X,t)$
-\\
-then
-$$
-dY =  \left( \di_t u + \sum_i \di_{x_i} u \cdot f_i + 
-\sum_{i,j} \di^2_{x_i x_j} u \cdot D_{ij} \right)
-\intd{t} +
- \left(  \sum_{ij} \di_{x_i} u  \cdot G_{ij} dW^{(j)} 
-\right)$$
-\end{thm}
-
-\subsection{Fokker-Planck and Kolmogorov's Backward Equations}
-The Fokker-Planck equation and Kolmogorov's Backward equation describe the
-forward (resp. backward)  evolution of the probability density of $X_t$
-
-For the rest of this section we will work in $ \R^n$, i.e. $X$ will be an $n-$
-dimensional vector that satisfies the stochastic differential 
-\begin{equation}
-dX_t = F(X,t) dt + G(X,t) dW,
-\label{eq:generic_Ito_SDE_Rn}
-\end{equation}
-$F \in \R^n, G \in \R^{n \times m}$,
-$dW = (dW^{(i)}) \in  \R^{  m}$ is an $m$-dimensional Brownian motion and we
-write $D_{ij} :=
-\tfrac{1}{2} 
-\left[{\sum_k G_{ik} G_{jk} } \right]_{ij} \in \R^{n\times n} $.
-
-% Let $$\F(x,t| y,s) = \Prob[X_t \preceq x| X_s = y] $$ be the forward transition
-% distribution and
-
-Let $$ \f(x,t| y,s) \intd{x} =  \Prob[X_t \in \intd{x} | X_s = y] $$ be the
-transition probability density. Then $\f$ satisfies:
-\begin{equation}
-\di_t \f= -\sum_i \di_{x_i} \left[ F_i(x,t) \f(x,t) \right] 
-+ 
-\sum_{i,j}  \di^2_{x_i x_j} \left[ D_{ij}(x,t) \f(x,t) \right]
-\label{eq:FP_pde}
-\end{equation}
-This is called the \emph{Fokker-Planck} or \emph{Forward Kolmogorov} equation.
-It can also be seen as a continuity euquation or a cosevation of probability
-equation. To this end define the probability current, $\p \in \R^n$ as:
-$$
-\phi_i =   F_i(x,t) \f(x,t) 
-+ 
-\sum_{j}  \di_{x_j} \left[ D_{ij}(x,t) \f(x,t)\right]
-$$
-then the Fokker-Planck equation, \cref{eq:FP_pde}, can be written as:
-$$
-\di_t \f= -\grad \cdot \phi
-$$
-which just says that the change in probability is the
-difference between the flow in and the flow out.
-
-Initial conditions for $\f$ are given by the distribution of $X_s$, if $s=0$
-then this is the initial distribution of $X$. Boundary conditions (BCs) depend
-on the domain of $X$ and what happens to $X$ once it hits its boundary. If the domain
-is all of $\R^n$ then we only insist that $\lim_{|x| \ra \infty} \f = 0$. If the
-domain has boundaries, then there are two common scenarios which we will also
-encounter in the sequel:
-\begin{enumerate}  
-  \item absorbing BCs
-  \item reflecting BCs
-\end{enumerate}
-At an absorbing boundary, the particle $X$ is removed and $\f=0$ there.
-In this situation, we will not have conservation of probability and the integral
-of $\f$ over $X$'s domain will monotonically decrease.
-
-At a reflecting boundary, the particle $X$ bounces back into its domain and we
-will have that the probability current $\p\cdot n = 0$ where $n$ is the outward
-normal at the reflecting boundary.
-
-We now consider $\f$ as functions of $y,s$, holding $x,t$ fixed, then
-\begin{equation}
--\di_s \f= \sum_i  \left[F_i(y,s) \cdot \di_{y_i}\f(|y,s) \right] 
-+ 
-\sum_{i,j}   \left[ D_{ij}(x,t) \di^2_{y_i y_j}\f(|y,s) \right]
-\label{eq:FP_backward_pde}
-\end{equation} 
-This is the \emph{Backward Kolmogorov} equation for $\f$. Note the minus sign
-in front $\di_s \f$. A mnemonic for the signs of the Backward vs. Forward
-equation is that as $t$ increases $\f$ diffuses and so $\di_t$ and $\di^2_{x}$
-have the same sign, but as $s$ increases $\f$ anti-diffuses and they have
-opposite signs.
-
-The differential operator on the right-hand side of
-\cref{eq:FP_backward_pde} occurs often in the study of SDEs and has its own
-name.
-\begin{defn}[Generator of an SDE] the Generator $A$ of $X$ is defined by:
-$$
-A[\psi(x)] = \lim_{t \searrow 0^+} \frac{\Exp[\psi(X_t)]  - \psi(x)}{t} ;  \quad X_0 =
-x \in \R^n
-$$
-\end{defn} 
-\begin{lemma} For an Ito SDE as in \cref{eq:generic_Ito_SDE_Rn} 
-$$
-A[\psi(x)] = \sum_i F_i(x,t) \cdot \di_{x_i} \psi + \sum_{i,j} D_{ij}(x,t)
-\di^2_{x_i x_j}\psi $$
-\end{lemma}
-
-\subsection{Stopping Times}
-\begin{defn}
-Let $\Fil(t)$ be some filtration, a random variable $\t$ is called a
-\emph{stopping time} if 
-$$
-\{\o : \t(\o) \leq t\} \in \Fil(t) \, \forall t  
-$$
-\end{defn}
-The colloquial way of describing stopping times is at any time we know
-whether $\t$ has occurred or not. For a counterexample, the time that a Wiener
-Process achieves its maximum over some interval is not a stopping time, since at any
-given time, we do not know if the maximum has occurred or not. The most common
-example of a stopping time is the first hitting-time, which is the first time
-$X_t$ leaves a set. 
-\begin{thm}[First-Hitting time] Let $E \subset \R^n, E \neq
-\phi$ be open or closed
-
-then $$\t := \inf\{ t \geq 0 | X_t \in E\}$$ is a stopping time.
-\end{thm}
-
-The reason stopping times are very useful is that all the facts so far quoted
-for Ito calculus using integrals $\int_0^T \intd{W}$ remain true if $T$ is
-replaced by a stopping time $\t$.
-
-Also it allows us to link SDE's and PDEs using the generator, $A$, of the
-diffusion: 
-\begin{thm}[Dynkin's formula] Given $\t$ a stopping time, $\Exp[\t] < \infty$.
-Then:
-$$
-\Exp[u(X_\t, \t)] =
-u(x, 0) + \Exp\left[\int_0^\t \di_t u + A[u] \intd{s}
-\right]
- $$
-\end{thm}
-This provides a link between PDEs and stochastic processes and allows us to go
-back and forth in that we can find probabilistic results by
-solving a PDE or we can approximate a PDE by simulating a stochastic process
-and averaging.
-
-\begin{ex}[\cite{Evansb} pg 99 - Expected hitting time to a boundary]
-\label{ex:mean_hitting_time}
- Let $\O
-\subset \R^n$ be a bounded open set with smooth boundary $\di \O$ then it is a
-basic fact from PDEs theory that
-\begin{equation}
-\begin{cases}
--\frac{1}{2} \grad^2 u = 1  & \text{over } \O
-\\
-u =  0 &\text{on } \di \O
-\end{cases}
-\end{equation}
-has a unique $C^\infty(\O)$ solution.
-
-Let $X = W_t + x$ for any $x \in \O$ and define
- $$\t_x := \text{first time } X \text{ hits } \di \O$$
-then the generator of $X$ is $A[\psi] = -\grad^2(\psi)/2$ and we will
-have:
-
-\begin{align*}
-\Exp[u(X_\t)] - \Exp[ u(x_0)] =& \Exp \left[ -\frac{1}{2}\int_0^\t \grad^2
-u\right]
-\\
-=& \Exp \left[ - \int_0^\t 1
-\right]
-\\
-=-&\Exp [\t]
-\end{align*}
-Finally, invoke $u$'s BCs, $u|_{\di E} = 0$, and $X$'s ICs, $X_0=0 = x$ to
-conclude that $$ u(x) = \Exp [\t]$$
-The solution to the PDE evaluated at $x$ is the expected exit time from
-$E$ for an $X_t$  starting at $x$.
-\end{ex}
-% Similarly we can calculate more moments of the hitting time.
-% \begin{ex}[Mean squared hitting time to a boundary] Let $E
-% \label{ex:mean_squared_hitting_time}
-%  \subset \R$ be a bounded open set and let $u$ solve:
-%  \begin{equation}
-%  \begin{cases}
-%  (\a - \tfrac{x}{\tc}) \di_x u + \frac{\b^2}{2} \di_x^2 u = -\frac t 2  &
-%  \text{over } E
-%  \\
-%  u =  0 &\text{on } \di E
-%  \end{cases}
-%  \end{equation}
-% 
-% Then $- t / 2 = A[u]$ for $A$, the generator, corresponding to $dX = (\a - x /
-% \tc) dt + \b dW$.
-% Define $$\t_x := \text{first time } X \text{ leaves }  E$$
-% then
-%  \begin{align*}
-%  \Exp[u(X_\t)] - \Exp[ u(X_0)] =& 
-%  \Exp \left[  \int_0^\t   (\a - \tfrac{x}{\tc}) \di_x u + \frac{\b^2}{2} \di_x^2
-%  u \right]
-%  \\
-%  =& \Exp \left[ - \int_0^\t -\frac t 2 
-%  \right]
-%  \\
-%  =-&\Exp [\t^2]
-%  \end{align*}
-% And we can conclude that 
-% $$ u(x) = \Exp [\t^2]$$
-% The solution to the PDE evaluated at $x$ is the expected squared exit time from
-% $E$ for an $X_t$  starting at $x$.
-% \end{ex}
-
-\subsection{Feynman-Kac formula}
-The Feynman-Kac formula is a generalization of the above idea. We state it in
-1-D. 
-\begin{thm}[Feynman-Kac Formula] Consider the (linear) PDE, with terminal
-conditions:
-\begin{equation}
-\begin{cases}
-\di_t u(x,t) + f \di_x u + \frac {g^2}2 \di_x^2 u - V u + q 
-&= 0
-\\
-u(x,\tf) &= \psi(x)
-\end{cases}
-\label{eq:PDE_feynman_kac}
-\end{equation}
-The dynamics can also be written:
-$$
-\di_t u(x,t) + A[u] - V u + q = 0
-$$
-for the generator $A = f \cdot \di_x + g^2/2 \cdot \di_x^2$ corresponding to the
-SDE: $$
-dX_t = f(X,t) dt + g(X,t) dW
-$$
-Then for $t< \tf$
-\begin{align*}
-u(x,t) =& \Exp\bigg[
-\int_t^\tf q(x,s)\cdot \exp \left(-\int_t^s V(X_\t) \intd{\t} \right)\intd{s} \\
-&\quad + \, \psi(X_t) \cdot \exp\left(-\int_t^\tf V(X_\t,\t) \intd{\t} \right)  
-\quad \Big| \,X_t = x \bigg]
-\end{align*}
-\begin{proof}[Proof]
-This can be seen as a more elaborate extension of example
-\ref{ex:mean_hitting_time}. We set: 
-$$
-Y_r = u(x,t)+ 
-\int_t^r q(x,s)\cdot\exp \left(-\int_t^s V(X_\t) \intd{\t} \right)\intd{s} +
-u(X_r,r)\cdot\exp\left(-\int_t^r V(X_\t,\t) \intd{\t} \right)   
-$$
-then 
-\begin{align*}
-dY =& q(x,r)\cdot\exp \left(-\int_t^r V(X_\t) \intd{\t} \right)
-\\
-& - u(X_r,r) \cdot V(X_r, r)  \exp\left(-\int_t^r V(X_\t,\t)
-\\
-& + du \cdot \exp\left(-\int_t^r V(X_\t,\t) \intd{\t} \right) 
-\intd{\t}\right)
-\end{align*}
-Now we recall Ito's lemma for $du(X,t)$, and replace it in the last line:
-\begin{align*}
-dY_r =& q(x,r)\cdot\exp \left(-\int_t^r V(X_\t) \intd{\t} \right) \intd{r}
-\\
-& - u(X_r,r) \cdot V(X_r, r)  \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t}\right)s
-\\
-& + \Big( \di_t u + A[u] \Big) \cdot \exp\left(-\int_t^r V(X_\t,\t) \intd{\t}
-\right) \intd{r}
-\\
-& + \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-\end{align*}
-Now we can group the terms in front of the exponential term on the first
-three lines to form:
-\begin{align*}
-dY_r =& \Big(\di_t u + A[u] - u(X_r,r) V(X_r, r) + q(X_r,r) \Big) \cdot\exp
-\left(-\int_t^r V(X_\t) \intd{\t} \right)
-\intd{r}
-\\
-& + \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-\end{align*}
-Since, $u$ satisfies the PDE, \cref{eq:PDE_feynman_kac}, the term on the first
-line is zero so $dY$ is just
-$$
-dY_r = \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-$$
-integrating we have: 
-$$
-Y_\tf - Y_t = \int^\tf_t \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot
-\exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-$$
-Now we need conditions such that the expectation of the stochastic integral is
-zero. For example, it suffices that $V$ is bounded, and often in applications
-$V$ is a positive rate.
-Then, we   take conditional expectations, conditional on $X_t=x$:
-\begin{equation*}
-\begin{gathered}
-\Exp\Big[Y_\tf | X_t = x\Big] - \Exp\Big[Y_t | X_t = x\Big]  
-= 
-\\
-\Exp\Bigg[\int^\tf_t \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot
-\exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W} \,\mid\, X_t = x\Bigg] 
-\\ = 0
-\end{gathered}
-\end{equation*}
-Unravelling the definition of $Y_t$, we get
-\begin{align*}
-\Exp\bigg[
-\int_t^\tf q(x,s)\cdot \exp \left(-\int_t^s V(X_\t) \intd{\t} \right)\intd{s} 
- \quad + &
-\\
- \, \psi(X_t) \cdot \exp\left(-\int_t^\tf V(X_\t,\t) \intd{\t} \right)  
-\, \Big| \,X_t = x \bigg] &=\, u(x,t)  
-\end{align*}
-
-\end{proof}
-A more detailed proof can be found in \cite{Oksendal2007} Thm. 8.2.1
-\end{thm}
-
-We will not use the Feynman-Kac formula in what follows, but
-we have included it here, because there are some very interesting developments in stochastic optimal
-control, which rely on solving a many-dimensional PDE using stochastic
-simulation, the link being made via the Feynman-Kac formula - see
-\cite{Kappen2005},\cite{Palmer2011}.
-
-% \subsection{Numerical Simulation of SDEs}
-
-\section{Deterministic Optimal Control in Finite Dimension}
-Optimal Control theory has three main components - a state, $x$, a control, $u$
-and an objective $J$ which is a functional of $x,u$ and which we try to either
-minimize or maximize. Here, we will be minimizing.
-
-We introduce the theory for a finite dimensional state, $x(t) \in \R^{n_x}$, but
-we will not address state constraints. Our control will also be
-finite-dimensional $u(t) \in \R^{n_u}$ and it will be allowed to take values in
-some closed domain, $\Udomain$, of allowed controls, e.g.\ $\Udomain = \{ u_t
-\in \R^{n_u} \, \st |u| \leq \Umax \}$ for a control constrained to a sphere.
-The objective $J$ will be given by a time-integral-plus-terminal cost:
-\begin{equation}
-J[u] = \int_0^\tf L(x_s, u_s) \intd{s} + M(x_\tf)
-\label{eq:generic_objective_function_deterministic}
-\end{equation}
-Where $\tf$ could be either a variable or fixed.
-
-The dynamics of $x$ are given by some controlled ordinary differential
-equation (ODE) given some initial conditions:
-\begin{equation}
-\dot{x} = f(x,t); \quad x(0) = x_0
-\label{eq:generic_dynamics_deterministic}
-\end{equation}
-
-Then we seek the optimal control $u^*$, such that:
-\begin{equation}
-u^* = \argmin_{u \in \Udomain} J[u]  
-\label{eq:generic_objective_argmin}
-\end{equation}
-
-Note that $x$ may also have final conditions, which we will deal with later,
-and/or path constraints. If there are path-constraints, we talk about a
-state-constrained problem. However, we will not face path constraints in the
-sequel and so will not say more about how to handle them.
-
-There are essentially three methods for solving equation
-\cref{eq:generic_objective_function_deterministic}. Two, Pontryagin's Minimum
-Principle and Dynamic Programing, are analytic in nature providing a set of
-equations that characterize the minimum and then solving these equations,
-numerically if necessary. The third, the direct method, simply discretizes the
-problem and solves the resulting nonlinear programing (NLP) problem via standard
-NLP techniques. We will describe both Pontryagin's Minimum Principle and Dynamic
-Programing below and indeed we will use both later on.
-
-\subsection{Pontryagin's Maximum Principle}
-Pontryagin's Principle is derived from on a variational argument, similar to the
-Euler-Lagrange equations from the Calculus of Variations, and it characterizes
-the optimal trajectory/control pair. It can be thought of as a generalization of
-the zero-tangent rule (Fermat's Rule) for finding optima in single-variable
-calculus and like both the Euler-Lagrange equations and Fermat's Rule it
-provides necessary, but not sufficient conditions for a minimum.
-
-Pontryagin's Principle also holds a similarity to Hamiltonian Mechanics, which
-should not be surprising, since Hamiltonian Mechanics is founded on
-the minimization of action.
-
-There are several versions of Pontryagin's Principle, depending on whether $x$ has
-terminal conditions or not, whether the final time, $\tf$ is specified or a
-variable in itself and most importantly whether the state has constraints or
-not. We will state the versions that are relevant for our needs, namely, $x$ has
-no constraints and $\tf$ is specified.
-
-We first introduce the Pontryagin Hamiltonian, $\H$, also called the control
-theory Hamiltonian and an adjoint state, $p \in \R^{n_x}$:
-\begin{defn} [$\H$] The control theory Hamiltonian is the function:
-$$
-\H(x,p, u) := f(x,u) \cdot p + L(x,u) \quad (x,p \in R^{n_x}, u \in \Udomain)
-$$ 
-\end{defn}
-
-\begin{thm}[Pontryagin's Fixed-Time, Free-End-Point] Assume $u^*(t)$ is
-optimal for
-\cref{eq:generic_objective_function_deterministic,eq:generic_dynamics_deterministic}
-and that $x^*$ is the corresponding trajectory.
-
-Then there exists a function $p^*:[0,\tf]:\ra \R^{n_x}$ st.
-\begin{equation}
-\begin{cases}
-\dot{x}^*(t) &=  \grad_p \H(x^*,p^*, u^*) = f(x^*, u^*)
-\\
-x(0) &= x_0
-\end{cases}
-\label{eq:pontryagin_state_ode}
-\end{equation}
-\begin{equation}
-\begin{cases}
-\dot{p}^*(t) &= -\grad_x \H(x^*,p^*, u^*) = -\grad_xL - p \grad_x f
-\\
-p^*(\tf) &= \grad_x M(x)
-\end{cases}
-\label{eq:pontryagin_adjoint_ode} 
-\end{equation}
-and
-\begin{equation}
-\H(x^*(t), p^*(t), u^*(t)) = \min_{u(t) \in \Udomain(t)}  \H(x^*(t), p^*(t),
-u(t))
-\label{eq:pontryaginH_optimality_condition} 
-\end{equation}
-\end{thm}
-We will not give a proof of this theorem (see A.2 in \cite{Evansb}),
-but we do provide a heuristic derivation in
-\cref{sec:Pontryagin_heuristic_derivation}, which is very helpful in deriving
-the necessary conditions for PDE dynamics and it explains the otherwise
-mysterious origins of $\H$ and $p$.
-
-At each time, $t$, we obtain the optimal control $u(t)$, by solving
-\cref{eq:pontryaginH_optimality_condition}. This can be easy, hard or useless.
-In the easy case, the minimum is analytically obvious, e.g.\ if the Hamiltonian
-is quadratic in $u$. If it is linear in $u$ then the minimum will occur on the
-boundaries of the control and we will talk about a bang-bang control.
-Conversely, solving the minimum of $\H$ with respect to $u$ may be impossible,
-since $x,p$ are unknown. In this case we need to have an iterative scheme which
-alternately solves
-\cref{eq:pontryaginH_optimality_condition} 
-and then \cref{eq:pontryagin_state_ode,eq:pontryagin_adjoint_ode}. Basic
-numerical techniques using this idea are discussed in the appendix in
-\cite{Kirk2004}. However, if we are going to be solving the problem numerically,
-then the direct method might be superior, see \cite{Ross2005} for a discussion.
-
-Finally, it is possible that \cref{eq:pontryaginH_optimality_condition} does not
-depend on $u$ explicitly (it will still depend  on $u$ implicitly through $x,p$)
-and so we cannot find $u$ at all by minimizing $\H$ wrt.\ $u$. This happens, for
-example, if the cost, $L$, does not depend on $u$, the dynamics, $f$, are linear
-in $u$ and $p=0$ over an interval of positive length. Such an optimal control
-problem is called singular and it is discussed briefly in \cite{Kirk2004}, and
-the theory can still be useful in finding a solution, but many complications can
-and do arise. We face a singular control problem in our own work later, however
-instead of tackling it head on, we merely regularize the problem - thus we will
-not say much more on Singular Control Problems.
-
-The version of Pontryagin's Principle for Fixed End Point, meaning $x$ has some
-terminal conditions, is very similar to the one with free end-point. Suppose 
-\begin{equation}
-\dot{x} = f(x,t); 
-\quad
-\begin{cases}
-x(0) = x_0
-\\
-x(\tf) = x_f
-\end{cases}
-\label{eq:generic_dynamics_deterministic_fixed_end_point}
-\end{equation}
-
-\begin{thm}[Pontryagin's Fixed-Time, Fixed-End-Point] Assume $u^*(t)$ is
-optimal for
-\cref{eq:generic_objective_function_deterministic,eq:generic_dynamics_deterministic_fixed_end_point}
-and that $x^*$ is the corresponding trajectory.
-
-Then there exists a function $p^*:[0,\tf]:\ra \R^{n_x}$ st.
-\begin{equation}
-\begin{cases}
-\dot{x}^*(t) &=  \grad_p \H(x^*,p^*, u^*) = f(x^*, u^*)
-\\
-x(0) &= x_0
-\\
-x(\tf) &= x_f
-\end{cases}
-\label{eq:pontryagin_state_ode}
-\end{equation}
-\begin{equation}
-\dot{p}^*(t) &= -\grad_x \H(x^*,p^*, u^*) = -\grad_xL - p \grad_x f
-\label{eq:pontryagin_adjoint_ode} 
-\end{equation}
-and
-\begin{equation}
-\H(x^*(t), p^*(t), u^*(t)) = \min_{u(t) \in \Udomain(t)}  \H(x^*(t), p^*(t),
-u(t))
-\label{eq:pontryaginH_optimality_condition} 
-\end{equation}
-\end{thm}
-I.e. $p$ loses its final conditions because $x$ has them. This relates to the
-variational nature of Pontryagin's Principle and the idea that $p$ represents
-the sensitivity of the objective wrt. small changes in $x$. Thus, if $x$ is
-fixed, there are no variations and $p$ can be anything.
-% To illustrate Pontryagin's Principle we will solve the moon
-% lander problem (Ex. 4.4.4 \cite{Evansb}), with the twist of holding the final
-% time fixed. 
-% \begin{ex}
-% 
-% \end{ex}
-
-\subsection{Dynamic Programing - the Hamilton-Jacobi-Bellman equation} 
-Dynamic Programing uses backwards recursion to tabulate the optimal control
-starting from the terminal time. The basic object in dynamic programing is the value function, $v$. In order to
-introduce it, we first extend our definition for the objective, $J$, to include later
-times, with different initial conditions.
-
-The running cost-to-go corresponding to
-\cref{eq:generic_objective_function_deterministic} is:
-\begin{equation}
-J[u; x, t] = \int_t^\tf L(x_s, u_s) \intd{s} + M(x_\tf), \quad x_t = x
-\label{eq:generic_cost_to_go_deterministic} 
-\end{equation}
-So that $J[u; x_0, 0] = J[u]$.
-
-The dynamic programming value function, $v$ is then defined by:
-$$
-v(x,t) = \inf_{u \in \Udomain} J[u; x,t] 
-$$
-it is also called the optimal cost-to-go.
-
-We immediately note the terminal conditions on $v$:
-\begin{equation}
-v(x,\tf ) = M(x)
-\end{equation}
-
-The Dynamic Programing approach is then to solve for $v$ backwards and in the
-process infer the control $u$.
-\begin{thm}[Hamilton-Jacobi-Bellman Equation] Assume that $v$ is $C^{1,1}$ in
-$x,t$. Then $v$ solves:
-\begin{equation}
-\begin{cases}
--\di_t v(x,t) &= \max_{u(t) \in \Udomain(t)} \left\{ f(x,t) \cdot \di_x v +
-L(x,u) \right\}
-\\
-v(x,\tf) &= M(x)
-\end{cases}
-\end{equation}
-\end{thm}
-We will not derive the HJB equation here, but we will derive a similar version
-later for the stochastic case, when the dynamics are given by SDEs.
-
-% \subsection{Model Predictive Control}
-% Solving the HJB equations and obtaining a feedback control is usually
-% impossible. Thus a practical alternative is to solve a simplified problem on a
-% limited horizon and use the simplified solution now and then after stepping
-% forward in time and observing the system in a new position, to solve a revised
-% limited horizon problem and use the new solution.
-% Let us make this more precise.
-
-\section{Stochastic Optimal Control}
-Here we extend the Dynamic Programing theory to the case where the dynamics are
-governed by an SDE:
-\begin{equation}
-dX = F(X,t, u) dt + G(X,t) dW
-\end{equation} 
-i.e.\ we assume control affects only the drift term. This 
-assumption is not necessary, but it corresponds to our problems.
-
-For stochastic dynamics one can only optimize probabilistically and so
-$J$ becomes:
-\begin{equation}
-J[u] = \Exp \left[ \int_0^\tf L(x_s, u_s) \intd{s} + M(X_\tf)
-\right] 
-\label{eq:generic_objective_function_stochastic}
-\end{equation}
-while the running cost-to-go is:
-\begin{equation}
-J[u; x, t] = \Exp \left[ \int_t^\tf L(X_s, u_s) \intd{s} + M(X_\tf) \right], 
-\quad X_t = x
-\label{eq:generic_cost_to_go_stochastic}
-\end{equation}
-the value function is:
-\begin{equation}
-v(x,t) := \inf_{u \in \Udomain} J[u; x,t]
-\end{equation}
-
-Finally, we again have that $v$ satisfies a certain PDE:
-\begin{thm}[Stochastic HJB] 
-\label{thm:stochastic_hjb}
-\begin{equation}
-\begin{cases}
--\di_t v(x,t) &=  \max_{u(t) \in \Udomain(t)} \big\{ L(x,u)  +
-F(x,t,u) \cdot \di_x v
-\big\} + \frac{G^2(x,t)}{2} \di_x^2v
-\\
-v(x,\tf) &= M(x)
-\end{cases}
-\end{equation}
-\begin{proof}[Heuristic Derivation (from \cite{Evansb})] Suppose we are at time
-$t$ and $X_t = x$.
-
-Take a time increment $[t, t+h]$ and assume that during that time we apply a
-control $u$ and subsequently we apply the the optimal control $u^*$. The
-running-cost will then break down as: 
-$$
-J[u; x,t] = \Exp \left[ \int_t^{t+h} L(X_s, u_s) \intd{s}  + v(X_{t+h},
-t+h) \right] $$
-
-Since $v(x,t) = \inf J(u; x,t)$, we must have that:
-$$
-v(x,t) \leq  \Exp \left[ \int_t^{t+h} L(X_s, u_s) \intd{s}  + v(X_{t+h},
-t+h) \right] $$
-or, rearranging, that
-\begin{align*}
-0 \leq^&  \Exp \left[ \int_t^{t+h} L(X_s, u_s) \intd{s} \right]  
-+ \underbrace{\Exp \left[ v(X_{t+h}, t+h) - v(x,t) \right]}_{\Exp\left[ dv
-\right]}
-\end{align*}
-Now, $\Exp\left[ dv \right]$ can be expressed using Dynkin's Formula, as:
-$$
-\Exp[dv] = \int_t^{t+h} \di_t v +  F\cdot \di_x v + \frac{G^2 }{2}\cdot \di_x^2
-v \intd{s} $$
-Plugging that back, we get
-$$
-0 \leq \int_t^{t+h} L(X,u) +  \di_tv +  F\cdot \di_x v + \frac{G^2 }{2}\cdot
-\di_x^2 v \intd{s} $$
-Taking $h \ra 0$ we get 
-$$
-0 \leq  L(x,u) +  \di_tv(x,t) +  F\cdot \di_x v(x,t) + \frac{G^2 }{2}\cdot
-\di_x^2 v(x,t) $$
-and we conjecture that for the actual optimal control, the inequality becomes an
-equality:
-\begin{equation}
-\begin{cases}
-0 &=   L(x,u^*)+ \di_t v(x,t) +
- F(x,t,u^*)\cdot \di_x v(x,t) + \frac{G^2(x,t)}{2}\cdot \di_x^2 v(x,t)
-\\
-u^*(t) &= \argmax_{u \in \Udomain(t)}  
-\big\{L(x,u) + F(x,t,u)\cdot \di_x v \big\}
-\end{cases}
-\end{equation}
-\end{proof}
-\end{thm}
-% A rigorous proof can be found in \cite{Krylov2008}, sec???ch???
-
-\section{Deterministic Optimal Control in Infinite Dimensions}
-Now we discuss Optimal Control when the dynamics are deterministic but of
-infinite-dimension, i.e. the state evolution is a PDE instead of an ODE.  The
-reason we discuss this, is that optimization of ordinary SDEs in an open-loop
-control is actually a deterministic optimal control problem in infinite
-dimensions.
-
-Why?
-
-Since the objective is probabilistic and the probability density evolution is
-given by the (deterministic) Fokker-Planck PDE, we need to optimize a
-deterministic system whose dynamics are given by the Fokker-Planck equation.
-
-There is no all-encompassing generalization of Pontryagin's Principle to
-infinite dimensions, but a Maximum Principle can be derived for certain systems.
-To illustrate the idea we will discuss a simple case as in \cite{Palmer2011}.
-Our controlled SDE will be given by: 
-$$ dX_t = F(X,t, u) dt + G(X,t) dW, $$
-The corresponding probability density is given by: $$ \di_t \f=
--\sum_i \di_{x_i} \left[ F_i(x,t,u) \f(x,t) \right] + \sum_{i,j}  \di^2_{x_i
-x_j} \left[ D_{ij}(x,t) \f(x,t) \right] $$ which we will write as $$ \dot{\f} =
-\L_{u} [\f] $$ where $\L_{u}$ is the differential operator corresponding to the
-RHS of the Fokker-Planck equation parametrized by the control $u$. For now we
-will assume that there are no BCs, and the domain of $X$ is all of $\R^n$.
-
-A natural inner-product, $ \langle.,. \rangle$ on functions in $L^2[\R^n]$ is
-given by: $$
- \langle\psi,\phi  \rangle = \int_{\R^n} \psi(x) \phi(x) \intd{x}
-$$
-Our objective will be to minimize:
-\begin{equation}
-J[u] = \int_0^\tf  \langle L(\cdot,s,u), \f_u(s) \rangle \intd{s} +  
-\langle M(\cdot),\f(\tf) \rangle
-\label{eq:generic_objective_function_probabilistic}  
-\end{equation}.
-
-\Cref{eq:generic_objective_function_probabilistic} is a very natural cost
-function in the stochastic dynamics case, because it can also be written as: 
-$$
-J[u] = \Exp \left[ \int_0^\tf  L(X_s, s,u) \intd{s} + M(X_\tf) \right] 
-$$ which
-is just \cref{eq:generic_objective_function_stochastic}. We use
-\cref{eq:generic_objective_function_stochastic} if we have some feedback and
-want to use dynamic programming. We use
-\cref{eq:generic_objective_function_probabilistic} if we have no observations
-and need to run an open-loop control.
-
-Since, \cref{eq:generic_objective_function_probabilistic} is the objective
-function for deterministic dynamics, we can use Pontryagin's Principle adapted
-to the case of infinite dimensional dynamics (PDEs).
-
-Introduce an adjoint co-state, $\l \in \R^n$ and let the Hamiltonian, $\H$ be:
-$$
-\H(\f, \l, u)= \langle \phi, L + \Lstar[\l] \rangle$$
-where $\Lstar$ is the adjoint operator to $\L$ defined via:
-$$
-\langle \L [\phi], \l \rangle =\langle \phi, \Lstar[\l] \rangle
-$$
-
-Let $\l$ evolve (backwards) according to:
-\begin{equation*}
-\begin{cases}
--\di_t \l =& \Lstar[\l] + L
-\\
-\l(\tf) =& M(x)  
-\end{cases}
-\end{equation*}
-then the optimal control, $u$, is given via
-$$
-u = \argmin_u {\H(\f, \l, u)}
-$$
-
-An important fact, useful for computations, is that the total derivative of the
-objective with respect to the control is given by the partial derivative of the
-Hamiltonian with respect to the control:
-\begin{equation}
-\grad_u \big[J[u] \big] = \di_u \H(\f, \l, u)
-\label{eq:objective_gradient_wrt_control}
-\end{equation}
-
-So for a given $u$, we can compute $\f,\l$, calculate the objective gradient
-using \cref{eq:objective_gradient_wrt_control} and then iterate in the direction
-of descent (if we are minimizing).
-
-All that remains is to compute $\Lstar$. It turns out that if we do not have any
-BCs, then
-$$
-\Lstar[\l] = \sum_i F_i(x,t) \cdot \di_{x_i} \l + \sum_{i,j} D_{ij}(x,t)
-\di^2_{x_i x_j}\l 
-$$ 
-which is just the generator of $X$!
-
-We will see later, that things become more complicated when we need to
-consider a PDE's BCs.
-
-\section{Closed-loop vs.\ Open-Loop control}
-At this point we stop to briefly discuss the difference between
-open-loop control and closed-loop control.
-
-By closed-loop control, we mean that at any time we have observations of the
-system state, $x$ and its value can influence the control. In open loop control,
-we decide on the value of $u(t)$ a priori and apply it independently of the
-value of $x(t)$, based only on the knowledge of the system dynamics and the
-initial conditions.
-
-If the system dynamics are deterministic there is no added value in closed-loop
-control, since the state evolution is perfectly predictable and one can infer
-where $x$ will be based only on the dynamics and the applied control. For
-stochastic dynamics this is no longer true and there is something to be gained
-by observing the state, since different controls can be applied depending on the
-different realizations of the stochastic process driving the dynamics.
-Naturally, if one has observations of the system, one would prefer to use this
-and apply a closed-loop control. Thus, an open-loop control would be used
-only if one has no reliable knowledge of the system state.
-
-\section{A Stochastic Optimal Control Problem from Mathematical Neuroscience}
-Neurons relay information by means of voltage spikes - sudden sharp increases in
-voltage. Although many details remain unclear, the information content is
-thought to be contained in the length of the time-interval between these spikes.
-In the simplest case, this can be thought of as a rate  -the average number of
-spikes per time interval, but more complicated coding schemes are hypothesized to
-exist.
-
-Many experiments allow for manipulating an individual neural cell. A natural
-goal then is to make a cell produce a given spike train. This may arise, for
-example, in brain-machine interfaces or in artificial prosthetics.
-
-We now consider this problem in the framework of Stochastic Optimal Control and
-give both a feedback solution, when the cell voltage is explicitly observable as
-well as an open-loop solution when only the occurrence of a spike is
-robustly observable.
-
-\subsection{Problem Formulation}
-The most basic representative model for a neuron is the noisy
-leaky-integrate-and-fire model:
+\section{Problem Formulation}
+Given the following OU Model:
 \begin{equation}
 \begin{gathered}
-dX_s = (\a(t) - \frac{X_s}{\tc} ) \intd{s} + \b \intd{W_s},
+dX_s = (\a - \frac{X_s}{\tc} ) \intd{s} + \b \intd{W_s},
 \\
-X(0) = .0,
+X(0) = x_0,
 \\
-X(\ts) = \xth \implies  
-\begin{cases}
-X(\ts^+) = .0 &  
-\end{cases}
+\a > 0
 \end{gathered}
@@
 \end{equation}
-In this $X$ represents the membrane electric potential which tries to revert to
-$0$ with a time-const of $\tc$, $dW$ is a Brownian motion increment with
-intensity $\b$ and $\a(t)$ is our external control. Having last spiked at
-time $0$, the potential hits $\xth$ at time $\ts$, we record a spike
-of duration $\ts$, the potential resets to $0$ and we start all over again.
+and a hitting time, $\t$, defined by
+\begin{equation}
+X_\t = \xth 
+\quad \textrm{AND} \quad
+X_{s < \t} < \xth
+\end{equation}
+we want an analytical expression for:
+\begin{equation}
+\Exp[\t^2|x_0]
+\end{equation}
+as a function of $\a,\tc,\b$
 
-Suppose we want to ensure that spikes occur exactly at the predetermined
-times $\tns$. Then given the renewal nature of our spiking model, we can work
-sequentially, focusing only on the first of these intervals, of duration $t_1 =
-\T$, and then ask: how do we choose $\a(t)$ st. $\ts = \T$. There are several
-ways to specify what we mean by 'make $\ts = \T$', all of which result in optimal
-control problems. A natural one is a least squares solution:
+We can assume $\xth = 1.0$, but $x_0$ is arbitrary, that is - it is not
+necessarily the 'equilibirum' value $.0$, indeed we will need to know
+$\Exp[\t^2|x_0]$ for a whole range of $x_0$.
+
+\subsection{Analytical Expressions for $\Exp[\t^2|x_0]$}
+
+We will use the formulas/notation directly from \cite{Inoue1995}.
 \begin{equation}
-\a(t) = \argmin_{\a(t)} \{\, \Exp[(\ts- \T)^2] \,\}
-\label{eq:OC_LS_variance}   
+\Exp[\t^2] = \tc^2 \left[
+2\phi_1^2(\eta) -
+  \phi_2(\eta) -
+  2\phi_1(\eta)\phi_1(\xi)+
+  \phi_2(\xi)
+  \right]
 \end{equation}
+where
+\begin{eqnarray}
+\xi &=& \boxed{(x_0 - \a\tc) \sqrt{\frac{2}{\b^2 \tc}}}
+\\
+\eta &=&
+\boxed{(\xth -\a \tc )\sqrt{\frac{2}{\b^2 \tc}}}
+\\
+\phi_1(z)&=&
+ \frac{1}{2} \sum_{n=1}^{\infty} \frac{(z \sqrt{2})^n}{n!}
+ \Gamma\left(\tfrac{n}{2}\right)
+\\
+\phi_2(z)&=& 
+ \frac{1}{2} \sum_{n=1}^{\infty} \frac{(z \sqrt{2})^n}{n!}
+ \Gamma\left(\tfrac{n}{2}\right) \left( \psi\left(\tfrac{n}{2} \right) -
+ \psi(1) \right)
+\end{eqnarray}
+where $\Gamma, \psi$ are the
+\href{http://docs.scipy.org/doc/scipy/reference/generated/scipy.special.gamma.html#scipy.special.gamma}{Gamma}
+and
+\href{http://docs.scipy.org/doc/scipy/reference/generated/scipy.special.psi.html#scipy.special.psi}{digamma}
+ functions respectively. Note that $\xth > x_0 \implies \eta > \xi$.
 
-It is also natural to assume that our control has certain limitations. The
-simplest and most common are simple bounded constraints:
-\begin{equation}
-\a(t) \in [\amin, \amax]
-\label{eq:bound_constraints_alpha}
-\end{equation}
+In the paper \cite{Inoue1995}, we are cautioned that evaluation of $\phi_k(z)$
+directly is efficient and accurate as long as $z$ is positive and small. If $z$
+is negative we will have an alternating series and a cancellation might cause
+numerical errors.
 
-% \subsection{Optimal Control Formulation}
-In addition to \cref{eq:OC_LS_variance}, our objective will also contain a
-running cost based on the control. This regularizes the problem eliminating the
-subtleties of singular-control situations and it might be argued appropritate
-to avoid excessive control.
+Actually, I cannot get this to work at all!!!
 
-So we will seek an optimal control, $\a$ that solves:
-\begin{equation}
-\a(t) = \argmin_{\a(t)} \Big\{ \,
-\Exp\left[
-\e \int_0^\ts  \a^2 \intd{s}
-+  
-(\ts - \T \big)^2 \right]  \, \Big\}
-\label{eq:OC_LS_variance_energy}   
-\end{equation}
-where $\e$ measure how much weight we put on minimizing the energy expenditure
-and if $\e \rightarrow 0$, then we do not care at all.
+% Instead we will use the Backward
+% Kolmogorov equation, which is more useful since we need to calculate
+% $\Exp[T^2|x_0]$ for a whole interval of $x_0$'s.
 
-Experimentally it is often difficult to observe precisely the exact voltage
-trajectory, but the occurrence of a spike is much more easily observable. As
-such a closed-loop approach is not implementable since the state is unknown. In
-the sequel we will perform three calculations. First we will derive the HJB
-equation for a closed-loop control if the system was observable, and use this as
-a benchmark for the best possible performance of an open-loop control. Then we
-will calculate the most simple-minded open-loop control, which is just the
-deterministic control of the dynamics when assuming there is no noise, $\b=0$.
-THen we will calculate the optimal open-loop control by optimizing the
-transition density. And finally we will compare the performance of all three
-controls.
+\subsection{Backward Kolmogorov equation}
 
-\section{Closed-Loop Solution - Dynamic Programing}
-Given $t, X_t = x$, let $\t$ be the (unknown) remaining time to spike. For
-example, given $t=0, X=.0$, $\t = \ts$, the entire duration of the inter-spike interval.
-Starting from a general $t, x$, our remaining-cost objective, $J[\a; x,t]$,
-will be:
-\begin{equation}
-J[\a, x,t]  = \Big\{ \,
-\Exp\left[
-\e \int_t^\ts  \a^2 \intd{s}
-+
-(\t - (\T-t) \big)^2 |x,t
-\right] \, \Big\}
-\label{eq:OC_LS_mean_objective}   
-\end{equation}
-i.e.\ if time $t$ has elapsed without a spike, we now want to minimize
-the difference between $\t$ and ($T-t$), given the current state $x$. For now we
-assume that we know $x$, later we will treat the problem when that is not so.
-
-We are now in a position to set up the HJB equation for this problem. First
-introduce the value function, $v(x,t)$ as:
-\begin{multline}
-v(x,t) = \argmin_{\a(s \geq t)} \{J[\a, x,t] \} =
-\\ 
-\argmin_{\a(s \geq t)} 
-\Big\{ \,
-\Exp\left[
-\e \int_t^\ts  \a^2 \intd{s}
-+
-(\t - (\T-t) \big)^2 |x,t
-\right] \, \Big\}
-\label{eq:OC_LS_mean_value}   
-\end{multline}
-
-The corresponding HJB PDE for $v$ is:
-\begin{equation}
-\di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + \min_{\a \in [\amin, \amax]}\big\{
-\e \a^2 + (\a-\tfrac{x}{\tc}) \di_x v \big\} = 0
-\label{eq:OC_LS_mean_HJB}
-\end{equation}
-
-We have been canny enough to set up a quadratic optimization problem in the
-control, which means that the optimal control is given analytically via:
-\begin{equation}
-\a(x,t) = \frac{-\di_x v}{2\e} 
-\label{eq:OC_LS_variance_energy_quadratic_control}
-\end{equation}
-However we need to recall the bounds on the control, $\amin < \a < \amax$, and
-thus
-\begin{equation}
-\astar(x,t) = \min \left(\amax, \max\left(\amin, -\frac{\di_x v}{2\e}\right)
-\right)
-\label{eq:OC_LS_variance_energy_quadratic_control_bounds}
-\end{equation}
-With this the HJB becomes
-\begin{equation}
-\di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + 
-\e (\astar)^2 + (\astar-\tfrac{x}{\tc}) \di_x v 
-= 0
-\label{eq:OC_LS_mean_HJB_bounded_control}
-\end{equation}
-
-In the sequel we will drop the superscript star on the optimal control and
-just write $\a(t)$ for the solution of
-\cref{eq:OC_LS_variance_energy_quadratic_control_bounds}. 
-% Thus the control can be solved out of the HJB equation, which now becomes:
-% \begin{equation}
-% \di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + \tfrac{(\di_x v)^2}{\e}
-% -\tfrac{x}{\tc}\di_x v= 0
-% \label{eq:OC_LS_mean_HJB}
-% \end{equation}
-
-Now, we need to consider boundary conditions (BCs). If $X_t =
-\xth$ then we have a spike now and we know that $\ts = t$. Thus:
-$$
-v(\xth,t) = (t-\T)^2
-$$
-Now for $x \ll .0$, we can assume that $v$ doesn't really depend on $x$ so say 
-$$
-\di_x v(\xmin, t) = .0
-$$
-This is legitimate if we impose artificial reflecting boundaries for the
-diffusion at $\xmin$. So we need to determine for which $\xmin$ is a reflecting
-boundary consistent with the dynamics. For example, we can take $\xmin$ to be
-two standard deviations of the stationary distribution of the non-controlled, $\a=0$, \cref{eq:X_evolution_uo}.
-% Ooops,
-% we need $\di_x v \neq 0$ and here we are setting $\di_x v = 0$ in one of the BCs\ldots oh well.
-
-Finally we come to the Terminal Conditions (TCs) for $v$. The idea is simple: if
-we reach $\T$ without having spiked we will just slam on the accelerator: i.e.\
-$$t>T \implies \a(t) = \amax$$ Thus: $$ v(x,T) = \Exp[\t^2|x, \amax] $$.
-
-Thus we restate the HJB equation in its fully specified form:
-\begin{equation}
-\begin{gathered}
-\di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + 
-\e \a^2 + (\a -\tfrac{x}{\tc}) \di_x v 
-= 0
-\\
-\a (x,t) = \min \left(\amax, \max\left(\amin, -\frac{\di_x v}{2\e}\right)
-\right)
-\\
-\begin{cases}
-v(\xth,t) = (t-\T)^2  \quad & \textrm{upper BC}
-\\
-\di_x v(\xmin, t)  = .0  \quad &\textrm{lower BC}
-\\
-v(x,T)  = \Exp[\t^2|x, \amax]  \quad& \textrm{TC}
-\end{cases}
-\end{gathered}
-\label{eq:OC_LS_HJB_full}
-\end{equation} 
-We are solving $v(x,t)$ over $[\xmin, \xth] \times [0,T]$.
-
-\subsection{Calculating the TCs} 
-\label{sec:calculating_TCs}
-At this point we are in a position to solve the HJB \cref{eq:OC_LS_HJB_full}.
-But first we need a more concrete expression for the TCs. Let
-\begin{equation}
-\begin{array}{lcl}
-\Tone(x_0) &=& \Exp[\t|x_0, \amax] 
-\\
-\Ttwo(x_0) &=& \Exp[\t^2|x_0, \amax]  
-\end{array}
-\end{equation}
-where as before, $\t$ is the remaining time-to-spike, given $X_0 = x_0, \a(t) =
-\amax$. Naturally, $\Ti$ depend on $x_0$ and are affected by the value of the
-various parameters, $\a,\b,\t$.
-
-Actually, we only  need $\Ttwo$ for the HJB equation above, but we'll see that
-$\Tone$ will be needed as an auxiliary.
-
-Indeed, analytical expressions exist for $\Ti$, \cite{Inoue1995}. However in my
-experience, they are very finicky numerically, especially as they involve
-infinite alternating sums which can easily overflow in numeric calculations.
-Moreover, they are not particulary convenient if you need $\Ti (x_0)$ for a wide
-range of $x$'s as we do here. Instead we will turn to:
-
-\subsubsection{the Backward Kolmogorov Equation}
 Here we follow Jacobs, \cite{Jacobs}, in deriving and solving a differential
-equation for $\Ti(x_0)$.
+equation for $\Ttwo = \Exp[\t^2|x_0]$.
 
@@
 
-\subsubsection{A BVP for $\Ti(x_0)$}
-If we solve the PDE for $\G$ over $t \in [0,\infty)$, then we can calculate
+If we solve the PDE for $\G$ for $t \in [0,\infty)$, then we can calculate
 $\Ttwo(x_0)$ as:
@@
 
-Since $\Ttwo$ is an integral over time, we can reduce the PDE in
+Since $\Tone$ is an integral over time, we can reduce the PDE in
 \cref{eq:backward_SDF} for the distribution to a BVP for the moment(s) by
@@
 
-\Cref{eq:BVP_Tone,eq:BVP_Ttwo} are seemingly second order ODEs, but are
-actually first order and can be written as:
+\Cref{eq:BVP_Tone,eq:BVP_Ttwo} are seemingly 2nd order, but are actually
+first order and can be written as:
 \begin{equation}
 f(y)
-= (\a - \frac y\tc) S(y) 
-- \frac {\b^2} 2 S'(y) ; \quad  S(\xmin) = .0
+=U(y) S(y) 
+- D S'(y) ; \quad  S(\xmin) = .0
 \label{eq:BVP_T_generic}
@@
 and we can recover $\Ti$ from $S$ via:
-$$\Ti(x_0) =   -\int_{x_0}^\xth S(y)\intd{y}$$.
+$$\Ti(x_0) =   -\int_{x_0}^\xth S(\xi)\intd{\xi}$$.
 It is clear that thus defined, $\Ti$ will satisfy the BCs of
@@
 
-% Solving \cref{eq:BVP_T_generic} for $S$ gives: \begin{eqnarray} S(y) &=& -
-% \exp\left(\smallint \tfrac UD \right) \cdot \int_\xmin^{y} \exp\left(-\int
-% \tfrac UD\right) \frac{f(\xi)}D \intd{\xi}
-% \\
-% &=& - \exp\left( \frac{\a y - \frac{y^2}{2\tc}}D \right) \cdot \int_\xmin^{y} 
-%  \exp\left(- \frac{\a \xi - \tfrac{\xi^2}{2\tc}}D \right) \frac{f(\xi)}D
-% \intd{\xi} \label{eq:BVP_T_generic_soln} \end{eqnarray}
-We can solve \cref{eq:BVP_T_generic} analytically, but the analytical solution
-turns out to be awkward for numerical calculations, so we
-are better off just calculating the numerical solutions to \cref{eq:BVP_Tone,eq:BVP_Ttwo} directly
-via an ODE solver, starting from $S(\xmin) = .0$ and integrating until $\xth$.
+Solving \cref{eq:BVP_T_generic} for $S$ gives:
+\begin{eqnarray}
+S(y) &=& - \exp\left(\smallint \tfrac UD \right) \cdot \int_\xmin^{y} 
+\exp\left(-\int \tfrac UD\right) \frac{f(\xi)}D
+\intd{\xi}
+\\
+ &=& - \exp\left( \frac{\a y - \frac{y^2}{2\tc}}D \right) 
+ 		\cdot 
+ 		\int_\xmin^{y}   \exp\left(- \frac{\a \xi - \tfrac{\xi^2}{2\tc}}D \right)
+ 		\frac{f(\xi)}D \intd{\xi}
+\label{eq:BVP_T_generic_soln}
+\end{eqnarray}
 
-With this we can compare $\Ttwo$ between simulations and the numerical
-solutions see \cref{fig:Ti_empirical_vs_analytic}.
-\begin{figure}[htp]
+Actually, the analytical solution is not very useful for numerical calculations,
+we are better off just calculating the numerical solutions to
+\cref{eq:BVP_Tone,eq:BVP_Ttwo} directly.
+
+Comparisons between simulations and the numerical solutions can be found in 
+\begin{figure}[h]
 \begin{center}
@@
 {
+\label{fig:ab1}
 \includegraphics[width=0.48\textwidth]
@@
 {
+\label{fig:ab2}
 \includegraphics[width=0.48\textwidth]
@@
 \caption[]{Comparison between solutions of equations
-\cref{eq:BVP_Tone,eq:BVP_Ttwo} and empirical statistics from simulating the
-hitting times of \cref{eq:X_evolution_uo} for four different pairs of $\a,\b$.
-In the top panel we show the first moment of the remaining time-to-spike
-starting from $x$, $\Tone(x)$, and in the bottom panel, we show the second
-moment, $\Ttwo(x)$. We have used only $N=128$ samples to form the moment
-statistics.}
-\label{fig:Ti_empirical_vs_analytic}
+\cref{eq:BVP_Tone,eq:BVP_Ttwo} and empirical samples from simulating
+\cref{eq:X_evolution_uo} }
+\label{fig:label}
 \end{center}
@@
 
-\subsection{Solving the HJB - the numerical method for
-\cref{eq:OC_LS_mean_HJB_bounded_control}}
-We now have a PDE for $v$ and an algorithm for computing all the BCs and TCs of
-this PDE. It is time to discuss the numerical method for solving
-\cref{eq:OC_LS_HJB_full}.
-Since we are in 1-d and we have the simplest geometry, the simplest thing to do
-is try a centred Finite Difference scheme, upwind wherever necessary and hope
-for the best.
-
-Things will break down if $\e$ is very small in which case the quadratic term,
-$(\di_x v)^2 / \e$, will dominate and cause 'havoc' or if $\b$ is very small in
-which case the smoothing effect of the second derivative will be negligible and
-we will be in the zone of nonlinear conservation equations. Let us hope for the
-best now and assume these two problems do not arise, for the parameters we
-select.
-
-Recall that we are solving backwards in time. Then for variables at the next
-time-step (already computed) we will write $(\cdot)^+$; while variables at the
-previous time step (to-be computed), will be denoted as$(\cdot)^-$.
-
-Similarly given some node in space, we will write
-$(\cdot)_-,(\cdot)_0,(\cdot)_+$ as the left node, the node itself and the right
-node.
-
-Then our finite-difference scheme in time looks like:
-\begin{equation}
-\frac{v^- - v^+}{\Delta t} = \tfrac{\b^2}2  .5*(\di_x^2 v^- + \di_x^2 v^+) +
-(\a^+ - \tfrac{x}{\tc})  .5*(\di_x v^- + \di_x v^+) + \e (\a ^+)^2
-\end{equation}
-
-I.e.\ we have removed the non-linearity in $v$ by computing the control, $\a$ at
-the next, already computed, time-step and used central-differencing
-otherwise.
-
-For the spatial derivatives, we will not upwind for now and also use central
-differencing:
-$$\di_x v = \frac{v_+ - v_-}{2\Delta x}$$
-and as is standard:
-$$\di^2_x v = \frac{v_+ - 2 v_0 + v_-}{(\Delta x)^2}$$
-
-We will not discuss a detailed study of the convergence properties of the
-numerical method, but just demonstrate its performance on a representative
-parameter set in \cref{fig:HJB_attempt} we see that the method is stable for a
-particular parameter set.
-
-% \begin{figure}[ht]
-% \begin{center}
-%   \includegraphics[width=.7\textwidth]{Figs/KolmogorovBVPMoments_a=10_t=9_b=8.png}
-%   \caption[]{TCs for alpha = 1.0;  tauchar = .9; beta = .75;}
-%   \label{fig:TCs_extended_x}
-% \end{center}
-% \end{figure}
-
-Let's discuss the most salient features of \cref{fig:HJB_attempt}. At the end,
-when $t = \T$, the value function, $v$ is monotonely decreasing. This makes
-sense, becaue the lower $x$ is the more time on average will it take us to spike
-and the more delay will we have from our desired spike time. However, as we go
-back in time $v$ inverts, in the sense that for $t \ll \T$, we would like to
-be somewhat far from the boundary to avoid the risk of spiking early.
-
-Now focus on the controls, \cref{fig:HJB_attempt_control}: Naturally, at the
-end we have the control assume its maximum value, $\a(t>=\T) = \amax$. As we
-start to go back in time however, two things start to happen, the control $\a$
-decreases for $x \ll \xth$, and it becomes negative for $x \lesssim \xth$. 
-That is fully consistent with the problem objective - for
-$t<\T$, we want to distance $X_t$ from the threshold to avoid an early spike.
-\begin{figure}[h!]
-\begin{center}
-\subfloat[Solution, $v$]
-{
-\label{fig:HJB_attempt_soln}
-\includegraphics[width=0.48\textwidth]
-{Figs/HJB/soln_al=-20_au=20__t=5_b=8.png}
-}
-\subfloat[Controls, $\a$]
-{
-\label{fig:HJB_attempt_control}
-\includegraphics[width=0.48\textwidth]
-{Figs/HJB/control_al=-20_au=20__t=5_b=8.png}
-}
-\caption[]{The solution and controls for the HJB PDE parametrized by:
-$\tc = 0.50,  \beta = 0.75,
-\amin = -2.00, \amax = 2.00, \eps=0.10,
-\T = 1.80$. The $k$ index indicates the time-step, and recall that we are
-running backwards in time, so the computation actually happens from right-to-left, bottom-to-top}
-\label{fig:HJB_attempt}
-\end{center}
-\end{figure}
-
-\section{Deterministic Control}
-\label{sec:deterministic_ode_control}
-Now we will compute the optimal control in the case that $\b = 0$. The reason we
-do this is that since we have assumed that we have no feedback, we attempt to
-devise an open-controller by controlling the non-random dynamics. The
-calculation of this control turns out to be very simple:
-
-First let us assume that we actually can get to $\xth$ from $0$ in $\T$ time
-given $\amax$ and no noise.
-
-Then 
-\begin{equation}
-\begin{gathered}
-J = \int_0^\T \e \a^2 \intd{t}
-\\
-\st
-\begin{cases}
-\dot{x} = (\a(t) - \frac{x}{\tc})
-\\
-x_\T = \xth
-\\
-x_0 = .0
-\end{cases} 
-\end{gathered}
-\end{equation}
-The (Pontryagin) Hamiltonian becomes:
-\begin{equation}
-\H(x,p, \a) = \e \a^2 + p \cdot (\a - x/\tc )
-\end{equation}
-the adjoint is given by:
-\begin{equation}
--\dot{p} = \di_x \H = - p / \tc 
-\end{equation} 
-and the (unconstrained) control by:
-\begin{equation}
-\a(t) = \frac p {2 \e}
-\end{equation}
-which looks very similar to the stochastic case, once you put $p = \di_x v$.
-
-We can actually solve for the adjoint analytically: 
-$$
-p = p_0 \exp(t / \tc)
-$$
-where we can freely choose $p_0$ to later recover the TCs for $x$.
-$$
-\a(t) = \frac{p_0 \exp(t / \tc)}{2\e}
-$$
-Finally the solution for $x$ is:
-\begin{align*}
-\dot{x} & = \a(t) -  \frac{x}{\tc}
-\\ & = \frac{p_0 \exp(t / \tc)}{2\e} - \frac{x}{\tc}
-\\
-e^{t/\tc}x &= \int^t e^{s/\tc}\frac{p_0 \exp(s / \tc)}{2\e} \intd{s}
-\\
-x &= e^{-t/\tc} \cdot \frac{p_0 }{2\e} \tc \left[  \exp(2 t / \tc) -1 \right] 
-\\
-&= \frac{p_0 }{\e} \tc \cdot  \sinh (t/ \tc)
-\end{align*}
-And so: 
-$$
-x(\T) = \xth \implies p_0 = \e \frac{\xth  }{ \tc \sinh (\T/ \tc) }
-$$
-And we have a full solution.
-
-If we have saturation, i.e. if $\a$ hits its upper bound, then the situation is
-mildly more complicated. Since $p$ is monotonically increasing, as long as $p_0
-> \amin$ then $\a(t) > \amin \forall t$. So the saturation is simplified to:
-\begin{equation}
-\a(t) = \min \left(\frac{p_0 \exp(t / \tc)}{2\e} , \amax \right)
-\label{eq:deterministic_control_bounds}
-\end{equation}
-In particular, $\a$ will start by following $p(t)$ and then eventually hit its
-upper bound and stay there until $\T$.
-
-The only complication then is to calculate when $\frac{p_0 \exp(t / \tc)}{2\e} =
-\amax$ and then redo the calculation for $x$ with the new control in order to
-obtain the new value of $p_0$. Or we can just brute-force solve for $p_0$, by a
-simple shooting method. Using the latter approach, a solution-control pair for
-the parameter set we will use later is shown in \cref{fig:deterministic_soln_controls}. 
-%\usepackage{graphics} is needed for \includegraphics
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.7\textwidth]{Figs/ControlSimulator/deterministic_example.png}
-  \caption[]{An example trajectory for the deterministic dynamics. Above is the
-  state evolution, $x(t)$ given $\b=0$ and below is the applied control,
-  $\a(t)$. }
-  \label{fig:deterministic_soln_controls}
-\end{center}
-\end{figure}
-
-\section{Open-loop Stochastic Control}
-As we have mentioned before, when the value of $X_t$ is unknown, the best one
-can do is to use the transition density to perform the optimization. Since the
-transition density follows a (deterministic) PDE, we will thus apply a Maximum
-Principle for PDEs as a method of obtaining the optimal control.
-
-\subsection{Fokker-Planck equation for the state density evolution}
-
-We will need the transition density of $X$, conditional on no
-spikes having occurred. We will write this as:
-\begin{align*}
-f(x,t) \intd{x} &= \Prob[X_t\in \intd{x} | X_0 = .0, X_{s<t} < \xth]  								  
-\end{align*}
-Then $f$ satisfies a Fokker-Planck equation with absorbing boundaries
-\begin{equation}
-\begin{gathered}
-\di_t f(x,t) =
-				\frac{\b^2 }{2}\cdot \di^2_x f -  
-				\di_x \Big[ (\a(t)- \frac{x}{\tc})  \cdot f \Big]
-\\
-\\
-\begin{array}{lll}
-	&\textrm{st.\ }& 
-	\left\{ \begin{array}{lcl}
-	 f(x,0) &=& \delta(x) \quad (\textrm{delta function}
-	\\
-	(\a(t)- \tfrac x \tc)f - \di_x \tfrac {\b^2}2 f] \big|_{x=\xmin} &\equiv& 0
-	\quad (\textrm{reflecting BCs at some } \xmin )
-	\\
-	f(x,t) |_{x=\xth} &\equiv& 0 \quad (\textrm{absorbing BCs at } \xth)
-\end{array} \right. 
-\end{array}
-\label{eq:FP_pde_OU_PDF}
-\end{gathered}
-\end{equation}
-In theory, $\xmin = -\infty$, but in the numerics below we will need to
-truncate it to some finite value.
-
-The probability flow, $\phi$, here is
-$$
-\phi(x,t) = (\a- \tfrac x \tc)f - \di_x [\tfrac {\b^2}2 f]
-$$
-as such the $f$ dynamics can also be written as:
-$$
-\di_t f(x,t) = - \di_x \phi(x,t)
-$$
-and the lower BC is:
-$$
-\phi(x,t) |_{x=\xmin} \equiv 0
-$$
-
-We will also need a short hand notation for the differential operator on the
-right side of \cref{eq:FP_pde_OU_PDF}. Let
-$$
-\L_\a[\cdot] := \di^2_x [ \frac{\b^2 }{2} \cdot] -  
-				\di_x  [ (\a(t)- \frac{x}{\tc}) \cdot]
-$$
-This is a linear and bounded differential operator. . We will usually omit the
-subscript $\a$, but have written it now, since the operator is parametrized by
-$\a$, that is by the control.
-
-\subsection{Restating the objective in terms of $f$}
-Let us recall our objective, \cref{eq:OC_LS_variance}:
-$$
-J(\a) = \Exp\left[
-(\ts - \T \big)^2 
-+  
-\e \int_0^\ts  \a^2(s) \intd{s}
-\right]
-$$
-Rewriting this in terms of $f$ reads:
-\begin{multline}
-J(\a) = 
-\int_\xmin^\xth \Ttwo(x) f(x,\T) \intd{x}
-\\
-+ \int_0^\ts \phi(\xth, s) (s-\T)^2 \intd{s}
-\\
-+  \e \int_0^\ts  \a^2(s)  \int_\xmin^\xth f(x,s) \intd{x} \intd{s}
-\label{eq:OC_LS_variance_density}
-\end{multline}
-
-Let us explain in more detail each term on the RHS in
-\cref{eq:OC_LS_variance_density}:
-
-The first term
-$$
-\int_\xmin^\xth \Ttwo(x) f(x,\T) \intd{x}
-$$
-counts the cost of trajectories which spike too late:
-this cost is the expected squared time-to-hit starting at $x$, with  $\a
-= \amax$, weighted by the probability of $X_\T = x$ , $f(x,\T)$. Recall
-that we calculated $\Ttwo(x)$ in \cref{sec:calculating_TCs}.
-
-The second term
-$$
-\int_0^\ts \phi(s,t) (s-\T)^2 \intd{s}
-$$
-counts the cost of trajectories which spike too soon, that
-is the squared difference between realized spike time, $t$ and desired spike
-time $\T$, weighted by the probability of spike at $t$ which is just the outflow of
-probability, $\phi(\xth,t)$. Note further that due to the homogeneous
-Dirichlet BC on $f$ at $\xth$, the outflow is simply:
-$$
-\phi(\xth, t) = -\frac{\b^2 }{2} \di_x f(\xth, t) 
-$$
-
-Finally, the third term
-$$
-\e \int_0^\ts  \a^2(s)  \left[  \int_\xmin^\xth f(x,s) \intd{x} \right] \intd{s}
-$$
-is the energy cost: the inner integral, $\int_\xmin^\xth f(x,s) \intd{x}$
-takes into account that we incur an energy cost only for those trajectories
-which have not yet spiked.
-
-With that our control $\a$ will naturally be found via:
-$$
-\a = \argmin J[\a]
-$$
-
-\subsection{Optimizing with $f$ using a Pontryagin-type Maximum Principle}
-\label{sec:PDE_max_principle_for_pdf}
-By now our ostensibly stochastic optimal control problem modelled by SDEs
-has become a deterministic optimal control problem modelled by PDEs. Our control
-$\a$ influences the evolution density $f$ and via $f$, the integrals in the
-objective, $J$.
-
-As in the ODE theory, we will introduce an adjoint function, $p$, which will be
-crucial in setting up necessary conditions for the optimal control.
-
-The basic formulation for the adjoint PDEs and the abstract Hamiltonian can be
-found in \cite{Fattorini1999,Palmer2011}. However our cost functional is a
-little strange as it involves values of $\di_x f$ at the boundary. As such we
-will derive the optimality condition heuristically, appealing to the
-references for rigorous proofs.
-
-Before proceeding, one might wish to read
-\cref{sec:Pontryagin_heuristic_derivation} for a simpler version of the
-following derivation with exactly the same ideas.
-
-We start by assuming that $f,\a$ are the optimum trajectory/control pair and we
-ask what necessary conditions must be true for $f,\a$. In particular if we take
-a small variation around $\a$, we will have:
-\begin{align*}
-\a_\e = \a + \e \da
-\\
-f_\e = f + \e \df
-\end{align*}
-
-With that the variation of the cost with respect to the control at the optimal
-$\a$ can be calculated as:
-\begin{equation}
-\frac {dJ}{d\e} \Big|_{\e = 0} =
-- \int_0^\T (s-\T)^2 \frac {\b^2}{2} \di_x  \df \intd{s}  
-+ \int_\xmin^\xth \Ttwo\df\intd{x}
-+ \tilde{\e} \int 2 \a \da \intd{s}  
-\label{eq:J_variation_PDE}
-\end{equation}
-This must be non-negative for a true minimum, but before we can say more, we
-need to eliminate $\df$ from $\frac {dJ}{d\e}$! In order to eliminate $\df$
-from \cref{eq:J_variation_PDE} we will introduce an adjoint variable, $p$. What
-we would want from this adjoint variable $p$ would be that:
-\begin{equation}
-p(x,\T) = \Ttwo(x)
-\label{eq:adjoint_TC_desideratum}
-\end{equation}  
-and 
-\begin{equation}
-\di_t \left[ \int_\xmin^\xth  p \cdot\df \intd{x} \right] = (s-\T)^2 \frac
-{\b^2}{2}
-\di_x
-\df
- + \text{stuff}
-\label{eq:adjoint_dynamics_desideratum}
-\end{equation}
-where 'stuff' does not depend on $\df$.
-
-If we had such a $p$ satisfying
-\cref{eq:adjoint_TC_desideratum,eq:adjoint_dynamics_desideratum} then
-\cref{eq:J_variation_PDE} would read:
-\begin{equation}
-\frac {dJ}{d\e} \Big|_{\e = 0} =
-\tilde{\e} \int 2 \a\da  \intd{s} +   \int_0^\T  \text{'stuff'}\intd{s}
-\label{eq:J_variation_PDE_nodf}
-\end{equation}
-which does not depend on $\df$! and we would be on our way. In order to find
-$p(x,t)$ we will need to
-\begin{enumerate}
-  \item calculate $\di_t \df$
-  \item specify $\di_t p $ st. \cref{eq:adjoint_dynamics_desideratum} holds.
-\end{enumerate}
-
-\subsection{The PDE for $\df$}
-First, calculate the PDE that $\df$ satisfies:
-\begin{multline}
-\di_t f_\e =  \L_{\a_\e} [f_\e]
-\\
-= -\di_x [ (\a_\e - \tfrac{x}\tc) f_\e  - D \di_x f_\e]
-\\
-=  -\di_x [ (\a + \e\da - \tfrac{x}\tc) f_\e  - D \di_x f_\e]
-\\
-=  -\di_x [ (\a + \e\da - \tfrac{x}\tc) (f + \e \df)  - D \di_x (f + \e \df)]
-\\
-=  -\di_x [ (\a + \e\da - \tfrac{x}\tc) (f)  - D \di_x (f)]
-- \di_x [ (\a + \e\da - \tfrac{x}\tc) (\e \df)  - D \di_x (\e \df)]
-\\
-= \L_\a f  -\di_x [ \e\da\cdot  f]
-  +\e \L_\a \df + o(\e) 
-\\
-= \di_t f + \e \di_t \df
-\end{multline}
-Of course, we have $\di_t f = \L_\a f$, so we conclude that:
-\begin{equation}
-\di_t \df = \L_\a \df -\di_x [\da \cdot f] 
-\end{equation}
-Now what are the ICs? Well, both $f_\e$ and $f$ must satisfy the system ICs in
-\cref{eq:FP_pde_OU_PDF}, so $$\df (x,0) = 0$$.
-Finally, what are the BCs?
-On the upper side:
-$$
-f_\e |_\xth = 0 = [ f  + \e \df |_\xth = 0 + \e \df|_\xth \implies \df |_\xth =
-0 $$
-While on the lower side:
-\begin{align*}
-0 =& [ (\a_\e - x/\tc)f_\e - D \di_x f_\e |_\xmin
-\\
-=& (\a +\e \da - x/\tc)(f + \e \df ) - D \di_x (f + \e \df) |_\xmin
-\\
-=& [(\a +\e \da - x/\tc)(f) - D \di_x (f) |_\xmin
-\\
-& + \e [ (\a +\e \da - x/\tc)(\df) - D \di_x (\df) |_\xmin
-\\
-=& \cancel{[(\a - x/\tc)f - D \di_x (f) |_\xmin}
-\\
-&+ \e \da f|_\xmin
-\\
-&+ \e [ (\a - x/\tc)(\df) - D \di_x (\df) |_\xmin   + o(\e)
-\end{align*}
-
-so $\df$ satisfies the following system of equations:
-\begin{equation}
-\begin{gathered}
-\begin{aligned}
-\di_t \df 
-&= \L_\a \df -\di_x [\da \cdot f]
-\\
-&= -\di_x \left[ (\a - \tfrac x\tc)\df - \tfrac{\b^2}2 \di_x \df \right]
--\di_xf \cdot\da
-\end{aligned}
-\\
-\begin{cases}
-\df(x,0) &= 0
-\\
-[ (\a - \frac x \tc)(\df) - D \di_x (\df) |_\xmin &= -\da \cdot f|_\xmin
-\\
-\df|_\xth &= 0
-\end{cases}
-\end{gathered}
-\label{eq:FP_variation_near_optimum}
-\end{equation} 
-With \cref{eq:FP_variation_near_optimum} we are ready to design the dynamics of
-$p$ such that $\df$ is eliminated from the optimality equation,
-\cref{eq:J_variation_PDE}.
-
-\subsection{Calculating the adjoint }
-Recall our desideratum, we want
-$$
-p(x,\T) = \Ttwo(x) \quad (\text{required TCs})
-$$
-and
-$$
-\int_0^\T \di_t (p \cdot \df) = \int_0^\T(s-\T)^2 \frac {\b^2}{2} \di_x 
-\df \intd{s} + \text{stuff}
-$$
-This is equivalent to having
-$$
-\di_t (p \cdot \df) =  (s-\T)^2 \frac {\b^2}{2} \di_x 
-\df  + \text{stuff}
-$$
-Let us calculate $\di_t (p \cdot \df)$
-\begin{align*}
-\di_t (p \cdot \df) =& \df \cdot \di_t p + p \cdot   \di_t \df
- \\
- =& \df \cdot \di_t p + p \cdot \left( \L [\df] -\di_xf \cdot\da \right)  
- \\
- =& \df \cdot \di_t p + p \cdot \L  [\df] - \underbrace{ p \cdot \di_xf \cdot\da}_{\text{stuff}}
-\end{align*}
-Now our job is clear - we want 
-$$
-\di_t p = -\Lstar[p]
-$$
-Where $\Lstar$ is defined by:
-\begin{equation}
-\begin{gathered}
-\int_\xmin^\xth  \Lstar [p] \df \intd{x} 
-=
-\int_\xmin^\xth p \L [\df] \intd{x} - (s-\T)^2 \frac {\b^2}{2}\di_x \df +
-\text{possibly more stuff}
-\\
-\\
-\forall \df \quad \st \begin{cases}
-\df(x,t) |_{x=\xth} & = 0 
-\\
-(\a - \tfrac x \tc)\df - \di_x \tfrac {\b^2}2 \df \big|_{x=\xmin} &=  -\da
-\cdot f
-\end{cases}
-\end{gathered}
-\end{equation}
-where 'possibly more stuff' should be independent of $\df$.
-
-Let us take a deep breath and calculate $\Lstar$:
-\begin{align*}
-\int p \L [\df] \intd{x} =&
-\int p \underbrace{\left(\frac{\b^2 }{2}\cdot \di^2_x \df -  
-			 \di_x \big[ (\a- \frac{x}{\tc})\cdot \df \big]\right)}_{\L [\df]} \intd{x}
-\\
-=& [ -p ((\a- \frac{x}{\tc})  \cdot \df)\quad  \vert^\xth_\xmin
-+ \int \di_xp ((\a- \frac{x}{\tc})  \cdot \df) \intd{x}
-\\
-& + [p (\frac{\b^2 }{2}\cdot \di_x \df \quad \vert ^\xth_\xmin
-- [ \di_x p \cdot \frac{\b^2 }{2}  \df\quad \vert^\xth_\xmin 
-+ \int \frac{\b^2 }{2} \di^2_xp    \cdot \df \intd{x} 
-\\
-=& \int \Lstar [p] \df \intd{x} 
-+\underbrace{[-p(\a- \frac x \tc) \cdot \df 
-+ p \frac{\b^2}{2} \di_x \df 
--\di_x p \frac{\b^2}{2} \df \,\vert^\xth_\xmin }_{\text{BC terms}}
-\end{align*}
-
-Now we want to make the 'BC terms' equal to $(s-T)^2 \frac{\b^2}{2} \di_x \df$
-+ 'possibly more stuff'.
-In doing so, we will discover the correct BCs associated with the adjoint
-operator, $\Lstar$:
-\begin{align*}
-\textrm{BC terms} =& 
-\Big[ -p (\a- \frac{x}{\tc})  \cdot \df 
-+ p \frac{\b^2 }{2}\cdot \di_x \df   
-- \di_x p \cdot \frac{\b^2 }{2}  \df   \quad ^\xth_\xmin\vert 
-\\
-=& -\cancel{p (\a- \frac{x}{\tc})  \cdot \df \Big|_\xth}
- + p (\a- \frac{x}{\tc}) \cdot \df \Big|_\xmin
-\\ 
-& + [p \frac{\b^2 }{2}\cdot \di_x \df \Big|_\xth
- - [p \frac{\b^2 }{2}\cdot \di_x \df \Big|_\xmin
-\\
-&- \cancel{[ \di_x p \cdot \frac{\b^2 }{2}   \df\Big|_\xth}
- + [ \di_x p \cdot \frac{\b^2 }{2}   \df\Big|_\xmin
-\\
-=& \cancelto{-p (-\da f)}{-p ((\a- \tfrac x \tc)\df - \di_x \tfrac {\b^2}2
-\df) \Big|_\xmin} 
-+ [p \frac{\b^2 }{2}\cdot \di_x f \Big|_\xth
-+ [ \di_x p \cdot \frac{\b^2 }{2}   f\Big|_\xmin
-\\
-=& (s-T)^2 \frac{\b^2 }{2}\cdot \di_x \df\Big|_\xth 
-+  p (\da f)\Big|_\xmin
-\end{align*}
-
-Since we want this to be true for arbitrary $\df$, we need to enforce the
-following (adjoint) BCs: 
-$$
-\begin{cases}
-p |_\xth &= (s-T)^2
-\\
-\di_xp |_\xmin &= 0
-\end{cases}
-$$
-
-In summary, the adjoint PDE equation is given by: 
-\begin{equation}
-\begin{gathered}
-\begin{aligned}
-\di_t p =& - \Lstar[p]
-\\
-		=&  
-			- \Big[ \frac{\b^2 }{2}\cdot \di^2_x p +  
-			(\a(t)- \frac{x}{\tc})  \cdot \di_x p \Big] 
-\end{aligned}
-\\
-\st  
-\begin{cases}
-	p(x,\T) &= \Ttwo(x)
-	\\ 
-	\di_x p  \big|_{x=\xmin} &\equiv 0
-	\\
-	p \big|_{x=\xth} &\equiv (s-T)^2
-\end{cases}
-\label{eq:adjoint_pde_OU}
-\end{gathered}
-\end{equation}
-
-And most importantly:
-\begin{equation}
-\int \Lstar [p] \df \intd{x}
-=  
-\int p \L [\df] \intd{x} 
-  - (s-T)^2 \frac{\b^2 }{2}\cdot \di_x \df\Big|_\xth 
-  -  p  f \cdot \da \Big|_\xmin
-\label{eq:L_to_Lstar_relation}
-\end{equation}
-
-\subsection{A Pontryagin-type Minimum Principle for the optimal control $\a$}
-Using \cref{eq:L_to_Lstar_relation} and the TCs in \cref{eq:adjoint_pde_OU}, we
-will have:
-\begin{align*}
-\int \Ttwo \df \intd{x} =& 
-\int_\xmin^\xth p \df \intd{x}
-\\ =& \int_0^\T \di_t \int_\xmin^\xth p \cdot \df \intd{x} \intd{s}
-\\
-=& \int_0^\T \int _\xmin^\xth - \df \cdot \Lstar[p]
- + p \cdot \L_\a [\df] 
- -  p \cdot \di_x [\da \cdot f]  \intd{x}  \intd{s}
-\\
-=& 
-\int_0^\T \Big\{  (s-T)^2 \frac{\b^2 }{2}\cdot \di_x \df\Big|_\xth 
-+  p (\da f)\Big|_\xmin
-\\
-& \quad
-- \int _\xmin^\xth p \cdot \di_x [\da \cdot f] \intd{x} \Big\} \intd{s}
-\end{align*}
-
-Now put this back into the equation of optimality, \cref{eq:J_variation_PDE}.
-\begin{align*}
-\frac {dJ}{d\e} \Big|_{\e = 0} =&
-- \int_0^\T (s-\T)^2 \frac {\b^2}{2} \di_x  \df \intd{s}  
-+ \int_\xmin^\xth \Ttwo\df\intd{x}
-+ \tilde{\e} \int 2 \a \da \intd{s}
-\\
-=& 
-\int_0^\T \Big\{
- p (\da f)\Big|_\xmin
-+ \tilde{\e}  2 \a \da
-- \int _\xmin^\xth p \cdot \di_x [\da \cdot f] \intd{x} 
- \Big\} \intd{s}
- \\
-=& 
-\int_0^\T \da \Big\{
- p f \Big|_\xmin
-+ \tilde{\e}  2 \a
-- \int _\xmin^\xth p \cdot \di_x f \intd{x} 
- \Big\} \intd{s}
-\end{align*}
-
-Now we pull the famous fundamental lemma of the calculus of variations out and
-conclude that
-\begin{equation}
-\Big\{
- \tilde{\e}  2 \a(t)
-+ p f \Big|_\xmin
-- \int _\xmin^\xth p \cdot \di_x f \intd{x} 
-\Big\} = 0
-\quad \forall t \in [0,T]
-\label{eq:J_necessary_condition}
-\end{equation}
-
-Working in reverse we can recognize a Hamiltonian, $\H$ as:
-$$
-\H(f,p,\a) =  \int \cref{eq:J_necessary_condition} \intd{\a} 
-$$
-
-Naturally, $f,p$ must satisfy their respective
-\cref{eq:FP_pde_OU_PDF,eq:adjoint_pde_OU}.
-
-And our optimal control, $\a$,  can be found via:
-\begin{equation}
-\a (t) = \frac { - p(\xmin,t) f(\xmin,t)
-+ \int _\xmin^\xth p(x,t) \cdot \di_x f(x,t) \intd{x} 
-}{2\e}
-\label{eq:optimal_alpha_4_PDF_optimization}
-\end{equation}
-As always, subject to the appropriate bounds, \cref{eq:bound_constraints_alpha}.
-
-\subsection{An Iterative Numerical Solution}
-We have an entangled situation: $f,p$ depend on $\a$ through
-\cref{eq:adjoint_pde_OU,eq:FP_pde_OU_PDF}, while $\a$, the optimal control, is
-given by \cref{eq:optimal_alpha_4_PDF_optimization} which requires  the
-solution to $f,p$ . So we need some kind of iteration to find $\a$ and the corresponding $f,p$.
-
-The key to such an iteration is that the quantity 
-$$\di_\a \H =  \Big\{
- \tilde{\e}  2 \a(t)
-+ p(x,t) f(x,t) \Big|_\xmin
-- \int _\xmin^\xth p \cdot \di_x f(x,t) \intd{x} 
-\Big\}
-$$
-gives us the direction of increase
-and we can use it as a gradient in a descent algorithm. 
-This we can also write as:
-\begin{align*}
-\di_\a \H =&  \Big\{
- \tilde{\e}  2 \a(t)
-+ p(x,t) f(x,t) \Big|_\xmin
-- [p(x,t) f(x,t) \, _\xmin^\xth\Big|
-+ \int _\xmin^\xth \di_x p \cdot f(x,t) \intd{x} 
-\Big\}
-\\
-=&
- \Big\{
- \tilde{\e}  2 \a(t)
-- p(x,t) f(x,t)\Big|_\xth
-+ 2p(x,t) f(x,t) \Big|_\xmin
-+ \int _\xmin^\xth \di_x p \cdot f(x,t) \intd{x} 
-\Big\}
-\end{align*}
-which has certain numerical advantages.
-
-So the plan is: we choose an initial $\a_1(t)$ - the deterministic solution
-from \cref{sec:deterministic_ode_control} is a natural choice. Then solve for $f,p$.
-Then evaluate $J$ and iterate until convergence. For our convergence
-criterion we will take that the incremental difference in $J$ is smaller than
-some difference small tolerance. The algorithm is specified more
-precisely in Algorithm \ref{alg:hamiltonian_descent_4_PDF_OC}.
-\begin{algorithm}
-\begin{algorithmic}
-\State $J_0 = \infty $ 
-\State $\a_1 = $ deterministic $\a$, the optimal control given $\b=0$ from
-\cref{sec:deterministic_ode_control}
-\For { $k= 1\dots k_{\max}$}
-	\State $\di_t f_{k} = \L_{a_{k}} [ f_{k}]$
-	\State $-\di_t p_{k} = \Lstar_{a_{k}} [ p_{k}]$
-	\State $J_{k} \gets J[\a_k, f_k]$
-	\If {$|J_{k} - J_{k-1}| < tol $}
-		\State BREAK
-	\EndIf
-	\State $\da = \di_a \H(f_{k} ,p_{k},  \a_{k})$
-
-	\If {$\sup_t |\da(t)| < tol$}
-		\State BREAK
-	\EndIf
-
-	\State $\a_{k+1} \gets \a_{k} - s \cdot \da  $
-\EndFor
-\State \Return $J_k, \a_k$
-\end{algorithmic}
-\caption{Optimal Control Descent Algorithm}
-\label{alg:hamiltonian_descent_4_PDF_OC}
-\end{algorithm}
-
-We show one realization of $\di_a \H$ in \cref{fig:grad_aH_deterministic_alpha}.
-\begin{figure}[htp]
-\begin{center}
-\subfloat[$\a_0$]
-{\includegraphics[width=.49\textwidth]{Figs/FP_Adjoint/Deterministic_alpha_di_aH.png}
- }
-  \subfloat[$\a_1$]
-{  \includegraphics[width=.49\textwidth]{Figs/FP_Adjoint/First_control_iteration.png}
-}
-\caption[labelInTOC]{Iteratively calculating $\a$ for the
-open-loop optimal control. Plotted is an example of
-$-\di_a \H$ calculated at $\a(t) = \a_1(t)$, where $a_1$ is the deterministic control. The green curve, $-\di_a \H$
-indicates the direction that we want to change the control, if it is negative
-we decrease and if it is positive we increase it. On the right panel, b), we
-show the result of applying the incremental change by also plotting the
-resulting $\a_1(t)$. }
-\label{fig:grad_aH_deterministic_alpha}
-\end{center}
-\end{figure}
-
-Finally, we are ready to calculate the open-loop stochastic optimal control for
-this parameter set. The results are in
-\cref{fig:FP_adjoint_objective_control_convergence}.
-
-\begin{figure}[h]
-\begin{center}
-\subfloat[$\a_k(t)$ ]
-{
-\includegraphics[width=0.48\textwidth]
-{Figs/FP_Adjoint/ExampleControlConvergence_control.png}
-}
-\subfloat[$J_k$ ]
-{
-\includegraphics[width=0.48\textwidth]
-{Figs/FP_Adjoint/ExampleControlConvergence_objective.png}
-}
-\caption[ ]{The result of an entire optimization iteration, on the left the
-iterates of the control, $\a_k(t)$, on the right the progress of the
-objective with each iteration. The final control, in purple, is inhibitory at
-the beginning of the time interval, $\a < 0$, unlike the the deterministic
-control solution. 
-We see a significant
-reduction in the objective value, $J$, and thus an improvement, on the
-order of 50\%.}
-\label{fig:FP_adjoint_objective_control_convergence}
-\end{center}
-\end{figure}
-
-% \begin{table}[h] 
-% \centering
-% \begin{tabular}{lc}
-% Control Law & Squared Error \\
-% \hline
-% Deterministic &  0.647 \\
-% Stochastic &  0.336\\
-% Theory (Value function) & 0.285
-% \end{tabular}
-% \caption{Realized performance of the different control laws and the theoretical
-% expected performance of the stochastic law (last row)}
-% \label{tab:realized_avg_errors_det_vs_stoch}
-% \end{table}
+% In order to do so we must convert the problem to one starting from $.0$ Set $$ Y
+% = X - x_0 $$
 % 
-% \begin{figure}[h]
-% \begin{center}
-% \subfloat[A]
-% {
-% \label{fig:controlled_traj_ex1}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/example_controlled_trajectories_id1.png}
-% }
-% \subfloat[B]
-% {
-% \label{fig:controlled_traj_ex2}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/example_controlled_trajectories_id4.png}
-% }
-% \subfloat[C]
-% {
-% \label{fig:controlled_traj_ex3}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/example_controlled_trajectories_id3.png}
-% }
-% \caption[]{Examples for the controlled trajectories using both the deterministic
-% and the stochastic control approaches. The red vertical line in the plots
-% indicates the desired spike-time, $\T$. In A, B the performance of both
-% control laws is essentially the same, but in C we see the advantage of the
-% stochastic approach. $\tc, \b = [.75, 1.25]$.}
-% \label{fig:control_trajectories_examples}
-% \end{center}
-% \end{figure}
-% \begin{figure}[htp]
-% \begin{center}
-%   \includegraphics[width=.9\textwidth]{Figs/ControlSimulator/example_controlled_trajectories_hists.png}
-%   \caption[labelInTOC]{Histogram of the spike timing error for the
-%   deterministic (left) vs. the stochastic (right) control laws.}
-%   \label{fig:error_histograms_det_vs_stoch}
-% \end{center}
-% \end{figure}
-
-\section{Test Comparison of Different Controls}
-\label{sec:probabilistic_numerical_test}
-We will now compare the performance of the three obtained controls: the
-closed-loop stochastic obtained via the HJB equation, the open-loop
-deterministic control obtained when setting $\b=0$ and the open-loop stochastic
-control obtained by optimizing using the transition density.
-
-For our comparison we will sample $128$ realizations of the controlled system
-and apply in turn each of the three controls. Naturally, for each path, we will
-apply exactly the same random disturbances to each controller.
-
-We will set $\e = .001$ and $\amax = 2.0$. As such the energy cost is of
-secondary importance and the paramount effect on the objective is the difference
-$(\ts - \T)$, which we are trying to minimize. We set $\tc, \b = [.75, 1.25]$.
-
-The performance of the three controllers is given in
-\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch}. We see that while the
-open-loop stochastic control is not as good as the feedback stochastic
-control, it is much closer in performance to it than to the blind
-deterministic control.
-
-For illustration sake, we also show the error
-histograms in \cref{fig:error_histograms_det_vs_openloop_vs_stoch} and visualize
-some trajectories in \cref{fig:control_trajectories_examples}.
-
-\begin{table}[h] 
-\centering
-\begin{tabular}{lcc}
-Control Law & Empirical Squared Error & Theoretical Squared Error \\
-\hline
-Deterministic &  0.621 & - \\
-Open-Loop Stochastic & 0.356 & 0.382\\
-Feedback Stochastic &  0.288& 0.285\\
-\hline
-\end{tabular}
-\caption{Realized and theoretical performance of the different control laws. The empirical
-performance is obtained using 128 sample paths. The theoretical performance is
-found using the $J$ in for the open-loop stochastic control and the value
-function $v(x=0, t =0)$ for the closed-loop stochastic control.}
-\label{tab:realized_avg_errors_det_vs_openloop_vs_stoch}
-\end{table}
-\begin{figure}[h]
-\begin{center}
-\subfloat[]
-{
-\label{fig:controlled_traj_ex1}
-\includegraphics[width=0.33\textwidth]
-{Figs/ControlSimulator/3controls_example_trajectories_id3.png}
-}
-\subfloat[ ]
-{
-\label{fig:controlled_traj_ex2}
-\includegraphics[width=0.33\textwidth]
-{Figs/ControlSimulator/3controls_example_trajectories_id5.png}
-}
-\subfloat[ ]
-{
-\label{fig:controlled_traj_ex3}
-\includegraphics[width=0.33\textwidth]
-{Figs/ControlSimulator/3controls_example_trajectories_id6.png}
-}
-\caption[]{Examples for the controlled trajectories using the deterministic,
-open-loop stochastic and feedback stochastic control approaches. The
-black vertical line in the plots indicates the desired spike-time, $\T$.
-For the parameter values are $\tc, \b = [.75, 1.25]$. The three panels are
-three different realizations of the model dynamics.}
-\label{fig:control_trajectories_examples}
-\end{center}
-\end{figure}
-
-\begin{figure}[h]
-\begin{center}
-  \includegraphics[width=.95\textwidth]
-  {Figs/ControlSimulator/3controls_example_trajectories_hists.png}
-  \caption[labelInTOC]{Histogram of the spike timing error for the
-  deterministic (left) vs. open-loop stochastic(centre) vs. feedback stochastic
-  (right) control laws. This is for the same problem as in
-  \cref{fig:control_trajectories_examples}. We have used $N=128$ sample paths
-  to form the statistics}
-  \label{fig:error_histograms_det_vs_openloop_vs_stoch} 
-\end{center}
-\end{figure}
-
-\section{Research Goals}
-I hope to be able to use the techniques of open-loop stochastic control
-discussed herein to address problems in stochastic control where for whatever
-reasons, the dynamic programing formalism is not applicable.
-
-One such reason might be that the state is unobservable. An example of such a
-system is the control of blood glucose for diabetes or intensive care
-patients, where the patients' blood glucose is modelled to follow a stochastic
-differential equation and actual observations are expensive and invasive and
-thus rare, on the order of a few hours. In this scenario, the applied external
-glucose or insulin to the patient in the time in-between measurements can be
-thought of as the solution to an optimal control problem where the objective is
-to minimize deviations of the patient's blood glucose from some baseline.
-
-An alternative reason to use the open-loop control techniques is when optimizing
-with multi-dimensional systems at which point the dynamic programing approach
-suffers from 'the curse of dimensionality' which is the colloquial name for the
-exponential increase in necessary computing resources. If the optimality
-conditions for the open-loop control lend themselves to a Feynman-Kac-type
-representation, then one might replace solving the PDEs by an appropriate
-Monte-Carlo process and thus make the computations tractable.
-
-\clearpage
-\appendix
-\section{Heuristic Derivation of Pontryagin's Minimum Principle}
-\label{sec:Pontryagin_heuristic_derivation}
-Here we will show a heuristic derivation for the Pontryagin Minimum Principle
-for linear ODE dynamics. It should motivate the manipulations in
-\cref{sec:PDE_max_principle_for_pdf}.
-
-The following is almost a direct transcription of A.1 in
-\cite{Evansb}, with the added simplification of linear dynamics
-and linear cost and with the notation adapted to what we have used
-hereto.
-
-The dynamics are given by:
-$$
-\dot{x} = A(t, \a) x
-$$
-where $A$ is a matrix in $\R^{n_x \times n_x}$ whose values depend on the
-time, $t$ and the control $\a$. The cost, $J$, is given by:
-$$
-J[\a] = \int_0^\tf L(t,\a) \cdot x \intd{s} + M \cdot x(\tf)
-$$
-where $L(t,\a), M$
-are vectors in $\R^{n_x}$. Both the dynamics and the cost are linear in the state $x$.
-Suppose $x,\a$ are the optimal trajectory - then small variations in $\a$ can
-only worsen $J$. Let such an acceptable variation be given by:
-$$
-\a_\e = \a + \e \b
-$$
-for some small $\e$. We will have a corresponding dynamic, $x_\e$ st:
-$$
-x_\e = x + \e \dx+ o(\e)
-$$
-Now obtain a differential equation for the variation $\dx$:
-\begin{align*}
-\dot{x_\e} &=  \dot{x} + \e \dot{\dx} 
-\\
-&= A(\a_\e) x_\e = A(\a) x + \e \dot{\dx}
-\end{align*}
-Now divide through by $\e$ and let it go to zero:
-$$
-\implies \dot{\dx} = A \dx + [\grad_\a [ A(t,\a)x]] \cdot\b
-$$
-Note that $$\dx(0) = 0$$ since $x(0)$ is specified and so there is no
-variation there.
-
-With that the derivative of $J$ at $\a$ will read:
-\begin{multline}
-\frac {dJ}{d\e} \Big|_{\a} = \lim \frac{ J[\a_\e] - J[\a]}{\e} 
-= \int L(\a,t) \dx +  \grad_\a L \cdot \b \intd{s} + M \dx(\tf)
-\geq 0 
-\label{eq:J_variation_ODE}
-\end{multline}
-The problem is that the state variation $\dx$ also appears in this derivative,
-whereas we just want the control variation $\b$ to appear there.
-
-The trick then is to introduce the adjoint variable $p$ with appropriate
-dynamics such that all terms involving $\dx$ in \cref{eq:J_variation_ODE} are
-eliminated.
-
-'Clearly', we want to replace the term $ M \cdot \dx $ first. Thus we set 
-$$
-p(\tf) = M
-$$
-
-Now notice the following:
-$$p(\tf)\dx(\tf) = \cancelto{0}{p(0)\dx(0)} + \int_0^\tf d_t (p \cdot \dx)$$
-
-So if we make $\int d_t (p \cdot \dx) = - \int r(\a,t) \dx$ we would be
-eliminate the $\dx$ terms in the expresion for $dJ/d\eps$ entirely!
-
-Now the way is clear:
-
-Let 
-$$ 
-\dot{p} = -A^T p - L(\a,t)
-$$
-Then indeed:
-\begin{multline}
-d_t (p \cdot \dx) = \dot{p} \dx + p \dot{\dx}
-\\
-= (-A^T p - r)\dx + p (A\dx+ [\grad_\a [ A(t,\a)x]] \cdot\b)
-\\
-= -r\dx + p ([\grad_\a [ A(t,\a)x]] \cdot\b)
-\end{multline}
-That is almost what we wanted and it will work, since it only contains
-variations wrt the control $\b$, not the state, $\dx$.
-
-Thus
-\cref{eq:J_variation_ODE} is transformed to read:
-\begin{multline}
-\frac {dJ}{d\e} \Big|_{\a} = 
-\int r(\a,t) \dx +  \grad_\a r \cdot \b \intd{s} + \psi \dx(\tf)
-\\
-=
-\int r(\a,t) \dx +  \grad_\a r \cdot \b \intd{s} + p(\tf) \dx(\tf)
-\\
-= \int r(\a,t) \dx +  \grad_\a r \cdot \b \intd{s} - \int r \dx - p ([\grad_\a [
-A(t,\a)x]] \cdot\b)
-\\
-=
-\int_0^\tf  
-[\underbrace{p ([\grad_\a [A(t,\a)x]] ) + \grad_a r)}_{\di_\a \H} ] 
-\b \intd{s}
-\end{multline}
-As always in these variational arguments, the final point is that since $\b$ is
-arbitrary, we must have that the rest of the integrand be identically zero -
-thus we see where the control theory Hamiltonian, $\H$, comes from.
-
-Except for the nitty-gritty manipulations of the BCs in
-\cref{sec:PDE_max_principle_for_pdf}, what is done there is exactly what is done
-here - a heuristic derivation for a necessary condition for a minimum.
+% Then we have:
+% \begin{equation}
+% \begin{gathered}
+% dY_s = (\atilde - \frac{Y_s}{\tc} ) \intd{s} + \b \intd{W_s},
+% \\
+% Y_0 = .0,
+% \\
+% \atilde =  \a - \tfrac{x_0}{\tc}
+% \\
+% \yth = \xth - x_0
+% \end{gathered}
+% \end{equation}
+% 
+% $X$ hits $\xth$ whenever $Y$ hits $\yth$ and so we use equations 15-19 in
+% \cite{Ditlevsen2008a}:
+% \begin{equation}
+% \Exp[\t^2] = \tc^2 \left[ 
+% 2\phi_1^2(\eta) -
+%   \phi_2(\eta) - 
+%   2\phi_1(\eta)\phi_1(\xi)+
+%   \phi_2(\xi)
+%   \right]
+% \end{equation}
+% where 
+% \begin{eqnarray}
+% \xi &=& \atilde\tc \sqrt{\tfrac{2}{\b^2 \tc} }
+% =  \boxed{(\a - \tfrac{x_0}{\tc} ) \tc \sqrt{\tfrac{2}{\b^2 \tc}}}
+% \\
+% \eta &=&
+% (\yth - \atilde\tc)\sqrt{\tfrac{2}{\b^2 \tc}}
+% = \boxed{(\xth -\a \tc )\sqrt{\tfrac{2}{\b^2 \tc}}}
+% \\
+% \phi_1(z)&=&
+%  \frac{1}{2} \sum_{n=1}^{\infty} \frac{(z \sqrt{2})^n}{n!}
+%  \Gamma\left(\frac{n}{2}\right)
+% \\
+% \phi_2(z)&=& \phi_1(z) \left[ \psi\left( \tfrac{n}{2} \right) - \psi(1) \right]
+% \end{eqnarray}
+% where $\Gamma, \psi$ are the
+% \href{http://docs.scipy.org/doc/scipy/reference/generated/scipy.special.gamma.html#scipy.special.gamma}{Gamma}
+% and
+% \href{http://docs.scipy.org/doc/scipy/reference/generated/scipy.special.psi.html#scipy.special.psi}{digamma}
+%  functions respectively.
 
@@
 \bibliography{library,local}
+% \bibliography{local}
 

### 2013-08-22 OptSpike.brief.tex
Initial commit

--- 2013-08-22 CompExam.aviolov.tex
@@
 \usepackage[super]{nth}
+
 % \usepackage{appendix}
@@
 
+\usepackage{embedfile}
+\embedfile{OptSpike.brief.tex}
+
 \begin{document}
 
-\title{Stochastic Optimal Control - with a view towards spike control in neurons
-\\
-{\small Advanced Comprehensive Exam in Mathematics}
-\\
-{\small Dept. of Math - University of Ottawa}
+\title{Stochastic Optimal Control for spike control in single neurons
+\\
+\vskip3pt
+brief - .tex attached (see attachmets in acrobat)
 }
@@
 
-\abstract{We review main concepts from stochastic optimal control theory,
-including stochastic differential equations, dynamic programing and Pontryagin's
-Maximum Principle. We then discuss an optimal control problem in a simple
-mathematical model of a neuron where the feedback is minimal and we have a
-quasi-open-loop control which necessitates the use of a Maximum Principle for
-the system's transition density.}
-
-\tableofcontents
-
-\section{Overview}
-Stochastic Optimal Control in continuous time deals with finding
-the input which optimizes the performance of a continuous-time dynamical systems
-that contain a significant random component in their evolutions.
-
-Here we present a brief overview of the subject starting from the basic notions
-of Stochastic Differential Equations (SDEs), 
-and then introducing the main tools of Optimal Control, namely the
-Hamilton-Jacobi-Bellman equation of Dynamic Programming and the state-adjoint
-equations arising from Pontryagin's Maximum Principle. We will not state
-proofs, although we sometimes provide heuristic derivations of the main
-results.
-
-If I had to learn the material all over again, I would proceed my readings as
-follows: First I would read the online notes by Evans on both
-SDEs,\cite{Evansa}, and Optimal Control, \cite{Evansb}. I would then read
-Jacobs' book, \cite{Jacobs}, especially for the chapter on the Fokker-Planck
-equation (ch 7) and then Oksendal \cite{Oksendal2007} for a
-more thorough mathematical treatment of SDEs. Having familiarized myself with SDE's I would
-return to Optimal Control in the Deterministic Case, starting with \cite{Kirk2004} for the
-finite dimensional deterministic case (ODEs) and then \cite{Fleming1975} for the
-extension to the stochastic case. I would keep the following for reference \cite{Krylov2008},
-\cite{Hanson2007} and finally \cite{Fattorini1999} for the
-deterministic infinite-dimensional (PDEs) case.
-
-I have not used any particularly sophisticated numerical
-methods, but, for example, Hanson \cite{Hanson2007} discusses both the
-Euler-Maruyama scheme for simulating SDEs and the Crank-Nicholson Method for
-solving PDEs which is all I have used in my codes.
-
-All of the theoretical results presented in the sequel will be from
-the above references.
-
-\section{Stochastic Differential Equations}
-In view of our ultimate goals, we will restrict ourselves to  stochastic
-processes whose sample paths have continuous paths, i.e.\ to SDEs driven by
-Brownian motion.
-
-Since the reader is more likely well familiar with the material in this section,
-we will not provide proofs, but only state the results with a view towards
-establishing the notation for the sequel.
-
-We will assume that the reader is familiar with the following concepts:
-\begin{itemize} 
-  \item a probability space, $\{\O, \sAlg, P\}$ consisting of a
-  probability space, a sigma-algebra and a probability measure
-  \item a continuous-time stochastic process, $X_t$
-  \item a filtration $\Fil(t)$ and in particular the filtration generated by a
-  stochastic process.
-\end{itemize}
-
-\subsection{The Wiener Process and the Ito Integral}
-The Wiener Process is the fundamental building block of the stochastic calculus,
-it is often called Brownian Motion and we will denote it $\{W_t\}_{t\geq 0}$. It
-satisfies:
-\begin{defn}Wiener Process, $W_t$:
-\begin{enumerate}
-  \item $W_0 = 0$
-  \item $W_t - W_s = N(0, |t-s|)$ , i.e. normally distributed with mean 0,
-  variance $|t-s|$)
-  \item $\forall \{t_i\}_1^N, \quad \{W_{t_i} - W_{t_{i-1}} \}_2^N \sim$
-  independent, i.e $W_t$ has independent increments
-\end{enumerate}
-\end{defn}
-The fact that the finite incremental distributions suffice to specify a unique
-continuous-time stochastic process is known as Kolmogorov's extension theorem.
-\begin{ex}[from \cite{Evansa} \#21.] Inverted Wiener Process:
-\\
-Let $W_t$ be a 1-d Wiener Process and define:
-\def \Wbar {{\bar{W} }}
-\begin{equation*}
-\Wbar(t) := \begin{cases}
-t W(1/t) & t >0
-\\
-0 & t = 0
-\end{cases}
-\end{equation*}
-Show $\Wbar(t)- \Wbar(s) \sim N(0, t-s)$.
-
-\begin{proof}[Soln]
-\begin{align*}
-\Wbar(t) - \Wbar(s) =& t W(1/t) - s W(1/s), \, t>s
-\\
-=& (t-s)W(1/t) + s W(1/t) - sW(1/s) & \text{recall:} 1/s > 1/t
-\\
-=& (t-s)N(0, 1/t) - sN(0, 1/s -1/t) & \text{by indnt increments}
-\\
-=& N\left(0, \frac{(t-s)^2}t + \frac{s^2 }{ 1/s -1/t} \right) & \text{
-addition of indnt normal RVs.}
-\end{align*}
-With some algebra, the variance in the last expression reduces to $t-s$
-\end{proof}
-\end{ex}
-
-Now, we collect a few more relevant properties of $W_t$:
-\begin{enumerate}
-  \item The sample paths $W_{[0,\infty)}(\o)$
-are almost surely (a.s.) continuous and are in fact Holder continuous for any
-exponent $\g < 1/2$
-\item The sample paths $W_{[0,\infty)}(\o)$ are nowhere differentiable
-\item $W_t$ is a Markov process: $\Prob[W_t \in B | \s(W_{s' \leq s })] =
-\Prob[W_t \in B \,| \,W_s]$, for any Borel set, $B \subset \R$, where $\s(W_{s'
-\leq s })$ is the filtration generated by the $W_t$
-\end{enumerate}
-
-We now turn to defining stochastic integrals based on the Wiener Process.
-
-\begin{defn} Progressively Measurable Functions:
-
-Let $\Fil_t$ be the filtration generated by the Wiener Process.
-
-Let $X_t$ be a stochastic process which is $\Fil_t$-measurable $\forall t$ and
-which is jointly measurable in $(t,\o)$. We call such an $X_t$
-\emph{progressively measurable}
-\end{defn}
-
-\begin{defn} $\Ltwopm, \Lonepm$
-
-We define $\Ltwopm[0,T]$ as the space of all progressively measurable
-$X$ such that
-\begin{equation*}
-\Exp \left[ \int_{[0,T]} X_t^2 \intd{t} \right] < \infty
-\end{equation*}
-
-Similarly, we define $\Lonepm[0,T]$ as the space of all progressively measurable
-$X$ such that
-\begin{equation*}
-\Exp \left[ \int_{[0,T]} X_t \intd{t} \right] < \infty
-\end{equation*} 
-\end{defn}
-
-$\Ltwopm$ will be the class of functions for which the Ito integral is
-well-defined. 
-\begin{defn} Ito Integral:
-\label{defn:ito_integral}
-
-Let $P^n := {a = t^n_1 \ldots t^n_{m_n} = b}$ be a partition of the interval
-$[a,b] \subset [0, \infty)$. Let $|P_n| = \sup_i|t_i - t_{i-1}|$. 
-Take $|P_n| \rightarrow_n 0 $ and consider an $ X_t \in \Ltwopm[a,b]$,
-
-then
-\begin{equation}
-\int_{[a,b]} X_t \intd{W} := \lim_{n \rightarrow \infty}  
-\sum_{i=1}^{m_n} X_{t_i}\left( W(t_{i+1}) - W(t_{i})\right)
-\end{equation}
-
-\end{defn}
-
-To be precise, our definition is actually a theorem, and the real
-definition is one that uses step functions and passes to the limit.  Also we
-will write the limits of integration $\int_0^T \cdot  \intd{W}$ or $\int_{[0,T]}
-\cdot  \intd{W}$ interchangeably.
-% To justify def'n \ref{defn:ito_integral}, we recall:
-% \begin{lemma}[Quadratic Variation]
-% \begin{equation}
-% \lim_{n \rightarrow \infty} \sum_{i=1}^{m_n} \left( W(t_i) - W(t_{i-1})\right)^2
-% = b-a \quad \text{in }  L^2(\O)
-% \end{equation}
-% \end{lemma}
-% The above can be read, heuristically as $\lim \sum (\Delta W)^2 \rightarrow \sum
-% \Delta t$, which is the meaning behind the colloquial $\intd{W} =
-% \sqrt{dt}$.
-
-Again, we state without proof a few interesting properties of $\int X \intd{W}$:
-\begin{thm} Ito Integral Properties
-
-\begin{enumerate}
-  \item $\Exp[\int_0^T X \intd{W} ] = 0$ 
-  \item $\Exp[\left(\int_0^T X \intd{W}\right)^2 ] = \Exp[\int_0^T X^2
-  \intd{t}]$
-  \item $I(t) = \int_0^t X \intd{W} $ is a martingale 
-  \item $I(t) = \int_0^t X \intd{W} $ has continuous sample paths
-\end{enumerate}
-\end{thm}
-
-\subsection{Ito SDEs and Ito's Lemma}
-\begin{defn} an Ito SDE:
-
-We write
-\begin{equation}
-dX_t = Fdt + G dW
-\end{equation}
-on $0 \leq t \leq T$, if $X$ is a real-valued stochastic process satisfying:
-\begin{equation*}
-X(r) = X(s) + \int_{[s,r]} F \intd{t} + \int_{[s,r]} G \intd{W}
-\quad 0\leq s \leq r \leq T
-\end{equation*}
-for some $F \in \Lonepm[0,T]$ and $G \in \Ltwopm[0,T]$
-\end{defn}
-
-We are now ready to present the celebrated Ito Lemma which is the chain-rule of
-stochastic calculus:
-
-\begin{thm}[Ito Lemma] Suppose $X$ satisfies the stochastic differential $dX_t =
-Fdt + G dW$ as above and take $u(x,t) \in C^{2,1}[ \R \times [0,T]]$.
-
-Set $Y(t) = u(X,t)$
-\\
-then
-$$
-dY =  \left( \di_t u + \di_x u \cdot F + \di^2_x u \cdot \frac{G^2}2 \right)
-\intd{t} + \left(   \di_x u\cdot G  \right)\intd{W}
-$$
-\end{thm}
-
-% We now present two examples, one fundamental and one more involved:
-% \begin{ex}[pg 67 in \cite{Evansa}] The 'Stochastic' Exponential:
-% 
-% Let $Y = \exp(\l W - \l^2 t / 2)$. In the notation of the above Thm, we have:
-% $dX = dW, F=0, G=1$ and $u(x,t) = \exp(\l x - \l^2 t / 2)$.
-% 
-% then
-% \begin{align*}
-% dY =& 
-% \left( \di_t u + \di_x u \cdot F + \di^2_x u \cdot \frac{G^2}2 \right)\intd{t} +
-% \left(   \di_x u \cdot G  \right)\intd{W}
-% \\
-% =&
-% \left( -\frac{\l^2 t}{2}\exp(\l x - \l^2 t / 2) + \l^2\exp(\l x - \l^2 t / 2)  \right)\intd{t}
-% \\ 
-% &+ \left(   \l \exp(\l x - \l^2 t / 2) \right)\intd{W}  
-% \\
-% =&  \l \left( \exp(\l x - \l^2 t / 2) \right)\intd{W}  
-% \\
-% =& \l Y dW
-% \end{align*}
-% \end{ex}
-% 
-% \begin{ex}[pg. 67 in \cite{Evansa}] Hermite Polynomials:
-% 
-% Let $h_n(x,t) = \frac{(-t)^n}{n!} e^{x^2/2t} \di^n_x \left( e^{x^2/2t} 
-% \right)$ be the $n$th Hermite polynomial
-% \\
-% then
-% $$
-% \int_0^t h_n(W, t) \intd{W} = h_{n+1}(W(t), t)) 
-% $$
-% i.e. in Ito calculus, the Hermite polynomials are like the power monomials
-% $\frac{t^n}{n!}$ in the ordinary calculus.
-% \begin{proof} Since
-% $$
-% d_\l^n [\exp(- \frac{(x-\l t)^2}{2t} ] |_{\l=0} =
-%  (-t)^n d_x^n [ \exp(-\frac{x^2}{2t})] $$
-% We will have:
-% \begin{align*}
-% d_\l^n [\exp(\l x - \l^2 t / 2) ] |_{\l=0} &= (-t)^n  e^{x^2/2t} d_x^n [
-% \exp(-\frac{x^2}{2t})]
-% \\
-% &= n! h_n(x,t)
-% \end{align*}
-% Thus we can expand $\exp(\l x - \l^2 t / 2) $ in a Taylor series at $\l=0$ as:
-% $$
-% \exp(\l x - \l^2 t / 2) = \sum_n \l^n h_n(x,t)
-% $$
-% Now recall $Y$ from the previous example:
-% \begin{equation*}
-% Y(t) = \exp(\l W(t) - \frac{\l^2 t }2) = \sum_n \l^n h_n(W(t),t)
-% \end{equation*}
-% which satisfies:
-% \begin{equation*}
-% Y(t) = 1 + \l \int Y \intd{W}
-% \end{equation*}
-% Naturally we will plug the series expansion in the SDE to obtain:
-% \begin{align*}
-% \sum_{n=0}^{\infty} \l^n h_n(x,t) =&
-% 1 + \l \int_0^t  \sum_{n=0}^{\infty} \l^n
-% h_n(x,t)\intd{W}
-% \\
-% =& 1 + \sum_{n=1}^{\infty} \l^n\int_0^t 
-% h_{n-1}(x,t)\intd{W} \quad {\textrm{bump $n$ up by one}}
-% \end{align*}
-% Equating the coefficients for $\l^n$ on both sides leads to the result
-% \end{proof}
-% \end{ex}
-
-Finally, we are in a position to define an Ito SDE:
-
-\begin{defn}[Ito SDE] Let
-$F(x,t):\R\times [0,T]\ra \R,  G(x,t):\R\times [0,T]\ra \R,$ be given functions.
-We say that a stochastic process $X$ satisfies:
-\begin{equation}
-dX =F dt + G dW
-\end{equation} 
-over $[0,T]$ if
-\begin{enumerate}
-  \item $X$ is progressively measurable wrt. $\Fil_t$
-  \item $F(X,t) \in \Lonepm$
-  \item $G(X,t) \in \Ltwopm$ 
-  \item $X_t = X_0 + \int_0^t F(X_s, s) \intd{s} + \int_0^t G(X_s, s) \intd{W_s}$
-\end{enumerate}
-\end{defn}
-
-% \begin{ex} The solution to 
-% $$
-% \begin{cases}
-% dX = F(t)X dt + G(t) X dW
-% \\
-% X_0 = 1
-% \end{cases}
-% $$
-% is 
-% $$
-% X_t = \exp\left( \int_0^t F(s) - \frac{G^2(s)}{2} \intd{s} + \int_0^t G
-% \intd{W} \right) $$
-% \begin{proof}
-% Let $Y = \int_0^t f(s) - \frac{g^2(s)}{2} \intd{s} + \int_0^t g
-% \intd{W} $
-% and guess that $X = e^Y$
-% then, using Ito's lemma:
-% \begin{align*}
-% dX =& (\di_y [e^y] dY + \frac{1}{2}\di^2_y[e^y]  g^2 dt
-% \\ =&
-% (e^y  f dt - e^y g^2 / 2) dt + e^y g dW + \frac{1}{2} e^y \cdot  g^2 dt
-% \\ =& e^y \left( f dt +g dW \right)
-% \\ =& X \left( f dt +g dW \right)
-% \end{align*}
-% \end{proof}
-% \end{ex}
-
-% \begin{ex}[Brownian Bridge, pg 80 in \cite{Evansa}] The solution of the SDE
-% $$
-% \begin{cases}
-% dB = -\frac{B}{1-t} dt + dW
-% \\
-% B_0= 0
-% \end{cases}
-% $$
-% is
-% $$
-% B(t) = (1-t) \int_0^t \frac{1}{1-s} dW
-% $$
-% \begin{proof}
-% It is not clear how to apply Ito's lemma to this problem, so instead we just
-% directly calculate the differential:
-% \begin{align*}
-% B(t^+) - B(t) =& (1-t^+) \int_0^{t^+} \frac{1}{1-s} dW - (1-t) \int_0^t
-% \frac{1}{1-s} dW
-% \\ =&
-% (1-t^+) \int_t^{t^+}\frac{1}{1-s} dW  - (t^+-t) \int_0^t
-% \frac{1}{1-s} dW
-% \\ \approx&
-% \int_t^{t^+} dW - \Delta t \frac{B(t)}{1-t}  
-% \end{align*}
-% \end{proof}
-% \end{ex}
-% Actually we can say a bit more about the Brownian bridge, in particular we
-% will calculate $\Exp[B^2]$
-%  \begin{ex}[Evans .37]
-% \begin{align*}
-% \Exp[B^2] =& \Exp\left[\left( (1-t) \int_0^t \frac{1}{1-s} dW \right)^2 \right]
-% \\
-% =&(1-t)^2 \Exp\left[ \int_0^t\left(  \frac{1}{1-s}\right)^2 \intd{s} \right]
-% \\
-% =&(1-t)^2 [ \frac{1}{1-t} - 1 ]
-% \\
-% =&(1-t) - (1-t)^2
-% \end{align*}
-% 
-% Further note that  $\lim_{t\ra 1^-}\Exp[B^2] = 0$ and that $B$ is a martingale
-% and so its first moment is always equal to its initial value, $0$. So indeed the 2nd
-% moment is the variance and with the variance converging to zero, we can use
-% Chebyshev's inequality to conclude that $\lim_{t\ra 1^-}B(t) = 0 $(a.s). This
-% justifies the name of the Brownian bridge - it is a normally distributed RV.
-% which is clamped to equal 0 at both $t= \{0,1\}$.
-% \end{ex}
-
-As an example we will discuss the Ornstein-Uhlenbeck process, which is the
-basis for the models we will face later on.
-\begin{ex}[O-U Process] Let $X$ follow:
-\begin{equation*}
-dX = \th (\a - X_t) dt + \b dW
-\end{equation*}
-with an initial condition, $X_0$, which may be an arbitrary distribution
-independent of the Wiener Process,$W$.
-We can solve this as follows:
-\begin{align*}
-dX =& \th (\a - X_t) dt + \b dW
-\\
-dX + \th X_t) dt=& \th \a dt + \b dW
-\\
-e^{\th t} dX + e^{\th t}\th X_t) dt=& \th e^{\th t}\a dt + \b e^{\th t} dW
-\\
-Xe^{\th t} - X_0 = & \int \th e^{\th t}\a dt +  \int \b e^{\th t} dW
-\\
-X_t =& e^{-\th t} X_0 + \a(1-e^{-\th t}) +  \frac{\b e^{-\th t}}{\sqrt{2\th}}
-W(e^{2\th t}-1)
-\end{align*}
-which means that $X$ forgets its initial conditions exponentially fast and
-converges to a normal random variable with mean $\a$ and variance
-$\tfrac{\b^2}{2\th}$
-\end{ex}
-
-At the end of this sub-section, we state Ito's Lemma with multiple Wiener
-processes and then for a multidimensional state.
-
-\begin{thm}[Ito Lemma for $dW \in  \R^{m}$] Suppose $X$ satisfies the stochastic
-differential $dX_t = F dt + G dW$, where $ F \in \R, G \in \R^{1\times m}$ and
-$dW = \left(dW^{(i)}\right) \in  \R^{m}$ is a vector of independent Wiener
-Processes. Take $u(x,t) \in C^{2,1}[ \R \times [0,T]]$. Let $D :=\left[{\sum_k
-G_{k} G_{k} } \right] \in \R$
-
-Set $Y(t) = u(X,t)$
-\\
-then
-$$
-dY =  \left( \di_t u +  \di_{x} u \cdot F + 
-  \di^2_{x} u \cdot D \right)
-\intd{t} +
- \left(  \di_{x} u  \cdot  G \right) dW  
-$$
-\end{thm}
-
-\begin{thm}[Ito Lemma for $X \in R^n$]
-Suppose $X$ satisfies the stochastic differential $dX_t = F dt + G dW$, where $
-F \in \R^n, G \in \R^{n\times m}$ and $dW = \left(dW^{(i)}\right) \in  \R^{ 
-m}$ is a vector of independent Wiener Processes. Take $u(x,t) \in C^{2,1}[ \R^n
-\times [0,T]]$. Let $D_{ij} := \tfrac{1}{2} \left[{\sum_k G_{ik} G_{jk} }
-\right]_{ij} \in \R^{n\times n} $
-
-Set $Y(t) = u(X,t)$
-\\
-then
-$$
-dY =  \left( \di_t u + \sum_i \di_{x_i} u \cdot f_i + 
-\sum_{i,j} \di^2_{x_i x_j} u \cdot D_{ij} \right)
-\intd{t} +
- \left(  \sum_{ij} \di_{x_i} u  \cdot G_{ij} dW^{(j)} 
-\right)$$
-\end{thm}
-
-\subsection{Fokker-Planck and Kolmogorov's Backward Equations}
-The Fokker-Planck equation and Kolmogorov's Backward equation describe the
-forward (resp. backward)  evolution of the probability density of $X_t$
-
-For the rest of this section we will work in $ \R^n$, i.e. $X$ will be an $n-$
-dimensional vector that satisfies the stochastic differential 
-\begin{equation}
-dX_t = F(X,t) dt + G(X,t) dW,
-\label{eq:generic_Ito_SDE_Rn}
-\end{equation}
-$F \in \R^n, G \in \R^{n \times m}$,
-$dW = (dW^{(i)}) \in  \R^{  m}$ is an $m$-dimensional Brownian motion and we
-write $D_{ij} :=
-\tfrac{1}{2} 
-\left[{\sum_k G_{ik} G_{jk} } \right]_{ij} \in \R^{n\times n} $.
-
-% Let $$\F(x,t| y,s) = \Prob[X_t \preceq x| X_s = y] $$ be the forward transition
-% distribution and
-
-Let $$ \f(x,t| y,s) \intd{x} =  \Prob[X_t \in \intd{x} | X_s = y] $$ be the
-transition probability density. Then $\f$ satisfies:
-\begin{equation}
-\di_t \f= -\sum_i \di_{x_i} \left[ F_i(x,t) \f(x,t) \right] 
-+ 
-\sum_{i,j}  \di^2_{x_i x_j} \left[ D_{ij}(x,t) \f(x,t) \right]
-\label{eq:FP_pde}
-\end{equation}
-This is called the \emph{Fokker-Planck} or \emph{Forward Kolmogorov} equation.
-It can also be seen as a continuity euquation or a cosevation of probability
-equation. To this end define the probability current, $\p \in \R^n$ as:
-$$
-\phi_i =   F_i(x,t) \f(x,t) 
-+ 
-\sum_{j}  \di_{x_j} \left[ D_{ij}(x,t) \f(x,t)\right]
-$$
-then the Fokker-Planck equation, \cref{eq:FP_pde}, can be written as:
-$$
-\di_t \f= -\grad \cdot \phi
-$$
-which just says that the change in probability is the
-difference between the flow in and the flow out.
-
-Initial conditions for $\f$ are given by the distribution of $X_s$, if $s=0$
-then this is the initial distribution of $X$. Boundary conditions (BCs) depend
-on the domain of $X$ and what happens to $X$ once it hits its boundary. If the domain
-is all of $\R^n$ then we only insist that $\lim_{|x| \ra \infty} \f = 0$. If the
-domain has boundaries, then there are two common scenarios which we will also
-encounter in the sequel:
-\begin{enumerate}  
-  \item absorbing BCs
-  \item reflecting BCs
-\end{enumerate}
-At an absorbing boundary, the particle $X$ is removed and $\f=0$ there.
-In this situation, we will not have conservation of probability and the integral
-of $\f$ over $X$'s domain will monotonically decrease.
-
-At a reflecting boundary, the particle $X$ bounces back into its domain and we
-will have that the probability current $\p\cdot n = 0$ where $n$ is the outward
-normal at the reflecting boundary.
-
-We now consider $\f$ as functions of $y,s$, holding $x,t$ fixed, then
-\begin{equation}
--\di_s \f= \sum_i  \left[F_i(y,s) \cdot \di_{y_i}\f(|y,s) \right] 
-+ 
-\sum_{i,j}   \left[ D_{ij}(x,t) \di^2_{y_i y_j}\f(|y,s) \right]
-\label{eq:FP_backward_pde}
-\end{equation} 
-This is the \emph{Backward Kolmogorov} equation for $\f$. Note the minus sign
-in front $\di_s \f$. A mnemonic for the signs of the Backward vs. Forward
-equation is that as $t$ increases $\f$ diffuses and so $\di_t$ and $\di^2_{x}$
-have the same sign, but as $s$ increases $\f$ anti-diffuses and they have
-opposite signs.
-
-The differential operator on the right-hand side of
-\cref{eq:FP_backward_pde} occurs often in the study of SDEs and has its own
-name.
-\begin{defn}[Generator of an SDE] the Generator $A$ of $X$ is defined by:
-$$
-A[\psi(x)] = \lim_{t \searrow 0^+} \frac{\Exp[\psi(X_t)]  - \psi(x)}{t} ;  \quad X_0 =
-x \in \R^n
-$$
-\end{defn} 
-\begin{lemma} For an Ito SDE as in \cref{eq:generic_Ito_SDE_Rn} 
-$$
-A[\psi(x)] = \sum_i F_i(x,t) \cdot \di_{x_i} \psi + \sum_{i,j} D_{ij}(x,t)
-\di^2_{x_i x_j}\psi $$
-\end{lemma}
-
-\subsection{Stopping Times}
-\begin{defn}
-Let $\Fil(t)$ be some filtration, a random variable $\t$ is called a
-\emph{stopping time} if 
-$$
-\{\o : \t(\o) \leq t\} \in \Fil(t) \, \forall t  
-$$
-\end{defn}
-The colloquial way of describing stopping times is at any time we know
-whether $\t$ has occurred or not. For a counterexample, the time that a Wiener
-Process achieves its maximum over some interval is not a stopping time, since at any
-given time, we do not know if the maximum has occurred or not. The most common
-example of a stopping time is the first hitting-time, which is the first time
-$X_t$ leaves a set. 
-\begin{thm}[First-Hitting time] Let $E \subset \R^n, E \neq
-\phi$ be open or closed
-
-then $$\t := \inf\{ t \geq 0 | X_t \in E\}$$ is a stopping time.
-\end{thm}
-
-The reason stopping times are very useful is that all the facts so far quoted
-for Ito calculus using integrals $\int_0^T \intd{W}$ remain true if $T$ is
-replaced by a stopping time $\t$.
-
-Also it allows us to link SDE's and PDEs using the generator, $A$, of the
-diffusion: 
-\begin{thm}[Dynkin's formula] Given $\t$ a stopping time, $\Exp[\t] < \infty$.
-Then:
-$$
-\Exp[u(X_\t, \t)] =
-u(x, 0) + \Exp\left[\int_0^\t \di_t u + A[u] \intd{s}
-\right]
- $$
-\end{thm}
-This provides a link between PDEs and stochastic processes and allows us to go
-back and forth in that we can find probabilistic results by
-solving a PDE or we can approximate a PDE by simulating a stochastic process
-and averaging.
-
-\begin{ex}[\cite{Evansb} pg 99 - Expected hitting time to a boundary]
-\label{ex:mean_hitting_time}
- Let $\O
-\subset \R^n$ be a bounded open set with smooth boundary $\di \O$ then it is a
-basic fact from PDEs theory that
-\begin{equation}
-\begin{cases}
--\frac{1}{2} \grad^2 u = 1  & \text{over } \O
-\\
-u =  0 &\text{on } \di \O
-\end{cases}
-\end{equation}
-has a unique $C^\infty(\O)$ solution.
-
-Let $X = W_t + x$ for any $x \in \O$ and define
- $$\t_x := \text{first time } X \text{ hits } \di \O$$
-then the generator of $X$ is $A[\psi] = -\grad^2(\psi)/2$ and we will
-have:
-
-\begin{align*}
-\Exp[u(X_\t)] - \Exp[ u(x_0)] =& \Exp \left[ -\frac{1}{2}\int_0^\t \grad^2
-u\right]
-\\
-=& \Exp \left[ - \int_0^\t 1
-\right]
-\\
-=-&\Exp [\t]
-\end{align*}
-Finally, invoke $u$'s BCs, $u|_{\di E} = 0$, and $X$'s ICs, $X_0=0 = x$ to
-conclude that $$ u(x) = \Exp [\t]$$
-The solution to the PDE evaluated at $x$ is the expected exit time from
-$E$ for an $X_t$  starting at $x$.
-\end{ex}
-% Similarly we can calculate more moments of the hitting time.
-% \begin{ex}[Mean squared hitting time to a boundary] Let $E
-% \label{ex:mean_squared_hitting_time}
-%  \subset \R$ be a bounded open set and let $u$ solve:
-%  \begin{equation}
-%  \begin{cases}
-%  (\a - \tfrac{x}{\tc}) \di_x u + \frac{\b^2}{2} \di_x^2 u = -\frac t 2  &
-%  \text{over } E
-%  \\
-%  u =  0 &\text{on } \di E
-%  \end{cases}
-%  \end{equation}
-% 
-% Then $- t / 2 = A[u]$ for $A$, the generator, corresponding to $dX = (\a - x /
-% \tc) dt + \b dW$.
-% Define $$\t_x := \text{first time } X \text{ leaves }  E$$
-% then
-%  \begin{align*}
-%  \Exp[u(X_\t)] - \Exp[ u(X_0)] =& 
-%  \Exp \left[  \int_0^\t   (\a - \tfrac{x}{\tc}) \di_x u + \frac{\b^2}{2} \di_x^2
-%  u \right]
-%  \\
-%  =& \Exp \left[ - \int_0^\t -\frac t 2 
-%  \right]
-%  \\
-%  =-&\Exp [\t^2]
-%  \end{align*}
-% And we can conclude that 
-% $$ u(x) = \Exp [\t^2]$$
-% The solution to the PDE evaluated at $x$ is the expected squared exit time from
-% $E$ for an $X_t$  starting at $x$.
-% \end{ex}
-
-\subsection{Feynman-Kac formula}
-The Feynman-Kac formula is a generalization of the above idea. We state it in
-1-D. 
-\begin{thm}[Feynman-Kac Formula] Consider the (linear) PDE, with terminal
-conditions:
-\begin{equation}
-\begin{cases}
-\di_t u(x,t) + f \di_x u + \frac {g^2}2 \di_x^2 u - V u + q 
-&= 0
-\\
-u(x,\tf) &= \psi(x)
-\end{cases}
-\label{eq:PDE_feynman_kac}
-\end{equation}
-The dynamics can also be written:
-$$
-\di_t u(x,t) + A[u] - V u + q = 0
-$$
-for the generator $A = f \cdot \di_x + g^2/2 \cdot \di_x^2$ corresponding to the
-SDE: $$
-dX_t = f(X,t) dt + g(X,t) dW
-$$
-Then for $t< \tf$
-\begin{align*}
-u(x,t) =& \Exp\bigg[
-\int_t^\tf q(x,s)\cdot \exp \left(-\int_t^s V(X_\t) \intd{\t} \right)\intd{s} \\
-&\quad + \, \psi(X_t) \cdot \exp\left(-\int_t^\tf V(X_\t,\t) \intd{\t} \right)  
-\quad \Big| \,X_t = x \bigg]
-\end{align*}
-\begin{proof}[Proof]
-This can be seen as a more elaborate extension of example
-\ref{ex:mean_hitting_time}. We set: 
-$$
-Y_r = u(x,t)+ 
-\int_t^r q(x,s)\cdot\exp \left(-\int_t^s V(X_\t) \intd{\t} \right)\intd{s} +
-u(X_r,r)\cdot\exp\left(-\int_t^r V(X_\t,\t) \intd{\t} \right)   
-$$
-then 
-\begin{align*}
-dY =& q(x,r)\cdot\exp \left(-\int_t^r V(X_\t) \intd{\t} \right)
-\\
-& - u(X_r,r) \cdot V(X_r, r)  \exp\left(-\int_t^r V(X_\t,\t)
-\\
-& + du \cdot \exp\left(-\int_t^r V(X_\t,\t) \intd{\t} \right) 
-\intd{\t}\right)
-\end{align*}
-Now we recall Ito's lemma for $du(X,t)$, and replace it in the last line:
-\begin{align*}
-dY_r =& q(x,r)\cdot\exp \left(-\int_t^r V(X_\t) \intd{\t} \right) \intd{r}
-\\
-& - u(X_r,r) \cdot V(X_r, r)  \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t}\right)s
-\\
-& + \Big( \di_t u + A[u] \Big) \cdot \exp\left(-\int_t^r V(X_\t,\t) \intd{\t}
-\right) \intd{r}
-\\
-& + \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-\end{align*}
-Now we can group the terms in front of the exponential term on the first
-three lines to form:
-\begin{align*}
-dY_r =& \Big(\di_t u + A[u] - u(X_r,r) V(X_r, r) + q(X_r,r) \Big) \cdot\exp
-\left(-\int_t^r V(X_\t) \intd{\t} \right)
-\intd{r}
-\\
-& + \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-\end{align*}
-Since, $u$ satisfies the PDE, \cref{eq:PDE_feynman_kac}, the term on the first
-line is zero so $dY$ is just
-$$
-dY_r = \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-$$
-integrating we have: 
-$$
-Y_\tf - Y_t = \int^\tf_t \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot
-\exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-$$
-Now we need conditions such that the expectation of the stochastic integral is
-zero. For example, it suffices that $V$ is bounded, and often in applications
-$V$ is a positive rate.
-Then, we   take conditional expectations, conditional on $X_t=x$:
-\begin{equation*}
+\abstract{We pose an optimal control problem to precisely specify the spike
+timing in a noisy leaky-integrate and fire model of a neuron. The optimal
+control problem is solved using Dynamic Programming if we have access to the
+voltage (closed-loop control) and using a Maximum Principle for the transition
+density if we only have access to the spike timings (open loop control)}
+
+\section{A Stochastic Optimal Control for Spike Timings Control}
+Assume your neuron can be represented by a leaky-integrate and fire model
+\begin{equation}
 \begin{gathered} 
-\Exp\Big[Y_\tf | X_t = x\Big] - \Exp\Big[Y_t | X_t = x\Big]  
-= 
-\\
-\Exp\Bigg[\int^\tf_t \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot
-\exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W} \,\mid\, X_t = x\Bigg] 
-\\ = 0
-\end{gathered}
-\end{equation*}
-Unravelling the definition of $Y_t$, we get
-\begin{align*}
-\Exp\bigg[
-\int_t^\tf q(x,s)\cdot \exp \left(-\int_t^s V(X_\t) \intd{\t} \right)\intd{s} 
- \quad + &
-\\
- \, \psi(X_t) \cdot \exp\left(-\int_t^\tf V(X_\t,\t) \intd{\t} \right)  
-\, \Big| \,X_t = x \bigg] &=\, u(x,t)  
-\end{align*}
-
-\end{proof}
-A more detailed proof can be found in \cite{Oksendal2007} Thm. 8.2.1
-\end{thm}
-
-We will not use the Feynman-Kac formula in what follows, but
-we have included it here, because there are some very interesting developments in stochastic optimal
-control, which rely on solving a many-dimensional PDE using stochastic
-simulation, the link being made via the Feynman-Kac formula - see
-\cite{Kappen2005},\cite{Palmer2011}.
-
-% \subsection{Numerical Simulation of SDEs}
-
-\section{Deterministic Optimal Control in Finite Dimension}
-Optimal Control theory has three main components - a state, $x$, a control, $u$
-and an objective $J$ which is a functional of $x,u$ and which we try to either
-minimize or maximize. Here, we will be minimizing.
-
-We introduce the theory for a finite dimensional state, $x(t) \in \R^{n_x}$, but
-we will not address state constraints. Our control will also be
-finite-dimensional $u(t) \in \R^{n_u}$ and it will be allowed to take values in
-some closed domain, $\Udomain$, of allowed controls, e.g.\ $\Udomain = \{ u_t
-\in \R^{n_u} \, \st |u| \leq \Umax \}$ for a control constrained to a sphere.
-The objective $J$ will be given by a time-integral-plus-terminal cost:
-\begin{equation}
-J[u] = \int_0^\tf L(x_s, u_s) \intd{s} + M(x_\tf)
-\label{eq:generic_objective_function_deterministic}
-\end{equation}
-Where $\tf$ could be either a variable or fixed.
-
-The dynamics of $x$ are given by some controlled ordinary differential
-equation (ODE) given some initial conditions:
-\begin{equation}
-\dot{x} = f(x,t); \quad x(0) = x_0
-\label{eq:generic_dynamics_deterministic}
-\end{equation}
-
-Then we seek the optimal control $u^*$, such that:
-\begin{equation}
-u^* = \argmin_{u \in \Udomain} J[u]  
-\label{eq:generic_objective_argmin}
-\end{equation}
-
-Note that $x$ may also have final conditions, which we will deal with later,
-and/or path constraints. If there are path-constraints, we talk about a
-state-constrained problem. However, we will not face path constraints in the
-sequel and so will not say more about how to handle them.
-
-There are essentially three methods for solving equation
-\cref{eq:generic_objective_function_deterministic}. Two, Pontryagin's Minimum
-Principle and Dynamic Programing, are analytic in nature providing a set of
-equations that characterize the minimum and then solving these equations,
-numerically if necessary. The third, the direct method, simply discretizes the
-problem and solves the resulting nonlinear programing (NLP) problem via standard
-NLP techniques. We will describe both Pontryagin's Minimum Principle and Dynamic
-Programing below and indeed we will use both later on.
-
-\subsection{Pontryagin's Maximum Principle}
-Pontryagin's Principle is derived from on a variational argument, similar to the
-Euler-Lagrange equations from the Calculus of Variations, and it characterizes
-the optimal trajectory/control pair. It can be thought of as a generalization of
-the zero-tangent rule (Fermat's Rule) for finding optima in single-variable
-calculus and like both the Euler-Lagrange equations and Fermat's Rule it
-provides necessary, but not sufficient conditions for a minimum.
-
-Pontryagin's Principle also holds a similarity to Hamiltonian Mechanics, which
-should not be surprising, since Hamiltonian Mechanics is founded on
-the minimization of action.
-
-There are several versions of Pontryagin's Principle, depending on whether $x$ has
-terminal conditions or not, whether the final time, $\tf$ is specified or a
-variable in itself and most importantly whether the state has constraints or
-not. We will state the versions that are relevant for our needs, namely, $x$ has
-no constraints and $\tf$ is specified.
-
-We first introduce the Pontryagin Hamiltonian, $\H$, also called the control
-theory Hamiltonian and an adjoint state, $p \in \R^{n_x}$:
-\begin{defn} [$\H$] The control theory Hamiltonian is the function:
-$$
-\H(x,p, u) := f(x,u) \cdot p + L(x,u) \quad (x,p \in R^{n_x}, u \in \Udomain)
-$$ 
-\end{defn}
-
-\begin{thm}[Pontryagin's Fixed-Time, Free-End-Point] Assume $u^*(t)$ is
-optimal for
-\cref{eq:generic_objective_function_deterministic,eq:generic_dynamics_deterministic}
-and that $x^*$ is the corresponding trajectory.
-
-Then there exists a function $p^*:[0,\tf]:\ra \R^{n_x}$ st.
-\begin{equation}
-\begin{cases}
-\dot{x}^*(t) &=  \grad_p \H(x^*,p^*, u^*) = f(x^*, u^*)
-\\
-x(0) &= x_0
-\end{cases}
-\label{eq:pontryagin_state_ode}
-\end{equation}
-\begin{equation}
-\begin{cases}
-\dot{p}^*(t) &= -\grad_x \H(x^*,p^*, u^*) = -\grad_xL - p \grad_x f
-\\
-p^*(\tf) &= \grad_x M(x)
-\end{cases}
-\label{eq:pontryagin_adjoint_ode} 
-\end{equation}
-and
-\begin{equation}
-\H(x^*(t), p^*(t), u^*(t)) = \min_{u(t) \in \Udomain(t)}  \H(x^*(t), p^*(t),
-u(t))
-\label{eq:pontryaginH_optimality_condition} 
-\end{equation}
-\end{thm}
-We will not give a proof of this theorem (see A.2 in \cite{Evansb}),
-but we do provide a heuristic derivation in
-\cref{sec:Pontryagin_heuristic_derivation}, which is very helpful in deriving
-the necessary conditions for PDE dynamics and it explains the otherwise
-mysterious origins of $\H$ and $p$.
-
-At each time, $t$, we obtain the optimal control $u(t)$, by solving
-\cref{eq:pontryaginH_optimality_condition}. This can be easy, hard or useless.
-In the easy case, the minimum is analytically obvious, e.g.\ if the Hamiltonian
-is quadratic in $u$. If it is linear in $u$ then the minimum will occur on the
-boundaries of the control and we will talk about a bang-bang control.
-Conversely, solving the minimum of $\H$ with respect to $u$ may be impossible,
-since $x,p$ are unknown. In this case we need to have an iterative scheme which
-alternately solves
-\cref{eq:pontryaginH_optimality_condition} 
-and then \cref{eq:pontryagin_state_ode,eq:pontryagin_adjoint_ode}. Basic
-numerical techniques using this idea are discussed in the appendix in
-\cite{Kirk2004}. However, if we are going to be solving the problem numerically,
-then the direct method might be superior, see \cite{Ross2005} for a discussion.
-
-Finally, it is possible that \cref{eq:pontryaginH_optimality_condition} does not
-depend on $u$ explicitly (it will still depend  on $u$ implicitly through $x,p$)
-and so we cannot find $u$ at all by minimizing $\H$ wrt.\ $u$. This happens, for
-example, if the cost, $L$, does not depend on $u$, the dynamics, $f$, are linear
-in $u$ and $p=0$ over an interval of positive length. Such an optimal control
-problem is called singular and it is discussed briefly in \cite{Kirk2004}, and
-the theory can still be useful in finding a solution, but many complications can
-and do arise. We face a singular control problem in our own work later, however
-instead of tackling it head on, we merely regularize the problem - thus we will
-not say much more on Singular Control Problems.
-
-The version of Pontryagin's Principle for Fixed End Point, meaning $x$ has some
-terminal conditions, is very similar to the one with free end-point. Suppose 
-\begin{equation}
-\dot{x} = f(x,t); 
-\quad
-\begin{cases}
-x(0) = x_0
-\\
-x(\tf) = x_f
-\end{cases}
-\label{eq:generic_dynamics_deterministic_fixed_end_point}
-\end{equation}
-
-\begin{thm}[Pontryagin's Fixed-Time, Fixed-End-Point] Assume $u^*(t)$ is
-optimal for
-\cref{eq:generic_objective_function_deterministic,eq:generic_dynamics_deterministic_fixed_end_point}
-and that $x^*$ is the corresponding trajectory.
-
-Then there exists a function $p^*:[0,\tf]:\ra \R^{n_x}$ st.
-\begin{equation}
-\begin{cases}
-\dot{x}^*(t) &=  \grad_p \H(x^*,p^*, u^*) = f(x^*, u^*)
-\\
-x(0) &= x_0
-\\
-x(\tf) &= x_f
-\end{cases}
-\label{eq:pontryagin_state_ode}
-\end{equation}
-\begin{equation}
-\dot{p}^*(t) &= -\grad_x \H(x^*,p^*, u^*) = -\grad_xL - p \grad_x f
-\label{eq:pontryagin_adjoint_ode} 
-\end{equation}
-and
-\begin{equation}
-\H(x^*(t), p^*(t), u^*(t)) = \min_{u(t) \in \Udomain(t)}  \H(x^*(t), p^*(t),
-u(t))
-\label{eq:pontryaginH_optimality_condition} 
-\end{equation}
-\end{thm}
-I.e. $p$ loses its final conditions because $x$ has them. This relates to the
-variational nature of Pontryagin's Principle and the idea that $p$ represents
-the sensitivity of the objective wrt. small changes in $x$. Thus, if $x$ is
-fixed, there are no variations and $p$ can be anything.
-% To illustrate Pontryagin's Principle we will solve the moon
-% lander problem (Ex. 4.4.4 \cite{Evansb}), with the twist of holding the final
-% time fixed. 
-% \begin{ex}
-% 
-% \end{ex}
-
-\subsection{Dynamic Programing - the Hamilton-Jacobi-Bellman equation} 
-Dynamic Programing uses backwards recursion to tabulate the optimal control
-starting from the terminal time. The basic object in dynamic programing is the value function, $v$. In order to
-introduce it, we first extend our definition for the objective, $J$, to include later
-times, with different initial conditions.
-
-The running cost-to-go corresponding to
-\cref{eq:generic_objective_function_deterministic} is:
-\begin{equation}
-J[u; x, t] = \int_t^\tf L(x_s, u_s) \intd{s} + M(x_\tf), \quad x_t = x
-\label{eq:generic_cost_to_go_deterministic} 
-\end{equation}
-So that $J[u; x_0, 0] = J[u]$.
-
-The dynamic programming value function, $v$ is then defined by:
-$$
-v(x,t) = \inf_{u \in \Udomain} J[u; x,t] 
-$$
-it is also called the optimal cost-to-go.
-
-We immediately note the terminal conditions on $v$:
-\begin{equation}
-v(x,\tf ) = M(x)
-\end{equation}
-
-The Dynamic Programing approach is then to solve for $v$ backwards and in the
-process infer the control $u$.
-\begin{thm}[Hamilton-Jacobi-Bellman Equation] Assume that $v$ is $C^{1,1}$ in
-$x,t$. Then $v$ solves:
-\begin{equation}
-\begin{cases}
--\di_t v(x,t) &= \max_{u(t) \in \Udomain(t)} \left\{ f(x,t) \cdot \di_x v +
-L(x,u) \right\}
-\\
-v(x,\tf) &= M(x)
-\end{cases}
-\end{equation}
-\end{thm}
-We will not derive the HJB equation here, but we will derive a similar version
-later for the stochastic case, when the dynamics are given by SDEs.
-
-% \subsection{Model Predictive Control}
-% Solving the HJB equations and obtaining a feedback control is usually
-% impossible. Thus a practical alternative is to solve a simplified problem on a
-% limited horizon and use the simplified solution now and then after stepping
-% forward in time and observing the system in a new position, to solve a revised
-% limited horizon problem and use the new solution.
-% Let us make this more precise.
-
-\section{Stochastic Optimal Control}
-Here we extend the Dynamic Programing theory to the case where the dynamics are
-governed by an SDE:
-\begin{equation}
-dX = F(X,t, u) dt + G(X,t) dW
-\end{equation} 
-i.e.\ we assume control affects only the drift term. This 
-assumption is not necessary, but it corresponds to our problems.
-
-For stochastic dynamics one can only optimize probabilistically and so
-$J$ becomes:
-\begin{equation}
-J[u] = \Exp \left[ \int_0^\tf L(x_s, u_s) \intd{s} + M(X_\tf)
-\right] 
-\label{eq:generic_objective_function_stochastic}
-\end{equation}
-while the running cost-to-go is:
-\begin{equation}
-J[u; x, t] = \Exp \left[ \int_t^\tf L(X_s, u_s) \intd{s} + M(X_\tf) \right], 
-\quad X_t = x
-\label{eq:generic_cost_to_go_stochastic}
-\end{equation}
-the value function is:
-\begin{equation}
-v(x,t) := \inf_{u \in \Udomain} J[u; x,t]
-\end{equation}
-
-Finally, we again have that $v$ satisfies a certain PDE:
-\begin{thm}[Stochastic HJB] 
-\label{thm:stochastic_hjb}
-\begin{equation}
-\begin{cases}
--\di_t v(x,t) &=  \max_{u(t) \in \Udomain(t)} \big\{ L(x,u)  +
-F(x,t,u) \cdot \di_x v
-\big\} + \frac{G^2(x,t)}{2} \di_x^2v
-\\
-v(x,\tf) &= M(x)
-\end{cases}
-\end{equation}
-\begin{proof}[Heuristic Derivation (from \cite{Evansb})] Suppose we are at time
-$t$ and $X_t = x$.
-
-Take a time increment $[t, t+h]$ and assume that during that time we apply a
-control $u$ and subsequently we apply the the optimal control $u^*$. The
-running-cost will then break down as: 
-$$
-J[u; x,t] = \Exp \left[ \int_t^{t+h} L(X_s, u_s) \intd{s}  + v(X_{t+h},
-t+h) \right] $$
-
-Since $v(x,t) = \inf J(u; x,t)$, we must have that:
-$$
-v(x,t) \leq  \Exp \left[ \int_t^{t+h} L(X_s, u_s) \intd{s}  + v(X_{t+h},
-t+h) \right] $$
-or, rearranging, that
-\begin{align*}
-0 \leq^&  \Exp \left[ \int_t^{t+h} L(X_s, u_s) \intd{s} \right]  
-+ \underbrace{\Exp \left[ v(X_{t+h}, t+h) - v(x,t) \right]}_{\Exp\left[ dv
-\right]}
-\end{align*}
-Now, $\Exp\left[ dv \right]$ can be expressed using Dynkin's Formula, as:
-$$
-\Exp[dv] = \int_t^{t+h} \di_t v +  F\cdot \di_x v + \frac{G^2 }{2}\cdot \di_x^2
-v \intd{s} $$
-Plugging that back, we get
-$$
-0 \leq \int_t^{t+h} L(X,u) +  \di_tv +  F\cdot \di_x v + \frac{G^2 }{2}\cdot
-\di_x^2 v \intd{s} $$
-Taking $h \ra 0$ we get 
-$$
-0 \leq  L(x,u) +  \di_tv(x,t) +  F\cdot \di_x v(x,t) + \frac{G^2 }{2}\cdot
-\di_x^2 v(x,t) $$
-and we conjecture that for the actual optimal control, the inequality becomes an
-equality:
-\begin{equation}
-\begin{cases}
-0 &=   L(x,u^*)+ \di_t v(x,t) +
- F(x,t,u^*)\cdot \di_x v(x,t) + \frac{G^2(x,t)}{2}\cdot \di_x^2 v(x,t)
-\\
-u^*(t) &= \argmax_{u \in \Udomain(t)}  
-\big\{L(x,u) + F(x,t,u)\cdot \di_x v \big\}
-\end{cases}
-\end{equation}
-\end{proof}
-\end{thm}
-% A rigorous proof can be found in \cite{Krylov2008}, sec???ch???
-
-\section{Deterministic Optimal Control in Infinite Dimensions}
-Now we discuss Optimal Control when the dynamics are deterministic but of
-infinite-dimension, i.e. the state evolution is a PDE instead of an ODE.  The
-reason we discuss this, is that optimization of ordinary SDEs in an open-loop
-control is actually a deterministic optimal control problem in infinite
-dimensions.
-
-Why?
-
-Since the objective is probabilistic and the probability density evolution is
-given by the (deterministic) Fokker-Planck PDE, we need to optimize a
-deterministic system whose dynamics are given by the Fokker-Planck equation.
-
-There is no all-encompassing generalization of Pontryagin's Principle to
-infinite dimensions, but a Maximum Principle can be derived for certain systems.
-To illustrate the idea we will discuss a simple case as in \cite{Palmer2011}.
-Our controlled SDE will be given by: 
-$$ dX_t = F(X,t, u) dt + G(X,t) dW, $$
-The corresponding probability density is given by: $$ \di_t \f=
--\sum_i \di_{x_i} \left[ F_i(x,t,u) \f(x,t) \right] + \sum_{i,j}  \di^2_{x_i
-x_j} \left[ D_{ij}(x,t) \f(x,t) \right] $$ which we will write as $$ \dot{\f} =
-\L_{u} [\f] $$ where $\L_{u}$ is the differential operator corresponding to the
-RHS of the Fokker-Planck equation parametrized by the control $u$. For now we
-will assume that there are no BCs, and the domain of $X$ is all of $\R^n$.
-
-A natural inner-product, $ \langle.,. \rangle$ on functions in $L^2[\R^n]$ is
-given by: $$
- \langle\psi,\phi  \rangle = \int_{\R^n} \psi(x) \phi(x) \intd{x}
-$$
-Our objective will be to minimize:
-\begin{equation}
-J[u] = \int_0^\tf  \langle L(\cdot,s,u), \f_u(s) \rangle \intd{s} +  
-\langle M(\cdot),\f(\tf) \rangle
-\label{eq:generic_objective_function_probabilistic}  
-\end{equation}.
-
-\Cref{eq:generic_objective_function_probabilistic} is a very natural cost
-function in the stochastic dynamics case, because it can also be written as: 
-$$
-J[u] = \Exp \left[ \int_0^\tf  L(X_s, s,u) \intd{s} + M(X_\tf) \right] 
-$$ which
-is just \cref{eq:generic_objective_function_stochastic}. We use
-\cref{eq:generic_objective_function_stochastic} if we have some feedback and
-want to use dynamic programming. We use
-\cref{eq:generic_objective_function_probabilistic} if we have no observations
-and need to run an open-loop control.
-
-Since, \cref{eq:generic_objective_function_probabilistic} is the objective
-function for deterministic dynamics, we can use Pontryagin's Principle adapted
-to the case of infinite dimensional dynamics (PDEs).
-
-Introduce an adjoint co-state, $\l \in \R^n$ and let the Hamiltonian, $\H$ be:
-$$
-\H(\f, \l, u)= \langle \phi, L + \Lstar[\l] \rangle$$
-where $\Lstar$ is the adjoint operator to $\L$ defined via:
-$$
-\langle \L [\phi], \l \rangle =\langle \phi, \Lstar[\l] \rangle
-$$
-
-Let $\l$ evolve (backwards) according to:
-\begin{equation*}
-\begin{cases}
--\di_t \l =& \Lstar[\l] + L
-\\
-\l(\tf) =& M(x)  
-\end{cases}
-\end{equation*}
-then the optimal control, $u$, is given via
-$$
-u = \argmin_u {\H(\f, \l, u)}
-$$
-
-An important fact, useful for computations, is that the total derivative of the
-objective with respect to the control is given by the partial derivative of the
-Hamiltonian with respect to the control:
-\begin{equation}
-\grad_u \big[J[u] \big] = \di_u \H(\f, \l, u)
-\label{eq:objective_gradient_wrt_control}
-\end{equation}
-
-So for a given $u$, we can compute $\f,\l$, calculate the objective gradient
-using \cref{eq:objective_gradient_wrt_control} and then iterate in the direction
-of descent (if we are minimizing).
-
-All that remains is to compute $\Lstar$. It turns out that if we do not have any
-BCs, then
-$$
-\Lstar[\l] = \sum_i F_i(x,t) \cdot \di_{x_i} \l + \sum_{i,j} D_{ij}(x,t)
-\di^2_{x_i x_j}\l 
-$$ 
-which is just the generator of $X$!
-
-We will see later, that things become more complicated when we need to
-consider a PDE's BCs.
-
-\section{Closed-loop vs.\ Open-Loop control}
-At this point we stop to briefly discuss the difference between
-open-loop control and closed-loop control.
-
-By closed-loop control, we mean that at any time we have observations of the
-system state, $x$ and its value can influence the control. In open loop control,
-we decide on the value of $u(t)$ a priori and apply it independently of the
-value of $x(t)$, based only on the knowledge of the system dynamics and the
-initial conditions.
-
-If the system dynamics are deterministic there is no added value in closed-loop
-control, since the state evolution is perfectly predictable and one can infer
-where $x$ will be based only on the dynamics and the applied control. For
-stochastic dynamics this is no longer true and there is something to be gained
-by observing the state, since different controls can be applied depending on the
-different realizations of the stochastic process driving the dynamics.
-Naturally, if one has observations of the system, one would prefer to use this
-and apply a closed-loop control. Thus, an open-loop control would be used
-only if one has no reliable knowledge of the system state.
-
-\section{A Stochastic Optimal Control Problem from Mathematical Neuroscience}
-Neurons relay information by means of voltage spikes - sudden sharp increases in
-voltage. Although many details remain unclear, the information content is
-thought to be contained in the length of the time-interval between these spikes.
-In the simplest case, this can be thought of as a rate  -the average number of
-spikes per time interval, but more complicated coding schemes are hypothesized to
-exist.
-
-Many experiments allow for manipulating an individual neural cell. A natural
-goal then is to make a cell produce a given spike train. This may arise, for
-example, in brain-machine interfaces or in artificial prosthetics.
-
-We now consider this problem in the framework of Stochastic Optimal Control and
-give both a feedback solution, when the cell voltage is explicitly observable as
-well as an open-loop solution when only the occurrence of a spike is
-robustly observable.
-
-\subsection{Problem Formulation}
-The most basic representative model for a neuron is the noisy
-leaky-integrate-and-fire model:
-\begin{equation}
-\begin{gathered}
-dX_s = (\a(t) - \frac{X_s}{\tc} ) \intd{s} + \b \intd{W_s},
+dX_t = \left(\a(t) - \frac{X_t}{\tc} \right) \intd{t} + \b \intd{W_t},
 \\
@@
 \end{equation}
-In this $X$ represents the membrane electric potential which tries to revert to
-$0$ with a time-const of $\tc$, $dW$ is a Brownian motion increment with
-intensity $\b$ and $\a(t)$ is our external control. Having last spiked at
-time $0$, the potential hits $\xth$ at time $\ts$, we record a spike
-of duration $\ts$, the potential resets to $0$ and we start all over again.
-
-Suppose we want to ensure that spikes occur exactly at the predetermined
-times $\tns$. Then given the renewal nature of our spiking model, we can work
-sequentially, focusing only on the first of these intervals, of duration $t_1 =
-\T$, and then ask: how do we choose $\a(t)$ st. $\ts = \T$. There are several
-ways to specify what we mean by 'make $\ts = \T$', all of which result in optimal
-control problems. A natural one is a least squares solution:
+We look for the control, $\a(t)$, which will minimize the squared distance
+between the realized spike time, $\ts$ and a desired, pre-determined time $\T$.
 \begin{equation}
@@
 \end{equation}
-
-It is also natural to assume that our control has certain limitations. The
-simplest and most common are simple bounded constraints:
-\begin{equation}
-\a(t) \in [\amin, \amax]
-\label{eq:bound_constraints_alpha}
-\end{equation}
-
-% \subsection{Optimal Control Formulation}
-In addition to \cref{eq:OC_LS_variance}, our objective will also contain a
-running cost based on the control. This regularizes the problem eliminating the
-subtleties of singular-control situations and it might be argued appropritate
-to avoid excessive control.
-
-So we will seek an optimal control, $\a$ that solves:
-\begin{equation}
-\a(t) = \argmin_{\a(t)} \Big\{ \,
-\Exp\left[
-\e \int_0^\ts  \a^2 \intd{s}
-+  
-(\ts - \T \big)^2 \right]  \, \Big\}
-\label{eq:OC_LS_variance_energy}   
-\end{equation}
-where $\e$ measure how much weight we put on minimizing the energy expenditure
-and if $\e \rightarrow 0$, then we do not care at all.
-
-Experimentally it is often difficult to observe precisely the exact voltage
-trajectory, but the occurrence of a spike is much more easily observable. As
-such a closed-loop approach is not implementable since the state is unknown. In
-the sequel we will perform three calculations. First we will derive the HJB
-equation for a closed-loop control if the system was observable, and use this as
-a benchmark for the best possible performance of an open-loop control. Then we
-will calculate the most simple-minded open-loop control, which is just the
-deterministic control of the dynamics when assuming there is no noise, $\b=0$.
-THen we will calculate the optimal open-loop control by optimizing the
-transition density. And finally we will compare the performance of all three
-controls.
+where expectation is taken with respect to the trajectories of $X_t$.
+
+We consider two scenarios, one, {\sl closed loop control}, in which the
+value of $X_t$ at $t$ is known and can be used in determining the control and
+,two, {\sl open loop control} in which only the times of spikes is known. In
+closed-loop scenario, we might better write the control as
+$$\a = \a(x,t)$$ to indicate its dependence on $X_t$. In the open-loop scenario,
+we would instead write $$\a = \a(t)$$ to indicate that we will always do the
+same thing regardless of the (unknown) behaviour of $X_t$.
+
+For the closed-loop scenario we will use {\sl Dynamic Programing},
+\cite{Fleming1975}, while for the open-loop scenario we will use a form
+of the {\sl Maximum Principle} for transition densities, \cite{Ahmed1981}.
 
@@
 +
-(\t - (\T-t) \big)^2 |x,t
+(\ts - \T) \big)^2 |x,t
 \right] \, \Big\}
@@
 \end{equation}
-i.e.\ if time $t$ has elapsed without a spike, we now want to minimize
-the difference between $\t$ and ($T-t$), given the current state $x$. For now we
-assume that we know $x$, later we will treat the problem when that is not so.
-
-We are now in a position to set up the HJB equation for this problem. First
-introduce the value function, $v(x,t)$ as:
-\begin{multline}
-v(x,t) = \argmin_{\a(s \geq t)} \{J[\a, x,t] \} =
-\\ 
+
+Setup the Hamilton-Jacobi-Bellman (HJB) equation
+\cite{Fleming1975} for this problem by first introducing the value function, $v(x,t)$ as: 
+\begin{align}
+v(x,t) =& \argmin_{\a(s \geq t)} \{J[\a, x,t] \}
+\notag 
+\\ 
+=&
 \argmin_{\a(s \geq t)} 
@@
 \label{eq:OC_LS_mean_value}   
-\end{multline}
+\end{align}
 
@@
 \end{equation}
-
-We have been canny enough to set up a quadratic optimization problem in the
-control, which means that the optimal control is given analytically via:
-\begin{equation}
-\a(x,t) = \frac{-\di_x v}{2\e} 
+and the optimal control is given by:
+\begin{align}
+\a(x,t)  =&\argmin_{\a \in [\amin, \amax]}\big\{
+\e \a^2 + (\a-\tfrac{x}{\tc}) \di_x v \big\}
+\notag 
+\\
+=& 
+\min \left(\amax, \max\left(\amin, -\frac{\di_x v}{2\e}\right)
 \label{eq:OC_LS_variance_energy_quadratic_control}
-\end{equation}
-However we need to recall the bounds on the control, $\amin < \a < \amax$, and
-thus
-\begin{equation}
-\astar(x,t) = \min \left(\amax, \max\left(\amin, -\frac{\di_x v}{2\e}\right)
-\right)
-\label{eq:OC_LS_variance_energy_quadratic_control_bounds}
-\end{equation}
+\end{align}
 With this the HJB becomes
@@
 \end{equation}
-
-In the sequel we will drop the superscript star on the optimal control and
-just write $\a(t)$ for the solution of
-\cref{eq:OC_LS_variance_energy_quadratic_control_bounds}. 
-% Thus the control can be solved out of the HJB equation, which now becomes:
-% \begin{equation}
-% \di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + \tfrac{(\di_x v)^2}{\e}
-% -\tfrac{x}{\tc}\di_x v= 0
-% \label{eq:OC_LS_mean_HJB}
-% \end{equation}
-
-Now, we need to consider boundary conditions (BCs). If $X_t =
-\xth$ then we have a spike now and we know that $\ts = t$. Thus:
-$$
-v(\xth,t) = (t-\T)^2
-$$
-Now for $x \ll .0$, we can assume that $v$ doesn't really depend on $x$ so say 
-$$
-\di_x v(\xmin, t) = .0
-$$
-This is legitimate if we impose artificial reflecting boundaries for the
-diffusion at $\xmin$. So we need to determine for which $\xmin$ is a reflecting
-boundary consistent with the dynamics. For example, we can take $\xmin$ to be
-two standard deviations of the stationary distribution of the non-controlled, $\a=0$, \cref{eq:X_evolution_uo}.
-% Ooops,
-% we need $\di_x v \neq 0$ and here we are setting $\di_x v = 0$ in one of the BCs\ldots oh well.
-
-Finally we come to the Terminal Conditions (TCs) for $v$. The idea is simple: if
-we reach $\T$ without having spiked we will just slam on the accelerator: i.e.\
-$$t>T \implies \a(t) = \amax$$ Thus: $$ v(x,T) = \Exp[\t^2|x, \amax] $$.
-
-Thus we restate the HJB equation in its fully specified form:
+\Cref{eq:OC_LS_mean_HJB_bounded_control,eq:OC_LS_variance_energy_quadratic_control}
+need to be solved simultaneously.
+
+Applying the right BCs and TCs for $v$ results in the fully specified HJB
+equation
 \begin{equation}
@@
 
-\subsection{Calculating the TCs} 
-\label{sec:calculating_TCs}
-At this point we are in a position to solve the HJB \cref{eq:OC_LS_HJB_full}.
-But first we need a more concrete expression for the TCs. Let
-\begin{equation}
-\begin{array}{lcl}
-\Tone(x_0) &=& \Exp[\t|x_0, \amax] 
-\\
-\Ttwo(x_0) &=& \Exp[\t^2|x_0, \amax]  
-\end{array}
-\end{equation}
-where as before, $\t$ is the remaining time-to-spike, given $X_0 = x_0, \a(t) =
-\amax$. Naturally, $\Ti$ depend on $x_0$ and are affected by the value of the
-various parameters, $\a,\b,\t$.
-
-Actually, we only  need $\Ttwo$ for the HJB equation above, but we'll see that
-$\Tone$ will be needed as an auxiliary.
-
-Indeed, analytical expressions exist for $\Ti$, \cite{Inoue1995}. However in my
-experience, they are very finicky numerically, especially as they involve
-infinite alternating sums which can easily overflow in numeric calculations.
-Moreover, they are not particulary convenient if you need $\Ti (x_0)$ for a wide
-range of $x$'s as we do here. Instead we will turn to:
-
-\subsubsection{the Backward Kolmogorov Equation}
-Here we follow Jacobs, \cite{Jacobs}, in deriving and solving a differential
-equation for $\Ti(x_0)$.
-
-Basically we are calculating the mean-time-squared to exit an interval $[\xmin,
-\xth]$. Theoretically, $\xmin = -\infty$, but for our purposes we will set it to
-some finite value and impose reflecting boundaries there - this is a very good
-approximation for $\xmin \ll 0$, as long as $\a>0$, which is what we have.
-
-Let $B(y,t|x) = \Prob[X_0 = y| X_t = x]$ for a fixed $x$. Then $B$ solves the
-following (backward Kolmogorov) PDE:
-\begin{equation}
--\di_t B = (\a - \frac y\tc)\di_yB - \frac {\b^2} 2 \di_y^2 B
-\label{eq:backward_Xdensity}
-\end{equation}
-
-Let $\G(t,y)$ be the survival function for $T$ for $X_0 = y$. 
-$$\G(t,y) = \Prob[T>t | X_0 = y] = \Prob[X_{s\leq t} \in [\xmin,
-\xth] | X_0 = y]$$.
-
-Because the dynamics do not depend explicitly on time, we have:
-$$
-\G(t,y) = \int_\xmin^\xth B(y,t|x) \intd{x}
-$$
-and so $\G$, being a definite integral of $B$ satisfies the same PDE:
-\begin{equation}
--\di_t \G = (\a - \frac y\tc) \di_y\G - \frac {\b^2} 2 \di_y^2 \G
-\label{eq:backward_SDF}
-\end{equation}
-
-With the BCs, ICs:
-\begin{equation}
-\begin{cases}
-\G(t,\xth) = .0 & \quad (T = t \implies \Prob[T>t] = 0)
-\\
-\di_x \G(t,\xmin) = .0  & \quad (\textrm{due to reflecting boundary for X})
-\\
-\G(0, y) = 1. & \quad (T \textrm{ is definitely} > 0 \textrm{ for } x_0 \in
-(\xmin, \xth) )
-\end{cases}
-\end{equation}
-
-\subsubsection{A BVP for $\Ti(x_0)$}
-If we solve the PDE for $\G$ over $t \in [0,\infty)$, then we can calculate
-$\Ttwo(x_0)$ as:
-\begin{eqnarray}
-\Ttwo(x_0) &=& \int_0^\infty t^2 \cdot (-\di_t\G(t,x_0)) \intd{t}
-\\
-		   &=& \int_0^\infty 2t \cdot \G(t,x_0) \intd{t}
-\end{eqnarray}
-assuming it exists.
-
-Since $\Ttwo$ is an integral over time, we can reduce the PDE in
-\cref{eq:backward_SDF} for the distribution to a BVP for the moment(s) by
-integrating over time. Here is how this is done: multiply both sides  of
-\cref{eq:backward_SDF} by $2t$, and integrate with respect to time:
-\begin{equation}
-\int_0^\infty 2t (-\di_t \G) \intd{t}
-= 
-\int_0^\infty  2t \left[ (\a - \frac y\tc) \di_y\G - \frac {\b^2} 2
-\di_y^2 \G\right]
-\intd{t}
-\end{equation}
-The LHS turns out to be $2\Tone$, so 
-\begin{equation}
-2\Tone
-=(\a - \frac y\tc)   \di_y \Ttwo 
-- \frac {\b^2} 2
-\di_y^2 \Ttwo
-\label{eq:BVP_Ttwo}
-\end{equation}
-Very similarly we can derive an equation for $\Tone$, namely:
-\begin{equation}
-1
-=(\a - \frac y\tc)   \di_y \Tone 
-- \frac {\b^2} 2
-\di_y^2 \Tone
-\label{eq:BVP_Tone}
-\end{equation}
-In both cases, the BCs will be:
-\begin{equation}
-\begin{cases}
-\Ti(\xth) = .0 & \quad (\textrm{we are already spking})
-\\
-\di_x \Ti(\xmin) = .0  & \quad (\textrm{due to reflecting boundary for }X)
-\end{cases}
-\label{eq:BVP_Ti_BCs}
-\end{equation}
-
-\Cref{eq:BVP_Tone,eq:BVP_Ttwo} are seemingly second order ODEs, but are
-actually first order and can be written as:
-\begin{equation}
-f(y)
-= (\a - \frac y\tc) S(y) 
-- \frac {\b^2} 2 S'(y) ; \quad  S(\xmin) = .0
-\label{eq:BVP_T_generic}
-\end{equation}
-and we can recover $\Ti$ from $S$ via:
-$$\Ti(x_0) =   -\int_{x_0}^\xth S(y)\intd{y}$$.
-It is clear that thus defined, $\Ti$ will satisfy the BCs of
-\cref{eq:BVP_Ti_BCs}.
-
-% Solving \cref{eq:BVP_T_generic} for $S$ gives: \begin{eqnarray} S(y) &=& -
-% \exp\left(\smallint \tfrac UD \right) \cdot \int_\xmin^{y} \exp\left(-\int
-% \tfrac UD\right) \frac{f(\xi)}D \intd{\xi}
-% \\
-% &=& - \exp\left( \frac{\a y - \frac{y^2}{2\tc}}D \right) \cdot \int_\xmin^{y} 
-%  \exp\left(- \frac{\a \xi - \tfrac{\xi^2}{2\tc}}D \right) \frac{f(\xi)}D
-% \intd{\xi} \label{eq:BVP_T_generic_soln} \end{eqnarray}
-We can solve \cref{eq:BVP_T_generic} analytically, but the analytical solution
-turns out to be awkward for numerical calculations, so we
-are better off just calculating the numerical solutions to \cref{eq:BVP_Tone,eq:BVP_Ttwo} directly
-via an ODE solver, starting from $S(\xmin) = .0$ and integrating until $\xth$.
-
-With this we can compare $\Ttwo$ between simulations and the numerical
-solutions see \cref{fig:Ti_empirical_vs_analytic}.
-\begin{figure}[htp]
-\begin{center}
-\subfloat[]
-{
-\includegraphics[width=0.48\textwidth]
-{/home/alex/Workspaces/Latex/OptSpike/Figs/Moments_a=10_b=1_N=128.png}
-}
-\subfloat[ ]
-{
-\includegraphics[width=0.48\textwidth]
-{/home/alex/Workspaces/Latex/OptSpike/Figs/Moments_a=10_b=10_N=128.png}
-}
-\\
-\subfloat[ ]
-{
-\includegraphics[width=0.48\textwidth]
-{/home/alex/Workspaces/Latex/OptSpike/Figs/Moments_a=15_b=1_N=128.png}
-}
-\subfloat[ ]
-{
-\includegraphics[width=0.48\textwidth]
-{/home/alex/Workspaces/Latex/OptSpike/Figs/Moments_a=15_b=10_N=128.png}
-}
-\caption[]{Comparison between solutions of equations
-\cref{eq:BVP_Tone,eq:BVP_Ttwo} and empirical statistics from simulating the
-hitting times of \cref{eq:X_evolution_uo} for four different pairs of $\a,\b$.
-In the top panel we show the first moment of the remaining time-to-spike
-starting from $x$, $\Tone(x)$, and in the bottom panel, we show the second
-moment, $\Ttwo(x)$. We have used only $N=128$ samples to form the moment
-statistics.}
-\label{fig:Ti_empirical_vs_analytic}
-\end{center}
-\end{figure}
-
-\subsection{Solving the HJB - the numerical method for
-\cref{eq:OC_LS_mean_HJB_bounded_control}}
-We now have a PDE for $v$ and an algorithm for computing all the BCs and TCs of
-this PDE. It is time to discuss the numerical method for solving
-\cref{eq:OC_LS_HJB_full}.
-Since we are in 1-d and we have the simplest geometry, the simplest thing to do
-is try a centred Finite Difference scheme, upwind wherever necessary and hope
-for the best.
-
-Things will break down if $\e$ is very small in which case the quadratic term,
-$(\di_x v)^2 / \e$, will dominate and cause 'havoc' or if $\b$ is very small in
-which case the smoothing effect of the second derivative will be negligible and
-we will be in the zone of nonlinear conservation equations. Let us hope for the
-best now and assume these two problems do not arise, for the parameters we
-select.
-
-Recall that we are solving backwards in time. Then for variables at the next
-time-step (already computed) we will write $(\cdot)^+$; while variables at the
-previous time step (to-be computed), will be denoted as$(\cdot)^-$.
-
-Similarly given some node in space, we will write
-$(\cdot)_-,(\cdot)_0,(\cdot)_+$ as the left node, the node itself and the right
-node.
-
-Then our finite-difference scheme in time looks like:
-\begin{equation}
-\frac{v^- - v^+}{\Delta t} = \tfrac{\b^2}2  .5*(\di_x^2 v^- + \di_x^2 v^+) +
-(\a^+ - \tfrac{x}{\tc})  .5*(\di_x v^- + \di_x v^+) + \e (\a ^+)^2
-\end{equation}
-
-I.e.\ we have removed the non-linearity in $v$ by computing the control, $\a$ at
-the next, already computed, time-step and used central-differencing
-otherwise.
-
-For the spatial derivatives, we will not upwind for now and also use central
-differencing:
-$$\di_x v = \frac{v_+ - v_-}{2\Delta x}$$
-and as is standard:
-$$\di^2_x v = \frac{v_+ - 2 v_0 + v_-}{(\Delta x)^2}$$
-
-We will not discuss a detailed study of the convergence properties of the
-numerical method, but just demonstrate its performance on a representative
-parameter set in \cref{fig:HJB_attempt} we see that the method is stable for a
-particular parameter set.
-
-% \begin{figure}[ht]
-% \begin{center}
-%   \includegraphics[width=.7\textwidth]{Figs/KolmogorovBVPMoments_a=10_t=9_b=8.png}
-%   \caption[]{TCs for alpha = 1.0;  tauchar = .9; beta = .75;}
-%   \label{fig:TCs_extended_x}
-% \end{center}
-% \end{figure}
-
-Let's discuss the most salient features of \cref{fig:HJB_attempt}. At the end,
-when $t = \T$, the value function, $v$ is monotonely decreasing. This makes
-sense, becaue the lower $x$ is the more time on average will it take us to spike
-and the more delay will we have from our desired spike time. However, as we go
-back in time $v$ inverts, in the sense that for $t \ll \T$, we would like to
-be somewhat far from the boundary to avoid the risk of spiking early.
-
-Now focus on the controls, \cref{fig:HJB_attempt_control}: Naturally, at the
-end we have the control assume its maximum value, $\a(t>=\T) = \amax$. As we
-start to go back in time however, two things start to happen, the control $\a$
-decreases for $x \ll \xth$, and it becomes negative for $x \lesssim \xth$. 
-That is fully consistent with the problem objective - for
-$t<\T$, we want to distance $X_t$ from the threshold to avoid an early spike.
+A few words on the terminal conditions (TCs). $\Exp[\t^2|x, \amax]$ is the
+expected time to spike (reach the threshold) starting at $x$ and applying
+$\amax$ throughout. We will not show it here, but this quantity can be found
+easily for all $x$ in the domain by solving a stationary backward Kolmogorov
+equation.
+
+An example solution for
+\cref{eq:OC_LS_mean_HJB_bounded_control,eq:OC_LS_variance_energy_quadratic_control} is shown in \cref{fig:HJB_attempt}.
+
 \begin{figure}[h!]
@@
 
-\section{Deterministic Control}
-\label{sec:deterministic_ode_control}
-Now we will compute the optimal control in the case that $\b = 0$. The reason we
-do this is that since we have assumed that we have no feedback, we attempt to
-devise an open-controller by controlling the non-random dynamics. The
-calculation of this control turns out to be very simple:
-
-First let us assume that we actually can get to $\xth$ from $0$ in $\T$ time
-given $\amax$ and no noise.
-
-Then 
-\begin{equation}
-\begin{gathered}
-J = \int_0^\T \e \a^2 \intd{t}
-\\
-\st
-\begin{cases}
-\dot{x} = (\a(t) - \frac{x}{\tc})
-\\
-x_\T = \xth
-\\
-x_0 = .0
-\end{cases} 
-\end{gathered}
-\end{equation}
-The (Pontryagin) Hamiltonian becomes:
-\begin{equation}
-\H(x,p, \a) = \e \a^2 + p \cdot (\a - x/\tc )
-\end{equation}
-the adjoint is given by:
-\begin{equation}
--\dot{p} = \di_x \H = - p / \tc 
-\end{equation} 
-and the (unconstrained) control by:
-\begin{equation}
-\a(t) = \frac p {2 \e}
-\end{equation}
-which looks very similar to the stochastic case, once you put $p = \di_x v$.
-
-We can actually solve for the adjoint analytically: 
-$$
-p = p_0 \exp(t / \tc)
-$$
-where we can freely choose $p_0$ to later recover the TCs for $x$.
-$$
-\a(t) = \frac{p_0 \exp(t / \tc)}{2\e}
-$$
-Finally the solution for $x$ is:
-\begin{align*}
-\dot{x} & = \a(t) -  \frac{x}{\tc}
-\\ & = \frac{p_0 \exp(t / \tc)}{2\e} - \frac{x}{\tc}
-\\
-e^{t/\tc}x &= \int^t e^{s/\tc}\frac{p_0 \exp(s / \tc)}{2\e} \intd{s}
-\\
-x &= e^{-t/\tc} \cdot \frac{p_0 }{2\e} \tc \left[  \exp(2 t / \tc) -1 \right] 
-\\
-&= \frac{p_0 }{\e} \tc \cdot  \sinh (t/ \tc)
-\end{align*}
-And so: 
-$$
-x(\T) = \xth \implies p_0 = \e \frac{\xth  }{ \tc \sinh (\T/ \tc) }
-$$
-And we have a full solution.
-
-If we have saturation, i.e. if $\a$ hits its upper bound, then the situation is
-mildly more complicated. Since $p$ is monotonically increasing, as long as $p_0
-> \amin$ then $\a(t) > \amin \forall t$. So the saturation is simplified to:
-\begin{equation}
-\a(t) = \min \left(\frac{p_0 \exp(t / \tc)}{2\e} , \amax \right)
-\label{eq:deterministic_control_bounds}
-\end{equation}
-In particular, $\a$ will start by following $p(t)$ and then eventually hit its
-upper bound and stay there until $\T$.
-
-The only complication then is to calculate when $\frac{p_0 \exp(t / \tc)}{2\e} =
-\amax$ and then redo the calculation for $x$ with the new control in order to
-obtain the new value of $p_0$. Or we can just brute-force solve for $p_0$, by a
-simple shooting method. Using the latter approach, a solution-control pair for
-the parameter set we will use later is shown in \cref{fig:deterministic_soln_controls}. 
-%\usepackage{graphics} is needed for \includegraphics
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.7\textwidth]{Figs/ControlSimulator/deterministic_example.png}
-  \caption[]{An example trajectory for the deterministic dynamics. Above is the
-  state evolution, $x(t)$ given $\b=0$ and below is the applied control,
-  $\a(t)$. }
-  \label{fig:deterministic_soln_controls}
-\end{center}
-\end{figure}
-
 \section{Open-loop Stochastic Control}
-As we have mentioned before, when the value of $X_t$ is unknown, the best one
-can do is to use the transition density to perform the optimization. Since the
-transition density follows a (deterministic) PDE, we will thus apply a Maximum
-Principle for PDEs as a method of obtaining the optimal control.
-
-\subsection{Fokker-Planck equation for the state density evolution}
-
-We will need the transition density of $X$, conditional on no
-spikes having occurred. We will write this as:
-\begin{align*}
-f(x,t) \intd{x} &= \Prob[X_t\in \intd{x} | X_0 = .0, X_{s<t} < \xth]  								  
-\end{align*}
-Then $f$ satisfies a Fokker-Planck equation with absorbing boundaries
+As we have mentioned before, when the value of $X_t$ is unknown, one should use
+the transition density to perform the optimization. The transition density, $f$,
+of $X$, conditional on no spikes having occurred satisfies a Fokker-Planck
+equation with absorbing boundaries
 \begin{equation}
@@
 				\frac{\b^2 }{2}\cdot \di^2_x f -  
-				\di_x \Big[ (\a(t)- \frac{x}{\tc})  \cdot f \Big]
+				\di_x \Big[ \left(\a(t)- \frac{x}{\tc}\right)  \cdot f \Big]
 \\
@@
 	\left\{ \begin{array}{lcl}
-	 f(x,0) &=& \delta(x) \quad (\textrm{delta function}
+	 f(x,0) &=& \delta(x) \quad \textrm{delta function}
 	\\
 	(\a(t)- \tfrac x \tc)f - \di_x \tfrac {\b^2}2 f] \big|_{x=\xmin} &\equiv& 0
-	\quad (\textrm{reflecting BCs at some } \xmin )
-	\\
-	f(x,t) |_{x=\xth} &\equiv& 0 \quad (\textrm{absorbing BCs at } \xth)
+	\quad \textrm{reflecting BCs at some } \xmin 
+	\\
+	f(x,t) |_{x=\xth} &\equiv& 0 \quad \textrm{absorbing BCs at } \xth
 \end{array} \right. 
@@
 \end{equation}
-In theory, $\xmin = -\infty$, but in the numerics below we will need to
-truncate it to some finite value.
-
-The probability flow, $\phi$, here is
+
+The $f$ dynamics can also be written as
+$$
+\di_t f(x,t) = - \di_x \phi(x,t)
+$$
+for the probability flow, $\phi$, 
 $$
@@
 $$
-as such the $f$ dynamics can also be written as:
-$$
-\di_t f(x,t) = - \di_x \phi(x,t)
-$$
-and the lower BC is:
+Then the lower BC is
 $$
@@
 
-We will also need a short hand notation for the differential operator on the
-right side of \cref{eq:FP_pde_OU_PDF}. Let
-$$
-\L_\a[\cdot] := \di^2_x [ \frac{\b^2 }{2} \cdot] -  
-				\di_x  [ (\a(t)- \frac{x}{\tc}) \cdot]
-$$
-This is a linear and bounded differential operator. . We will usually omit the
-subscript $\a$, but have written it now, since the operator is parametrized by
-$\a$, that is by the control.
-
-\subsection{Restating the objective in terms of $f$}
-Let us recall our objective, \cref{eq:OC_LS_variance}:
+With this, our objective, \cref{eq:OC_LS_variance}, becomes
 $$
@@
 Rewriting this in terms of $f$ reads:
-\begin{multline}
-J(\a) = 
-\int_\xmin^\xth \Ttwo(x) f(x,\T) \intd{x}
-\\
-+ \int_0^\ts \phi(\xth, s) (s-\T)^2 \intd{s}
-\\
-+  \e \int_0^\ts  \a^2(s)  \int_\xmin^\xth f(x,s) \intd{x} \intd{s}
+
+\begin{align}
+J(\a) =& 
+\int_\xmin^\xth \Exp[\t^2|x, \amax] \cdot f(x,\T) \intd{x}
+\notag
+\\
+&+ \int_0^\T \phi(\xth, t) (t-\T)^2 \intd{t}
+\notag
+\\
+&+  \e \int_0^\ts  \a^2(t)  \int_\xmin^\xth f(x,t) \intd{x} \intd{t}
 \label{eq:OC_LS_variance_density}
-\end{multline}
-
-Let us explain in more detail each term on the RHS in
-\cref{eq:OC_LS_variance_density}:
-
-The first term
-$$
-\int_\xmin^\xth \Ttwo(x) f(x,\T) \intd{x}
-$$
-counts the cost of trajectories which spike too late:
-this cost is the expected squared time-to-hit starting at $x$, with  $\a
-= \amax$, weighted by the probability of $X_\T = x$ , $f(x,\T)$. Recall
-that we calculated $\Ttwo(x)$ in \cref{sec:calculating_TCs}.
-
-The second term
-$$
-\int_0^\ts \phi(s,t) (s-\T)^2 \intd{s}
-$$
-counts the cost of trajectories which spike too soon, that
-is the squared difference between realized spike time, $t$ and desired spike
-time $\T$, weighted by the probability of spike at $t$ which is just the outflow of
-probability, $\phi(\xth,t)$. Note further that due to the homogeneous
-Dirichlet BC on $f$ at $\xth$, the outflow is simply:
-$$
-\phi(\xth, t) = -\frac{\b^2 }{2} \di_x f(\xth, t) 
-$$
-
-Finally, the third term
-$$
-\e \int_0^\ts  \a^2(s)  \left[  \int_\xmin^\xth f(x,s) \intd{x} \right] \intd{s}
-$$
-is the energy cost: the inner integral, $\int_\xmin^\xth f(x,s) \intd{x}$
-takes into account that we incur an energy cost only for those trajectories
-which have not yet spiked.
-
-With that our control $\a$ will naturally be found via:
-$$
-\a = \argmin J[\a]
-$$
-
-\subsection{Optimizing with $f$ using a Pontryagin-type Maximum Principle}
-\label{sec:PDE_max_principle_for_pdf}
-By now our ostensibly stochastic optimal control problem modelled by SDEs
-has become a deterministic optimal control problem modelled by PDEs. Our control
-$\a$ influences the evolution density $f$ and via $f$, the integrals in the
-objective, $J$.
-
-As in the ODE theory, we will introduce an adjoint function, $p$, which will be
-crucial in setting up necessary conditions for the optimal control.
-
-The basic formulation for the adjoint PDEs and the abstract Hamiltonian can be
-found in \cite{Fattorini1999,Palmer2011}. However our cost functional is a
-little strange as it involves values of $\di_x f$ at the boundary. As such we
-will derive the optimality condition heuristically, appealing to the
-references for rigorous proofs.
-
-Before proceeding, one might wish to read
-\cref{sec:Pontryagin_heuristic_derivation} for a simpler version of the
-following derivation with exactly the same ideas.
-
-We start by assuming that $f,\a$ are the optimum trajectory/control pair and we
-ask what necessary conditions must be true for $f,\a$. In particular if we take
-a small variation around $\a$, we will have:
-\begin{align*}
-\a_\e = \a + \e \da
-\\
-f_\e = f + \e \df
-\end{align*}
-
-With that the variation of the cost with respect to the control at the optimal
-$\a$ can be calculated as:
-\begin{equation}
-\frac {dJ}{d\e} \Big|_{\e = 0} =
-- \int_0^\T (s-\T)^2 \frac {\b^2}{2} \di_x  \df \intd{s}  
-+ \int_\xmin^\xth \Ttwo\df\intd{x}
-+ \tilde{\e} \int 2 \a \da \intd{s}  
-\label{eq:J_variation_PDE}
-\end{equation}
-This must be non-negative for a true minimum, but before we can say more, we
-need to eliminate $\df$ from $\frac {dJ}{d\e}$! In order to eliminate $\df$
-from \cref{eq:J_variation_PDE} we will introduce an adjoint variable, $p$. What
-we would want from this adjoint variable $p$ would be that:
-\begin{equation}
-p(x,\T) = \Ttwo(x)
-\label{eq:adjoint_TC_desideratum}
-\end{equation}  
-and 
-\begin{equation}
-\di_t \left[ \int_\xmin^\xth  p \cdot\df \intd{x} \right] = (s-\T)^2 \frac
-{\b^2}{2}
-\di_x
-\df
- + \text{stuff}
-\label{eq:adjoint_dynamics_desideratum}
-\end{equation}
-where 'stuff' does not depend on $\df$.
-
-If we had such a $p$ satisfying
-\cref{eq:adjoint_TC_desideratum,eq:adjoint_dynamics_desideratum} then
-\cref{eq:J_variation_PDE} would read:
-\begin{equation}
-\frac {dJ}{d\e} \Big|_{\e = 0} =
-\tilde{\e} \int 2 \a\da  \intd{s} +   \int_0^\T  \text{'stuff'}\intd{s}
-\label{eq:J_variation_PDE_nodf}
-\end{equation}
-which does not depend on $\df$! and we would be on our way. In order to find
-$p(x,t)$ we will need to
-\begin{enumerate}
-  \item calculate $\di_t \df$
-  \item specify $\di_t p $ st. \cref{eq:adjoint_dynamics_desideratum} holds.
-\end{enumerate}
-
-\subsection{The PDE for $\df$}
-First, calculate the PDE that $\df$ satisfies:
-\begin{multline}
-\di_t f_\e =  \L_{\a_\e} [f_\e]
-\\
-= -\di_x [ (\a_\e - \tfrac{x}\tc) f_\e  - D \di_x f_\e]
-\\
-=  -\di_x [ (\a + \e\da - \tfrac{x}\tc) f_\e  - D \di_x f_\e]
-\\
-=  -\di_x [ (\a + \e\da - \tfrac{x}\tc) (f + \e \df)  - D \di_x (f + \e \df)]
-\\
-=  -\di_x [ (\a + \e\da - \tfrac{x}\tc) (f)  - D \di_x (f)]
-- \di_x [ (\a + \e\da - \tfrac{x}\tc) (\e \df)  - D \di_x (\e \df)]
-\\
-= \L_\a f  -\di_x [ \e\da\cdot  f]
-  +\e \L_\a \df + o(\e) 
-\\
-= \di_t f + \e \di_t \df
-\end{multline}
-Of course, we have $\di_t f = \L_\a f$, so we conclude that:
-\begin{equation}
-\di_t \df = \L_\a \df -\di_x [\da \cdot f] 
-\end{equation}
-Now what are the ICs? Well, both $f_\e$ and $f$ must satisfy the system ICs in
-\cref{eq:FP_pde_OU_PDF}, so $$\df (x,0) = 0$$.
-Finally, what are the BCs?
-On the upper side:
-$$
-f_\e |_\xth = 0 = [ f  + \e \df |_\xth = 0 + \e \df|_\xth \implies \df |_\xth =
-0 $$
-While on the lower side:
-\begin{align*}
-0 =& [ (\a_\e - x/\tc)f_\e - D \di_x f_\e |_\xmin
-\\
-=& (\a +\e \da - x/\tc)(f + \e \df ) - D \di_x (f + \e \df) |_\xmin
-\\
-=& [(\a +\e \da - x/\tc)(f) - D \di_x (f) |_\xmin
-\\
-& + \e [ (\a +\e \da - x/\tc)(\df) - D \di_x (\df) |_\xmin
-\\
-=& \cancel{[(\a - x/\tc)f - D \di_x (f) |_\xmin}
-\\
-&+ \e \da f|_\xmin
-\\
-&+ \e [ (\a - x/\tc)(\df) - D \di_x (\df) |_\xmin   + o(\e)
-\end{align*}
-
-so $\df$ satisfies the following system of equations:
+\end{align}
+
+To find the optimal control, $\a = \argmin J[\a]$, the Maximum Principle,
+\cite{Ahmed1981} introduces an adjoint variable, $p$,
+which solves a PDE related to the PDE satisfied by the density $f$ and then calculates the optimal
+control, $\a$ as a function(al) of $f$ and $p$.
+
+In short, the equation for the adjoint, $p$, is 
 \begin{equation}
@@
 \begin{aligned}
-\di_t \df 
-&= \L_\a \df -\di_x [\da \cdot f]
-\\
-&= -\di_x \left[ (\a - \tfrac x\tc)\df - \tfrac{\b^2}2 \di_x \df \right]
--\di_xf \cdot\da
-\end{aligned}
-\\
-\begin{cases}
-\df(x,0) &= 0
-\\
-[ (\a - \frac x \tc)(\df) - D \di_x (\df) |_\xmin &= -\da \cdot f|_\xmin
-\\
-\df|_\xth &= 0
-\end{cases}
-\end{gathered}
-\label{eq:FP_variation_near_optimum}
-\end{equation} 
-With \cref{eq:FP_variation_near_optimum} we are ready to design the dynamics of
-$p$ such that $\df$ is eliminated from the optimality equation,
-\cref{eq:J_variation_PDE}.
-
-\subsection{Calculating the adjoint }
-Recall our desideratum, we want
-$$
-p(x,\T) = \Ttwo(x) \quad (\text{required TCs})
-$$
-and
-$$
-\int_0^\T \di_t (p \cdot \df) = \int_0^\T(s-\T)^2 \frac {\b^2}{2} \di_x 
-\df \intd{s} + \text{stuff}
-$$
-This is equivalent to having
-$$
-\di_t (p \cdot \df) =  (s-\T)^2 \frac {\b^2}{2} \di_x 
-\df  + \text{stuff}
-$$
-Let us calculate $\di_t (p \cdot \df)$
-\begin{align*}
-\di_t (p \cdot \df) =& \df \cdot \di_t p + p \cdot   \di_t \df
- \\
- =& \df \cdot \di_t p + p \cdot \left( \L [\df] -\di_xf \cdot\da \right)  
- \\
- =& \df \cdot \di_t p + p \cdot \L  [\df] - \underbrace{ p \cdot \di_xf \cdot\da}_{\text{stuff}}
-\end{align*}
-Now our job is clear - we want 
-$$
-\di_t p = -\Lstar[p]
-$$
-Where $\Lstar$ is defined by:
-\begin{equation}
-\begin{gathered}
-\int_\xmin^\xth  \Lstar [p] \df \intd{x} 
-=
-\int_\xmin^\xth p \L [\df] \intd{x} - (s-\T)^2 \frac {\b^2}{2}\di_x \df +
-\text{possibly more stuff}
-\\
-\\
-\forall \df \quad \st \begin{cases}
-\df(x,t) |_{x=\xth} & = 0 
-\\
-(\a - \tfrac x \tc)\df - \di_x \tfrac {\b^2}2 \df \big|_{x=\xmin} &=  -\da
-\cdot f
-\end{cases}
-\end{gathered}
-\end{equation}
-where 'possibly more stuff' should be independent of $\df$.
-
-Let us take a deep breath and calculate $\Lstar$:
-\begin{align*}
-\int p \L [\df] \intd{x} =&
-\int p \underbrace{\left(\frac{\b^2 }{2}\cdot \di^2_x \df -  
-			 \di_x \big[ (\a- \frac{x}{\tc})\cdot \df \big]\right)}_{\L [\df]} \intd{x}
-\\
-=& [ -p ((\a- \frac{x}{\tc})  \cdot \df)\quad  \vert^\xth_\xmin
-+ \int \di_xp ((\a- \frac{x}{\tc})  \cdot \df) \intd{x}
-\\
-& + [p (\frac{\b^2 }{2}\cdot \di_x \df \quad \vert ^\xth_\xmin
-- [ \di_x p \cdot \frac{\b^2 }{2}  \df\quad \vert^\xth_\xmin 
-+ \int \frac{\b^2 }{2} \di^2_xp    \cdot \df \intd{x} 
-\\
-=& \int \Lstar [p] \df \intd{x} 
-+\underbrace{[-p(\a- \frac x \tc) \cdot \df 
-+ p \frac{\b^2}{2} \di_x \df 
--\di_x p \frac{\b^2}{2} \df \,\vert^\xth_\xmin }_{\text{BC terms}}
-\end{align*}
-
-Now we want to make the 'BC terms' equal to $(s-T)^2 \frac{\b^2}{2} \di_x \df$
-+ 'possibly more stuff'.
-In doing so, we will discover the correct BCs associated with the adjoint
-operator, $\Lstar$:
-\begin{align*}
-\textrm{BC terms} =& 
-\Big[ -p (\a- \frac{x}{\tc})  \cdot \df 
-+ p \frac{\b^2 }{2}\cdot \di_x \df   
-- \di_x p \cdot \frac{\b^2 }{2}  \df   \quad ^\xth_\xmin\vert 
-\\
-=& -\cancel{p (\a- \frac{x}{\tc})  \cdot \df \Big|_\xth}
- + p (\a- \frac{x}{\tc}) \cdot \df \Big|_\xmin
-\\ 
-& + [p \frac{\b^2 }{2}\cdot \di_x \df \Big|_\xth
- - [p \frac{\b^2 }{2}\cdot \di_x \df \Big|_\xmin
-\\
-&- \cancel{[ \di_x p \cdot \frac{\b^2 }{2}   \df\Big|_\xth}
- + [ \di_x p \cdot \frac{\b^2 }{2}   \df\Big|_\xmin
-\\
-=& \cancelto{-p (-\da f)}{-p ((\a- \tfrac x \tc)\df - \di_x \tfrac {\b^2}2
-\df) \Big|_\xmin} 
-+ [p \frac{\b^2 }{2}\cdot \di_x f \Big|_\xth
-+ [ \di_x p \cdot \frac{\b^2 }{2}   f\Big|_\xmin
-\\
-=& (s-T)^2 \frac{\b^2 }{2}\cdot \di_x \df\Big|_\xth 
-+  p (\da f)\Big|_\xmin
-\end{align*}
-
-Since we want this to be true for arbitrary $\df$, we need to enforce the
-following (adjoint) BCs: 
-$$
-\begin{cases}
-p |_\xth &= (s-T)^2
-\\
-\di_xp |_\xmin &= 0
-\end{cases}
-$$
-
-In summary, the adjoint PDE equation is given by: 
-\begin{equation}
-\begin{gathered}
-\begin{aligned}
-\di_t p =& - \Lstar[p]
-\\
-		=&  
+\di_t p =&   
 			- \Big[ \frac{\b^2 }{2}\cdot \di^2_x p +  
@@
 
-And most importantly:
-\begin{equation}
-\int \Lstar [p] \df \intd{x}
-=  
-\int p \L [\df] \intd{x} 
-  - (s-T)^2 \frac{\b^2 }{2}\cdot \di_x \df\Big|_\xth 
-  -  p  f \cdot \da \Big|_\xmin
-\label{eq:L_to_Lstar_relation}
-\end{equation}
-
-\subsection{A Pontryagin-type Minimum Principle for the optimal control $\a$}
-Using \cref{eq:L_to_Lstar_relation} and the TCs in \cref{eq:adjoint_pde_OU}, we
-will have:
-\begin{align*}
-\int \Ttwo \df \intd{x} =& 
-\int_\xmin^\xth p \df \intd{x}
-\\ =& \int_0^\T \di_t \int_\xmin^\xth p \cdot \df \intd{x} \intd{s}
-\\
-=& \int_0^\T \int _\xmin^\xth - \df \cdot \Lstar[p]
- + p \cdot \L_\a [\df] 
- -  p \cdot \di_x [\da \cdot f]  \intd{x}  \intd{s}
-\\
-=& 
-\int_0^\T \Big\{  (s-T)^2 \frac{\b^2 }{2}\cdot \di_x \df\Big|_\xth 
-+  p (\da f)\Big|_\xmin
-\\
-& \quad
-- \int _\xmin^\xth p \cdot \di_x [\da \cdot f] \intd{x} \Big\} \intd{s}
-\end{align*}
-
-Now put this back into the equation of optimality, \cref{eq:J_variation_PDE}.
-\begin{align*}
-\frac {dJ}{d\e} \Big|_{\e = 0} =&
-- \int_0^\T (s-\T)^2 \frac {\b^2}{2} \di_x  \df \intd{s}  
-+ \int_\xmin^\xth \Ttwo\df\intd{x}
-+ \tilde{\e} \int 2 \a \da \intd{s}
-\\
-=& 
-\int_0^\T \Big\{
- p (\da f)\Big|_\xmin
-+ \tilde{\e}  2 \a \da
-- \int _\xmin^\xth p \cdot \di_x [\da \cdot f] \intd{x} 
- \Big\} \intd{s}
- \\
-=& 
-\int_0^\T \da \Big\{
- p f \Big|_\xmin
-+ \tilde{\e}  2 \a
-- \int _\xmin^\xth p \cdot \di_x f \intd{x} 
- \Big\} \intd{s}
-\end{align*}
-
-Now we pull the famous fundamental lemma of the calculus of variations out and
-conclude that
+and then $\a$ can be found via
 \begin{equation}
@@
 
-Working in reverse we can recognize a Hamiltonian, $\H$ as:
-$$
-\H(f,p,\a) =  \int \cref{eq:J_necessary_condition} \intd{\a} 
-$$
-
-Naturally, $f,p$ must satisfy their respective
-\cref{eq:FP_pde_OU_PDF,eq:adjoint_pde_OU}.
-
-And our optimal control, $\a$,  can be found via:
-\begin{equation}
-\a (t) = \frac { - p(\xmin,t) f(\xmin,t)
-+ \int _\xmin^\xth p(x,t) \cdot \di_x f(x,t) \intd{x} 
-}{2\e}
-\label{eq:optimal_alpha_4_PDF_optimization}
-\end{equation}
-As always, subject to the appropriate bounds, \cref{eq:bound_constraints_alpha}.
-
-\subsection{An Iterative Numerical Solution}
-We have an entangled situation: $f,p$ depend on $\a$ through
-\cref{eq:adjoint_pde_OU,eq:FP_pde_OU_PDF}, while $\a$, the optimal control, is
-given by \cref{eq:optimal_alpha_4_PDF_optimization} which requires  the
-solution to $f,p$ . So we need some kind of iteration to find $\a$ and the corresponding $f,p$.
-
-The key to such an iteration is that the quantity 
-$$\di_\a \H =  \Big\{
+More practically, the quantity, 
+$$\di_\a J =  \Big\{
  \tilde{\e}  2 \a(t)
@@
 $$
-gives us the direction of increase
-and we can use it as a gradient in a descent algorithm. 
-This we can also write as:
-\begin{align*}
-\di_\a \H =&  \Big\{
- \tilde{\e}  2 \a(t)
-+ p(x,t) f(x,t) \Big|_\xmin
-- [p(x,t) f(x,t) \, _\xmin^\xth\Big|
-+ \int _\xmin^\xth \di_x p \cdot f(x,t) \intd{x} 
-\Big\}
-\\
-=&
- \Big\{
- \tilde{\e}  2 \a(t)
-- p(x,t) f(x,t)\Big|_\xth
-+ 2p(x,t) f(x,t) \Big|_\xmin
-+ \int _\xmin^\xth \di_x p \cdot f(x,t) \intd{x} 
-\Big\}
-\end{align*}
-which has certain numerical advantages.
-
-So the plan is: we choose an initial $\a_1(t)$ - the deterministic solution
-from \cref{sec:deterministic_ode_control} is a natural choice. Then solve for $f,p$.
-Then evaluate $J$ and iterate until convergence. For our convergence
-criterion we will take that the incremental difference in $J$ is smaller than
-some difference small tolerance. The algorithm is specified more
-precisely in Algorithm \ref{alg:hamiltonian_descent_4_PDF_OC}.
-\begin{algorithm}
-\begin{algorithmic}
-\State $J_0 = \infty $ 
-\State $\a_1 = $ deterministic $\a$, the optimal control given $\b=0$ from
-\cref{sec:deterministic_ode_control}
-\For { $k= 1\dots k_{\max}$}
-	\State $\di_t f_{k} = \L_{a_{k}} [ f_{k}]$
-	\State $-\di_t p_{k} = \Lstar_{a_{k}} [ p_{k}]$
-	\State $J_{k} \gets J[\a_k, f_k]$
-	\If {$|J_{k} - J_{k-1}| < tol $}
-		\State BREAK
-	\EndIf
-	\State $\da = \di_a \H(f_{k} ,p_{k},  \a_{k})$
-
-	\If {$\sup_t |\da(t)| < tol$}
-		\State BREAK
-	\EndIf
-
-	\State $\a_{k+1} \gets \a_{k} - s \cdot \da  $
-\EndFor
-\State \Return $J_k, \a_k$
-\end{algorithmic}
-\caption{Optimal Control Descent Algorithm}
-\label{alg:hamiltonian_descent_4_PDF_OC}
-\end{algorithm}
-
-We show one realization of $\di_a \H$ in \cref{fig:grad_aH_deterministic_alpha}.
-\begin{figure}[htp]
-\begin{center}
-\subfloat[$\a_0$]
-{\includegraphics[width=.49\textwidth]{Figs/FP_Adjoint/Deterministic_alpha_di_aH.png}
- }
-  \subfloat[$\a_1$]
-{  \includegraphics[width=.49\textwidth]{Figs/FP_Adjoint/First_control_iteration.png}
-}
-\caption[labelInTOC]{Iteratively calculating $\a$ for the
-open-loop optimal control. Plotted is an example of
-$-\di_a \H$ calculated at $\a(t) = \a_1(t)$, where $a_1$ is the deterministic control. The green curve, $-\di_a \H$
-indicates the direction that we want to change the control, if it is negative
-we decrease and if it is positive we increase it. On the right panel, b), we
-show the result of applying the incremental change by also plotting the
-resulting $\a_1(t)$. }
-\label{fig:grad_aH_deterministic_alpha}
-\end{center}
-\end{figure}
-
-Finally, we are ready to calculate the open-loop stochastic optimal control for
-this parameter set. The results are in
+gives us the direction of increase of $J$ at $\a(t)$
+and we can use it as a gradient in a descent algorithm, given some initial
+guess, $\a_0(t)$, for the control.
+
+We show an iteration for finding the optimal control in
 \cref{fig:FP_adjoint_objective_control_convergence}.
-
 \begin{figure}[h]
@@
 \label{sec:probabilistic_numerical_test}
-We will now compare the performance of the three obtained controls: the
-closed-loop stochastic obtained via the HJB equation, the open-loop
-deterministic control obtained when setting $\b=0$ and the open-loop stochastic
-control obtained by optimizing using the transition density.
-
-For our comparison we will sample $128$ realizations of the controlled system
-and apply in turn each of the three controls. Naturally, for each path, we will
-apply exactly the same random disturbances to each controller.
-
-We will set $\e = .001$ and $\amax = 2.0$. As such the energy cost is of
+We now compare the performance of the optimal closed-loop and open-loop controls
+as well as a very naive open-loop controller, which is obtained by assuming
+$\b=0$. This naive controller is called 'Deterministic', since it assumes
+deterministic dynamics in $X_t$.
+
+For the comparison we sample $128$ realizations of the controlled system
+and apply in turn each of the three controls.
+
+We set $\e = .001$ and $\amax = 2.0$. As such the energy cost is of
 secondary importance and the paramount effect on the objective is the difference
@@
 The performance of the three controllers is given in
-\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch}. We see that while the
-open-loop stochastic control is not as good as the feedback stochastic
-control, it is much closer in performance to it than to the blind
-deterministic control.
-
-For illustration sake, we also show the error
-histograms in \cref{fig:error_histograms_det_vs_openloop_vs_stoch} and visualize
-some trajectories in \cref{fig:control_trajectories_examples}.
+\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch} and in
+\cref{fig:error_histograms_det_vs_openloop_vs_stoch}.
 
@@
 \begin{tabular}{lcc}
-Control Law & Empirical Squared Error & Theoretical Squared Error \\
+Control Law & Squared Error &  Squared Error \\
+ & Empirical & Theoretical\\
 \hline
@@
 \end{table}
+
 \begin{figure}[h]
@@
   deterministic (left) vs. open-loop stochastic(centre) vs. feedback stochastic
-  (right) control laws. This is for the same problem as in
-  \cref{fig:control_trajectories_examples}. We have used $N=128$ sample paths
+  (right) control laws, using the same parameter set as in \cref{fig:control_trajectories_examples}. We have used $N=128$ sample paths
   to form the statistics}
@@
 
-\section{Research Goals}
-I hope to be able to use the techniques of open-loop stochastic control
-discussed herein to address problems in stochastic control where for whatever
-reasons, the dynamic programing formalism is not applicable.
-
-One such reason might be that the state is unobservable. An example of such a
-system is the control of blood glucose for diabetes or intensive care
-patients, where the patients' blood glucose is modelled to follow a stochastic
-differential equation and actual observations are expensive and invasive and
-thus rare, on the order of a few hours. In this scenario, the applied external
-glucose or insulin to the patient in the time in-between measurements can be
-thought of as the solution to an optimal control problem where the objective is
-to minimize deviations of the patient's blood glucose from some baseline.
-
-An alternative reason to use the open-loop control techniques is when optimizing
-with multi-dimensional systems at which point the dynamic programing approach
-suffers from 'the curse of dimensionality' which is the colloquial name for the
-exponential increase in necessary computing resources. If the optimality
-conditions for the open-loop control lend themselves to a Feynman-Kac-type
-representation, then one might replace solving the PDEs by an appropriate
-Monte-Carlo process and thus make the computations tractable.
-
-\clearpage
-\appendix
-\section{Heuristic Derivation of Pontryagin's Minimum Principle}
-\label{sec:Pontryagin_heuristic_derivation}
-Here we will show a heuristic derivation for the Pontryagin Minimum Principle
-for linear ODE dynamics. It should motivate the manipulations in
-\cref{sec:PDE_max_principle_for_pdf}.
-
-The following is almost a direct transcription of A.1 in
-\cite{Evansb}, with the added simplification of linear dynamics
-and linear cost and with the notation adapted to what we have used
-hereto.
-
-The dynamics are given by:
-$$
-\dot{x} = A(t, \a) x
-$$
-where $A$ is a matrix in $\R^{n_x \times n_x}$ whose values depend on the
-time, $t$ and the control $\a$. The cost, $J$, is given by:
-$$
-J[\a] = \int_0^\tf L(t,\a) \cdot x \intd{s} + M \cdot x(\tf)
-$$
-where $L(t,\a), M$
-are vectors in $\R^{n_x}$. Both the dynamics and the cost are linear in the state $x$.
-Suppose $x,\a$ are the optimal trajectory - then small variations in $\a$ can
-only worsen $J$. Let such an acceptable variation be given by:
-$$
-\a_\e = \a + \e \b
-$$
-for some small $\e$. We will have a corresponding dynamic, $x_\e$ st:
-$$
-x_\e = x + \e \dx+ o(\e)
-$$
-Now obtain a differential equation for the variation $\dx$:
-\begin{align*}
-\dot{x_\e} &=  \dot{x} + \e \dot{\dx} 
-\\
-&= A(\a_\e) x_\e = A(\a) x + \e \dot{\dx}
-\end{align*}
-Now divide through by $\e$ and let it go to zero:
-$$
-\implies \dot{\dx} = A \dx + [\grad_\a [ A(t,\a)x]] \cdot\b
-$$
-Note that $$\dx(0) = 0$$ since $x(0)$ is specified and so there is no
-variation there.
-
-With that the derivative of $J$ at $\a$ will read:
-\begin{multline}
-\frac {dJ}{d\e} \Big|_{\a} = \lim \frac{ J[\a_\e] - J[\a]}{\e} 
-= \int L(\a,t) \dx +  \grad_\a L \cdot \b \intd{s} + M \dx(\tf)
-\geq 0 
-\label{eq:J_variation_ODE}
-\end{multline}
-The problem is that the state variation $\dx$ also appears in this derivative,
-whereas we just want the control variation $\b$ to appear there.
-
-The trick then is to introduce the adjoint variable $p$ with appropriate
-dynamics such that all terms involving $\dx$ in \cref{eq:J_variation_ODE} are
-eliminated.
-
-'Clearly', we want to replace the term $ M \cdot \dx $ first. Thus we set 
-$$
-p(\tf) = M
-$$
-
-Now notice the following:
-$$p(\tf)\dx(\tf) = \cancelto{0}{p(0)\dx(0)} + \int_0^\tf d_t (p \cdot \dx)$$
-
-So if we make $\int d_t (p \cdot \dx) = - \int r(\a,t) \dx$ we would be
-eliminate the $\dx$ terms in the expresion for $dJ/d\eps$ entirely!
-
-Now the way is clear:
-
-Let 
-$$ 
-\dot{p} = -A^T p - L(\a,t)
-$$
-Then indeed:
-\begin{multline}
-d_t (p \cdot \dx) = \dot{p} \dx + p \dot{\dx}
-\\
-= (-A^T p - r)\dx + p (A\dx+ [\grad_\a [ A(t,\a)x]] \cdot\b)
-\\
-= -r\dx + p ([\grad_\a [ A(t,\a)x]] \cdot\b)
-\end{multline}
-That is almost what we wanted and it will work, since it only contains
-variations wrt the control $\b$, not the state, $\dx$.
-
-Thus
-\cref{eq:J_variation_ODE} is transformed to read:
-\begin{multline}
-\frac {dJ}{d\e} \Big|_{\a} = 
-\int r(\a,t) \dx +  \grad_\a r \cdot \b \intd{s} + \psi \dx(\tf)
-\\
-=
-\int r(\a,t) \dx +  \grad_\a r \cdot \b \intd{s} + p(\tf) \dx(\tf)
-\\
-= \int r(\a,t) \dx +  \grad_\a r \cdot \b \intd{s} - \int r \dx - p ([\grad_\a [
-A(t,\a)x]] \cdot\b)
-\\
-=
-\int_0^\tf  
-[\underbrace{p ([\grad_\a [A(t,\a)x]] ) + \grad_a r)}_{\di_\a \H} ] 
-\b \intd{s}
-\end{multline}
-As always in these variational arguments, the final point is that since $\b$ is
-arbitrary, we must have that the rest of the integrand be identically zero -
-thus we see where the control theory Hamiltonian, $\H$, comes from.
-
-Except for the nitty-gritty manipulations of the BCs in
-\cref{sec:PDE_max_principle_for_pdf}, what is done there is exactly what is done
-here - a heuristic derivation for a necessary condition for a minimum.
-
 \bibliographystyle{plain}

### 2013-08-22 SingularStochasticControl.tex
Initial commit

\documentclass{article}

\usepackage{amsmath}
\usepackage{amscd}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsthm}

\usepackage{circuitikz}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows,snakes,backgrounds}
% \usetikz
\usepackage{subfig}

\usepackage[super]{nth}
% \usepackage{appendix}
% \usepackage{listings}
% \usepackage{color}

\usepackage{hyperref}
%\usepackage{url}

\usepackage{cleveref}

\usepackage{aviolov_style}
\usepackage{local_style}

\begin{document}

\title{Singular Stochastic Control} \author{Alexandre Iolov 
$<$\href{mailto:aiolo040@uottawa.ca}
		{aiolo040 at uottawa dot ca}$>$}

\date{\today}

\maketitle

\abstract{How to solve HJB equations, whose Hamiltonian is linear in the
control and there is no guarantee that $\di_x v \neq 0$?}

\section{Problem Statement}

Consider a 2nd order HJB equation in 1-d with no running costs and fixed
terminal time $\tf$:
\begin{equation}
\begin{gathered}
dX_s =  (\a(s) - X_s) \intd{s} + \b \intd{W_s} 
\\
J[\a] = M(X_\tf)
\\
\di_t v(x,t) + \inf_{\a} \left\{ \tfrac{\b^2}{2} \di_x^2 v + 
(\a(t) - x) \di_x v \right\} = 0
\\
\ldots + \textrm{BCs, TCs}
\end{gathered}
\label{eq:HJB_singular}
\end{equation} 
We are solving $v(x,t)$ on some closed, bounded domain $\O \subset \R \times
[0,\tf]$.

As long as $\di_x v \neq 0$, we will have a 'bang-bang' type control for
$\a(t)$:
\begin{equation}
\a(t) =
\begin{cases}
\amin & \di_x v > 0
\\
\amax & \di_x v < 0
\end{cases}
\end{equation}
But what to do if $\di_x v= 0$?

There is a solution for this problem in the deterministic optimal control
theory. In the deterministic optimal control, the quantity:
$$
H(x,p) = \inf_{\a} \left\{
(\a(t) - x) p \right\}
$$
is called the 'Hamiltonian'. The problem is called 'singular', when $p=0$, since
then it is impossible to know the value of $\a$ that attains the minimum. The
remedy is to note that along optimal trajectories, $x^*,p^*$ Pontryagin's
Maximum Principle is satisfied, $x,p$ form a Hamiltonian Dynamical System and
the Hamiltonian is an integral of motion - $H = \const$. Thus along optimal
trajectories the full time derivative of $H$ is zero, $d_t H=0$. Thus we can
differentiate $H$ until $\a$ props up again and then set the resulting
expression to zero to obtain an equation for $\a$.

But is there an analogue to this in the Stochastic Case?

For our very simple problem with no running costs, the stochastic Hamiltonian
is simply:
$$
H(x,p,q) = \inf_{\a} \left\{\tfrac{\b^2}{2} q +
(\a(t) - x) p \right\}
$$

\bibliographystyle{plain}
\bibliography{library,local}
% \bibliography{local}

\end{document}

### 2013-08-22 draft.tex
Initial commit

% \documentclass[12pt]{iopart}
\documentclass{article}

\usepackage{amsmath}
\usepackage{amscd}
\usepackage{amssymb}
\usepackage{amsfonts} 
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsthm}

\usepackage{tikz}
\usetikzlibrary{arrows,snakes,backgrounds}
% \usepackage{circuitikz}
% \usepackage{pgf}
% \usetikz
\usepackage{subfig}

\usepackage{algpseudocode}
\usepackage{algorithm}

\usepackage[super]{nth}
\usepackage{appendix}
% \usepackage{listings}
\usepackage{color}

\usepackage{hyperref}
%\usepackage{url}

\usepackage{cleveref}
\usepackage{cancel}
\usepackage{slashbox}

\usepackage{aviolov_style}
\usepackage{local_style}

\newtheorem{thm}{Theorem}[section]
\newtheorem{lemma}{Theorem}[thm]
% \theoremstyle{definition}
\newtheorem{ex}{Example}[thm]
\newtheorem{defn}{Definition}[thm]

% \def\includegraphic{}
% \def\includegraphics{}

\begin{document}

\title{Stochastic Optimal Control of Single Neuron Spike Trains
\\
---
\vskip5pt
Draft
}
\author{Alexandre Iolov, Susanne Ditlevsen, Andr\'e Longtin,
		}		
% \address{Department of Mathematics and Statistics, University of Ottawa, Ottawa, Canada 
% \\ Department of Mathematical Sciences, University of Copenhagen, Copenhagen,
% Denmark }
% \ead{$<$\href{mailto:aiolo040@uottawa.ca}
% 		{aiolo040 at uottawa dot ca}$>$}

\date{\today}

\maketitle

\abstract{
{\em Objective.} External control of exact spike timings in single neurons is
important for basic understanding of neuronal activity, and has
potential for improving neural prostheses and in medical applications
related to neuronal disorders, such as Parkinson's disease or epilepsy.
The goal of this paper is the design of optimal controlled
electrical stimulation patterns to a cell to achieve a target
spike train under physiological constraints not to damage tissue and
minimize possible side effects.
{\em Approach.} We pose an optimal control problem to precisely specify
the spike timing in a noisy leaky integrate-and-fire model of a
neuron. In particular, we allow for the noise to be of arbitrary intensity. The optimal control problem is solved using Dynamic Programming if the controller has access to the voltage (closed-loop control) and using a Maximum Principle for the transition density if the controller only has access to the spike timings (open loop
control).
{\em Main results.} We have developed a stochastic optimal control
scheme to obtain precise spike timings, which is not limited
by the intensity of the noise in the system. Given this scope, the
numerical demands are non-trivial, but simulations show that the
algorithms are potentially feasible. It however remains to see if
they can be successfully applied in a real-time experimental setting.
Furthermore, we show how wrong the control strategy can perform if we
assume that the noise is negligible.
{\em Significance.} More precise targeting and control of neural
activity to achieve optimal performance in a noisy environment, in
which most neurons are embedded, have great potential in health
care. The main contribution is the online feedback control of a noisy
system through
modulation of the input, taking into account physiological constraints
on the control, as well as incurred costs to diminish adverse effects.
}

\tableofcontents

MSC KEYS: 49N90 90C39 93E20 93C20 93C95

\section{Introduction}
Manipulation of individual neurons through external electrical
stimulation provides a means to
controlling the spiking activity. Applications may arise, for
example, in brain-machine interfaces or in artificial prosthetics.
A natural goal is to make a cell produce a pre-specified spike train,
under certain constraints on the modulation of the input.
We consider this problem in the framework of Stochastic Optimal Control and
give both a feedback solution (closed-loop), when the cell voltage is
explicitly observable, as well as an open-loop solution when only the
occurrence of spikes is observable. Importantly, we allow for an
arbitrary noise intensity.

Both theoretically and in practice related problems have been
addressed in the literature. One objective has been to
obtain either minimum or maximum interspike intervals lengths,
when the input is constrained to be between some prespecified upper
and lower bounds, see e.g. \cite{Lee1994,Lefebvre1987} for a
mathematical treatment, or \cite{Danzl2009,Nabi2012,Wu2009} in a neuronal
context. Another objective is to break a pathological synchronous
firing pattern in clusters of neurons, highly relevant for neurological
disorders such as epilepsy and Parkinson's disease,
\cite{Nabi2013a,Nabi2011}, see also \cite{Feng2007b}.

Our objective, namely targeting exact spike times in single neurons, has been
considered mainly in the open-loop context, and in either absence of or for
small noise. In \cite{Ahmadian2011} they use the Spike Response Model,
\cite{Gerstner2002}, to control output target spike trains and implement their
scheme on pyramidal cells in mouse cortical slices. Their method is numerically
efficient and allows for the simultaneous control of many neurons. However the
efficiency of the numerics strongly relies on the assumption that the diffusion
noise, the value of $\b$ in \cref{eq:X_evolution_uo} below, is small enough, see
the discussion above and below equation 12 in \cite{Ahmadian2011}. They only
work with open-loop control. The difference between the objective in
\cite{Ahmadian2011} and what we consider is that they maximize the probability
of spiking at some given time $t^*$, whereas we minimize the mean squared
difference between the realized spike time, $T_{sp}$ and the desired, $t^*$.
Moehlis et al. \cite{Moehlis2006} work with a Phase Response Model to obtain
spikes at exact times, while keeping the root mean square of the input to a
minimum. A similar approach is taken in \cite{Dasanayake2011}. They do not
consider noise, though. In \cite{Nabi2013}, the methods are implemented on brain
slices of pyramidal neurons of rat hippocampus. [[[ALEX, Is it closed- or open
loop? Could you SHORTLY describe the difference of what they do and what we
do?]]] \{ALEX: They use Phase Response Curves, assuming that a spike coincides
with $\th=0$; no noise, Pontryagin Maximum Principle, solved numerically $=>$
open-loop. Short Description: Since they do not consider noise, it is
straight-forward to make the neuron spike when you want it, and in fact that can
be done in many ways. Thus their optimization reduces to finding the way to make
the neuron spike at a pre-specified time, using least energy (the integral of
$u^2$.\}. While the Phase Response Curve Model is a parsimonious and effective
way to describe a neuron's response to a stimulus, it is only valid in the
supra-threshold regime, when the unstimulated neuron is periodically spiking.
In \cite{Feng2003} they use a leaky integrate-and-fire model and also
use open-loop control, but to obtain exact spike timings, they choose an
objective of minimizing the variance of the membrane potential at the desired
spike time, while forcing the mean of the membrane potential at this time point
to equal the threshold. This provides exact solutions since it does not involve
first-passage times, but has the drawback of getting a non-negligible
probability of being way-off the objective.

Our objective of imposing a certain timing sequence of the spike train using an
externally applied control is obtained in both the closed- and
open-loop setting, and we specifically include the noise in the
calculations of the controls. We restrict the controlled input to stay
within pre-specified bounds, and also include a cost function to
minimize intervention.
Our main contributions are... (((what we have done which has not been
done before, numerical merits...)))
\{ALEX: Basically, we deal with a significant noise component. Everything
else follows, i.e. the objective only makes sense if the ISI is a random variable.
The PDEs (HJB and Max Principle) only occur because of the noise
component, o/w we would have ODEs. The numerical schemes are simple given the
PDEs, we just use the Crank-Nicholson method for the PDEs and a basic
gradient-descent for the optimization in the open-loop case. The numerics take a
non-trivial time to run - hard to see them coming down to real-time, i.e.
10-20 ms, \ldots but possible. We also explore a wide spectrum of parameter
values, in particular we do  not restrict our attention to the
autonomously-spiking, supra-threshold regime.\}

[[[Susanne, we should mention Whittle \cite{Whittle1996}, but here, or together
with other monographs on optimal control?]]] \{ALEX: Indeed we should mention
Whittle, in particular mention that he discusses analytically the problem in the
closed-loop case (only), but not the open-loop case. He does not provide
numerics in either case. I would refrain from citing other books on
stochastic optimal control, here. Books commonly cited are
\cite{Fleming1975,Krylov2008} and the chapter on HJB equation in Oksendal,
\cite{Oksendal2007}, some of them are cited in later sections\ldots\}

The paper is structured as follows: First we describe the neuron model and
formalize the control objective. Then we describe a feedback-based solution,
which assumes that the controller has detailed access to the voltage trajectory.
Then we relax the observation assumption so that the controller only has access
to the spike timings. Finally, we compare the two methods through
simulations against a simple-minded control technique which ignores
the stochastic input to the neuron.

\section{Problem Formulation}
A basic but useful model for the neural membrane potential evolution is the
noisy leaky-integrate-and-fire model:
\begin{equation}
\begin{gathered} 
dX_s = \left(\Iext(t) - \frac{X_s}{\tc} \right) \intd{s} + \b \intd{W_s},
\\
X(0) = 0,
\\
X(\ts) = \xth \implies   
\begin{cases}
X(\ts^+) = 0 &  
\end{cases}
\end{gathered}
\label{eq:X_evolution_uo}
\end{equation}
Here $X$ represents the membrane electric potential which tries to revert to $0$
with a time-const of $\tc$, $dW$ is a Brownian motion increment scaled by $\b$
and $\Iext(t)$ is the deterministic external input to the cell. Having last
spiked at time $0$, the potential hits $\xth$ at some time $\ts$, the potential
resets to $0$ and we start all over again. We write $\ts^+$ in the formula for
the reset since we are working in continuous time, i.e.\ it is really the
voltage's limit from the right at $\ts$ that is 0. For our purposes, we will
always have that the threshold is set to one, $\xth = 1$.

Suppose that we have some control over the external current such that it can be
decomposed as
\begin{equation}
\Iext(t) = \m + \a(t)
\label{eq:current_mu_alpha}
\end{equation}
where $\m$ is an uncontrollable, but constant part of the external current and
$\a(t)$ is controllable, i.e.\ it can be chosen to achieve some goal. 
A natural goal is to attempt to control the spike time, $\ts$. 
That is we would like to ask how do we choose $\a(t)$ such that $\ts = \T$.
A natural optimal control objective to achieve this is the least squares solution
\begin{equation}
\a(t) = \argmin_{\a(t)} \{\, \Exp[(\ts- \T)^2] \,\}
\label{eq:OC_LS_variance}   
\end{equation}
where expectation is taken with respect to the distribution of the trajectories
of $X_t$.

Often the control has certain constraints. The most common are simple box
constraints:
\begin{equation}
\a(t) \in [\amin, \amax] \quad \forall t
\label{eq:bound_constraints_alpha}
\end{equation}

% \subsection{Optimal Control Formulation}
In addition to \cref{eq:OC_LS_variance}, we will also add to our objective a
running cost based on the control. This regularizes the problem eliminating the
subtleties of singular-control situations and it might be argued appropriate to
avoid excessive control. WE SHOULD DISCCUSS THIS IN GREATER DETAIL, BUT PERHAPS
LATER. (ALEX:\cite{Ahmadian2011} state that they impose a similar control
penalty to avoid excessive charge building up on the cell.)

So we seek an optimal control, $\a^*$, that solves
\begin{align}
J[\a(\cdot)] =&  
\Exp\left[
\e \int_0^\ts  \a^2(s) \intd{s}
+  
(\ts - \T \big)^2 \right]  
\label{eq:OC_LS_variance_energy}   
\\ \notag
\a^*(\cdot) =& \argmin_{\a(\cdot)} J[\a(\cdot)]. 
\end{align}
where $\e$ measure how much weight we put on minimizing the energy expenditure.
If $\e = 0$, then we do not care at all about the expended
energy.

It will often be the case that we write $\a$ and mean either a function or a
value of that function at a particular time. And that this function could be
random, i.e.\ a stochastic process. Since $\a$ is used to control the behaviour
of $X$, it will be at least measurable with respect to the filtration of $X$,
see \cite{Krylov2008}. We will try to make clear below when we are refering to
$\a$ as a stochastic process and when we are merely refering to its particular
value at some time. For example, in \cref{eq:OC_LS_variance_energy}, $\a(\cdot)$
refers to the function, while $\a(s)$ refers to that function's value, possibly
random, at time $s$.

We will consider two control contexts -- {\sl closed-loop} and  {\sl open-loop}
control. In closed-loop control, the value of $X_t$ at time $t$ is observable
and can be used in determining the control. In open-loop control only the spike
times are observable. In the closed-loop context, we might better write the
control as $$\a = \a(x,t)$$ to indicate its dependence on $X_t$ and to express
that it will be updated based on the time-course of $X$. 
In the open-loop context, we would instead write $$\a = \a(t)$$ to indicate that
$\a(\cdot)$ is decided for all times at time 0. 
The techniques used in the two scenarios will be different. For the closed-loop
scenario we will use Dynamic Programing, \cite{Fleming1975}, in order to obtain
the optimal control, while for the open-loop scenario we will use a form of the
Maximum Principle applied on the transition density of the controlled process,
\cite{Ahmed1981,Borzi2012}.

Crucially, we assume that the model parameters, $\m,\tc,\b$ in
\cref{eq:X_evolution_uo} are known.

\subsection{Parameter Regimes}
Different  parameter regimes can be envisioned given \cref{eq:X_evolution_uo},
depending on whether the noise intensity, $\b$, is
relatively high or low, and whether the external, uncontrollable bias current,
$\m$, induces spikes in the absence of noise or not.

Spikes will occur in the absence of noise, if and only
if $$ \m > \frac 1\tc,$$ which is called the supra-threshold regime. When $\m
\leq 1/\tc$, the regime is called the sub-threshold regime.
In addition we will impose two values of $\b$, which we
will call high-noise and low-noise, respectively.
Example values for each parameter regime are given in \cref{tab:regimes}, and we
visualize a single path for each regime in \cref{fig:regime_path_examples}.
\begin{table}
\begin{tabular}{|l||{c}|{c}|}
\hline
\backslashbox{$\m$}{$\b$}
& $1.5$ & $0.3$ \\
\hline
$1.5 / \tc $ &Supra-threshold-High-noise & Supra-threshold-Low-noise \\ 
\hline
$0.1 / \tc$   &Sub-threshold-High-noise & Sub-threshold-Low-noise \\
\hline
\end{tabular}
\caption{Regime labels and example values. Note that for the numerical
experiments below, we use $\tc = 0.5$}
\label{tab:regimes}
\end{table}
%\usepackage{graphics} is needed for \includegraphics
\begin{figure}[htp]
\begin{center} 
  \includegraphics[width=0.99\textwidth]{Figs/PathSimulator/path_T=14_combined.pdf}
  \caption[labelInTOC]{Example Trajectories from \cref{eq:X_evolution_uo}
  using the parameter values from \cref{tab:regimes}. A)
  Supra-Threshold-High-Noise B) Supra-Threshold-Low-Noise C) Sub-Threshold-High-Noise D) Sub-Threshold-Low-Noise}
  \label{fig:regime_path_examples} 
\end{center} 
\end{figure}

\section{Closed-Loop Solution - Dynamic Programing} 
We now detail the Dynamic Programing approach to obtaining a feedback-optimal
control. What this amounts to is that the controller can be continuously updated
depending on the realization of the stochastic process, $X_t$.

Given a time $t$ and a realized voltage value $X_t = x$, let $\trem$ be the
(unknown) remaining time to spike, $\trem = \ts - t$. 
Starting from an arbitrary $t, x$, our remaining-cost
objective, $J[\a(\cdot); x,t]$, will be
\begin{equation}
J[\a(\cdot); x,t]  = 
\Exp\left[
\big(\trem - (\T-t) \big)^2 
+
\e \int_t^\ts  \a^2(s) \intd{s}
\,\Big|\, X_t = x
\right] 
\label{eq:OC_LS_cost_to_go}
\end{equation}
i.e.\ if time $t$ has elapsed without a spike, we now want to minimize
the difference between $\t = \ts - t$ and ($\T-t$), given the current state $x$.
Recall that in the closed-loop scenario, we assume that the value of $x = X_t$
is known to the controller.

\subsection{Hamilton-Jacobi-Bellman equation}
The Hamilton-Jacobi-Bellman (HJB) equation,
see \cite{Fleming1975,Evansb,Krylov2008}, \{ALEX: Should we also cite
Moehlis' papers that use/discuss HJB \cite{Nabi2013a,Danzl2009}???\}
associated with \cref{eq:OC_LS_cost_to_go} is obtained as follows. We introduce the value function, $\v(x,t)$
% \footnote{Traditionally, the value function is denoted as $v$. We
% use $\v$ since we are working in a neural context, where $v$ is so often used
% to denote a voltage}
:
\begin{align}
\v(x,t) =&
 \min_{\a(\cdot)_{s \geq t}}
 \{J[\a(\cdot); x,t] \}
\notag
\\
=&
\min_{\a(\cdot)_{s \geq t}}
\Exp\left[
\big(\trem - (\T-t) \big)^2 
+
\e \int_t^\ts  \a^2(s) \intd{s}
\,\Big|\,X_t = x
\right]
\label{eq:OC_LS_mean_value}
\end{align}

Then $\v$ satisfies the following HJB partial differential equation (PDE) (see
\cite{Fleming1975,Evansb,Krylov2008} for details):
\begin{equation}
\begin{gathered}
\di_t \v(x,t) + \tfrac{\b^2}{2} \di_x^2 \v(x,t) + \\
\min_{\a(x,t) \in [\amin,
\amax]}\Big\{ \e \a^2 (x,t)+ \left(\m + \a(x,t)-\tfrac{x}{\tc}\right) \di_x
\v(x,t) \Big\} = 0
\label{eq:OC_LS_mean_HJB}
\end{gathered}
\end{equation}
The special feature of \cref{eq:OC_LS_mean_HJB} in contrast to a generic
parabolic PDE is that it contains an embedded optimisation that depends on the
solution, $\v$. That is for each $x,t$ in the computational domain,
$\a$ is chosen as to minimize $\{\e \a^2 + \left(\m + \a - \tfrac{x}{\tc}\right) \di_x \v \}$. 
Here the we can solve for the optimal control, $\a^*(x,t)$ analytically as:
\begin{align}
\a^*(x,t)  =& \argmin_{\a \in [\amin, \amax]}\big\{\e \a^2 + (\m + \a-\tfrac{x}{\tc}) \di_x\v\big\}
\notag
\\
=&
\min \left(\amax, \max\left(\amin, -\frac{\di_x\v(x,t)}{2\e}\right)\right)
\label{eq:OC_LS_variance_energy_quadratic_control}
\end{align}
With this the HJB PDE becomes
\begin{equation}
\di_t \v(x,t) + \tfrac{\b^2}{2} \di_x^2\v+
\e (\astar(x,t))^2 + (\m + \astar(x,t)-\tfrac{x}{\tc}) \di_x\v
= 0
\label{eq:OC_LS_mean_HJB_bounded_control}
\end{equation}

Now, we need to consider boundary conditions (BCs) for $\v$. If $X_t = \xth$
then we have a spike now and we know that $\ts = t$. Thus
$$ \v(\xth,t) =
(t-\T)^2.$$ 
At the threshold, the value function equals the squared difference
between the desired spike time and the realized one.
Now for $x \ll .0$, we can assume that $\v$ is not
significantly affected by the change in $x$, i.e.\ that $$ \di_x \v(\xmin, t) =
0 $$ for some lower boundary $\xmin$. Such a boundary condition will be
justified if we impose artificial reflecting boundaries for the diffusion at $\xmin$. So we need to
determine for which $\xmin$ is a reflecting boundary consistent with the
dynamics. For example, we can take $\xmin$ to be two standard deviations
below the mean of the stationary distribution of the maximally
inhibited, $\a=\amin$, \cref{eq:current_mu_alpha}. That is we set $\xmin =
\tc(\mu + \amin) - 2 \b / \sqrt{\tc/2}$. In fact, we further enforce that
$\xmin \leq -0.5$.
% Ooops, we need $\di_x\v\neq 0$ and here we are setting $\di_x\v= 0$ in one of
% the BCs\ldots oh well.

Now we come to the Terminal Conditions (TCs) for $\v$. The idea is simple:
if we reach $\T$ without having spiked we apply maximum control in the
positive direction, i.e.\ $$t>T \implies \a(t) = \amax$$ Thus: 
\begin{equation}\v(x,\T) =
\Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big].
\label{eq:valuefun_TCs}
\end{equation} 
Note that we are making an approximation here
- that is we are ignoring the energy term, $\eps u^2$ in the objective for
$t>T$. Naturally, this approximation is ever more accurate for $\eps \ll 1$.
We discuss in more detail the validity of the approximation in
\cref{sec:effect_of_eps}.
Alternatively, we could impose this approximate terminal condition at some $t^+
> \T$.

We will see the quantity on the right hand side of \cref{eq:valuefun_TCs}
repeatedly so we will give it a special name.
\begin{equation}
\Ttwo(x) := \Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big].
\label{eq:Trem_squared}
\end{equation} 
$\Ttwo$ is the second moment of the remaining time to reach the
threshold starting at $X_\t = x$ and applying $\amax$ throughout. This quantity can be found
easily for all $x$ in the domain by solving a stationary backward Kolmogorov
equation. We show the details in \cref{sec:valuefun_TCs}.

Thus, we restate the HJB equation in its fully specified form:
\begin{equation}
\begin{gathered}
\di_t \v(x,t) + \tfrac{\b^2}{2} \di_x^2\v(x,t)+
\e \a^2(x,t) + \big(\m + \a(x,t) -\tfrac{x}{\tc}\big)\cdot \di_x\v(x,t)
= 0
\\
\a (x,t) = \min \left(\amax, \max\left(\amin, -\frac{\di_x \v(x,t)}{2\e}\right)
\right)
\\
\begin{cases}
\v(\xth,t) = (t-\T)^2  \quad & \textrm{upper BC}
\\
\di_x \v(\xmin, t)  = 0  \quad &\textrm{lower BC}
\\
\v(x,\T)  =
% \Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big] 
% \Exp[\t^2|x, \amax]  \quad
\Ttwo(x)
\,& \textrm{TC}
\end{cases}
\end{gathered}
\label{eq:OC_LS_HJB_full}
\end{equation}
We are solving $\v(x,t)$ over the domain $[\xmin, \xth] \times [0,T]$.

\subsection{The numerical method for the HJB equation}
We now have a PDE for $\v$ and an algorithm for computing all the BCs and TCs of
this PDE. It is time to discuss the numerical method for solving \cref{eq:OC_LS_HJB_full}. Since we are in 1-d the simplest thing to do is try a
centred Finite Difference scheme using Crank-Nicholson to step in time.
Although, it might become necessary, we do not upwind the first order
derivatives for now. Also to resolve the non-linearity, $(\di_x\v)^2$, in the
PDE, we treat it as a mixed implicit-explicit term
$$(\di_x \v(x,t_k))^2 = \di_x
\v(x,t_k) \cdot \di_x \v(x,t_k) \approx \underbrace{\di_x \v(x,t_k)}_{\textrm{implicit}}
\cdot \underbrace{\di_x\v(x,t_{k+1})}_{\textrm{explicit}}.$$ Note that the
implicit term is in the previous time $t_k$ instead of, as is conventional,
the next $t_{k+1}$, because we are solving for $\v$ {\sl backwards} in time. 
TODO: Give justification for this trick.

The numerical scheme is implemented in Python, using the Scipy/Numpy library,
\cite{Scipy}.

TODO(ALEX): Mention time- space- discretization

We next demonstrate solutions for the value function and the associated
control law for a parameter set from each regime in \cref{tab:regimes}.

\subsection{Solutions of the HJB equation}
We set $\T=1.5$ and $\eps = .001$, so that the primary focus is the accurate
spiking and energy minimization considerations are secondary.

In \cref{fig:HJB_4regimes_value_control_cuts}, we depict three snapshots
in time of the value function and the control, $\v$ and $\a$, at $t = [0,\T/2,
\T]$, i.e.\ at the beginning, middle and end of the relevant time interval. Then in
\cref{fig:HJB_4regimes_value_surf} we show the entire value function, $\v(x,t)$,
the solution to \cref{eq:OC_LS_HJB_full}, for each of the parameter regimes.
% THE CUTS:
\begin{figure}[htp]
\begin{center}
  \includegraphics[width=.99\textwidth]{Figs/HJB/Regimes_vc_cuts.pdf}
  \caption[labelInTOC]{Snapshots of the value function, $\v$, on the left and on
  the optimal control, $\a$, on the right for $t$ fixed, at the start (blue),
  mid-point (green) and end-point (red), for each of the four regimes. The desired spike time is set to $\T=1.5$, the energy penalty, $\eps
  = .001$. The control bounds are $\a \in [-2,2]$. Note that the green and blue
  curves in E) are lying on top of each other for all $x$.
  \\
  A,E) Supra-Threshold-High-Noise B,F) Supra-Threshold-Low-Noise 
  C,G) Sub-Threshold-High-Noise D,H) Sub-Threshold-Low-Noise
  }   
\label{fig:HJB_4regimes_value_control_cuts}
\end{center}
\end{figure}  
% THE SURFS:    
\begin{figure}[h!]  
\begin{center}
\includegraphics[width=0.99\textwidth]{Figs/HJB/Regimes_valuesurf.pdf}
\caption{The numerical solution for the value function $\v(x,t)$ HJB PDE,
\cref{eq:OC_LS_HJB_full}, for the four different parameter regimes. Recall that
the solution is solved backwards in time.
The desired spike time is set to $\T=1.5$.  
The control bounds are $\a \in [-2,2]$. 
The contours for fixed $t$ plotted in thick correspond to the graphs in \cref{fig:HJB_4regimes_value_control_cuts}.
\\   
A) Supra-Threshold-High-Noise B) Supra-Threshold-Low-Noise 
C) Sub-Threshold-High-Noise   D) Sub-Threshold-Low-Noise  }
\label{fig:HJB_4regimes_value_surf}
\end{center} 
\end{figure}

Let us discuss the most salient features in
\cref{fig:HJB_4regimes_value_surf,fig:HJB_4regimes_value_control_cuts}. Recall
that lower values for the value function, $\v(x,t)$, are preferred, since we are
minimizing. We see the same general shape in all four space-time surface plots
of the value function in \cref{fig:HJB_4regimes_value_surf}. At the end of the
interval, $\v(x,\T)$ is monotonically decreasing in $x$, which is to be expected
given its terminal condition in \cref{eq:valuefun_TCs}. That is the lower the
value of $X_\T$ the more on average will we have to wait for the spike to occur.
% This same basic shape represents the idea that we would like to be close to
% the threshold, $x = 1$, near the end of the interval when $t \approx \T$, but
% we would like to be far from threshold, for $t \ll \T$. Indeed, at the very
% end of the interval, when $t = \T$, the value function, $v(x,\T)$, is
% monotonically decreasing in $x$, because the lower the value of $X_\T$ the
% more time on average will it take us to spike and the more delay will we have
% from our desired spike time.
As we go back in time $\v$ inverts, with a clear peak near the upper, threshold
boundary. That is, for intermediate values of $t$ we have to consider the risk
of spiking too early, represented by the high value of $\v$ near the upper
boundary {\sl and} the risk of spiking too late, represented by the high values
of $\v$ at the lower boundary. As we progress even further back in time to the
beginning of the interval, the peak near the threshold rises further, since we are spiking
{\sl earlier}, while the peak at the lower end flattens, since now there is
enough time to reach threshold despite starting far from it. The full surface
plots for $\v(x,t)$ in \cref{fig:HJB_4regimes_value_surf} can be thought of as
the interpolation between these three basic phases.

% Note, that exact shape is generated by Since we have put a very small weight
% on the energy penalty, $\e = .001$, and we have a significantly high bound on
% the maximum / minimum of the applied control, the value function becomes quite
% flat for $x \ll 1$ and $t \ll \T$, since it is likely that we can apply enough
% control to avoid an early spike and still drive the system close to threshold
% at the desired time, $t = \T$
Now focus on the controls, on the right in
\cref{fig:HJB_4regimes_value_control_cuts}. Naturally, at the end of the
interval, the control takes its maximum value, $\a(x,\T) = \amax$ for all $x$,
i.e.\ it gives the maximum available push for the neuron to spike. This is
implied \cref{eq:valuefun_TCs}. As we go back in time, however, the control $\a$
decreases for $x \ll \xth$, and it becomes negative, i.e.\ inhibitory for $x
\lesssim \xth$. That is intuitively consistent with the problem objective - for
$t<\T$, we want to bring $X_t$ closer to the threshold, but not too close, to
avoid early spiking!

Now let us consider the differences between the individual regimes, i.e.\ the
effect of changing the bias and the noise intensity, $\m,\b$. All things being
equal, the effect of increasing the noise intensity, $\b$, is to lift the value
function, i.e.\ to make our objective worse, at the beginning of the interval
and to decrease it at the end. This is to be expected since we are attempting
to minimize the variance in the spike time and without noise there would be no
variance at all. However, at the end of the interval, noise only helps with the
spiking, since now we are only interested in spiking as soon as possible, and a
higher noise intensity, will tend to decrease the spike time on average.
Increasing $\b$ also has the effect of increasing the size of the boundary layer
near $\xth$, where the value function rises steeply. I.e.\ for small $\b$ that
layer is small since the risk of spiking early is only significant close to
the threshold. For larger $\b$ this layer increases, see panels A,C, with a thin
layer, vs.\ B,D, with a thicker layer, in
\cref{fig:HJB_4regimes_value_control_cuts}. Similarly, increasing the bias,
$\mu$, tends to decrease the value function, especially at the end since that
has the unequivocal effect of preventing late spikes.
\clearpage

\section{Open-loop Stochastic Control}
When the value of $X_t$ is unobservable, the best
one can do is to use the transition density to perform the optimization. Since the
transition density follows a (deterministic) PDE, we apply a Maximum
Principle for PDEs as a method of obtaining the optimal control,\cite{Borzi2012}.

\subsection{Fokker-Planck equation for the state density evolution}
We write the transition density of $X$, conditional on no
spikes having occurred as:
\begin{align*}
f(x,t) \intd{x} &= \Prob[X_t\in \intd{x} \,\big|\, X_0 = 0,\, X_{s} < \xth \quad
\forall s < t]
\end{align*}
Then $f$ satisfies a Fokker-Planck equation with absorbing boundaries,
see Ch.\ 7 in \cite{Jacobs},
\begin{equation}
\begin{gathered}
\di_t f(x,t) =
				\frac{\b^2 }{2}\cdot \di^2_x f -
				\di_x \Big[ \left( \m + \a(t)- \frac{x}{\tc}\right)  \cdot f \Big]
\\
\\
\begin{array}{lll}
	&\textrm{st.\ }&
	\left\{ \begin{array}{lcl}
	 f(x,0) &=& \delta(x) \quad \textrm{delta function}
	\\
	(\m + \a(t)- \tfrac x \tc)f - \di_x \tfrac {\b^2}2 f] \big|_{x=\xmin} &\equiv&
	0 \quad \textrm{reflecting BCs at some } \xmin
	\\
	f(x,t) |_{x=\xth} &\equiv& 0 \quad \textrm{absorbing BCs at } \xth
\end{array} \right.
\end{array}
\label{eq:FP_pde_OU_PDF}
\end{gathered}
\end{equation}
In theory, $\xmin = -\infty$, but in the numerics below we will need to
truncate it to some finite value, exactly as in the HJB equation,
\cref{eq:OC_LS_HJB_full}.
Note that we write $\a(t)$ here instead of $\a(x,t)$ since we cannot use the
value of $X_t$ in choosing the optimal control.

The $f$ dynamics can also be written as
$$
\di_t f(x,t) = - \di_x \phi(x,t)
$$
for the probability flow
$$
\phi(x,t) = (\m + \a(t) - \tfrac x \tc)f - \di_x [\tfrac {\b^2}2 f]
$$
Then the lower BC is
$$
\phi(x,t) |_{x=\xmin} \equiv 0
$$

We will also need a short hand notation for the differential operator on the
right side of \cref{eq:FP_pde_OU_PDF}. Let
$$ \L_\a[\cdot] := \di^2_x \Big[\frac{\b^2 }{2} \cdot\Big] -
 \di_x \Big[ (\m + \a(t)- \frac{x}{\tc}) \cdot \Big] $$where
 $[\cdot]$ indicates the argument of the operator $\L_\a$. We will usually
 omit the subscript $\a$, but have written it now to emphasize that the
 differential operator is parametrized by the control, $\a$.

\subsection{Restating the objective in terms of the transition density}
Let us recall our objective, \cref{eq:OC_LS_variance}: 
$$
J[\a(\cdot)] = \Exp\left[
(\ts - \T \big)^2
+
\e \int_0^\ts  \a^2(s) \intd{s}
\right].
$$
Rewriting this in terms of the transition density, $f$, reads:
\begin{align}
J[\a(\cdot)] =&
\int_\xmin^\xth \Ttwo(x) \cdot f(x,\T) \intd{x}
\notag
\\
&+ \int_0^\T \phi(\xth, t) (t-\T)^2 \intd{t}
\label{eq:OC_LS_variance_density}
\\
&+  \e \int_0^\ts  \a^2(t)  \int_\xmin^\xth f(x,t) \intd{x} \intd{t}
\notag
\end{align}
Let us explain in more detail each term on the RHS in
\cref{eq:OC_LS_variance_density}.

The first term $$ \int_\xmin^\xth \Ttwo(x) \cdot f(x,\T)\intd{x} $$ counts the cost of trajectories which spike too late. This cost is the expected
squared time-to-hit starting at $x$, with  $\a = \amax$, weighted by the
probability of $X_\T = x$, which is just $f(x,\T)$. Recall that $\Ttwo$ is
defined in \cref{eq:Trem_squared}.

The second term $$ \int_0^\T \phi(\xth,t) (t-\T)^2 \intd{t} $$ counts the cost
of trajectories which spike too early, that is the squared difference between
some realized spike time, $\ts = t$, and desired spike time $\T$, weighted by
the probability of a spike at $t$ which is just the outflow of probability,
$\phi(\xth,t)$. Recall that we assume that $\xth = 1$ throughout. Note further
that due to the homogeneous Dirichlet BC at $\xth$, $f(\xth) = 0$, the outflow
is simply: $$ \phi(\xth, t) = -\frac{\b^2 }{2} \di_x f(\xth, t).$$

Finally, the third term
$$
\e \int_0^\ts  \a^2(t)  \left[  \int_\xmin^\xth f(x,t) \intd{x} \right] \intd{t}
$$
is the energy cost: the inner integral, $\int_\xmin^\xth f(x,t) \intd{x}$
takes into account that we incur an energy cost only for those trajectories
which have not yet spiked.

With that our optimal control $\a^*(\cdot)$ will naturally be found via:
$$
\a^*(\cdot) = \argmin_{\a(\cdot)} J[\a(\cdot)]
$$

\subsection{Optimizing using a Maximum Principle}
\label{sec:PDE_max_principle_for_pdf}
By now our stochastic optimal control problem modelled by SDEs
has become a deterministic optimal control problem modelled by PDEs. Our control
$\a$ influences the evolution density $f$ and via $f$, the integrals in the
objective, $J$.

The Maximum Principle for PDEs, which is an extension to the famous Pontryagin
Maximum Principle from finite dimensional systems, introduces an adjoint
variable, $p$, which solves a PDE related to the PDE satisfied by the density
$f$ and then calculates the optimal control, $\a(\cdot)$, as a functional of $f$
and $p$.

In short, the equation for the adjoint function, $p$, is
\begin{equation}
\begin{gathered}
\begin{aligned}
\di_t p =& - \Lstar[p]
\\
		=&
			- \Big[ \frac{\b^2 }{2}\cdot \di^2_x p +
			(\m + \a(t)- \frac{x}{\tc})  \cdot \di_x p \Big]
\end{aligned}
\\
\st
\begin{cases}
	p(x,\T) &= \Ttwo(x)
	\\
	\di_x p  \big|_{x=\xmin} &\equiv 0
	\\
	p \big|_{x=\xth} &\equiv (s-T)^2
\end{cases}
\label{eq:adjoint_pde_OU}
\end{gathered}
\end{equation}

and then $\a$ can be found via
\begin{equation}
\Big\{
 \tilde{\e}  2 \a(t)
+ p f \Big|_\xmin
- \int _\xmin^\xth p \cdot \di_x f \intd{x}
\Big\} = 0
\quad \forall t \in [0,T]
\label{eq:J_necessary_condition}
\end{equation}

More practically, the quantity,
$$\di_\a J =  \Big\{
 \tilde{\e}  2 \a(t)
+ p(x,t) f(x,t) \Big|_\xmin
- \int _\xmin^\xth p \cdot \di_x f(x,t) \intd{x}
\Big\}
$$
gives the direction of increase of $J$ at $\a(t)$
and we can use it as a gradient in a descent algorithm, given some initial
guess, $\a_0(t)$, for the control.

Finally, we are ready to calculate the open-loop stochastic optimal control for
the four parameter regimes. The calculated optimal controls for each regime are
in \cref{fig:FBK_Regimes_cs}. Recall that the optimal control, $\a^*(t)$, is
open-loop and thus it is only a function of time.
% \usepackage{graphics} is needed for \includegraphics 
\begin{figure}[htp]
\begin{center}
  \includegraphics[width=.99\textwidth]{Figs/FP_Adjoint/Regimes_cs.pdf}  
  \caption[labelInTOC]{The deterministic optimal controls for each parameter
  regime as functions of time, $t \in [0, \T]$.
  The desired spike time is set to $\T=1.5$, the energy penalty, $\eps
  = .001$ and the bounds are $\a \in [-2,2]$.
  A) Supra-Threshold-High-Noise B) Supra-Threshold-Low-Noise 
C) Sub-Threshold-High-Noise   D) Sub-Threshold-Low-Noise  }
  \label{fig:FBK_Regimes_cs}   
\end{center}
\end{figure}

Let us discuss the most salient features of the controls calculated in
\cref{fig:FBK_Regimes_cs}. In general, the strategy is to inhibit the neuron at
the beginning of the interval, $\a(t) = \amin$ for $t \ll \T$ and to excite it
near the end, $\a(t) = \amax$ for $t \approx \T$. The smooth portion that connects these
two segments in the middle of the interval is due to the energy penalty, $\e \int
\a^2(s) \intd{s}$. The behaviour of the control in different regimes
is relatively simple to explain and we see what we would expect. In a
sense the bias current $\mu$ can be absorbed by the control and so increasing
$\mu$ amounts to decreasing $\a$ modulo its bounds. That is exactly the
difference between plots A) vs.\ C) and B) vs.\ D) in \cref{fig:FBK_Regimes_cs},
where the reduction of $\mu$ in the lower plots, C) and D), results in an
increase in $\a(t)$ compared to A) and B), respectively. The effect of the noise
intensity, $\b$, is more subtle and it is not the same in the Supra vs.\ the
Sub-Threshold regimes A,B) vs.\ C,D). In the Supra-Threshold regime,  panels
A,B), reducing $\b$, reduces the need for excitatory  control towards the end,
since the bias alone should put the neuron over the threshold. Thus the main
part of the control with low noise in the Supra-Threshold regime is to stop the
neuron from spiking too early. In the Sub-Threshold regime, panels C,D), the
situation is somewhat reversed - reducing $\b$, obviates the need to apply an
inhibitory control in the early part of the interval and also prompts the
controller to apply excitatory input earlier, since it is the controller which
is now the main drive for a spike.

% An example for . The results are in
% \cref{fig:FP_adjoint_objective_control_convergence}.
% \begin{figure}[h]
% \begin{center}
% \subfloat[$\a_k(t)$ ]
% {
% \includegraphics[width=0.48\textwidth]
% {Figs/FP_Adjoint/ExampleControlConvergence_control.png}
% }
% \subfloat[$J_k$ ]
% {
% \includegraphics[width=0.48\textwidth]
% {Figs/FP_Adjoint/ExampleControlConvergence_objective.png}
% }
% \caption[ ]{The result of an entire optimization iteration, on the left the
% iterates of the control, $\a_k(t)$, on the right the progress of the
% objective with each iteration. The final control, in purple, is inhibitory at
% the beginning of the time interval, $\a < 0$, unlike the the deterministic
% control solution.
% We see a significant
% reduction in the objective value, $J$, and thus an improvement, on the
% order of 50\%.}
% \label{fig:FP_adjoint_objective_control_convergence}
% \end{center}
% \end{figure}
% 
% \begin{table}[h]
% \centering
% \begin{tabular}{lc}
% Control Law & Squared Error \\
% \hline
% Deterministic &  0.647 \\
% Stochastic &  0.336\\
% Theory (Value function) & 0.285
% \end{tabular}
% \caption{Realized performance of the different control laws and the theoretical
% expected performance of the stochastic law (last row)}
% \label{tab:realized_avg_errors_det_vs_stoch}
% \end{table}
% 
% \begin{figure}[h]
% \begin{center}
% \subfloat[A]
% {
% \label{fig:controlled_traj_ex1}
% \includegraphics[width=0.33\textwidth]
% {Figs/ControlSimulator/example_controlled_trajectories_id1.png}
% }
% \subfloat[B]
% {
% \label{fig:controlled_traj_ex2}
% \includegraphics[width=0.33\textwidth]
% {Figs/ControlSimulator/example_controlled_trajectories_id4.png}
% }
% \subfloat[C]
% {
% \label{fig:controlled_traj_ex3}
% \includegraphics[width=0.33\textwidth]
% {Figs/ControlSimulator/example_controlled_trajectories_id3.png}
% }
% \caption[]{Examples for the controlled trajectories using both the deterministic
% and the stochastic control approaches. The red vertical line in the plots
% indicates the desired spike-time, $\T$. In A, B the performance of both
% control laws is essentially the same, but in C we see the advantage of the
% stochastic approach. $\tc, \b = [.75, 1.25]$.}
% \label{fig:control_trajectories_examples}
% \end{center}
% \end{figure}
% \begin{figure}[htp]
% \begin{center}
%   \includegraphics[width=.9\textwidth]{Figs/ControlSimulator/example_controlled_trajectories_hists.png}
%   \caption[labelInTOC]{Histogram of the spike timing error for the
%   deterministic (left) vs. the stochastic (right) control laws.}
%   \label{fig:error_histograms_det_vs_stoch}
% \end{center}
% \end{figure}

\clearpage
\section{Effect of energy penalty}
\label{sec:effect_of_eps}
So far we have assumed a low value for the energy penalty parameter,
$\eps$, in the objective defined in \cref{eq:OC_LS_variance_energy}. Our
approach has been that we are most concerned with the accurate spiking and that
penalizing energy is rather intended to regularize the control and avoid
excessive chattering of the control between its extreme values, than as a goal
in minimizing energy in its own right. Now, we take some time to explore the
effect of a higher $\eps$ on the optimal controls. Intuitively, a
higher value of $\eps$ will tend to bring the optimal control, $\a^*$, closer
to zero. We make a comparison for the four regimes for both the
closed-loop control and the open-loop control in
\cref{fig:HJB_4regimes_control_different_eps,fig:FBK_Regimes_cs_different_es}.
Indeed, for both the closed-loop and open-loop optimal controls, increasing
$\eps$ tends to reduce the absolute value of $\a^*(t)$. In particular, for the
closed-loop control it tends to broaden the area of transition for
decreasing $x$ when the control swings from its minimal, i.e.\ inhibitory value,
to its maximal, i.e.\ excitatory value. Similarly for the open-loop control,
instead of banging from its minimal bound at the beginning of the interval to
its maximal bound at the end, the optimal control for $\e = .1$ tends to have a
more mild transition from a slightly inhibitory to slightly excitatory
values.

Note however that increasing $\eps$ has an important effect on the validity of
the approximation, used to form the terminal conditions for
the value function and the adjoint variable, $\v,p$. We assumed that applying
$\amax$ is optimal for $t>\T$. This is indeed true in the
limit $\eps \ra 0$. However, even for a finite value of $\eps$, this may still
be the optimal thing to do. Indeed, in the original calculations,
where $\eps=.001$, see the panels on the left in
\cref{fig:HJB_4regimes_control_different_eps,fig:FBK_Regimes_cs_different_es},
the so-obtained value function {\sl implied} that $\a(x,\T) = \amax.$. I.e.\ our
guess that $\a(x, t) = \amax$ for $t > \t$, is self-consistent with the
so-obtained value function. In a sense, this is like confirming an ansatz. For a higher
value of $\eps$, like $\eps = .1$, this is no longer the case. The red
curves on the right-side of \cref{fig:HJB_4regimes_control_different_eps} are no
longer pushed up at $\a = \amax$, and as such it is no longer valid to
assume that the value function can be obtained by assuming $\a(t>\T) = \amax$
and then just calculating the expected remaining time-to-spike.
The exact same conclusion can be drawn from the open-loop control. There the
optimal control always (almost) reaches its maximum, $\amax$ while choosing
$\eps = .001$, but no longer so for the higher $\eps = .1$.
In fact, as we mentioned after \cref{eq:valuefun_TCs}, the correct thing to do
for higher values of $\eps$, is to push the
calculation interval to some $t^+ > \T$, apply the same terminal condition at
$t^+$ instead of $\T$ and then solve backwards (either for the closed- or
open-loop control). One will be guaranteed to find some $t_f > \T$ that works
since eventually the quadratic term $(t-\T)^2$ will dominate the energy term in
the objective, which is linear in $t$. We do not explore this further here. 
% CLOSED_LOOP EPS: 
\begin{figure}[htp]
\begin{center}
  \includegraphics[width=.99\textwidth]{Figs/HJB/Regimes_eps_comparison.pdf}
  \caption[labelInTOC]{Effect of $\e$ on the closed-loop control. Snapshots of
  $\a(x,t)$ for $t$ fixed, at the start (blue), mid-point (green) and end-point
  (red), for each of the four regimes and different values of $\eps$.
  On the left, we use $\eps = .001$ and on the right $\eps = .1$. 
  The desired spike time is set to
  $\T=1.5$ and the bounds are $\a \in [-2,2]$.
  A,B) Supra-Threshold-High-Noise C,D) Supra-Threshold-Low-Noise 
  E,F) Sub-Threshold-High-Noise G,H) Sub-Threshold-Low-Noise}
\label{fig:HJB_4regimes_control_different_eps} 
\end{center}
\end{figure}
% OPEN_LOOP EPS:
\begin{figure}[htp]
\begin{center}
  \includegraphics[width=.99\textwidth]{Figs/FP_Adjoint/Regimes_eps_comparison.pdf}
  \caption[labelInTOC]{Effect of $\e$ on the open-loop control. 
  The deterministic time-trajectory of $\a(t)$ for each
  parameter regime as function of time, $t \in [0, \T]$.
  On the left, we use $\eps = .001$ and on the right $\eps = .1$.
  The desired spike time is set to $\T=1.5$ and the bounds are $\a \in [-2,2]$.
  A,B) Supra-Threshold-High-Noise C,D) Supra-Threshold-Low-Noise 
  E,F) Sub-Threshold-High-Noise G,H) Sub-Threshold-Low-Noise}
  \label{fig:FBK_Regimes_cs_different_es}
\end{center}
\end{figure}

\clearpage

\section{Test Comparison of Different Controls}
\label{sec:probabilistic_numerical_test}
Having obtained the optimal controllers, both closed- and open-loop, we evaluate
their performance with simulated realizations of the voltage process,
\cref{eq:X_evolution_uo}. In particular, while they both minimize the expected
squared deviation of the spike-time from some desired spike time, we analyze
the distribution of the squared deviation and the behaviour of the
controls for different parameter regimes.

In addition to the optimal controllers, we also show the behaviour of
another control law - perhaps the most naive one - which is obtained by ignoring
the noise and the energy penalty. That is we find a {\sl constant} value of $\a$
which satisfies the desired boundary conditions, $x(0) = .0, x(\T) = \xth$. This naive
controller will be called 'Deterministic', since it assumes deterministic
dynamics in $X_t$, i.e.\ $\b = 0$.

For the comparison, we sample $N$ realizations of the controlled system
and apply in turn each of the three controls. Naturally, we reuse the same
realization of the underlying stochastic process, $W_t$, for each of the
three different controls.

We set $\e = .001$ and $\amax = 2.0$. As such the energy cost is of secondary
importance and the paramount effect on the objective is the difference $(\ts -
\T)$, which we are trying to minimize.

The performance of the three controllers for each parameter regime is given in
\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch} and  
\cref{fig:error_histograms_det_vs_openloop_vs_stoch}. Naturally, the closed-loop
achieves a lower error than the open-loop, and the naive controller fares worst.
The difference in performance mostly depends on the strength of the noise. Thus,
for low noise, the performance of the stochastic controllers, be they open- or
closed-loop is not much superior to the naive deterministic controller. On the
contrary - in the high-noise regime, using a stochastic controller gives a much
lower error on average between the desired $\T$ and the realized $\ts$.

For illustration sake, we also visualize some trajectories from the
Sub-Threshold-High-Noise regime in \cref{fig:control_trajectories_examples}.

\begin{table}[h]
\begin{center}
\subfloat[Supra-Threshold High-Noise]{
\input{Figs/ControlSimulator/MeanSquaredErrors__SUPT_HN.txt}}\\
\subfloat[Supra-Threshold Low-Noise]{
   \input{Figs/ControlSimulator/MeanSquaredErrors__SUPT_ln.txt}}\\
 \subfloat[Sub-Threshold High-Noise]{
 \input{Figs/ControlSimulator/MeanSquaredErrors__subt_HN.txt}}\\
 \subfloat[Sub-Threshold Low-Noise]{
 \input{Figs/ControlSimulator/MeanSquaredErrors__subt_ln.txt}}\\
%  \begin{tabular}{lcc}
% Control Law &  Squared Error &  Squared Error \\
%  & Empirical & Theoretical\\% \hline
% Deterministic &  0.621 & - \\
% Open-Loop Stochastic & 0.356 & 0.382\\
% Feedback Stochastic &  0.288& 0.285\\
% \hline % \end{tabular} -
\caption{TODO(ALEX): THIS TABLE IS MOSTLY A SANITY CHECK TO SEE THAT THE CODE
WORKS - CONSIDER OMITTTING IN DRAFT.
Realized and theoretical performance of the different control laws. The
empirical performance is obtained using $N=10000$ sample paths. The theoretical
performance is found using the optimal value for $J$ for the open-loop
stochastic control and the value function $\v(x=0, t =0)$ for the closed-loop
stochastic control.}
\label{tab:realized_avg_errors_det_vs_openloop_vs_stoch}  
\end{center}
\end{table}

\begin{figure}[h]
\begin{center}
\subfloat[Supra-Threshold High-Noise]{
  \includegraphics[width=.99\textwidth]
  {Figs/ControlSimulator/Regimes_SUPT_HN_errors_hist.pdf}
  }\\
\subfloat[Supra-Threshold Low-Noise]{
  \includegraphics[width=.99\textwidth] 
  {Figs/ControlSimulator/Regimes_SUPT_ln_errors_hist.pdf}
}\\
\subfloat[Sub-Threshold High-Noise]{
  \includegraphics[width=.99\textwidth]
  {Figs/ControlSimulator/Regimes_subt_HN_errors_hist.pdf}
}\\   
\subfloat[Sub-Threshold Low-Noise]{
  \includegraphics[width=.99\textwidth]
  {Figs/ControlSimulator/Regimes_subt_ln_errors_hist.pdf}
}
  \caption[labelInTOC]{Histogram of the spike timing error for the
  deterministic (left) vs. open-loop stochastic(centre) vs. closed-loop
  stochastic (right) control laws. This is for the same problem as in
  \cref{fig:control_trajectories_examples}. 
  We have used $N=10000$ sample paths
  to form the statistics}
  \label{fig:error_histograms_det_vs_openloop_vs_stoch}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\subfloat[]
{
\label{fig:controlled_traj_ex1}
\includegraphics[width=0.33\textwidth]
{Figs/ControlSimulator/SubTHighNoise_Traj5.pdf}
}
\subfloat[ ]
{
\label{fig:controlled_traj_ex2}
\includegraphics[width=0.33\textwidth] 
{Figs/ControlSimulator/SubTHighNoise_Traj3.pdf}
}
\subfloat[ ]
{
\label{fig:controlled_traj_ex3}
\includegraphics[width=0.33\textwidth]
{Figs/ControlSimulator/SubTHighNoise_Traj7.pdf}
}
\caption[]{Examples for the controlled trajectories using the
deterministic, open-loop stochastic and closed-loop stochastic control
approaches. The black vertical line in the plots indicates the desired spike-time, $\T$.
For the parameter values are $\m, \tc, \b = [.2, .5,  1.5]$ (Sub-Threshold
High-Noise regime). $\T = 1.5, \eps = .001$. The three panels from left to
right are three different realizations of the model dynamics. On the upper
plots, we show the voltage evolution, $X_t$, on the lower plots we show the
applied control, $\a(t)$. Note that $\a(t)$ obtained from the
deterministic and open-loop controls are the same for all three samples.}
\label{fig:control_trajectories_examples} 
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\subfloat[]
{
\label{fig:controlled_traj_ex1}
\includegraphics[width=0.33\textwidth]
{Figs/ControlSimulator/HighEps_Traj5.pdf}
}
\subfloat[ ]
{
\label{fig:controlled_traj_ex2}
\includegraphics[width=0.33\textwidth]
{Figs/ControlSimulator/HighEps_Traj6.pdf}
}
\subfloat[ ]
{
\label{fig:controlled_traj_ex3}
\includegraphics[width=0.33\textwidth]
{Figs/ControlSimulator/HighEps_Traj0.pdf}  
}
\caption[]{The effect of higher energy penalty, $\eps = .1$!
Examples for the controlled trajectories using the deterministic,
open-loop stochastic and closed-loop stochastic control approaches. The black vertical line in the plots indicates the desired
spike-time, $\T$. For the parameter values are $\m, \tc, \b = [.2, .5,  1.5]$ (Sub-Threshold
High-Noise regime). $\T = 1.5, \eps = .1$. The three panels from left to
right are three different realizations of the model dynamics. On the upper
plots, we show the voltage evolution, $X_t$, on the lower plots we show the
applied control, $\a(t)$. Note that $\a(t)$ obtained from the
deterministic and open-loop controls are the same for all three samples.}
\label{fig:control_trajectories_examples_high_epsilon} 
\end{center}
\end{figure}

\clearpage

\section{Control Robustness}
We have so far assumed that we know the parameters determining the dynamics in
\cref{eq:X_evolution_uo} exactly. That is we assume that the values for $\tc,
\b$ used to obtain the controls are the same as the ones that drive the
controlled dynamics. That is slightly naive, as in any realistic application
these will have to be estimated and the estimates will have some error
associated with them. 
Note that misestimating $\mu$ has the most immediate and clear effect on the
control policy as that amounts to a direct shift of the control and so for now,
let us assume that $\mu$ is estimated correctly.

Let us then discuss one of the regimes, the Sub-Threshold,
High-Noise regime, and see what is the impact of misspecifying the parameters.

We need to assume that the errors large but not too large\ldots

So say, $$\b_{est} = \b / 1.5$$
and 
$$\tc_{est} = \tc / 1.5$$

So we will have:
$$\mu, \tc_{est}, \b_{est} = \{.2, .5, 1.5 \}$$ (which is the control
calculated already for the Sub-Threshold,
High-Noise regime), while the actual simulated dynamics will be driven by
$$\mu, \tc, \b = \{.2, .75, 2.25 \}$$

We show the effect in
\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch_misspecified}.
\begin{table}[h]
\begin{center}
 \subfloat[Sub-Threshold High-Noise]{
 \input{Figs/ControlSimulator/Misspec_MeanSquaredErrors__subt_HN.txt}}\\
 \caption{The effect of misspecifying the parameters. On the left, we
 system parameters and the parameters used to obtain the control are the
 same, i.e.\ accurate, on the right, they are misspecified.}
\label{tab:realized_avg_errors_det_vs_openloop_vs_stoch_misspecified}
\end{center}   
\end{table}

\section{Trains Simulation }
We now turn to the ultimate goal of our analysis, the control of {\sl spike
trains}. In particular, we will generate $M=10$ realizations of $N=20$ spikes
each in attempt to meet a prescribed spike train. We will focus on two
regimes, the sub-threshold regime with either low or high noise. 
The results
for parameters from the Sub-Threshold, Low-Noise regime are shown in
results in \cref{fig:targettrain_cl_lownoise,fig:targettrain_ol_lownoise}.
While, the results
for parameters from the Sub-Threshold, High-Noise regime are shown in
results in \cref{fig:targettrain_cl_highnoise,fig:targettrain_ol_highnoise}.
%\usepackage{graphics} is needed for \includegraphics
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%% CRIT REGIME LOW-NOISe:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htp]     
\begin{center}  
  \includegraphics[width=.99\textwidth]{Figs/TrainController/CRITLN_cl_trains_sim_10.pdf}
  \caption[ ]{How good the {\sl closed-loop} controller is at hitting a target
  spike train. Uses the parameters from the CRITICAL,Low-Noise regime. The
  target spike train has ISIs that are exponentially distributed with a
  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
  control strategy}
  \label{fig:targettrain_cl_critlownoise}  
\end{center}
\end{figure}    
\begin{figure}[htp]     
\begin{center}  
  \includegraphics[width=.99\textwidth]{Figs/TrainController/CRITLN_ol_trains_sim_10.pdf}
  \caption[ ]{How good the {\sl open-loop} controller is at hitting a target
  spike train. Uses the parameters from the CRITICAL,Low-Noise regime. The
  target spike train has ISIs that are exponentially distributed with a
  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
  control strategy} 
  \label{fig:targettrain_ol_critlownoise}    
\end{center} 
\end{figure}   
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%% CRIT REGIME HIGH-NOISe:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htp]     
\begin{center}  
  \includegraphics[width=.99\textwidth]{Figs/TrainController/CRITHN_cl_trains_sim_10.pdf}
  \caption[ ]{How good the {\sl closed-loop} controller is at hitting a target
  spike train. Uses the parameters from the CRITICAL, High-Noise regime. The
  target spike train has ISIs that are exponentially distributed with a
  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
  control strategy}
  \label{fig:targettrain_cl_critlownoise}  
\end{center}
\end{figure}   
\begin{figure}[htp]     
\begin{center}  
  \includegraphics[width=.99\textwidth]{Figs/TrainController/CRITHN_ol_trains_sim_10.pdf}
  \caption[ ]{How good the {\sl open-loop} controller is at hitting a target
  spike train. Uses the parameters from the CRITICAL, High-Noise regime. The
  target spike train has ISIs that are exponentially distributed with a
  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
  control strategy}
  \label{fig:targettrain_ol_critlownoise}   
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%% CRIT REGIME LOW-NOISe (AHmadian):
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htp]     
\begin{center}  
  \includegraphics[width=.99\textwidth]{Figs/TrainController/CRITLN_Ahmadian_cl_trains_sim_10.pdf}
  \caption[ ]{AHMADIAN COMPARES: How good the {\sl closed-loop} controller is at
  hitting a target spike train. Uses the parameters from the CRITICAL, High-Noise regime. The
  target spike train has ISIs that are exponentially distributed with a
  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
  control strategy}
  \label{fig:targettrain_cl_critlownoise}  
\end{center}
\end{figure}   
\begin{figure}[htp]     
\begin{center}  
  \includegraphics[width=.99\textwidth]{Figs/TrainController/CRITLN_Ahmadian_ol_trains_sim_10.pdf}
  \caption[ ]{AHMADIAN COMPARES: How good the {\sl open-loop} controller is at
  hitting a target spike train. Uses the parameters from the CRITICAL, High-Noise regime. The
  target spike train has ISIs that are exponentially distributed with a
  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
  control strategy}
  \label{fig:targettrain_ol_critlownoise}   
\end{center}
\end{figure}  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%% FIGS:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htp]     
\begin{center}  
\subfloat[DO NOT SKIP MISSED]{
  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTLN_cl_trains_sim_10.pdf}
  }\\
 \subfloat[SKIP MISSED]{
  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTLN_cl_skip_trains_sim_10.pdf}
  }
  \caption[ ]{How good the {\sl closed-loop} controller is at hitting a target
  spike train. Uses the parameters from the Sub-Threshold,Low-Noise regime. The
  target spike train has ISIs that are exponentially distributed with a
  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
  control strategy}
  \label{fig:targettrain_cl_lownoise}  
\end{center}
\end{figure}   
\begin{figure}[htp]  
\begin{center}
\subfloat[DO NOT SKIP MISSED]{
  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTLN_ol_trains_sim_10.pdf}
  }\\
 \subfloat[SKIP MISSED]{
 \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTLN_ol_skip_trains_sim_10.pdf}
 }     
  \caption[ ]{How good the {\sl open-loop} controller is at hitting a target
  spike train. Uses the parameters from the Sub-Threshold,Low-Noise regime. The
  target spike train has ISIs that are exponentially distributed with a
  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the   
  control strategy}
  \label{fig:targettrain_ol_lownoise}
\end{center}
\end{figure}
\begin{figure}[htp]  
\begin{center}    \subfloat[DO NOT SKIP MISSED]{ 
  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTHN_cl_trains_sim_10.pdf}  }\\
 \subfloat[SKIP MISSED]{
 \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTHN_cl_skip_trains_sim_10.pdf}
 }  
  \caption[ ]{How good the {\sl closed-loop} controller is at hitting a target
  spike train. Uses the parameters from the Sub-Threshold,High-Noise regime. The
  target spike train has ISIs that are exponentially distributed with a 
  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
  control strategy}
  \label{fig:targettrain_cl_highnoise}
\end{center}
\end{figure}  
\begin{figure}[htp]  
\begin{center}    \subfloat[DO NOT SKIP MISSED]{ 
  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTHN_ol_trains_sim_10.pdf}  }\\
 \subfloat[SKIP MISSED]{
 \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTHN_ol_skip_trains_sim_10.pdf}
 }  
  \caption[ ]{How good the {\sl open-loop} controller is at hitting a target
  spike train. Uses the parameters from the Sub-Threshold,High-Noise regime. The
  target spike train has ISIs that are exponentially distributed with a
  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
  control strategy}
  \label{fig:targettrain_ol_highnoise}
\end{center}
\end{figure}    
\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%% SKIP MISSED FIGS:
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{figure}[htp]     
% \begin{center}  
%   \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTLN_cl_skip_trains_sim_10.pdf}
%   \caption[ ]{SKIPPING MISSED TARGET. How good the {\sl closed-loop} controller is at hitting a target
%   spike train. Uses the parameters from the Sub-Threshold,Low-Noise regime. The
%   target spike train has ISIs that are exponentially distributed with a
%   mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
%   control strategy}
%   \label{fig:targettrain_cl_lownoise}  
% \end{center}
% \end{figure}   
% \begin{figure}[htp]  
% \begin{center}     
%   \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTLN_ol_skip_trains_sim_10.pdf}
%   \caption[ ]{SKIPPING MISSED TARGET. How good the {\sl open-loop} controller is at hitting a target
%   spike train. Uses the parameters from the Sub-Threshold,Low-Noise regime. The
%   target spike train has ISIs that are exponentially distributed with a
%   mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the   
%   control strategy}
%   \label{fig:targettrain_ol_lownoise}
% \end{center}
% \end{figure}
% \begin{figure}[htp]  
% \begin{center}     
%   \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTHN_cl_skip_trains_sim_10.pdf}
%   \caption[ ]{SKIPPING MISSED TARGET. How good the {\sl closed-loop} controller is at hitting a target
%   spike train. Uses the parameters from the Sub-Threshold,High-Noise regime. The
%   target spike train has ISIs that are exponentially distributed with a
%   mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
%   control strategy}
%   \label{fig:targettrain_cl_highnoise}
% \end{center}
% \end{figure}  
% \begin{figure}[htp]  
% \begin{center}     
%   \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTHN_ol_skip_trains_sim_10.pdf}
%   \caption[ ]{SKIPPING MISSED TARGET. How good the {\sl open-loop} controller is at hitting a target
%   spike train. Uses the parameters from the Sub-Threshold,High-Noise regime. The
%   target spike train has ISIs that are exponentially distributed with a
%   mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
%   control strategy}
%   \label{fig:targettrain_ol_highnoise}
% \end{center}
% \end{figure}  
\clearpage
\section{Conclusions / Discussion} 
We have analyzed and computed the optimal control of spikes in two scenarios -
one where the underlying voltage of the neuron is observable to the controller
(closed-loop control) and one where only the spikes are observable (open-loop
control).

ALEX: SUBSECTIONS ARE FOR DRAFT ONLY - THEY WILL BE FUSED
TOGETHER IN FINAL SUBMISSION
\subsection{different regimes}
Naturally, the ability to control the system is most clearly affected by the
level of the noise. In a sense, the noise acts like an adversary - in our
context it has no beneficial role, but to obstruct precise spiking.

The bias current has a more helpful role at least in our examples, where a
high value of the bias tends to help precise spiking. Of course, where it helps
or hinders depends on the value of the desired spike time - a combination of a
high positive bias and a 'distant' spike time, will tend to be difficult to
control as the system will naturally tend to spike earlier than desired.

\subsection{computational matters}
In both cases finding the optimal control is computationally intensive as we
numerically solve partial differential equations. If these algorithms were to be
practical in an online setting, some more work would have to be done in order to
ensure they can be computed very quickly (in the millisecond range) or that
they can be pre-computed.

\subsection{objective criterion}
As our criterion to optimize, we have chosen the squared deviation between the
realized time and some {\sl a priori} chosen time. We have done so as it
appeared to us the simplest objective, representing the idea. Other
objectives such as the absolute value of the difference, $|\ts - \T|$ or its
exponential, $\exp(|\ts - \T|)$, are directly applicable. A different approach
may be to require that $\Exp[\ts] = \T$. It is non-trivial to do this in the
dynamic programing framework, while it is fairly straightforward while working
with the forward equation in the open-loop context. A similarly comment may be
made about maximizing the
likelihood of $\ts = T$, which is the objective chosen in \cite{Ahmadian2011}.

\subsection{relation to other works}
Our work has strayed pretty close to the paradigm of \cite{Ahmadian2011} in
trying to make the neuron spike at a particular time. Other goals have been
formulated, most notably the idea of desynchronizing a population of neurons,
\cite{Nabi2011}, where it is not so important when the neurons spike as long as
they do not spike at the same time. We have also incorporated into our objective
the minimization of total energy used to achieve our goal, which as discussed in
\cite{Ahmadian2011} is sensible given the potential damage on the cell of
accumulating charge. Furthermore, we have assumed that our control is
constrained in magnitude, which is natural given physical limitations on
equipment and safety considerations. Another possible constraint on the control
is that it is charge-balanced, meaning that its time integral is zero, $\int u
\intd{t} = 0$. Such constraints are most easily posed in the context of
deterministic spiking models like the Phase-Response Curve, see e.g. Danzl et
al. \cite{Danzl2010}. It is possible to add them to the open-loop control, since
the control is still deterministic, but it is not-trivial. It is even less
trivial when one uses Dynamic Programing \{ALEX: IT IS WORTH THINKING ABOUT, AS
IT MAKES FOR A VERY COOL MATH PROBLEM, BUT WHAT IS THE PAY-OFF?\}. In
particular, insisting on charge-balance in our schemes will make it much more
difficult to apply the Terminal Conditions used in deriving both the closed- and
open-loop controls.

\subsection{novelty}
We have developed a stochastic optimal control scheme which is not limited
by the intensity of the noise in the system. Given this scope, the
numerical demands on our scheme are non-trivial, but our experience and
experiments show that these demands are potentially feasible. It
however remains to be seen whether they can be successfully applied in the lab
and beyond.

\subsection{{\sl Stochastic} effects} 
We show how wrong the control strategy can be if we assume that the noise is
negligible.

\appendix
\section{Calculating the Terminal Conditions for the HJB equation}
\label{sec:valuefun_TCs}
Here we explain how to calculate the first two moments of the time to-spike,
$\Exp \big[\trem^p \,\Big|\, X_\T = x, \a(t) = \amax \big].$, which is needed in setting the terminal conditions for both,
$\v(x,\T)$ and $p(x,\T)$, see \cref{eq:OC_LS_HJB_full} and \cref{eq:adjoint_pde_OU}.

Let 
\begin{equation} 
\begin{array}{lcl}
\Tone(x_0) &=& \Exp \Big[\trem \,\Big|\, X_\T = x, \a(t) = \amax
\Big]
\\
\Ttwo(x_0) &=&
\Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big],
\end{array}
\end{equation}
where as before, $\t$ is the remaining time-to-spike in
\cref{eq:X_evolution_uo}, at the desired spike-time $t=\T$, given $X_\T = x_\T$
and $\a(t) = \amax$.
Naturally, the $\Ti$'s depend on $x_\T$ and are affected by the value of the
various parameters, $\a,\b,\tc$.

Actually, we only  need $\Ttwo$ for the TCs in \cref{eq:OC_LS_HJB_full} and \cref{eq:adjoint_pde_OU}. 
but $\Tone$ is needed as an auxiliary.

Indeed, analytical expressions exist for $\Ti$, e.g.\ see \cite{Inoue1995}.
However in our experience, they are awkward to work with numerically, especially
as they involve infinite alternating sums which can easily overflow in numeric
calculations. Moreover, they are not particularly convenient if you need $\Ti
(x_0)$ for a wide range of $x_0$'s as we do here, since they calculate the value
independently for each $x$. Instead we will follow and slightly extend Jacobs\footnote{Jacobs only discusses how to calculate the first moment
$\Tone$, however it is quite clear from his discussion how to proceed
recursively to obtain any higher moment $\Ti$, by using previously calculated
lower moments. This is what we do here. Moreover the relation between mean exit
times and boundary value problems is discussed in most
introductory books on stochastic processes, see \cite{Oksendal2007,Evansa} etc.},
\cite{Jacobs}, in deriving and solving a differential equation for
$\Ti(x_0)$.

Basically we are calculating the mean-time-squared to exit an interval $[\xmin,
\xth]$. Theoretically, $\xmin = -\infty$, but for our purposes we will set it to
some finite value and impose reflecting boundaries there - this is a very good
approximation for $\xmin \ll 0$, as long as $\a = \amax >0$, which is what we
have.

Let $B(y,t|x) = \Prob[X_0 = y| X_t = x]$ for a fixed $x$. Then $B$ solves the
following (backward Kolmogorov) PDE
\begin{equation}
-\di_t B = (\a - \frac y\tc)\di_yB - \frac {\b^2} 2 \di_y^2 B
\label{eq:backward_Xdensity}
\end{equation}

Let $\G(t,y)$ be the survival function for $T$ for $X_0 = y$. 
$$\G(t,y) = \Prob[T>t | X_0 = y] = \Prob[X_{s\leq t} \in [\xmin,
\xth] | X_0 = y]$$.

Because the dynamics do not depend explicitly on time, we have:
$$
\G(t,y) = \int_\xmin^\xth B(y,t|x) \intd{x}
$$
and so $\G$, being a definite integral of $B$ satisfies the same PDE:
\begin{equation}
-\di_t \G = (\a - \frac y\tc) \di_y\G - \frac {\b^2} 2 \di_y^2 \G
\label{eq:backward_SDF}
\end{equation}

With the BCs, ICs:
\begin{equation}
\begin{cases}
\G(t,\xth) = .0 & \quad (T = t \implies \Prob[T>t] = 0)
\\
\di_x \G(t,\xmin) = .0  & \quad (\textrm{due to reflecting boundary for X})
\\
\G(0, y) = 1. & \quad (T \textrm{ is definitely} > 0 \textrm{ for } x_0 \in
(\xmin, \xth) )
\end{cases}
\end{equation}

If we solve the PDE for $\G$ over $t \in [0,\infty)$, then we can calculate
$\Ttwo(x_0)$ as:
\begin{eqnarray}
\Ttwo(x_0) &=& \int_0^\infty t^2 \cdot (-\di_t\G(t,x_0)) \intd{t}
\\
		   &=& \int_0^\infty 2t \cdot \G(t,x_0) \intd{t}
\end{eqnarray}
assuming it exists.

Since $\Ttwo$ is an integral over time, we can reduce the PDE in
\cref{eq:backward_SDF} for the distribution to a BVP for the moment(s) by
integrating over time. Here is how this is done: multiply both sides  of
\cref{eq:backward_SDF} by $2t$, and integrate with respect to time:
\begin{equation}
\int_0^\infty 2t (-\di_t \G) \intd{t}
= 
\int_0^\infty  2t \left[ (\a - \frac y\tc) \di_y\G - \frac {\b^2} 2
\di_y^2 \G\right]
\intd{t}
\end{equation}
The LHS turns out to be $2\Tone$, so 
\begin{equation}
2\Tone 
=(\a - \frac y\tc)   \di_y \Ttwo 
- \frac {\b^2} 2
\di_y^2 \Ttwo
\label{eq:BVP_Ttwo}
\end{equation}
Similarly we can derive an equation for $\Tone$, namely:
\begin{equation}
1
=(\a - \frac y\tc)   \di_y \Tone 
- \frac {\b^2} 2
\di_y^2 \Tone
\label{eq:BVP_Tone} 
\end{equation}
In both cases, the BCs will be:
\begin{equation}
\begin{cases}
\Ti(\xth) = .0 & \quad (\textrm{we are already spking})
\\
\di_x \Ti(\xmin) = .0  & \quad (\textrm{due to reflecting boundary for }X)
\end{cases}
\label{eq:BVP_Ti_BCs}
\end{equation}
Equations \ref{eq:BVP_Tone} and \ref{eq:BVP_Ttwo} can be reduced to first order
differential equations and solved analytically, at least up to some definite integral,
but the analytical solution is not particularly useful or necessary for
numerical calculations. Instead, we opt to calculate the solutions to
\ref{eq:BVP_Tone} and \ref{eq:BVP_Ttwo} directly via an ODE solver, starting
from $T_i(\xth) = .0$ and integrating down for $x  \in [\xmin, \xth]$.
That is how we have obtained the terminal conditions for both $\v(x,\T)$ and
$p(x,\T)$ in our calculations above.

\bibliographystyle{plain}
% \bibliographystyle{unsrt}
% \bibliographystyle{bmc_article}
% \section*{References}  \bibliographystyle{iopartnum}   
\bibliography{library,local}

\end{document}

### 2013-08-22 log.tex
Initial commit

--- 2013-08-22 CompExam.aviolov.tex
@@
 
+\usepackage{algpseudocode}
+\usepackage{algorithm}
+
+\usepackage{cancel}
 \usepackage{cleveref}
-\usepackage{cancel}
+% \crefname{algocf}{alg.}{algs.}
+% \Crefname{algocf}{Algorithm}{Algorithms}
 
@@
 
-\newtheorem{thm}{Theorem}[section]
-\newtheorem{lemma}{Theorem}[thm]
-% \theoremstyle{definition}
-\newtheorem{ex}{Example}[thm]
-\newtheorem{defn}{Definition}[thm]
-
 \begin{document}
 
-\title{Stochastic Optimal Control - with a view towards spike control in neurons
-\\
-{\small Advanced Comprehensive Exam in Mathematics}
-\\
-{\small Dept. of Math - University of Ottawa}
-}
+\title{ISI Control in stochastic LIF models}
 \author{Alexandre Iolov 
@@
 
-\abstract{We review main concepts from stochastic optimal control theory,
-including stochastic differential equations, dynamic programing and Pontryagin's
-Maximum Principle. We then discuss an optimal control problem in a simple
-mathematical model of a neuron where the feedback is minimal and we have a
-quasi-open-loop control which necessitates the use of a Maximum Principle for
-the system's transition density.}
+\abstract{Given a noisy and leaky integrate-and-fire neuronal model - we pose an
+optimal control problem - how do we ensure a certain spike sequence. We then ask
+whether and how can we do this online, in 'real-time'?}
 
@@
 
-\section{Overview}
-Stochastic Optimal Control in continuous time deals with finding
-the input which optimizes the performance of a continuous-time dynamical systems
-that contain a significant random component in their evolutions.
-
-Here we present a brief overview of the subject starting from the basic notions
-of Stochastic Differential Equations (SDEs), 
-and then introducing the main tools of Optimal Control, namely the
-Hamilton-Jacobi-Bellman equation of Dynamic Programming and the state-adjoint
-equations arising from Pontryagin's Maximum Principle. We will not state
-proofs, although we sometimes provide heuristic derivations of the main
-results.
-
-If I had to learn the material all over again, I would proceed my readings as
-follows: First I would read the online notes by Evans on both
-SDEs,\cite{Evansa}, and Optimal Control, \cite{Evansb}. I would then read
-Jacobs' book, \cite{Jacobs}, especially for the chapter on the Fokker-Planck
-equation (ch 7) and then Oksendal \cite{Oksendal2007} for a
-more thorough mathematical treatment of SDEs. Having familiarized myself with SDE's I would
-return to Optimal Control in the Deterministic Case, starting with \cite{Kirk2004} for the
-finite dimensional deterministic case (ODEs) and then \cite{Fleming1975} for the
-extension to the stochastic case. I would keep the following for reference \cite{Krylov2008},
-\cite{Hanson2007} and finally \cite{Fattorini1999} for the
-deterministic infinite-dimensional (PDEs) case.
-
-I have not used any particularly sophisticated numerical
-methods, but, for example, Hanson \cite{Hanson2007} discusses both the
-Euler-Maruyama scheme for simulating SDEs and the Crank-Nicholson Method for
-solving PDEs which is all I have used in my codes.
-
-All of the theoretical results presented in the sequel will be from
-the above references.
-
-\section{Stochastic Differential Equations}
-In view of our ultimate goals, we will restrict ourselves to  stochastic
-processes whose sample paths have continuous paths, i.e.\ to SDEs driven by
-Brownian motion.
-
-Since the reader is more likely well familiar with the material in this section,
-we will not provide proofs, but only state the results with a view towards
-establishing the notation for the sequel.
-
-We will assume that the reader is familiar with the following concepts:
-\begin{itemize} 
-  \item a probability space, $\{\O, \sAlg, P\}$ consisting of a
-  probability space, a sigma-algebra and a probability measure
-  \item a continuous-time stochastic process, $X_t$
-  \item a filtration $\Fil(t)$ and in particular the filtration generated by a
-  stochastic process.
-\end{itemize}
-
-\subsection{The Wiener Process and the Ito Integral}
-The Wiener Process is the fundamental building block of the stochastic calculus,
-it is often called Brownian Motion and we will denote it $\{W_t\}_{t\geq 0}$. It
-satisfies:
-\begin{defn}Wiener Process, $W_t$:
-\begin{enumerate}
-  \item $W_0 = 0$
-  \item $W_t - W_s = N(0, |t-s|)$ , i.e. normally distributed with mean 0,
-  variance $|t-s|$)
-  \item $\forall \{t_i\}_1^N, \quad \{W_{t_i} - W_{t_{i-1}} \}_2^N \sim$
-  independent, i.e $W_t$ has independent increments
-\end{enumerate}
-\end{defn}
-The fact that the finite incremental distributions suffice to specify a unique
-continuous-time stochastic process is known as Kolmogorov's extension theorem.
-\begin{ex}[from \cite{Evansa} \#21.] Inverted Wiener Process:
-\\
-Let $W_t$ be a 1-d Wiener Process and define:
-\def \Wbar {{\bar{W} }}
-\begin{equation*}
-\Wbar(t) := \begin{cases}
-t W(1/t) & t >0
-\\
-0 & t = 0
-\end{cases}
-\end{equation*}
-Show $\Wbar(t)- \Wbar(s) \sim N(0, t-s)$.
-
-\begin{proof}[Soln]
-\begin{align*}
-\Wbar(t) - \Wbar(s) =& t W(1/t) - s W(1/s), \, t>s
-\\
-=& (t-s)W(1/t) + s W(1/t) - sW(1/s) & \text{recall:} 1/s > 1/t
-\\
-=& (t-s)N(0, 1/t) - sN(0, 1/s -1/t) & \text{by indnt increments}
-\\
-=& N\left(0, \frac{(t-s)^2}t + \frac{s^2 }{ 1/s -1/t} \right) & \text{
-addition of indnt normal RVs.}
-\end{align*}
-With some algebra, the variance in the last expression reduces to $t-s$
-\end{proof}
-\end{ex}
-
-Now, we collect a few more relevant properties of $W_t$:
-\begin{enumerate}
-  \item The sample paths $W_{[0,\infty)}(\o)$
-are almost surely (a.s.) continuous and are in fact Holder continuous for any
-exponent $\g < 1/2$
-\item The sample paths $W_{[0,\infty)}(\o)$ are nowhere differentiable
-\item $W_t$ is a Markov process: $\Prob[W_t \in B | \s(W_{s' \leq s })] =
-\Prob[W_t \in B \,| \,W_s]$, for any Borel set, $B \subset \R$, where $\s(W_{s'
-\leq s })$ is the filtration generated by the $W_t$
-\end{enumerate}
-
-We now turn to defining stochastic integrals based on the Wiener Process.
-
-\begin{defn} Progressively Measurable Functions:
-
-Let $\Fil_t$ be the filtration generated by the Wiener Process.
-
-Let $X_t$ be a stochastic process which is $\Fil_t$-measurable $\forall t$ and
-which is jointly measurable in $(t,\o)$. We call such an $X_t$
-\emph{progressively measurable}
-\end{defn}
-
-\begin{defn} $\Ltwopm, \Lonepm$
-
-We define $\Ltwopm[0,T]$ as the space of all progressively measurable
-$X$ such that
-\begin{equation*}
-\Exp \left[ \int_{[0,T]} X_t^2 \intd{t} \right] < \infty
-\end{equation*}
-
-Similarly, we define $\Lonepm[0,T]$ as the space of all progressively measurable
-$X$ such that
-\begin{equation*}
-\Exp \left[ \int_{[0,T]} X_t \intd{t} \right] < \infty
-\end{equation*} 
-\end{defn}
-
-$\Ltwopm$ will be the class of functions for which the Ito integral is
-well-defined. 
-\begin{defn} Ito Integral:
-\label{defn:ito_integral}
-
-Let $P^n := {a = t^n_1 \ldots t^n_{m_n} = b}$ be a partition of the interval
-$[a,b] \subset [0, \infty)$. Let $|P_n| = \sup_i|t_i - t_{i-1}|$. 
-Take $|P_n| \rightarrow_n 0 $ and consider an $ X_t \in \Ltwopm[a,b]$,
-
-then
-\begin{equation}
-\int_{[a,b]} X_t \intd{W} := \lim_{n \rightarrow \infty}  
-\sum_{i=1}^{m_n} X_{t_i}\left( W(t_{i+1}) - W(t_{i})\right)
-\end{equation}
-
-\end{defn}
-
-To be precise, our definition is actually a theorem, and the real
-definition is one that uses step functions and passes to the limit.  Also we
-will write the limits of integration $\int_0^T \cdot  \intd{W}$ or $\int_{[0,T]}
-\cdot  \intd{W}$ interchangeably.
-% To justify def'n \ref{defn:ito_integral}, we recall:
-% \begin{lemma}[Quadratic Variation]
-% \begin{equation}
-% \lim_{n \rightarrow \infty} \sum_{i=1}^{m_n} \left( W(t_i) - W(t_{i-1})\right)^2
-% = b-a \quad \text{in }  L^2(\O)
-% \end{equation}
-% \end{lemma}
-% The above can be read, heuristically as $\lim \sum (\Delta W)^2 \rightarrow \sum
-% \Delta t$, which is the meaning behind the colloquial $\intd{W} =
-% \sqrt{dt}$.
-
-Again, we state without proof a few interesting properties of $\int X \intd{W}$:
-\begin{thm} Ito Integral Properties
-
-\begin{enumerate}
-  \item $\Exp[\int_0^T X \intd{W} ] = 0$ 
-  \item $\Exp[\left(\int_0^T X \intd{W}\right)^2 ] = \Exp[\int_0^T X^2
-  \intd{t}]$
-  \item $I(t) = \int_0^t X \intd{W} $ is a martingale 
-  \item $I(t) = \int_0^t X \intd{W} $ has continuous sample paths
-\end{enumerate}
-\end{thm}
-
-\subsection{Ito SDEs and Ito's Lemma}
-\begin{defn} an Ito SDE:
-
-We write
-\begin{equation}
-dX_t = Fdt + G dW
-\end{equation}
-on $0 \leq t \leq T$, if $X$ is a real-valued stochastic process satisfying:
-\begin{equation*}
-X(r) = X(s) + \int_{[s,r]} F \intd{t} + \int_{[s,r]} G \intd{W}
-\quad 0\leq s \leq r \leq T
-\end{equation*}
-for some $F \in \Lonepm[0,T]$ and $G \in \Ltwopm[0,T]$
-\end{defn}
-
-We are now ready to present the celebrated Ito Lemma which is the chain-rule of
-stochastic calculus:
-
-\begin{thm}[Ito Lemma] Suppose $X$ satisfies the stochastic differential $dX_t =
-Fdt + G dW$ as above and take $u(x,t) \in C^{2,1}[ \R \times [0,T]]$.
-
-Set $Y(t) = u(X,t)$
-\\
-then
-$$
-dY =  \left( \di_t u + \di_x u \cdot F + \di^2_x u \cdot \frac{G^2}2 \right)
-\intd{t} + \left(   \di_x u\cdot G  \right)\intd{W}
-$$
-\end{thm}
-
-% We now present two examples, one fundamental and one more involved:
-% \begin{ex}[pg 67 in \cite{Evansa}] The 'Stochastic' Exponential:
-% 
-% Let $Y = \exp(\l W - \l^2 t / 2)$. In the notation of the above Thm, we have:
-% $dX = dW, F=0, G=1$ and $u(x,t) = \exp(\l x - \l^2 t / 2)$.
-% 
-% then
-% \begin{align*}
-% dY =& 
-% \left( \di_t u + \di_x u \cdot F + \di^2_x u \cdot \frac{G^2}2 \right)\intd{t} +
-% \left(   \di_x u \cdot G  \right)\intd{W}
-% \\
-% =&
-% \left( -\frac{\l^2 t}{2}\exp(\l x - \l^2 t / 2) + \l^2\exp(\l x - \l^2 t / 2)  \right)\intd{t}
-% \\ 
-% &+ \left(   \l \exp(\l x - \l^2 t / 2) \right)\intd{W}  
-% \\
-% =&  \l \left( \exp(\l x - \l^2 t / 2) \right)\intd{W}  
-% \\
-% =& \l Y dW
-% \end{align*}
-% \end{ex}
-% 
-% \begin{ex}[pg. 67 in \cite{Evansa}] Hermite Polynomials:
-% 
-% Let $h_n(x,t) = \frac{(-t)^n}{n!} e^{x^2/2t} \di^n_x \left( e^{x^2/2t} 
-% \right)$ be the $n$th Hermite polynomial
-% \\
-% then
-% $$
-% \int_0^t h_n(W, t) \intd{W} = h_{n+1}(W(t), t)) 
-% $$
-% i.e. in Ito calculus, the Hermite polynomials are like the power monomials
-% $\frac{t^n}{n!}$ in the ordinary calculus.
-% \begin{proof} Since
-% $$
-% d_\l^n [\exp(- \frac{(x-\l t)^2}{2t} ] |_{\l=0} =
-%  (-t)^n d_x^n [ \exp(-\frac{x^2}{2t})] $$
-% We will have:
-% \begin{align*}
-% d_\l^n [\exp(\l x - \l^2 t / 2) ] |_{\l=0} &= (-t)^n  e^{x^2/2t} d_x^n [
-% \exp(-\frac{x^2}{2t})]
-% \\
-% &= n! h_n(x,t)
-% \end{align*}
-% Thus we can expand $\exp(\l x - \l^2 t / 2) $ in a Taylor series at $\l=0$ as:
-% $$
-% \exp(\l x - \l^2 t / 2) = \sum_n \l^n h_n(x,t)
-% $$
-% Now recall $Y$ from the previous example:
-% \begin{equation*}
-% Y(t) = \exp(\l W(t) - \frac{\l^2 t }2) = \sum_n \l^n h_n(W(t),t)
-% \end{equation*}
-% which satisfies:
-% \begin{equation*}
-% Y(t) = 1 + \l \int Y \intd{W}
-% \end{equation*}
-% Naturally we will plug the series expansion in the SDE to obtain:
-% \begin{align*}
-% \sum_{n=0}^{\infty} \l^n h_n(x,t) =&
-% 1 + \l \int_0^t  \sum_{n=0}^{\infty} \l^n
-% h_n(x,t)\intd{W}
-% \\
-% =& 1 + \sum_{n=1}^{\infty} \l^n\int_0^t 
-% h_{n-1}(x,t)\intd{W} \quad {\textrm{bump $n$ up by one}}
-% \end{align*}
-% Equating the coefficients for $\l^n$ on both sides leads to the result
-% \end{proof}
-% \end{ex}
-
-Finally, we are in a position to define an Ito SDE:
-
-\begin{defn}[Ito SDE] Let
-$F(x,t):\R\times [0,T]\ra \R,  G(x,t):\R\times [0,T]\ra \R,$ be given functions.
-We say that a stochastic process $X$ satisfies:
-\begin{equation}
-dX =F dt + G dW
-\end{equation} 
-over $[0,T]$ if
-\begin{enumerate}
-  \item $X$ is progressively measurable wrt. $\Fil_t$
-  \item $F(X,t) \in \Lonepm$
-  \item $G(X,t) \in \Ltwopm$ 
-  \item $X_t = X_0 + \int_0^t F(X_s, s) \intd{s} + \int_0^t G(X_s, s) \intd{W_s}$
-\end{enumerate}
-\end{defn}
-
-% \begin{ex} The solution to 
-% $$
-% \begin{cases}
-% dX = F(t)X dt + G(t) X dW
-% \\
-% X_0 = 1
-% \end{cases}
-% $$
-% is 
-% $$
-% X_t = \exp\left( \int_0^t F(s) - \frac{G^2(s)}{2} \intd{s} + \int_0^t G
-% \intd{W} \right) $$
-% \begin{proof}
-% Let $Y = \int_0^t f(s) - \frac{g^2(s)}{2} \intd{s} + \int_0^t g
-% \intd{W} $
-% and guess that $X = e^Y$
-% then, using Ito's lemma:
-% \begin{align*}
-% dX =& (\di_y [e^y] dY + \frac{1}{2}\di^2_y[e^y]  g^2 dt
-% \\ =&
-% (e^y  f dt - e^y g^2 / 2) dt + e^y g dW + \frac{1}{2} e^y \cdot  g^2 dt
-% \\ =& e^y \left( f dt +g dW \right)
-% \\ =& X \left( f dt +g dW \right)
-% \end{align*}
-% \end{proof}
-% \end{ex}
-
-% \begin{ex}[Brownian Bridge, pg 80 in \cite{Evansa}] The solution of the SDE
-% $$
-% \begin{cases}
-% dB = -\frac{B}{1-t} dt + dW
-% \\
-% B_0= 0
-% \end{cases}
-% $$
-% is
-% $$
-% B(t) = (1-t) \int_0^t \frac{1}{1-s} dW
-% $$
-% \begin{proof}
-% It is not clear how to apply Ito's lemma to this problem, so instead we just
-% directly calculate the differential:
-% \begin{align*}
-% B(t^+) - B(t) =& (1-t^+) \int_0^{t^+} \frac{1}{1-s} dW - (1-t) \int_0^t
-% \frac{1}{1-s} dW
-% \\ =&
-% (1-t^+) \int_t^{t^+}\frac{1}{1-s} dW  - (t^+-t) \int_0^t
-% \frac{1}{1-s} dW
-% \\ \approx&
-% \int_t^{t^+} dW - \Delta t \frac{B(t)}{1-t}  
-% \end{align*}
-% \end{proof}
-% \end{ex}
-% Actually we can say a bit more about the Brownian bridge, in particular we
-% will calculate $\Exp[B^2]$
-%  \begin{ex}[Evans .37]
-% \begin{align*}
-% \Exp[B^2] =& \Exp\left[\left( (1-t) \int_0^t \frac{1}{1-s} dW \right)^2 \right]
-% \\
-% =&(1-t)^2 \Exp\left[ \int_0^t\left(  \frac{1}{1-s}\right)^2 \intd{s} \right]
-% \\
-% =&(1-t)^2 [ \frac{1}{1-t} - 1 ]
-% \\
-% =&(1-t) - (1-t)^2
-% \end{align*}
-% 
-% Further note that  $\lim_{t\ra 1^-}\Exp[B^2] = 0$ and that $B$ is a martingale
-% and so its first moment is always equal to its initial value, $0$. So indeed the 2nd
-% moment is the variance and with the variance converging to zero, we can use
-% Chebyshev's inequality to conclude that $\lim_{t\ra 1^-}B(t) = 0 $(a.s). This
-% justifies the name of the Brownian bridge - it is a normally distributed RV.
-% which is clamped to equal 0 at both $t= \{0,1\}$.
-% \end{ex}
-
-As an example we will discuss the Ornstein-Uhlenbeck process, which is the
-basis for the models we will face later on.
-\begin{ex}[O-U Process] Let $X$ follow:
-\begin{equation*}
-dX = \th (\a - X_t) dt + \b dW
-\end{equation*}
-with an initial condition, $X_0$, which may be an arbitrary distribution
-independent of the Wiener Process,$W$.
-We can solve this as follows:
-\begin{align*}
-dX =& \th (\a - X_t) dt + \b dW
-\\
-dX + \th X_t) dt=& \th \a dt + \b dW
-\\
-e^{\th t} dX + e^{\th t}\th X_t) dt=& \th e^{\th t}\a dt + \b e^{\th t} dW
-\\
-Xe^{\th t} - X_0 = & \int \th e^{\th t}\a dt +  \int \b e^{\th t} dW
-\\
-X_t =& e^{-\th t} X_0 + \a(1-e^{-\th t}) +  \frac{\b e^{-\th t}}{\sqrt{2\th}}
-W(e^{2\th t}-1)
-\end{align*}
-which means that $X$ forgets its initial conditions exponentially fast and
-converges to a normal random variable with mean $\a$ and variance
-$\tfrac{\b^2}{2\th}$
-\end{ex}
-
-At the end of this sub-section, we state Ito's Lemma with multiple Wiener
-processes and then for a multidimensional state.
-
-\begin{thm}[Ito Lemma for $dW \in  \R^{m}$] Suppose $X$ satisfies the stochastic
-differential $dX_t = F dt + G dW$, where $ F \in \R, G \in \R^{1\times m}$ and
-$dW = \left(dW^{(i)}\right) \in  \R^{m}$ is a vector of independent Wiener
-Processes. Take $u(x,t) \in C^{2,1}[ \R \times [0,T]]$. Let $D :=\left[{\sum_k
-G_{k} G_{k} } \right] \in \R$
-
-Set $Y(t) = u(X,t)$
-\\
-then
-$$
-dY =  \left( \di_t u +  \di_{x} u \cdot F + 
-  \di^2_{x} u \cdot D \right)
-\intd{t} +
- \left(  \di_{x} u  \cdot  G \right) dW  
-$$
-\end{thm}
-
-\begin{thm}[Ito Lemma for $X \in R^n$]
-Suppose $X$ satisfies the stochastic differential $dX_t = F dt + G dW$, where $
-F \in \R^n, G \in \R^{n\times m}$ and $dW = \left(dW^{(i)}\right) \in  \R^{ 
-m}$ is a vector of independent Wiener Processes. Take $u(x,t) \in C^{2,1}[ \R^n
-\times [0,T]]$. Let $D_{ij} := \tfrac{1}{2} \left[{\sum_k G_{ik} G_{jk} }
-\right]_{ij} \in \R^{n\times n} $
-
-Set $Y(t) = u(X,t)$
-\\
-then
-$$
-dY =  \left( \di_t u + \sum_i \di_{x_i} u \cdot f_i + 
-\sum_{i,j} \di^2_{x_i x_j} u \cdot D_{ij} \right)
-\intd{t} +
- \left(  \sum_{ij} \di_{x_i} u  \cdot G_{ij} dW^{(j)} 
-\right)$$
-\end{thm}
-
-\subsection{Fokker-Planck and Kolmogorov's Backward Equations}
-The Fokker-Planck equation and Kolmogorov's Backward equation describe the
-forward (resp. backward)  evolution of the probability density of $X_t$
-
-For the rest of this section we will work in $ \R^n$, i.e. $X$ will be an $n-$
-dimensional vector that satisfies the stochastic differential 
-\begin{equation}
-dX_t = F(X,t) dt + G(X,t) dW,
-\label{eq:generic_Ito_SDE_Rn}
-\end{equation}
-$F \in \R^n, G \in \R^{n \times m}$,
-$dW = (dW^{(i)}) \in  \R^{  m}$ is an $m$-dimensional Brownian motion and we
-write $D_{ij} :=
-\tfrac{1}{2} 
-\left[{\sum_k G_{ik} G_{jk} } \right]_{ij} \in \R^{n\times n} $.
-
-% Let $$\F(x,t| y,s) = \Prob[X_t \preceq x| X_s = y] $$ be the forward transition
-% distribution and
-
-Let $$ \f(x,t| y,s) \intd{x} =  \Prob[X_t \in \intd{x} | X_s = y] $$ be the
-transition probability density. Then $\f$ satisfies:
-\begin{equation}
-\di_t \f= -\sum_i \di_{x_i} \left[ F_i(x,t) \f(x,t) \right] 
-+ 
-\sum_{i,j}  \di^2_{x_i x_j} \left[ D_{ij}(x,t) \f(x,t) \right]
-\label{eq:FP_pde}
-\end{equation}
-This is called the \emph{Fokker-Planck} or \emph{Forward Kolmogorov} equation.
-It can also be seen as a continuity euquation or a cosevation of probability
-equation. To this end define the probability current, $\p \in \R^n$ as:
-$$
-\phi_i =   F_i(x,t) \f(x,t) 
-+ 
-\sum_{j}  \di_{x_j} \left[ D_{ij}(x,t) \f(x,t)\right]
-$$
-then the Fokker-Planck equation, \cref{eq:FP_pde}, can be written as:
-$$
-\di_t \f= -\grad \cdot \phi
-$$
-which just says that the change in probability is the
-difference between the flow in and the flow out.
-
-Initial conditions for $\f$ are given by the distribution of $X_s$, if $s=0$
-then this is the initial distribution of $X$. Boundary conditions (BCs) depend
-on the domain of $X$ and what happens to $X$ once it hits its boundary. If the domain
-is all of $\R^n$ then we only insist that $\lim_{|x| \ra \infty} \f = 0$. If the
-domain has boundaries, then there are two common scenarios which we will also
-encounter in the sequel:
-\begin{enumerate}  
-  \item absorbing BCs
-  \item reflecting BCs
-\end{enumerate}
-At an absorbing boundary, the particle $X$ is removed and $\f=0$ there.
-In this situation, we will not have conservation of probability and the integral
-of $\f$ over $X$'s domain will monotonically decrease.
-
-At a reflecting boundary, the particle $X$ bounces back into its domain and we
-will have that the probability current $\p\cdot n = 0$ where $n$ is the outward
-normal at the reflecting boundary.
-
-We now consider $\f$ as functions of $y,s$, holding $x,t$ fixed, then
-\begin{equation}
--\di_s \f= \sum_i  \left[F_i(y,s) \cdot \di_{y_i}\f(|y,s) \right] 
-+ 
-\sum_{i,j}   \left[ D_{ij}(x,t) \di^2_{y_i y_j}\f(|y,s) \right]
-\label{eq:FP_backward_pde}
-\end{equation} 
-This is the \emph{Backward Kolmogorov} equation for $\f$. Note the minus sign
-in front $\di_s \f$. A mnemonic for the signs of the Backward vs. Forward
-equation is that as $t$ increases $\f$ diffuses and so $\di_t$ and $\di^2_{x}$
-have the same sign, but as $s$ increases $\f$ anti-diffuses and they have
-opposite signs.
-
-The differential operator on the right-hand side of
-\cref{eq:FP_backward_pde} occurs often in the study of SDEs and has its own
-name.
-\begin{defn}[Generator of an SDE] the Generator $A$ of $X$ is defined by:
-$$
-A[\psi(x)] = \lim_{t \searrow 0^+} \frac{\Exp[\psi(X_t)]  - \psi(x)}{t} ;  \quad X_0 =
-x \in \R^n
-$$
-\end{defn} 
-\begin{lemma} For an Ito SDE as in \cref{eq:generic_Ito_SDE_Rn} 
-$$
-A[\psi(x)] = \sum_i F_i(x,t) \cdot \di_{x_i} \psi + \sum_{i,j} D_{ij}(x,t)
-\di^2_{x_i x_j}\psi $$
-\end{lemma}
-
-\subsection{Stopping Times}
-\begin{defn}
-Let $\Fil(t)$ be some filtration, a random variable $\t$ is called a
-\emph{stopping time} if 
-$$
-\{\o : \t(\o) \leq t\} \in \Fil(t) \, \forall t  
-$$
-\end{defn}
-The colloquial way of describing stopping times is at any time we know
-whether $\t$ has occurred or not. For a counterexample, the time that a Wiener
-Process achieves its maximum over some interval is not a stopping time, since at any
-given time, we do not know if the maximum has occurred or not. The most common
-example of a stopping time is the first hitting-time, which is the first time
-$X_t$ leaves a set. 
-\begin{thm}[First-Hitting time] Let $E \subset \R^n, E \neq
-\phi$ be open or closed
-
-then $$\t := \inf\{ t \geq 0 | X_t \in E\}$$ is a stopping time.
-\end{thm}
-
-The reason stopping times are very useful is that all the facts so far quoted
-for Ito calculus using integrals $\int_0^T \intd{W}$ remain true if $T$ is
-replaced by a stopping time $\t$.
-
-Also it allows us to link SDE's and PDEs using the generator, $A$, of the
-diffusion: 
-\begin{thm}[Dynkin's formula] Given $\t$ a stopping time, $\Exp[\t] < \infty$.
-Then:
-$$
-\Exp[u(X_\t, \t)] =
-u(x, 0) + \Exp\left[\int_0^\t \di_t u + A[u] \intd{s}
-\right]
- $$
-\end{thm}
-This provides a link between PDEs and stochastic processes and allows us to go
-back and forth in that we can find probabilistic results by
-solving a PDE or we can approximate a PDE by simulating a stochastic process
-and averaging.
-
-\begin{ex}[\cite{Evansb} pg 99 - Expected hitting time to a boundary]
-\label{ex:mean_hitting_time}
- Let $\O
-\subset \R^n$ be a bounded open set with smooth boundary $\di \O$ then it is a
-basic fact from PDEs theory that
-\begin{equation}
-\begin{cases}
--\frac{1}{2} \grad^2 u = 1  & \text{over } \O
-\\
-u =  0 &\text{on } \di \O
-\end{cases}
-\end{equation}
-has a unique $C^\infty(\O)$ solution.
-
-Let $X = W_t + x$ for any $x \in \O$ and define
- $$\t_x := \text{first time } X \text{ hits } \di \O$$
-then the generator of $X$ is $A[\psi] = -\grad^2(\psi)/2$ and we will
-have:
-
-\begin{align*}
-\Exp[u(X_\t)] - \Exp[ u(x_0)] =& \Exp \left[ -\frac{1}{2}\int_0^\t \grad^2
-u\right]
-\\
-=& \Exp \left[ - \int_0^\t 1
-\right]
-\\
-=-&\Exp [\t]
-\end{align*}
-Finally, invoke $u$'s BCs, $u|_{\di E} = 0$, and $X$'s ICs, $X_0=0 = x$ to
-conclude that $$ u(x) = \Exp [\t]$$
-The solution to the PDE evaluated at $x$ is the expected exit time from
-$E$ for an $X_t$  starting at $x$.
-\end{ex}
-% Similarly we can calculate more moments of the hitting time.
-% \begin{ex}[Mean squared hitting time to a boundary] Let $E
-% \label{ex:mean_squared_hitting_time}
-%  \subset \R$ be a bounded open set and let $u$ solve:
-%  \begin{equation}
-%  \begin{cases}
-%  (\a - \tfrac{x}{\tc}) \di_x u + \frac{\b^2}{2} \di_x^2 u = -\frac t 2  &
-%  \text{over } E
-%  \\
-%  u =  0 &\text{on } \di E
-%  \end{cases}
-%  \end{equation}
-% 
-% Then $- t / 2 = A[u]$ for $A$, the generator, corresponding to $dX = (\a - x /
-% \tc) dt + \b dW$.
-% Define $$\t_x := \text{first time } X \text{ leaves }  E$$
-% then
-%  \begin{align*}
-%  \Exp[u(X_\t)] - \Exp[ u(X_0)] =& 
-%  \Exp \left[  \int_0^\t   (\a - \tfrac{x}{\tc}) \di_x u + \frac{\b^2}{2} \di_x^2
-%  u \right]
-%  \\
-%  =& \Exp \left[ - \int_0^\t -\frac t 2 
-%  \right]
-%  \\
-%  =-&\Exp [\t^2]
-%  \end{align*}
-% And we can conclude that 
-% $$ u(x) = \Exp [\t^2]$$
-% The solution to the PDE evaluated at $x$ is the expected squared exit time from
-% $E$ for an $X_t$  starting at $x$.
-% \end{ex}
-
-\subsection{Feynman-Kac formula}
-The Feynman-Kac formula is a generalization of the above idea. We state it in
-1-D. 
-\begin{thm}[Feynman-Kac Formula] Consider the (linear) PDE, with terminal
-conditions:
-\begin{equation}
-\begin{cases}
-\di_t u(x,t) + f \di_x u + \frac {g^2}2 \di_x^2 u - V u + q 
-&= 0
-\\
-u(x,\tf) &= \psi(x)
-\end{cases}
-\label{eq:PDE_feynman_kac}
-\end{equation}
-The dynamics can also be written:
-$$
-\di_t u(x,t) + A[u] - V u + q = 0
-$$
-for the generator $A = f \cdot \di_x + g^2/2 \cdot \di_x^2$ corresponding to the
-SDE: $$
-dX_t = f(X,t) dt + g(X,t) dW
-$$
-Then for $t< \tf$
-\begin{align*}
-u(x,t) =& \Exp\bigg[
-\int_t^\tf q(x,s)\cdot \exp \left(-\int_t^s V(X_\t) \intd{\t} \right)\intd{s} \\
-&\quad + \, \psi(X_t) \cdot \exp\left(-\int_t^\tf V(X_\t,\t) \intd{\t} \right)  
-\quad \Big| \,X_t = x \bigg]
-\end{align*}
-\begin{proof}[Proof]
-This can be seen as a more elaborate extension of example
-\ref{ex:mean_hitting_time}. We set: 
-$$
-Y_r = u(x,t)+ 
-\int_t^r q(x,s)\cdot\exp \left(-\int_t^s V(X_\t) \intd{\t} \right)\intd{s} +
-u(X_r,r)\cdot\exp\left(-\int_t^r V(X_\t,\t) \intd{\t} \right)   
-$$
-then 
-\begin{align*}
-dY =& q(x,r)\cdot\exp \left(-\int_t^r V(X_\t) \intd{\t} \right)
-\\
-& - u(X_r,r) \cdot V(X_r, r)  \exp\left(-\int_t^r V(X_\t,\t)
-\\
-& + du \cdot \exp\left(-\int_t^r V(X_\t,\t) \intd{\t} \right) 
-\intd{\t}\right)
-\end{align*}
-Now we recall Ito's lemma for $du(X,t)$, and replace it in the last line:
-\begin{align*}
-dY_r =& q(x,r)\cdot\exp \left(-\int_t^r V(X_\t) \intd{\t} \right) \intd{r}
-\\
-& - u(X_r,r) \cdot V(X_r, r)  \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t}\right)s
-\\
-& + \Big( \di_t u + A[u] \Big) \cdot \exp\left(-\int_t^r V(X_\t,\t) \intd{\t}
-\right) \intd{r}
-\\
-& + \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-\end{align*}
-Now we can group the terms in front of the exponential term on the first
-three lines to form:
-\begin{align*}
-dY_r =& \Big(\di_t u + A[u] - u(X_r,r) V(X_r, r) + q(X_r,r) \Big) \cdot\exp
-\left(-\int_t^r V(X_\t) \intd{\t} \right)
-\intd{r}
-\\
-& + \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-\end{align*}
-Since, $u$ satisfies the PDE, \cref{eq:PDE_feynman_kac}, the term on the first
-line is zero so $dY$ is just
-$$
-dY_r = \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-$$
-integrating we have: 
-$$
-Y_\tf - Y_t = \int^\tf_t \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot
-\exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-$$
-Now we need conditions such that the expectation of the stochastic integral is
-zero. For example, it suffices that $V$ is bounded, and often in applications
-$V$ is a positive rate.
-Then, we   take conditional expectations, conditional on $X_t=x$:
-\begin{equation*}
-\begin{gathered}
-\Exp\Big[Y_\tf | X_t = x\Big] - \Exp\Big[Y_t | X_t = x\Big]  
-= 
-\\
-\Exp\Bigg[\int^\tf_t \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot
-\exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W} \,\mid\, X_t = x\Bigg] 
-\\ = 0
-\end{gathered}
-\end{equation*}
-Unravelling the definition of $Y_t$, we get
-\begin{align*}
-\Exp\bigg[
-\int_t^\tf q(x,s)\cdot \exp \left(-\int_t^s V(X_\t) \intd{\t} \right)\intd{s} 
- \quad + &
-\\
- \, \psi(X_t) \cdot \exp\left(-\int_t^\tf V(X_\t,\t) \intd{\t} \right)  
-\, \Big| \,X_t = x \bigg] &=\, u(x,t)  
-\end{align*}
-
-\end{proof}
-A more detailed proof can be found in \cite{Oksendal2007} Thm. 8.2.1
-\end{thm}
-
-We will not use the Feynman-Kac formula in what follows, but
-we have included it here, because there are some very interesting developments in stochastic optimal
-control, which rely on solving a many-dimensional PDE using stochastic
-simulation, the link being made via the Feynman-Kac formula - see
-\cite{Kappen2005},\cite{Palmer2011}.
-
-% \subsection{Numerical Simulation of SDEs}
-
-\section{Deterministic Optimal Control in Finite Dimension}
-Optimal Control theory has three main components - a state, $x$, a control, $u$
-and an objective $J$ which is a functional of $x,u$ and which we try to either
-minimize or maximize. Here, we will be minimizing.
-
-We introduce the theory for a finite dimensional state, $x(t) \in \R^{n_x}$, but
-we will not address state constraints. Our control will also be
-finite-dimensional $u(t) \in \R^{n_u}$ and it will be allowed to take values in
-some closed domain, $\Udomain$, of allowed controls, e.g.\ $\Udomain = \{ u_t
-\in \R^{n_u} \, \st |u| \leq \Umax \}$ for a control constrained to a sphere.
-The objective $J$ will be given by a time-integral-plus-terminal cost:
-\begin{equation}
-J[u] = \int_0^\tf L(x_s, u_s) \intd{s} + M(x_\tf)
-\label{eq:generic_objective_function_deterministic}
-\end{equation}
-Where $\tf$ could be either a variable or fixed.
-
-The dynamics of $x$ are given by some controlled ordinary differential
-equation (ODE) given some initial conditions:
-\begin{equation}
-\dot{x} = f(x,t); \quad x(0) = x_0
-\label{eq:generic_dynamics_deterministic}
-\end{equation}
-
-Then we seek the optimal control $u^*$, such that:
-\begin{equation}
-u^* = \argmin_{u \in \Udomain} J[u]  
-\label{eq:generic_objective_argmin}
-\end{equation}
-
-Note that $x$ may also have final conditions, which we will deal with later,
-and/or path constraints. If there are path-constraints, we talk about a
-state-constrained problem. However, we will not face path constraints in the
-sequel and so will not say more about how to handle them.
-
-There are essentially three methods for solving equation
-\cref{eq:generic_objective_function_deterministic}. Two, Pontryagin's Minimum
-Principle and Dynamic Programing, are analytic in nature providing a set of
-equations that characterize the minimum and then solving these equations,
-numerically if necessary. The third, the direct method, simply discretizes the
-problem and solves the resulting nonlinear programing (NLP) problem via standard
-NLP techniques. We will describe both Pontryagin's Minimum Principle and Dynamic
-Programing below and indeed we will use both later on.
-
-\subsection{Pontryagin's Maximum Principle}
-Pontryagin's Principle is derived from on a variational argument, similar to the
-Euler-Lagrange equations from the Calculus of Variations, and it characterizes
-the optimal trajectory/control pair. It can be thought of as a generalization of
-the zero-tangent rule (Fermat's Rule) for finding optima in single-variable
-calculus and like both the Euler-Lagrange equations and Fermat's Rule it
-provides necessary, but not sufficient conditions for a minimum.
-
-Pontryagin's Principle also holds a similarity to Hamiltonian Mechanics, which
-should not be surprising, since Hamiltonian Mechanics is founded on
-the minimization of action.
-
-There are several versions of Pontryagin's Principle, depending on whether $x$ has
-terminal conditions or not, whether the final time, $\tf$ is specified or a
-variable in itself and most importantly whether the state has constraints or
-not. We will state the versions that are relevant for our needs, namely, $x$ has
-no constraints and $\tf$ is specified.
-
-We first introduce the Pontryagin Hamiltonian, $\H$, also called the control
-theory Hamiltonian and an adjoint state, $p \in \R^{n_x}$:
-\begin{defn} [$\H$] The control theory Hamiltonian is the function:
-$$
-\H(x,p, u) := f(x,u) \cdot p + L(x,u) \quad (x,p \in R^{n_x}, u \in \Udomain)
-$$ 
-\end{defn}
-
-\begin{thm}[Pontryagin's Fixed-Time, Free-End-Point] Assume $u^*(t)$ is
-optimal for
-\cref{eq:generic_objective_function_deterministic,eq:generic_dynamics_deterministic}
-and that $x^*$ is the corresponding trajectory.
-
-Then there exists a function $p^*:[0,\tf]:\ra \R^{n_x}$ st.
-\begin{equation}
-\begin{cases}
-\dot{x}^*(t) &=  \grad_p \H(x^*,p^*, u^*) = f(x^*, u^*)
-\\
-x(0) &= x_0
-\end{cases}
-\label{eq:pontryagin_state_ode}
-\end{equation}
-\begin{equation}
-\begin{cases}
-\dot{p}^*(t) &= -\grad_x \H(x^*,p^*, u^*) = -\grad_xL - p \grad_x f
-\\
-p^*(\tf) &= \grad_x M(x)
-\end{cases}
-\label{eq:pontryagin_adjoint_ode} 
-\end{equation}
-and
-\begin{equation}
-\H(x^*(t), p^*(t), u^*(t)) = \min_{u(t) \in \Udomain(t)}  \H(x^*(t), p^*(t),
-u(t))
-\label{eq:pontryaginH_optimality_condition} 
-\end{equation}
-\end{thm}
-We will not give a proof of this theorem (see A.2 in \cite{Evansb}),
-but we do provide a heuristic derivation in
-\cref{sec:Pontryagin_heuristic_derivation}, which is very helpful in deriving
-the necessary conditions for PDE dynamics and it explains the otherwise
-mysterious origins of $\H$ and $p$.
-
-At each time, $t$, we obtain the optimal control $u(t)$, by solving
-\cref{eq:pontryaginH_optimality_condition}. This can be easy, hard or useless.
-In the easy case, the minimum is analytically obvious, e.g.\ if the Hamiltonian
-is quadratic in $u$. If it is linear in $u$ then the minimum will occur on the
-boundaries of the control and we will talk about a bang-bang control.
-Conversely, solving the minimum of $\H$ with respect to $u$ may be impossible,
-since $x,p$ are unknown. In this case we need to have an iterative scheme which
-alternately solves
-\cref{eq:pontryaginH_optimality_condition} 
-and then \cref{eq:pontryagin_state_ode,eq:pontryagin_adjoint_ode}. Basic
-numerical techniques using this idea are discussed in the appendix in
-\cite{Kirk2004}. However, if we are going to be solving the problem numerically,
-then the direct method might be superior, see \cite{Ross2005} for a discussion.
-
-Finally, it is possible that \cref{eq:pontryaginH_optimality_condition} does not
-depend on $u$ explicitly (it will still depend  on $u$ implicitly through $x,p$)
-and so we cannot find $u$ at all by minimizing $\H$ wrt.\ $u$. This happens, for
-example, if the cost, $L$, does not depend on $u$, the dynamics, $f$, are linear
-in $u$ and $p=0$ over an interval of positive length. Such an optimal control
-problem is called singular and it is discussed briefly in \cite{Kirk2004}, and
-the theory can still be useful in finding a solution, but many complications can
-and do arise. We face a singular control problem in our own work later, however
-instead of tackling it head on, we merely regularize the problem - thus we will
-not say much more on Singular Control Problems.
-
-The version of Pontryagin's Principle for Fixed End Point, meaning $x$ has some
-terminal conditions, is very similar to the one with free end-point. Suppose 
-\begin{equation}
-\dot{x} = f(x,t); 
-\quad
-\begin{cases}
-x(0) = x_0
-\\
-x(\tf) = x_f
-\end{cases}
-\label{eq:generic_dynamics_deterministic_fixed_end_point}
-\end{equation}
-
-\begin{thm}[Pontryagin's Fixed-Time, Fixed-End-Point] Assume $u^*(t)$ is
-optimal for
-\cref{eq:generic_objective_function_deterministic,eq:generic_dynamics_deterministic_fixed_end_point}
-and that $x^*$ is the corresponding trajectory.
-
-Then there exists a function $p^*:[0,\tf]:\ra \R^{n_x}$ st.
-\begin{equation}
-\begin{cases}
-\dot{x}^*(t) &=  \grad_p \H(x^*,p^*, u^*) = f(x^*, u^*)
-\\
-x(0) &= x_0
-\\
-x(\tf) &= x_f
-\end{cases}
-\label{eq:pontryagin_state_ode}
-\end{equation}
-\begin{equation}
-\dot{p}^*(t) &= -\grad_x \H(x^*,p^*, u^*) = -\grad_xL - p \grad_x f
-\label{eq:pontryagin_adjoint_ode} 
-\end{equation}
-and
-\begin{equation}
-\H(x^*(t), p^*(t), u^*(t)) = \min_{u(t) \in \Udomain(t)}  \H(x^*(t), p^*(t),
-u(t))
-\label{eq:pontryaginH_optimality_condition} 
-\end{equation}
-\end{thm}
-I.e. $p$ loses its final conditions because $x$ has them. This relates to the
-variational nature of Pontryagin's Principle and the idea that $p$ represents
-the sensitivity of the objective wrt. small changes in $x$. Thus, if $x$ is
-fixed, there are no variations and $p$ can be anything.
-% To illustrate Pontryagin's Principle we will solve the moon
-% lander problem (Ex. 4.4.4 \cite{Evansb}), with the twist of holding the final
-% time fixed. 
-% \begin{ex}
-% 
-% \end{ex}
-
-\subsection{Dynamic Programing - the Hamilton-Jacobi-Bellman equation} 
-Dynamic Programing uses backwards recursion to tabulate the optimal control
-starting from the terminal time. The basic object in dynamic programing is the value function, $v$. In order to
-introduce it, we first extend our definition for the objective, $J$, to include later
-times, with different initial conditions.
-
-The running cost-to-go corresponding to
-\cref{eq:generic_objective_function_deterministic} is:
-\begin{equation}
-J[u; x, t] = \int_t^\tf L(x_s, u_s) \intd{s} + M(x_\tf), \quad x_t = x
-\label{eq:generic_cost_to_go_deterministic} 
-\end{equation}
-So that $J[u; x_0, 0] = J[u]$.
-
-The dynamic programming value function, $v$ is then defined by:
-$$
-v(x,t) = \inf_{u \in \Udomain} J[u; x,t] 
-$$
-it is also called the optimal cost-to-go.
-
-We immediately note the terminal conditions on $v$:
-\begin{equation}
-v(x,\tf ) = M(x)
-\end{equation}
-
-The Dynamic Programing approach is then to solve for $v$ backwards and in the
-process infer the control $u$.
-\begin{thm}[Hamilton-Jacobi-Bellman Equation] Assume that $v$ is $C^{1,1}$ in
-$x,t$. Then $v$ solves:
-\begin{equation}
-\begin{cases}
--\di_t v(x,t) &= \max_{u(t) \in \Udomain(t)} \left\{ f(x,t) \cdot \di_x v +
-L(x,u) \right\}
-\\
-v(x,\tf) &= M(x)
-\end{cases}
-\end{equation}
-\end{thm}
-We will not derive the HJB equation here, but we will derive a similar version
-later for the stochastic case, when the dynamics are given by SDEs.
-
-% \subsection{Model Predictive Control}
-% Solving the HJB equations and obtaining a feedback control is usually
-% impossible. Thus a practical alternative is to solve a simplified problem on a
-% limited horizon and use the simplified solution now and then after stepping
-% forward in time and observing the system in a new position, to solve a revised
-% limited horizon problem and use the new solution.
-% Let us make this more precise.
-
-\section{Stochastic Optimal Control}
-Here we extend the Dynamic Programing theory to the case where the dynamics are
-governed by an SDE:
-\begin{equation}
-dX = F(X,t, u) dt + G(X,t) dW
-\end{equation} 
-i.e.\ we assume control affects only the drift term. This 
-assumption is not necessary, but it corresponds to our problems.
-
-For stochastic dynamics one can only optimize probabilistically and so
-$J$ becomes:
-\begin{equation}
-J[u] = \Exp \left[ \int_0^\tf L(x_s, u_s) \intd{s} + M(X_\tf)
-\right] 
-\label{eq:generic_objective_function_stochastic}
-\end{equation}
-while the running cost-to-go is:
-\begin{equation}
-J[u; x, t] = \Exp \left[ \int_t^\tf L(X_s, u_s) \intd{s} + M(X_\tf) \right], 
-\quad X_t = x
-\label{eq:generic_cost_to_go_stochastic}
-\end{equation}
-the value function is:
-\begin{equation}
-v(x,t) := \inf_{u \in \Udomain} J[u; x,t]
-\end{equation}
-
-Finally, we again have that $v$ satisfies a certain PDE:
-\begin{thm}[Stochastic HJB] 
-\label{thm:stochastic_hjb}
-\begin{equation}
-\begin{cases}
--\di_t v(x,t) &=  \max_{u(t) \in \Udomain(t)} \big\{ L(x,u)  +
-F(x,t,u) \cdot \di_x v
-\big\} + \frac{G^2(x,t)}{2} \di_x^2v
-\\
-v(x,\tf) &= M(x)
-\end{cases}
-\end{equation}
-\begin{proof}[Heuristic Derivation (from \cite{Evansb})] Suppose we are at time
-$t$ and $X_t = x$.
-
-Take a time increment $[t, t+h]$ and assume that during that time we apply a
-control $u$ and subsequently we apply the the optimal control $u^*$. The
-running-cost will then break down as: 
-$$
-J[u; x,t] = \Exp \left[ \int_t^{t+h} L(X_s, u_s) \intd{s}  + v(X_{t+h},
-t+h) \right] $$
-
-Since $v(x,t) = \inf J(u; x,t)$, we must have that:
-$$
-v(x,t) \leq  \Exp \left[ \int_t^{t+h} L(X_s, u_s) \intd{s}  + v(X_{t+h},
-t+h) \right] $$
-or, rearranging, that
-\begin{align*}
-0 \leq^&  \Exp \left[ \int_t^{t+h} L(X_s, u_s) \intd{s} \right]  
-+ \underbrace{\Exp \left[ v(X_{t+h}, t+h) - v(x,t) \right]}_{\Exp\left[ dv
-\right]}
-\end{align*}
-Now, $\Exp\left[ dv \right]$ can be expressed using Dynkin's Formula, as:
-$$
-\Exp[dv] = \int_t^{t+h} \di_t v +  F\cdot \di_x v + \frac{G^2 }{2}\cdot \di_x^2
-v \intd{s} $$
-Plugging that back, we get
-$$
-0 \leq \int_t^{t+h} L(X,u) +  \di_tv +  F\cdot \di_x v + \frac{G^2 }{2}\cdot
-\di_x^2 v \intd{s} $$
-Taking $h \ra 0$ we get 
-$$
-0 \leq  L(x,u) +  \di_tv(x,t) +  F\cdot \di_x v(x,t) + \frac{G^2 }{2}\cdot
-\di_x^2 v(x,t) $$
-and we conjecture that for the actual optimal control, the inequality becomes an
-equality:
-\begin{equation}
-\begin{cases}
-0 &=   L(x,u^*)+ \di_t v(x,t) +
- F(x,t,u^*)\cdot \di_x v(x,t) + \frac{G^2(x,t)}{2}\cdot \di_x^2 v(x,t)
-\\
-u^*(t) &= \argmax_{u \in \Udomain(t)}  
-\big\{L(x,u) + F(x,t,u)\cdot \di_x v \big\}
-\end{cases}
-\end{equation}
-\end{proof}
-\end{thm}
-% A rigorous proof can be found in \cite{Krylov2008}, sec???ch???
-
-\section{Deterministic Optimal Control in Infinite Dimensions}
-Now we discuss Optimal Control when the dynamics are deterministic but of
-infinite-dimension, i.e. the state evolution is a PDE instead of an ODE.  The
-reason we discuss this, is that optimization of ordinary SDEs in an open-loop
-control is actually a deterministic optimal control problem in infinite
-dimensions.
-
-Why?
-
-Since the objective is probabilistic and the probability density evolution is
-given by the (deterministic) Fokker-Planck PDE, we need to optimize a
-deterministic system whose dynamics are given by the Fokker-Planck equation.
-
-There is no all-encompassing generalization of Pontryagin's Principle to
-infinite dimensions, but a Maximum Principle can be derived for certain systems.
-To illustrate the idea we will discuss a simple case as in \cite{Palmer2011}.
-Our controlled SDE will be given by: 
-$$ dX_t = F(X,t, u) dt + G(X,t) dW, $$
-The corresponding probability density is given by: $$ \di_t \f=
--\sum_i \di_{x_i} \left[ F_i(x,t,u) \f(x,t) \right] + \sum_{i,j}  \di^2_{x_i
-x_j} \left[ D_{ij}(x,t) \f(x,t) \right] $$ which we will write as $$ \dot{\f} =
-\L_{u} [\f] $$ where $\L_{u}$ is the differential operator corresponding to the
-RHS of the Fokker-Planck equation parametrized by the control $u$. For now we
-will assume that there are no BCs, and the domain of $X$ is all of $\R^n$.
-
-A natural inner-product, $ \langle.,. \rangle$ on functions in $L^2[\R^n]$ is
-given by: $$
- \langle\psi,\phi  \rangle = \int_{\R^n} \psi(x) \phi(x) \intd{x}
-$$
-Our objective will be to minimize:
-\begin{equation}
-J[u] = \int_0^\tf  \langle L(\cdot,s,u), \f_u(s) \rangle \intd{s} +  
-\langle M(\cdot),\f(\tf) \rangle
-\label{eq:generic_objective_function_probabilistic}  
-\end{equation}.
-
-\Cref{eq:generic_objective_function_probabilistic} is a very natural cost
-function in the stochastic dynamics case, because it can also be written as: 
-$$
-J[u] = \Exp \left[ \int_0^\tf  L(X_s, s,u) \intd{s} + M(X_\tf) \right] 
-$$ which
-is just \cref{eq:generic_objective_function_stochastic}. We use
-\cref{eq:generic_objective_function_stochastic} if we have some feedback and
-want to use dynamic programming. We use
-\cref{eq:generic_objective_function_probabilistic} if we have no observations
-and need to run an open-loop control.
-
-Since, \cref{eq:generic_objective_function_probabilistic} is the objective
-function for deterministic dynamics, we can use Pontryagin's Principle adapted
-to the case of infinite dimensional dynamics (PDEs).
-
-Introduce an adjoint co-state, $\l \in \R^n$ and let the Hamiltonian, $\H$ be:
-$$
-\H(\f, \l, u)= \langle \phi, L + \Lstar[\l] \rangle$$
-where $\Lstar$ is the adjoint operator to $\L$ defined via:
-$$
-\langle \L [\phi], \l \rangle =\langle \phi, \Lstar[\l] \rangle
-$$
-
-Let $\l$ evolve (backwards) according to:
-\begin{equation*}
-\begin{cases}
--\di_t \l =& \Lstar[\l] + L
-\\
-\l(\tf) =& M(x)  
-\end{cases}
-\end{equation*}
-then the optimal control, $u$, is given via
-$$
-u = \argmin_u {\H(\f, \l, u)}
-$$
-
-An important fact, useful for computations, is that the total derivative of the
-objective with respect to the control is given by the partial derivative of the
-Hamiltonian with respect to the control:
-\begin{equation}
-\grad_u \big[J[u] \big] = \di_u \H(\f, \l, u)
-\label{eq:objective_gradient_wrt_control}
-\end{equation}
-
-So for a given $u$, we can compute $\f,\l$, calculate the objective gradient
-using \cref{eq:objective_gradient_wrt_control} and then iterate in the direction
-of descent (if we are minimizing).
-
-All that remains is to compute $\Lstar$. It turns out that if we do not have any
-BCs, then
-$$
-\Lstar[\l] = \sum_i F_i(x,t) \cdot \di_{x_i} \l + \sum_{i,j} D_{ij}(x,t)
-\di^2_{x_i x_j}\l 
-$$ 
-which is just the generator of $X$!
-
-We will see later, that things become more complicated when we need to
-consider a PDE's BCs.
-
-\section{Closed-loop vs.\ Open-Loop control}
-At this point we stop to briefly discuss the difference between
-open-loop control and closed-loop control.
-
-By closed-loop control, we mean that at any time we have observations of the
-system state, $x$ and its value can influence the control. In open loop control,
-we decide on the value of $u(t)$ a priori and apply it independently of the
-value of $x(t)$, based only on the knowledge of the system dynamics and the
-initial conditions.
-
-If the system dynamics are deterministic there is no added value in closed-loop
-control, since the state evolution is perfectly predictable and one can infer
-where $x$ will be based only on the dynamics and the applied control. For
-stochastic dynamics this is no longer true and there is something to be gained
-by observing the state, since different controls can be applied depending on the
-different realizations of the stochastic process driving the dynamics.
-Naturally, if one has observations of the system, one would prefer to use this
-and apply a closed-loop control. Thus, an open-loop control would be used
-only if one has no reliable knowledge of the system state.
-
-\section{A Stochastic Optimal Control Problem from Mathematical Neuroscience}
-Neurons relay information by means of voltage spikes - sudden sharp increases in
-voltage. Although many details remain unclear, the information content is
-thought to be contained in the length of the time-interval between these spikes.
-In the simplest case, this can be thought of as a rate  -the average number of
-spikes per time interval, but more complicated coding schemes are hypothesized to
-exist.
-
-Many experiments allow for manipulating an individual neural cell. A natural
-goal then is to make a cell produce a given spike train. This may arise, for
-example, in brain-machine interfaces or in artificial prosthetics.
-
-We now consider this problem in the framework of Stochastic Optimal Control and
-give both a feedback solution, when the cell voltage is explicitly observable as
-well as an open-loop solution when only the occurrence of a spike is
-robustly observable.
-
-\subsection{Problem Formulation}
-The most basic representative model for a neuron is the noisy
-leaky-integrate-and-fire model:
+\section{Problem Formulation}
+Given a noisy LIF neuronal model:
 \begin{equation}
@@
 \end{equation}
-In this $X$ represents the membrane electric potential which tries to revert to
-$0$ with a time-const of $\tc$, $dW$ is a Brownian motion increment with
-intensity $\b$ and $\a(t)$ is our external control. Having last spiked at
-time $0$, the potential hits $\xth$ at time $\ts$, we record a spike
-of duration $\ts$, the potential resets to $0$ and we start all over again.
-
-Suppose we want to ensure that spikes occur exactly at the predetermined
+
+So $X$ represents the membrane el. potential which tries to revert to $0$ with a
+time-const of $\tc$, $dW$ is a Brownian motion increment with intensity $\b$.
+$\a(t)$ is our external control and having last spiked at time $0$, the
+potential hits $\xth$ at time
+$t$, then we record the next 'spike' of duration $t$ and reset the potential to
+$0$.
+
+Suppose we want want to ensure that spikes occur exactly at the predetermined
 times $\tns$. Then given the renewal nature of our spiking model, we can work
-sequentially, focusing only on the first of these intervals, of duration $t_1 =
-\T$, and then ask: how do we choose $\a(t)$ st. $\ts = \T$. There are several
-ways to specify what we mean by 'make $\ts = \T$', all of which result in optimal
-control problems. A natural one is a least squares solution:
+sequentially, focus only on the first of these intervals, of duration $\T$ and
+then ask: how do we choose $\a(t)$ st. $\ts= \T$. This is not specific enough
+since the probabilistic aspects of the problem ensure that we will not be able
+to hit $\T$ precisely, unless we take extreme forms for $\a$, which will
+naturally be disallowed in the sequel.
+
+In the probabilistic context, there are several ways to specify to what we
+mean by 'make $\ts = \T$', all of which result in optimal control problems.
+
+The first is to ensure that we maximize the probability of $\T$. Let:
+\begin{equation}
+\begin{array}{lcll}
+g_{\a}(\t) \intd{\t} &:=& \Prob[\ts \in [\t, \t + \intd{\t}] ] &
+ \textrm{(probability density)} 
+\\
+G_{\a}(t) &:=& \Prob[\ts \leq t ] = \int_0^t g_{\phi}(\t) \intd{\t}
+&
+ \textrm{(cumulative distribution)}
+\\
+\G_{\a}(t) &:= & \Prob[\ts>t] = 1 - G_{\a}(t)
+&
+ \textrm{(survivor distribution)}
+\end{array}
+\label{eq:ISI_distribution_functions}
+\end{equation}
+be the various distributions for the spike time $\ts$. Then given $\T$, we want
+to take:
+\begin{equation}
+\a(t) = \argmax_{\a(t)}  g_{\a}(\T)
+\label{eq:OC_MLE_density} 
+\end{equation}
+or in a softer sense:
+\begin{multline}
+\a(t) = \argmax_{\a(t)}  [G_{\a}(\T+\delta T) - G_{\a}(\T-\delta T)]
+\\ 
+	  = \argmin_{\a(t)} [\G_{\a}(\T+\delta T) - \G_{\a}(\T-\delta T)]
+\label{eq:OC_MLE_distribution} 
+\end{multline}
+for some small positive variation,  $\delta T$.
+
+Alternatively, we can desire a least squares solution:
+\begin{equation}
+\a(t) = \argmin_{\a(t)} \{ \,(\Exp[\ts] - \T)^2  \, \}
+\label{eq:OC_LS_mean}   
+\end{equation}
+or 
 \begin{equation}
@@
 \end{equation}
-
-It is also natural to assume that our control has certain limitations. The
-simplest and most common are simple bounded constraints:
+i.e we can minimize the mean or the mean square; or alternatively we can
+minimize both at the same time:
+\begin{equation}
+\a(t) = \argmin_{\a(t)} \{\, \Exp[(\ts- \T)^2  ] + m \Exp[(\ts- \T)^2 \,\}
+\label{eq:OC_LS_mean-variance}   
+\end{equation}
+where $m$ sets our preference for mean-square minimization over minimization
+just in the mean.
+
+% Note that all three of
+% \cref{eq:OC_LS_mean,eq:OC_LS_variance,eq:OC_LS_mean-variance} contain the $\Exp$
+% operator and as such seemingly require us to know the distributions of $\ts$,
+% i.e.\ $g_{\a}(\t)$, $G_{\a}(t)$ and $\G_{\a}(t)$ or do we\ldots?
+
+% Since we have no hope of finding the density, $g_{\a}$, analytically, we will
+% naturally resort to numerical methods both for calculating $g_{\a}(\t)$ and for
+% solving the various optimal control problems.
+
+So far we have not said anything about the constraints on the control. For now
+we will just take simple bounded constraints:
 \begin{equation}
@@
 
-% \subsection{Optimal Control Formulation}
-In addition to \cref{eq:OC_LS_variance}, our objective will also contain a
-running cost based on the control. This regularizes the problem eliminating the
-subtleties of singular-control situations and it might be argued appropritate
-to avoid excessive control.
-
-So we will seek an optimal control, $\a$ that solves:
+\subsection{Dynamic Programing Basics}
+We want to use Dynamic Programing to solve for the optimal control $\a(t)$.
+Dynamic Programming has a well-developed theory for the control of both
+diffusion- and jump- stochastic differential equations, see e.g.\
+\cite{Hanson2007,Øksendal2005}. Here we just state the main result for
+the diffusion case, with a non-controlled diffusion term.
+
+Let the controlled dynamics be:
+\begin{equation}
+dX = a(x,t, \a) \intd{t} + b(x,t)\intd{W}; \quad X_0 = x_0
+\end{equation} 
+where $X$ is the 'state' and $\a$ is the 'control'.
+
+Let the objective be:
+\begin{equation}
+J = \int_0^T L(s,x(t),\a(t)) \intd{s} + M(X_T)
+\end{equation} 
+Suppose we are minimizing, and set the value function, also called the
+optimal cost-to-go, be:
+\begin{equation}
+v(x,t) = \inf_{\a(s>t)} \int_t^T L(s,x,\a) \intd{s} + M(X_T); \quad X_t = x
+\end{equation}
+then $v$ satisfies the following  backward parabolic PDE, the
+Hamilton-Jacobi-Bellman (HJB) equation:
+\begin{equation}
+\di_t v(x,t) + \inf_{\a}  \{ L(t,x,a)  + a(x,t,\a)\di_x v + \tfrac{b^2(x,t)}{2}
+\di^2_x v \}
+\end{equation}
+i.e.\ we tabulate the optimal value for any given $x$ in the state-space at any
+given time $t$. With the terminal conditions:
+\begin{equation}
+v(x,T) = M(x)
+\end{equation}
+
+The boundary conditions on $v$ are quite context-specific and we will see how
+they look for our problem below.
+
+The optimal control is embedded in the evolution of $v$ and given $x,t$
+and $v(x,t)$ we will have: 
+\begin{equation}
+\a(t) = \inf_{\a}  \{ L(t,x,a)  + a(x,t,\a)\di_x v + \tfrac{b^2(x,t)}{2}
+\di^2_x v \}
+\end{equation}
+
+\subsection{The 'feedback' for our problem}
+The key product of an application of Dynamic Programing is a feedback control
+$\a(t) = \a(x(t))$, i.e.\ the current control depends on the current state (the
+observe-react paradigm). However, in our particular problem, we cannot observe
+the state.
+
+We are not completely blind - at time $t$ we know whether $X$ has spiked
+or not! So if $t > \T$, and we haven't spiked, we know we must put $\a =
+\amax$. But for $t < \T$, we don't know where $X_t$ is.
+
+To repeat - we are in some limbo state between fully observing the state and
+knowing nothing about the state - we know whether the spike has occurred.
+
+If we can fully observed the state, we will know what $X_t$ is; if we know
+nothing about the state we only have its unconstrained transition density
+$\rho(x,t)$ of $X_t$. However, in our case, we know the transition density,
+$f(x,t)$, of $X_t$, conditioned on $X_{s <=t} < \xth$. The main difference
+between the $\rho$ and $f$ - is that $\rho(x,t)$ gives a non-zero probability to
+$X_t > \xth$, whereas we have the knowledge that this is impossible. And so
+our actual optimal control law - the one that will be a candidate for
+'real-world' implementation will be one that controls not $X_t$, but the
+computable transition density $f(x,t)$.
+
+% \section{Deterministic Control}
+% \label{sec:deterministic_control}
+% Let us first discuss the problem in the simpler, deterministic setting. We will
+% assume that given $\T$ it is possible to get from $x_0$ to $\xth$ in time $t =
+% \T$. Our problem then is:
+% 
+% Find $\a(t) \in [\amin, \amax]$ st. 
+% \begin{equation}
+% \begin{gathered}
+% \dot{x} = \a(t) - x_t/\tc
+% \\
+% \begin{cases}
+% x(.0) = .0 \\
+% x(\T) = \xth \\
+% x(t < \T) < \xth
+% \end{cases}
+% \end{gathered}
+% \label{eq:X_evolution_deterministic}
+% \end{equation}
+% Ok, this problem is entirely ill-posed (there are very many solutions, $\a(t)$)
+% and so to pick out one in particular we need to put some extra constraints on the control,
+% $\a$.
+
+\section{Maximum-Likelihood Control}
+We will not dwell much on this here, except to make the following point. Suppose
+the MLE distribution $g_{\a^*}$ is very slanted to the right, say it looks like:
+\begin{equation}
+g_{\a^*}(t) \propto 
+\begin{cases}
+0 & t < \T
+\\
+e^{-(t-T)}&  \ow
+\end{cases}
+\end{equation}
+then the most likely $\ts$ will indeed be $\T$, but the bias etc. will be 'bad'.
+However, since I am not an ML expert, I shouldn't comment much
+more on whether this outcome is possible - maybe such a pathological situation
+cannot occur and ML control is the right thing to do\ldots
+
+\section{Minimization-in-the-Variance Control}
+Here we discuss finding the control that minimizes \cref{eq:OC_LS_variance} plus
+a running energy cost:
 \begin{equation}
@@
 where $\e$ measure how much weight we put on minimizing the energy expenditure
-and if $\e \rightarrow 0$, then we do not care at all.
-
-Experimentally it is often difficult to observe precisely the exact voltage
-trajectory, but the occurrence of a spike is much more easily observable. As
-such a closed-loop approach is not implementable since the state is unknown. In
-the sequel we will perform three calculations. First we will derive the HJB
-equation for a closed-loop control if the system was observable, and use this as
-a benchmark for the best possible performance of an open-loop control. Then we
-will calculate the most simple-minded open-loop control, which is just the
-deterministic control of the dynamics when assuming there is no noise, $\b=0$.
-THen we will calculate the optimal open-loop control by optimizing the
-transition density. And finally we will compare the performance of all three
-controls.
-
-\section{Closed-Loop Solution - Dynamic Programing}
-Given $t, X_t = x$, let $\t$ be the (unknown) remaining time to spike. For
-example, given $t=0, X=.0$, $\t = \ts$, the entire duration of the inter-spike interval.
-Starting from a general $t, x$, our remaining-cost objective, $J[\a; x,t]$,
+and if $\e \rightarrow 0$, then we don't care at all. This running cost is
+similar to the regularization term $R(\x)$ in eq.\ 3 in \cite{Ahmadian2011}.
+
+\subsection{Dynamic Programing Formulation}
+Given $x,t$, let $\t$ be the (unknown) remaining time to spike. For example,
+given $t=0, X=.0$, $\t = \ts$, the entire duration of the inter-spike interval.
+Starting from a general $t, X_t=x$, our remaining-cost objective, $J[\a, x,t]$,
 will be:
@@
 the difference between $\t$ and ($T-t$), given the current state $x$. For now we
-assume that we know $x$, later we will treat the problem when that is not so.
+assume we know $x$, later we will admit that we don't.
 
@@
 \end{equation}
-However we need to recall the bounds on the control, $\amin < \a < \amax$, and
-thus
-\begin{equation}
-\astar(x,t) = \min \left(\amax, \max\left(\amin, -\frac{\di_x v}{2\e}\right)
-\right)
-\label{eq:OC_LS_variance_energy_quadratic_control_bounds}
-\end{equation}
-With this the HJB becomes
-\begin{equation}
-\di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + 
-\e (\astar)^2 + (\astar-\tfrac{x}{\tc}) \di_x v 
-= 0
-\label{eq:OC_LS_mean_HJB_bounded_control}
-\end{equation}
-
-In the sequel we will drop the superscript star on the optimal control and
-just write $\a(t)$ for the solution of
-\cref{eq:OC_LS_variance_energy_quadratic_control_bounds}. 
-% Thus the control can be solved out of the HJB equation, which now becomes:
-% \begin{equation}
-% \di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + \tfrac{(\di_x v)^2}{\e}
-% -\tfrac{x}{\tc}\di_x v= 0
-% \label{eq:OC_LS_mean_HJB}
-% \end{equation}
+Thus the control can be solved out of the HJB equation, which now becomes:
+\begin{equation}
+\di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + \tfrac{(\di_x v)^2}{\e}
+-\tfrac{x}{\tc}\di_x v= 0
+\label{eq:OC_LS_mean_HJB}
+\end{equation}
 
@@
 Finally we come to the Terminal Conditions (TCs) for $v$. The idea is simple: if
-we reach $\T$ without having spiked we will just slam on the accelerator: i.e.\
-$$t>T \implies \a(t) = \amax$$ Thus: $$ v(x,T) = \Exp[\t^2|x, \amax] $$.
+we reach $\T$ without having spiked we will just slam on the accelerator: i.e.
+$$t>T \implies \a(t) = \amax$$ Thus: $$ v(x,T) = \Exp[\t^2|x, \amax] $$ Note
+that there are formulas for the second moment of the first-passage time of an OU
+process and they can be used to compute $v(x,T)$ without solving for the
+density, $g_{\amax}(\t)$, see \cite{Ditlevsen2008a},\cite{Ricciardi1988}.
+Moreover in a real-time setting these can be pre-computed, without the need to
+know $\T$.
 
@@
 \begin{gathered}
-\di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + 
-\e \a^2 + (\a -\tfrac{x}{\tc}) \di_x v 
-= 0
-\\
-\a (x,t) = \min \left(\amax, \max\left(\amin, -\frac{\di_x v}{2\e}\right)
-\right)
+% \di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + 
+% (\a(t) -\tfrac{x}{\tc}) \di_x v = 0
+\di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + \tfrac{(\di_x v)^2}{\e}
+-\tfrac{x}{\tc}\di_x v = 0
+\\
+\a(t) =
+% \begin{cases}
+% \amin & \di_x v > 0
+% \\
+% \amax & \di_x v < 0
+% \end{cases}
+\frac{-\di_x v(x(t),t)}{2\e}
 \\
@@
 \end{equation} 
-We are solving $v(x,t)$ over $[\xmin, \xth] \times [0,T]$.
+It should be clear by now that we are solving $v(x,t)$ on the domain $\O =
+[\xmin, \xth] \times [0,T]$.
 
@@
 
-% Solving \cref{eq:BVP_T_generic} for $S$ gives: \begin{eqnarray} S(y) &=& -
-% \exp\left(\smallint \tfrac UD \right) \cdot \int_\xmin^{y} \exp\left(-\int
-% \tfrac UD\right) \frac{f(\xi)}D \intd{\xi}
-% \\
-% &=& - \exp\left( \frac{\a y - \frac{y^2}{2\tc}}D \right) \cdot \int_\xmin^{y} 
-%  \exp\left(- \frac{\a \xi - \tfrac{\xi^2}{2\tc}}D \right) \frac{f(\xi)}D
-% \intd{\xi} \label{eq:BVP_T_generic_soln} \end{eqnarray}
-We can solve \cref{eq:BVP_T_generic} analytically, but the analytical solution
-turns out to be awkward for numerical calculations, so we
-are better off just calculating the numerical solutions to \cref{eq:BVP_Tone,eq:BVP_Ttwo} directly
-via an ODE solver, starting from $S(\xmin) = .0$ and integrating until $\xth$.
+Solving \cref{eq:BVP_T_generic} for $S$ gives:
+\begin{eqnarray}
+S(y) &=& - \exp\left(\smallint \tfrac UD \right) \cdot \int_\xmin^{y} 
+\exp\left(-\int \tfrac UD\right) \frac{f(\xi)}D
+\intd{\xi}
+\\
+ &=& - \exp\left( \frac{\a y - \frac{y^2}{2\tc}}D \right) 
+ 		\cdot 
+ 		\int_\xmin^{y}   \exp\left(- \frac{\a \xi - \tfrac{\xi^2}{2\tc}}D \right)
+ 		\frac{f(\xi)}D \intd{\xi}
+\label{eq:BVP_T_generic_soln}
+\end{eqnarray}
+
+But the analytical solution is not very useful for numerical calculations,
+we are better off just calculating the numerical solutions to
+\cref{eq:BVP_Tone,eq:BVP_Ttwo} directly via an ODE solver, starting from
+$S(\xmin) = .0$ and integrating until $\xth$.
 
@@
 solutions see \cref{fig:Ti_empirical_vs_analytic}.
-\begin{figure}[htp]
+\begin{figure}[h]
 \begin{center}
@@
 \cref{eq:BVP_Tone,eq:BVP_Ttwo} and empirical statistics from simulating the
-hitting times of \cref{eq:X_evolution_uo} for four different pairs of $\a,\b$.
-In the top panel we show the first moment of the remaining time-to-spike
-starting from $x$, $\Tone(x)$, and in the bottom panel, we show the second
-moment, $\Ttwo(x)$. We have used only $N=128$ samples to form the moment
-statistics.}
+hitting times of \cref{eq:X_evolution_uo}. We've used only $N=128$ samples to
+form the moment statistics.}
 \label{fig:Ti_empirical_vs_analytic}
@@
 
-\subsection{Solving the HJB - the numerical method for
-\cref{eq:OC_LS_mean_HJB_bounded_control}}
+\subsection{Solving the HJB}
 We now have a PDE for $v$ and an algorithm for computing all the BCs and TCs of
@@
 
-Things will break down if $\e$ is very small in which case the quadratic term,
-$(\di_x v)^2 / \e$, will dominate and cause 'havoc' or if $\b$ is very small in
-which case the smoothing effect of the second derivative will be negligible and
-we will be in the zone of nonlinear conservation equations. Let us hope for the
-best now and assume these two problems do not arise, for the parameters we
-select.
+Things will break down if $\e$ is very small in which case the quadratic term
+will dominate and cause 'havoc' or if $\b$ is very small in which case the
+smoothing effect of the second derivative will be negligible and we will be in
+the zone of nonlinear conservation equations.
+
+Let us hope for the best now and assume these two problems do not arise.
+
+\subsection{Bounded Control}
+So far we have assumed that $\a$ is not constrained. However, physically, we
+will have some bounds, i.e. $\a \in [\amin, \amax]$. Then the optimal control
+will look like:
+\begin{equation}
+\astar(x,t) = \min \left(\amax, \max\left(\amin, -\frac{\di_x v}{2\e}\right)
+\right)
+\label{eq:OC_LS_variance_energy_quadratic_control_bounds}
+\end{equation}
+And the HJB becomes:
+\begin{equation}
+\di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + 
+\e (\astar)^2 + (\astar-\tfrac{x}{\tc}) \di_x v 
+= 0
+\label{eq:OC_LS_mean_HJB_bounded_control}
+\end{equation}
+
+In the sequel we will drop the superscript $*$ on the optimal control and just
+assume $\a$ is given by
+\cref{eq:OC_LS_variance_energy_quadratic_control_bounds}.
+
+\subsection{The numerical method for
+\cref{eq:OC_LS_mean_HJB_bounded_control}}
+ We will try the simplest thing and go from there:
 
@@
 
-Then our finite-difference scheme in time looks like:
+Then our finite-difference scheme looks like:
 \begin{equation}
@@
 
-I.e.\ we have removed the non-linearity in $v$ by computing the control, $\a$ at
+I.e. we have removed the non-linearity in $v$ by computing the control, $\a$ at
 the next, already computed, time-step and used central-differencing
-otherwise.
-
-For the spatial derivatives, we will not upwind for now and also use central
-differencing:
+otherwise. We will not upwind \emph{al all} for now\ldots I.e. 
 $$\di_x v = \frac{v_+ - v_-}{2\Delta x}$$
 and as is standard:
-$$\di^2_x v = \frac{v_+ - 2 v_0 + v_-}{(\Delta x)^2}$$
-
-We will not discuss a detailed study of the convergence properties of the
-numerical method, but just demonstrate its performance on a representative
-parameter set in \cref{fig:HJB_attempt} we see that the method is stable for a
-particular parameter set.
-
-% \begin{figure}[ht]
-% \begin{center}
-%   \includegraphics[width=.7\textwidth]{Figs/KolmogorovBVPMoments_a=10_t=9_b=8.png}
-%   \caption[]{TCs for alpha = 1.0;  tauchar = .9; beta = .75;}
-%   \label{fig:TCs_extended_x}
-% \end{center}
-% \end{figure}
-
-Let's discuss the most salient features of \cref{fig:HJB_attempt}. At the end,
-when $t = \T$, the value function, $v$ is monotonely decreasing. This makes
-sense, becaue the lower $x$ is the more time on average will it take us to spike
-and the more delay will we have from our desired spike time. However, as we go
-back in time $v$ inverts, in the sense that for $t \ll \T$, we would like to
-be somewhat far from the boundary to avoid the risk of spiking early.
-
-Now focus on the controls, \cref{fig:HJB_attempt_control}: Naturally, at the
-end we have the control assume its maximum value, $\a(t>=\T) = \amax$. As we
-start to go back in time however, two things start to happen, the control $\a$
-decreases for $x \ll \xth$, and it becomes negative for $x \lesssim \xth$. 
-That is fully consistent with the problem objective - for
-$t<\T$, we want to distance $X_t$ from the threshold to avoid an early spike.
-\begin{figure}[h!]
+$$\di_x v = \frac{v_+ - 2 v_0 + v_-}{(\Delta x)^2}$$
+
+Will that work? Let's see - \cref{fig:HJB_attempt} shows that it works for a
+particular parameter set, which sounds representative (to me). 
+\begin{figure}[ht]
 \begin{center}
-\subfloat[Solution, $v$]
+\subfloat[Sol'n, $v$]
 {
@@
 \end{figure}
-
-\section{Deterministic Control}
+% \begin{figure}[ht]
+% \begin{center}
+%   \includegraphics[width=.7\textwidth]{Figs/KolmogorovBVPMoments_a=10_t=9_b=8.png}
+%   \caption[]{TCs for alpha = 1.0;  tauchar = .9; beta = .75;}
+%   \label{fig:TCs_extended_x}
+% \end{center}
+% \end{figure}
+
+Let's discuss the most salient features of \cref{fig:HJB_attempt}. At the end,
+when $t = \T$, the value function, $v$ is monotonely decreasing. This makes
+sense, becaue the lower $x$ is the more time on average will it take us to hit spike
+and the more delay will we have from our desired spike time. However, as we go
+back in time $v$ inverts, in the sense that for $t \ll \T$, we would like to
+be somewhat far from the boundary to avoid the risk of spiking early.
+
+Now focus on the controls, \cref{fig:HJB_attempt_control}: Naturally, at the
+end we have pedal-to-the-metal type control, $\a(t>=T) = \amax$, as we start to
+go back however, two things start to happen, the velocity $\a$ decreases for $x
+\ll \xth$, and it even becomes negative for $x \approx \xth$. 
+That is fully consistent with the problem objective - near $\xth$ and for
+$t<\T$, we want to distance $X_t$ from the threshold to avoid an early spike.
+
+So far so good. And we can be glad that for this parameter set -
+there were no numerical instabilities, despite our simplistic numerical method.
+
+\section{Deterministic vs. Fully-Informed Stochastic Control}
+At this point we are ready to attempt a comparison between a blind deterministic
+type control and our stochastic control. But first let's calculate the controls
+for our problem with deterministic dynamics:
+
+\subsection{Computing the Deterministic Control}
 \label{sec:deterministic_ode_control}
-Now we will compute the optimal control in the case that $\b = 0$. The reason we
-do this is that since we have assumed that we have no feedback, we attempt to
-devise an open-controller by controlling the non-random dynamics. The
-calculation of this control turns out to be very simple:
-
-First let us assume that we actually can get to $\xth$ from $0$ in $\T$ time
+Since we are ignoring the noise, $\a(t)$ can be pre-calculated, for a given
+$\T$. i.e. we can provide an open-loop scheme to the 'controller'.
+
+Let's quickly derive this deterministic-case $\a$; it turns out to be very
+simple:
+
+First let's assume that we actually can get to $\xth$ from $0$ in $\T$ time
 given $\amax$ and no noise.
@@
 \begin{cases}
-\dot{x} = (\a(t) - \frac{x}{\tc})
+\dot{x} = (\m + \a(t) - \frac{x}{\tc})
 \\
@@
 \begin{equation}
-\H(x,p, \a) = \e \a^2 + p \cdot (\a - x/\tc )
+\H(x,p, \a) = \e \a^2 + p \cdot (\m + \a - x/\tc )
 \end{equation}
@@
 \begin{align*}
-\dot{x} & = \a(t) -  \frac{x}{\tc}
-\\ & = \frac{p_0 \exp(t / \tc)}{2\e} - \frac{x}{\tc}
-\\
-e^{t/\tc}x &= \int^t e^{s/\tc}\frac{p_0 \exp(s / \tc)}{2\e} \intd{s}
-\\
-x &= e^{-t/\tc} \cdot \frac{p_0 }{2\e} \tc \left[  \exp(2 t / \tc) -1 \right] 
-\\
-&= \frac{p_0 }{\e} \tc \cdot  \sinh (t/ \tc)
+\dot{x} & = \m + \a(t) -  \frac{x}{\tc}
+\\ & = \m +  \frac{p_0 \exp(t / \tc)}{2\e} - \frac{x}{\tc}
+\\
+e^{t/\tc}x &= \int^t e^{s/\tc}\frac{p_0 \exp(s / \tc)}{2\e} \intd{s} +
+				\int^t e^{s/\tc}\m \intd{s}  
+\\
+x &= e^{-t/\tc} \cdot \left( \frac{p_0 }{2\e} \tc \left[  e^{2 t / \tc} -1
+\right] + \m\tc (e^{t/\tc} - 1)\right)
+\\
+&= \frac{p_0 }{\e} \tc \cdot  \sinh (t/ \tc) + \m\tc \cdot( 1 - e^{-t/\tc}))
 \end{align*}
@@
 $$
-x(\T) = \xth \implies p_0 = \e \frac{\xth  }{ \tc \sinh (\T/ \tc) }
+x(\T) = \xth \implies p_0 = \e \frac{ \xth - \m\tc  }{ \tc \sinh (\T/ \tc) }
 $$
@@
 
-If we have saturation, i.e. if $\a$ hits its upper bound, then the situation is
-mildly more complicated. Since $p$ is monotonically increasing, as long as $p_0
-> \amin$ then $\a(t) > \amin \forall t$. So the saturation is simplified to:
+If we have saturation, i.e. if $\a$ hits its bound, then the situation is mildly
+more complicated. Since $p$ is monotonely increasing, as long as $p_0 >
+\amin$ then $\a(t) > \amin \forall t$. So the saturation is simplified to:
 \begin{equation}
@@
 obtain the new value of $p_0$. Or we can just brute-force solve for $p_0$, by a
-simple shooting method. Using the latter approach, a solution-control pair for
-the parameter set we will use later is shown in \cref{fig:deterministic_soln_controls}. 
+simple shooting method.
+
+using the latter approach, a solution-control pair for the parameter set we will
+use later is shown in \cref{fig:deterministic_soln_controls}. 
 %\usepackage{graphics} is needed for \includegraphics
@@
   \includegraphics[width=.7\textwidth]{Figs/ControlSimulator/deterministic_example.png}
-  \caption[]{An example trajectory for the deterministic dynamics. Above is the
-  state evolution, $x(t)$ given $\b=0$ and below is the applied control,
-  $\a(t)$. }
+  \caption[]{An example trajectory for the deterministic dynamics. }
   \label{fig:deterministic_soln_controls}
@@
 
-\section{Open-loop Stochastic Control}
-As we have mentioned before, when the value of $X_t$ is unknown, the best one
-can do is to use the transition density to perform the optimization. Since the
-transition density follows a (deterministic) PDE, we will thus apply a Maximum
-Principle for PDEs as a method of obtaining the optimal control.
-
-\subsection{Fokker-Planck equation for the state density evolution}
-
-We will need the transition density of $X$, conditional on no
-spikes having occurred. We will write this as:
+ALTERNATIVELY:
+We can also ignore the energy cost. And just choose a constant $\a$ which will
+get us to threshold at time $t=\T$. Then the calculations are:
+Finally the solution for $x$ is:
 \begin{align*}
-f(x,t) \intd{x} &= \Prob[X_t\in \intd{x} | X_0 = .0, X_{s<t} < \xth]  								  
+\dot{x} & = \m + \a  -  \frac{x}{\tc}
+\\
+e^{t/\tc}x &=  \int^t e^{s/\tc}(\m+\a) \intd{s}   
+\\
+x &= ( \a + \m)\cdot \tc (1 - e^{-t/\tc}) 
 \end{align*}
-Then $f$ satisfies a Fokker-Planck equation with absorbing boundaries
+And so: 
+$$
+x(\T) = \xth \implies \a = \frac{\xth}{\tc(1 - e^{-\T/\tc})} - \m$$  
+Assuming that $\mu$ and $\amin,\amax$ allow for a solution to this equation.
+\subsection{Test Comparison}
+\label{sec:deterministic_vs_feedback_numerical_test}
+Let us  now make, the rather unfair, comparison of the deterministic control
+against the stochastic control.
+
+We will sample 512 realizations of the controlled system and apply either the
+deterministic or the stochastic controller. Naturally, we will apply exactly
+the same random disturbances to both controllers.
+
+We will set $\e = .001$ and $\amax = 2.0$. As such the energy cost is of
+secondary importance and the paramount effect on the objective is the difference
+$(\ts - \T)$.
+
+We show three random examples of controlled trajectories in
+\cref{fig:control_trajectories_examples} and the statistical performance of the
+two control laws is shown in \cref{fig:error_histograms_det_vs_stoch}. The
+statistical performance of the two controllers is compared in
+\cref{tab:realized_avg_errors_det_vs_stoch}. There are two important things to
+note in \cref{tab:realized_avg_errors_det_vs_stoch}: first, the difference
+between the deterministic-control-law and the stochastic-control-law error,
+row 1 vs row 2, can be thought of as the 'added value' of feedback; second,
+we see that the stochastic controller performs in practice pretty close to its
+theoretical value, that is the difference between row 2 and row 3.
+
+The most pertinent information in 
+\cref{tab:realized_avg_errors_det_vs_stoch} is that the controls we can
+implement for our problem with its feedback peculiarities can never achieve a
+lower error than what is achieved in rows 2,3 and should never produce a higher
+error than what is in row 1.
+
+\begin{table}[h] 
+\centering
+\begin{tabular}{lc}
+Control Law & Squared Error \\
+\hline
+Deterministic &  0.647 \\
+Stochastic &  0.336\\
+Theory (Value function) & 0.285
+\end{tabular}
+\caption{Realized performance of the different control laws and the theoretical
+expected performance of the stochastic law (last row)}
+\label{tab:realized_avg_errors_det_vs_stoch}
+\end{table}
+
+\begin{figure}[h]
+\begin{center}
+\subfloat[A]
+{
+\label{fig:controlled_traj_ex1}
+\includegraphics[width=0.33\textwidth]
+{Figs/ControlSimulator/example_controlled_trajectories_id1.png}
+}
+\subfloat[B]
+{
+\label{fig:controlled_traj_ex2}
+\includegraphics[width=0.33\textwidth]
+{Figs/ControlSimulator/example_controlled_trajectories_id4.png}
+}
+\subfloat[C]
+{
+\label{fig:controlled_traj_ex3}
+\includegraphics[width=0.33\textwidth]
+{Figs/ControlSimulator/example_controlled_trajectories_id3.png}
+}
+\caption[]{Examples for the controlled trajectories using both the deterministic
+and the stochastic control approaches. The red vertical line in the plots
+indicates the desired spike-time, $\T$. In A, B the performance of both
+control laws is essentially the same, but in C we see the advantage of the
+stochastic approach. $\tc, \b = [.75, 1.25]$.}
+\label{fig:control_trajectories_examples}
+\end{center}
+\end{figure}
+\begin{figure}[htp]
+\begin{center}
+  \includegraphics[width=.9\textwidth]{Figs/ControlSimulator/example_controlled_trajectories_hists.png}
+  \caption[labelInTOC]{Histogram of the spike timing error for the
+  deterministic (left) vs. the stochastic (right) control laws.}
+  \label{fig:error_histograms_det_vs_stoch}
+\end{center}
+\end{figure}
+
+\section{Uninformed Stochastic Control}
+As we have mentioned before, it is unrealistic to assume we know the value of
+$X_t$.
+
+So what should we do?
+
+Note that our main criteria - the minimization of the distance between $\ts,\T$
+- is a probabilistic criterion. Since we cannot act reactively, for each
+$X$ trajectory, the next most sensible thing to do is to optimize using the
+transition density of $X$. We will describe how, now.
+
+\subsection{Fokker-Plank equation for the state density evolution}
+We will need the conditional transitions densities of $X$, conditional on no
+spikes having occurred. We will write these as:
+\begin{align*}
+F(x,t) &= \Prob[X_t \leq x| X_0 = .0, X_{s<t} < \xth] &\textrm{Conditional
+Dist'n}
+\\
+f(x,t) &= \di_x F  								   &\textrm{Conditional Density}
+\end{align*}
+
+Note that:
+$$
+F(\xth,t) = \int^\xth f(x,t) = \Prob[X_t \textrm{ has not spiked yet}]
+$$
+
+It is well known that for a It\^o process as in \cref{eq:X_evolution_uo},
+$f$ obeys the following (forward) PDE:
 \begin{equation}
@@
 
-The probability flow, $\phi$, here is
+We will often encounter the probability flow term, $\phi$,:
 $$
@@
 $$
-This is a linear and bounded differential operator. . We will usually omit the
+This is a linear and bounded differential operator (linearity is obvious, but
+we would need to prove the boundedness) and, if we must, we will call it the
+Fokker-Plank or Forward Kolmogorov differential operator. We will usually omit the
 subscript $\a$, but have written it now, since the operator is parametrized by
 $\a$, that is by the control.
+
+With this, we are on our way to optimize using the density $f$ instead of a
+particular trajectory $X_t$.
 
@@
 J(\a) = 
-\int_\xmin^\xth \Ttwo(x) f(x,\T) \intd{x}
+\int_\xmin^\xth \Ttwo(x) f(x,T) \intd{x}
 \\
@@
 $$
-\int_\xmin^\xth \Ttwo(x) f(x,\T) \intd{x}
+\int_\xmin^\xth \Ttwo(x) f(x,T) \intd{x}
 $$
@@
 this cost is the expected squared time-to-hit starting at $x$, with  $\a
-= \amax$, weighted by the probability of $X_\T = x$ , $f(x,\T)$. Recall
-that we calculated $\Ttwo(x)$ in \cref{sec:calculating_TCs}.
+= \amax$, weighted by the probability of being at $x$, $f(x,T)$. Recall that we
+calculated $\Ttwo(x)$ in \cref{sec:calculating_TCs}.
 
@@
 counts the cost of trajectories which spike too soon, that
-is the squared difference between realized spike time, $t$ and desired spike
-time $\T$, weighted by the probability of spike at $t$ which is just the outflow of
+is the squared difference between spike time, $t$ and desired spike time $\T$,
+weighted by the probability of spike at $t$ which is just the outflow of
 probability, $\phi(\xth,t)$. Note further that due to the homogeneous
@@
 \label{sec:PDE_max_principle_for_pdf}
-By now our ostensibly stochastic optimal control problem modelled by SDEs
-has become a deterministic optimal control problem modelled by PDEs. Our control
-$\a$ influences the evolution density $f$ and via $f$, the integrals in the
+By now our ostensibly stochastic optimal control problem on SDEs
+has become a deterministic optimal control problem on PDEs. Our control $\a$
+influences the evolution density $f$ and via $f$, the integrals in the
 objective, $J$.
 
-As in the ODE theory, we will introduce an adjoint function, $p$, which will be
-crucial in setting up necessary conditions for the optimal control.
+We have already had a taste of deterministic control in
+\cref{sec:deterministic_ode_control}, but the current problem is much more
+complicated as it involves PDEs instead of ODEs. Thankfully we are still in 1-D
+so we can solve these PDEs with simple and cheap numerical methods.
+
+As in the ODE theory, we will introduce a Hamiltonian functional, $\H$, and an
+adjoint function, $p$, which will be crucial in setting up necessary
+conditions for the optimal control.
 
@@
 little strange as it involves values of $\di_x f$ at the boundary. As such we
-will derive the optimality condition heuristically, appealing to the
-references for rigorous proofs.
+will 'derive' the optimality condition heuristically, appealing to the
+references for rigorous proofs later.
 
@@
 \begin{equation}
-\di_t \left[ \int_\xmin^\xth  p \cdot\df \intd{x} \right] = (s-\T)^2 \frac
-{\b^2}{2}
-\di_x
+\di_t [ \int_\xmin^\xth  p \cdot\df \intd{x} ] = (s-\T)^2 \frac {\b^2}{2} \di_x 
 \df
@@
 \end{equation}
-which does not depend on $\df$! and we would be on our way. In order to find
+which does not depend on $\df$! and we'd be on our way. In order to find
 $p(x,t)$ we will need to
@@
 With \cref{eq:FP_variation_near_optimum} we are ready to design the dynamics of
-$p$ such that $\df$ is eliminated from the optimality equation,
+$p$ st. $\df$ is eliminated from the optimality equation,
 \cref{eq:J_variation_PDE}.
 
-\subsection{Calculating the adjoint }
-Recall our desideratum, we want
+\subsection{Calculating the adjoint(s)}
+Recall our desideratum:
 $$
@@
 $$
-\int_0^\T \di_t (p \cdot \df) = \int_0^\T(s-\T)^2 \frac {\b^2}{2} \di_x 
-\df \intd{s} + \text{stuff}
-$$
-This is equivalent to having
-$$
-\di_t (p \cdot \df) =  (s-\T)^2 \frac {\b^2}{2} \di_x 
+\int_\xmin^\xth \di_t (p \cdot \df) = (s-\T)^2 \frac {\b^2}{2} \di_x 
 \df + \text{stuff}
 $$
-Let us calculate $\di_t (p \cdot \df)$
+Well,
 \begin{align*}
@@
 \begin{gathered}
-\int_\xmin^\xth  \Lstar [p] \df \intd{x} 
+\int  \Lstar [p] \df \intd{x} 
 =
-\int_\xmin^\xth p \L [\df] \intd{x} - (s-\T)^2 \frac {\b^2}{2}\di_x \df +
-\text{possibly more stuff}
+\int p \L [\df] \intd{x} - (s-\T)^2 \frac {\b^2}{2}\di_x \df + \text{possibly
+more stuff}
 \\
@@
 \int p \L [\df] \intd{x} =&
-\int p \underbrace{\left(\frac{\b^2 }{2}\cdot \di^2_x \df -  
-			 \di_x \big[ (\a- \frac{x}{\tc})\cdot \df \big]\right)}_{\L [\df]} \intd{x}
-\\
-=& [ -p ((\a- \frac{x}{\tc})  \cdot \df)\quad  \vert^\xth_\xmin
+\int p (\frac{\b^2 }{2}\cdot \di^2_x \df -  
+									\di_x \big[ (\a- \frac{x}{\tc})  \cdot \df \big]) \intd{x}
+\\
+=& [ -p ((\a- \frac{x}{\tc})  \cdot \df)\,\vert^\xth_\xmin
 + \int \di_xp ((\a- \frac{x}{\tc})  \cdot \df) \intd{x}
 \\
-& + [p (\frac{\b^2 }{2}\cdot \di_x \df \quad \vert ^\xth_\xmin
-- [ \di_x p \cdot \frac{\b^2 }{2}  \df\quad \vert^\xth_\xmin 
+& + [p (\frac{\b^2 }{2}\cdot \di_x \df \, \vert^\xth_\xmin 
+- [ \di_x p \cdot \frac{\b^2 }{2}  \df\, \vert^\xth_\xmin 
 + \int \frac{\b^2 }{2} \di^2_xp    \cdot \df \intd{x} 
@@
 + p \frac{\b^2}{2} \di_x \df 
--\di_x p \frac{\b^2}{2} \df \,\vert^\xth_\xmin }_{\text{BC terms}}
+-\di_x p \frac{\b^2}{2} \df \, \vert^\xth_\xmin }_{\text{BC terms}}
 \end{align*}
@@
 \\
+\\
 =& -\cancel{p (\a- \frac{x}{\tc})  \cdot \df \Big|_\xth}
@@
 \\
+\\
 &- \cancel{[ \di_x p \cdot \frac{\b^2 }{2}   \df\Big|_\xth}
@@
 \\
+\\
 =& \cancelto{-p (-\da f)}{-p ((\a- \tfrac x \tc)\df - \di_x \tfrac {\b^2}2
 \df) \Big|_\xmin} 
-+ [p \frac{\b^2 }{2}\cdot \di_x f \Big|_\xth
-+ [ \di_x p \cdot \frac{\b^2 }{2}   f\Big|_\xmin
++ [p \frac{\b^2 }{2}\cdot \di_x \df \Big|_\xth
++ [ \di_x p \cdot \frac{\b^2 }{2}   \df\Big|_\xmin
+\\
 \\
@@
 \subsection{An Iterative Numerical Solution}
-We have an entangled situation: $f,p$ depend on $\a$ through
-\cref{eq:adjoint_pde_OU,eq:FP_pde_OU_PDF}, while $\a$, the optimal control, is
-given by \cref{eq:optimal_alpha_4_PDF_optimization} which requires  the
-solution to $f,p$ . So we need some kind of iteration to find $\a$ and the corresponding $f,p$.
-
-The key to such an iteration is that the quantity 
+We have an entangled situation: $f,p$ depend on $\a$, but $\a$, the optimal
+control, is given by the solution to $f,p$. So we need some kind of iteration to
+find $\a$ and the corresponding $f,p$.
+
+The key to such an iteration is that the quantity, 
 $$\di_\a \H =  \Big\{
@@
 \Big\}
-$$
-gives us the direction of increase
+$$ gives us the direction of increase
 and we can use it as a gradient in a descent algorithm. 
@@
 from \cref{sec:deterministic_ode_control} is a natural choice. Then solve for $f,p$.
-Then evaluate $J$ and iterate until convergence. For our convergence
-criterion we will take that the incremental difference in $J$ is smaller than
-some difference small tolerance. The algorithm is specified more
-precisely in Algorithm \ref{alg:hamiltonian_descent_4_PDF_OC}.
+Then evaluate $J$ and iterate until convergence. (Until $J$ stops getting smaller.)
+
+More specifically see the pseudo-code in
+Algorithm \ref{alg:hamiltonian_descent_4_PDF_OC}:
 \begin{algorithm}
@@
 \State $J_0 = \infty $ 
-\State $\a_1 = $ deterministic $\a$, the optimal control given $\b=0$ from
-\cref{sec:deterministic_ode_control}
+\State $\a_1 = $ Deterministic $\a$
 \For { $k= 1\dots k_{\max}$}
@@
 We show one realization of $\di_a \H$ in \cref{fig:grad_aH_deterministic_alpha}.
+Also, we can calculate $J$ for the deterministic $\a(t)$. We find it is
+$J[\a_{\text{det}}] = .709$, which is quite close to the experimentally observed
+$J$ in the first row of \cref{tab:realized_avg_errors_det_vs_stoch} and so we
+know we are on the right path.
+%\usepackage{graphics} is needed for \includegraphics
 \begin{figure}[htp]
@@
 \subfloat[$\a_0$]
-{\includegraphics[width=.49\textwidth]{Figs/FP_Adjoint/Deterministic_alpha_di_aH.png}
+{
+  \includegraphics[width=.49\textwidth]{Figs/FP_Adjoint/Deterministic_alpha_di_aH.png}
   }
   \subfloat[$\a_1$]
-{  \includegraphics[width=.49\textwidth]{Figs/FP_Adjoint/First_control_iteration.png}
+{
+  \includegraphics[width=.49\textwidth]{Figs/FP_Adjoint/First_control_iteration.png}
   }
-\caption[labelInTOC]{Iteratively calculating $\a$ for the
-open-loop optimal control. Plotted is an example of
-$-\di_a \H$ calculated at $\a(t) = \a_1(t)$, where $a_1$ is the deterministic control. The green curve, $-\di_a \H$
-indicates the direction that we want to change the control, if it is negative
-we decrease and if it is positive we increase it. On the right panel, b), we
-show the result of applying the incremental change by also plotting the
-resulting $\a_1(t)$. }
+  \caption[labelInTOC]{Example of $-\di_a \H$ calculated at $\a(t)$ being the
+  deterministic control, the information is that we need to decrease $\a$ at the beginning and increase it
+  towards the end (if possible due to the bound) and the resulting incremental
+  change, $\a_1$, is shown on the right. }
   \label{fig:grad_aH_deterministic_alpha}
@@
 
-Finally, we are ready to calculate the open-loop stochastic optimal control for
-this parameter set. The results are in
-\cref{fig:FP_adjoint_objective_control_convergence}.
+Finally, we are ready to calculate the open-loop, stochastic optimal control
+(for this parameter set) - the results are in
+\cref{fig:FP_adjoint_objective_control_convergence}. We see that all our work
+was not for naught:-)
 
@@
 }
-\caption[ ]{The result of an entire optimization iteration, on the left the
-iterates of the control, $\a_k(t)$, on the right the progress of the
-objective with each iteration. The final control, in purple, is inhibitory at
-the beginning of the time interval, $\a < 0$, unlike the the deterministic
-control solution. 
-We see a significant
-reduction in the objective value, $J$, and thus an improvement, on the
-order of 50\%.}
+\caption[ ]{The result of an optimization iteration, on the left the various
+controls, on the right the progress of the objective. THe final control is much
+more inhibitory at the beginning, than the deterministic control. Moreover, we
+see that there is a significant improvement in the objective (about 43\%)}
 \label{fig:FP_adjoint_objective_control_convergence}
@@
 
-% \begin{table}[h] 
-% \centering
-% \begin{tabular}{lc}
-% Control Law & Squared Error \\
-% \hline
-% Deterministic &  0.647 \\
-% Stochastic &  0.336\\
-% Theory (Value function) & 0.285
-% \end{tabular}
-% \caption{Realized performance of the different control laws and the theoretical
-% expected performance of the stochastic law (last row)}
-% \label{tab:realized_avg_errors_det_vs_stoch}
-% \end{table}
-% 
-% \begin{figure}[h]
-% \begin{center}
-% \subfloat[A]
-% {
-% \label{fig:controlled_traj_ex1}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/example_controlled_trajectories_id1.png}
-% }
-% \subfloat[B]
-% {
-% \label{fig:controlled_traj_ex2}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/example_controlled_trajectories_id4.png}
-% }
-% \subfloat[C]
-% {
-% \label{fig:controlled_traj_ex3}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/example_controlled_trajectories_id3.png}
-% }
-% \caption[]{Examples for the controlled trajectories using both the deterministic
-% and the stochastic control approaches. The red vertical line in the plots
-% indicates the desired spike-time, $\T$. In A, B the performance of both
-% control laws is essentially the same, but in C we see the advantage of the
-% stochastic approach. $\tc, \b = [.75, 1.25]$.}
-% \label{fig:control_trajectories_examples}
-% \end{center}
-% \end{figure}
-% \begin{figure}[htp]
-% \begin{center}
-%   \includegraphics[width=.9\textwidth]{Figs/ControlSimulator/example_controlled_trajectories_hists.png}
-%   \caption[labelInTOC]{Histogram of the spike timing error for the
-%   deterministic (left) vs. the stochastic (right) control laws.}
-%   \label{fig:error_histograms_det_vs_stoch}
-% \end{center}
-% \end{figure}
-
-\section{Test Comparison of Different Controls}
+\subsection{Test Comparison}
 \label{sec:probabilistic_numerical_test}
-We will now compare the performance of the three obtained controls: the
-closed-loop stochastic obtained via the HJB equation, the open-loop
-deterministic control obtained when setting $\b=0$ and the open-loop stochastic
-control obtained by optimizing using the transition density.
-
-For our comparison we will sample $128$ realizations of the controlled system
-and apply in turn each of the three controls. Naturally, for each path, we will
-apply exactly the same random disturbances to each controller.
-
-We will set $\e = .001$ and $\amax = 2.0$. As such the energy cost is of
-secondary importance and the paramount effect on the objective is the difference
-$(\ts - \T)$, which we are trying to minimize. We set $\tc, \b = [.75, 1.25]$.
-
-The performance of the three controllers is given in
-\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch}. We see that while the
-open-loop stochastic control is not as good as the feedback stochastic
-control, it is much closer in performance to it than to the blind
-deterministic control.
-
-For illustration sake, we also show the error
-histograms in \cref{fig:error_histograms_det_vs_openloop_vs_stoch} and visualize
-some trajectories in \cref{fig:control_trajectories_examples}.
+We will now run our open-loop stochastic control obtained in \cref{fig:FP_adjoint_objective_control_convergence}
+through the same comparison test as we did for the deterministic open-loop control and the stochastic
+feedback (closed-loop) control in
+\cref{sec:deterministic_vs_feedback_numerical_test}, with the same parameter
+values, $\e, \T, \amax, \beta, \tc$. From
+\cref{fig:FP_adjoint_objective_control_convergence}, we expect that our
+open-loop control should do quite better than the deterministic control, and not
+too much worse than the feedback stochastic control - let us see if that is
+indeed the case. As before we report the gross results in
+\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch}, show the error
+histograms in \cref{fig:error_histograms_det_vs_openloop_vs_stoch} and
+visualize some trajectories in
+\cref{fig:control_trajectories_examples_with_openloop}.
+
+The basic conclusion is that it was all worth it! The open-loop stochastic
+control is certainly not as good as the feedback stochastic control, but it is much closer in performance to
+it, than to the blind deterministic control.
 
@@
 \centering
-\begin{tabular}{lcc}
-Control Law & Empirical Squared Error & Theoretical Squared Error \\
+\begin{tabular}{lc}
+Control Law & Squared Error \\
 \hline
-Deterministic &  0.621 & - \\
-Open-Loop Stochastic & 0.356 & 0.382\\
-Feedback Stochastic &  0.288& 0.285\\
+Deterministic &  0.621 \\
+Open-Loop Stochastic & 0.356 \\
+Feedback Stochastic &  0.288\\
 \hline
+Openloop Theory (J) & 0.382
+\\
+Feedback Theory (Value function) & 0.285
 \end{tabular}
-\caption{Realized and theoretical performance of the different control laws. The empirical
-performance is obtained using 128 sample paths. The theoretical performance is
-found using the $J$ in for the open-loop stochastic control and the value
-function $v(x=0, t =0)$ for the closed-loop stochastic control.}
+\caption{Realized performance of the different control laws and the theoretical
+expected performance for the stochastic laws (last 2 rows)}
 \label{tab:realized_avg_errors_det_vs_openloop_vs_stoch}
@@
 \begin{center}
-\subfloat[]
+\subfloat[A]
 {
@@
 }
-\subfloat[ ]
+\subfloat[B]
 {
@@
 }
-\subfloat[ ]
+\subfloat[C]
 {
@@
 open-loop stochastic and feedback stochastic control approaches. The
-black vertical line in the plots indicates the desired spike-time, $\T$.
-For the parameter values are $\tc, \b = [.75, 1.25]$. The three panels are
-three different realizations of the model dynamics.}
-\label{fig:control_trajectories_examples}
+black vertical line in the plots indicates the desired spike-time, $\T$. $\tc,
+\b = [.75, 1.25]$.}
+\label{fig:control_trajectories_examples_with_openloop}
 \end{center}
@@
   deterministic (left) vs. open-loop stochastic(centre) vs. feedback stochastic
-  (right) control laws. This is for the same problem as in
-  \cref{fig:control_trajectories_examples}. We have used $N=128$ sample paths
-  to form the statistics}
+  (right) control laws. $N=128$}
   \label{fig:error_histograms_det_vs_openloop_vs_stoch} 
@@
 
-\section{Research Goals}
-I hope to be able to use the techniques of open-loop stochastic control
-discussed herein to address problems in stochastic control where for whatever
-reasons, the dynamic programing formalism is not applicable.
-
-One such reason might be that the state is unobservable. An example of such a
-system is the control of blood glucose for diabetes or intensive care
-patients, where the patients' blood glucose is modelled to follow a stochastic
-differential equation and actual observations are expensive and invasive and
-thus rare, on the order of a few hours. In this scenario, the applied external
-glucose or insulin to the patient in the time in-between measurements can be
-thought of as the solution to an optimal control problem where the objective is
-to minimize deviations of the patient's blood glucose from some baseline.
-
-An alternative reason to use the open-loop control techniques is when optimizing
-with multi-dimensional systems at which point the dynamic programing approach
-suffers from 'the curse of dimensionality' which is the colloquial name for the
-exponential increase in necessary computing resources. If the optimality
-conditions for the open-loop control lend themselves to a Feynman-Kac-type
-representation, then one might replace solving the PDEs by an appropriate
-Monte-Carlo process and thus make the computations tractable.
-
-\clearpage
+% \subsection{Optimizing with $f$ - the Direct Approach}
+% The adjoint approach to optimizing with PDEs is actually more appropriate when
+% we have a distributed control\ldots
+% 
+% Here we have a much more crude single-variable control - the value of $\a$ at a
+% given time. So it might be simpler to just apply the direct approach of Optimal
+% Control and discretize the system and converge to the minimum. We have not
+% explored this for now.
+
+\newpage
+% \section{Hacked reuse of the HJB control}
+% A more simple-minded approach is to use the HJB controller with an estimate,
+% $\xest$, for the state. Since there are no observations, the only sensible
+% estimate would be the mean of the process. 
+% We will write 
+% \begin{align*}
+% \mcond &= \Exp [X_t | X_{s<t} < \xth] = \int x f(x,t)
+% \intd{x} &(\textrm{Conditional Mean})
+% \end{align*}
+% then 
+% \begin{align*}
+% \Exp[X_t | X_{s<t} < \xth] &= \frac{\int_\xmin^\xth x \di_xF(x) \intd{x} }{
+% \int_\xmin^\xth  \di_xF(x) \intd{x}}
+% \\
+% &= \frac{[x F(x) |_\xmin^\xth \quad - \quad \int_\xmin^\xth  F(x) \intd{x} }{
+% F(x) |_\xmin^\xth}
+% \\
+% &= \xth - \frac{\int_\xmin^\xth  F(x) \intd{x}} {F(\xth)}
+% \\
+% &= \mcond
+% \end{align*}
+% For a consistency check, note that $\xest = .0$ at $t = 0$, since
+% $F(\xth,0) = 1.$ and $\int_\xmin^\xth H(x) \intd{x} = \xth$, as long as $\xmin <.0$, which it is.
+
+\section{Sample-based solution to Open-Loop control}
+In the previous section we have set up and numerically solved a
+state-adjoint pair of PDEs which helps us find the Optimal Control using
+a simple and efficient gradient-based iterative scheme.
+
+We now discuss how to perform the solution using trajectory samples instead of
+numericallly solving a pair of PDEs. To this end we will use the Feynman-Kac
+formula, once we realize that the adjoint PDE has a Feynman-Kac representation.
+
+\subsection{Fokker-Planck and Feynman-Kac}
+We know that the solution to the forward PDE for $f(x,t)$,
+\cref{eq:FP_pde_OU_PDF} represents the density of samples at $(x,t)$
+that start from $x=0$ at $t= 0$. 
+$$
+f(x,t) = \Prob[ X_t = x| X_0 = 0]
+$$
+
+What is less obvious is that the adjoint PDE
+for $p(x,t)$, \cref{eq:adjoint_pde_OU}, has the following stochastic
+representation:
+\begin{align}
+p(x,t) =&
+	\Exp\Big[(\ts - \tf)^2 + \e\int_0^\tf u(s) \intd{s} \,|\, X_t = x \text{
+	AND } \ts \leq \tf \Big] \notag
+\\ 	\,	&
+	\Exp \Big[\Ttwo(X_\tf) + \e\int_0^\ts u(s) \intd{s} \,|\, X_t = x \text{ AND }
+	\ts> \tf \Big] +\,
+\label{eq:adjoint_FeynmanKac_representation}
+\end{align}
+i.e. $p(x,t)$ gives the expected cost of a trajectory that goes through $x$ at
+time $t$.
+
+Recall that we need $f,p$ in order to calculate the gradient:
+$$\di_\a \H =  \Big\{
+ \tilde{\e}  2 \a(t)
++ p(x,t) f(x,t) \Big|_\xmin
+- \int _\xmin^\xth p \cdot \di_x f(x,t) \intd{x} 
+\Big\}
+$$
+If we use samples, we do not need to artificially truncate the domain from below
+so we can drop the $p(x,t) f(x,t) \Big|_\xmin$ term. And we just need to
+calculate either:
+$$ - \int_{x < \xth} p(x,t) \cdot \di_x f(x,t) \intd{x}$$
+or, equivalently, by integrating-by-parts:
+$$   \int_{x < \xth} \di_x p(x,t) \cdot f(x,t) \intd{x}$$
+If we sample then this is just an expectation with respect to $X$ and can be
+written as: 
+\begin{equation}
+\int_{x < \xth} \di_x p(x,t) \cdot f(x,t) \intd{x} = \Exp[\di_x p(X,t) ]
+\approx 1/N\sum_i \di_x p(X^{(i)}_t,t) 
+\label{eq:f_sampled_vs_f_solved}
+\end{equation}   
+where $\{ X^{(i)}_t \}_{i=1}^N$ are the samples. This expression would be
+particularly useful if we had a closed form expression for $\di_x p$, but that
+is usually not the case.
+
+We show a comparison of the two ways of computing
+\cref{eq:f_sampled_vs_f_solved} simulation vs. PDE in
+\cref{fig:f_sampled_vs_f_solved}. We see a ball-park agreement and of course,
+the simulation-based approximation is noisy.
+% \usepackage{graphics} is needed for \includegraphics
+\begin{figure}[htp]
+\begin{center}
+  \includegraphics[width=.9\textwidth]{Figs/FeynmanKac/fsampled_psolved_t=8_b=12.png}
+  \caption{Simulation vs.\ the PDE mode of computing $\di_\a \H$.
+  Recall $p$ is solved using PDE numerics, but $f$ is simulated using trajectory
+  samples. Thus, $\int \di_x p f $ at time $t$ is approximated as $1/N
+  \sum_{X_t < 1} \di_x p(X_t)$.}
+  \label{fig:f_sampled_vs_f_solved}
+\end{center}
+\end{figure}
+
+If we would like to dispense with solving for $p$ as well then we need a way to
+compute $\di_x p$ from the samples. As we said, the Feynman-Kac formula gives us
+a stochastic representation for the solution of a PDE. But we need the
+partial derivative of that solution with respect to $x_t$\ldots It's not obvious
+what to do, we can obtain a PDE for $\psi = \di_x p$, by just differentiating
+the $p$ PDE and that will give us a Feynman-Kac type evolution, but we don't
+know the right BCs. We would know the TCs: $\psi(\tf) = \di_x \Ttwo(x)$, but we
+don't know what to do at the threshold boundary $x=\xth$\ldots and using a
+finite difference approximation would be proxibitive, b/c then for every point
+in space-time we would need to start a bunch of simulations. Maybe we shold go
+back to trying to evaluate $\int \di_x f \cdot p$ and look for sample-based
+approximations of $\di_x f$?
+
+\section{Spike Response Model}
+An alternative model to the hard-threshold LIF model is the Spike Response
+Model. The difference is that here a neuron spikes with instantaneous rate
+$\lambda$:
+where usually
+$$
+\lambda(x) \propto exp(x)  
+$$
+that is, we want $\lim_{-\inf} \l = 0$ and $\lim_{\inf} \l = \infty$, meaning
+that for very low voltages spikes are impossible and for very  high voltages,
+spikes are inevitable and instantaneous.
+
+With this the Fokker-Planck PDE becomes:
+
+\begin{equation}
+\begin{gathered}
+\di_t f(x,t) =
+				\frac{\b^2 }{2}\cdot \di^2_x f -  
+				\di_x \Big[ (\a(t)- \frac{x}{\tc})  \cdot f \Big]
+				-\l(x) f 
+\\
+\\
+\begin{array}{lll}
+	&\textrm{st.\ }& 
+	\left\{ \begin{array}{lcl}
+	 f(x,0) &=& \delta(x) \quad (\textrm{delta function}
+	\\
+	(\a(t)- \tfrac x \tc)f - \di_x \tfrac {\b^2}2 f] \big|_{x=\xmin} &\equiv& 0
+	\quad (\textrm{reflecting BCs at some } \xmin )
+	\\
+	(\a(t)- \tfrac x \tc)f - \di_x \tfrac {\b^2}2 f] \big|_{x=\xmax} &\equiv& 0
+	\quad (\textrm{reflecting BCs at some } \xmax )
+\end{array} \right. 
+\end{array}
+\label{eq:FP_pde_OU_PDF_SRM}
+\end{gathered}
+\end{equation}
+The biggest difference between \cref{eq:FP_pde_OU_PDF_SRM} and
+\cref{eq:FP_pde_OU_PDF} is the addition of the killing term $-\l f$ and the
+different domain and boundary conditions.  For the Spike Response Model, we need
+a bigger domain, since the voltage can become arbitrary high, but the decay term
+will pull it back towards $\a$, meaning by symmetry we should have $\xmax +
+\xmin = 2\alpha$.
+
+Now spikes happen throughout the domain (with frequency $\l f$) and thus the
+objective function $
+J(\a) = \Exp\left[
+(\ts - \T)^2 + \e \int_0^\ts  \a^2(s) \intd{s}
+\right]
+$ in terms of $f$ now reads:
+\begin{align}
+J(\a) = &
+\int_\xmin^\xmax \Ttwol(x) f(x,T) \intd{x}
+\\
+&+ \int_0^\ts (s-\T)^2  \int_\xmin^\xmax \l(x) f(x,s) \intd{x}\intd{s}
+\notag
+\\
+&+  \e \int_0^\ts  \a^2(s)  \int_\xmin^\xmax f(x,s) \intd{x} \intd{s}
+\notag
+\label{eq:OC_LS_variance_density_SRM}
+\end{align}
+here we have $\Ttwol(x)$ as the squared time to spike in the soft-threshold
+framework starting at $x$. This is, of course, different from $\Ttwo(x)$
+which is the squared time to spike in the hard-threshold model.
+
+In fact one can see that the soft-threshold includes the hard-threshold in the
+special case that $\l(x) = \delta (x-\xth)$.
+
+\subsection{Calculating  $\Ttwol(x)$}
+OMIT FOR NOW, assume it can be dones as easily as $\Ttwo(x)$.
+
+\subsection{The SRM Adjoint}
+We will skip deriving the BCs for the adjoint $p$ and just state the PDE:
+\begin{equation}
+\begin{gathered}
+\begin{aligned}
+\di_t p(x,t) =& - [ \Lstar[p] + (s-\T)^2 \l(x) ]   
+\\
+		=&  
+			-\Big[ \frac{\b^2 }{2}\cdot \di^2_x p +  
+				(\a(t)- \frac{x}{\tc})  \cdot \di_x p 
+				- \l(x) p
+				+ (s-\T)^2 \l(x) 
+			\Big] 
+\end{aligned}
+\\
+\st  
+\begin{cases}
+	p(x,\T) &= \Ttwol(x)
+	\\ 
+	??? \di_x p  \big|_{x=\xmin} &\equiv 0 ???
+	\\
+	??? \di_x p  \big|_{x=\xmin} &\equiv 0 ???
+\end{cases}
+\label{eq:adjoint_pde_OU_SRM}
+\end{gathered}
+\end{equation}
+
+Optimal principles and objective gradients follow exactly as before:
+
+\subsection{Feynman-Kac for the SRM Model}
+Recall the difficulty we had with the hard-threshold model and implementing the
+Feynman-Kac formula. In particular, since we needed the value of $\di_x p =:
+\psi$ we were stuck.
+
+However, with the soft-threshold formula we have:
+$$
+\di_x \di_t p(x,t) = 
+			-\di_x \Big[ \frac{\b^2 }{2}\cdot \di^2_x p +  
+				(\a(t)- \frac{x}{\tc})  \cdot \di_x p 
+				- \l(x) p
+				+ (s-\T)^2 \l(x) 
+			\Big] 
+$$
+so $\psi = \di_x p$ obeys the following F-K type equation:
+$$
+\di_t \psi  
++ \frac{\b^2 }{2} \di^2_x \psi 
++ \left(\a(t)- \frac{x}{\tc}\right) \di_x \psi
+- \frac{1}{\tc}    \psi
+-  \l(x) \psi
+- \boxed{\l'(x) p}
++ (s-\T)^2 \l'(x)
+= 0
+$$
+Ooops! We still have a $p$ term in there, which we have to calculate
+separately\ldots but we can do that again with a F-K type expectation. For the
+$\psi$  TCs, we have:
+$$
+ \psi(\tf) = \di_x \Ttwol(x)
+$$
+ Let us rewrite the PDE for $\psi $ in a standard form:
+ $$
+ \di_t \psi  
+ +  U(x,t) \di_x \psi 
+ +  D(x,t) \di^2_x \psi
+ - \underbrace{V(x,t)}_{ 1 / \tc + \l(x)} \psi 
+ + \underbrace{f(x,t)}_{( (s-\T)^2  - p) \l'(x)} = 0
+ $$
+
+% $$
+% \di_t p(x,t) + \Big[ \frac{\b^2 }{2}\cdot \di^2_x p +  
+% 				(\a(t)- \frac{x}{\tc})  \cdot \di_x p 
+% 				- \l(x) p
+% 				+ (s-\T)^2 \l(x) 
+% 			\Big] = 0
+% 			$$
+then we can evaluate $\psi$ as conditional expectations\footnote{I need a
+better reference than: WP \url{http://en.wikipedia.org/wiki/Feynman-Kac_formula}}:
+$$
+\psi(x,t) = 
+\Exp \left[
+\int_t^\tf  f(X_s, s) e^{-\int_t^sV(X_\t,\t) \intd{\t}}
+\intd{s}
++
+e^{-\int_t^\tf V(X_\t,\t) \intd{\t}} \psi(X_\tf)) | X_t = x
+\right] $$ 
+where the expectation is done over the RV $X_t$  with the dynamics: $$
+dX = U(x,t)dt + \sqrt{2D} \intd{W}
+$$
+which are exactly the same dynamics as the ones for the forward process. This
+means that the same trajectory can be used to compute both the distribution $f$
+and the adjoint and its derivative $p$ and $\psi$. 
+This means that we can perform a simulation for $X$ and then unroll the
+calculation backwards to obtain $p, \psi$ and 
+$\int_x f(x,t) \di_x p(x,t) \intd{x} \approx \sum_i \psi(X_t^{(i)} )$
+
+% \section{Refractoriness}
+% Our models have so far not included an expl
+\section{The multivariate case}
+Our methods are only worth anything if we can scale them to the multivariate and
+indeed to the massive multivariate case, with more than 10 neurons, i.e\
+states. This makes the direct pde approach very difficult, since a PDE
+
+\section{The Big Questions}
+Where are we going with all this?
+
+So far, we have a cleanly set-up problem. It has been numerically solved for a
+limited range of parameters and there is no obvious difficulty in extending the
+solution scheme to the entire relevant parameter range. A curious challenge
+would be to what extent the implementation can be pushed to be 'real-time',
+i.e.\ whether this optimization can be done online with neurons spiking at
+frequencies on the order of 20-50 kHz (inter-spike intervals of 20-50 ms.), at
+least in {\sl silico}.
+
+A much more ambitious goal is to see whether this can be done for many neurons
+(say $N=25$) which have coupled control/dynamics. This is very challenging,
+because a naive application of most of the techniques used so far will result in
+us having to solve a 25-dimensional PDE and that is essentially
+impossible\footnote{with 10 nodes discretizing each neuron state we will need
+$10^{16}$ Gbytes to store one time-step of $v$ so, for one, we won't have enough
+memory}. Thus PDE methods cannot be directly applied and either some kind of
+independence between the neurons needs to be assumed or a different control
+technique should be applied - perhaps, and very speculatively, a path-integral
+control approach as in \cite{Kappen2007,Theodoru2011}.
+
+Note that $N=26$ was accomplished in \cite{Ahmadian2011}, but they used a
+different spiking model, which allowed for certain analytical tricks to greatly
+simplify the mathematics. Perhaps a conclusion that will come out of this whole
+research is that one should use their models (known as Spike Response
+Models) and not the hard-threshold LIF model if one wants to do optimal control
+over spike trains.
+
+We don't know until we try.
+
+% \begin{quote}
+% \raggedleft
+% Life can only be understood backwards; but it must be lived forwards.
+% \\
+% \textsl{Soren Kierkegaard}
+% \end{quote}
+
+\newpage
 \appendix
@@
 
-The following is almost a direct transcription of A.1 in
+The following is almost a direct transcription A.1 in
 \cite{Evansb}, with the added simplification of linear dynamics
-and linear cost and with the notation adapted to what we have used
-hereto.
+and linear cost and adapted to our notation.
 
@@
 $$
-where $A$ is a matrix in $\R^{n_x \times n_x}$ whose values depend on the
-time, $t$ and the control $\a$. The cost, $J$, is given by:
-$$
-J[\a] = \int_0^\tf L(t,\a) \cdot x \intd{s} + M \cdot x(\tf)
-$$
-where $L(t,\a), M$
-are vectors in $\R^{n_x}$. Both the dynamics and the cost are linear in the state $x$.
+where A is a matrix.
+The cost is given by:
+$$
+J[\a] = \int_0^\tf r(t,\a) \cdot x \intd{s} + \psi \cdot x(\tf)
+$$
+
 Suppose $x,\a$ are the optimal trajectory - then small variations in $\a$ can
-only worsen $J$. Let such an acceptable variation be given by:
+only worsen $J$.
+
+Let such an acceptable variation be given by:
 $$
@@
 $$
-for some small $\e$. We will have a corresponding dynamic, $x_\e$ st:
+for some small $\e$.
+
+We will have a corresponding dynamic, $x_\e$ st:
 $$
@@
 $$
-Note that $$\dx(0) = 0$$ since $x(0)$ is specified and so there is no
-variation there.
+Note that $\dx(0) = 0$ and that this is also a linear equation. (Even if $x$ had
+non-linear dynamics, $\dx$ would satisfy a linear ODE).
 
@@
 \frac {dJ}{d\e} \Big|_{\a} = \lim \frac{ J[\a_\e] - J[\a]}{\e} 
-= \int L(\a,t) \dx +  \grad_\a L \cdot \b \intd{s} + M \dx(\tf)
+= \int r(\a,t) \dx +  \grad_\a r \cdot \b \intd{s} + \psi \dx(\tf)
 \geq 0 
@@
 
-'Clearly', we want to replace the term $ M \cdot \dx $ first. Thus we set 
-$$
-p(\tf) = M
+'Clearly', we want to replace the term $ \psi \cdot \dx $ first. Thus we set 
+$$
+p(\tf) = \psi
 $$
@@
 Now notice the following:
-$$p(\tf)\dx(\tf) = \cancelto{0}{p(0)\dx(0)} + \int_0^\tf d_t (p \cdot \dx)$$
-
-So if we make $\int d_t (p \cdot \dx) = - \int r(\a,t) \dx$ we would be
-eliminate the $\dx$ terms in the expresion for $dJ/d\eps$ entirely!
+$$p(\tf)\dx(\tf) = \cancelto{0}{p(0)\dx(0)} + \int d_t (p \cdot \dx)$$
+
+So if we can make $\int d_t (p \cdot \dx) = - \int r(\a,t) \dx$ we would be set!
 
@@
 $$ 
-\dot{p} = -A^T p - L(\a,t)
+\dot{p} = -A^T p - r(\a,t)
 $$

### 2013-08-22 phd_thesis_plan.tex
Initial commit

\documentclass{article}

\usepackage{amsmath}
\usepackage{amscd}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm} 
\usepackage{amsfonts}
\usepackage{amsthm}

\usepackage{circuitikz}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows,snakes,backgrounds}
% \usetikz
\usepackage{subfig}

\usepackage[super]{nth}
% \usepackage{appendix}
% \usepackage{listings}
% \usepackage{color}

\usepackage{hyperref}
%\usepackage{url}

\usepackage{cleveref}

\usepackage{aviolov_style}  
\usepackage{local_style}

\begin{document}

\title{Math Thesis Plan}
\author{alexandre iolov\\ $<$\href{mailto:aiolo040@uottawa.ca}
		{aiolo040 at uottawa dot ca}$>$
		}

\date{\today}

\maketitle

\tableofcontents

\section{Thesis Proposal}
I propose we call the thesis:

{\large \sl ``Estimation and Control in Single Neural Models''}

I propose that the thesis consists of three distinct sections

\begin{enumerate}
  \item Our estimation scheme for the sinusoidal LIF problem 
  \item The spike time optimal control for a noisy LIF 
  \item A discustion on Optimal Design which combines elements from the two
  studies above. That is the control is used to obtain better estimates.
\end{enumerate}

\section{Timelines / Milestones}
I would like to submit my thesis ideally by {\sl January 22nd, 2014}. This is
the last day to do so without signing up for the winter semester.

Failing that, I would aim for {\sl May 22nd, 2014}

What kind of milestones / schedule would make it likely that one of the
above happens?

\end{document}

### 2013-08-22 short_description.tex
Initial commit

--- 2013-08-22 CompExam.aviolov.tex
@@
 
-\title{Stochastic Optimal Control - with a view towards spike control in neurons
-\\
-{\small Advanced Comprehensive Exam in Mathematics}
-\\
-{\small Dept. of Math - University of Ottawa}
+\title{stochastic optimal control for spike control in single neurons
+\\
+---
+\vskip5pt
+summary
 }
@@
 $<$\href{mailto:aiolo040@uottawa.ca}
-		{aiolo040 at uottawa dot ca}$>$}
-
-\date{January 10, 2013}
+		{aiolo040 at uottawa dot ca}$>$, 
+		Andr\'e Longtin}
+
+\date{\today}
 
@@
 
-\abstract{We review main concepts from stochastic optimal control theory,
-including stochastic differential equations, dynamic programing and Pontryagin's
-Maximum Principle. We then discuss an optimal control problem in a simple
-mathematical model of a neuron where the feedback is minimal and we have a
-quasi-open-loop control which necessitates the use of a Maximum Principle for
-the system's transition density.}
+\abstract{We pose an optimal control problem to precisely specify the spike
+timing in a noisy leaky-integrate and fire model of a neuron. The optimal
+control problem is solved using Dynamic Programming if the controller has access
+to the voltage (closed-loop control) and using a Maximum Principle for the
+transition density if the controller only has access to the spike timings (open
+loop control)}
 
@@
 
-\section{Overview}
-Stochastic Optimal Control in continuous time deals with finding
-the input which optimizes the performance of a continuous-time dynamical systems
-that contain a significant random component in their evolutions.
-
-Here we present a brief overview of the subject starting from the basic notions
-of Stochastic Differential Equations (SDEs), 
-and then introducing the main tools of Optimal Control, namely the
-Hamilton-Jacobi-Bellman equation of Dynamic Programming and the state-adjoint
-equations arising from Pontryagin's Maximum Principle. We will not state
-proofs, although we sometimes provide heuristic derivations of the main
-results.
-
-If I had to learn the material all over again, I would proceed my readings as
-follows: First I would read the online notes by Evans on both
-SDEs,\cite{Evansa}, and Optimal Control, \cite{Evansb}. I would then read
-Jacobs' book, \cite{Jacobs}, especially for the chapter on the Fokker-Planck
-equation (ch 7) and then Oksendal \cite{Oksendal2007} for a
-more thorough mathematical treatment of SDEs. Having familiarized myself with SDE's I would
-return to Optimal Control in the Deterministic Case, starting with \cite{Kirk2004} for the
-finite dimensional deterministic case (ODEs) and then \cite{Fleming1975} for the
-extension to the stochastic case. I would keep the following for reference \cite{Krylov2008},
-\cite{Hanson2007} and finally \cite{Fattorini1999} for the
-deterministic infinite-dimensional (PDEs) case.
-
-I have not used any particularly sophisticated numerical
-methods, but, for example, Hanson \cite{Hanson2007} discusses both the
-Euler-Maruyama scheme for simulating SDEs and the Crank-Nicholson Method for
-solving PDEs which is all I have used in my codes.
-
-All of the theoretical results presented in the sequel will be from
-the above references.
-
-\section{Stochastic Differential Equations}
-In view of our ultimate goals, we will restrict ourselves to  stochastic
-processes whose sample paths have continuous paths, i.e.\ to SDEs driven by
-Brownian motion.
-
-Since the reader is more likely well familiar with the material in this section,
-we will not provide proofs, but only state the results with a view towards
-establishing the notation for the sequel.
-
-We will assume that the reader is familiar with the following concepts:
-\begin{itemize} 
-  \item a probability space, $\{\O, \sAlg, P\}$ consisting of a
-  probability space, a sigma-algebra and a probability measure
-  \item a continuous-time stochastic process, $X_t$
-  \item a filtration $\Fil(t)$ and in particular the filtration generated by a
-  stochastic process.
-\end{itemize}
-
-\subsection{The Wiener Process and the Ito Integral}
-The Wiener Process is the fundamental building block of the stochastic calculus,
-it is often called Brownian Motion and we will denote it $\{W_t\}_{t\geq 0}$. It
-satisfies:
-\begin{defn}Wiener Process, $W_t$:
-\begin{enumerate}
-  \item $W_0 = 0$
-  \item $W_t - W_s = N(0, |t-s|)$ , i.e. normally distributed with mean 0,
-  variance $|t-s|$)
-  \item $\forall \{t_i\}_1^N, \quad \{W_{t_i} - W_{t_{i-1}} \}_2^N \sim$
-  independent, i.e $W_t$ has independent increments
-\end{enumerate}
-\end{defn}
-The fact that the finite incremental distributions suffice to specify a unique
-continuous-time stochastic process is known as Kolmogorov's extension theorem.
-\begin{ex}[from \cite{Evansa} \#21.] Inverted Wiener Process:
-\\
-Let $W_t$ be a 1-d Wiener Process and define:
-\def \Wbar {{\bar{W} }}
-\begin{equation*}
-\Wbar(t) := \begin{cases}
-t W(1/t) & t >0
-\\
-0 & t = 0
-\end{cases}
-\end{equation*}
-Show $\Wbar(t)- \Wbar(s) \sim N(0, t-s)$.
-
-\begin{proof}[Soln]
-\begin{align*}
-\Wbar(t) - \Wbar(s) =& t W(1/t) - s W(1/s), \, t>s
-\\
-=& (t-s)W(1/t) + s W(1/t) - sW(1/s) & \text{recall:} 1/s > 1/t
-\\
-=& (t-s)N(0, 1/t) - sN(0, 1/s -1/t) & \text{by indnt increments}
-\\
-=& N\left(0, \frac{(t-s)^2}t + \frac{s^2 }{ 1/s -1/t} \right) & \text{
-addition of indnt normal RVs.}
-\end{align*}
-With some algebra, the variance in the last expression reduces to $t-s$
-\end{proof}
-\end{ex}
-
-Now, we collect a few more relevant properties of $W_t$:
-\begin{enumerate}
-  \item The sample paths $W_{[0,\infty)}(\o)$
-are almost surely (a.s.) continuous and are in fact Holder continuous for any
-exponent $\g < 1/2$
-\item The sample paths $W_{[0,\infty)}(\o)$ are nowhere differentiable
-\item $W_t$ is a Markov process: $\Prob[W_t \in B | \s(W_{s' \leq s })] =
-\Prob[W_t \in B \,| \,W_s]$, for any Borel set, $B \subset \R$, where $\s(W_{s'
-\leq s })$ is the filtration generated by the $W_t$
-\end{enumerate}
-
-We now turn to defining stochastic integrals based on the Wiener Process.
-
-\begin{defn} Progressively Measurable Functions:
-
-Let $\Fil_t$ be the filtration generated by the Wiener Process.
-
-Let $X_t$ be a stochastic process which is $\Fil_t$-measurable $\forall t$ and
-which is jointly measurable in $(t,\o)$. We call such an $X_t$
-\emph{progressively measurable}
-\end{defn}
-
-\begin{defn} $\Ltwopm, \Lonepm$
-
-We define $\Ltwopm[0,T]$ as the space of all progressively measurable
-$X$ such that
-\begin{equation*}
-\Exp \left[ \int_{[0,T]} X_t^2 \intd{t} \right] < \infty
-\end{equation*}
-
-Similarly, we define $\Lonepm[0,T]$ as the space of all progressively measurable
-$X$ such that
-\begin{equation*}
-\Exp \left[ \int_{[0,T]} X_t \intd{t} \right] < \infty
-\end{equation*} 
-\end{defn}
-
-$\Ltwopm$ will be the class of functions for which the Ito integral is
-well-defined. 
-\begin{defn} Ito Integral:
-\label{defn:ito_integral}
-
-Let $P^n := {a = t^n_1 \ldots t^n_{m_n} = b}$ be a partition of the interval
-$[a,b] \subset [0, \infty)$. Let $|P_n| = \sup_i|t_i - t_{i-1}|$. 
-Take $|P_n| \rightarrow_n 0 $ and consider an $ X_t \in \Ltwopm[a,b]$,
-
-then
-\begin{equation}
-\int_{[a,b]} X_t \intd{W} := \lim_{n \rightarrow \infty}  
-\sum_{i=1}^{m_n} X_{t_i}\left( W(t_{i+1}) - W(t_{i})\right)
-\end{equation}
-
-\end{defn}
-
-To be precise, our definition is actually a theorem, and the real
-definition is one that uses step functions and passes to the limit.  Also we
-will write the limits of integration $\int_0^T \cdot  \intd{W}$ or $\int_{[0,T]}
-\cdot  \intd{W}$ interchangeably.
-% To justify def'n \ref{defn:ito_integral}, we recall:
-% \begin{lemma}[Quadratic Variation]
-% \begin{equation}
-% \lim_{n \rightarrow \infty} \sum_{i=1}^{m_n} \left( W(t_i) - W(t_{i-1})\right)^2
-% = b-a \quad \text{in }  L^2(\O)
-% \end{equation}
-% \end{lemma}
-% The above can be read, heuristically as $\lim \sum (\Delta W)^2 \rightarrow \sum
-% \Delta t$, which is the meaning behind the colloquial $\intd{W} =
-% \sqrt{dt}$.
-
-Again, we state without proof a few interesting properties of $\int X \intd{W}$:
-\begin{thm} Ito Integral Properties
-
-\begin{enumerate}
-  \item $\Exp[\int_0^T X \intd{W} ] = 0$ 
-  \item $\Exp[\left(\int_0^T X \intd{W}\right)^2 ] = \Exp[\int_0^T X^2
-  \intd{t}]$
-  \item $I(t) = \int_0^t X \intd{W} $ is a martingale 
-  \item $I(t) = \int_0^t X \intd{W} $ has continuous sample paths
-\end{enumerate}
-\end{thm}
-
-\subsection{Ito SDEs and Ito's Lemma}
-\begin{defn} an Ito SDE:
-
-We write
-\begin{equation}
-dX_t = Fdt + G dW
-\end{equation}
-on $0 \leq t \leq T$, if $X$ is a real-valued stochastic process satisfying:
-\begin{equation*}
-X(r) = X(s) + \int_{[s,r]} F \intd{t} + \int_{[s,r]} G \intd{W}
-\quad 0\leq s \leq r \leq T
-\end{equation*}
-for some $F \in \Lonepm[0,T]$ and $G \in \Ltwopm[0,T]$
-\end{defn}
-
-We are now ready to present the celebrated Ito Lemma which is the chain-rule of
-stochastic calculus:
-
-\begin{thm}[Ito Lemma] Suppose $X$ satisfies the stochastic differential $dX_t =
-Fdt + G dW$ as above and take $u(x,t) \in C^{2,1}[ \R \times [0,T]]$.
-
-Set $Y(t) = u(X,t)$
-\\
-then
-$$
-dY =  \left( \di_t u + \di_x u \cdot F + \di^2_x u \cdot \frac{G^2}2 \right)
-\intd{t} + \left(   \di_x u\cdot G  \right)\intd{W}
-$$
-\end{thm}
-
-% We now present two examples, one fundamental and one more involved:
-% \begin{ex}[pg 67 in \cite{Evansa}] The 'Stochastic' Exponential:
-% 
-% Let $Y = \exp(\l W - \l^2 t / 2)$. In the notation of the above Thm, we have:
-% $dX = dW, F=0, G=1$ and $u(x,t) = \exp(\l x - \l^2 t / 2)$.
-% 
-% then
-% \begin{align*}
-% dY =& 
-% \left( \di_t u + \di_x u \cdot F + \di^2_x u \cdot \frac{G^2}2 \right)\intd{t} +
-% \left(   \di_x u \cdot G  \right)\intd{W}
-% \\
-% =&
-% \left( -\frac{\l^2 t}{2}\exp(\l x - \l^2 t / 2) + \l^2\exp(\l x - \l^2 t / 2)  \right)\intd{t}
-% \\ 
-% &+ \left(   \l \exp(\l x - \l^2 t / 2) \right)\intd{W}  
-% \\
-% =&  \l \left( \exp(\l x - \l^2 t / 2) \right)\intd{W}  
-% \\
-% =& \l Y dW
-% \end{align*}
-% \end{ex}
-% 
-% \begin{ex}[pg. 67 in \cite{Evansa}] Hermite Polynomials:
-% 
-% Let $h_n(x,t) = \frac{(-t)^n}{n!} e^{x^2/2t} \di^n_x \left( e^{x^2/2t} 
-% \right)$ be the $n$th Hermite polynomial
-% \\
-% then
-% $$
-% \int_0^t h_n(W, t) \intd{W} = h_{n+1}(W(t), t)) 
-% $$
-% i.e. in Ito calculus, the Hermite polynomials are like the power monomials
-% $\frac{t^n}{n!}$ in the ordinary calculus.
-% \begin{proof} Since
-% $$
-% d_\l^n [\exp(- \frac{(x-\l t)^2}{2t} ] |_{\l=0} =
-%  (-t)^n d_x^n [ \exp(-\frac{x^2}{2t})] $$
-% We will have:
-% \begin{align*}
-% d_\l^n [\exp(\l x - \l^2 t / 2) ] |_{\l=0} &= (-t)^n  e^{x^2/2t} d_x^n [
-% \exp(-\frac{x^2}{2t})]
-% \\
-% &= n! h_n(x,t)
-% \end{align*}
-% Thus we can expand $\exp(\l x - \l^2 t / 2) $ in a Taylor series at $\l=0$ as:
-% $$
-% \exp(\l x - \l^2 t / 2) = \sum_n \l^n h_n(x,t)
-% $$
-% Now recall $Y$ from the previous example:
-% \begin{equation*}
-% Y(t) = \exp(\l W(t) - \frac{\l^2 t }2) = \sum_n \l^n h_n(W(t),t)
-% \end{equation*}
-% which satisfies:
-% \begin{equation*}
-% Y(t) = 1 + \l \int Y \intd{W}
-% \end{equation*}
-% Naturally we will plug the series expansion in the SDE to obtain:
-% \begin{align*}
-% \sum_{n=0}^{\infty} \l^n h_n(x,t) =&
-% 1 + \l \int_0^t  \sum_{n=0}^{\infty} \l^n
-% h_n(x,t)\intd{W}
-% \\
-% =& 1 + \sum_{n=1}^{\infty} \l^n\int_0^t 
-% h_{n-1}(x,t)\intd{W} \quad {\textrm{bump $n$ up by one}}
-% \end{align*}
-% Equating the coefficients for $\l^n$ on both sides leads to the result
-% \end{proof}
-% \end{ex}
-
-Finally, we are in a position to define an Ito SDE:
-
-\begin{defn}[Ito SDE] Let
-$F(x,t):\R\times [0,T]\ra \R,  G(x,t):\R\times [0,T]\ra \R,$ be given functions.
-We say that a stochastic process $X$ satisfies:
-\begin{equation}
-dX =F dt + G dW
-\end{equation} 
-over $[0,T]$ if
-\begin{enumerate}
-  \item $X$ is progressively measurable wrt. $\Fil_t$
-  \item $F(X,t) \in \Lonepm$
-  \item $G(X,t) \in \Ltwopm$ 
-  \item $X_t = X_0 + \int_0^t F(X_s, s) \intd{s} + \int_0^t G(X_s, s) \intd{W_s}$
-\end{enumerate}
-\end{defn}
-
-% \begin{ex} The solution to 
-% $$
-% \begin{cases}
-% dX = F(t)X dt + G(t) X dW
-% \\
-% X_0 = 1
-% \end{cases}
-% $$
-% is 
-% $$
-% X_t = \exp\left( \int_0^t F(s) - \frac{G^2(s)}{2} \intd{s} + \int_0^t G
-% \intd{W} \right) $$
-% \begin{proof}
-% Let $Y = \int_0^t f(s) - \frac{g^2(s)}{2} \intd{s} + \int_0^t g
-% \intd{W} $
-% and guess that $X = e^Y$
-% then, using Ito's lemma:
-% \begin{align*}
-% dX =& (\di_y [e^y] dY + \frac{1}{2}\di^2_y[e^y]  g^2 dt
-% \\ =&
-% (e^y  f dt - e^y g^2 / 2) dt + e^y g dW + \frac{1}{2} e^y \cdot  g^2 dt
-% \\ =& e^y \left( f dt +g dW \right)
-% \\ =& X \left( f dt +g dW \right)
-% \end{align*}
-% \end{proof}
-% \end{ex}
-
-% \begin{ex}[Brownian Bridge, pg 80 in \cite{Evansa}] The solution of the SDE
-% $$
-% \begin{cases}
-% dB = -\frac{B}{1-t} dt + dW
-% \\
-% B_0= 0
-% \end{cases}
-% $$
-% is
-% $$
-% B(t) = (1-t) \int_0^t \frac{1}{1-s} dW
-% $$
-% \begin{proof}
-% It is not clear how to apply Ito's lemma to this problem, so instead we just
-% directly calculate the differential:
-% \begin{align*}
-% B(t^+) - B(t) =& (1-t^+) \int_0^{t^+} \frac{1}{1-s} dW - (1-t) \int_0^t
-% \frac{1}{1-s} dW
-% \\ =&
-% (1-t^+) \int_t^{t^+}\frac{1}{1-s} dW  - (t^+-t) \int_0^t
-% \frac{1}{1-s} dW
-% \\ \approx&
-% \int_t^{t^+} dW - \Delta t \frac{B(t)}{1-t}  
-% \end{align*}
-% \end{proof}
-% \end{ex}
-% Actually we can say a bit more about the Brownian bridge, in particular we
-% will calculate $\Exp[B^2]$
-%  \begin{ex}[Evans .37]
-% \begin{align*}
-% \Exp[B^2] =& \Exp\left[\left( (1-t) \int_0^t \frac{1}{1-s} dW \right)^2 \right]
-% \\
-% =&(1-t)^2 \Exp\left[ \int_0^t\left(  \frac{1}{1-s}\right)^2 \intd{s} \right]
-% \\
-% =&(1-t)^2 [ \frac{1}{1-t} - 1 ]
-% \\
-% =&(1-t) - (1-t)^2
-% \end{align*}
-% 
-% Further note that  $\lim_{t\ra 1^-}\Exp[B^2] = 0$ and that $B$ is a martingale
-% and so its first moment is always equal to its initial value, $0$. So indeed the 2nd
-% moment is the variance and with the variance converging to zero, we can use
-% Chebyshev's inequality to conclude that $\lim_{t\ra 1^-}B(t) = 0 $(a.s). This
-% justifies the name of the Brownian bridge - it is a normally distributed RV.
-% which is clamped to equal 0 at both $t= \{0,1\}$.
-% \end{ex}
-
-As an example we will discuss the Ornstein-Uhlenbeck process, which is the
-basis for the models we will face later on.
-\begin{ex}[O-U Process] Let $X$ follow:
-\begin{equation*}
-dX = \th (\a - X_t) dt + \b dW
-\end{equation*}
-with an initial condition, $X_0$, which may be an arbitrary distribution
-independent of the Wiener Process,$W$.
-We can solve this as follows:
-\begin{align*}
-dX =& \th (\a - X_t) dt + \b dW
-\\
-dX + \th X_t) dt=& \th \a dt + \b dW
-\\
-e^{\th t} dX + e^{\th t}\th X_t) dt=& \th e^{\th t}\a dt + \b e^{\th t} dW
-\\
-Xe^{\th t} - X_0 = & \int \th e^{\th t}\a dt +  \int \b e^{\th t} dW
-\\
-X_t =& e^{-\th t} X_0 + \a(1-e^{-\th t}) +  \frac{\b e^{-\th t}}{\sqrt{2\th}}
-W(e^{2\th t}-1)
-\end{align*}
-which means that $X$ forgets its initial conditions exponentially fast and
-converges to a normal random variable with mean $\a$ and variance
-$\tfrac{\b^2}{2\th}$
-\end{ex}
-
-At the end of this sub-section, we state Ito's Lemma with multiple Wiener
-processes and then for a multidimensional state.
-
-\begin{thm}[Ito Lemma for $dW \in  \R^{m}$] Suppose $X$ satisfies the stochastic
-differential $dX_t = F dt + G dW$, where $ F \in \R, G \in \R^{1\times m}$ and
-$dW = \left(dW^{(i)}\right) \in  \R^{m}$ is a vector of independent Wiener
-Processes. Take $u(x,t) \in C^{2,1}[ \R \times [0,T]]$. Let $D :=\left[{\sum_k
-G_{k} G_{k} } \right] \in \R$
-
-Set $Y(t) = u(X,t)$
-\\
-then
-$$
-dY =  \left( \di_t u +  \di_{x} u \cdot F + 
-  \di^2_{x} u \cdot D \right)
-\intd{t} +
- \left(  \di_{x} u  \cdot  G \right) dW  
-$$
-\end{thm}
-
-\begin{thm}[Ito Lemma for $X \in R^n$]
-Suppose $X$ satisfies the stochastic differential $dX_t = F dt + G dW$, where $
-F \in \R^n, G \in \R^{n\times m}$ and $dW = \left(dW^{(i)}\right) \in  \R^{ 
-m}$ is a vector of independent Wiener Processes. Take $u(x,t) \in C^{2,1}[ \R^n
-\times [0,T]]$. Let $D_{ij} := \tfrac{1}{2} \left[{\sum_k G_{ik} G_{jk} }
-\right]_{ij} \in \R^{n\times n} $
-
-Set $Y(t) = u(X,t)$
-\\
-then
-$$
-dY =  \left( \di_t u + \sum_i \di_{x_i} u \cdot f_i + 
-\sum_{i,j} \di^2_{x_i x_j} u \cdot D_{ij} \right)
-\intd{t} +
- \left(  \sum_{ij} \di_{x_i} u  \cdot G_{ij} dW^{(j)} 
-\right)$$
-\end{thm}
-
-\subsection{Fokker-Planck and Kolmogorov's Backward Equations}
-The Fokker-Planck equation and Kolmogorov's Backward equation describe the
-forward (resp. backward)  evolution of the probability density of $X_t$
-
-For the rest of this section we will work in $ \R^n$, i.e. $X$ will be an $n-$
-dimensional vector that satisfies the stochastic differential 
-\begin{equation}
-dX_t = F(X,t) dt + G(X,t) dW,
-\label{eq:generic_Ito_SDE_Rn}
-\end{equation}
-$F \in \R^n, G \in \R^{n \times m}$,
-$dW = (dW^{(i)}) \in  \R^{  m}$ is an $m$-dimensional Brownian motion and we
-write $D_{ij} :=
-\tfrac{1}{2} 
-\left[{\sum_k G_{ik} G_{jk} } \right]_{ij} \in \R^{n\times n} $.
-
-% Let $$\F(x,t| y,s) = \Prob[X_t \preceq x| X_s = y] $$ be the forward transition
-% distribution and
-
-Let $$ \f(x,t| y,s) \intd{x} =  \Prob[X_t \in \intd{x} | X_s = y] $$ be the
-transition probability density. Then $\f$ satisfies:
-\begin{equation}
-\di_t \f= -\sum_i \di_{x_i} \left[ F_i(x,t) \f(x,t) \right] 
-+ 
-\sum_{i,j}  \di^2_{x_i x_j} \left[ D_{ij}(x,t) \f(x,t) \right]
-\label{eq:FP_pde}
-\end{equation}
-This is called the \emph{Fokker-Planck} or \emph{Forward Kolmogorov} equation.
-It can also be seen as a continuity euquation or a cosevation of probability
-equation. To this end define the probability current, $\p \in \R^n$ as:
-$$
-\phi_i =   F_i(x,t) \f(x,t) 
-+ 
-\sum_{j}  \di_{x_j} \left[ D_{ij}(x,t) \f(x,t)\right]
-$$
-then the Fokker-Planck equation, \cref{eq:FP_pde}, can be written as:
-$$
-\di_t \f= -\grad \cdot \phi
-$$
-which just says that the change in probability is the
-difference between the flow in and the flow out.
-
-Initial conditions for $\f$ are given by the distribution of $X_s$, if $s=0$
-then this is the initial distribution of $X$. Boundary conditions (BCs) depend
-on the domain of $X$ and what happens to $X$ once it hits its boundary. If the domain
-is all of $\R^n$ then we only insist that $\lim_{|x| \ra \infty} \f = 0$. If the
-domain has boundaries, then there are two common scenarios which we will also
-encounter in the sequel:
-\begin{enumerate}  
-  \item absorbing BCs
-  \item reflecting BCs
-\end{enumerate}
-At an absorbing boundary, the particle $X$ is removed and $\f=0$ there.
-In this situation, we will not have conservation of probability and the integral
-of $\f$ over $X$'s domain will monotonically decrease.
-
-At a reflecting boundary, the particle $X$ bounces back into its domain and we
-will have that the probability current $\p\cdot n = 0$ where $n$ is the outward
-normal at the reflecting boundary.
-
-We now consider $\f$ as functions of $y,s$, holding $x,t$ fixed, then
-\begin{equation}
--\di_s \f= \sum_i  \left[F_i(y,s) \cdot \di_{y_i}\f(|y,s) \right] 
-+ 
-\sum_{i,j}   \left[ D_{ij}(x,t) \di^2_{y_i y_j}\f(|y,s) \right]
-\label{eq:FP_backward_pde}
-\end{equation} 
-This is the \emph{Backward Kolmogorov} equation for $\f$. Note the minus sign
-in front $\di_s \f$. A mnemonic for the signs of the Backward vs. Forward
-equation is that as $t$ increases $\f$ diffuses and so $\di_t$ and $\di^2_{x}$
-have the same sign, but as $s$ increases $\f$ anti-diffuses and they have
-opposite signs.
-
-The differential operator on the right-hand side of
-\cref{eq:FP_backward_pde} occurs often in the study of SDEs and has its own
-name.
-\begin{defn}[Generator of an SDE] the Generator $A$ of $X$ is defined by:
-$$
-A[\psi(x)] = \lim_{t \searrow 0^+} \frac{\Exp[\psi(X_t)]  - \psi(x)}{t} ;  \quad X_0 =
-x \in \R^n
-$$
-\end{defn} 
-\begin{lemma} For an Ito SDE as in \cref{eq:generic_Ito_SDE_Rn} 
-$$
-A[\psi(x)] = \sum_i F_i(x,t) \cdot \di_{x_i} \psi + \sum_{i,j} D_{ij}(x,t)
-\di^2_{x_i x_j}\psi $$
-\end{lemma}
-
-\subsection{Stopping Times}
-\begin{defn}
-Let $\Fil(t)$ be some filtration, a random variable $\t$ is called a
-\emph{stopping time} if 
-$$
-\{\o : \t(\o) \leq t\} \in \Fil(t) \, \forall t  
-$$
-\end{defn}
-The colloquial way of describing stopping times is at any time we know
-whether $\t$ has occurred or not. For a counterexample, the time that a Wiener
-Process achieves its maximum over some interval is not a stopping time, since at any
-given time, we do not know if the maximum has occurred or not. The most common
-example of a stopping time is the first hitting-time, which is the first time
-$X_t$ leaves a set. 
-\begin{thm}[First-Hitting time] Let $E \subset \R^n, E \neq
-\phi$ be open or closed
-
-then $$\t := \inf\{ t \geq 0 | X_t \in E\}$$ is a stopping time.
-\end{thm}
-
-The reason stopping times are very useful is that all the facts so far quoted
-for Ito calculus using integrals $\int_0^T \intd{W}$ remain true if $T$ is
-replaced by a stopping time $\t$.
-
-Also it allows us to link SDE's and PDEs using the generator, $A$, of the
-diffusion: 
-\begin{thm}[Dynkin's formula] Given $\t$ a stopping time, $\Exp[\t] < \infty$.
-Then:
-$$
-\Exp[u(X_\t, \t)] =
-u(x, 0) + \Exp\left[\int_0^\t \di_t u + A[u] \intd{s}
-\right]
- $$
-\end{thm}
-This provides a link between PDEs and stochastic processes and allows us to go
-back and forth in that we can find probabilistic results by
-solving a PDE or we can approximate a PDE by simulating a stochastic process
-and averaging.
-
-\begin{ex}[\cite{Evansb} pg 99 - Expected hitting time to a boundary]
-\label{ex:mean_hitting_time}
- Let $\O
-\subset \R^n$ be a bounded open set with smooth boundary $\di \O$ then it is a
-basic fact from PDEs theory that
-\begin{equation}
-\begin{cases}
--\frac{1}{2} \grad^2 u = 1  & \text{over } \O
-\\
-u =  0 &\text{on } \di \O
-\end{cases}
-\end{equation}
-has a unique $C^\infty(\O)$ solution.
-
-Let $X = W_t + x$ for any $x \in \O$ and define
- $$\t_x := \text{first time } X \text{ hits } \di \O$$
-then the generator of $X$ is $A[\psi] = -\grad^2(\psi)/2$ and we will
-have:
-
-\begin{align*}
-\Exp[u(X_\t)] - \Exp[ u(x_0)] =& \Exp \left[ -\frac{1}{2}\int_0^\t \grad^2
-u\right]
-\\
-=& \Exp \left[ - \int_0^\t 1
-\right]
-\\
-=-&\Exp [\t]
-\end{align*}
-Finally, invoke $u$'s BCs, $u|_{\di E} = 0$, and $X$'s ICs, $X_0=0 = x$ to
-conclude that $$ u(x) = \Exp [\t]$$
-The solution to the PDE evaluated at $x$ is the expected exit time from
-$E$ for an $X_t$  starting at $x$.
-\end{ex}
-% Similarly we can calculate more moments of the hitting time.
-% \begin{ex}[Mean squared hitting time to a boundary] Let $E
-% \label{ex:mean_squared_hitting_time}
-%  \subset \R$ be a bounded open set and let $u$ solve:
-%  \begin{equation}
-%  \begin{cases}
-%  (\a - \tfrac{x}{\tc}) \di_x u + \frac{\b^2}{2} \di_x^2 u = -\frac t 2  &
-%  \text{over } E
-%  \\
-%  u =  0 &\text{on } \di E
-%  \end{cases}
-%  \end{equation}
-% 
-% Then $- t / 2 = A[u]$ for $A$, the generator, corresponding to $dX = (\a - x /
-% \tc) dt + \b dW$.
-% Define $$\t_x := \text{first time } X \text{ leaves }  E$$
-% then
-%  \begin{align*}
-%  \Exp[u(X_\t)] - \Exp[ u(X_0)] =& 
-%  \Exp \left[  \int_0^\t   (\a - \tfrac{x}{\tc}) \di_x u + \frac{\b^2}{2} \di_x^2
-%  u \right]
-%  \\
-%  =& \Exp \left[ - \int_0^\t -\frac t 2 
-%  \right]
-%  \\
-%  =-&\Exp [\t^2]
-%  \end{align*}
-% And we can conclude that 
-% $$ u(x) = \Exp [\t^2]$$
-% The solution to the PDE evaluated at $x$ is the expected squared exit time from
-% $E$ for an $X_t$  starting at $x$.
-% \end{ex}
-
-\subsection{Feynman-Kac formula}
-The Feynman-Kac formula is a generalization of the above idea. We state it in
-1-D. 
-\begin{thm}[Feynman-Kac Formula] Consider the (linear) PDE, with terminal
-conditions:
-\begin{equation}
-\begin{cases}
-\di_t u(x,t) + f \di_x u + \frac {g^2}2 \di_x^2 u - V u + q 
-&= 0
-\\
-u(x,\tf) &= \psi(x)
-\end{cases}
-\label{eq:PDE_feynman_kac}
-\end{equation}
-The dynamics can also be written:
-$$
-\di_t u(x,t) + A[u] - V u + q = 0
-$$
-for the generator $A = f \cdot \di_x + g^2/2 \cdot \di_x^2$ corresponding to the
-SDE: $$
-dX_t = f(X,t) dt + g(X,t) dW
-$$
-Then for $t< \tf$
-\begin{align*}
-u(x,t) =& \Exp\bigg[
-\int_t^\tf q(x,s)\cdot \exp \left(-\int_t^s V(X_\t) \intd{\t} \right)\intd{s} \\
-&\quad + \, \psi(X_t) \cdot \exp\left(-\int_t^\tf V(X_\t,\t) \intd{\t} \right)  
-\quad \Big| \,X_t = x \bigg]
-\end{align*}
-\begin{proof}[Proof]
-This can be seen as a more elaborate extension of example
-\ref{ex:mean_hitting_time}. We set: 
-$$
-Y_r = u(x,t)+ 
-\int_t^r q(x,s)\cdot\exp \left(-\int_t^s V(X_\t) \intd{\t} \right)\intd{s} +
-u(X_r,r)\cdot\exp\left(-\int_t^r V(X_\t,\t) \intd{\t} \right)   
-$$
-then 
-\begin{align*}
-dY =& q(x,r)\cdot\exp \left(-\int_t^r V(X_\t) \intd{\t} \right)
-\\
-& - u(X_r,r) \cdot V(X_r, r)  \exp\left(-\int_t^r V(X_\t,\t)
-\\
-& + du \cdot \exp\left(-\int_t^r V(X_\t,\t) \intd{\t} \right) 
-\intd{\t}\right)
-\end{align*}
-Now we recall Ito's lemma for $du(X,t)$, and replace it in the last line:
-\begin{align*}
-dY_r =& q(x,r)\cdot\exp \left(-\int_t^r V(X_\t) \intd{\t} \right) \intd{r}
-\\
-& - u(X_r,r) \cdot V(X_r, r)  \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t}\right)s
-\\
-& + \Big( \di_t u + A[u] \Big) \cdot \exp\left(-\int_t^r V(X_\t,\t) \intd{\t}
-\right) \intd{r}
-\\
-& + \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-\end{align*}
-Now we can group the terms in front of the exponential term on the first
-three lines to form:
-\begin{align*}
-dY_r =& \Big(\di_t u + A[u] - u(X_r,r) V(X_r, r) + q(X_r,r) \Big) \cdot\exp
-\left(-\int_t^r V(X_\t) \intd{\t} \right)
-\intd{r}
-\\
-& + \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-\end{align*}
-Since, $u$ satisfies the PDE, \cref{eq:PDE_feynman_kac}, the term on the first
-line is zero so $dY$ is just
-$$
-dY_r = \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot \exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-$$
-integrating we have: 
-$$
-Y_\tf - Y_t = \int^\tf_t \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot
-\exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W}
-$$
-Now we need conditions such that the expectation of the stochastic integral is
-zero. For example, it suffices that $V$ is bounded, and often in applications
-$V$ is a positive rate.
-Then, we   take conditional expectations, conditional on $X_t=x$:
-\begin{equation*}
-\begin{gathered}
-\Exp\Big[Y_\tf | X_t = x\Big] - \Exp\Big[Y_t | X_t = x\Big]  
-= 
-\\
-\Exp\Bigg[\int^\tf_t \Big(g(X_r,r)\di_x u(X_r,r)\Big) \cdot
-\exp\left(-\int_t^r V(X_\t,\t)
-\intd{\t} \right) \intd{W} \,\mid\, X_t = x\Bigg] 
-\\ = 0
-\end{gathered}
-\end{equation*}
-Unravelling the definition of $Y_t$, we get
-\begin{align*}
-\Exp\bigg[
-\int_t^\tf q(x,s)\cdot \exp \left(-\int_t^s V(X_\t) \intd{\t} \right)\intd{s} 
- \quad + &
-\\
- \, \psi(X_t) \cdot \exp\left(-\int_t^\tf V(X_\t,\t) \intd{\t} \right)  
-\, \Big| \,X_t = x \bigg] &=\, u(x,t)  
-\end{align*}
-
-\end{proof}
-A more detailed proof can be found in \cite{Oksendal2007} Thm. 8.2.1
-\end{thm}
-
-We will not use the Feynman-Kac formula in what follows, but
-we have included it here, because there are some very interesting developments in stochastic optimal
-control, which rely on solving a many-dimensional PDE using stochastic
-simulation, the link being made via the Feynman-Kac formula - see
-\cite{Kappen2005},\cite{Palmer2011}.
-
-% \subsection{Numerical Simulation of SDEs}
-
-\section{Deterministic Optimal Control in Finite Dimension}
-Optimal Control theory has three main components - a state, $x$, a control, $u$
-and an objective $J$ which is a functional of $x,u$ and which we try to either
-minimize or maximize. Here, we will be minimizing.
-
-We introduce the theory for a finite dimensional state, $x(t) \in \R^{n_x}$, but
-we will not address state constraints. Our control will also be
-finite-dimensional $u(t) \in \R^{n_u}$ and it will be allowed to take values in
-some closed domain, $\Udomain$, of allowed controls, e.g.\ $\Udomain = \{ u_t
-\in \R^{n_u} \, \st |u| \leq \Umax \}$ for a control constrained to a sphere.
-The objective $J$ will be given by a time-integral-plus-terminal cost:
-\begin{equation}
-J[u] = \int_0^\tf L(x_s, u_s) \intd{s} + M(x_\tf)
-\label{eq:generic_objective_function_deterministic}
-\end{equation}
-Where $\tf$ could be either a variable or fixed.
-
-The dynamics of $x$ are given by some controlled ordinary differential
-equation (ODE) given some initial conditions:
-\begin{equation}
-\dot{x} = f(x,t); \quad x(0) = x_0
-\label{eq:generic_dynamics_deterministic}
-\end{equation}
-
-Then we seek the optimal control $u^*$, such that:
-\begin{equation}
-u^* = \argmin_{u \in \Udomain} J[u]  
-\label{eq:generic_objective_argmin}
-\end{equation}
-
-Note that $x$ may also have final conditions, which we will deal with later,
-and/or path constraints. If there are path-constraints, we talk about a
-state-constrained problem. However, we will not face path constraints in the
-sequel and so will not say more about how to handle them.
-
-There are essentially three methods for solving equation
-\cref{eq:generic_objective_function_deterministic}. Two, Pontryagin's Minimum
-Principle and Dynamic Programing, are analytic in nature providing a set of
-equations that characterize the minimum and then solving these equations,
-numerically if necessary. The third, the direct method, simply discretizes the
-problem and solves the resulting nonlinear programing (NLP) problem via standard
-NLP techniques. We will describe both Pontryagin's Minimum Principle and Dynamic
-Programing below and indeed we will use both later on.
-
-\subsection{Pontryagin's Maximum Principle}
-Pontryagin's Principle is derived from on a variational argument, similar to the
-Euler-Lagrange equations from the Calculus of Variations, and it characterizes
-the optimal trajectory/control pair. It can be thought of as a generalization of
-the zero-tangent rule (Fermat's Rule) for finding optima in single-variable
-calculus and like both the Euler-Lagrange equations and Fermat's Rule it
-provides necessary, but not sufficient conditions for a minimum.
-
-Pontryagin's Principle also holds a similarity to Hamiltonian Mechanics, which
-should not be surprising, since Hamiltonian Mechanics is founded on
-the minimization of action.
-
-There are several versions of Pontryagin's Principle, depending on whether $x$ has
-terminal conditions or not, whether the final time, $\tf$ is specified or a
-variable in itself and most importantly whether the state has constraints or
-not. We will state the versions that are relevant for our needs, namely, $x$ has
-no constraints and $\tf$ is specified.
-
-We first introduce the Pontryagin Hamiltonian, $\H$, also called the control
-theory Hamiltonian and an adjoint state, $p \in \R^{n_x}$:
-\begin{defn} [$\H$] The control theory Hamiltonian is the function:
-$$
-\H(x,p, u) := f(x,u) \cdot p + L(x,u) \quad (x,p \in R^{n_x}, u \in \Udomain)
-$$ 
-\end{defn}
-
-\begin{thm}[Pontryagin's Fixed-Time, Free-End-Point] Assume $u^*(t)$ is
-optimal for
-\cref{eq:generic_objective_function_deterministic,eq:generic_dynamics_deterministic}
-and that $x^*$ is the corresponding trajectory.
-
-Then there exists a function $p^*:[0,\tf]:\ra \R^{n_x}$ st.
-\begin{equation}
-\begin{cases}
-\dot{x}^*(t) &=  \grad_p \H(x^*,p^*, u^*) = f(x^*, u^*)
-\\
-x(0) &= x_0
-\end{cases}
-\label{eq:pontryagin_state_ode}
-\end{equation}
-\begin{equation}
-\begin{cases}
-\dot{p}^*(t) &= -\grad_x \H(x^*,p^*, u^*) = -\grad_xL - p \grad_x f
-\\
-p^*(\tf) &= \grad_x M(x)
-\end{cases}
-\label{eq:pontryagin_adjoint_ode} 
-\end{equation}
-and
-\begin{equation}
-\H(x^*(t), p^*(t), u^*(t)) = \min_{u(t) \in \Udomain(t)}  \H(x^*(t), p^*(t),
-u(t))
-\label{eq:pontryaginH_optimality_condition} 
-\end{equation}
-\end{thm}
-We will not give a proof of this theorem (see A.2 in \cite{Evansb}),
-but we do provide a heuristic derivation in
-\cref{sec:Pontryagin_heuristic_derivation}, which is very helpful in deriving
-the necessary conditions for PDE dynamics and it explains the otherwise
-mysterious origins of $\H$ and $p$.
-
-At each time, $t$, we obtain the optimal control $u(t)$, by solving
-\cref{eq:pontryaginH_optimality_condition}. This can be easy, hard or useless.
-In the easy case, the minimum is analytically obvious, e.g.\ if the Hamiltonian
-is quadratic in $u$. If it is linear in $u$ then the minimum will occur on the
-boundaries of the control and we will talk about a bang-bang control.
-Conversely, solving the minimum of $\H$ with respect to $u$ may be impossible,
-since $x,p$ are unknown. In this case we need to have an iterative scheme which
-alternately solves
-\cref{eq:pontryaginH_optimality_condition} 
-and then \cref{eq:pontryagin_state_ode,eq:pontryagin_adjoint_ode}. Basic
-numerical techniques using this idea are discussed in the appendix in
-\cite{Kirk2004}. However, if we are going to be solving the problem numerically,
-then the direct method might be superior, see \cite{Ross2005} for a discussion.
-
-Finally, it is possible that \cref{eq:pontryaginH_optimality_condition} does not
-depend on $u$ explicitly (it will still depend  on $u$ implicitly through $x,p$)
-and so we cannot find $u$ at all by minimizing $\H$ wrt.\ $u$. This happens, for
-example, if the cost, $L$, does not depend on $u$, the dynamics, $f$, are linear
-in $u$ and $p=0$ over an interval of positive length. Such an optimal control
-problem is called singular and it is discussed briefly in \cite{Kirk2004}, and
-the theory can still be useful in finding a solution, but many complications can
-and do arise. We face a singular control problem in our own work later, however
-instead of tackling it head on, we merely regularize the problem - thus we will
-not say much more on Singular Control Problems.
-
-The version of Pontryagin's Principle for Fixed End Point, meaning $x$ has some
-terminal conditions, is very similar to the one with free end-point. Suppose 
-\begin{equation}
-\dot{x} = f(x,t); 
-\quad
-\begin{cases}
-x(0) = x_0
-\\
-x(\tf) = x_f
-\end{cases}
-\label{eq:generic_dynamics_deterministic_fixed_end_point}
-\end{equation}
-
-\begin{thm}[Pontryagin's Fixed-Time, Fixed-End-Point] Assume $u^*(t)$ is
-optimal for
-\cref{eq:generic_objective_function_deterministic,eq:generic_dynamics_deterministic_fixed_end_point}
-and that $x^*$ is the corresponding trajectory.
-
-Then there exists a function $p^*:[0,\tf]:\ra \R^{n_x}$ st.
-\begin{equation}
-\begin{cases}
-\dot{x}^*(t) &=  \grad_p \H(x^*,p^*, u^*) = f(x^*, u^*)
-\\
-x(0) &= x_0
-\\
-x(\tf) &= x_f
-\end{cases}
-\label{eq:pontryagin_state_ode}
-\end{equation}
-\begin{equation}
-\dot{p}^*(t) &= -\grad_x \H(x^*,p^*, u^*) = -\grad_xL - p \grad_x f
-\label{eq:pontryagin_adjoint_ode} 
-\end{equation}
-and
-\begin{equation}
-\H(x^*(t), p^*(t), u^*(t)) = \min_{u(t) \in \Udomain(t)}  \H(x^*(t), p^*(t),
-u(t))
-\label{eq:pontryaginH_optimality_condition} 
-\end{equation}
-\end{thm}
-I.e. $p$ loses its final conditions because $x$ has them. This relates to the
-variational nature of Pontryagin's Principle and the idea that $p$ represents
-the sensitivity of the objective wrt. small changes in $x$. Thus, if $x$ is
-fixed, there are no variations and $p$ can be anything.
-% To illustrate Pontryagin's Principle we will solve the moon
-% lander problem (Ex. 4.4.4 \cite{Evansb}), with the twist of holding the final
-% time fixed. 
-% \begin{ex}
-% 
-% \end{ex}
-
-\subsection{Dynamic Programing - the Hamilton-Jacobi-Bellman equation} 
-Dynamic Programing uses backwards recursion to tabulate the optimal control
-starting from the terminal time. The basic object in dynamic programing is the value function, $v$. In order to
-introduce it, we first extend our definition for the objective, $J$, to include later
-times, with different initial conditions.
-
-The running cost-to-go corresponding to
-\cref{eq:generic_objective_function_deterministic} is:
-\begin{equation}
-J[u; x, t] = \int_t^\tf L(x_s, u_s) \intd{s} + M(x_\tf), \quad x_t = x
-\label{eq:generic_cost_to_go_deterministic} 
-\end{equation}
-So that $J[u; x_0, 0] = J[u]$.
-
-The dynamic programming value function, $v$ is then defined by:
-$$
-v(x,t) = \inf_{u \in \Udomain} J[u; x,t] 
-$$
-it is also called the optimal cost-to-go.
-
-We immediately note the terminal conditions on $v$:
-\begin{equation}
-v(x,\tf ) = M(x)
-\end{equation}
-
-The Dynamic Programing approach is then to solve for $v$ backwards and in the
-process infer the control $u$.
-\begin{thm}[Hamilton-Jacobi-Bellman Equation] Assume that $v$ is $C^{1,1}$ in
-$x,t$. Then $v$ solves:
-\begin{equation}
-\begin{cases}
--\di_t v(x,t) &= \max_{u(t) \in \Udomain(t)} \left\{ f(x,t) \cdot \di_x v +
-L(x,u) \right\}
-\\
-v(x,\tf) &= M(x)
-\end{cases}
-\end{equation}
-\end{thm}
-We will not derive the HJB equation here, but we will derive a similar version
-later for the stochastic case, when the dynamics are given by SDEs.
-
-% \subsection{Model Predictive Control}
-% Solving the HJB equations and obtaining a feedback control is usually
-% impossible. Thus a practical alternative is to solve a simplified problem on a
-% limited horizon and use the simplified solution now and then after stepping
-% forward in time and observing the system in a new position, to solve a revised
-% limited horizon problem and use the new solution.
-% Let us make this more precise.
-
-\section{Stochastic Optimal Control}
-Here we extend the Dynamic Programing theory to the case where the dynamics are
-governed by an SDE:
-\begin{equation}
-dX = F(X,t, u) dt + G(X,t) dW
-\end{equation} 
-i.e.\ we assume control affects only the drift term. This 
-assumption is not necessary, but it corresponds to our problems.
-
-For stochastic dynamics one can only optimize probabilistically and so
-$J$ becomes:
-\begin{equation}
-J[u] = \Exp \left[ \int_0^\tf L(x_s, u_s) \intd{s} + M(X_\tf)
-\right] 
-\label{eq:generic_objective_function_stochastic}
-\end{equation}
-while the running cost-to-go is:
-\begin{equation}
-J[u; x, t] = \Exp \left[ \int_t^\tf L(X_s, u_s) \intd{s} + M(X_\tf) \right], 
-\quad X_t = x
-\label{eq:generic_cost_to_go_stochastic}
-\end{equation}
-the value function is:
-\begin{equation}
-v(x,t) := \inf_{u \in \Udomain} J[u; x,t]
-\end{equation}
-
-Finally, we again have that $v$ satisfies a certain PDE:
-\begin{thm}[Stochastic HJB] 
-\label{thm:stochastic_hjb}
-\begin{equation}
-\begin{cases}
--\di_t v(x,t) &=  \max_{u(t) \in \Udomain(t)} \big\{ L(x,u)  +
-F(x,t,u) \cdot \di_x v
-\big\} + \frac{G^2(x,t)}{2} \di_x^2v
-\\
-v(x,\tf) &= M(x)
-\end{cases}
-\end{equation}
-\begin{proof}[Heuristic Derivation (from \cite{Evansb})] Suppose we are at time
-$t$ and $X_t = x$.
-
-Take a time increment $[t, t+h]$ and assume that during that time we apply a
-control $u$ and subsequently we apply the the optimal control $u^*$. The
-running-cost will then break down as: 
-$$
-J[u; x,t] = \Exp \left[ \int_t^{t+h} L(X_s, u_s) \intd{s}  + v(X_{t+h},
-t+h) \right] $$
-
-Since $v(x,t) = \inf J(u; x,t)$, we must have that:
-$$
-v(x,t) \leq  \Exp \left[ \int_t^{t+h} L(X_s, u_s) \intd{s}  + v(X_{t+h},
-t+h) \right] $$
-or, rearranging, that
-\begin{align*}
-0 \leq^&  \Exp \left[ \int_t^{t+h} L(X_s, u_s) \intd{s} \right]  
-+ \underbrace{\Exp \left[ v(X_{t+h}, t+h) - v(x,t) \right]}_{\Exp\left[ dv
-\right]}
-\end{align*}
-Now, $\Exp\left[ dv \right]$ can be expressed using Dynkin's Formula, as:
-$$
-\Exp[dv] = \int_t^{t+h} \di_t v +  F\cdot \di_x v + \frac{G^2 }{2}\cdot \di_x^2
-v \intd{s} $$
-Plugging that back, we get
-$$
-0 \leq \int_t^{t+h} L(X,u) +  \di_tv +  F\cdot \di_x v + \frac{G^2 }{2}\cdot
-\di_x^2 v \intd{s} $$
-Taking $h \ra 0$ we get 
-$$
-0 \leq  L(x,u) +  \di_tv(x,t) +  F\cdot \di_x v(x,t) + \frac{G^2 }{2}\cdot
-\di_x^2 v(x,t) $$
-and we conjecture that for the actual optimal control, the inequality becomes an
-equality:
-\begin{equation}
-\begin{cases}
-0 &=   L(x,u^*)+ \di_t v(x,t) +
- F(x,t,u^*)\cdot \di_x v(x,t) + \frac{G^2(x,t)}{2}\cdot \di_x^2 v(x,t)
-\\
-u^*(t) &= \argmax_{u \in \Udomain(t)}  
-\big\{L(x,u) + F(x,t,u)\cdot \di_x v \big\}
-\end{cases}
-\end{equation}
-\end{proof}
-\end{thm}
-% A rigorous proof can be found in \cite{Krylov2008}, sec???ch???
-
-\section{Deterministic Optimal Control in Infinite Dimensions}
-Now we discuss Optimal Control when the dynamics are deterministic but of
-infinite-dimension, i.e. the state evolution is a PDE instead of an ODE.  The
-reason we discuss this, is that optimization of ordinary SDEs in an open-loop
-control is actually a deterministic optimal control problem in infinite
-dimensions.
-
-Why?
-
-Since the objective is probabilistic and the probability density evolution is
-given by the (deterministic) Fokker-Planck PDE, we need to optimize a
-deterministic system whose dynamics are given by the Fokker-Planck equation.
-
-There is no all-encompassing generalization of Pontryagin's Principle to
-infinite dimensions, but a Maximum Principle can be derived for certain systems.
-To illustrate the idea we will discuss a simple case as in \cite{Palmer2011}.
-Our controlled SDE will be given by: 
-$$ dX_t = F(X,t, u) dt + G(X,t) dW, $$
-The corresponding probability density is given by: $$ \di_t \f=
--\sum_i \di_{x_i} \left[ F_i(x,t,u) \f(x,t) \right] + \sum_{i,j}  \di^2_{x_i
-x_j} \left[ D_{ij}(x,t) \f(x,t) \right] $$ which we will write as $$ \dot{\f} =
-\L_{u} [\f] $$ where $\L_{u}$ is the differential operator corresponding to the
-RHS of the Fokker-Planck equation parametrized by the control $u$. For now we
-will assume that there are no BCs, and the domain of $X$ is all of $\R^n$.
-
-A natural inner-product, $ \langle.,. \rangle$ on functions in $L^2[\R^n]$ is
-given by: $$
- \langle\psi,\phi  \rangle = \int_{\R^n} \psi(x) \phi(x) \intd{x}
-$$
-Our objective will be to minimize:
-\begin{equation}
-J[u] = \int_0^\tf  \langle L(\cdot,s,u), \f_u(s) \rangle \intd{s} +  
-\langle M(\cdot),\f(\tf) \rangle
-\label{eq:generic_objective_function_probabilistic}  
-\end{equation}.
-
-\Cref{eq:generic_objective_function_probabilistic} is a very natural cost
-function in the stochastic dynamics case, because it can also be written as: 
-$$
-J[u] = \Exp \left[ \int_0^\tf  L(X_s, s,u) \intd{s} + M(X_\tf) \right] 
-$$ which
-is just \cref{eq:generic_objective_function_stochastic}. We use
-\cref{eq:generic_objective_function_stochastic} if we have some feedback and
-want to use dynamic programming. We use
-\cref{eq:generic_objective_function_probabilistic} if we have no observations
-and need to run an open-loop control.
-
-Since, \cref{eq:generic_objective_function_probabilistic} is the objective
-function for deterministic dynamics, we can use Pontryagin's Principle adapted
-to the case of infinite dimensional dynamics (PDEs).
-
-Introduce an adjoint co-state, $\l \in \R^n$ and let the Hamiltonian, $\H$ be:
-$$
-\H(\f, \l, u)= \langle \phi, L + \Lstar[\l] \rangle$$
-where $\Lstar$ is the adjoint operator to $\L$ defined via:
-$$
-\langle \L [\phi], \l \rangle =\langle \phi, \Lstar[\l] \rangle
-$$
-
-Let $\l$ evolve (backwards) according to:
-\begin{equation*}
-\begin{cases}
--\di_t \l =& \Lstar[\l] + L
-\\
-\l(\tf) =& M(x)  
-\end{cases}
-\end{equation*}
-then the optimal control, $u$, is given via
-$$
-u = \argmin_u {\H(\f, \l, u)}
-$$
-
-An important fact, useful for computations, is that the total derivative of the
-objective with respect to the control is given by the partial derivative of the
-Hamiltonian with respect to the control:
-\begin{equation}
-\grad_u \big[J[u] \big] = \di_u \H(\f, \l, u)
-\label{eq:objective_gradient_wrt_control}
-\end{equation}
-
-So for a given $u$, we can compute $\f,\l$, calculate the objective gradient
-using \cref{eq:objective_gradient_wrt_control} and then iterate in the direction
-of descent (if we are minimizing).
-
-All that remains is to compute $\Lstar$. It turns out that if we do not have any
-BCs, then
-$$
-\Lstar[\l] = \sum_i F_i(x,t) \cdot \di_{x_i} \l + \sum_{i,j} D_{ij}(x,t)
-\di^2_{x_i x_j}\l 
-$$ 
-which is just the generator of $X$!
-
-We will see later, that things become more complicated when we need to
-consider a PDE's BCs.
-
-\section{Closed-loop vs.\ Open-Loop control}
-At this point we stop to briefly discuss the difference between
-open-loop control and closed-loop control.
-
-By closed-loop control, we mean that at any time we have observations of the
-system state, $x$ and its value can influence the control. In open loop control,
-we decide on the value of $u(t)$ a priori and apply it independently of the
-value of $x(t)$, based only on the knowledge of the system dynamics and the
-initial conditions.
-
-If the system dynamics are deterministic there is no added value in closed-loop
-control, since the state evolution is perfectly predictable and one can infer
-where $x$ will be based only on the dynamics and the applied control. For
-stochastic dynamics this is no longer true and there is something to be gained
-by observing the state, since different controls can be applied depending on the
-different realizations of the stochastic process driving the dynamics.
-Naturally, if one has observations of the system, one would prefer to use this
-and apply a closed-loop control. Thus, an open-loop control would be used
-only if one has no reliable knowledge of the system state.
-
-\section{A Stochastic Optimal Control Problem from Mathematical Neuroscience}
+\section{Introduction}
 Neurons relay information by means of voltage spikes - sudden sharp increases in
@@
 thought to be contained in the length of the time-interval between these spikes.
-In the simplest case, this can be thought of as a rate  -the average number of
+In the simplest case, this can be thought of as a rate-the average number of
 spikes per time interval, but more complicated coding schemes are hypothesized to
@@
 We now consider this problem in the framework of Stochastic Optimal Control and
-give both a feedback solution, when the cell voltage is explicitly observable as
-well as an open-loop solution when only the occurrence of a spike is
-robustly observable.
-
-\subsection{Problem Formulation}
-The most basic representative model for a neuron is the noisy
+give both a feedback solution, when the cell voltage is explicitly and
+perfectly observable as well as an open-loop solution when only the occurrence
+of a spike is robustly observable.
+
+The paper is structured as follows: First we describe the neuron model and
+formalize the control objective. Then we describe a feedback-based solution
+which assumes that the controller has detailed access to the voltage trajectory.
+Then we relax the observation assumption so that the controller only has access
+to the spike timings. Finally we compare the two methods against a
+simple-minded control technique which ignores the stochastic input to the
+neuron.
+
+\section{Problem Formulation}
+A basic but useful model for a neural membrane potential is the noisy
 leaky-integrate-and-fire model:
@@
 \begin{gathered} 
-dX_s = (\a(t) - \frac{X_s}{\tc} ) \intd{s} + \b \intd{W_s},
+dX_s = \left(\a(t) - \frac{X_s}{\tc} \right) \intd{s} + \b \intd{W_s},
 \\
@@
 
-Suppose we want to ensure that spikes occur exactly at the predetermined
-times $\tns$. Then given the renewal nature of our spiking model, we can work
-sequentially, focusing only on the first of these intervals, of duration $t_1 =
-\T$, and then ask: how do we choose $\a(t)$ st. $\ts = \T$. There are several
-ways to specify what we mean by 'make $\ts = \T$', all of which result in optimal
-control problems. A natural one is a least squares solution:
+Suppose we want to ensure that spikes occur exactly at the predetermined times
+$\tns$ and that we are at least capable of knowing when a spike has occurred.
+Then given the renewal nature of our spiking model, we can work sequentially,
+focusing only on the first of these intervals, of duration $t_1 = \T$, and then
+ask: how do we choose $\a(t)$ st. $\ts = \T$. A natural optimal control
+objective formulating this is to require a least squares solution
 \begin{equation}
@@
 \end{equation}
+where expectation is taken with respect to the trajectories of $X_t$.
 
@@
 \begin{equation}
-\a(t) \in [\amin, \amax]
+\a(t) \in [\amin, \amax] \quad \forall t
 \label{eq:bound_constraints_alpha}
@@
 
-Experimentally it is often difficult to observe precisely the exact voltage
-trajectory, but the occurrence of a spike is much more easily observable. As
-such a closed-loop approach is not implementable since the state is unknown. In
-the sequel we will perform three calculations. First we will derive the HJB
-equation for a closed-loop control if the system was observable, and use this as
-a benchmark for the best possible performance of an open-loop control. Then we
-will calculate the most simple-minded open-loop control, which is just the
-deterministic control of the dynamics when assuming there is no noise, $\b=0$.
-THen we will calculate the optimal open-loop control by optimizing the
-transition density. And finally we will compare the performance of all three
-controls.
+We will consider two scenarios, one, {\sl closed loop control}, in which the
+value of $X_t$ at $t$ is known and can be used in determining the control and
+,two, {\sl open loop control} in which only the times of spikes is known. In
+closed-loop scenario, we might better write the control as
+$$\a = \a(x,t)$$ to indicate its dependence on $X_t$. In the open-loop scenario,
+we would instead write $$\a = \a(t)$$ to indicate that we will always do the
+same thing regardless of the (unknown) behavious of $X_t$. The techniques used
+in both scenarios will be different. For the closed-loop scenario we will use Dynamic Programing,
+\cite{Fleming1975}, while for the open-loop scenario we will use a form
+of the Maximum Principle for transition densities, \cite{Ahmed1981}.
+
+Crucially, we assume that the model parameters in \cref{eq:X_evolution_uo} are
+known and given.
 
@@
 i.e.\ if time $t$ has elapsed without a spike, we now want to minimize
-the difference between $\t$ and ($T-t$), given the current state $x$. For now we
-assume that we know $x$, later we will treat the problem when that is not so.
-
-We are now in a position to set up the HJB equation for this problem. First
-introduce the value function, $v(x,t)$ as:
-\begin{multline}
-v(x,t) = \argmin_{\a(s \geq t)} \{J[\a, x,t] \} =
-\\ 
+the difference between $\t = \ts - t$ and ($T-t$), given the current state $x$.
+Recall that in the closed-loop scenario, we assume that $x = X_t$ is known
+to the controller.
+
+\subsection{Hamilton-Jacobi-Bellman equation for the value function}
+We are now in a position to set up the Hamilton-Jacobi-Bellman (HJB) equation
+\cite{Fleming1975,Evansb} for this problem. First introduce the value function,
+$v(x,t)$ as: 
+\begin{align}
+v(x,t) =& \argmin_{\a(s \geq t)} \{J[\a, x,t] \}
+\notag 
+\\ 
+=&
 \argmin_{\a(s \geq t)} 
@@
 
-We have been canny enough to set up a quadratic optimization problem in the
-control, which means that the optimal control is given analytically via:
-\begin{equation}
-\a(x,t) = \frac{-\di_x v}{2\e} 
+In this set-up we can solve for the optimal control analytically via:
+\begin{align}
+\a(x,t)  =&\argmin_{\a \in [\amin, \amax]}\big\{
+\e \a^2 + (\a-\tfrac{x}{\tc}) \di_x v \big\}
+\notag 
+\\
+=& 
+\min \left(\amax, \max\left(\amin, -\frac{\di_x v}{2\e}\right)
 \label{eq:OC_LS_variance_energy_quadratic_control}
-\end{equation}
-However we need to recall the bounds on the control, $\amin < \a < \amax$, and
-thus
-\begin{equation}
-\astar(x,t) = \min \left(\amax, \max\left(\amin, -\frac{\di_x v}{2\e}\right)
-\right)
-\label{eq:OC_LS_variance_energy_quadratic_control_bounds}
-\end{equation}
+\end{align}
 With this the HJB becomes
@@
 \end{equation}
-
-In the sequel we will drop the superscript star on the optimal control and
-just write $\a(t)$ for the solution of
-\cref{eq:OC_LS_variance_energy_quadratic_control_bounds}. 
-% Thus the control can be solved out of the HJB equation, which now becomes:
-% \begin{equation}
-% \di_t v(x,t) + \tfrac{\b^2}{2} \di_x^2 v + \tfrac{(\di_x v)^2}{\e}
-% -\tfrac{x}{\tc}\di_x v= 0
-% \label{eq:OC_LS_mean_HJB}
-% \end{equation}
 
@@
 Finally we come to the Terminal Conditions (TCs) for $v$. The idea is simple: if
-we reach $\T$ without having spiked we will just slam on the accelerator: i.e.\
-$$t>T \implies \a(t) = \amax$$ Thus: $$ v(x,T) = \Exp[\t^2|x, \amax] $$.
+we reach $\T$ without having spiked we will just apply maximum control in
+the positive direction, i.e.\ $$t>T \implies \a(t) =
+\amax$$ Thus: $$ v(x,T) = \Exp[\t^2|x, \amax] $$. This is indeed optimal only if
+we ignore the energy term, $\eps u^2$ in the objective, and is a good
+approximation for $\epx \ll 1$.
 
@@
 
-\subsection{Calculating the TCs} 
-\label{sec:calculating_TCs}
-At this point we are in a position to solve the HJB \cref{eq:OC_LS_HJB_full}.
-But first we need a more concrete expression for the TCs. Let
-\begin{equation}
-\begin{array}{lcl}
-\Tone(x_0) &=& \Exp[\t|x_0, \amax] 
-\\
-\Ttwo(x_0) &=& \Exp[\t^2|x_0, \amax]  
-\end{array}
-\end{equation}
-where as before, $\t$ is the remaining time-to-spike, given $X_0 = x_0, \a(t) =
-\amax$. Naturally, $\Ti$ depend on $x_0$ and are affected by the value of the
-various parameters, $\a,\b,\t$.
-
-Actually, we only  need $\Ttwo$ for the HJB equation above, but we'll see that
-$\Tone$ will be needed as an auxiliary.
-
-Indeed, analytical expressions exist for $\Ti$, \cite{Inoue1995}. However in my
-experience, they are very finicky numerically, especially as they involve
-infinite alternating sums which can easily overflow in numeric calculations.
-Moreover, they are not particulary convenient if you need $\Ti (x_0)$ for a wide
-range of $x$'s as we do here. Instead we will turn to:
-
-\subsubsection{the Backward Kolmogorov Equation}
-Here we follow Jacobs, \cite{Jacobs}, in deriving and solving a differential
-equation for $\Ti(x_0)$.
-
-Basically we are calculating the mean-time-squared to exit an interval $[\xmin,
-\xth]$. Theoretically, $\xmin = -\infty$, but for our purposes we will set it to
-some finite value and impose reflecting boundaries there - this is a very good
-approximation for $\xmin \ll 0$, as long as $\a>0$, which is what we have.
-
-Let $B(y,t|x) = \Prob[X_0 = y| X_t = x]$ for a fixed $x$. Then $B$ solves the
-following (backward Kolmogorov) PDE:
-\begin{equation}
--\di_t B = (\a - \frac y\tc)\di_yB - \frac {\b^2} 2 \di_y^2 B
-\label{eq:backward_Xdensity}
-\end{equation}
-
-Let $\G(t,y)$ be the survival function for $T$ for $X_0 = y$. 
-$$\G(t,y) = \Prob[T>t | X_0 = y] = \Prob[X_{s\leq t} \in [\xmin,
-\xth] | X_0 = y]$$.
-
-Because the dynamics do not depend explicitly on time, we have:
-$$
-\G(t,y) = \int_\xmin^\xth B(y,t|x) \intd{x}
-$$
-and so $\G$, being a definite integral of $B$ satisfies the same PDE:
-\begin{equation}
--\di_t \G = (\a - \frac y\tc) \di_y\G - \frac {\b^2} 2 \di_y^2 \G
-\label{eq:backward_SDF}
-\end{equation}
-
-With the BCs, ICs:
-\begin{equation}
-\begin{cases}
-\G(t,\xth) = .0 & \quad (T = t \implies \Prob[T>t] = 0)
-\\
-\di_x \G(t,\xmin) = .0  & \quad (\textrm{due to reflecting boundary for X})
-\\
-\G(0, y) = 1. & \quad (T \textrm{ is definitely} > 0 \textrm{ for } x_0 \in
-(\xmin, \xth) )
-\end{cases}
-\end{equation}
-
-\subsubsection{A BVP for $\Ti(x_0)$}
-If we solve the PDE for $\G$ over $t \in [0,\infty)$, then we can calculate
-$\Ttwo(x_0)$ as:
-\begin{eqnarray}
-\Ttwo(x_0) &=& \int_0^\infty t^2 \cdot (-\di_t\G(t,x_0)) \intd{t}
-\\
-		   &=& \int_0^\infty 2t \cdot \G(t,x_0) \intd{t}
-\end{eqnarray}
-assuming it exists.
-
-Since $\Ttwo$ is an integral over time, we can reduce the PDE in
-\cref{eq:backward_SDF} for the distribution to a BVP for the moment(s) by
-integrating over time. Here is how this is done: multiply both sides  of
-\cref{eq:backward_SDF} by $2t$, and integrate with respect to time:
-\begin{equation}
-\int_0^\infty 2t (-\di_t \G) \intd{t}
-= 
-\int_0^\infty  2t \left[ (\a - \frac y\tc) \di_y\G - \frac {\b^2} 2
-\di_y^2 \G\right]
-\intd{t}
-\end{equation}
-The LHS turns out to be $2\Tone$, so 
-\begin{equation}
-2\Tone
-=(\a - \frac y\tc)   \di_y \Ttwo 
-- \frac {\b^2} 2
-\di_y^2 \Ttwo
-\label{eq:BVP_Ttwo}
-\end{equation}
-Very similarly we can derive an equation for $\Tone$, namely:
-\begin{equation}
-1
-=(\a - \frac y\tc)   \di_y \Tone 
-- \frac {\b^2} 2
-\di_y^2 \Tone
-\label{eq:BVP_Tone}
-\end{equation}
-In both cases, the BCs will be:
-\begin{equation}
-\begin{cases}
-\Ti(\xth) = .0 & \quad (\textrm{we are already spking})
-\\
-\di_x \Ti(\xmin) = .0  & \quad (\textrm{due to reflecting boundary for }X)
-\end{cases}
-\label{eq:BVP_Ti_BCs}
-\end{equation}
-
-\Cref{eq:BVP_Tone,eq:BVP_Ttwo} are seemingly second order ODEs, but are
-actually first order and can be written as:
-\begin{equation}
-f(y)
-= (\a - \frac y\tc) S(y) 
-- \frac {\b^2} 2 S'(y) ; \quad  S(\xmin) = .0
-\label{eq:BVP_T_generic}
-\end{equation}
-and we can recover $\Ti$ from $S$ via:
-$$\Ti(x_0) =   -\int_{x_0}^\xth S(y)\intd{y}$$.
-It is clear that thus defined, $\Ti$ will satisfy the BCs of
-\cref{eq:BVP_Ti_BCs}.
-
-% Solving \cref{eq:BVP_T_generic} for $S$ gives: \begin{eqnarray} S(y) &=& -
-% \exp\left(\smallint \tfrac UD \right) \cdot \int_\xmin^{y} \exp\left(-\int
-% \tfrac UD\right) \frac{f(\xi)}D \intd{\xi}
-% \\
-% &=& - \exp\left( \frac{\a y - \frac{y^2}{2\tc}}D \right) \cdot \int_\xmin^{y} 
-%  \exp\left(- \frac{\a \xi - \tfrac{\xi^2}{2\tc}}D \right) \frac{f(\xi)}D
-% \intd{\xi} \label{eq:BVP_T_generic_soln} \end{eqnarray}
-We can solve \cref{eq:BVP_T_generic} analytically, but the analytical solution
-turns out to be awkward for numerical calculations, so we
-are better off just calculating the numerical solutions to \cref{eq:BVP_Tone,eq:BVP_Ttwo} directly
-via an ODE solver, starting from $S(\xmin) = .0$ and integrating until $\xth$.
-
-With this we can compare $\Ttwo$ between simulations and the numerical
-solutions see \cref{fig:Ti_empirical_vs_analytic}.
-\begin{figure}[htp]
-\begin{center}
-\subfloat[]
-{
-\includegraphics[width=0.48\textwidth]
-{/home/alex/Workspaces/Latex/OptSpike/Figs/Moments_a=10_b=1_N=128.png}
-}
-\subfloat[ ]
-{
-\includegraphics[width=0.48\textwidth]
-{/home/alex/Workspaces/Latex/OptSpike/Figs/Moments_a=10_b=10_N=128.png}
-}
-\\
-\subfloat[ ]
-{
-\includegraphics[width=0.48\textwidth]
-{/home/alex/Workspaces/Latex/OptSpike/Figs/Moments_a=15_b=1_N=128.png}
-}
-\subfloat[ ]
-{
-\includegraphics[width=0.48\textwidth]
-{/home/alex/Workspaces/Latex/OptSpike/Figs/Moments_a=15_b=10_N=128.png}
-}
-\caption[]{Comparison between solutions of equations
-\cref{eq:BVP_Tone,eq:BVP_Ttwo} and empirical statistics from simulating the
-hitting times of \cref{eq:X_evolution_uo} for four different pairs of $\a,\b$.
-In the top panel we show the first moment of the remaining time-to-spike
-starting from $x$, $\Tone(x)$, and in the bottom panel, we show the second
-moment, $\Ttwo(x)$. We have used only $N=128$ samples to form the moment
-statistics.}
-\label{fig:Ti_empirical_vs_analytic}
-\end{center}
-\end{figure}
+A few words on the terminal conditions (TCs). $\Exp[\t^2|x, \amax]$ is the
+expected time to spike (reach the threshold) starting at $x$ and applying
+$\amax$ throughout. We will not show it here, but this quantity can be found
+easily for all $x$ in the domain by solving a stationary backward Kolmogorov
+equation.
 
@@
 this PDE. It is time to discuss the numerical method for solving
-\cref{eq:OC_LS_HJB_full}.
-Since we are in 1-d and we have the simplest geometry, the simplest thing to do
-is try a centred Finite Difference scheme, upwind wherever necessary and hope
-for the best.
-
-Things will break down if $\e$ is very small in which case the quadratic term,
-$(\di_x v)^2 / \e$, will dominate and cause 'havoc' or if $\b$ is very small in
-which case the smoothing effect of the second derivative will be negligible and
-we will be in the zone of nonlinear conservation equations. Let us hope for the
-best now and assume these two problems do not arise, for the parameters we
-select.
-
-Recall that we are solving backwards in time. Then for variables at the next
-time-step (already computed) we will write $(\cdot)^+$; while variables at the
-previous time step (to-be computed), will be denoted as$(\cdot)^-$.
-
-Similarly given some node in space, we will write
-$(\cdot)_-,(\cdot)_0,(\cdot)_+$ as the left node, the node itself and the right
-node.
-
-Then our finite-difference scheme in time looks like:
-\begin{equation}
-\frac{v^- - v^+}{\Delta t} = \tfrac{\b^2}2  .5*(\di_x^2 v^- + \di_x^2 v^+) +
-(\a^+ - \tfrac{x}{\tc})  .5*(\di_x v^- + \di_x v^+) + \e (\a ^+)^2
-\end{equation}
-
-I.e.\ we have removed the non-linearity in $v$ by computing the control, $\a$ at
-the next, already computed, time-step and used central-differencing
-otherwise.
-
-For the spatial derivatives, we will not upwind for now and also use central
-differencing:
-$$\di_x v = \frac{v_+ - v_-}{2\Delta x}$$
-and as is standard:
-$$\di^2_x v = \frac{v_+ - 2 v_0 + v_-}{(\Delta x)^2}$$
+\cref{eq:OC_LS_HJB_full}. Since we are in 1-d the simplest thing to do is try a
+centred Finite Difference scheme, upwind wherever necessary and hope for the
+best.
 
@@
 numerical method, but just demonstrate its performance on a representative
-parameter set in \cref{fig:HJB_attempt} we see that the method is stable for a
-particular parameter set.
-
-% \begin{figure}[ht]
-% \begin{center}
-%   \includegraphics[width=.7\textwidth]{Figs/KolmogorovBVPMoments_a=10_t=9_b=8.png}
-%   \caption[]{TCs for alpha = 1.0;  tauchar = .9; beta = .75;}
-%   \label{fig:TCs_extended_x}
-% \end{center}
-% \end{figure}
-
-Let's discuss the most salient features of \cref{fig:HJB_attempt}. At the end,
+parameter set in \cref{fig:HJB_attempt}.
+
+Let us discuss the most salient features of \cref{fig:HJB_attempt}. At the end,
 when $t = \T$, the value function, $v$ is monotonely decreasing. This makes
@@
 
-\section{Deterministic Control}
-\label{sec:deterministic_ode_control}
-Now we will compute the optimal control in the case that $\b = 0$. The reason we
-do this is that since we have assumed that we have no feedback, we attempt to
-devise an open-controller by controlling the non-random dynamics. The
-calculation of this control turns out to be very simple:
-
-First let us assume that we actually can get to $\xth$ from $0$ in $\T$ time
-given $\amax$ and no noise.
-
-Then 
-\begin{equation}
-\begin{gathered}
-J = \int_0^\T \e \a^2 \intd{t}
-\\
-\st
-\begin{cases}
-\dot{x} = (\a(t) - \frac{x}{\tc})
-\\
-x_\T = \xth
-\\
-x_0 = .0
-\end{cases} 
-\end{gathered}
-\end{equation}
-The (Pontryagin) Hamiltonian becomes:
-\begin{equation}
-\H(x,p, \a) = \e \a^2 + p \cdot (\a - x/\tc )
-\end{equation}
-the adjoint is given by:
-\begin{equation}
--\dot{p} = \di_x \H = - p / \tc 
-\end{equation} 
-and the (unconstrained) control by:
-\begin{equation}
-\a(t) = \frac p {2 \e}
-\end{equation}
-which looks very similar to the stochastic case, once you put $p = \di_x v$.
-
-We can actually solve for the adjoint analytically: 
-$$
-p = p_0 \exp(t / \tc)
-$$
-where we can freely choose $p_0$ to later recover the TCs for $x$.
-$$
-\a(t) = \frac{p_0 \exp(t / \tc)}{2\e}
-$$
-Finally the solution for $x$ is:
-\begin{align*}
-\dot{x} & = \a(t) -  \frac{x}{\tc}
-\\ & = \frac{p_0 \exp(t / \tc)}{2\e} - \frac{x}{\tc}
-\\
-e^{t/\tc}x &= \int^t e^{s/\tc}\frac{p_0 \exp(s / \tc)}{2\e} \intd{s}
-\\
-x &= e^{-t/\tc} \cdot \frac{p_0 }{2\e} \tc \left[  \exp(2 t / \tc) -1 \right] 
-\\
-&= \frac{p_0 }{\e} \tc \cdot  \sinh (t/ \tc)
-\end{align*}
-And so: 
-$$
-x(\T) = \xth \implies p_0 = \e \frac{\xth  }{ \tc \sinh (\T/ \tc) }
-$$
-And we have a full solution.
-
-If we have saturation, i.e. if $\a$ hits its upper bound, then the situation is
-mildly more complicated. Since $p$ is monotonically increasing, as long as $p_0
-> \amin$ then $\a(t) > \amin \forall t$. So the saturation is simplified to:
-\begin{equation}
-\a(t) = \min \left(\frac{p_0 \exp(t / \tc)}{2\e} , \amax \right)
-\label{eq:deterministic_control_bounds}
-\end{equation}
-In particular, $\a$ will start by following $p(t)$ and then eventually hit its
-upper bound and stay there until $\T$.
-
-The only complication then is to calculate when $\frac{p_0 \exp(t / \tc)}{2\e} =
-\amax$ and then redo the calculation for $x$ with the new control in order to
-obtain the new value of $p_0$. Or we can just brute-force solve for $p_0$, by a
-simple shooting method. Using the latter approach, a solution-control pair for
-the parameter set we will use later is shown in \cref{fig:deterministic_soln_controls}. 
-%\usepackage{graphics} is needed for \includegraphics
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.7\textwidth]{Figs/ControlSimulator/deterministic_example.png}
-  \caption[]{An example trajectory for the deterministic dynamics. Above is the
-  state evolution, $x(t)$ given $\b=0$ and below is the applied control,
-  $\a(t)$. }
-  \label{fig:deterministic_soln_controls}
-\end{center}
-\end{figure}
-
 \section{Open-loop Stochastic Control}
-As we have mentioned before, when the value of $X_t$ is unknown, the best one
+As we have mentioned before, when the value of $X_t$ is unknown , the best one
 can do is to use the transition density to perform the optimization. Since the
 transition density follows a (deterministic) PDE, we will thus apply a Maximum
-Principle for PDEs as a method of obtaining the optimal control.
+Principle for PDEs as a method of obtaining the optimal
+control, \cite{Ahmed1981,Borzi2012}.
 
 \subsection{Fokker-Planck equation for the state density evolution} 
-
 We will need the transition density of $X$, conditional on no
@@
 In theory, $\xmin = -\infty$, but in the numerics below we will need to
-truncate it to some finite value.
-
-The probability flow, $\phi$, here is
+truncate it to some finite value, exactly as in the HJB equation,
+\cref{eq:OC_LS_HJB_full}.
+
+The $f$ dynamics can also be written as
+$$
+\di_t f(x,t) = - \di_x \phi(x,t)
+$$
+for the probability flow, $\phi$, 
 $$
@@
 We will also need a short hand notation for the differential operator on the
-right side of \cref{eq:FP_pde_OU_PDF}. Let
-$$
-\L_\a[\cdot] := \di^2_x [ \frac{\b^2 }{2} \cdot] -  
-				\di_x  [ (\a(t)- \frac{x}{\tc}) \cdot]
-$$
-This is a linear and bounded differential operator. . We will usually omit the
-subscript $\a$, but have written it now, since the operator is parametrized by
-$\a$, that is by the control.
+right side of \cref{eq:FP_pde_OU_PDF}. Let $$ \L_\a[\cdot] := \di^2_x [
+\frac{\b^2 }{2} \cdot] - \di_x  [ (\a(t)- \frac{x}{\tc}) \cdot] $$ We will
+usually omit the subscript $\a$, but have written it now, since the operator is
+parametrized by $\a$, that is by the control.
 
@@
 Rewriting this in terms of $f$ reads:
-\begin{multline}
-J(\a) = 
-\int_\xmin^\xth \Ttwo(x) f(x,\T) \intd{x}
-\\
-+ \int_0^\ts \phi(\xth, s) (s-\T)^2 \intd{s}
-\\
-+  \e \int_0^\ts  \a^2(s)  \int_\xmin^\xth f(x,s) \intd{x} \intd{s}
+
+\begin{align}
+J(\a) =& 
+\int_\xmin^\xth \Exp[\t^2|x, \amax] \cdot f(x,\T) \intd{x}
+\notag
+\\
+&+ \int_0^\T \phi(\xth, t) (t-\T)^2 \intd{t}
+\notag
+\\
+&+  \e \int_0^\ts  \a^2(t)  \int_\xmin^\xth f(x,t) \intd{x} \intd{t}
 \label{eq:OC_LS_variance_density}
-\end{multline}
-
+\end{align}
 Let us explain in more detail each term on the RHS in
-\cref{eq:OC_LS_variance_density}:
-
-The first term
-$$
-\int_\xmin^\xth \Ttwo(x) f(x,\T) \intd{x}
-$$
-counts the cost of trajectories which spike too late:
-this cost is the expected squared time-to-hit starting at $x$, with  $\a
-= \amax$, weighted by the probability of $X_\T = x$ , $f(x,\T)$. Recall
-that we calculated $\Ttwo(x)$ in \cref{sec:calculating_TCs}.
+\cref{eq:OC_LS_variance_density}.
+
+The first term $$ \int_\xmin^\xth \Exp[\t^2|x, \amax] \cdot f(x,\T) \intd{x} $$
+counts the cost of trajectories which spike too late. This cost is the expected
+squared time-to-hit starting at $x$, with  $\a = \amax$, weighted by the
+probability of $X_\T = x$, which is just $f(x,\T)$.
 
@@
 $$
-\int_0^\ts \phi(s,t) (s-\T)^2 \intd{s}
+\int_0^\T \phi(s,t) (t-\T)^2 \intd{t}
 $$
@@
 $$
-\e \int_0^\ts  \a^2(s)  \left[  \int_\xmin^\xth f(x,s) \intd{x} \right] \intd{s}
+\e \int_0^\ts  \a^2(t)  \left[  \int_\xmin^\xth f(x,t) \intd{x} \right] \intd{t}
 $$
@@
 
-As in the ODE theory, we will introduce an adjoint function, $p$, which will be
-crucial in setting up necessary conditions for the optimal control.
-
-The basic formulation for the adjoint PDEs and the abstract Hamiltonian can be
-found in \cite{Fattorini1999,Palmer2011}. However our cost functional is a
-little strange as it involves values of $\di_x f$ at the boundary. As such we
-will derive the optimality condition heuristically, appealing to the
-references for rigorous proofs.
-
-Before proceeding, one might wish to read
-\cref{sec:Pontryagin_heuristic_derivation} for a simpler version of the
-following derivation with exactly the same ideas.
-
-We start by assuming that $f,\a$ are the optimum trajectory/control pair and we
-ask what necessary conditions must be true for $f,\a$. In particular if we take
-a small variation around $\a$, we will have:
-\begin{align*}
-\a_\e = \a + \e \da
-\\
-f_\e = f + \e \df
-\end{align*}
-
-With that the variation of the cost with respect to the control at the optimal
-$\a$ can be calculated as:
-\begin{equation}
-\frac {dJ}{d\e} \Big|_{\e = 0} =
-- \int_0^\T (s-\T)^2 \frac {\b^2}{2} \di_x  \df \intd{s}  
-+ \int_\xmin^\xth \Ttwo\df\intd{x}
-+ \tilde{\e} \int 2 \a \da \intd{s}  
-\label{eq:J_variation_PDE}
-\end{equation}
-This must be non-negative for a true minimum, but before we can say more, we
-need to eliminate $\df$ from $\frac {dJ}{d\e}$! In order to eliminate $\df$
-from \cref{eq:J_variation_PDE} we will introduce an adjoint variable, $p$. What
-we would want from this adjoint variable $p$ would be that:
-\begin{equation}
-p(x,\T) = \Ttwo(x)
-\label{eq:adjoint_TC_desideratum}
-\end{equation}  
-and 
-\begin{equation}
-\di_t \left[ \int_\xmin^\xth  p \cdot\df \intd{x} \right] = (s-\T)^2 \frac
-{\b^2}{2}
-\di_x
-\df
- + \text{stuff}
-\label{eq:adjoint_dynamics_desideratum}
-\end{equation}
-where 'stuff' does not depend on $\df$.
-
-If we had such a $p$ satisfying
-\cref{eq:adjoint_TC_desideratum,eq:adjoint_dynamics_desideratum} then
-\cref{eq:J_variation_PDE} would read:
-\begin{equation}
-\frac {dJ}{d\e} \Big|_{\e = 0} =
-\tilde{\e} \int 2 \a\da  \intd{s} +   \int_0^\T  \text{'stuff'}\intd{s}
-\label{eq:J_variation_PDE_nodf}
-\end{equation}
-which does not depend on $\df$! and we would be on our way. In order to find
-$p(x,t)$ we will need to
-\begin{enumerate}
-  \item calculate $\di_t \df$
-  \item specify $\di_t p $ st. \cref{eq:adjoint_dynamics_desideratum} holds.
-\end{enumerate}
-
-\subsection{The PDE for $\df$}
-First, calculate the PDE that $\df$ satisfies:
-\begin{multline}
-\di_t f_\e =  \L_{\a_\e} [f_\e]
-\\
-= -\di_x [ (\a_\e - \tfrac{x}\tc) f_\e  - D \di_x f_\e]
-\\
-=  -\di_x [ (\a + \e\da - \tfrac{x}\tc) f_\e  - D \di_x f_\e]
-\\
-=  -\di_x [ (\a + \e\da - \tfrac{x}\tc) (f + \e \df)  - D \di_x (f + \e \df)]
-\\
-=  -\di_x [ (\a + \e\da - \tfrac{x}\tc) (f)  - D \di_x (f)]
-- \di_x [ (\a + \e\da - \tfrac{x}\tc) (\e \df)  - D \di_x (\e \df)]
-\\
-= \L_\a f  -\di_x [ \e\da\cdot  f]
-  +\e \L_\a \df + o(\e) 
-\\
-= \di_t f + \e \di_t \df
-\end{multline}
-Of course, we have $\di_t f = \L_\a f$, so we conclude that:
-\begin{equation}
-\di_t \df = \L_\a \df -\di_x [\da \cdot f] 
-\end{equation}
-Now what are the ICs? Well, both $f_\e$ and $f$ must satisfy the system ICs in
-\cref{eq:FP_pde_OU_PDF}, so $$\df (x,0) = 0$$.
-Finally, what are the BCs?
-On the upper side:
-$$
-f_\e |_\xth = 0 = [ f  + \e \df |_\xth = 0 + \e \df|_\xth \implies \df |_\xth =
-0 $$
-While on the lower side:
-\begin{align*}
-0 =& [ (\a_\e - x/\tc)f_\e - D \di_x f_\e |_\xmin
-\\
-=& (\a +\e \da - x/\tc)(f + \e \df ) - D \di_x (f + \e \df) |_\xmin
-\\
-=& [(\a +\e \da - x/\tc)(f) - D \di_x (f) |_\xmin
-\\
-& + \e [ (\a +\e \da - x/\tc)(\df) - D \di_x (\df) |_\xmin
-\\
-=& \cancel{[(\a - x/\tc)f - D \di_x (f) |_\xmin}
-\\
-&+ \e \da f|_\xmin
-\\
-&+ \e [ (\a - x/\tc)(\df) - D \di_x (\df) |_\xmin   + o(\e)
-\end{align*}
-
-so $\df$ satisfies the following system of equations:
-\begin{equation}
-\begin{gathered}
-\begin{aligned}
-\di_t \df 
-&= \L_\a \df -\di_x [\da \cdot f]
-\\
-&= -\di_x \left[ (\a - \tfrac x\tc)\df - \tfrac{\b^2}2 \di_x \df \right]
--\di_xf \cdot\da
-\end{aligned}
-\\
-\begin{cases}
-\df(x,0) &= 0
-\\
-[ (\a - \frac x \tc)(\df) - D \di_x (\df) |_\xmin &= -\da \cdot f|_\xmin
-\\
-\df|_\xth &= 0
-\end{cases}
-\end{gathered}
-\label{eq:FP_variation_near_optimum}
-\end{equation} 
-With \cref{eq:FP_variation_near_optimum} we are ready to design the dynamics of
-$p$ such that $\df$ is eliminated from the optimality equation,
-\cref{eq:J_variation_PDE}.
-
-\subsection{Calculating the adjoint }
-Recall our desideratum, we want
-$$
-p(x,\T) = \Ttwo(x) \quad (\text{required TCs})
-$$
-and
-$$
-\int_0^\T \di_t (p \cdot \df) = \int_0^\T(s-\T)^2 \frac {\b^2}{2} \di_x 
-\df \intd{s} + \text{stuff}
-$$
-This is equivalent to having
-$$
-\di_t (p \cdot \df) =  (s-\T)^2 \frac {\b^2}{2} \di_x 
-\df  + \text{stuff}
-$$
-Let us calculate $\di_t (p \cdot \df)$
-\begin{align*}
-\di_t (p \cdot \df) =& \df \cdot \di_t p + p \cdot   \di_t \df
- \\
- =& \df \cdot \di_t p + p \cdot \left( \L [\df] -\di_xf \cdot\da \right)  
- \\
- =& \df \cdot \di_t p + p \cdot \L  [\df] - \underbrace{ p \cdot \di_xf \cdot\da}_{\text{stuff}}
-\end{align*}
-Now our job is clear - we want 
-$$
-\di_t p = -\Lstar[p]
-$$
-Where $\Lstar$ is defined by:
-\begin{equation}
-\begin{gathered}
-\int_\xmin^\xth  \Lstar [p] \df \intd{x} 
-=
-\int_\xmin^\xth p \L [\df] \intd{x} - (s-\T)^2 \frac {\b^2}{2}\di_x \df +
-\text{possibly more stuff}
-\\
-\\
-\forall \df \quad \st \begin{cases}
-\df(x,t) |_{x=\xth} & = 0 
-\\
-(\a - \tfrac x \tc)\df - \di_x \tfrac {\b^2}2 \df \big|_{x=\xmin} &=  -\da
-\cdot f
-\end{cases}
-\end{gathered}
-\end{equation}
-where 'possibly more stuff' should be independent of $\df$.
-
-Let us take a deep breath and calculate $\Lstar$:
-\begin{align*}
-\int p \L [\df] \intd{x} =&
-\int p \underbrace{\left(\frac{\b^2 }{2}\cdot \di^2_x \df -  
-			 \di_x \big[ (\a- \frac{x}{\tc})\cdot \df \big]\right)}_{\L [\df]} \intd{x}
-\\
-=& [ -p ((\a- \frac{x}{\tc})  \cdot \df)\quad  \vert^\xth_\xmin
-+ \int \di_xp ((\a- \frac{x}{\tc})  \cdot \df) \intd{x}
-\\
-& + [p (\frac{\b^2 }{2}\cdot \di_x \df \quad \vert ^\xth_\xmin
-- [ \di_x p \cdot \frac{\b^2 }{2}  \df\quad \vert^\xth_\xmin 
-+ \int \frac{\b^2 }{2} \di^2_xp    \cdot \df \intd{x} 
-\\
-=& \int \Lstar [p] \df \intd{x} 
-+\underbrace{[-p(\a- \frac x \tc) \cdot \df 
-+ p \frac{\b^2}{2} \di_x \df 
--\di_x p \frac{\b^2}{2} \df \,\vert^\xth_\xmin }_{\text{BC terms}}
-\end{align*}
-
-Now we want to make the 'BC terms' equal to $(s-T)^2 \frac{\b^2}{2} \di_x \df$
-+ 'possibly more stuff'.
-In doing so, we will discover the correct BCs associated with the adjoint
-operator, $\Lstar$:
-\begin{align*}
-\textrm{BC terms} =& 
-\Big[ -p (\a- \frac{x}{\tc})  \cdot \df 
-+ p \frac{\b^2 }{2}\cdot \di_x \df   
-- \di_x p \cdot \frac{\b^2 }{2}  \df   \quad ^\xth_\xmin\vert 
-\\
-=& -\cancel{p (\a- \frac{x}{\tc})  \cdot \df \Big|_\xth}
- + p (\a- \frac{x}{\tc}) \cdot \df \Big|_\xmin
-\\ 
-& + [p \frac{\b^2 }{2}\cdot \di_x \df \Big|_\xth
- - [p \frac{\b^2 }{2}\cdot \di_x \df \Big|_\xmin
-\\
-&- \cancel{[ \di_x p \cdot \frac{\b^2 }{2}   \df\Big|_\xth}
- + [ \di_x p \cdot \frac{\b^2 }{2}   \df\Big|_\xmin
-\\
-=& \cancelto{-p (-\da f)}{-p ((\a- \tfrac x \tc)\df - \di_x \tfrac {\b^2}2
-\df) \Big|_\xmin} 
-+ [p \frac{\b^2 }{2}\cdot \di_x f \Big|_\xth
-+ [ \di_x p \cdot \frac{\b^2 }{2}   f\Big|_\xmin
-\\
-=& (s-T)^2 \frac{\b^2 }{2}\cdot \di_x \df\Big|_\xth 
-+  p (\da f)\Big|_\xmin
-\end{align*}
-
-Since we want this to be true for arbitrary $\df$, we need to enforce the
-following (adjoint) BCs: 
-$$
-\begin{cases}
-p |_\xth &= (s-T)^2
-\\
-\di_xp |_\xmin &= 0
-\end{cases}
-$$
-
-In summary, the adjoint PDE equation is given by: 
+The Maximum Principle introduces an adjoint variable, $p$, which solves a PDE
+related to the PDE satisfied by the density $f$ and then calculates the optimal
+control, $\a$ as a function(al) of $f$ and $p$.
+
+In short, the equation for the adjoint, $p$, is 
 \begin{equation}
@@
 $$
-gives us the direction of increase
-and we can use it as a gradient in a descent algorithm. 
-This we can also write as:
-\begin{align*}
-\di_\a \H =&  \Big\{
- \tilde{\e}  2 \a(t)
-+ p(x,t) f(x,t) \Big|_\xmin
-- [p(x,t) f(x,t) \, _\xmin^\xth\Big|
-+ \int _\xmin^\xth \di_x p \cdot f(x,t) \intd{x} 
-\Big\}
-\\
-=&
- \Big\{
- \tilde{\e}  2 \a(t)
-- p(x,t) f(x,t)\Big|_\xth
-+ 2p(x,t) f(x,t) \Big|_\xmin
-+ \int _\xmin^\xth \di_x p \cdot f(x,t) \intd{x} 
-\Big\}
-\end{align*}
-which has certain numerical advantages.
-
-So the plan is: we choose an initial $\a_1(t)$ - the deterministic solution
-from \cref{sec:deterministic_ode_control} is a natural choice. Then solve for $f,p$.
-Then evaluate $J$ and iterate until convergence. For our convergence
-criterion we will take that the incremental difference in $J$ is smaller than
-some difference small tolerance. The algorithm is specified more
-precisely in Algorithm \ref{alg:hamiltonian_descent_4_PDF_OC}.
-\begin{algorithm}
-\begin{algorithmic}
-\State $J_0 = \infty $ 
-\State $\a_1 = $ deterministic $\a$, the optimal control given $\b=0$ from
-\cref{sec:deterministic_ode_control}
-\For { $k= 1\dots k_{\max}$}
-	\State $\di_t f_{k} = \L_{a_{k}} [ f_{k}]$
-	\State $-\di_t p_{k} = \Lstar_{a_{k}} [ p_{k}]$
-	\State $J_{k} \gets J[\a_k, f_k]$
-	\If {$|J_{k} - J_{k-1}| < tol $}
-		\State BREAK
-	\EndIf
-	\State $\da = \di_a \H(f_{k} ,p_{k},  \a_{k})$
-
-	\If {$\sup_t |\da(t)| < tol$}
-		\State BREAK
-	\EndIf
-
-	\State $\a_{k+1} \gets \a_{k} - s \cdot \da  $
-\EndFor
-\State \Return $J_k, \a_k$
-\end{algorithmic}
-\caption{Optimal Control Descent Algorithm}
-\label{alg:hamiltonian_descent_4_PDF_OC}
-\end{algorithm}
-
-We show one realization of $\di_a \H$ in \cref{fig:grad_aH_deterministic_alpha}.
-\begin{figure}[htp]
-\begin{center}
-\subfloat[$\a_0$]
-{\includegraphics[width=.49\textwidth]{Figs/FP_Adjoint/Deterministic_alpha_di_aH.png}
- }
-  \subfloat[$\a_1$]
-{  \includegraphics[width=.49\textwidth]{Figs/FP_Adjoint/First_control_iteration.png}
-}
-\caption[labelInTOC]{Iteratively calculating $\a$ for the
-open-loop optimal control. Plotted is an example of
-$-\di_a \H$ calculated at $\a(t) = \a_1(t)$, where $a_1$ is the deterministic control. The green curve, $-\di_a \H$
-indicates the direction that we want to change the control, if it is negative
-we decrease and if it is positive we increase it. On the right panel, b), we
-show the result of applying the incremental change by also plotting the
-resulting $\a_1(t)$. }
-\label{fig:grad_aH_deterministic_alpha}
-\end{center}
-\end{figure}
+gives us the direction of increase of $J$ at $\a(t)$
+and we can use it as a gradient in a descent algorithm, given some initial
+guess, $\a_0(t)$, for the control.
 
@@
 \cref{fig:FP_adjoint_objective_control_convergence}.
-
 \begin{figure}[h]
@@
 open-loop stochastic control is not as good as the feedback stochastic
-control, it is much closer in performance to it than to the blind
+control, it is much closer in performance to it than to the naive
 deterministic control.
@@
 
-\section{Research Goals}
-I hope to be able to use the techniques of open-loop stochastic control
-discussed herein to address problems in stochastic control where for whatever
-reasons, the dynamic programing formalism is not applicable.
-
-One such reason might be that the state is unobservable. An example of such a
-system is the control of blood glucose for diabetes or intensive care
-patients, where the patients' blood glucose is modelled to follow a stochastic
-differential equation and actual observations are expensive and invasive and
-thus rare, on the order of a few hours. In this scenario, the applied external
-glucose or insulin to the patient in the time in-between measurements can be
-thought of as the solution to an optimal control problem where the objective is
-to minimize deviations of the patient's blood glucose from some baseline.
-
-An alternative reason to use the open-loop control techniques is when optimizing
-with multi-dimensional systems at which point the dynamic programing approach
-suffers from 'the curse of dimensionality' which is the colloquial name for the
-exponential increase in necessary computing resources. If the optimality
-conditions for the open-loop control lend themselves to a Feynman-Kac-type
-representation, then one might replace solving the PDEs by an appropriate
-Monte-Carlo process and thus make the computations tractable.
-
-\clearpage
-\appendix
-\section{Heuristic Derivation of Pontryagin's Minimum Principle}
-\label{sec:Pontryagin_heuristic_derivation}
-Here we will show a heuristic derivation for the Pontryagin Minimum Principle
-for linear ODE dynamics. It should motivate the manipulations in
-\cref{sec:PDE_max_principle_for_pdf}.
-
-The following is almost a direct transcription of A.1 in
-\cite{Evansb}, with the added simplification of linear dynamics
-and linear cost and with the notation adapted to what we have used
-hereto.
-
-The dynamics are given by:
-$$
-\dot{x} = A(t, \a) x
-$$
-where $A$ is a matrix in $\R^{n_x \times n_x}$ whose values depend on the
-time, $t$ and the control $\a$. The cost, $J$, is given by:
-$$
-J[\a] = \int_0^\tf L(t,\a) \cdot x \intd{s} + M \cdot x(\tf)
-$$
-where $L(t,\a), M$
-are vectors in $\R^{n_x}$. Both the dynamics and the cost are linear in the state $x$.
-Suppose $x,\a$ are the optimal trajectory - then small variations in $\a$ can
-only worsen $J$. Let such an acceptable variation be given by:
-$$
-\a_\e = \a + \e \b
-$$
-for some small $\e$. We will have a corresponding dynamic, $x_\e$ st:
-$$
-x_\e = x + \e \dx+ o(\e)
-$$
-Now obtain a differential equation for the variation $\dx$:
-\begin{align*}
-\dot{x_\e} &=  \dot{x} + \e \dot{\dx} 
-\\
-&= A(\a_\e) x_\e = A(\a) x + \e \dot{\dx}
-\end{align*}
-Now divide through by $\e$ and let it go to zero:
-$$
-\implies \dot{\dx} = A \dx + [\grad_\a [ A(t,\a)x]] \cdot\b
-$$
-Note that $$\dx(0) = 0$$ since $x(0)$ is specified and so there is no
-variation there.
-
-With that the derivative of $J$ at $\a$ will read:
-\begin{multline}
-\frac {dJ}{d\e} \Big|_{\a} = \lim \frac{ J[\a_\e] - J[\a]}{\e} 
-= \int L(\a,t) \dx +  \grad_\a L \cdot \b \intd{s} + M \dx(\tf)
-\geq 0 
-\label{eq:J_variation_ODE}
-\end{multline}
-The problem is that the state variation $\dx$ also appears in this derivative,
-whereas we just want the control variation $\b$ to appear there.
-
-The trick then is to introduce the adjoint variable $p$ with appropriate
-dynamics such that all terms involving $\dx$ in \cref{eq:J_variation_ODE} are
-eliminated.
-
-'Clearly', we want to replace the term $ M \cdot \dx $ first. Thus we set 
-$$
-p(\tf) = M
-$$
-
-Now notice the following:
-$$p(\tf)\dx(\tf) = \cancelto{0}{p(0)\dx(0)} + \int_0^\tf d_t (p \cdot \dx)$$
-
-So if we make $\int d_t (p \cdot \dx) = - \int r(\a,t) \dx$ we would be
-eliminate the $\dx$ terms in the expresion for $dJ/d\eps$ entirely!
-
-Now the way is clear:
-
-Let 
-$$ 
-\dot{p} = -A^T p - L(\a,t)
-$$
-Then indeed:
-\begin{multline}
-d_t (p \cdot \dx) = \dot{p} \dx + p \dot{\dx}
-\\
-= (-A^T p - r)\dx + p (A\dx+ [\grad_\a [ A(t,\a)x]] \cdot\b)
-\\
-= -r\dx + p ([\grad_\a [ A(t,\a)x]] \cdot\b)
-\end{multline}
-That is almost what we wanted and it will work, since it only contains
-variations wrt the control $\b$, not the state, $\dx$.
-
-Thus
-\cref{eq:J_variation_ODE} is transformed to read:
-\begin{multline}
-\frac {dJ}{d\e} \Big|_{\a} = 
-\int r(\a,t) \dx +  \grad_\a r \cdot \b \intd{s} + \psi \dx(\tf)
-\\
-=
-\int r(\a,t) \dx +  \grad_\a r \cdot \b \intd{s} + p(\tf) \dx(\tf)
-\\
-= \int r(\a,t) \dx +  \grad_\a r \cdot \b \intd{s} - \int r \dx - p ([\grad_\a [
-A(t,\a)x]] \cdot\b)
-\\
-=
-\int_0^\tf  
-[\underbrace{p ([\grad_\a [A(t,\a)x]] ) + \grad_a r)}_{\di_\a \H} ] 
-\b \intd{s}
-\end{multline}
-As always in these variational arguments, the final point is that since $\b$ is
-arbitrary, we must have that the rest of the integrand be identically zero -
-thus we see where the control theory Hamiltonian, $\H$, comes from.
-
-Except for the nitty-gritty manipulations of the BCs in
-\cref{sec:PDE_max_principle_for_pdf}, what is done there is exactly what is done
-here - a heuristic derivation for a necessary condition for a minimum.
-
 \bibliographystyle{plain}
-\bibliography{library,local}
+\bibliography{library}
 

### 2013-08-29 draft.tex
Commiting old multi-train figs before pruning down

@@
-% \documentclass[12pt]{iopart}
-\documentclass{article}
+\documentclass[12pt]{iopart}
+%\documentclass{article}
 
@@
 
-\newtheorem{thm}{Theorem}[section]
-\newtheorem{lemma}{Theorem}[thm]
+\usepackage{iopams}
+
+% \newtheorem{thm}{Theorem}[section]
+% \newtheorem{lemma}{Theorem}[thm]
 % \theoremstyle{definition}
-\newtheorem{ex}{Example}[thm]
-\newtheorem{defn}{Definition}[thm]
+% \newtheorem{ex}{Example}[thm]
+% \newtheorem{defn}{Definition}[thm]
 
@@
 
-\title{Stochastic Optimal Control of Single Neuron Spike Trains
-\\
----
-\vskip5pt
-Draft
+\title{Stochastic Optimal Control of Single Neuron Spike Trains}
+\author{Alexandre Iolov$^1$, Susanne Ditlevsen$^2$, Andr\'e Longtin$^1$
 		}	
-\author{Alexandre Iolov, Susanne Ditlevsen, Andr\'e Longtin,
+\address{%
+     (1) Department of Mathematics and Statistics, University of Ottawa, Ottawa,
+    Canada
+\\
+    (2) Department of Mathematical Sciences, University of Copenhagen,
+Copenhagen, Denmark
 }
+\ead{aiolo040@uottawa.ca}			
 % \address{Department of Mathematics and Statistics, University of Ottawa, Ottawa, Canada 
@@
 
-\date{\today}
+% \date{\today}
+
+\begin{abstract}
+
+{\em Objective.} External control of exact spike timings in single neurons is
+important for basic understanding of neuronal activity, and has potential in
+brain-machine interfaces, for improving neural prostheses, and in medical
+applications related to neuronal disorders, such as Parkinson's disease or
+epilepsy. The goal of this paper is the design of optimal controlled electrical
+stimulation patterns to a cell to achieve a target spike train under
+physiological constraints not to damage tissue and minimize possible side
+effects. {\em Approach.} We pose an optimal control problem to precisely specify
+the spike timing in a noisy leaky integrate-and-fire model of a neuron. In
+particular, we allow for the noise to be of arbitrary intensity. The optimal
+control problem is solved using Dynamic Programming if the controller has access
+to the voltage (closed-loop control) and using a Maximum Principle for the
+transition density if the controller only has access to the spike timings (open
+loop control). {\em Main results.} We have developed a stochastic optimal
+control scheme to obtain precise spike timings, which is not limited by the
+intensity of the noise in the system. Given this scope, the numerical demands
+are non-trivial, but simulations show that the algorithms are potentially
+feasible. It however remains to see if they can be successfully applied in a
+real-time experimental setting. Furthermore, we show how wrong the control
+strategy can perform if we assume that the noise is negligible. {\em
+Significance.} More precise targeting and control of neural activity to achieve
+optimal performance in a noisy environment, in which most neurons are embedded,
+have great potential in health care. The main contribution is the online
+feedback control of a noisy system through modulation of the input, taking into
+account physiological constraints on the control, as well as incurred costs to
+diminish adverse effects.
+
+\end{abstract}
+% \tableofcontents MSC KEYS: 49N90 90C39 93E20 93C20 93C95
 
 \maketitle
-
-\abstract{
-{\em Objective.} External control of exact spike timings in single neurons is
-important for basic understanding of neuronal activity, and has
-potential for improving neural prostheses and in medical applications
-related to neuronal disorders, such as Parkinson's disease or epilepsy.
-The goal of this paper is the design of optimal controlled
-electrical stimulation patterns to a cell to achieve a target
-spike train under physiological constraints not to damage tissue and
-minimize possible side effects.
-{\em Approach.} We pose an optimal control problem to precisely specify
-the spike timing in a noisy leaky integrate-and-fire model of a
-neuron.
-In particular, we allow for the noise to be of arbitrary intensity.
-The optimal control problem is solved using Dynamic Programming if the controller has access to the voltage (closed-loop control) and using a Maximum Principle for the transition density if the controller only has access to the spike timings (open loop
-control).
-{\em Main results.} We have developed a stochastic optimal control
-scheme to obtain precise spike timings, which is not limited
-by the intensity of the noise in the system. Given this scope, the
-numerical demands are non-trivial, but simulations show that the
-algorithms are potentially feasible. It however remains to see if
-they can be successfully applied in a real-time experimental setting.
-Furthermore, we show how wrong the control strategy can perform if we
-assume that the noise is negligible.
-{\em Significance.} More precise targeting and control of neural
-activity to achieve optimal performance in a noisy environment, in
-which most neurons are embedded, have great potential in health
-care. The main contribution is the online feedback control of a noisy
-system through
-modulation of the input, taking into account physiological constraints
-on the control, as well as incurred costs to diminish adverse effects.
-}
-
-\tableofcontents
-
-MSC KEYS: 49N90 90C39 93E20 93C20 93C95
 
@@
 minimum. A similar approach is taken in \cite{Dasanayake2011}. They do not
-consider noise, though. In \cite{Nabi2013}, the methods are implemented on brain
-slices of pyramidal neurons of rat hippocampus. [[[ALEX, Is it closed- or open
-loop? Could you SHORTLY describe the difference of what they do and what we
-do?]]] \{ALEX: They use Phase Response Curves, assuming that a spike coincides
-with $\th=0$; no noise, Pontryagin Maximum Principle, solved numerically $=>$
-open-loop. Short Description: Since they do not consider noise, it is
-straight-forward to make the neuron spike when you want it, and in fact that can
-be done in many ways. Thus their optimization reduces to finding the way to make
-the neuron spike at a pre-specified time, using least energy (the integral of
-$u^2$.\}. While the Phase Response Curve Model is a parsimonious and effective
-way to describe a neuron's response to a stimulus, it is only valid in the
-supra-threshold regime, when the unstimulated neuron is periodically spiking.
-In \cite{Feng2003} they use a leaky integrate-and-fire model and also
-use open-loop control, but to obtain exact spike timings, they choose an
-objective of minimizing the variance of the membrane potential at the desired
-spike time, while forcing the mean of the membrane potential at this time point
-to equal the threshold. This provides exact solutions since it does not involve
-first-passage times, but has the drawback of getting a non-negligible
-probability of being way-off the objective.
+consider noise, though, and only work with open-loop control. In
+\cite{Nabi2013}, the methods are implemented on brain slices of pyramidal
+neurons of rat hippocampus. While the
+Phase Response Curve Model is a parsimonious and effective way to describe a
+neuron's response to a stimulus, it is only valid in the supra-threshold regime,
+where the unstimulated neuron is periodically spiking. In \cite{Feng2003} they
+use a leaky integrate-and-fire model and open-loop control. To
+obtain a given spike time, they choose an objective of minimizing the
+variance of the membrane potential at the desired spike time, while forcing the mean of
+the membrane potential at this time point to equal the threshold. This provides
+exact solutions since it does not involve first-passage times, but has the
+drawback that there is a non-negligible probability that the obtained
+spike time will be far from the desired spike time.
 
@@
 minimize intervention.
-Our main contributions are... (((what we have done which has not been
-done before, numerical merits...)))
-\{ALEX: Basically, we deal with a significant noise component. Everything
-else follows, i.e. the objective only makes sense if the ISI is a random variable.
-The PDEs (HJB and Max Principle) only occur because of the noise
-component, o/w we would have ODEs. The numerical schemes are simple given the
-PDEs, we just use the Crank-Nicholson method for the PDEs and a basic
-gradient-descent for the optimization in the open-loop case. The numerics take a
-non-trivial time to run - hard to see them coming down to real-time, i.e.
-10-20 ms, \ldots but possible. We also explore a wide spectrum of parameter
-values, in particular we do  not restrict our attention to the
-autonomously-spiking, supra-threshold regime.\}
-
-[[[Susanne, we should mention Whittle \cite{Whittle1996}, but here, or together
-with other monographs on optimal control?]]] \{ALEX: Indeed we should mention
-Whittle, in particular mention that he discusses analytically the problem in the
-closed-loop case (only), but not the open-loop case. He does not provide
-numerics in either case. I would refrain from citing other books on
-stochastic optimal control, here. Books commonly cited are
-\cite{Fleming1975,Krylov2008} and the chapter on HJB equation in Oksendal,
-\cite{Oksendal2007}, some of them are cited in later sections\ldots\}
+Our main contributions are that we specifically allow for a non-negligible
+noise component in the neural activity when calculating the control, and
+consider both open- and closed-loop controls. In particular we do not
+restrict our attention to the autonomously-spiking, supra-threshold regime.
 
@@
 \end{equation}
-Here $X$ represents the membrane electric potential which tries to revert to $0$
-with a time-const of $\tc$, $dW$ is a Brownian motion increment scaled by $\b$
-and $\Iext(t)$ is the deterministic external input to the cell. Having last
+Here $X$ represents the membrane electric potential which decays to $0$
+with a time constant of $\tc$, $dW$ is a Brownian motion increment scaled by
+$\b$ and $\Iext(t)$ is the deterministic external input to the cell. Having last
 spiked at time $0$, the potential hits $\xth$ at some time $\ts$, the potential
-resets to $0$ and we start all over again. We write $\ts^+$ in the formula for
-the reset since we are working in continuous time, i.e.\ it is really the
-voltage's limit from the right at $\ts$ that is 0. For our purposes, we will
-always have that the threshold is set to one, $\xth = 1$.
+resets to $0$ and the process starts all over again. 
+We write $\ts^+$ for the voltage's limit from the right at $\ts$.
+Throughout the paper the threshold is set to one, $\xth = 1$.
 
@@
 A natural goal is to attempt to control the spike time, $\ts$.
-That is we would like to ask how do we choose $\a(t)$ such that $\ts = \T$.
+That is how do we choose $\a(t)$ such that $\ts = \T$.
 A natural optimal control objective to achieve this is the least squares solution
@@
 \begin{equation}
-\a(t) \in [\amin, \amax] \quad \forall t
+\a(t) \in [\amin, \amax] \quad \forall t.
 \label{eq:bound_constraints_alpha}
@@
 running cost based on the control. This regularizes the problem eliminating the
-subtleties of singular-control situations and it might be argued appropriate to
-avoid excessive control. WE SHOULD DISCCUSS THIS IN GREATER DETAIL, BUT PERHAPS
-LATER. (ALEX:\cite{Ahmadian2011} state that they impose a similar control
-penalty to avoid excessive charge building up on the cell.)
+subtleties of singular-control situations and it serves to avoid excessive
+control as well as to avoid excessive charge building up on the cell.
 
@@
 \end{align}
-where $\e$ measure how much weight we put on minimizing the energy expenditure.
-If $\e = 0$, then we do not care at all about the expended
-energy.
-
-It will often be the case that we write $\a$ and mean either a function or a
-value of that function at a particular time. And that this function could be
+where $\e$ measures how much weight we put on minimizing the energy cost.
+If $\e = 0$, then we do not care at all about the expended cost.
+
+It will often be the case that $\a$ means either a function or a
+value of that function at a particular time. This function could be
 random, i.e.\ a stochastic process. Since $\a$ is used to control the behaviour
 of $X$, it will be at least measurable with respect to the filtration of $X$,
-see \cite{Krylov2008}. We will try to make clear below when we are refering to
-$\a$ as a stochastic process and when we are merely refering to its particular
+see \cite{Oksendal2007} for the details. We will try to make clear below when we
+are refering to $\a$ as a stochastic process and when we are merely refering to its particular
 value at some time. For example, in \cref{eq:OC_LS_variance_energy}, $\a(\cdot)$
@@
 and can be used in determining the control. In open-loop control only the spike
-times are observable. In the closed-loop context, we might better write the
-control as $$\a = \a(x,t)$$ to indicate its dependence on $X_t$ and to express
-that it will be updated based on the time-course of $X$. 
-In the open-loop context, we would instead write $$\a = \a(t)$$ to indicate that
-$\a(\cdot)$ is decided for all times at time 0. 
-The techniques used in the two scenarios will be different. For the closed-loop
-scenario we will use Dynamic Programing, \cite{Fleming1975}, in order to obtain
-the optimal control, while for the open-loop scenario we will use a form of the
+times are observable. In the closed-loop context, we write the control as $$\a =
+\a(x,t)$$ to indicate its dependence on $X_t$ and to express that it will be
+updated based on the time-course of $X$. In the open-loop context, we write $$\a
+= \a(t)$$ to indicate that $\a(\cdot)$ is decided for all times at time 0. The
+techniques used to obtain the optimal controls in the two scenarios will be
+different. For the closed-loop scenario we use Dynamic Programing,
+see \cite{Fleming1975}, while for the open-loop scenario we use a form of the
 Maximum Principle applied on the transition density of the controlled process,
-\cite{Ahmed1981,Borzi2012}.
+see \cite{Borzi2012}.
 
 Crucially, we assume that the model parameters, $\m,\tc,\b$ in
-\cref{eq:X_evolution_uo} are known.
+\cref{eq:X_evolution_uo} are known. This is a strong assumption and will be
+discussed later.
 
@@
 $\m$, induces spikes in the absence of noise or not. 
-
 Spikes will occur in the absence of noise, if and only
-if $$ \m > \frac 1\tc,$$ which is called the supra-threshold regime. When $\m
+if $ \m > \frac 1\tc,$ which is called the supra-threshold regime. When $\m
 \leq 1/\tc$, the regime is called the sub-threshold regime.
@@
 Given a time $t$ and a realized voltage value $X_t = x$, let $\trem$ be the
-(unknown) remaining time to spike, $\trem = \ts - t$. 
-Starting from an arbitrary $t, x$, our remaining-cost
-objective, $J[\a(\cdot); x,t]$, will be
+(unknown) remaining time to spike, $\trem = \ts - t$. Starting from an arbitrary
+$t, x$, our remaining-cost objective, $J[\a(\cdot); x,t]$, will be
 \begin{equation}
@@
 the difference between $\t = \ts - t$ and ($\T-t$), given the current state $x$.
-Recall that in the closed-loop scenario, we assume that the value of $x = X_t$
-is known to the controller.
+Recall that in the closed-loop scenario, we assume that the value of $X_t$
+is known to the controller. A similar problem is discussed analytically at
+length in the book on optimal control by Peter Whittle, \cite{Whittle1996},
+although there is no discussion there of the numerics required to solve it.
 
@@
 The Hamilton-Jacobi-Bellman (HJB) equation,
-see \cite{Fleming1975,Evansb,Krylov2008}, \{ALEX: Should we also cite
-Moehlis' papers that use/discuss HJB \cite{Nabi2013a,Danzl2009}???\}
-associated with \cref{eq:OC_LS_cost_to_go} is obtained as follows. We introduce the value function, $\v(x,t)$
-% \footnote{Traditionally, the value function is denoted as $v$. We
-% use $\v$ since we are working in a neural context, where $v$ is so often used
-% to denote a voltage}
-:
+see \cite{Fleming1975,Whittle1996} or \cite{Danzl2009,Nabi2013a} for a neuroscience
+context, associated with
+\cref{eq:OC_LS_cost_to_go} is obtained as follows. We introduce the value function, $\v(x,t)$:
 \begin{align}
@@
 
-Then $\v$ satisfies the following HJB partial differential equation (PDE) (see
-\cite{Fleming1975,Evansb,Krylov2008} for details):
+Then $\v$ satisfies the following HJB partial differential equation (PDE):
 \begin{equation}
@@
 parabolic PDE is that it contains an embedded optimisation that depends on the
-solution, $\v$. That is for each $x,t$ in the computational domain,
+solution, $\v$. For each $x,t$ in the computational domain,
 $\a$ is chosen as to minimize $\{\e \a^2 + \left(\m + \a - \tfrac{x}{\tc}\right) \di_x \v \}$.
-Here the we can solve for the optimal control, $\a^*(x,t)$ analytically as:
+Here we can solve for the optimal control, $\a^*(x,t)$ analytically as:
 \begin{align}
@@
 
-Now, we need to consider boundary conditions (BCs) for $\v$. If $X_t = \xth$
-then we have a spike now and we know that $\ts = t$. Thus
-$$ \v(\xth,t) =
-(t-\T)^2.$$ 
-At the threshold, the value function equals the squared difference
-between the desired spike time and the realized one.
-Now for $x \ll .0$, we can assume that $\v$ is not
+We need to consider boundary conditions (BCs) for $\v$. If $X_t = \xth$ then we
+have a spike now and $\ts = t$. Thus $$ \v(\xth,t) = (t-\T)^2.$$ At the
+threshold, the value function equals the squared difference between the desired
+spike time and the realized one. For $x \ll 0$, we assume that $\v$ is not
 significantly affected by the change in $x$, i.e.\ that $$ \di_x \v(\xmin, t) =
 0 $$ for some lower boundary $\xmin$. Such a boundary condition will be
-justified if we impose artificial reflecting boundaries for the diffusion at $\xmin$. So we need to
-determine for which $\xmin$ is a reflecting boundary consistent with the
-dynamics. For example, we can take $\xmin$ to be two standard deviations
-below the mean of the stationary distribution of the maximally
-inhibited, $\a=\amin$, \cref{eq:current_mu_alpha}. That is we set $\xmin =
-\tc(\mu + \amin) - 2 \b / \sqrt{\tc/2}$. In fact, we further enforce that
-$\xmin \leq -0.5$.
+justified if we impose artificial reflecting boundaries for the diffusion at
+$\xmin$. So we need to determine for which $\xmin$ is a reflecting boundary
+consistent with the dynamics. For example, we can take $\xmin$ to be two
+standard deviations below the mean of the stationary distribution of the
+maximally inhibited, $\a=\amin$, \cref{eq:current_mu_alpha}. That is we set
+$\xmin = \tc(\mu + \amin) - 2 \b / \sqrt{\tc/2}$. We further enforce
+that $\xmin \leq -0.5$.
 % Ooops, we need $\di_x\v\neq 0$ and here we are setting $\di_x\v= 0$ in one of
@@
 
-Now we come to the Terminal Conditions (TCs) for $\v$. The idea is simple:
-if we reach $\T$ without having spiked we apply maximum control in the
+To determine the Terminal Conditions (TCs) for $\v$, the idea is
+simple: if we reach $\T$ without having spiked we apply maximum control in the
 positive direction, i.e.\ $$t>T \implies \a(t) = \amax$$ Thus:
@@
 Note that we are making an approximation here
-- that is we are ignoring the energy term, $\eps u^2$ in the objective for
+- we are ignoring the energy term, $\eps u^2$, in the objective for
 $t>T$. Naturally, this approximation is ever more accurate for $\eps \ll 1$.
@@
 easily for all $x$ in the domain by solving a stationary backward Kolmogorov
-equation. We show the details in \cref{sec:valuefun_TCs}.
+equation. We show the details in \ref{sec:valuefun_TCs}.
 
@@
 \subsection{The numerical method for the HJB equation}
+
 We now have a PDE for $\v$ and an algorithm for computing all the BCs and TCs of
-this PDE. It is time to discuss the numerical method for solving \cref{eq:OC_LS_HJB_full}. Since we are in 1-d the simplest thing to do is try a
-centred Finite Difference scheme using Crank-Nicholson to step in time.
-Although, it might become necessary, we do not upwind the first order
-derivatives for now. Also to resolve the non-linearity, $(\di_x\v)^2$, in the
-PDE, we treat it as a mixed implicit-explicit term
-$$(\di_x \v(x,t_k))^2 = \di_x
-\v(x,t_k) \cdot \di_x \v(x,t_k) \approx \underbrace{\di_x \v(x,t_k)}_{\textrm{implicit}}
-\cdot \underbrace{\di_x\v(x,t_{k+1})}_{\textrm{explicit}}.$$ Note that the
-implicit term is in the previous time $t_k$ instead of, as is conventional,
-the next $t_{k+1}$, because we are solving for $\v$ {\sl backwards} in time. 
-TODO: Give justification for this trick.
+this PDE. It is time to discuss the numerical method for solving
+\cref{eq:OC_LS_HJB_full}. Since it is one-dimensional, it is straight forward to
+apply the standard centred finite difference using the Crank-Nicholson scheme to
+step in time. We have not found it necessary to upwind the first order
+derivatives for the regimes we have explored. To resolve the non-linearity,
+$(\di_x\v)^2$, in the PDE, we treat it as a mixed implicit-explicit term
+$$(\di_x \v(x,t_k))^2 = \di_x \v(x,t_k) \cdot \di_x \v(x,t_k) \approx
+\underbrace{\di_x \v(x,t_k)}_{\textrm{implicit}} \cdot
+\underbrace{\di_x\v(x,t_{k+1})}_{\textrm{explicit}}.$$ Note that the implicit
+term is in the previous time $t_k$ instead of, as is conventional, the next
+$t_{k+1}$, because we are solving for $\v$ {\sl backwards} in time.
 
 The numerical scheme is implemented in Python, using the Scipy/Numpy library,
-\cite{Scipy}.
-
-TODO(ALEX): Mention time- space- discretization
+\cite{Scipy}. For the discretization in time and space, we choose $\Delta x$ and
+$\Delta t$ in relation to the parameter regimes. for example in the
+Supra-Threshold Low-Noise regime, we have $\Delta x = 0.0043$ and $\Delta t =
+0.0012$. We consider these to be a fairly fine discretization.
 
@@
 \subsection{Solutions of the HJB equation}
-We set $\T=1.5$ and $\eps = .001$, so that the primary focus is the accurate
+We set $\T=1.5$ and $\eps = 0.001$, so that the primary focus is the accurate
 spiking and energy minimization considerations are secondary.
@@
   mid-point (green) and end-point (red), for each of the four regimes. The desired spike time is set to $\T=1.5$, the energy penalty, $\eps
-  = .001$. The control bounds are $\a \in [-2,2]$. Note that the green and blue
+  = 0.001$. The control bounds are $\a \in [-2,2]$. Note that the green and blue
   curves in E) are lying on top of each other for all $x$.
   \\
-  A,E) Supra-Threshold-High-Noise B,F) Supra-Threshold-Low-Noise 
-  C,G) Sub-Threshold-High-Noise D,H) Sub-Threshold-Low-Noise
+  A,E) Supra-Threshold-Low-Noise
+  B,F) Supra-Threshold-High-Noise
+  C,G) Sub-Threshold-Low-Noise
+  D,H) Sub-Threshold-High-Noise
   }
@@
 \caption{The numerical solution for the value function $\v(x,t)$ HJB PDE,
-\cref{eq:OC_LS_HJB_full}, for the four different parameter regimes. Recall that
-the solution is solved backwards in time.
+\cref{eq:OC_LS_HJB_full}, for the four different parameter regimes.
 The desired spike time is set to $\T=1.5$.
@@
 \\
-A) Supra-Threshold-High-Noise B) Supra-Threshold-Low-Noise 
-C) Sub-Threshold-High-Noise   D) Sub-Threshold-Low-Noise  }
+A) Supra-Threshold-Low-Noise
+B) Supra-Threshold-High-Noise
+C) Sub-Threshold-Low-Noise
+D) Sub-Threshold-High-Noise  }
 \label{fig:HJB_4regimes_value_surf}
@@
 interval, $\v(x,\T)$ is monotonically decreasing in $x$, which is to be expected
-given its terminal condition in \cref{eq:valuefun_TCs}. That is the lower the
+given its terminal condition in \cref{eq:valuefun_TCs}. That is, the lower the
 value of $X_\T$ the more on average will we have to wait for the spike to occur.
@@
 decreases for $x \ll \xth$, and it becomes negative, i.e.\ inhibitory for $x
-\lesssim \xth$. That is intuitively consistent with the problem objective - for
-$t<\T$, we want to bring $X_t$ closer to the threshold, but not too close, to
-avoid early spiking!
-
-Now let us consider the differences between the individual regimes, i.e.\ the
+\lesssim \xth$. That is intuitively consistent with the problem objective. 
+For $t<\T$, we want to bring $X_t$ closer to the threshold, but not too close, to
+avoid early spiking.
+
+Now consider the differences between the individual regimes, i.e.\ the
 effect of changing the bias and the noise intensity, $\m,\b$. All things being
@@
 function, i.e.\ to make our objective worse, at the beginning of the interval
-and to decrease it at the end. This is to be expected since we are attempting
-to minimize the variance in the spike time and without noise there would be no
-variance at all. However, at the end of the interval, noise only helps with the
-spiking, since now we are only interested in spiking as soon as possible, and a
-higher noise intensity, will tend to decrease the spike time on average.
-Increasing $\b$ also has the effect of increasing the size of the boundary layer
-near $\xth$, where the value function rises steeply. I.e.\ for small $\b$ that
-layer is small since the risk of spiking early is only significant close to
-the threshold. For larger $\b$ this layer increases, see panels A,C, with a thin
-layer, vs.\ B,D, with a thicker layer, in
-\cref{fig:HJB_4regimes_value_control_cuts}. Similarly, increasing the bias,
-$\mu$, tends to decrease the value function, especially at the end since that
-has the unequivocal effect of preventing late spikes.
-\clearpage
+and to decrease it at the end. This is to be expected since we are attempting to
+minimize the variance in the spike time and without noise there would be no
+variance at all. However, at the end of the interval, noise only helps, since
+now we are only interested in spiking as soon as possible, and a higher noise
+intensity will tend to decrease the spike time on average. Increasing $\b$ also
+has the effect of increasing the size of the boundary layer near $\xth$, where
+the value function rises steeply. I.e.\ for small $\b$ that layer is small since
+the risk of spiking early is only significant close to the threshold. For larger
+$\b$ this layer increases, see panels A,C, with a thin layer, vs.\ B,D, with a
+thicker layer, in \cref{fig:HJB_4regimes_value_control_cuts}. Similarly,
+increasing the bias, $\mu$, tends to decrease the value function, especially at
+the end since that has the unequivocal effect of preventing late spikes.
+% \clearpage
 
@@
   = .001$ and the bounds are $\a \in [-2,2]$.
-  A) Supra-Threshold-High-Noise B) Supra-Threshold-Low-Noise 
-C) Sub-Threshold-High-Noise   D) Sub-Threshold-Low-Noise  }
+  A) Supra-Threshold-Low-Noise B) Supra-Threshold-High-Noise
+C) Sub-Threshold-Low-Noise   D) Sub-Threshold-High-Noise  }
   \label{fig:FBK_Regimes_cs}
@@
 
-\clearpage
-\section{Effect of energy penalty}
-\label{sec:effect_of_eps}
-So far we have assumed a low value for the energy penalty parameter,
-$\eps$, in the objective defined in \cref{eq:OC_LS_variance_energy}. Our
-approach has been that we are most concerned with the accurate spiking and that
-penalizing energy is rather intended to regularize the control and avoid
-excessive chattering of the control between its extreme values, than as a goal
-in minimizing energy in its own right. Now, we take some time to explore the
-effect of a higher $\eps$ on the optimal controls. Intuitively, a
-higher value of $\eps$ will tend to bring the optimal control, $\a^*$, closer
-to zero. We make a comparison for the four regimes for both the
-closed-loop control and the open-loop control in
-\cref{fig:HJB_4regimes_control_different_eps,fig:FBK_Regimes_cs_different_es}.
-Indeed, for both the closed-loop and open-loop optimal controls, increasing
-$\eps$ tends to reduce the absolute value of $\a^*(t)$. In particular, for the
-closed-loop control it tends to broaden the area of transition for
-decreasing $x$ when the control swings from its minimal, i.e.\ inhibitory value,
-to its maximal, i.e.\ excitatory value. Similarly for the open-loop control,
-instead of banging from its minimal bound at the beginning of the interval to
-its maximal bound at the end, the optimal control for $\e = .1$ tends to have a
-more mild transition from a slightly inhibitory to slightly excitatory
-values.
-
-Note however that increasing $\eps$ has an important effect on the validity of
-the approximation, used to form the terminal conditions for
-the value function and the adjoint variable, $\v,p$. We assumed that applying
-$\amax$ is optimal for $t>\T$. This is indeed true in the
-limit $\eps \ra 0$. However, even for a finite value of $\eps$, this may still
-be the optimal thing to do. Indeed, in the original calculations,
-where $\eps=.001$, see the panels on the left in
-\cref{fig:HJB_4regimes_control_different_eps,fig:FBK_Regimes_cs_different_es},
-the so-obtained value function {\sl implied} that $\a(x,\T) = \amax.$. I.e.\ our
-guess that $\a(x, t) = \amax$ for $t > \t$, is self-consistent with the
-so-obtained value function. In a sense, this is like confirming an ansatz. For a higher
-value of $\eps$, like $\eps = .1$, this is no longer the case. The red
-curves on the right-side of \cref{fig:HJB_4regimes_control_different_eps} are no
-longer pushed up at $\a = \amax$, and as such it is no longer valid to
-assume that the value function can be obtained by assuming $\a(t>\T) = \amax$
-and then just calculating the expected remaining time-to-spike.
-The exact same conclusion can be drawn from the open-loop control. There the
-optimal control always (almost) reaches its maximum, $\amax$ while choosing
-$\eps = .001$, but no longer so for the higher $\eps = .1$.
-In fact, as we mentioned after \cref{eq:valuefun_TCs}, the correct thing to do
-for higher values of $\eps$, is to push the
-calculation interval to some $t^+ > \T$, apply the same terminal condition at
-$t^+$ instead of $\T$ and then solve backwards (either for the closed- or
-open-loop control). One will be guaranteed to find some $t_f > \T$ that works
-since eventually the quadratic term $(t-\T)^2$ will dominate the energy term in
-the objective, which is linear in $t$. We do not explore this further here. 
-% CLOSED_LOOP EPS: 
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/HJB/Regimes_eps_comparison.pdf}
-  \caption[labelInTOC]{Effect of $\e$ on the closed-loop control. Snapshots of
-  $\a(x,t)$ for $t$ fixed, at the start (blue), mid-point (green) and end-point
-  (red), for each of the four regimes and different values of $\eps$.
-  On the left, we use $\eps = .001$ and on the right $\eps = .1$. 
-  The desired spike time is set to
-  $\T=1.5$ and the bounds are $\a \in [-2,2]$.
-  A,B) Supra-Threshold-High-Noise C,D) Supra-Threshold-Low-Noise 
-  E,F) Sub-Threshold-High-Noise G,H) Sub-Threshold-Low-Noise}
-\label{fig:HJB_4regimes_control_different_eps} 
-\end{center}
-\end{figure}
-% OPEN_LOOP EPS:
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/FP_Adjoint/Regimes_eps_comparison.pdf}
-  \caption[labelInTOC]{Effect of $\e$ on the open-loop control. 
-  The deterministic time-trajectory of $\a(t)$ for each
-  parameter regime as function of time, $t \in [0, \T]$.
-  On the left, we use $\eps = .001$ and on the right $\eps = .1$.
-  The desired spike time is set to $\T=1.5$ and the bounds are $\a \in [-2,2]$.
-  A,B) Supra-Threshold-High-Noise C,D) Supra-Threshold-Low-Noise 
-  E,F) Sub-Threshold-High-Noise G,H) Sub-Threshold-Low-Noise}
-  \label{fig:FBK_Regimes_cs_different_es}
-\end{center}
-\end{figure}
-
-\clearpage
-
-\section{Test Comparison of Different Controls}
+\section{Simulations} 
+\subsection{Single-Spike Control}
 \label{sec:probabilistic_numerical_test}
@@
 \begin{center}
+\subfloat[Supra-Threshold Low-Noise]{
+  \includegraphics[width=.99\textwidth]
+  {Figs/ControlSimulator/Regimes_SUPT_ln_errors_hist.pdf}
+}
+\\
 \subfloat[Supra-Threshold High-Noise]{
@@
   {Figs/ControlSimulator/Regimes_SUPT_HN_errors_hist.pdf}
-  }\\
-\subfloat[Supra-Threshold Low-Noise]{
+  }
+  \\
+\subfloat[Sub-Threshold Low-Noise]{
   \includegraphics[width=.99\textwidth]
-  {Figs/ControlSimulator/Regimes_SUPT_ln_errors_hist.pdf}
-}\\
+  {Figs/ControlSimulator/Regimes_subt_ln_errors_hist.pdf}
+}
+\\
 \subfloat[Sub-Threshold High-Noise]{
@@
 }\\
-\subfloat[Sub-Threshold Low-Noise]{
-  \includegraphics[width=.99\textwidth]
-  {Figs/ControlSimulator/Regimes_subt_ln_errors_hist.pdf}
-}
   \caption[labelInTOC]{Histogram of the spike timing error for the
@@
 \includegraphics[width=0.33\textwidth]
-{Figs/ControlSimulator/SubTHighNoise_Traj5.pdf}
+{Figs/ControlSimulator/SubTHighNoise_Traj0.pdf}
 }
@@
 \includegraphics[width=0.33\textwidth]
-{Figs/ControlSimulator/SubTHighNoise_Traj3.pdf}
+{Figs/ControlSimulator/SubTHighNoise_Traj4.pdf}
 }
@@
 }
-\caption[]{The effect of higher energy penalty, $\eps = .1$!
+\caption[]{The effect of higher energy penalty, $\eps = 0.1$.
 Examples for the controlled trajectories using the deterministic,
@@
 plots, we show the voltage evolution, $X_t$, on the lower plots we show the
-applied control, $\a(t)$. Note that $\a(t)$ obtained from the
-deterministic and open-loop controls are the same for all three samples.}
+applied control, $\a(t)$. Note that the optimal control, $\a(t)$, obtained from
+the deterministic and from the open-loop controls are the same across all three
+samples.}
 \label{fig:control_trajectories_examples_high_epsilon}
@@
 
-\clearpage
-
-\section{Control Robustness}
-We have so far assumed that we know the parameters determining the dynamics in
-\cref{eq:X_evolution_uo} exactly. That is we assume that the values for $\tc,
-\b$ used to obtain the controls are the same as the ones that drive the
-controlled dynamics. That is slightly naive, as in any realistic application
-these will have to be estimated and the estimates will have some error
-associated with them. 
-Note that misestimating $\mu$ has the most immediate and clear effect on the
-control policy as that amounts to a direct shift of the control and so for now,
-let us assume that $\mu$ is estimated correctly.
-
-Let us then discuss one of the regimes, the Sub-Threshold,
-High-Noise regime, and see what is the impact of misspecifying the parameters.
-
-We need to assume that the errors large but not too large\ldots
-
-So say, $$\b_{est} = \b / 1.5$$
-and 
-$$\tc_{est} = \tc / 1.5$$
-
-So we will have:
-$$\mu, \tc_{est}, \b_{est} = \{.2, .5, 1.5 \}$$ (which is the control
-calculated already for the Sub-Threshold,
-High-Noise regime), while the actual simulated dynamics will be driven by
-$$\mu, \tc, \b = \{.2, .75, 2.25 \}$$
-
-We show the effect in
-\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch_misspecified}.
-\begin{table}[h]
-\begin{center}
- \subfloat[Sub-Threshold High-Noise]{
- \input{Figs/ControlSimulator/Misspec_MeanSquaredErrors__subt_HN.txt}}\\
- \caption{The effect of misspecifying the parameters. On the left, we
- system parameters and the parameters used to obtain the control are the
- same, i.e.\ accurate, on the right, they are misspecified.}
-\label{tab:realized_avg_errors_det_vs_openloop_vs_stoch_misspecified}
-\end{center}   
-\end{table}
-
-\section{Trains Simulation }
+\subsection{Multi-Spike Control}
 We now turn to the ultimate goal of our analysis, the control of {\sl spike
@@
   \caption[ ]{AHMADIAN COMPARES: How good the {\sl closed-loop} controller is at
-  hitting a target spike train. Uses the parameters from the CRITICAL, High-Noise regime. The
-  target spike train has ISIs that are exponentially distributed with a
+  hitting a target spike train. Uses the parameters from the CRITICAL, Low-Noise
+  regime. The target spike train has ISIs that are exponentially distributed with a
   mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
@@
   \caption[ ]{AHMADIAN COMPARES: How good the {\sl open-loop} controller is at
-  hitting a target spike train. Uses the parameters from the CRITICAL, High-Noise regime. The
-  target spike train has ISIs that are exponentially distributed with a
+  hitting a target spike train. Uses the parameters from the CRITICAL, Low-Noise
+  regime. The target spike train has ISIs that are exponentially distributed with a
   mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
@@
   \label{fig:targettrain_ol_highnoise}
+\end{center}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%% SUPER-THRESHOLD REGIME LOW-NOISe (Ahmadian):
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[htp]
+\begin{center}
+  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPERTLN_Ahmadian_cl_trains_sim_10.pdf}
+  \caption[ ]{AHMADIAN COMPARES (SUPER-T): How good the {\sl closed-loop}
+  controller is at hitting a target spike train. Uses the parameters from the
+  SUPER-THRESH, High-Noise regime. The target spike train has ISIs that are exponentially distributed with a
+  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
+  control strategy}
+  \label{fig:targettrain_cl_critlownoise}
+\end{center}
+\end{figure}
+\begin{figure}[htp]
+\begin{center}
+  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPERTLN_Ahmadian_ol_trains_sim_10.pdf}
+  \caption[ ]{AHMADIAN COMPARES (SUPER-T): How good the {\sl open-loop}
+  controller is at hitting a target spike train. Uses the parameters from the
+  SUPER-THRESH, High-Noise regime. The target spike train has ISIs that are
+  exponentially distributed with a mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
+  control strategy}
+  \label{fig:targettrain_ol_critlownoise}
 \end{center}
@@
 % \end{figure}
-\clearpage
-\section{Conclusions / Discussion} 
+
+\section{Miscleaneous Analysis???}
+TODO: WE DON''T KNOW WHAT TO CALL THIS SECTION!
+\subsection{Effect of energy penalty}
+\label{sec:effect_of_eps}
+So far we have assumed a low value for the energy penalty parameter,
+$\eps$, in the objective defined in \cref{eq:OC_LS_variance_energy}. Our
+approach has been that we are most concerned with the accurate spiking and that
+penalizing energy is rather intended to regularize the control and avoid
+excessive chattering of the control between its extreme values, than as a goal
+in minimizing energy in its own right. Now, we take some time to explore the
+effect of a higher $\eps$ on the optimal controls. Intuitively, a
+higher value of $\eps$ will tend to bring the optimal control, $\a^*$, closer
+to zero. We make a comparison for the four regimes for both the
+closed-loop control and the open-loop control in
+\cref{fig:HJB_4regimes_control_different_eps,fig:FBK_Regimes_cs_different_es}.
+Indeed, for both the closed-loop and open-loop optimal controls, increasing
+$\eps$ tends to reduce the absolute value of $\a^*(t)$. In particular, for the
+closed-loop control it tends to broaden the area of transition for
+decreasing $x$ when the control swings from its minimal, i.e.\ inhibitory value,
+to its maximal, i.e.\ excitatory value. Similarly for the open-loop control,
+instead of banging from its minimal bound at the beginning of the interval to
+its maximal bound at the end, the optimal control for $\e = .1$ tends to have a
+more mild transition from a slightly inhibitory to slightly excitatory
+values.
+
+Note however that increasing $\eps$ has an important effect on the validity of
+the approximation, used to form the terminal conditions for
+the value function and the adjoint variable, $\v,p$. We assumed that applying
+$\amax$ is optimal for $t>\T$. This is indeed true in the
+limit $\eps \ra 0$. However, even for a finite value of $\eps$, this may still
+be the optimal thing to do. Indeed, in the original calculations,
+where $\eps=.001$, see the panels on the left in
+\cref{fig:HJB_4regimes_control_different_eps,fig:FBK_Regimes_cs_different_es},
+the so-obtained value function {\sl implied} that $\a(x,\T) = \amax.$. I.e.\ our
+guess that $\a(x, t) = \amax$ for $t > \t$, is self-consistent with the
+so-obtained value function. In a sense, this is like confirming an ansatz. For a higher
+value of $\eps$, like $\eps = .1$, this is no longer the case. The red
+curves on the right-side of \cref{fig:HJB_4regimes_control_different_eps} are no
+longer pushed up at $\a = \amax$, and as such it is no longer valid to
+assume that the value function can be obtained by assuming $\a(t>\T) = \amax$
+and then just calculating the expected remaining time-to-spike.
+The exact same conclusion can be drawn from the open-loop control. There the
+optimal control always (almost) reaches its maximum, $\amax$ while choosing
+$\eps = .001$, but no longer so for the higher $\eps = .1$.
+In fact, as we mentioned after \cref{eq:valuefun_TCs}, the correct thing to do
+for higher values of $\eps$, is to push the
+calculation interval to some $t^+ > \T$, apply the same terminal condition at
+$t^+$ instead of $\T$ and then solve backwards (either for the closed- or
+open-loop control). One will be guaranteed to find some $t_f > \T$ that works
+since eventually the quadratic term $(t-\T)^2$ will dominate the energy term in
+the objective, which is linear in $t$. We do not explore this further here.
+% CLOSED_LOOP EPS:
+\begin{figure}[htp]
+\begin{center}
+  \includegraphics[width=.99\textwidth]{Figs/HJB/Regimes_eps_comparison.pdf}
+  \caption[labelInTOC]{Effect of $\e$ on the closed-loop control. Snapshots of
+  $\a(x,t)$ for $t$ fixed, at the start (blue), mid-point (green) and end-point
+  (red), for each of the four regimes and different values of $\eps$.
+  On the left, we use $\eps = .001$ and on the right $\eps = .1$.
+  The desired spike time is set to
+  $\T=1.5$ and the bounds are $\a \in [-2,2]$.
+  A,B) Supra-Threshold-High-Noise C,D) Supra-Threshold-Low-Noise
+  E,F) Sub-Threshold-High-Noise G,H) Sub-Threshold-Low-Noise}
+\label{fig:HJB_4regimes_control_different_eps}
+\end{center}
+\end{figure}
+% OPEN_LOOP EPS:
+\begin{figure}[htp]
+\begin{center}
+  \includegraphics[width=.99\textwidth]{Figs/FP_Adjoint/Regimes_eps_comparison.pdf}
+  \caption[labelInTOC]{Effect of $\e$ on the open-loop control.
+  The deterministic time-trajectory of $\a(t)$ for each
+  parameter regime as function of time, $t \in [0, \T]$.
+  On the left, we use $\eps = .001$ and on the right $\eps = .1$.
+  The desired spike time is set to $\T=1.5$ and the bounds are $\a \in [-2,2]$.
+  A,B) Supra-Threshold-High-Noise C,D) Supra-Threshold-Low-Noise
+  E,F) Sub-Threshold-High-Noise G,H) Sub-Threshold-Low-Noise}
+  \label{fig:FBK_Regimes_cs_different_es}
+\end{center}
+\end{figure}
+
+% \clearpage
+
+\subsection{Control Robustness}
+We have so far assumed that we know the parameters determining the dynamics in
+\cref{eq:X_evolution_uo} exactly. That is we assume that the values for $\tc,
+\b$ used to obtain the controls are the same as the ones that drive the
+controlled dynamics. That is slightly naive, as in any realistic application
+these will have to be estimated and the estimates will have some error
+associated with them.
+Note that misestimating $\mu$ has the most immediate and clear effect on the
+control policy as that amounts to a direct shift of the control and so for now,
+let us assume that $\mu$ is estimated correctly.
+
+Let us then discuss one of the regimes, the Sub-Threshold,
+High-Noise regime, and see what is the impact of misspecifying the parameters.
+
+We need to assume that the errors large but not too large\ldots
+
+So say, $$\b_{est} = \b / 1.5$$
+and
+$$\tc_{est} = \tc / 1.5$$
+
+So we will have:
+$$\mu, \tc_{est}, \b_{est} = \{.2, .5, 1.5 \}$$ (which is the control
+calculated already for the Sub-Threshold,
+High-Noise regime), while the actual simulated dynamics will be driven by
+$$\mu, \tc, \b = \{.2, .75, 2.25 \}$$
+
+We show the effect in
+\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch_misspecified}.
+\begin{table}[h]
+\begin{center}
+ \subfloat[Sub-Threshold High-Noise]{
+ \input{Figs/ControlSimulator/Misspec_MeanSquaredErrors__subt_HN.txt}}\\
+ \caption{The effect of misspecifying the parameters. On the left, we
+ system parameters and the parameters used to obtain the control are the
+ same, i.e.\ accurate, on the right, they are misspecified.}
+\label{tab:realized_avg_errors_det_vs_openloop_vs_stoch_misspecified}
+\end{center}
+\end{table}
+
+\section{Discussion}
 We have analyzed and computed the optimal control of spikes in two scenarios -
@@
 
+\section{Acknowledgements}
+
 \appendix
@@
 times and boundary value problems is discussed in most
-introductory books on stochastic processes, see \cite{Oksendal2007,Evansa} etc.},
-\cite{Jacobs}, in deriving and solving a differential equation for
+introductory books on stochastic processes, see \cite{Oksendal2007} for
+example.}, \cite{Jacobs}, in deriving and solving a differential equation for
 $\Ti(x_0)$.
@@
 \bibliographystyle{plain}
-% \bibliographystyle{unsrt}
-% \bibliographystyle{bmc_article}
-% \section*{References}  \bibliographystyle{iopartnum}   
+\bibliographystyle{unsrt}
+\bibliographystyle{bmc_article}
+\section*{References}
+\bibliographystyle{iopartnum}
 \bibliography{library,local}

### 2013-09-02 draft.tex
Camera-ready version for S+A approval + corrections

@@
 \maketitle
+
+\tableofcontents
 
@@
   \caption[labelInTOC]{Example Trajectories from \cref{eq:X_evolution_uo}
-  using the parameter values from \cref{tab:regimes}. A)
-  Supra-Threshold-High-Noise B) Supra-Threshold-Low-Noise C) Sub-Threshold-High-Noise D) Sub-Threshold-Low-Noise}
+  using the parameter values from \cref{tab:regimes}. 
+  A) Supra-Threshold-Low-Noise 
+  B)Supra-Threshold-High-Noise 
+  C) Sub-Threshold-Low-Noise
+  D) Sub-Threshold-High-Noise 
+  }
   \label{fig:regime_path_examples}
@@
 the end since that has the unequivocal effect of preventing late spikes.
-% \clearpage
 
 \section{Open-loop Stochastic Control}
-When the value of $X_t$ is unobservable, the best
-one can do is to use the transition density to perform the optimization. Since the
-transition density follows a (deterministic) PDE, we apply a Maximum
-Principle for PDEs as a method of obtaining the optimal control,\cite{Borzi2012}.
+When the value of $X_t$ is unobservable, the transition density can be used to
+perform the optimization. Since the transition density follows a deterministic
+PDE, we apply a Maximum Principle for PDEs as a method of obtaining the optimal
+control, see \cite{Borzi2012} for details on optimization with PDEs.
 
@@
 Note that we write $\a(t)$ here instead of $\a(x,t)$ since we cannot use the
-value of $X_t$ in choosing the optimal control.
+value of $X_t$.
 
@@
 \end{align}
-Let us explain in more detail each term on the RHS in
+Let us explain in more detail each term on the right-hand side in
 \cref{eq:OC_LS_variance_density}.
@@
   The desired spike time is set to $\T=1.5$, the energy penalty, $\eps
-  = .001$ and the bounds are $\a \in [-2,2]$.
+  = 0.001$ and the bounds are $\a \in [-2,2]$.
   A) Supra-Threshold-Low-Noise B) Supra-Threshold-High-Noise
@@
 the noise and the energy penalty. That is we find a {\sl constant} value of $\a$
-which satisfies the desired boundary conditions, $x(0) = .0, x(\T) = \xth$. This naive
+which satisfies the desired boundary conditions, $x(0) = 0, x(\T) = \xth$. This naive
 controller will be called 'Deterministic', since it assumes deterministic
@@
 
-We set $\e = .001$ and $\amax = 2.0$. As such the energy cost is of secondary
+We set $\e = 0.001$ and $\amax = 2.0$. As such the energy cost is of secondary
 importance and the paramount effect on the objective is the difference $(\ts -
@@
 For illustration sake, we also visualize some trajectories from the
-Sub-Threshold-High-Noise regime in \cref{fig:control_trajectories_examples}.
+Sub-Threshold-High-Noise regime in \cref{fig:control_trajectories_examples},
+see panels A),C),E).
 
@@
 \begin{center}  
+\subfloat[Supra-Threshold Low-Noise]{
+   \input{Figs/ControlSimulator/MeanSquaredErrors__SUPT_ln.txt}}\\ 
 \subfloat[Supra-Threshold High-Noise]{   
 \input{Figs/ControlSimulator/MeanSquaredErrors__SUPT_HN.txt}}\\
-\subfloat[Supra-Threshold Low-Noise]{
-   \input{Figs/ControlSimulator/MeanSquaredErrors__SUPT_ln.txt}}\\
+ \subfloat[Sub-Threshold Low-Noise]{
+ \input{Figs/ControlSimulator/MeanSquaredErrors__subt_ln.txt}}\\ 
  \subfloat[Sub-Threshold High-Noise]{
  \input{Figs/ControlSimulator/MeanSquaredErrors__subt_HN.txt}}\\
- \subfloat[Sub-Threshold Low-Noise]{
- \input{Figs/ControlSimulator/MeanSquaredErrors__subt_ln.txt}}\\
 %  \begin{tabular}{lcc}
@@
 % \hline % \end{tabular} - 
-\caption{TODO(ALEX): THIS TABLE IS MOSTLY A SANITY CHECK TO SEE THAT THE CODE
-WORKS - CONSIDER OMITTTING IN DRAFT.
+\caption{
 Realized and theoretical performance of the different control laws. The
@@
 
-\begin{figure}[h]
-\begin{center}
-\subfloat[]
-{
-\label{fig:controlled_traj_ex1}
-\includegraphics[width=0.33\textwidth]
-{Figs/ControlSimulator/SubTHighNoise_Traj0.pdf}
-}
-\subfloat[ ]
-{
-\label{fig:controlled_traj_ex2}
-\includegraphics[width=0.33\textwidth]
-{Figs/ControlSimulator/SubTHighNoise_Traj4.pdf}
-}
-\subfloat[ ]
-{
-\label{fig:controlled_traj_ex3}
-\includegraphics[width=0.33\textwidth]
-{Figs/ControlSimulator/SubTHighNoise_Traj7.pdf}
-}
-\caption[]{Examples for the controlled trajectories using the
-deterministic, open-loop stochastic and closed-loop stochastic control
-approaches. The black vertical line in the plots indicates the desired spike-time, $\T$.
-For the parameter values are $\m, \tc, \b = [.2, .5,  1.5]$ (Sub-Threshold
-High-Noise regime). $\T = 1.5, \eps = .001$. The three panels from left to
-right are three different realizations of the model dynamics. On the upper
-plots, we show the voltage evolution, $X_t$, on the lower plots we show the
-applied control, $\a(t)$. Note that $\a(t)$ obtained from the
-deterministic and open-loop controls are the same for all three samples.}
-\label{fig:control_trajectories_examples}
-\end{center}
-\end{figure}
-
-\begin{figure}[h]
-\begin{center}
-\subfloat[]
-{
-\label{fig:controlled_traj_ex1}
-\includegraphics[width=0.33\textwidth]
-{Figs/ControlSimulator/HighEps_Traj5.pdf}
-}
-\subfloat[ ]
-{
-\label{fig:controlled_traj_ex2}
-\includegraphics[width=0.33\textwidth]
-{Figs/ControlSimulator/HighEps_Traj6.pdf}
-}
-\subfloat[ ]
-{
-\label{fig:controlled_traj_ex3}
-\includegraphics[width=0.33\textwidth]
-{Figs/ControlSimulator/HighEps_Traj0.pdf}
-}
-\caption[]{The effect of higher energy penalty, $\eps = 0.1$.
-Examples for the controlled trajectories using the deterministic,
-open-loop stochastic and closed-loop stochastic control approaches. The black vertical line in the plots indicates the desired
-spike-time, $\T$. For the parameter values are $\m, \tc, \b = [.2, .5,  1.5]$ (Sub-Threshold
-High-Noise regime). $\T = 1.5, \eps = .1$. The three panels from left to
-right are three different realizations of the model dynamics. On the upper
-plots, we show the voltage evolution, $X_t$, on the lower plots we show the
-applied control, $\a(t)$. Note that the optimal control, $\a(t)$, obtained from
-the deterministic and from the open-loop controls are the same across all three
-samples.}
-\label{fig:control_trajectories_examples_high_epsilon}
-\end{center}
-\end{figure}
+% \begin{figure}[h]
+% \begin{center}
+% \subfloat[]
+% {
+% \label{fig:controlled_traj_ex1}
+% \includegraphics[width=0.33\textwidth]
+% {Figs/ControlSimulator/SubTHighNoise_Traj0.pdf}
+% }
+% \subfloat[ ]
+% {
+% \label{fig:controlled_traj_ex2}
+% \includegraphics[width=0.33\textwidth]
+% {Figs/ControlSimulator/SubTHighNoise_Traj4.pdf}
+% }
+% \subfloat[ ]
+% {
+% \label{fig:controlled_traj_ex3}
+% \includegraphics[width=0.33\textwidth]
+% {Figs/ControlSimulator/SubTHighNoise_Traj7.pdf}
+% }
+% \caption[]{Examples for the controlled trajectories using the
+% deterministic, open-loop stochastic and closed-loop stochastic control
+% approaches. The black vertical line in the plots indicates the desired spike-time, $\T$.
+% For the parameter values are $\m, \tc, \b = [.2, .5,  1.5]$ (Sub-Threshold
+% High-Noise regime). $\T = 1.5, \eps = 0.001$. The three panels from left to
+% right are three different realizations of the model dynamics. On the upper
+% plots, we show the voltage evolution, $X_t$, on the lower plots we show the
+% applied control, $\a(t)$. Note that $\a(t)$ obtained from the
+% deterministic and open-loop controls are the same for all three samples.}
+% \label{fig:control_trajectories_examples}
+% \end{center}
+% \end{figure}
 
@@
 trains}. In particular, we will generate $M=10$ realizations of $N=20$ spikes
-each in attempt to meet a prescribed spike train. We will focus on two
+each in an attempt to meet a prescribed spike train. We will focus on two
 regimes, the sub-threshold regime with either low or high noise.
@@
 for parameters from the Sub-Threshold, High-Noise regime are shown in
-results in \cref{fig:targettrain_cl_highnoise,fig:targettrain_ol_highnoise}.
+results in
+\cref{fig:targettrain_cl_SUPT_HN,fig:targettrain_cl_SUPT_ln,fig:targettrain_ol_SUPT_ln,fig:targettrain_ol_SUPT_HN}.
 %\usepackage{graphics} is needed for \includegraphics
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%% CRIT REGIME LOW-NOISe:
+%%%%%%%%%% SUPER-THRESHOLD REGIME LOW-NOISe:
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/CRITLN_cl_trains_sim_10.pdf}
+  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_ln_cl_trains_sim_10.pdf}
   \caption[ ]{How good the {\sl closed-loop} controller is at hitting a target
-  spike train. Uses the parameters from the CRITICAL,Low-Noise regime. The
-  target spike train has ISIs that are exponentially distributed with a
-  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-  control strategy}
-  \label{fig:targettrain_cl_critlownoise}
+  spike train. Uses the parameters from the Super-Threshold,Low-Noise regime.
+  The target spike train was generated using the model itself, with
+  parameters from the Super-Threshold, High-Noise regime, without an applied
+  control, $\a=0$.
+  There are $N=16$ spikes and $M=10$ realizations of the control strategy} 
+  \label{fig:targettrain_cl_SUPT_ln}
 \end{center}
@@
 \begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/CRITLN_ol_trains_sim_10.pdf}
+  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_ln_ol_trains_sim_10.pdf}
   \caption[ ]{How good the {\sl open-loop} controller is at hitting a target
-  spike train. Uses the parameters from the CRITICAL,Low-Noise regime. The
-  target spike train has ISIs that are exponentially distributed with a
-  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-  control strategy}
-  \label{fig:targettrain_ol_critlownoise}
+  spike train. Uses the parameters from the Super-Threshold,Low-Noise regime.
+  The target spike train was generated using the model itself, with
+  parameters from the Super-Threshold, High-Noise regime, without an applied
+  control, $\a=0$.
+  There are $N=16$ spikes and $M=10$ realizations of the control strategy}
+  \label{fig:targettrain_ol_SUPT_ln}
 \end{center} 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%% CRIT REGIME HIGH-NOISe:
+%%%%%%%%%% SUPER-THRESHOLD REGIME HIGH-NOISe:
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/CRITHN_cl_trains_sim_10.pdf}
+  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_HN_cl_trains_sim_10.pdf}
   \caption[ ]{How good the {\sl closed-loop} controller is at hitting a target
-  spike train. Uses the parameters from the CRITICAL, High-Noise regime. The
-  target spike train has ISIs that are exponentially distributed with a
-  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-  control strategy}
-  \label{fig:targettrain_cl_critlownoise}
+  spike train. Uses the parameters from the Super-Threshold,High-Noise regime.
+  The target spike train was generated using the model itself, with
+  parameters from the Super-Threshold, High-Noise regime, without an applied
+  control, $\a=0$.
+  There are $N=16$ spikes and $M=10$ realizations of the control strategy}
+  \label{fig:targettrain_cl_SUPT_HN}
 \end{center} 
@@
 \begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/CRITHN_ol_trains_sim_10.pdf}
+  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_HN_ol_trains_sim_10.pdf}
   \caption[ ]{How good the {\sl open-loop} controller is at hitting a target
-  spike train. Uses the parameters from the CRITICAL, High-Noise regime. The
-  target spike train has ISIs that are exponentially distributed with a
-  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-  control strategy}
-  \label{fig:targettrain_ol_critlownoise}
+  spike train. Uses the parameters from the Super-Threshold,High-Noise regime.
+  The target spike train was generated using the model itself, with
+  parameters from the Super-Threshold, High-Noise regime, without an applied
+  control, $\a=0$.
+  There are $N=16$ spikes and $M=10$ realizations of the control strategy}
+  \label{fig:targettrain_ol_SUPT_HN}    
 \end{center}
 \end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%% CRIT REGIME LOW-NOISe (AHmadian):
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/CRITLN_Ahmadian_cl_trains_sim_10.pdf}
-  \caption[ ]{AHMADIAN COMPARES: How good the {\sl closed-loop} controller is at
-  hitting a target spike train. Uses the parameters from the CRITICAL, Low-Noise
-  regime. The target spike train has ISIs that are exponentially distributed with a
-  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-  control strategy}
-  \label{fig:targettrain_cl_critlownoise}
-\end{center}
-\end{figure}
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/CRITLN_Ahmadian_ol_trains_sim_10.pdf}
-  \caption[ ]{AHMADIAN COMPARES: How good the {\sl open-loop} controller is at
-  hitting a target spike train. Uses the parameters from the CRITICAL, Low-Noise
-  regime. The target spike train has ISIs that are exponentially distributed with a
-  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-  control strategy}
-  \label{fig:targettrain_ol_critlownoise}
-\end{center}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%% FIGS:
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[htp]
-\begin{center}
-\subfloat[DO NOT SKIP MISSED]{
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTLN_cl_trains_sim_10.pdf}
-  }\\
- \subfloat[SKIP MISSED]{
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTLN_cl_skip_trains_sim_10.pdf}
-  }
-  \caption[ ]{How good the {\sl closed-loop} controller is at hitting a target
-  spike train. Uses the parameters from the Sub-Threshold,Low-Noise regime. The
-  target spike train has ISIs that are exponentially distributed with a
-  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-  control strategy}
-  \label{fig:targettrain_cl_lownoise}
-\end{center}
-\end{figure}
-\begin{figure}[htp]
-\begin{center}
-\subfloat[DO NOT SKIP MISSED]{
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTLN_ol_trains_sim_10.pdf}
-  }\\
- \subfloat[SKIP MISSED]{
- \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTLN_ol_skip_trains_sim_10.pdf}
- }
-  \caption[ ]{How good the {\sl open-loop} controller is at hitting a target
-  spike train. Uses the parameters from the Sub-Threshold,Low-Noise regime. The
-  target spike train has ISIs that are exponentially distributed with a
-  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-  control strategy}
-  \label{fig:targettrain_ol_lownoise}
-\end{center}
-\end{figure}
-\begin{figure}[htp]
-\begin{center}    \subfloat[DO NOT SKIP MISSED]{
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTHN_cl_trains_sim_10.pdf}  }\\
- \subfloat[SKIP MISSED]{
- \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTHN_cl_skip_trains_sim_10.pdf}
- }
-  \caption[ ]{How good the {\sl closed-loop} controller is at hitting a target
-  spike train. Uses the parameters from the Sub-Threshold,High-Noise regime. The
-  target spike train has ISIs that are exponentially distributed with a
-  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-  control strategy}
-  \label{fig:targettrain_cl_highnoise}
-\end{center}
-\end{figure}
-\begin{figure}[htp]
-\begin{center}    \subfloat[DO NOT SKIP MISSED]{
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTHN_ol_trains_sim_10.pdf}  }\\
- \subfloat[SKIP MISSED]{
- \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTHN_ol_skip_trains_sim_10.pdf}
- }
-  \caption[ ]{How good the {\sl open-loop} controller is at hitting a target
-  spike train. Uses the parameters from the Sub-Threshold,High-Noise regime. The
-  target spike train has ISIs that are exponentially distributed with a
-  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-  control strategy}
-  \label{fig:targettrain_ol_highnoise}
-\end{center}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%% SUPER-THRESHOLD REGIME LOW-NOISe (Ahmadian):
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPERTLN_Ahmadian_cl_trains_sim_10.pdf}
-  \caption[ ]{AHMADIAN COMPARES (SUPER-T): How good the {\sl closed-loop}
-  controller is at hitting a target spike train. Uses the parameters from the
-  SUPER-THRESH, High-Noise regime. The target spike train has ISIs that are exponentially distributed with a
-  mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-  control strategy}
-  \label{fig:targettrain_cl_critlownoise}
-\end{center}
-\end{figure}
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPERTLN_Ahmadian_ol_trains_sim_10.pdf}
-  \caption[ ]{AHMADIAN COMPARES (SUPER-T): How good the {\sl open-loop}
-  controller is at hitting a target spike train. Uses the parameters from the
-  SUPER-THRESH, High-Noise regime. The target spike train has ISIs that are
-  exponentially distributed with a mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-  control strategy}
-  \label{fig:targettrain_ol_critlownoise}
-\end{center}
-\end{figure}
+
 \clearpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%% SKIP MISSED FIGS:
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \begin{figure}[htp]
-% \begin{center}
-%   \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTLN_cl_skip_trains_sim_10.pdf}
-%   \caption[ ]{SKIPPING MISSED TARGET. How good the {\sl closed-loop} controller is at hitting a target
-%   spike train. Uses the parameters from the Sub-Threshold,Low-Noise regime. The
-%   target spike train has ISIs that are exponentially distributed with a
-%   mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-%   control strategy}
-%   \label{fig:targettrain_cl_lownoise}
-% \end{center}
-% \end{figure}
-% \begin{figure}[htp]
-% \begin{center}
-%   \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTLN_ol_skip_trains_sim_10.pdf}
-%   \caption[ ]{SKIPPING MISSED TARGET. How good the {\sl open-loop} controller is at hitting a target
-%   spike train. Uses the parameters from the Sub-Threshold,Low-Noise regime. The
-%   target spike train has ISIs that are exponentially distributed with a
-%   mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-%   control strategy}
-%   \label{fig:targettrain_ol_lownoise}
-% \end{center}
-% \end{figure}
-% \begin{figure}[htp]
-% \begin{center}
-%   \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTHN_cl_skip_trains_sim_10.pdf}
-%   \caption[ ]{SKIPPING MISSED TARGET. How good the {\sl closed-loop} controller is at hitting a target
-%   spike train. Uses the parameters from the Sub-Threshold,High-Noise regime. The
-%   target spike train has ISIs that are exponentially distributed with a
-%   mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-%   control strategy}
-%   \label{fig:targettrain_cl_highnoise}
-% \end{center}
-% \end{figure}
-% \begin{figure}[htp]
-% \begin{center}
-%   \includegraphics[width=.99\textwidth]{Figs/TrainController/SUBTHN_ol_skip_trains_sim_10.pdf}
-%   \caption[ ]{SKIPPING MISSED TARGET. How good the {\sl open-loop} controller is at hitting a target
-%   spike train. Uses the parameters from the Sub-Threshold,High-Noise regime. The
-%   target spike train has ISIs that are exponentially distributed with a
-%   mean of $1.5$ a.u. There are $N=20$ spikes and $M=10$ realizations of the
-%   control strategy}
-%   \label{fig:targettrain_ol_highnoise}
-% \end{center}
-% \end{figure}
 
@@
 instead of banging from its minimal bound at the beginning of the interval to
-its maximal bound at the end, the optimal control for $\e = .1$ tends to have a
+its maximal bound at the end, the optimal control for $\e = 0.1$ tends to have a
 more mild transition from a slightly inhibitory to slightly excitatory
@@
 be the optimal thing to do. Indeed, in the original calculations,
-where $\eps=.001$, see the panels on the left in
+where $\eps=0.001$, see the panels on the left in
 \cref{fig:HJB_4regimes_control_different_eps,fig:FBK_Regimes_cs_different_es},
-the so-obtained value function {\sl implied} that $\a(x,\T) = \amax.$. I.e.\ our
+the so-obtained value function {\sl implied} that $\a(x,\T) = \amax$.
+I.e.\ our
 guess that $\a(x, t) = \amax$ for $t > \t$, is self-consistent with the
-so-obtained value function. In a sense, this is like confirming an ansatz. For a higher
-value of $\eps$, like $\eps = .1$, this is no longer the case. The red
+so-obtained value function. This is akin to confirming an ansatz.
+For a higher value of $\eps$, like $\eps = 0.1$, this is no longer the case. The red
 curves on the right-side of \cref{fig:HJB_4regimes_control_different_eps} are no
@@
 optimal control always (almost) reaches its maximum, $\amax$ while choosing
-$\eps = .001$, but no longer so for the higher $\eps = .1$.
+$\eps = 0.001$, but no longer so for the higher $\eps = 0.1$.
 In fact, as we mentioned after \cref{eq:valuefun_TCs}, the correct thing to do
@@
   (red), for each of the four regimes and different values of $\eps$.
-  On the left, we use $\eps = .001$ and on the right $\eps = .1$.
+  On the left, we use $\eps = 0.001$ and on the right $\eps = 0.1$.
   The desired spike time is set to
   $\T=1.5$ and the bounds are $\a \in [-2,2]$. 
-  A,B) Supra-Threshold-High-Noise C,D) Supra-Threshold-Low-Noise
-  E,F) Sub-Threshold-High-Noise G,H) Sub-Threshold-Low-Noise}
+  A,B)
+   Supra-Threshold-Low-Noise 
+  C,D) 
+   Supra-Threshold-High-Noise 
+  E,F)
+   Sub-Threshold-Low-Noise 
+  G,H)
+   Sub-Threshold-High-Noise}
 \label{fig:HJB_4regimes_control_different_eps} 
@@
   parameter regime as function of time, $t \in [0, \T]$.
-  On the left, we use $\eps = .001$ and on the right $\eps = .1$.
+  On the left, we use $\eps = 0.001$ and on the right $\eps =0.1$.
   The desired spike time is set to $\T=1.5$ and the bounds are $\a \in [-2,2]$.
-  A,B) Supra-Threshold-High-Noise C,D) Supra-Threshold-Low-Noise
-  E,F) Sub-Threshold-High-Noise G,H) Sub-Threshold-Low-Noise}
+    A,B)
+   Supra-Threshold-Low-Noise
+  C,D) 
+   Supra-Threshold-High-Noise
+  E,F)
+   Sub-Threshold-Low-Noise
+  G,H)
+   Sub-Threshold-High-Noise}
   \label{fig:FBK_Regimes_cs_different_es} 
@@
 
-% \clearpage
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.99\textwidth]
+{Figs/ControlSimulator/Composite_Traj.pdf}
+\caption[]{Three examples for the controlled trajectories using the
+deterministic, open-loop stochastic and closed-loop stochastic control approaches.
+The plots on the left are obtained using a value of $\eps=0.001$, while the
+plots on the right use $\eps=0.1$.
+The black vertical line in the plots indicates the desired spike-time, $\T$.
+The parameter values are $\m, \tc, \b = [0.2, 0.5,  1.5]$ (Sub-Threshold
+High-Noise regime), with $\T = 1.5$. 
+On the upper plots, we show the voltage evolution, $X_t$, on the lower plots we
+show the applied control, $\a(t)$. Note that the optimal control, $\a(t)$, obtained from
+the deterministic and from the open-loop controls are the same across all three
+samples.}
+\label{fig:control_trajectories_examples}
+\end{center} 
+\end{figure}
+
+% 
+% \begin{figure}[h]
+% \begin{center}
+% \subfloat[]
+% {
+% \label{fig:controlled_traj_ex1}
+% \includegraphics[width=0.33\textwidth]
+% {Figs/ControlSimulator/HighEps_Traj5.pdf}
+% }
+% \subfloat[ ]
+% {
+% \label{fig:controlled_traj_ex2}
+% \includegraphics[width=0.33\textwidth]
+% {Figs/ControlSimulator/HighEps_Traj6.pdf}
+% }
+% \subfloat[ ]
+% {
+% \label{fig:controlled_traj_ex3}
+% \includegraphics[width=0.33\textwidth]
+% {Figs/ControlSimulator/HighEps_Traj0.pdf}
+% }
+% \caption[]{The effect of higher energy penalty, $\eps = 0.1$.
+% Examples for the controlled trajectories using the deterministic,
+% open-loop stochastic and closed-loop stochastic control approaches. The black vertical line in the plots indicates the desired
+% spike-time, $\T$.
+% The parameter values are $\m, \tc, \b = [0.2, 0.5,  1.5]$ (Sub-Threshold
+% High-Noise regime). $\T = 1.5, \eps = 0.1$. The three panels from left to
+% right are three different realizations of the model dynamics. On the upper
+% plots, we show the voltage evolution, $X_t$, on the lower plots we show the
+% applied control, $\a(t)$. Note that the optimal control, $\a(t)$, obtained from
+% the deterministic and from the open-loop controls are the same across all three
+% samples.}
+% \label{fig:control_trajectories_examples_high_epsilon}
+% \end{center}
+% \end{figure}
 
 \subsection{Control Robustness}
-We have so far assumed that we know the parameters determining the dynamics in
-\cref{eq:X_evolution_uo} exactly. That is we assume that the values for $\tc,
-\b$ used to obtain the controls are the same as the ones that drive the
+We have so far assumed that we know the parameters, $\m, \tc, \b$, determining
+the dynamics in \cref{eq:X_evolution_uo,eq:current_mu_alpha} exactly. That is we
+assume that the values for $\tc, \b$ used to obtain the controls are the same as the ones that drive the
 controlled dynamics. That is slightly naive, as in any realistic application
@@
 
-We need to assume that the errors large but not too large\ldots
-
-So say, $$\b_{est} = \b / 1.5$$
-and
-$$\tc_{est} = \tc / 1.5$$
+Assume that $\b$ and $\tc$ are underestimated, so that $\b_{est} = \b / 1.5$
+and $\tc_{est} = \tc / 1.5$.
 
 So we will have:
-$$\mu, \tc_{est}, \b_{est} = \{.2, .5, 1.5 \}$$ (which is the control
+$$\mu, \tc_{est}, \b_{est} = \{.2, .5, 1.5 \}$$ which is the control
 calculated already for the Sub-Threshold,
-High-Noise regime), while the actual simulated dynamics will be driven by
+High-Noise regime, while the actual simulated dynamics will be driven by
 $$\mu, \tc, \b = \{.2, .75, 2.25 \}$$
@@
 We show the effect in
-\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch_misspecified}.
+\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch_misspecified}. Note that
+a misspecification of $\b$ should not have a direct influence on the behaviour
+of the 'deterministic' controller since it already assumes in any case that $\b=0$.
 \begin{table}[h]
@@
  \input{Figs/ControlSimulator/Misspec_MeanSquaredErrors__subt_HN.txt}}\\
- \caption{The effect of misspecifying the parameters. On the left, we
+ \caption{The effect of misspecifying the parameters. On the left, the
  system parameters and the parameters used to obtain the control are the
@@
 
-ALEX: SUBSECTIONS ARE FOR DRAFT ONLY - THEY WILL BE FUSED
-TOGETHER IN FINAL SUBMISSION
-\subsection{different regimes}
 Naturally, the ability to control the system is most clearly affected by the
@@
 
-\subsection{computational matters}
 In both cases finding the optimal control is computationally intensive as we
@@
 
-\subsection{objective criterion}
-As our criterion to optimize, we have chosen the squared deviation between the
-realized time and some {\sl a priori} chosen time. We have done so as it
-appeared to us the simplest objective, representing the idea. Other
-objectives such as the absolute value of the difference, $|\ts - \T|$ or its
-exponential, $\exp(|\ts - \T|)$, are directly applicable. A different approach
-may be to require that $\Exp[\ts] = \T$. It is non-trivial to do this in the
-dynamic programing framework, while it is fairly straightforward while working
-with the forward equation in the open-loop context. A similarly comment may be
-made about maximizing the
-likelihood of $\ts = T$, which is the objective chosen in \cite{Ahmadian2011}.
-
-\subsection{relation to other works}
 Our work has strayed pretty close to the paradigm of \cite{Ahmadian2011} in
@@
 the control is still deterministic, but it is not-trivial. It is even less
-trivial when one uses Dynamic Programing \{ALEX: IT IS WORTH THINKING ABOUT, AS
-IT MAKES FOR A VERY COOL MATH PROBLEM, BUT WHAT IS THE PAY-OFF?\}. In
+trivial when one uses Dynamic Programing. In
 particular, insisting on charge-balance in our schemes will make it much more
@@
 
-\subsection{novelty}
-We have developed a stochastic optimal control scheme which is not limited
-by the intensity of the noise in the system. Given this scope, the
-numerical demands on our scheme are non-trivial, but our experience and
-experiments show that these demands are potentially feasible. It
+The novelty in our work is primarily in obtaining a stochastic control scheme,
+which is not limited by the intensity of the noise in the system. In fact, by
+comparison with the naive 'deterministic' controller, we have shown how wrong
+the control strategy can be if we assume that the noise is negligible. Given
+this scope, the numerical demands on our scheme are non-trivial, but our
+experience and experiments show that these demands are potentially feasible. It
 however remains to be seen whether they can be successfully applied in the lab
 and beyond.
-
-\subsection{{\sl Stochastic} effects}
-We show how wrong the control strategy can be if we assume that the noise is
-negligible.
-
-\section{Acknowledgements}
 
@@
 \bibliographystyle{bmc_article}
+\section*{Acknowledgements}
+   \ifthenelse{\boolean{publ}}{\small}{}
+SD is supported by the Danish Council for Independent Research $\mid$ 
+Natural Sciences.
+\\ 
+AL acknowledges support from NSERC Canada.  
+\\
+AI is supported by a PGS-D scholarship from NSERC Canada.
+
 \section*{References}

### 2013-09-02 draft.tex
sec 5.2 revision

@@
 each in an attempt to meet a prescribed spike train. We will focus on two
-regimes, the sub-threshold regime with either low or high noise.
+regimes, the Supra-Threshold regime with either low or high noise.
 The results
-for parameters from the Sub-Threshold, Low-Noise regime are shown in
-results in \cref{fig:targettrain_cl_lownoise,fig:targettrain_ol_lownoise}.
-While, the results
-for parameters from the Sub-Threshold, High-Noise regime are shown in
+for parameters from the low-noise regime are shown in
+results in \cref{fig:targettrain_cl_lownoise,fig:targettrain_ol_lownoise}, 
+while, the results
+for the high-noise regime are shown in
 results in
-\cref{fig:targettrain_cl_SUPT_HN,fig:targettrain_cl_SUPT_ln,fig:targettrain_ol_SUPT_ln,fig:targettrain_ol_SUPT_HN}.
-%\usepackage{graphics} is needed for \includegraphics
+\cref{fig:targettrain_cl_highnoise,fig:targettrain_ol_highnoise}. In all cases,
+the target train is obtained by a simulation of the Supra-Threshold High Noise
+regime with no additional control, i.e.\ $\a(\cdot) = 0$.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
   There are $N=16$ spikes and $M=10$ realizations of the control strategy} 
-  \label{fig:targettrain_cl_SUPT_ln}
+  \label{fig:targettrain_cl_lownoise}
 \end{center}
@@
   There are $N=16$ spikes and $M=10$ realizations of the control strategy}
-  \label{fig:targettrain_ol_SUPT_ln}
+  \label{fig:targettrain_ol_lownoise}
 \end{center} 
@@
   There are $N=16$ spikes and $M=10$ realizations of the control strategy}
-  \label{fig:targettrain_cl_SUPT_HN}
+  \label{fig:targettrain_cl_highnoise}
 \end{center} 
@@
   There are $N=16$ spikes and $M=10$ realizations of the control strategy}
-  \label{fig:targettrain_ol_SUPT_HN}    
+  \label{fig:targettrain_ol_highnoise}    
 \end{center}

### 2014-01-22 draft.tex
Adding Refreply.tex

@@
 \title{Stochastic Optimal Control of Single Neuron Spike Trains}
-\author{Alexandre Iolov$^1$, Susanne Ditlevsen$^2$, Andr\'e Longtin$^1$
+\author{Alexandre Iolov$^{1,2}$, Susanne Ditlevsen$^2$, Andr\'e Longtin$^{1,3}$ 
 		}	
@@
 Copenhagen, Denmark
+\\
+    (3) Department of Physics, University of Ottawa,
+    Ottawa, Canada
 }
@@
 
-{\em Objective.} External control of exact spike timings in single neurons is
+{\em Objective.} External control of exact spike times in single neurons is
 important for basic understanding of neuronal activity, and has potential in
@@
 stimulation patterns to a cell to achieve a target spike train under
-physiological constraints not to damage tissue and minimize possible side
-effects. {\em Approach.} We pose an optimal control problem to precisely specify
-the spike timing in a noisy leaky integrate-and-fire model of a neuron. In
-particular, we allow for the noise to be of arbitrary intensity. The optimal
-control problem is solved using Dynamic Programming if the controller has access
-to the voltage (closed-loop control) and using a Maximum Principle for the
-transition density if the controller only has access to the spike timings (open
-loop control). {\em Main results.} We have developed a stochastic optimal
-control scheme to obtain precise spike timings, which is not limited by the
-intensity of the noise in the system. Given this scope, the numerical demands
-are non-trivial, but simulations show that the algorithms are potentially
-feasible. It however remains to see if they can be successfully applied in a
-real-time experimental setting. Furthermore, we show how wrong the control
-strategy can perform if we assume that the noise is negligible. {\em
-Significance.} More precise targeting and control of neural activity to achieve
-optimal performance in a noisy environment, in which most neurons are embedded,
-have great potential in health care. The main contribution is the online
-feedback control of a noisy system through modulation of the input, taking into
-account physiological constraints on the control, as well as incurred costs to
-diminish adverse effects.
+physiological constraints not to damage tissue. {\em Approach.} We pose a
+stochastic optimal control problem to precisely specify the spike times in a
+noisy leaky integrate-and-fire model of a neuron. In particular, we allow for
+the noise to be of arbitrary intensity. The optimal control problem is solved
+using Dynamic Programming if the controller has access to the voltage
+(closed-loop control) and using a Maximum Principle for the transition density
+if the controller only has access to the spike times (open-loop control). {\em
+Main results.} We have developed a stochastic optimal control scheme to obtain
+precise spike times, which is not limited by the intensity of the noise in the
+system and is applicable in both the supra-threshold and sub-threshold regimes.
+Given this scope, our simulations show that our algorithms produce the desired results. They further reveal that the numerical demands are non-trivial,
+such that the algorithms may require efficiency refinements to achieve real-time
+control, in particular the open-loop context is more numerically demanding than
+the closed-loop one. Furthermore, we show how wrong the control strategy can
+perform if we assume that the noise is negligible. {\em Significance.} More
+precise targeting and control of neural activity to achieve optimal performance
+in a noisy environment, in which most neurons are embedded, have great potential
+in health care. The main contribution is the online feedback control of a noisy
+system through modulation of the input, taking into account physiological
+constraints on the control, as well as incurred costs to diminish adverse
+effects.
 
@@
 
-\tableofcontents
+% \tableofcontents
+% TODO: Remove ToCs (for internal consumption only)
 
@@
 work with open-loop control. The difference between the objective in
-\cite{Ahmadian2011} and what we consider is that they maximize the probability
-of spiking at some given time $t^*$, whereas we minimize the mean squared
+\cite{Ahmadian2011} and what we consider below is that they maximize the
+probability of spiking at some given time $t^*$, whereas we minimize the mean squared
 difference between the realized spike time, $T_{sp}$ and the desired, $t^*$.
@@
 consider noise, though, and only work with open-loop control. In
-\cite{Nabi2013}, the methods are implemented on brain slices of pyramidal
+\cite{Nabi2013}, their methods are implemented on brain slices of pyramidal
 neurons of rat hippocampus. While the
@@
 neuron's response to a stimulus, it is only valid in the supra-threshold regime,
-where the unstimulated neuron is periodically spiking. In \cite{Feng2003} they
-use a leaky integrate-and-fire model and open-loop control. To
-obtain a given spike time, they choose an objective of minimizing the
-variance of the membrane potential at the desired spike time, while forcing the mean of
-the membrane potential at this time point to equal the threshold. This provides
-exact solutions since it does not involve first-passage times, but has the
-drawback that there is a non-negligible probability that the obtained
-spike time will be far from the desired spike time.
-
-Our objective of imposing a certain timing sequence of the spike train using an
+where the unstimulated neuron is periodically spiking. Reference
+\cite{Feng2003} investigates the control of the firing times of a
+leaky integrate-and-fire neuron by varying the intensity of a noise process that
+drives the voltage (there is no other intrinsic noise in the model). 
+To obtain a given spike time, they choose the objective of minimizing the
+variance of the membrane potential at the desired spike time, while forcing the
+mean of the membrane potential at this time point to equal the threshold. This
+provides exact solutions since it does not involve first-passage times, but has
+the drawback that there is a non-negligible probability that the obtained spike
+time will be far from the desired spike time.
+
+Our objective of imposing a certain timing sequence for the spike train using an
 externally applied control is obtained in both the closed- and
@@
 Then we relax the observation assumption so that the controller only has access
-to the spike timings. Finally, we compare the two methods through
+to the spike times. Finally, we compare the two methods through
 simulations against a simple-minded control technique which ignores
@@
 \begin{gathered}
-dX_s = \left(\Iext(t) - \frac{X_s}{\tc} \right) \intd{s} + \b \intd{W_s},
+dX(t) = \left(\Iext(t) - \frac{X(t)}{\tc} \right) \intd{t} + \b \intd{W},
 \\
@@
 \begin{cases}
-X(\ts^+) = 0 &
+X(\ts^+) = 0. & 
 \end{cases}
@@
 \end{equation}
-Here $X$ represents the membrane electric potential which decays to $0$
-with a time constant of $\tc$, $dW$ is a Brownian motion increment scaled by
-$\b$ and $\Iext(t)$ is the deterministic external input to the cell. Having last
-spiked at time $0$, the potential hits $\xth$ at some time $\ts$, the potential
-resets to $0$ and the process starts all over again. 
-We write $\ts^+$ for the voltage's limit from the right at $\ts$.
+Here $X(t)$ represents the membrane electric potential at time $t$, which in
+absence of input decays to $0$ with a time constant of $\tc$, $dW$ is a Brownian
+motion increment scaled by $\b$ and $\Iext(t)$ is the deterministic external input to the cell. Having last
+spiked at time $0$, the potential hits $\xth$ at some random time $\ts$, the
+potential resets to $0$ and the process starts all over again. 
+We write $\ts^+$ for the limit from the right at $\ts$.
 Throughout the paper the threshold is set to one, $\xth = 1$.
@@
 \begin{equation}
-\Iext(t) = \m + \a(t)
+\Iext(t) = \m + \a(t),
 \label{eq:current_mu_alpha}
@@
 A natural goal is to attempt to control the spike time, $\ts$.
-That is how do we choose $\a(t)$ such that $\ts = \T$.
-A natural optimal control objective to achieve this is the least squares solution
-\begin{equation}
-\a(t) = \argmin_{\a(t)} \{\, \Exp[(\ts- \T)^2] \,\}
+That is, how do we choose $\a(t)$ such that $\ts \approx \T$, where $\T$ is the
+desired spike-time. A natural optimal control objective to achieve this is the
+least squares solution
+\begin{equation}
+\a(\cdot) = \argmin_{\a(\cdot)} \{\, \Exp[(\ts- \T)^2] \,\},
 \label{eq:OC_LS_variance}
@@
 where expectation is taken with respect to the distribution of the trajectories
-of $X_t$.
+of $X$.
 
@@
 In addition to \cref{eq:OC_LS_variance}, we will also add to our objective a
-running cost based on the control. This regularizes the problem eliminating the
-subtleties of singular-control situations and it serves to avoid excessive
-control as well as to avoid excessive charge building up on the cell.
+running energy cost based on the control. This regularizes the problem
+eliminating the subtleties of singular-control situations and it serves to avoid excessive
+control as well as to avoid excessive charge building up on the cell, see
+\cite{Ahmadian2011}.
 
@@
 \\ \notag
-\a^*(\cdot) =& \argmin_{\a(\cdot)} J[\a(\cdot)].
+\a^*(\cdot) =& \argmin_{\a(\cdot)} J[\a(\cdot)],
 \end{align}
 where $\e$ measures how much weight we put on minimizing the energy cost.
-If $\e = 0$, then we do not care at all about the expended cost.
+If $\e = 0$, then we do not care at all about the expended energy cost.
 
@@
 of $X$, it will be at least measurable with respect to the filtration of $X$,
-see \cite{Oksendal2007} for the details. We will try to make clear below when we
-are refering to $\a$ as a stochastic process and when we are merely refering to its particular
+see Ch.\ 11 in \cite{Oksendal2007} for the details. We will try to make clear
+below when we are refering to $\a$ as a stochastic process and when we are merely refering to its particular
 value at some time. For example, in \cref{eq:OC_LS_variance_energy}, $\a(\cdot)$
@@
 We will consider two control contexts -- {\sl closed-loop} and  {\sl open-loop}
-control. In closed-loop control, the value of $X_t$ at time $t$ is observable
+control. In closed-loop control, the value of $X$ at time $t$ is observable
 and can be used in determining the control. In open-loop control only the spike
 times are observable. In the closed-loop context, we write the control as $$\a =
-\a(x,t)$$ to indicate its dependence on $X_t$ and to express that it will be
+\a(x,t)$$ to indicate its dependence on $X(t)$ and to express that it will be
 updated based on the time-course of $X$. In the open-loop context, we write $$\a
@@
 see \cite{Fleming1975}, while for the open-loop scenario we use a form of the
-Maximum Principle applied on the transition density of the controlled process,
+Maximum Principle applied to the transition density of the controlled process,
 see \cite{Borzi2012}.
@@
 Crucially, we assume that the model parameters, $\m,\tc,\b$ in
-\cref{eq:X_evolution_uo} are known. This is a strong assumption and will be
+\cref{eq:X_evolution_uo,eq:current_mu_alpha} are known. This is a strong assumption and will be
 discussed later.
@@
 Spikes will occur in the absence of noise, if and only
-if $ \m > \frac 1\tc,$ which is called the supra-threshold regime. When $\m
+if $ \m > 1 / \tc,$ which is called the supra-threshold regime. When $\m
 \leq 1/\tc$, the regime is called the sub-threshold regime.
-In addition we will impose two values of $\b$, which we
+In addition we will investigate two values of $\b$, which we
 will call high-noise and low-noise, respectively.
@@
   A) Supra-Threshold-Low-Noise 
-  B)Supra-Threshold-High-Noise 
+  B) Supra-Threshold-High-Noise 
   C) Sub-Threshold-Low-Noise
-  D) Sub-Threshold-High-Noise 
-  }
+  D) Sub-Threshold-High-Noise. 
+  Note the multiple crossings very close together in the high-noise regimes
+  in B and D. }
   \label{fig:regime_path_examples}
@@
 \section{Closed-Loop Solution - Dynamic Programing}
-We now detail the Dynamic Programing approach to obtaining a feedback-optimal
-control. What this amounts to is that the controller can be continuously updated
-depending on the realization of the stochastic process, $X_t$.
-
-Given a time $t$ and a realized voltage value $X_t = x$, let $\trem$ be the
-(unknown) remaining time to spike, $\trem = \ts - t$. Starting from an arbitrary
-$t, x$, our remaining-cost objective, $J[\a(\cdot); x,t]$, will be
+We now detail the Dynamic Programing approach to obtaining an optimal feedback
+control. In closed-loop, the controller can be continuously updated depending on
+the realization of the stochastic process, $X$.
+
+Given a time $t$ and a realized voltage value $X(t) = x$, let $(\ts-t)$ be the
+unknown and random remaining time to spike, note that $\ts$ is a random
+variable. Given arbitrary $t, x$, our remaining-cost objective, $J[\a(\cdot);
+x,t]$, will be
 \begin{equation}
@@
 \Exp\left[
-\big(\trem - (\T-t) \big)^2
+\big((\ts-t) - (\T-t) \big)^2  
 +
 \e \int_t^\ts  \a^2(s) \intd{s} 
-\,\Big|\, X_t = x
-\right]
+\,\Big|\, X(t) = x
+\right],
 \label{eq:OC_LS_cost_to_go}
@@
 i.e.\ if time $t$ has elapsed without a spike, we now want to minimize
-the difference between $\t = \ts - t$ and ($\T-t$), given the current state $x$.
-Recall that in the closed-loop scenario, we assume that the value of $X_t$
-is known to the controller. A similar problem is discussed analytically at
-length in the book on optimal control by Peter Whittle, \cite{Whittle1996},
-although there is no discussion there of the numerics required to solve it.
+the difference between $(\ts - t)$ and $(\T-t)$, given the current state $x$.
+The mean in \cref{eq:OC_LS_cost_to_go} is taken over the distribution of
+hitting times, $\ts$, conditional on $X(t)$ or equivalently over the
+distribution of forward trajectories of $X$ starting at $x$ and ending at the
+threshold. 
+Recall that in the closed-loop scenario, we assume that
+the value of $X(t)$ is known to the controller. A similar problem is discussed
+analytically at length in the book on optimal control by Peter Whittle,
+\cite{Whittle1996}, although there is no discussion there of the numerics
+required to solve it.
 
 \subsection{Hamilton-Jacobi-Bellman equation}
-The Hamilton-Jacobi-Bellman (HJB) equation,
-see \cite{Fleming1975,Whittle1996} or \cite{Danzl2009,Nabi2013a} for a neuroscience
-context, associated with
-\cref{eq:OC_LS_cost_to_go} is obtained as follows. We introduce the value function, $\v(x,t)$:
+The Hamilton-Jacobi-Bellman (HJB) equation, see \cite{Fleming1975,Whittle1996}
+or the articles \cite{Danzl2009,Nabi2013a} in a neuroscience context, associated
+with the optimal control for \cref{eq:OC_LS_cost_to_go} is obtained as follows.
+We introduce the value function, $\v(x,t)$, as the minimum of the remaining-cost objective, i.e.\ of
+the cost function between the current time $t$ and the desired spike time
+$\T$:
 \begin{align}
@@
 \Exp\left[
-\big(\trem - (\T-t) \big)^2
+\big((\ts-t) - (\T-t) \big)^2
 +
 \e \int_t^\ts  \a^2(s) \intd{s}
-\,\Big|\,X_t = x
-\right]
+\,\Big|\,X(t) = x
+\right].
 \label{eq:OC_LS_mean_value}
 \end{align}
-
 Then $\v$ satisfies the following HJB partial differential equation (PDE):
@@
 \amax]}\Big\{ \e \a^2 (x,t)+ \left(\m + \a(x,t)-\tfrac{x}{\tc}\right) \di_x
-\v(x,t) \Big\} = 0
+\v(x,t) \Big\} = 0.
 \label{eq:OC_LS_mean_HJB}
@@
 solution, $\v$. For each $x,t$ in the computational domain,
-$\a$ is chosen as to minimize $\{\e \a^2 + \left(\m + \a - \tfrac{x}{\tc}\right) \di_x \v \}$.
-Here we can solve for the optimal control, $\a^*(x,t)$ analytically as:
+$\a$ is chosen such as to minimize $\{\e \a^2 + \left(\m + \a -
+\tfrac{x}{\tc}\right) \di_x \v \}$. Here we can solve for the optimal control, $\a^*(x,t)$ analytically as:
 \begin{align}
@@
 =&
-\min \left(\amax, \max\left(\amin, -\frac{\di_x\v(x,t)}{2\e}\right)\right)
+\min \left(\amax, \max\left(\amin, -\frac{\di_x\v(x,t)}{2\e}\right)\right).
 \label{eq:OC_LS_variance_energy_quadratic_control}
@@
 \e (\astar(x,t))^2 + (\m + \astar(x,t)-\tfrac{x}{\tc}) \di_x\v
-= 0
+= 0.
 \label{eq:OC_LS_mean_HJB_bounded_control}
@@
 
-We need to consider boundary conditions (BCs) for $\v$. If $X_t = \xth$ then we
+We need to consider boundary conditions (BCs) for $\v$. If $X(t) = \xth$ then we
 have a spike now and $\ts = t$. Thus $$ \v(\xth,t) = (t-\T)^2.$$ At the
@@
 
-To determine the Terminal Conditions (TCs) for $\v$, the idea is
-simple: if we reach $\T$ without having spiked we apply maximum control in the
-positive direction, i.e.\ $$t>T \implies \a(t) = \amax$$ Thus:
-\begin{equation}\v(x,\T) =
-\Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big].
+Note that Dynamic Programing and the HJB equation work backwards. Thus the
+evolution of the value function proceeds from the future to the past and we need
+some Terminal Condition (TCs) at some point in the future, possibly infinity,
+from which to start incrementing $\v$ using the dynamics and the BCs. To
+determine TCs for $\v$, our idea is simple: if we reach $\T$ without having
+spiked we apply maximum control in the positive direction, i.e.\ 
+$$t>\T \implies \a(t) = \amax.$$ Thus: \begin{equation}\v(x,\T) = \Exp
+\Big[\trem^2 \,\Big|\, X(\T) = x, \a(t) = \amax \Big].
 \label{eq:valuefun_TCs}
@@
 - we are ignoring the energy term, $\eps u^2$, in the objective for
-$t>T$. Naturally, this approximation is ever more accurate for $\eps \ll 1$.
+$t>\T$. Naturally, this approximation is ever more accurate for $\eps \ll 1$.
 We discuss in more detail the validity of the approximation in
@@
 \begin{equation}
-\Ttwo(x) := \Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big].
+\Ttwo(x) := \Exp \Big[\trem^2 \,\Big|\, X(\T) = x, \a(t) = \amax \Big].
 \label{eq:Trem_squared}
@@
 $\Ttwo$ is the second moment of the remaining time to reach the
-threshold starting at $X_\t = x$ and applying $\amax$ throughout. This quantity can be found
-easily for all $x$ in the domain by solving a stationary backward Kolmogorov
-equation. We show the details in \ref{sec:valuefun_TCs}.
+threshold starting at $X(\T) = x$ and applying $\amax$ throughout. This quantity
+can be found easily for all $x$ in the domain by solving a stationary backward Kolmogorov
+equation. This is an extension to the calculation of the first moment of an
+exit-time as seen in textbooks such as \cite{Jacobs} and we give the
+details in \ref{sec:valuefun_TCs}.
 
@@
 \e \a^2(x,t) + \big(\m + \a(x,t) -\tfrac{x}{\tc}\big)\cdot \di_x\v(x,t)
-= 0
+= 0,
 \\
 \a (x,t) = \min \left(\amax, \max\left(\amin, -\frac{\di_x \v(x,t)}{2\e}\right)
-\right)
+\right).
 \\
@@
 \end{equation}
-We are solving $\v(x,t)$ over the domain $[\xmin, \xth] \times [0,T]$.
+We are solving $\v(x,t)$ over the domain $[\xmin, \xth] \times [0,\T]$.
 
 \subsection{The numerical method for the HJB equation}
-
+\label{sec:hjb_numerix}
 We now have a PDE for $\v$ and an algorithm for computing all the BCs and TCs of
 this PDE. It is time to discuss the numerical method for solving
-\cref{eq:OC_LS_HJB_full}. Since it is one-dimensional, it is straight forward to
-apply the standard centred finite difference using the Crank-Nicholson scheme to
-step in time. We have not found it necessary to upwind the first order
-derivatives for the regimes we have explored. To resolve the non-linearity,
-$(\di_x\v)^2$, in the PDE, we treat it as a mixed implicit-explicit term
-$$(\di_x \v(x,t_k))^2 = \di_x \v(x,t_k) \cdot \di_x \v(x,t_k) \approx
-\underbrace{\di_x \v(x,t_k)}_{\textrm{implicit}} \cdot
+\cref{eq:OC_LS_HJB_full}. Since it is one-dimensional in space, it is straight
+forward to apply the standard centred finite difference using the
+Crank-Nicholson scheme to step in time, see ch.\ 19.2 in \cite{Press1992}. To
+resolve the non-linearity, $(\di_x\v)^2$, in the PDE, we treat it as a mixed
+implicit-explicit term $$(\di_x \v(x,t_k))^2 = \di_x \v(x,t_k) \cdot \di_x
+\v(x,t_k) \approx \underbrace{\di_x \v(x,t_k)}_{\textrm{implicit}} \cdot
 \underbrace{\di_x\v(x,t_{k+1})}_{\textrm{explicit}}.$$ Note that the implicit
 term is in the previous time $t_k$ instead of, as is conventional, the next
-$t_{k+1}$, because we are solving for $\v$ {\sl backwards} in time.
+$t_{k+1}$, because we are solving for $\v$ {\sl backwards} in time from
+$t_{k+1}$ to $t_k$ down to  $t_0 = 0$.
 
@@
 \cite{Scipy}. For the discretization in time and space, we choose $\Delta x$ and
-$\Delta t$ in relation to the parameter regimes. for example in the
-Supra-Threshold Low-Noise regime, we have $\Delta x = 0.0043$ and $\Delta t =
-0.0012$. We consider these to be a fairly fine discretization.
+$\Delta t$ in relation to the parameter regimes. In particular, we take $\Delta
+x$ to be less than $\b$ divided by the largest possible absolute value of $(\m +
+\a - x/\tc)$ for $\a \in [\amin, \amax], x \in [\xmin, \xth]$; in our examples
+this always be equal to $\mu + \amax - \xmin/\tc$. This is an attempt to ensure
+that numerically we are in the diffusion-dominated regime. We set $\Delta t$ to
+be twice the value of $\Delta x$ divided by the largest possible absolute
+value of $(\m + \a - x/\tc)$. For example in the Supra-Threshold Low-Noise regime, we
+have $\Delta x = 0.0043$ and $\Delta t = 0.0013$. In all regimes, our
+discretization results in point-wise convergence of up to at least three
+significant digits for the value function. We consider this to be a sufficiently
+fine discretization.
 
@@
 
-In \cref{fig:HJB_4regimes_value_control_cuts}, we depict three snapshots
-in time of the value function and the control, $\v$ and $\a$, at $t = [0,\T/2,
-\T]$, i.e.\ at the beginning, middle and end of the relevant time interval. Then in
+In \cref{fig:HJB_4regimes_value_control_cuts}, we depict three snapshots in time
+of the value function and the control, $\v$ and $\a$, at $t = [0,\T/2, \T]$,
+i.e.\ at the beginning, middle and end of the relevant time interval. Then in
 \cref{fig:HJB_4regimes_value_surf} we show the entire value function, $\v(x,t)$,
@@
 \begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/HJB/Regimes_vc_cuts.pdf}
+  \includegraphics[width=\textwidth]{Figs/HJB/Regimes_vc_cuts.pdf}
   \caption[labelInTOC]{Snapshots of the value function, $\v$, on the left and on
   the optimal control, $\a$, on the right for $t$ fixed, at the start (blue),
-  mid-point (green) and end-point (red), for each of the four regimes. The desired spike time is set to $\T=1.5$, the energy penalty, $\eps
-  = 0.001$. The control bounds are $\a \in [-2,2]$. Note that the green and blue
-  curves in E) are lying on top of each other for all $x$.
-  \\
-  A,E) Supra-Threshold-Low-Noise
-  B,F) Supra-Threshold-High-Noise
-  C,G) Sub-Threshold-Low-Noise
-  D,H) Sub-Threshold-High-Noise
+  mid-point (green) and end-point (red), for each of the four regimes.
+  The desired spike time is set to $\T=1.5$ assuming the previous spike was at
+  $t=0$, the energy penalty, $\eps = 0.001$. The control bounds are $\a \in
+  [-2,2]$. 
+  Note that all red value function curves go to zero at the threshold to satisfy
+  the upper BC in \cref{eq:OC_LS_HJB_full}.
+  Note that the green and blue
+  curves in F) are lying on top of each other for all $x$.
+  \\
+  A,B) Supra-Threshold-Low-Noise
+  C,D) Supra-Threshold-High-Noise
+  E,F) Sub-Threshold-Low-Noise
+  G,H) Sub-Threshold-High-Noise
   }
@@
 The control bounds are $\a \in [-2,2]$. 
-The contours for fixed $t$ plotted in thick correspond to the graphs in \cref{fig:HJB_4regimes_value_control_cuts}.
+The coloured contours for fixed $t$ plotted in thick correspond to the
+curves plotted in the left panels of \cref{fig:HJB_4regimes_value_control_cuts}
+(blue: $t=0$, green: $t=t*/2$, red: $t=t*$).
 \\
@@
 given its terminal condition in \cref{eq:valuefun_TCs}. That is, the lower the
-value of $X_\T$ the more on average will we have to wait for the spike to occur.
+value of $X(\T)$ the more on average will we have to wait for the spike to
+occur.
 % This same basic shape represents the idea that we would like to be close to
@@
 i.e.\ it gives the maximum available push for the neuron to spike. This is
-implied \cref{eq:valuefun_TCs}. As we go back in time, however, the control $\a$
-decreases for $x \ll \xth$, and it becomes negative, i.e.\ inhibitory for $x
+consistent with \cref{eq:valuefun_TCs}. As we go back in time,
+however, the control $\a$ decreases for $x \ll \xth$, and it becomes negative, i.e.\ inhibitory for $x
 \lesssim \xth$. That is intuitively consistent with the problem objective. 
-For $t<\T$, we want to bring $X_t$ closer to the threshold, but not too close, to
+For $t<\T$, we want to bring $X(t)$ closer to the threshold, but not too close, to
 avoid early spiking.
@@
 \section{Open-loop Stochastic Control}
-When the value of $X_t$ is unobservable, the transition density can be used to
+When the value of $X(t)$ is unobservable, the transition density can be used to
 perform the optimization. Since the transition density follows a deterministic
 PDE, we apply a Maximum Principle for PDEs as a method of obtaining the optimal
-control, see \cite{Borzi2012} for details on optimization with PDEs.
+control, see \cite{Borzi2012} for details on optimization with PDEs, or the
+article, \cite{Annunziato2010}, for optimization with a
+Fokker-Planck-type of system.
 
@@
 We write the transition density of $X$, conditional on no
-spikes having occurred as:
+spikes having occurred since $t=0$ as:
 \begin{align*}
-f(x,t) \intd{x} &= \Prob[X_t\in \intd{x} \,\big|\, X_0 = 0,\, X_{s} < \xth \quad
-\forall s < t]
+f(x,t) \intd{x} &= \Prob[X(t)\in \intd{x} \,\big|\, X(0) = 0,\, X(s) < \xth
+\quad \forall s < t].
 \end{align*}
@@
 				\frac{\b^2 }{2}\cdot \di^2_x f -
-				\di_x \Big[ \left( \m + \a(t)- \frac{x}{\tc}\right)  \cdot f \Big]
-\\
-\\
-\begin{array}{lll}
-	&\textrm{st.\ }&
+				\di_x \Big[ \left( \m + \a(t)- \frac{x}{\tc}\right)  \cdot f \Big].
+\\
+\\
+\begin{array}{ll}
+	& 
 	\left\{ \begin{array}{lcl}
@@
 Note that we write $\a(t)$ here instead of $\a(x,t)$ since we cannot use the
-value of $X_t$.
+value of $X(t)$.
 
@@
 $$
-for the probability flow
+for the probability flux
 $$
-\phi(x,t) = (\m + \a(t) - \tfrac x \tc)f - \di_x [\tfrac {\b^2}2 f]
+\phi(x,t) = (\m + \a(t) - \tfrac x \tc)f - \di_x [\tfrac {\b^2}2 f].
 $$
@@
 $$
-\phi(x,t) |_{x=\xmin} \equiv 0
+\phi(x,t) |_{x=\xmin} \equiv 0.
 $$
@@
 $$ \L_\a[\cdot] := \di^2_x \Big[\frac{\b^2 }{2} \cdot\Big] -
- \di_x \Big[ (\m + \a(t)- \frac{x}{\tc}) \cdot \Big] $$where
+ \di_x \Big[ (\m + \a(t)- \frac{x}{\tc}) \cdot \Big],$$
+ where
  $[\cdot]$ indicates the argument of the operator $\L_\a$. We will usually
@@
 \subsection{Restating the objective in terms of the transition density}
-Let us recall our objective, \cref{eq:OC_LS_variance}:
-$$
-J[\a(\cdot)] = \Exp\left[
-(\ts - \T \big)^2
-+
-\e \int_0^\ts  \a^2(s) \intd{s}
-\right].
-$$
-Rewriting this in terms of the transition density, $f$, reads:
+We now derive the optimality conditions for the optimal control $\a^*$ for the
+open-loop context. Recall our objective, \cref{eq:OC_LS_variance}.  
+% $$
+% J[\a(\cdot)] = \Exp\left[
+% (\ts - \T \big)^2
+% +
+% \e \int_0^\ts  \a^2(s) \intd{s}
+% \right].
+% $$
+We can write this in terms of the transition density, $f$, as:
 \begin{align}
@@
 \\
-&+  \e \int_0^\ts  \a^2(t)  \int_\xmin^\xth f(x,t) \intd{x} \intd{t}
+&+  \e \int_0^\ts  \a^2(t)  \cdot \left(  \int_\xmin^\xth f(x,t) \intd{x}
+\right)
+\intd{t}.
 \notag
@@
 
-The first term $$ \int_\xmin^\xth \Ttwo(x) \cdot f(x,\T)\intd{x} $$ counts the cost of trajectories which spike too late. This cost is the expected
-squared time-to-hit starting at $x$, with  $\a = \amax$, weighted by the
-probability of $X_\T = x$, which is just $f(x,\T)$. Recall that $\Ttwo$ is
+The first term, $$ \int_\xmin^\xth \Ttwo(x) \cdot f(x,\T)\intd{x}, $$ counts the
+cost of trajectories which spike too late. This cost is the expected squared time-to-hit starting at $x$, with  $\a = \amax$, weighted by the
+probability of $X(\T) = x$, which is just $f(x,\T)$. Recall that $\Ttwo$ is
 defined in \cref{eq:Trem_squared}.
 
-The second term $$ \int_0^\T \phi(\xth,t) (t-\T)^2 \intd{t} $$ counts the cost
+The second term, $$ \int_0^\T \phi(\xth,t) (t-\T)^2 \intd{t}, $$ counts the cost
 of trajectories which spike too early, that is the squared difference between
 some realized spike time, $\ts = t$, and desired spike time $\T$, weighted by
-the probability of a spike at $t$ which is just the outflow of probability,
-$\phi(\xth,t)$. Recall that we assume that $\xth = 1$ throughout. Note further
-that due to the homogeneous Dirichlet BC at $\xth$, $f(\xth) = 0$, the outflow
-is simply: $$ \phi(\xth, t) = -\frac{\b^2 }{2} \di_x f(\xth, t).$$
-
-Finally, the third term
+the probability of a spike at $t$ which is just the outward probability flux at
+the threshold, $\phi(\xth,t)$. Recall that we assume that $\xth = 1$ throughout.
+Note further that due to the homogeneous Dirichlet BC at $\xth$, $f(\xth) = 0$, the outward flux is simply $$ \phi(\xth, t) = -\frac{\b^2 }{2} \di_x f(\xth,
+t).$$
+
+Finally, the third term,
 $$
-\e \int_0^\ts  \a^2(t)  \left[  \int_\xmin^\xth f(x,t) \intd{x} \right] \intd{t}
-$$
-is the energy cost: the inner integral, $\int_\xmin^\xth f(x,t) \intd{x}$
+\e \int_0^\ts  \a^2(t)  \cdot \left(  \int_\xmin^\xth f(x,t) \intd{x} \right)
+\intd{t}, $$
+is the energy cost; the inner integral, $\int_\xmin^\xth f(x,t) \intd{x}$,
 takes into account that we incur an energy cost only for those trajectories
@@
 $$
-\a^*(\cdot) = \argmin_{\a(\cdot)} J[\a(\cdot)]
+\a^*(\cdot) = \argmin_{\a(\cdot)} J[\a(\cdot)].
 $$
@@
 			- \Big[ \frac{\b^2 }{2}\cdot \di^2_x p +
-			(\m + \a(t)- \frac{x}{\tc})  \cdot \di_x p \Big]
+			(\m + \a(t)- \frac{x}{\tc})  \cdot \di_x p \Big].
 \end{aligned}
 \\
-\st
 \begin{cases}
+	p \big|_{x=\xth} &= (t-\T)^2
+	\\
+	\di_x p  \big|_{x=\xmin} &= 0
+	\\
 	p(x,\T) &= \Ttwo(x)
-	\\
-	\di_x p  \big|_{x=\xmin} &\equiv 0
-	\\
-	p \big|_{x=\xth} &\equiv (s-T)^2
 \end{cases}
@@
 \Big\{
- \tilde{\e}  2 \a(t)
+ \e 2 \a(t)
 + p f \Big|_\xmin
@@
 \Big\} = 0
-\quad \forall t \in [0,T]
+\quad \forall t \in [0,T].
 \label{eq:J_necessary_condition}
 \end{equation}
-
-More practically, the quantity,
-$$\di_\a J =  \Big\{
- \tilde{\e}  2 \a(t)
+Roughly speaking, \cref{eq:J_necessary_condition} corresponds to setting the
+derivative of the objective $J$  with
+respect to $\a$ to zero, in \cref{eq:OC_LS_variance_density} and the
+adjoint state $p$ acts as a Lagrange multiplier corresponding to the constraints of
+the transition density dynamics.
+
+More practically, the quantity
+
+\begin{equation}
+\delta_\a J (t) =  \Big\{
+- \int _\xmin^\xth p(x,t) \cdot \di_x f(x,t) \intd{x}
 + p(x,t) f(x,t) \Big|_\xmin
-- \int _\xmin^\xth p \cdot \di_x f(x,t) \intd{x}
++ \e  2 \a(t)
 \Big\}
-$$
+\label{eq:hamiltonian_gradient}
+\end{equation}
 gives the direction of increase of $J$ at $\a(t)$
@@
 Finally, we are ready to calculate the open-loop stochastic optimal control for
-the four parameter regimes. The calculated optimal controls for each regime are
-in \cref{fig:FBK_Regimes_cs}. Recall that the optimal control, $\a^*(t)$, is
-open-loop and thus it is only a function of time.
+the four parameter regimes. We use the same numerical method for solving the
+PDEs as described in \cref{sec:hjb_numerix}, since the PDEs for $f,p$ are very
+similar in structure to the PDE for $\v$ except they do not contain a nonlinear
+term. The details of the simple gradient descent algorithm for obtaining the
+optimal open-loop control are given in algorithm
+\ref{alg:gradient_descent_4_OC}. See \cite{Annunziato2013} for a more
+sophisticated descent method based on the conjugate-gradient method. For our
+applications, the algorithm in \ref{alg:gradient_descent_4_OC} typically
+converges in less than 10 iterations, although sometimes it can take longer. 
+\begin{algorithm}
+\begin{algorithmic}
+\State Fix $\T, \e, \ldots$ the problem parameters
+\State Fix $\{t_n\}_0^{N_t}$ a time-discretization of $[0,\T]$
+\State Fix $g_{tol}$ a convergence tolerance for the gradient
+\State Fix $K_{\max}, J_{\max}$ number of maximum iterations in outer, resp.
+inner descent loops
+\State Fix $s$ the initial step-size. 
+\\ {\itshape $\#$ we use $g_{tol}=1e-5,K_{\max}=100,J_{\max}=10,s=10$}
+\State $\a_1(t) \gets (\amax-\amin) \cdot t / \T + \amin$ 
+\\{\itshape  $\#$ $\a_1(t) \sim$ initial guess for the control, linear
+interpolate between $\amin, \amax$}
+\For { $k= 1\dots K_{\max}$} 
+\\ {\itshape $\#$ This is the outer loop where we descend down different
+gradients}
+ \State Calculate $f_k,J_{k},p_k, \delta_\a J_k$ corresponding to
+	$\a_k$ from
+	\cref{eq:hamiltonian_gradient,eq:adjoint_pde_OU,eq:FP_pde_OU_PDF,eq:OC_LS_variance_density}
+	\State $N_{active}\gets$   Number of time nodes $t_n$, where either
+	$\a_k(t_n) \neq \{\amin, \amax\}$ or $\delta_\a J_k(t_n)$ points inwards
+	\If{ $|| \delta_\a J_k||_{\R^{N_{active}}} \leq g_{tol}\cdot N_{active}$}
+		  \\ {\itshape  $\#$ 'Active' gradient is small enough,
+		 consider converged:}
+		 \State BREAK
+	\EndIf
+	\\ {\itshape $\#$ Find the step size, $s$, of how far to move $\a$ in the
+	direction $\delta_\a J_k$:}
+	\For { $j= 1\dots K_{\max}$}
+	\\ {\itshape $\#$ This is the inner loop where we find how much to descend down
+	the current gradient}
+	\State $\a_{k,j} \gets \a_{k} - s_j \cdot \delta_\a J_k  $
+	\\ {\itshape $\#$ $\a_{k,j}$ is the new control strategy to try}
+	\State Calculate $f_{k,j}, J_{k,j}$ corresponding to
+		$\a_{k,j}$ from	\cref{eq:FP_pde_OU_PDF,eq:OC_LS_variance_density}
+	\\ {\itshape $\#$ Recall $f_{k,j}$ is a probability density resulting from the
+	control $a_{k,j}$ and $J_{k,j}$ is the objective value resulting from
+	$f_{k,j}$}
+	\If {$J_{k,j} < J_k$}
+		\\ {\itshape $\#$ We found a better (smaller) objective value}
+		\State $s \gets 2 s_j$ {\itshape $\#$ try a more aggressive step in the
+		next iteration}
+		\State BREAK
+		\EndIf
+	\If {$j == J_{\max}$}
+		\\ {\itshape $\#$ We exhausted the step search without finding a 
+		smaller $J$, return current values}
+		\State Return $J_k, \a_k$
+	\EndIf
+ 	\\ {\itshape  $ \#$ Continue the inner looop, now try with a smaller step:}
+	 	\State $s_{j+1}$ \gets $s_j / 2.$
+    \EndFor  {\itshape $\quad \#$ single step loop}
+	\If {$k == K_{\max}$}
+		\State ERROR 'Could not converge'
+	\EndIf
+    \\{\itshape $\#$ Assign the new candidate for the optimal control and
+    re-loop}
+		\State $\a_{k+1} \gets \a_{k,j}$
+\EndFor {\itshape $\quad \#$ gradient descent loop}
+\State \Return $J_k, \a_k$
+\end{algorithmic}
+\caption{ Gradient descent algorithm for obtaining the optimal open-loop 
+control}
+\label{alg:gradient_descent_4_OC}
+\end{algorithm}
+
+The optimal controls obtained using algorithm \ref{alg:gradient_descent_4_OC} for each regime are in
+\cref{fig:FBK_Regimes_cs}. Recall that the optimal control, $\a^*(t)$, is open-loop and thus it is only a function of time.
 % \usepackage{graphics} is needed for \includegraphics
@@
 \begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/FP_Adjoint/Regimes_cs.pdf}
+%   \includegraphics[width=\textwidth]{Figs/FP_Adjoint/Regimes_cs.pdf}
+  \includegraphics[width=\textwidth]{Figs/FP_Adjoint/Regimes_cs_singleplot.pdf}
   \caption[labelInTOC]{The deterministic optimal controls for each parameter
@@
   = 0.001$ and the bounds are $\a \in [-2,2]$.
-  A) Supra-Threshold-Low-Noise B) Supra-Threshold-High-Noise
-C) Sub-Threshold-Low-Noise   D) Sub-Threshold-High-Noise  }
+  From left-to-right, the control curves correspond to the
+  Sub-Threshold-Low-Noise, Sub-Threshold-High-Noise, Super-Threshold-High-Noise,
+  Sub-Threshold-Low-Noise regimes
+%   A) Supra-Threshold-Low-Noise 
+%   B) Supra-Threshold-High-Noise
+%   C) Sub-Threshold-Low-Noise
+%   D) Sub-Threshold-High-Noise  
+  }
   \label{fig:FBK_Regimes_cs}
@@
 the beginning of the interval, $\a(t) = \amin$ for $t \ll \T$ and to excite it
-near the end, $\a(t) = \amax$ for $t \approx \T$. The smooth portion that connects these
-two segments in the middle of the interval is due to the energy penalty, $\e \int
-\a^2(s) \intd{s}$. The behaviour of the control in different regimes
-is relatively simple to explain and we see what we would expect. In a
+near the end, $\a(t) = \amax$ for $t \approx \T$. The smooth portion that
+connects these two segments in the middle of the interval is due to the energy
+penalty, $\e \int \a^2(s) \intd{s}$. The behaviour of the control in different
+regimes is relatively simple to explain and we see what we would expect. In a
 sense the bias current $\mu$ can be absorbed by the control and so increasing
 $\mu$ amounts to decreasing $\a$ modulo its bounds. That is exactly the
-difference between plots A) vs.\ C) and B) vs.\ D) in \cref{fig:FBK_Regimes_cs},
-where the reduction of $\mu$ in the lower plots, C) and D), results in an
-increase in $\a(t)$ compared to A) and B), respectively. The effect of the noise
-intensity, $\b$, is more subtle and it is not the same in the Supra vs.\ the
-Sub-Threshold regimes A,B) vs.\ C,D). In the Supra-Threshold regime,  panels
-A,B), reducing $\b$, reduces the need for excitatory  control towards the end,
-since the bias alone should put the neuron over the threshold. Thus the main
-part of the control with low noise in the Supra-Threshold regime is to stop the
-neuron from spiking too early. In the Sub-Threshold regime, panels C,D), the
-situation is somewhat reversed - reducing $\b$, obviates the need to apply an
-inhibitory control in the early part of the interval and also prompts the
-controller to apply excitatory input earlier, since it is the controller which
-is now the main drive for a spike.
+difference between the Super-Threshold vs.\ Sub-Threshold plots in
+\cref{fig:FBK_Regimes_cs} when holding the noise intensity fixed, either
+$\b=0.3$ or $\b=1.5$. Both for low and high noise intensity, a reduction of $\mu$
+results in an increase in $\a(t)$. 
+The effect of varying the noise intensity, $\b$, is more subtle and it is not
+the same in the Supra vs.\ the Sub-Threshold regimes. In the
+Supra-Threshold regime, reducing $\b$ reduces the need for
+excitatory control towards the end, since the bias alone should put the neuron
+over the threshold. Thus the main part of the control with low noise in the
+Supra-Threshold regime is to stop the neuron from spiking too early. In the
+Sub-Threshold regime, the situation is somewhat reversed - reducing
+$\b$ obviates the need to apply an inhibitory control in the early part of the
+interval and also prompts the controller to apply excitatory input earlier,
+since it is the controller which is now the main drive for a spike.
 
@@
 another control law - perhaps the most naive one - which is obtained by ignoring
-the noise and the energy penalty. That is we find a {\sl constant} value of $\a$
-which satisfies the desired boundary conditions, $x(0) = 0, x(\T) = \xth$. This naive
+the noise and the energy penalty. That is, we find a {\sl constant} value of
+$\a$ which satisfies the desired boundary conditions, $x(0) = 0, x(\T) = \xth$. This naive
 controller will be called 'Deterministic', since it assumes deterministic
-dynamics in $X_t$, i.e.\ $\b = 0$.
+dynamics in $X$, i.e.\ $\b = 0$.
 
@@
 For illustration sake, we also visualize some trajectories from the
-Sub-Threshold-High-Noise regime in \cref{fig:control_trajectories_examples},
-see panels A),C),E).
+Sub-Threshold-High-Noise regime in \cref{fig:control_trajectories_examples}, see
+panels A),C),E). The most notable feature of the $\a(\cdot)$ plots in
+\cref{fig:control_trajectories_examples} is that especially for $\eps=0.001$,
+there is a high level of fluctuations in the closed-loop $\a$. That is to be
+expected from the optimality condition in
+\cref{eq:OC_LS_variance_energy_quadratic_control}. Since the control is
+proportional to the derivative of the value function and we are minimizing,
+roughly speaking, the control attempts to push the process, $X$, to the bottom
+of the valley that is formed by the value function, $\v$. However as the
+stochastic fluctuations of $X(t)$ push it randomly to either side of that
+'valley' and in turn force the control to change signs to counter-act.
+These fluctuations are then amplified by the division by the small parameter
+$\eps$ so that the control mostly bangs up or down to its extremes
+$\amin, \amax$.
 
@@
 \subfloat[Supra-Threshold Low-Noise]{
-  \includegraphics[width=.99\textwidth]
+  \includegraphics[width=\textwidth]
   {Figs/ControlSimulator/Regimes_SUPT_ln_errors_hist.pdf}
@@
 \subfloat[Supra-Threshold High-Noise]{
-  \includegraphics[width=.99\textwidth]
+  \includegraphics[width=\textwidth]
   {Figs/ControlSimulator/Regimes_SUPT_HN_errors_hist.pdf}
@@
 \subfloat[Sub-Threshold Low-Noise]{
-  \includegraphics[width=.99\textwidth]
+  \includegraphics[width=\textwidth]
   {Figs/ControlSimulator/Regimes_subt_ln_errors_hist.pdf}
@@
 \subfloat[Sub-Threshold High-Noise]{
-  \includegraphics[width=.99\textwidth]
+  \includegraphics[width=\textwidth]
   {Figs/ControlSimulator/Regimes_subt_HN_errors_hist.pdf}
@@
   \caption[labelInTOC]{Histogram of the spike timing error for the
-  deterministic (left) vs. open-loop stochastic(centre) vs. closed-loop
-  stochastic (right) control laws. This is for the same problem as in
+  deterministic (left) vs. open-loop stochastic (centre) vs. closed-loop
+  stochastic (right) control laws.
+  Recall that $\ts$ is the random realized spike time and $\T$ is the target
+  spike time.  
+  This is for the same problem as in
   \cref{fig:control_trajectories_examples}.
   We have used $N=10000$ sample paths
-  to form the statistics}
+  to form the statistics.}
   \label{fig:error_histograms_det_vs_openloop_vs_stoch}
+\end{center}
+\end{figure}
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.99\textwidth]
+{Figs/ControlSimulator/Composite_Traj.pdf} 
+\caption[]{Three examples for the controlled trajectories using the
+deterministic, open-loop stochastic and closed-loop stochastic control approaches.
+The plots on the left, A,C,E, are obtained using a value of $\eps=0.001$,
+while the plots on the right, B,D,F, use $\eps=0.1$ (for a discussion on the
+effect of $\eps$ see \cref{sec:effect_of_eps}). The black vertical line in the plots indicates the
+desired spike-time, $\T$. The parameter values are $\m, \tc, \b = [0.2, 0.5,  1.5]$ (Sub-Threshold
+High-Noise regime), with $\T = 1.5$. 
+The upper plots show the voltage evolution, $X(t)$, while the lower
+plots show the applied control, $\a(t)$. Note that the optimal control, $\a(t)$, obtained from
+the deterministic and from the open-loop controls are the same across all three
+samples.}
+\label{fig:control_trajectories_examples}   
 \end{center}  
@@
 
+\clearpage
+
 \subsection{Multi-Spike Control}
 We now turn to the ultimate goal of our analysis, the control of {\sl spike
-trains}. In particular, we will generate $M=10$ realizations of $N=20$ spikes
+trains}. In particular, we will generate $M=50$ realizations of $N=16$ spikes
 each in an attempt to meet a prescribed spike train. We will focus on two
-regimes, the Supra-Threshold regime with either low or high noise.
-The results
+regimes, the Supra-Threshold regime with either low or high noise. The results
 for parameters from the low-noise regime are shown in
-results in \cref{fig:targettrain_cl_lownoise,fig:targettrain_ol_lownoise}, 
-while, the results
-for the high-noise regime are shown in
-results in
+\cref{fig:targettrain_cl_lownoise,fig:targettrain_ol_lownoise}, while, the
+results for the high-noise regime are shown in
 \cref{fig:targettrain_cl_highnoise,fig:targettrain_ol_highnoise}. In all cases,
@@
 regime with no additional control, i.e.\ $\a(\cdot) = 0$.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%% SUPER-THRESHOLD REGIME LOW-NOISe:
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+It is immediately clear that while the evoked trains closely follow the target
+train in the low-noise case,
+\cref{fig:targettrain_cl_lownoise,fig:targettrain_ol_lownoise}, this is a lot
+more difficult in the high-noise case,
+\cref{fig:targettrain_cl_highnoise,fig:targettrain_ol_highnoise}.
+
+In each figure,
+\cref{fig:targettrain_cl_lownoise,fig:targettrain_ol_lownoise,fig:targettrain_cl_highnoise,fig:targettrain_ol_highnoise}
+on the top panel we show a smoothed version of the target train vs.\ an
+empirical firing rate from the simulations. The red curve is a smoothed Gaussian
+kernel applied at the target times where the standard deviation of the kernel is
+$\sigma = 0.1$. The blue curve is the empirical firing rate equal to the number
+of spikes in a small window divided by the number of realizations and divided by
+the width of the window $w_b$. We have used $w_b = 0.1$. In the bottom panels of
+\cref{fig:targettrain_cl_lownoise,fig:targettrain_ol_lownoise,fig:targettrain_cl_highnoise,fig:targettrain_ol_highnoise},
+we show in detail $10$ out of the $M=50$ realizations.
+
+In the high-noise case, the restricted control is obviously not able to produce
+reliable results. A stronger control is needed to obtain the same level of
+accuracy in the target. In \cref{fig:targettrain_cl_highnoise_aplus}, we relax
+the bound constraints on the control $[\amin, \amax]$ from $[-2; 2]$ to $[-4;
+4]$. Then a significantly better tracking of the target train is achieved. Thus,
+to control a system under higher noise comes at the price of allowing a more
+powerful control signal.
+
+The main question in this section is whether the separate optimization
+of each interval independently of the others is a suitable approach when
+targeting a whole spike train. In particular, we do not account 
+for the possibility that two target spikes occur closely together. Then
+it might make sense to attempt generating the first
+one a little earlier, on average, to give more time to generate the
+second. For example, in figs. 6 and 7, the 13th target spike is
+successfully produced, but the 14th target spike occurs so soon after
+that in all trials it is delayed. However, if the target interspike
+intervals are within reasonable reach of the neuron, this is not a problem.
+
+% In \cref{fig:targettrain_cl_highnoise_aplus}, we see that once we relax the
+% bound constrains on the control $\amin, \amax$ from $[-2,2]$ to $[-4,4]$, we can
+% achieve significantly better tracking of the target train.
+% 
+% Note that the main question in this section is whether the separate optimization
+% of each interval independently of the others 'works' when we are working with a
+% whole train. In particular, we do not account for the following possible
+% occurence, if two target spikes occur closely together, then it might make sense
+% to attempt generating the first one a little earlier, on average, to give more
+% time to generate the second. At least visually however, we see that this is not
+% necessarily a problem in
+% \cref{fig:targettrain_cl_lownoise,fig:targettrain_ol_lownoise,fig:targettrain_cl_highnoise,fig:targettrain_ol_highnoise}.
+% 
+% TODO: I DON''T REALLY KNOW HOW TO ARGUE THIS - IT'S A POSSIBLE PROBLEM, BUT IT
+% IS HARD TO SAY WHETHER IT OCCURS IN PRACTICE, WHEN IT OCCURS, DOES IT OCCUR IN
+% OUR SIMULATIONS  OR DOES IT NOT\ldots?
+
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%
+% SUPER-THRESHOLD REGIME LOW-NOISe: %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[htp]
 \begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_ln_cl_trains_sim_10.pdf}
+  \includegraphics[width=\textwidth]{Figs/TrainController/SUPT_ln_cl_trains_sim_50.pdf}
   \caption[ ]{How good the {\sl closed-loop} controller is at hitting a target
-  spike train. Uses the parameters from the Super-Threshold,Low-Noise regime.
-  The target spike train was generated using the model itself, with
-  parameters from the Super-Threshold, High-Noise regime, without an applied
+  spike train. In the panel above, the blue curve is the empirical firing
+  rate of $M$ controlled trains.   strategy, the red curve is a
+  smoothed-version of the target train. (See text for details of how they are calculated). In the panel below, the
+  dashed, red lines indicate the target times, $\t_n$, while the blue dots are
+  the realizations of the controlled trains.  
+  The parameters from the Super-Threshold, Low-Noise
+  regime are used. The target spike train was generated using the model itself,
+  with parameters from the Super-Threshold, High-Noise regime, without an applied
   control, $\a=0$.
-  There are $N=16$ spikes and $M=10$ realizations of the control strategy} 
+  There are $N=16$ target spikes and $M=50$ realizations of the control strategy} 
   \label{fig:targettrain_cl_lownoise}
@@
 \begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_ln_ol_trains_sim_10.pdf}
+  \includegraphics[width=\textwidth]{Figs/TrainController/SUPT_ln_ol_trains_sim_50.pdf}
   \caption[ ]{How good the {\sl open-loop} controller is at hitting a target
-  spike train. Uses the parameters from the Super-Threshold,Low-Noise regime.
-  The target spike train was generated using the model itself, with
-  parameters from the Super-Threshold, High-Noise regime, without an applied
-  control, $\a=0$.
-  There are $N=16$ spikes and $M=10$ realizations of the control strategy}
+  spike train.
+  The drawn curves have  the same meaning as in
+  \cref{fig:targettrain_cl_lownoise}.
+  The parameters from the Super-Threshold, Low-Noise regime are used. 
+  The target spike train was generated using the model itself, with parameters
+  from the Super-Threshold, High-Noise regime, without an applied control, $\a=0$.
+  There are $N=16$ target spikes and $M=50$ realizations of the control strategy}
   \label{fig:targettrain_ol_lownoise}
@@
 \begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_HN_cl_trains_sim_10.pdf}
+  \includegraphics[width=\textwidth]{Figs/TrainController/SUPT_HN_cl_trains_sim_50.pdf}
   \caption[ ]{How good the {\sl closed-loop} controller is at hitting a target
-  spike train. Uses the parameters from the Super-Threshold,High-Noise regime.
-  The target spike train was generated using the model itself, with
+  spike train. 
+  The drawn curves have  the same meaning as in
+  \cref{fig:targettrain_cl_lownoise}.
+  The parameters from the Super-Threshold, High-Noise regime
+  are used. The target spike train was generated using the model itself, with
   parameters from the Super-Threshold, High-Noise regime, without an applied
   control, $\a=0$.
-  There are $N=16$ spikes and $M=10$ realizations of the control strategy}
+  There are $N=16$ target spikes and $M=50$ realizations of the control strategy}
   \label{fig:targettrain_cl_highnoise}
@@
 \begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_HN_ol_trains_sim_10.pdf}
-  \caption[ ]{How good the {\sl open-loop} controller is at hitting a target
-  spike train. Uses the parameters from the Super-Threshold,High-Noise regime.
-  The target spike train was generated using the model itself, with
+  \includegraphics[width=\textwidth]{Figs/TrainController/SUPT_HN_ol_trains_sim_50.pdf}
+  \caption[]{How good the {\sl open-loop} controller is at hitting a target
+  spike train. 
+  The drawn curves have  the same meaning as in
+  \cref{fig:targettrain_cl_lownoise}.
+  The parameters from the Super-Threshold, High-Noise regime
+  are used. The target spike train was generated using the model itself, with
   parameters from the Super-Threshold, High-Noise regime, without an applied
   control, $\a=0$.
-  There are $N=16$ spikes and $M=10$ realizations of the control strategy}
+  There are $N=16$ target spikes and $M=50$ realizations of the control strategy}
   \label{fig:targettrain_ol_highnoise}    
@@
 
-\clearpage
-
-\section{Miscleaneous Analysis???}
-TODO: WE DON''T KNOW WHAT TO CALL THIS SECTION!
-\subsection{Effect of energy penalty}
+\begin{figure}[htp]
+\begin{center}
+  \includegraphics[width=\textwidth]{Figs/TrainController/SUPT_HN_cl_aplus_trains_sim_50.pdf}
+  \caption[ ]{Same as \cref{fig:targettrain_cl_highnoise}, but with the control
+  constraints, $[\amin, \amax]$ increased to $[-4,4]$ instead of $[-2,2]$.}
+  \label{fig:targettrain_cl_highnoise_aplus}    
+\end{center}
+\end{figure} 
+% \clearpage
+
+\section{Effect of energy penalty}
 \label{sec:effect_of_eps}
@@
 $\eps$, in the objective defined in \cref{eq:OC_LS_variance_energy}. Our
-approach has been that we are most concerned with the accurate spiking and that
-penalizing energy is rather intended to regularize the control and avoid
+approach has been to be most concerned with the accurate spiking and
+that penalizing energy is rather intended to regularize the control and avoid
 excessive chattering of the control between its extreme values, than as a goal
@@
 Note however that increasing $\eps$ has an important effect on the validity of
-the approximation, used to form the terminal conditions for
-the value function and the adjoint variable, $\v,p$. We assumed that applying
-$\amax$ is optimal for $t>\T$. This is indeed true in the
-limit $\eps \ra 0$. However, even for a finite value of $\eps$, this may still
-be the optimal thing to do. Indeed, in the original calculations,
-where $\eps=0.001$, see the panels on the left in
+the approximation, used to form the terminal conditions for the value function,
+$\v$, and the adjoint variable, $p$. We assumed that applying $\amax$ is
+optimal for $t>\T$. This is indeed true in the limit $\eps \ra 0$. However, even for a
+finite value of $\eps$, this may still be the optimal thing to do. Indeed, in
+the original calculations, where $\eps=0.001$, see the panels on the left in
 \cref{fig:HJB_4regimes_control_different_eps,fig:FBK_Regimes_cs_different_es},
-the so-obtained value function {\sl implied} that $\a(x,\T) = \amax$.
-I.e.\ our
-guess that $\a(x, t) = \amax$ for $t > \t$, is self-consistent with the
-so-obtained value function. This is akin to confirming an ansatz.
-For a higher value of $\eps$, like $\eps = 0.1$, this is no longer the case. The red
-curves on the right-side of \cref{fig:HJB_4regimes_control_different_eps} are no
-longer pushed up at $\a = \amax$, and as such it is no longer valid to
-assume that the value function can be obtained by assuming $\a(t>\T) = \amax$
-and then just calculating the expected remaining time-to-spike.
-The exact same conclusion can be drawn from the open-loop control. There the
-optimal control always (almost) reaches its maximum, $\amax$ while choosing
-$\eps = 0.001$, but no longer so for the higher $\eps = 0.1$.
-In fact, as we mentioned after \cref{eq:valuefun_TCs}, the correct thing to do
-for higher values of $\eps$, is to push the
-calculation interval to some $t^+ > \T$, apply the same terminal condition at
-$t^+$ instead of $\T$ and then solve backwards (either for the closed- or
-open-loop control). One will be guaranteed to find some $t_f > \T$ that works
-since eventually the quadratic term $(t-\T)^2$ will dominate the energy term in
-the objective, which is linear in $t$. We do not explore this further here.
+the so-obtained value function {\sl implied} that $\a(x,\T) = \amax$. In
+other words, our guess that $\a(x, t) = \amax$ for $t > \T$, is
+self-consistent with the so-obtained value function. This is akin to confirming an ansatz. For a higher
+value of $\eps$, like $\eps = 0.1$, this is no longer the case. The red curves
+on the right-side of \cref{fig:HJB_4regimes_control_different_eps} are no longer
+pushed up at $\a = \amax$, and as such it is no longer valid, strictly speaking,
+to assume that the value function can be obtained by assuming $\a(t)|_{t>\T} =
+\amax$ and then calculating the expected remaining time-to-spike. The exact same
+conclusion can be drawn from the open-loop control. There the optimal control
+always (almost) reaches its maximum, $\amax$ while choosing $\eps = 0.001$,
+this is no longer so for the higher $\eps = 0.1$. In fact, as we mentioned
+after \cref{eq:valuefun_TCs}, the correct thing to do for higher values of $\eps$ is
+to push the calculation interval to some $t^+ > \T$, apply the same terminal
+condition at $t^+$ instead of $\T$ and then solve backwards (either for the
+closed- or open-loop control). One will be guaranteed to find some $t^+ > \T$
+that works since eventually the quadratic term $(t-\T)^2$ will dominate the
+energy term in the objective, which is linear in $t$. We do not explore this
+further here.
 % CLOSED_LOOP EPS: 
@@
 \begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/HJB/Regimes_eps_comparison.pdf}
+  \includegraphics[width=\textwidth]{Figs/HJB/Regimes_eps_comparison.pdf}
   \caption[labelInTOC]{Effect of $\e$ on the closed-loop control. Snapshots of
@@
 \begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/FP_Adjoint/Regimes_eps_comparison.pdf}
+  \includegraphics[width=\textwidth]{Figs/FP_Adjoint/Regimes_eps_comparison.pdf}
   \caption[labelInTOC]{Effect of $\e$ on the open-loop control.
@@
   \label{fig:FBK_Regimes_cs_different_es} 
-\end{center}
-\end{figure}
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=0.99\textwidth]
-{Figs/ControlSimulator/Composite_Traj.pdf}
-\caption[]{Three examples for the controlled trajectories using the
-deterministic, open-loop stochastic and closed-loop stochastic control approaches.
-The plots on the left are obtained using a value of $\eps=0.001$, while the
-plots on the right use $\eps=0.1$.
-The black vertical line in the plots indicates the desired spike-time, $\T$.
-The parameter values are $\m, \tc, \b = [0.2, 0.5,  1.5]$ (Sub-Threshold
-High-Noise regime), with $\T = 1.5$. 
-On the upper plots, we show the voltage evolution, $X_t$, on the lower plots we
-show the applied control, $\a(t)$. Note that the optimal control, $\a(t)$, obtained from
-the deterministic and from the open-loop controls are the same across all three
-samples.}
-\label{fig:control_trajectories_examples}
 \end{center}
@@
 
-\subsection{Control Robustness}
+\section{Robustness of Control}
 We have so far assumed that we know the parameters, $\m, \tc, \b$, determining
-the dynamics in \cref{eq:X_evolution_uo,eq:current_mu_alpha} exactly. That is we
-assume that the values for $\tc, \b$ used to obtain the controls are the same as the ones that drive the
-controlled dynamics. That is slightly naive, as in any realistic application
-these will have to be estimated and the estimates will have some error
-associated with them.
-Note that misestimating $\mu$ has the most immediate and clear effect on the
-control policy as that amounts to a direct shift of the control and so for now,
-let us assume that $\mu$ is estimated correctly.
+the dynamics in \cref{eq:X_evolution_uo,eq:current_mu_alpha} exactly. That is,
+we assume that the values for $\tc, \b$ used to obtain the controls are the same
+as the ones that drive the controlled dynamics. That is slightly naive, as in
+any realistic application these will have to be estimated and the estimates will
+have some error associated with them. Note that misestimating $\mu$ has the most
+immediate and clear effect on the control policy as that amounts to a direct
+shift of the control and so for now, let us assume that $\mu$ is estimated
+correctly.
 
@@
 Assume that $\b$ and $\tc$ are underestimated, so that $\b_{est} = \b / 1.5$
-and $\tc_{est} = \tc / 1.5$.
-
-So we will have:
-$$\mu, \tc_{est}, \b_{est} = \{.2, .5, 1.5 \}$$ which is the control
+and $\tc_{est} = \tc / 1.5$. We will then have: 
+$$\mu, \tc_{est}, \b_{est} = \{0.2, 0.5, 1.5 \},$$ which is the control
 calculated already for the Sub-Threshold,
 High-Noise regime, while the actual simulated dynamics will be driven by
-$$\mu, \tc, \b = \{.2, .75, 2.25 \}$$
+$$\mu, \tc, \b = \{0.2, 0.75, 2.25 \}.$$
 
@@
 a misspecification of $\b$ should not have a direct influence on the behaviour
-of the 'deterministic' controller since it already assumes in any case that $\b=0$.
+of the 'deterministic' controller since it assumes in any case that $\b=0$.
 \begin{table}[h]
 \begin{center}
- \subfloat[Sub-Threshold High-Noise]{
- \input{Figs/ControlSimulator/Misspec_MeanSquaredErrors__subt_HN.txt}}\\
+ \input{Figs/ControlSimulator/Misspec_MeanSquaredErrors__subt_HN.txt}
  \caption{The effect of misspecifying the parameters. On the left, the
  system parameters and the parameters used to obtain the control are the
- same, i.e.\ accurate, on the right, they are misspecified.}
+ same, i.e.\ correct, on the right, they are misspecified.}
 \label{tab:realized_avg_errors_det_vs_openloop_vs_stoch_misspecified}
@@
 level of the noise. In a sense, the noise acts like an adversary - in our
-context it has no beneficial role, but to obstruct precise spiking.
+context it has no beneficial role, except to obstruct precise spiking.
+When the level of the noise is high, a stronger control is needed, and
+the trade-off between accuracy (hitting the target spikes), possible
+damage of the system (the limits of the control) and energy expenditure
+has to be considered.
 
@@
 control as the system will naturally tend to spike earlier than desired.
-
-In both cases finding the optimal control is computationally intensive as we
-numerically solve partial differential equations. If these algorithms were to be
-practical in an online setting, some more work would have to be done in order to
-ensure they can be computed very quickly (in the millisecond range) or that
-they can be pre-computed.
-
-Our work has strayed pretty close to the paradigm of \cite{Ahmadian2011} in
-trying to make the neuron spike at a particular time. Other goals have been
-formulated, most notably the idea of desynchronizing a population of neurons,
-\cite{Nabi2011}, where it is not so important when the neurons spike as long as
-they do not spike at the same time. We have also incorporated into our objective
+In summary, we have found that systems in the Super-Threshold, Low-Noise regime
+tend to be most accurately controlled, while the Super-Threshold, High-Noise
+regime are the least accurate.
+
+It should be again noted that we do not require that the controlled cell is
+already in the Supra-Threshold regime - the control scheme applies in the same
+manner in the Sub-Threshold regime as well.
+
+In both contexts, open-loop and closed-loop, finding the optimal control is
+computationally intensive as we numerically solve partial differential
+equations. If these algorithms were to be practical in an online setting, some
+more work would have to be done in order to ensure they can be computed very
+quickly (in the millisecond range) or that they can be pre-computed.
+
+Our work has stayed close to the paradigm of \cite{Ahmadian2011} in
+trying to make the neuron spike at a particular time. 
+Similiar to them, we have also incorporated into our objective
 the minimization of total energy used to achieve our goal, which as discussed in
-\cite{Ahmadian2011} is sensible given the potential damage on the cell of
+\cite{Ahmadian2011} is sensible given the potential damage to the cell of
 accumulating charge. Furthermore, we have assumed that our control is
@@
 al. \cite{Danzl2010}. It is possible to add them to the open-loop control, since
-the control is still deterministic, but it is not-trivial. It is even less
+the control is still deterministic, but it is non-trivial. It is even less
 trivial when one uses Dynamic Programing. In
@@
 open-loop controls.
+
+% Alternative objectives to spike-timing control in the literature have been
+% formulated, most notably the idea of desynchronizing a population of neurons,
+% \cite{Nabi2011}, where it is not so important when the neurons spike as long as
+% they do not spike at the same time. It is  Maybe you would like to
+% write that as a future application, in which case there would be many neurons in parallel to be
+% controlled. Alternately, one can imagine that each neuron is firing in the
+% suprathreshold regime with a bit of noise, and you wish to apply independent
+% controls that makes each one fire the most Poisson-like. If this were achieved
+% for each neuron in a population, it would surely have the result of
+% desynchronizing them.
+
+As points of outlook, we mention that our scheme can potentially be applied to
+produce a periodic train in the face of noise with applications to various
+biological pacemakers, or it can also be used to produce the opposite effect - a
+random behaviour at the single cell level, for example by specifying a target
+train with Poisson-like characteristics.
 
@@
 this scope, the numerical demands on our scheme are non-trivial, but our
-experience and experiments show that these demands are potentially feasible. It
-however remains to be seen whether they can be successfully applied in the lab
+experience and experiments show that these demands can be addressed.
+It however remains to be seen whether they can be successfully applied in the lab
 and beyond.
@@
 \label{sec:valuefun_TCs}
-Here we explain how to calculate the first two moments of the time to-spike,
-$\Exp \big[\trem^p \,\Big|\, X_\T = x, \a(t) = \amax \big].$, which is needed in setting the terminal conditions for both,
-$\v(x,\T)$ and $p(x,\T)$, see \cref{eq:OC_LS_HJB_full} and \cref{eq:adjoint_pde_OU}.
+Here we explain how to calculate the first two moments of the remaining
+time-to-spike, $\Exp \big[\trem^n \,\Big|\, X(\T) = x, \a(t) = \amax
+\big]$, $n=1,2$, which is needed for the terminal conditions for $\v(x,\T)$
+and $p(x,\T)$ in, respectively, \cref{eq:OC_LS_HJB_full} and
+\cref{eq:adjoint_pde_OU}.
 
@@
 \begin{array}{lcl}
-\Tone(x_0) &=& \Exp \Big[\trem \,\Big|\, X_\T = x, \a(t) = \amax
-\Big]
-\\
-\Ttwo(x_0) &=&
-\Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big],
+\Tone(x) &=& \Exp \Big[\trem \,\,\Big|\, X(\T) = x, \a(t) = \amax
+\Big],
+\\
+\Ttwo(x) &=&
+\Exp \Big[\trem^2 \,\Big|\, X(\T) = x, \a(t) = \amax \Big],
 \end{array}
 \end{equation}
-where as before, $\t$ is the remaining time-to-spike in
-\cref{eq:X_evolution_uo}, at the desired spike-time $t=\T$, given $X_\T = x_\T$
-and $\a(t) = \amax$.
-Naturally, the $\Ti$'s depend on $x_\T$ and are affected by the value of the
-various parameters, $\a,\b,\tc$.
-
-Actually, we only  need $\Ttwo$ for the TCs in \cref{eq:OC_LS_HJB_full} and \cref{eq:adjoint_pde_OU}.
-but $\Tone$ is needed as an auxiliary.
-
-Indeed, analytical expressions exist for $\Ti$, e.g.\ see \cite{Inoue1995}.
+where as before, $\ts-\T$ is the remaining time-to-spike in
+\cref{eq:X_evolution_uo}, with the voltage being $X(\T) = x$ at the desired
+spike-time $t=\T$ and applying the control $\a(t) = \amax$ afterwards.
+Naturally, the $\Ti$'s depend on $x$ and are affected by the value of the
+various parameters, $\a,\b,\tc$. We only  need $\Ttwo$ for the TCs in
+\cref{eq:OC_LS_HJB_full} and \cref{eq:adjoint_pde_OU}, but $\Tone$ is needed as
+an auxiliary.
+
+Analytical expressions exist for $\Ti$, e.g.\ see \cite{Inoue1995}.
 However in our experience, they are awkward to work with numerically, especially
@@
 calculations. Moreover, they are not particularly convenient if you need $\Ti
-(x_0)$ for a wide range of $x_0$'s as we do here, since they calculate the value
+(x)$ for a wide range of $x$'s as we do here, since they calculate the value
 independently for each $x$. Instead we will follow and slightly extend Jacobs\footnote{Jacobs only discusses how to calculate the first moment
@@
 example.}, \cite{Jacobs}, in deriving and solving a differential equation for
-$\Ti(x_0)$.
-
-Basically we are calculating the mean-time-squared to exit an interval $[\xmin,
-\xth]$. Theoretically, $\xmin = -\infty$, but for our purposes we will set it to
-some finite value and impose reflecting boundaries there - this is a very good
-approximation for $\xmin \ll 0$, as long as $\a = \amax >0$, which is what we
-have.
-
-Let $B(y,t|x) = \Prob[X_0 = y| X_t = x]$ for a fixed $x$. Then $B$ solves the
+$\Ti(x)$.
+
+Basically we are calculating the mean squared time to exit an interval
+$[\xmin, \xth]$. Theoretically, $\xmin = -\infty$, but for our purposes we will set it to
+some finite value and impose reflecting boundaries there. This is a very
+good approximation for $\xmin \ll 0$, as long as $\a = \amax >0$.
+
+Let $B(y,t|x) = \Prob[X_0 = y| X(t) = x]$ for a fixed $x$. Then $B$ solves the
 following (backward Kolmogorov) PDE
 \begin{equation}
--\di_t B = (\a - \frac y\tc)\di_yB - \frac {\b^2} 2 \di_y^2 B
+-\di_t B = (\m + \a - \frac y\tc)\di_yB - \frac {\b^2} 2 \di_y^2 B.
 \label{eq:backward_Xdensity}
 \end{equation}
-
-Let $\G(t,y)$ be the survival function for $T$ for $X_0 = y$.
-$$\G(t,y) = \Prob[T>t | X_0 = y] = \Prob[X_{s\leq t} \in [\xmin,
-\xth] | X_0 = y]$$.
-
-Because the dynamics do not depend explicitly on time, we have:
+Let $\G(t,y)$ be the survival function for $\ts$ given $X(\T) = y$,
+$$\G(t,y) = \Prob[\T>t | X(\T) = y] = \Prob[X(s)|_{s \leq t} \in [\xmin,
+\xth] \,\big|\, X(\T) = y].$$
+Because the dynamics do not depend explicitly on time, we have
 $$
@@
 \begin{equation}
--\di_t \G = (\a - \frac y\tc) \di_y\G - \frac {\b^2} 2 \di_y^2 \G
+-\di_t \G = (\m + \a - \frac y\tc) \di_y\G - \frac {\b^2} 2 \di_y^2 \G,
 \label{eq:backward_SDF}
 \end{equation}
-
-With the BCs, ICs:
+with the BCs, ICs:
 \begin{equation}
 \begin{cases}
-\G(t,\xth) = .0 & \quad (T = t \implies \Prob[T>t] = 0)
-\\
-\di_x \G(t,\xmin) = .0  & \quad (\textrm{due to reflecting boundary for X})
-\\
-\G(0, y) = 1. & \quad (T \textrm{ is definitely} > 0 \textrm{ for } x_0 \in
-(\xmin, \xth) )
+\G(t,\xth) = 0 & \quad (\ts = t \implies \Prob[\ts>t] = 0)
+\\
+\di_x \G(t,\xmin) = 0  & \quad (\textrm{due to reflecting boundary for X})
+\\
+\G(\T, y) = 1. & \quad (\ts \textrm{ is definitely} > \t \textrm{ for } X(\T)
+\in (\xmin, \xth) )
 \end{cases}
@@
 If we solve the PDE for $\G$ over $t \in [0,\infty)$, then we can calculate
-$\Ttwo(x_0)$ as:
+$\Ttwo(x)$ as:
 \begin{eqnarray}
-\Ttwo(x_0) &=& \int_0^\infty t^2 \cdot (-\di_t\G(t,x_0)) \intd{t}
-\\
-		   &=& \int_0^\infty 2t \cdot \G(t,x_0) \intd{t}
+\Ttwo(x) &=& \int_0^\infty t^2 \cdot (-\di_t\G(t,x)) \intd{t}
+\\
+		   &=& \int_0^\infty 2t \cdot \G(t,x) \intd{t},
 \end{eqnarray}
@@
 Since $\Ttwo$ is an integral over time, we can reduce the PDE in
-\cref{eq:backward_SDF} for the distribution to a BVP for the moment(s) by
-integrating over time. Here is how this is done: multiply both sides  of
+\cref{eq:backward_SDF} for the distribution to an ordinary differential equation
+(ODE) for the moments by integrating over time. Multiply both sides of
 \cref{eq:backward_SDF} by $2t$, and integrate with respect to time:
@@
 =
-\int_0^\infty  2t \left[ (\a - \frac y\tc) \di_y\G - \frac {\b^2} 2
+\int_0^\infty  2t \left[ (\m + \amax - \frac y\tc) \di_y\G - \frac {\b^2} 2
 \di_y^2 \G\right]
-\intd{t}
-\end{equation}
-The LHS turns out to be $2\Tone$, so
+\intd{t}.
+\end{equation}
+The left-hand side turns out to be $2\Tone$, so
 \begin{equation}  
 2\Tone
-=(\a - \frac y\tc)   \di_y \Ttwo
+=(\m + \amax - \frac y\tc)   \di_y \Ttwo
 - \frac {\b^2} 2
-\di_y^2 \Ttwo
+\di_y^2 \Ttwo.
 \label{eq:BVP_Ttwo}
@@
 1
-=(\a - \frac y\tc)   \di_y \Tone
+=(\m + \amax - \frac y\tc)   \di_y \Tone
 - \frac {\b^2} 2
-\di_y^2 \Tone
+\di_y^2 \Tone.
 \label{eq:BVP_Tone}
 \end{equation}
-In both cases, the BCs will be:
+In both cases, the boundary condtions will be:
 \begin{equation}
 \begin{cases}
-\Ti(\xth) = .0 & \quad (\textrm{we are already spking})
-\\
-\di_x \Ti(\xmin) = .0  & \quad (\textrm{due to reflecting boundary for }X)
+\Ti(\xth) = 0 & \quad (\textrm{we are already spking})
 \end{cases}
@@
 \ref{eq:BVP_Tone} and \ref{eq:BVP_Ttwo} directly via an ODE solver, starting
-from $T_i(\xth) = .0$ and integrating down for $x  \in [\xmin, \xth]$.
+from $T_i(\xth) = 0$ and integrating down for $x  \in [\xmin, \xth]$.
 That is how we have obtained the terminal conditions for both $\v(x,\T)$ and
-$p(x,\T)$ in our calculations above.
+$p(x,\T)$ to apply in, respectively, \cref{eq:OC_LS_HJB_full} and
+\cref{eq:adjoint_pde_OU}.
 

### 2014-01-22 ref_reply.tex
Adding Refreply.tex

\documentclass{article}
\usepackage{amsmath}
\usepackage{amscd}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsthm} 
%\usepackage{circuitikz}
% \usepackage{pgf}
% \usepackage{tikz}
% \usetikzlibrary{arrows,snakes,backgrounds}
% \usepackage{verbatim} 
% \usetikz
% \usepackage{subfig}
% \usepackage{graphicx}
\newtheorem*{rep}{Reply}
% \usepackage[super]{nth}
% \usepackage{appendix}
% \usepackage{listings}
% \usepackage{color}

% \usepackage{hyperref}
%\usepackage{url}

%\usepackage{cleveref}
% \usepackage{aviolov_style}
% \usepackage{local_style}

\begin{document}

\title{Stochastic Optimal Control of Single Neuron Spike Trains\\
--\\
Reply to Referees at the JNE}

\author{Alexandre Iolov, 
Susanne Ditlevsen 
and
Andr\'e Longtin
}

\date{\today}

\maketitle

%\maketitle

%\tableofcontents

\vskip 20pt

We want to thank the 4 referees for the thoughtful and concrete points made on
our submission. We will attempt to answer all their concerns in a point-by-point
fashion. Below, we list the referees comments and then our replies immediately
following each comment with '\begin{rep} TODO: \end{rep}' to indicate the reply.

\tableofcontents

\clearpage
\section{Referee: 1}
This is a nice study in which the authors study optimal control of spike timing in the presence of noise. I only have a few comments which the authors should consider:

\begin{enumerate}
\item the authors already compare their methodology with that of other in the
introduction, these statements should be moved to the discussion.
\begin{rep}

\end{rep}

\item can the authors quantify the performance of their algorithm (e.g. \% of
correct spikes) as a function of model parameters (e.g. noise intensity) in
different regimes. I believe that these will be more informative to the reader.
\begin{rep}

\end{rep}
\end{enumerate}

\clearpage
\section{Referee: 2}

In this manuscript, the problem of controlling the spike firing times in response to stimulating waveforms was considered in the closed- and open-loop control schemes. In particular, the authors have proposed a procedure for determining the stimulating waveform in the open-loop situation which is useful for clinical situations. The reviewer thinks that the results of the manuscript are interesting and the manuscript would be well organized. However, some issues need to be clarified before recommending the manuscript for publication in Journal of Neural Engineering.

\subsection{Major issue}
Many readers would wonder if it would be appropriate to determine the
stimulating waveform on the basis of a criteria of the inter-spike intervals, as
the decision should be made from the knowledge on the all spike times,
i.e.,“spike trains”. The error of the spike firing times determined on the basis
of only ISIs seems to be accumulated as time goes. The authors have not cited
anything about how to determine the succeeding spike time interval for the
multi-spike control in the discussion section or anywhere else. The authors
should have clearly stated whether or not the spike time error could be canceled
by the succeeding spike time after the inaccurate preceding spike time given, in
the multi-spike control.
\begin{rep}

\end{rep}

\subsection{Some other issues}
\begin{enumerate}
  \item 
--In l.6, Page 4 of 33, the authors stated “a non-neglegible noise component
...”. What kind of noise would the authors assume? Is that a weak or strong one?
Is that Gaussian or not? Is that white or colored one?
\begin{rep}
It is clear taht the referee is confused. By non-negligible noise component, we
mean a value of $\b >> 0 $.
\end{rep}

\item In 3.1 HJB equation subsection, the authors mentioned repeatedly the HJB 3
times in Eqs. (8), (10), and (13). It would be good enough to specify Eq.(13) so
as to simplify the problem under consideration.
\begin{rep}
TODO:
\end{rep}

\item In Page 11 of 33, Figure 2 shows w(x,t) and alpha(x,t) at t=0, t*/2, or t*.
However, the left column of (A, C, E, and G) looks redundant as Figure 3 shows
w(x,t) as functions of x and t. Many readers might want to look at the shape of
alpha(x,t) as functions of x and t, instead of the right column of (B, D, F, and
H) in Figure 2.
\begin{rep}
We can replace the surface plots for the value function with surface plots for
the control instead?
\end{rep}

\item In Simulations section, Page 18 of 33, the results of the naive control seem
unnecessary.
\begin{rep}
Both Susanne and Alex disagree. The point of the naive control as far as Alex is
concerned is to provide a base-case against which to compare the 'optimal'
stimulus. If the optimal stimulus improves things by 5,10\% then it is  hard to
justify its computational cost. 
\end{rep}

\item In the right column of Figure 6 in Page 22 of 33, 3 sample realizations of
x(t) and alpha(t) at epsilon=0.1 are depicted, but any explanation about those
graphs could not be found in section 6 or anywhere else, although the authors
have stated in the legend of Fig.6 “... see section 6)”.
\begin{rep}

\end{rep}

\item In l. 7 in Page 23 of 33, “High Noise” should be deleted.
\begin{rep}
TODO:
\end{rep}

\item In the fourth paragraph in Page 23 of 33, the authors mentioned “In fig. 11,
we relax the bound constrains on the control $[\amin, \amax]$ from
[-2;2] to [-4;4].”. However they depicted the firing rates and dot raster plots only in
the case of the closed-loop control in Figure 11. They might want to show the
simulation results of the open-loop control scheme in relaxing the constraints
on the alpha's boundaries. Also, they might want to depict the sample
realizations of alpha(t) or the mean of alpha(t) as a function of time as well
as the firing rates and dot raster plots in the closed- and open-control
schemes.
\begin{rep}
TODO:
\end{rep}

\item In 3-4 lines before the section 6, “For example, in figs. 6 and 7” should be
replaced by “For example, in figs 7 and 8”.
\begin{rep}
TODO:
\end{rep}

\item In the last sentence of 5.2 Multi-Spike Control subsection in Page 23 of 33,
the authors stated “if the target interspike intervals are within reasonable
reach of the neuron, this is not a problem.” However, many readers would wonder
how much reasonable reach could be, since they might think of if the reach would
be longer or shorter than the refractory period.
\begin{rep}
TODO:
\end{rep}
\item In Page 28 of 33. the left column of Figure 12 should be deleted because
of a duplication of the right column of Figure 2. Instead, the right column of Figure
12, alpha(x,t), should be depicted explicitly in 3 dim as functions of x and t.
\begin{rep}
TODO:
\end{rep}

\item In the left column of Figure 13 in Page 29 of 33, the traces were same as
those of Figure 4. The authors might want to refine how to show us the effect of
energy penalty.
\begin{rep}
The two traces are the same! They are added here to compare against
the traces on the right, so that people don't flip back and forth for figures to
appreciate what has changed
\end{rep}

\item In the fourth paragraph of discussion section in Page 30 of 33, the
authors stated “the control scheme applies in the same manner in the Sub-Threshold
regime as well”. However, it would not be obvious for many readers as in the
sub-threshold regime, nobody knows if noise-enhanced stimuli elicit spikes or if
noise itself elicits spikes.
\begin{rep}
TODO: I don't understand the referee. Susanne pointed out that we need to
clarify that in all cases, the maximum value of the control puts the model in
the Supra-Threshold regime. That is using the control one can always have
Supra-Threshold behaviour. The Supra / Sub- distinction only applies to the
intrinsic (uncontrolled dynamics)
\end{rep}

\end{enumerate}

\clearpage
\section{Referee: 3}
The authors have considered two optimal control strategies on a single neuron
level using a noisy leaky integrate-and-fire model: Closed-loop control where
they use the Dynamic Programming technique, and Open-loop control where they use
a variation of Pontryagin's  Maximum Principle. They mention, as an advantage of
their work to others, that their control method is not limited by the magnitude
of the noise present in the model. They also consider the effect of energy
penalty and touch on the robustness of their algorithm. They have very clearly
demonstrated a background of recent literature and pointed out how the
mathematical formulation of their work is different.

This is a well written paper and, in my opinion, a nice extension to the
previous literature. I do however have some comments and questions that should
be addressed before the manuscript could be considered for publication in the
Journal of Neural Engineering (JNE).

\subsection{Major comments/questions:}
1. The most important shortcoming of this manuscript is the lack of adequate
discussion in applicability of the methods presented in a practical situation,
as would be most interesting for the majority of this journal's readers. In the
JNE, we are always looking for new ideas that are applicable in practice. The
authors need to discuss how these methods can be applied to a neuron, what are
the potential limitations (perhaps the choice of model), and how can one find a
way around those limitations. The authors allude to this in the discussion
section, but I think this needs a more elaborate discussion with possible
suggestions.
\begin{rep}
TODO:
\end{rep}

2. Related to 1 above, the authors mention several highly sophisticated
network-level dependent interactions/diseases, such as brain-machine interface,
neural prostheses, the Parkinson's disease, and epilepsy, as potential
beneficiaries of their method. Given the fact the network-level activity in
neurons are much different than single neurons, and that the focus of the
current manuscript is on single neuron behaviour, I strongly suggest that the
authors revise the application sections of their paper and include other
physiological applications that are more relevant.
\begin{rep}
TODO:
\end{rep}
3. The authors need to acknowledge the difference between the Maximum Principle
as an optimal control theory that they have used and Dynamic Programming as an
efficient numerical method for finding the global minimum for an optimal control
problem. The way the manuscript is written, suggests that the authors use the
Maximum Principle only for the open-loop control. However, in the closed-loop
control problem, the fact that the control is bounded, means that one needs to
use the Maximum Principle to find the optimal control, just as the authors have
in (9). The Maximum Principle yields the control (9) and Dynamic Programming is
just a numerical method for solving for the optimal trajectories.
\begin{rep}
TODO:
\end{rep}

4. How can these methods extend to networks of neurons efficiently? For
instance, Dynamic Programming suffers from the "curse of dimensionality". If one
were to make a coupled network of integrate-and-fire neurons, perhaps in an
effort to model network-level pathology, how can this network be solved
numerically in a time-efficient manner?
\begin{rep}
TODO:
\end{rep}

5. When epsilon is small, some of the controls that the authors have computed
are very similar to a bang-bang type control and even seem to have a net area
under curve of close to zero. I think it will help the comprehensiveness of the
paper if the authors made a somewhat elaborate connection between their
formulism to that of a bang-bang control formulism. And also explain why these
controls come out almost balanced in negative and positive values.
\begin{rep}
TODO:
\end{rep}

\subsection{Minor comments/questions:}

Some of the control/mathematical jargon that is used in this paper could be
defined for the readers of JNE's better understanding. Examples of these include
"measurable with respect to the filtration of X", "transition density",
"artificial reflecting boundaries", "absorbing boundaries" 
\begin{rep}
TODO:
\end{rep}
The bounds on control should also appear in the minimization in (7)
\begin{rep}
TODO:
\end{rep}

P7,L44: Why do the authors assume for $x<<0$, w is not significantly affected?
What do they mean by $x<<0$? Looking at Figure 1, min(x) is usually around zero,
unless there is high noise, in which case it becomes in the order of -1 to -10 probably. 
Please elaborate a little.
\begin{rep}
TODO:
\end{rep}
P10, Caption of Figure 2: "Note that the green and blue curves in F) are lying on top of each other for all x." This is not true. Did the authors mean subplot (D)? And can you explain why it is that the green and blue curves are on top of each other?
\begin{rep}
TODO:
\end{rep}
The initial condition for the computations seems to be the point of previous spike, x=0. Is that right? Please make note of this where ever appropriate for clarity, for example, when showing Figure 4.
\begin{rep}
TODO:
\end{rep}
Can the authors elaborate on the noise that they implemented in their simulations, and discuss the shape of the distributions in Figure 5, especially, for the deterministic case?
\begin{rep}
TODO:
\end{rep}
Figure 7: What does the solid horizontal line in the lower plot represent?
\begin{rep}
TODO:
\end{rep}
"i.e." usually has commas before and after. Also, at the beginning of sentence, please consider using the words "That is" instead of "I.e." (P12,L19).
For consistency, please consider using either Supra- or Super-Threshold throughout.
\begin{rep}
TODO:
\end{rep}

Typos and trivial mistakes: \\
P3,L06: setting $->$ settings\\
P4,L25: means $->$ is\\
P8,L22: optimisation $->$ optimization\\
P9,L14: be equal to $->$ equals\\
P10,L46: and on the optimal $->$ and the optimal\\
P15,L51: Sub-Threshold-Low-Noise regimes $- >$ Supra-Threshold-Low-Noise regimes
?
\begin{rep}
TODO:
\end{rep}

\clearpage
\section{Referee: 4}
In this manuscript, Iolov and colleagues develop and
test optimal algorithms for imposing predetermined spike patterns on simple
neural models.  The manuscript is generally clear and describes well the
advances over past methods.  However, I am left with the impression that this
first paper may be better suited for an applied math journal, with follow-up
work to be published in JNE.  Alternatively, the authors may wish to include
more applications-focused work in this manuscript.  This step would create a
paper that is better matched to JNE, at the possible cost of creating a
manuscript of unwieldy length.

\subsection{General points}
1. I am not convinced that this manuscript as presented is
a good match for JNE.  Certainly, stochastic optimal control is important for
neural devices, and the math is not beyond a significant fraction of JNE’s
readership.  The potential mismatch lies in the rhetoric, the assumptions, and
the examples given. 
\\1a. Rhetorically, the paper reads more like an applied
math paper than a neural engineering paper.  The authors are encouraged to place more
emphasis on the potential for application, especially in the results and
discussion sections. 
\begin{rep} TODO: \end{rep}

1b. For JNE, the authors should place more emphasis on robustness, which is
considered only very briefly in section 7, with no figures. In any application,
the experimenter is going to know very little about the underlying model, and
will have to build algorithms that work around this problem.

\begin{rep} TODO: \end{rep}

1c. On a related front, the authors assume intracellular access  and
use highly oversimplified integrate-and-fire models.  It would be interesting to
test the effectiveness of the authors’ approach with more realistic models and
with explicitly extracellular forms of stimulation.

\begin{rep} TODO: \end{rep}

2. The results and discussion sections could be improved by more context, to
connect specific findings to broader conclusions from the abstract and
introduction.

\begin{rep} TODO: \end{rep}

\subsection{Specific points}

Abstract: The authors state that their goal was to design a control scheme to
achieve a target spike train with minimal damage to tissue.  I don’t think this
goal was achieved, because of the level of abstraction of the work.

\begin{rep} TODO: \end{rep}

p. 2, first par.  The goal mentioned by the authors of control via extracellular
stimulation does not seem to be achieved by the following work.  Also, the
examples noted by the authors (BMIs and prosthetics) are typically thought of as
problems for which one needs to interpret spike trains, not impose them.

\begin{rep} TODO: \end{rep}

p. 2, par. 3.  The sentence beginning “However the efficiency of the numerics …”
is nearly impossible to parse and understand.

\begin{rep} TODO: \end{rep}

p. 5, section 3.  The sentence beginning “Given a time t …” ends very awkwardly and should be rewritten.

\begin{rep} TODO: \end{rep}

p. 16.  I suggest putting Algorithm 1 in an appendix.

\begin{rep} TODO: \end{rep}

Figure legends, especially for Figs. 7-10, begin awkwardly (e.g., “How good the
closed-loop controller is at hitting a target spike train.”)  I suggest
beginning the figure legends with declarative statements about the quality of
the outcome for open- and closed-loop algorithms in the low- and high-noise
cases.

\begin{rep} TODO: \end{rep}

% \bibliographystyle{plain}
% \bibliography{library}
\end{document}

### 2014-02-18 MLnote.tex
Adding ML Note

\documentclass{article}
\usepackage{amsmath}
\usepackage{amscd}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsthm} 
%\usepackage{circuitikz}
% \usepackage{pgf}
% \usepackage{tikz} 
% \usetikzlibrary{arrows,snakes,backgrounds}
% \usetikz
% \usepackage{verbatim} 
\usepackage{subfig} 
\usepackage{graphicx}
\newtheorem*{rep}{Reply}
% \usepackage[super]{nth}
% \usepackage{appendix}
% \usepackage{listings}
% \usepackage{color}

\usepackage{hyperref}
\usepackage{url}

\usepackage{cleveref}
\usepackage{aviolov_style}
\usepackage{local_style}

\begin{document}

\title{Spike-Controlling the Noisy ML Model}

\author{Alexandre Iolov, 
Susanne Ditlevsen 
and
Andr\'e Longtin
}
\date{\today}

\maketitle

\tableofcontents

\section{Stochastic Morris-Lecar Model}
\subsection{Diffusion model}
\def \Vt {{ U_s }} \def \Wt {{ W_s }} \def \Vz {{ U_z}} \def \Wz {{ W_z}} The
stochastic Morris-Lecar model including both current and channel noise is
defined as the solution to
\begin{equation}
\left\{
\begin{array}{ccl}
d\Vt &=& f(\Vt,\Wt)dt +\gamma d\tilde{B}_s,\\
d\Wt&=&b(\Vt,\Wt) dt  \, + \sigma(\Vt,\Wt)dB_s,
\end{array}
\right.
\tag{ML}
\label{eq:ML}
\end{equation}
where 
\begin{eqnarray*}
f(\Vt,\Wt)&=&\frac{1}{C}\Big(-g_{Ca}m_\infty(\Vt) (\Vt-V_{C_a})
\\ && - g_K\Wt(\Vt-V_K)
-g_L(\Vt-V_L)+I + A(s) \Big),
\\ b(\Vt,\Wt)&=&
\left(\alpha(\Vt)(1-\Wt) - \beta(\Vt)\Wt\right),\\ m_\infty(v)&=&\frac{1}{2}\left(1+\tanh\left(\frac{v-V_1}{V_2}\right)\right),\\
\alpha(v) &=& \frac{1}{2}\phi \cosh\left(\frac{v-V_3}{2V_4}\right)\left(1+\tanh\left(\frac{v-V_3}{V_4}\right)\right),\\
\beta(v) &=& \frac{1}{2}\phi \cosh\left(\frac{v-V_3}{2V_4}\right)\left(1-\tanh\left(\frac{v-V_3}{V_4}\right)\right),
\end{eqnarray*}

and the initial condition $(\Vz, \Wz)$ is random  with density $p(\Vz, \Wz)$.
 Processes $(\tilde{B}_s)_{s\geq s_0}$ and $(B_s)_{s\geq s_0}$ are independent
 Brownian motions.
The variable $\Vt$ represents the membrane potential of the neuron at time $s$,
and $\Wt$ represents the normalized conductance of the K$^+$ current. It varies
between 0 and 1, and can be interpreted as the probability that a K$^+$ ion
channel is open at time $t$. The equation for $f(\cdot)$ describing the dynamics
of $V_t$ contains four terms, corresponding to Ca$^{2+}$ current, K$^+$ current,
a general leak current, the input current $I$ AND THE EXTERNAL CONTROL CURRENT
$A(s)$!

 The functions $\alpha
  (\cdot )$ and $\beta (\cdot )$ model the rates of opening and closing of the
  K$^+$ ion channels. The function $m_{\infty}(\cdot)$ represents the
  equilibrium value of the normalized Ca$^{2+}$ conductance for a given value of
  the membrane potential. The parameters $V_1, V_2, V_3$ and $V_4$ are scaling
  parameters; $g_{Ca}, g_K$ and $g_L$ are conductances associated with
  Ca$^{2+}$, K$^{+}$ and leak currents; $V_{Ca}, V_K$ and $V_L$ are reversal
  potentials for Ca$^{2+}$, K$^+$ and leak currents; $C$ is the membrane
  capacitance; $\phi$ is a rate scaling parameter for the opening and closing of
  the K$^+$ ion channels; and $I$ is the input current.

\begin{center}
\begin{figure}
\subfloat[Type 1]{
\includegraphics[width=.5\textwidth]{Figs/MLBox/ML_type1_example_traj.pdf}
\label{fig:simu_ML_1} 
}
\subfloat[Type 2]{
\includegraphics[width=.5\textwidth]{Figs/MLBox/ML_type2_example_traj.pdf}
\label{fig:simu_ML_2}
}

\caption{Simulated trajectory of the stochastic 
  Morris-Lecar model: $(\Vt)$ as a function of time (top),
  $(\Wt)$ as a function of time (bottom). Parameters are given in Section
  \ref{sec:simulation}. Time is measured in ms, voltage in mV,
  the conductance is normalized between 0 and 1. }
 \label{fig:simu_ML} 
\end{figure}
\end{center}

Various noise sources are present in single neurons, and they act on many
different spatial and temporal scales \citep{Gerstner2005,Longtin2013}. A main
component arises from the synaptic bombardment from other neurons in the
network, and in the diffusion limit appears as an additive noise on the current
equation. Parameter $\gamma$ scales this current noise. Conductance fluctuations
caused by random opening and closing of ion channels leads to multiplicative
noise on the conductance equation. Function $\sigma(\Vt,\Wt)$ models this
channel or conductance noise. We consider the following function that ensures
that $\Wt$ stays bounded in the unit interval  if $\sigma \leq 1$
\citep{DitlevsenGreenwood2011}: $$\sigma(\Vt, \Wt ) = \sigma \sqrt{2
\frac{\alpha(\Vt) \beta(\Vt)}{\alpha(\Vt)+ \beta(\Vt)} \Wt (1-\Wt)}. $$
A strajectory of the model is simulated in Fig. \ref{fig:simu_ML}. The peaks of
$(\Vt)$ correspond to spikes of the neuron.

\subsection{Simulation study}\label{sec:simulation}
Parameter values of the Morris-Lecar model used in the simulations are the same
as those of \citep{RinzelErmentrout1989,Tateno2004} for a class I and for a
class II membrane.

The values are:\\ 
% \begin{verbatim}
\input{Figs/MLBox/MLParams_type1.txt}
% \end{verbatim}
AND \\
% \begin{verbatim}
\input{Figs/MLBox/MLParams_type2.txt}
% \end{verbatim}

% $V_K = -84$ mV, $V_L = -60$ mV, $V_{Ca} = 120$ mV, $C = 1\,
% \mu$F/cm$^2$, $g_L = 0.1\, \mu$S/cm$^2$, $g_{Ca} = 0.22 \,
% \mu$S/cm$^2$, $g_K = 0.4\, \mu$S/cm$^2$, $V_1 = -1.2$ mV, $V_2 = 18$ mV,
% $V_3 = 2$ mV, $V_4 = 30$ mV, $\phi = 0.04$ $\mbox{ms}^{-1}$.  
% Input is chosen to be $I = 90 \, \mu$A/cm$^2$.

(The values are from Susanne's code, in Andre's code the values are normalized
st $C= 1$ and there is no noise, whereas in Susanne's code the parameter values
for $\s, \g$ are given and so I don't have to rederive them)

Initial conditions for the Morris-Lecar model are $V_{t_0} = -26$ mV, $W_{t_0} =
0.2$.

Trajectories are simulated with time step $\Delta
t=0.01$ ms.  An example of a simulated trajectory is
given in Figure \ref{fig:simu_ML}.

IMPORTANT: From now on we will use data generated from a TYPE I (One) Model, as
it is deemed more suitable to be well-fitted by an LIF model.

\section{Morris Lecar $<->$ LIF conversion}
We consider how to go about converting $V,W,s$ data from the Morris-Lecar model
into $Y,s$ data from the (dimensional) LIF model,

\begin{equation}
\begin{gathered}
dY(s) = \frac 1C \left( m + A(s) - Y  \right) \intd{s} + \s \intd{B_s},
\\
Y(\ss) = \yth \implies
\begin{cases}
Y(\ss^+) = \yreset & 
\end{cases}
\end{gathered}
\tag{OU:dim}
\label{eq:LIF_dimenisonal}
\end{equation}
Which in turn can be transformed into the non-dimensional LIF model, which is
used to determine the control algorithm:
\begin{equation}
\begin{gathered}
dX(t) = \left(\m + \a(t) - \frac{X(t)}{\tc} \right) \intd{t} + \b \intd{B_t},
\\
X(\ts) = 1 \implies
\begin{cases}
X(\ts^+) = 0 & 
\end{cases}
\end{gathered}
\tag{OU:non-dim}
\label{eq:LIF_non_dimenisonal}
\end{equation}
We use the same symbol $C$ in both \cref{eq:ML} and \cref{eq:LIF_dimenisonal}
because the two models never really 'talk' to each other through their parameters. The
two models only see each other through their data, 
In particular given data from \cref{eq:ML} we estimate the
parameters $C, m,\s, \yth, \yreset$ in \cref{eq:LIF_dimenisonal}. But we never
estimate (know) the parameter $C$ in the Morris-Lecar model, in fact as far as
we are concerned, the M-L data comes from a Black Box.

We will need to know how to perform the following tasks:
\begin{enumerate}
  \item Given $V,s$ how to estimate $m, Cc, \s$ (In general estimating
  Ornstein-Uhlenbeck parameters is fairly straightforward, but only given OU
  data)

  \item determine the physical threshold and reset $\yth, \yreset$

  \item Given an optimal control $\a(x,t)$ determined in the $X,t$ domain,
  convert to $A(Y, s)$ in the $Y,s$ domain. (And then apply $A$ to $V$ in the
  ML model).
\end{enumerate}

\subsection{Converting the OU model from dimensional to non-dimensional form
and back}
Let's start with the translation between $Y,s$ and $X,t$. First we
normalize $X$ to reset to 0 and spike at $1$. With that we have that $X_t =
\frac{Y_s - \yreset}{\yth - \yreset} $. We also non-dimensionalize time, using $\meanI$,
which is the mean observed ISI. (Alternatively, we could normalize by the mean
ISI of the target train. We'll use the same symbol in any case) $$ t = \frac
s\meanI$$

Let us write, $G$ for gap.
$$G = \yth - \yreset$$
b/c the expression $\yth - \yreset$  occurs quite often.

Now we can do the conversions:
\begin{eqnarray*}
\Delta Y =&
	\frac 1 C (m + A - Y) \Delta s +
 	\s \sqrt{\Delta s} \cdot \xi \\
\frac {\Delta Y}{G} =&
	\frac {\meanI}{GC} (m - \yreset + A - (Y-\yreset))\frac{\Delta s}{\meanI} +
	\frac{\s}{G} \cdot \sqrt{\meanI} \sqrt{\frac{\Delta s}{\meanI}} \cdot \xi
\\
\Delta X  = &
	(\frac{\meanI (m - \yreset)}{GC} + \frac {\meanI\cdot A}{GC} -
	\frac{\meanI}{C} X) \Delta t +  
	\frac{\s}{G} \cdot \sqrt{\meanI} \sqrt{\Delta t} \cdot \xi
\end{eqnarray*}
From this we can read off, $\m, \tc, \b, \a$. We have:
\begin{subequations}
\begin{eqnarray}
X_t = \frac{Y_s - \yreset}{\yth - \yreset} 
& \iff  &
 Y_s =  \ldots
\\ 
t = \frac s \meanI &
\iff & s = \meanI t
\\
\m = \frac{\meanI (m - \yreset)}{C\cdot G}
& \iff &
m = \ldots   
\\
\a() = \frac{\meanI \cdot A()}{GC}
&\iff&
A() = \frac{GC \cdot \alpha() }{\meanI}
\\
\tc = \frac{ C }{\meanI }
&\iff&
C = \ldots
\\
\b = \frac{\s \sqrt{\meanI} }{\yth - \yreset}
&\iff&
\s =\ldots
\end{eqnarray}
\end{subequations}

We've used \ldots to indicate conversions that will never be made, so they do
not matter. (We will never go $\mu \ra m$).

To indicate that we've done everything correct we show two trajectories $Y_s,
X_t$ in \cref{fig:OU_dim2nondim_conversions}.
%\usepackage{graphics} is needed for \includegraphics
\begin{figure}[htp]
\begin{center}
  \includegraphics[width=\textwidth]{Figs/MLBox/OUDim2NonDimConversion.pdf}
  \caption[labelInTOC]{Conversions between $Y_s\ra X_t$. 
  Top plot is a chunk of $Y$ data. Below is a chunk of $X$ data using the conversion params.
  Also below we plot the original $Y$ chunk from above converted to $X$.
  Obviously we've used the same stream of random numbers. The parameters work
  out to be:
  meanISI = 189.11,\\
$m = -16.91 C = 132.72 \sigma = 1.00$ \\ 
$\mu = 1.39 \tc = 0.70 \beta = 0.40 $.\\ 
The physical parameters, $m,C,\sigma $ are obtained from
\cref{sec:estimaing_LIF_from_ML}}
\label{fig:OU_dim2nondim_conversions}
\end{center}
\end{figure}

\subsection{Estimating the LIF params from ML data}
\label{sec:estimaing_LIF_from_ML}
This estimation and in general the relation between $Y$ and $V$ is the most
challenging aspect of the whole problem!

It is difficult to unambiguously determine what if anything is the threshold,
the point-of-no-return for $V$ in \cref{fig:simu_ML}. But it looks to be about
-14.0. In addition we stipulate a lower threshold, at -22, which is the point
after which we no longer consider the data as suitable to be represented as an
OU process. Thus rises the question of what to do between (-22 and -14). We will
say that this is a grey zone where the diffusion has almost spiked: $X = 1-
\epsilon$, say $X = .999$

Let's plow ahead! \Cref{fig:ML_estimates_data} shows the original $V,s$ data,
plus an identification of the spikes, using a completely arbitrarily chosen
threshold, $\yth = -14$. A spike is recorded when $V$ (equivalently $Y$) crosses
$\yth$.

Once a spike is recorded, we wait an absolute refractory time of $$t_{ref} =
40 ms$$ at which point we restart. As such the time between $\ts$ and
$\ts+t_{ref}$ is compactified to a point as far as the LIF is concerned. I've
played around with different values of $t_{ref}$ and $40 ms$ seems to give the
best results. In particular notice on the bottom plot in
\cref{fig:ML_estimates_data} how all trajectories converge more or less to the
same point. Visually, this uniform reset effect is achieved for any $t_{ref} \in
[37, 43]ms$.

The data between the reset time and the OU bound crossing time is used for
parameter estimation. (It is plotted in the middle plot in \cref{fig:ML_estimates_data})

\begin{figure}[htp]
\begin{center}
  \includegraphics[width=\textwidth]{Figs/MLBox/ML_estimates_annotated.pdf} 
  \caption[labelInTOC]{Annotating (spike-segmenting) the Raw ML data. 
  This is the same data as in \cref{fig:simu_ML_1} (redrawn in top plot).
  The middle plot shows the data segments that are used to estimate the OU
  parameters from ML data.
  The bottom plot shows the stereotypical spike shape and justifies why 40 ms
  is a reasonable refractory time}
  \label{fig:ML_estimates_data} 
\end{center}
\end{figure}

Now we want to estimate $m, C, \s$ from a chunk of Morris-Lecar generated data.

Let's recall the MaxLikelihood formulas for the parameters of the non-threshold,
$A\equiv 0$ OU process in \cref{eq:LIF_dimenisonal} given discrete
observations, $\{Y_n\}_{n=0}^{N}$:
\begin{subequations}
\begin{eqnarray}
\hat{ m} &=&  \frac{ \sum_{n=1}^{N} \left( Y_n - e^{-\frac {\Delta} {\hat C}}
Y_{n-1} \right)} {N ( 1-e^{-\frac {\Delta} {\hat C}})}
\\
 0 &=& e^{-\frac {\Delta}{\hat{C}} } -
  \frac { \sum_{n=1}^{N} ( Y_n - \hat m)( Y_{n-1} - \hat m) }
  	   { \sum_{n=1}^{N} \left( Y_{n-1} - \hat m \right)^2 } 
\\
\hat\s^2 &=& 
2\frac{ \sum_{n=2}^{N}  \left( Y_n - \hat m - (Y_{n-1} -
\hat m) e^{-\frac {\Delta} {\hat C}} \right)^2 } 
	  {N \cdot (1-e^{-2\frac {\Delta} {\hat C}}) \hat C}
\end{eqnarray}
\end{subequations}
This is for one uninterupted sample. However, we have $K$ distinct paths
each consisting of $N_k+1$ data points ($N_k$ transitions). Thus we must
generalize the formulas:
\begin{eqnarray*}
\hat{ m} &=& 
\frac{ \sum_{k=1}^K \sum_{n=1}^{N_k} \left( Y^{(k)}_n - e^{-\frac{\Delta_n} {\hat C}} Y^{(k)}_{n-1} \right)} 
	 { N( 1-e^{-\frac {\Delta_n} {\hat C})}}
\\
0 &=& e^{-K\frac {\Delta}{\hat{C}} } - 
\prod_{k=1}^K\left[ \frac { \sum_{n=1}^{N_k} ( Y^{(k)}_n - \hat m)(Y^{(k)}_{n-1} - \hat m) }
						  { \sum_{n=1}^{N_k} \left( Y^{(k)}_{n-1} - \hat m \right)^2 } 
  	    \right]
\\
\hat\s^2 &=&  
2\frac{ \sum_{k=1}^K \sum_{n=2}^{N}  \left( Y^{(k)}_n - \hat m - (Y^{(k)}_{n-1} -
\hat m) e^{-\frac {\Delta} {\hat C}} \right)^2 } 
	  { N (1-e^{-2\frac {\Delta} {\hat C}}) \hat C}
\end{eqnarray*}
Where $N= \sum_k N_k$

Now we are ready to estimate $m, C, \s$ given data from the ML trajectories.
Given a stream of ML-Type 1 data points, we get the estimates in
\cref{tab:OU_from_ML1_ests}.
\begin{table}
\begin{tabular}{l|cc}
\input{Figs/MLBox/OU_from_ML1_estimates.txt}
\end{tabular}
\caption{Estimates for the three OU parameters $m, C, \s$ given data from an ML
type 1 model (middle-column) and then from an OU model ran with the estimated
parameters of the ML 1 model (right column)}
\label{tab:OU_from_ML1_ests}
\end{table}
Then, if we run the OU model with the inferred parameters from
\cref{tab:OU_from_ML1_ests} (the ML1 column), we get the OU trajectories in
\cref{fig:OU_vs_ML1}. Finally, recalling the conversions of $Y$ to $X$ to
non-dimensionalize the OU process gives us the values for $\m,\tc, \s$ in
\cref{tab:OU_nondim_vs_OU_dim}. This is a very interesting parameter regime. 
It is sub-threshold, $\bar X \approx .66$, but the noise strength is halfway
between what we called Low-Noise $\b=.3$ and High-Noise $\b=1.5$. We know our
control is very good

In \cref{tab:OU_nondim_vs_OU_dim}, we can also consider the bounds on our
control. We've assumed that $$A(s) \in [\pm 20]$$. We see that this gives
us quite wide bounds on $\a(\cdot)$. Recall that so far we have used $\a \in
[-2,2]$ and we've only needed more 'oomph' in the case of high-noise, i.e.\
$\b=1.5$, whereas here we have the milder $\b=0.7$. Also the time-constant is
pretty small, meaning, we have a pretty stiff system, i.e.\ one which quickly
mean-reverts to the base-line $0$.

\begin{figure}[htp]
\begin{center}
  \includegraphics[width=\textwidth]{Figs/MLBox/OU_from_ML_type1_Tf2000.pdf}
  \caption[labelInTOC]{The original Type 1 ML sub-threshold dynamics vs.\ OU
  dynamics. Notice that the OU dynamics are generated using the exact same
  $Y_0$ and the SAME stream of random numbers as those that generate the ML1
  dynamics. It is clear that the match (OU of ML1) is satisfactory.} 
  \label{fig:OU_vs_ML1}
\end{center} 
\end{figure}

\begin{table}
\begin{align*}
\input{Figs/MLBox/OUDims2Nondims.txt}
\end{align*} 
\caption{Estimates for the three OU parameters $m, C, \s$ and their 
non-dimensional counterparts, $\mu , \tc ,\beta$} 
\label{tab:OU_nondim_vs_OU_dim}
\end{table}

\clearpage

\section{Controlling the Morris-Lecar model}
\label{sec:controlling_ML}
We now have all the pieces to control the Morris-Lecar model. Here is a summary:
\begin{enumerate}
  \item Before doing anything estimate the LIF params, $m, C, \s$ (we've
  already done this) and their conversions to non-dimensional form $\mu , \tc
  ,\beta$
  \item Take a series of target spike times, we will use the spikes generated by
  the model in 
  \item Given $V_s$, treat it like $Y_s$ and convert, $Y_s$ into $X_t$. Convert
  the remaining time-to-spike from $s \ra t$. 
  \item determine the optimal $\a(t, X_t)$ from closed or open-loop. 
  \item convert $\a()$ into $A(s)$
  \item Apply $A(s)$ to the $dV_s$ system.
  \item Rinse and repeat  
\end{enumerate}

Ok, let's skip to the chase - the results are in
\cref{fig:ML_controled_simulation_cl} and \cref{fig:ML_controled_simulation_ol}.
It is clear that the control strategy (for closed-loop) is highly accurate! This
might have much to do with the fact that we have assumed such wide $A$ bounds.
$$A(s) \in [\pm 20]$$ $$\implies \a(t) \in [\pm 4.7]$$

Should we reduce the power bounds\ldots? In general how should we come up with
the values for $A_{min}, A_{max}$?!

For reference
\cref{fig:ML_controled_simulation_cl_A10,fig:ML_controled_simulation_ol_A10}
give the results for $$A_{min}, A_{max} = [-10,10]$$.

%%%%%%%%%%%% CL %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\subfloat[example controlled trajectory]
{
\label{fig:ML_example_controled_traj}
\includegraphics[width=\textwidth]
{Figs/MLTrainController/MLSim_example_cl.pdf} 
}\\
\subfloat[raster plot for controlled simulations]
{
\label{fig:ML_raster_plot}
\includegraphics[width=\textwidth] 
{Figs/MLTrainController/MLSim_raster_cl.pdf}
}
\caption[labelInTOC]{Example of a controlled ML trajectory and the raster plot
of realized spikes vs.\ controlled spikes, $N=5$ 	(Closed-Loop!)}
\label{fig:ML_controled_simulation_cl}
\end{center} 
\end{figure}
%%%%%%%%%%%% OL %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h] 
\begin{center}
\subfloat[example controlled trajectory]
{
\label{fig:ML_example_controled_traj}
\includegraphics[width=\textwidth]
{Figs/MLTrainController/MLSim_example_ol.pdf} 
}\\
\subfloat[raster plot for controlled simulations]
{
\label{fig:ML_raster_plot}
\includegraphics[width=\textwidth]
{Figs/MLTrainController/MLSim_raster_ol.pdf} 
}
\caption[labelInTOC]{Example of a controlled ML trajectory and the raster plot
of realized spikes vs.\ controlled spikes, $N=4$ 	(Open-Loop!)}
\label{fig:ML_controled_simulation_ol}
\end{center} 
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% REDUCED POWER BOUNDS: %%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%% CL %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\subfloat[example controlled trajectory]
{
\label{fig:ML_example_controled_traj}
\includegraphics[width=\textwidth]
{Figs/MLTrainController/MLSim_example_cl_Amax10.pdf} 
}\\
\subfloat[raster plot for controlled simulations]
{
\label{fig:ML_raster_plot}
\includegraphics[width=\textwidth] 
{Figs/MLTrainController/MLSim_raster_cl_Amax10.pdf}
}
\caption[labelInTOC]{Same as \cref{fig:ML_controled_simulation_cl}
(closed-loop) but with reduced power bounds $A \in [-10,10]$}
\label{fig:ML_controled_simulation_cl_A10} 
\end{center}
\end{figure}
%%%%%%%%%%%% OL %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h] 
\begin{center}
\subfloat[example controlled trajectory]
{
\label{fig:ML_example_controled_traj}
\includegraphics[width=\textwidth]
{Figs/MLTrainController/MLSim_example_ol_Amax10.pdf} 
}\\
\subfloat[raster plot for controlled simulations]
{
\label{fig:ML_raster_plot}
\includegraphics[width=\textwidth]
{Figs/MLTrainController/MLSim_raster_ol_Amax10.pdf} 
}
\caption[labelInTOC]{Same as \cref{fig:ML_controled_simulation_ol}
(open-loop) but with reduced power bounds $A \in [-10,10]$}
\label{fig:ML_controled_simulation_ol_A10}
\end{center}
\end{figure}

\clearpage

\section{Morris Lecar $<->$ LIF conversion - Spikes-Only}
It remains to do one more thing - estimate the LIF parameters from Spike-Data
only.

However, we know that estimating $\tc$ from spikes is problemeatic. To keep
things simple at first. We assume we know $\tc$ (we'll use the estimated value
from the \cref{tab:OU_from_ML1_ests}) and only estimate $\m, \b$, the mean
bias current and the noise intensity.

We use the Ditlevsen-Lansky scheme that utilizes the Fortet eqn to estimate
$\m, \b$ - \cite{Ditlevsen2007}.

In short, first we scale time by $C$ and we introduce a unit-tau-char variable
$Z$
$$
dZ = (a- Z) dt + bdW_t$$
with the relation between $a,b$ being
$$
a = \frac{m - \yreset}{\yth - \yreset} 
	\quad \iff \quad 
		 m = a \cdot (\yth - \yreset) + \yreset
$$
$$
b  = \frac{\sigma \sqrt{C}}{\yth - \yreset} 
	\quad \iff \quad  
		\s = \frac{b \cdot (\yth - \yreset)}{\sqrt{C}} 
$$

We estimate $a,b$ by minimizing the Fortet error and off we go. The estimates
are given in \cref{tab:OU_from_ML1spikes_ests} ($N=92$ spikes (20 secs.)). We
see that the estimates are quite accurate meaning that the spikes only estimates are no more
than $10$\% off from the estimates using detailed trajectories (in
\cref{tab:OU_from_ML1_ests}.

It remains to be seen whether this error has any discernible effect on the
control performance (It is plausible that the performance is imporved)

An aside/caveat: the Fortet estimator seems to be quite sensitive to
initial conditions, meaning that if you give it a too small guess for $\s$ is estimates
the noise intensity as $0$ (Possible bug of mine, I don't know).
\begin{table}
\begin{tabular}{l|c}
\input{Figs/MLBox/OU_from_ML1_estimates_spikes_only.txt}
\end{tabular}
\caption{SPIKES-ONLY: Estimates for the three OU parameters $m, C, \s$ given
data from an ML type 1 model (middle-column) and then from an OU model ran with the estimated
parameters of the ML 1 model (right column)}
\label{tab:OU_from_ML1spikes_ests}
\end{table}

\begin{table}
\begin{align*}
\mu &= 2.99 &\tc &= 0.22 &\beta &= 0.70
\\
\mu &=3.64  &\tc &= 0.22  &\beta &= 0.57
\end{align*} 
\caption{Estimates for the three non-dim OU parameters $\mu , \tc ,\beta$
Top: Voltage tracings, Bottom (Spikes-Only). recall that $\tc$ is artificially
set in the spikes only, which is why the two are the same!}
\label{tab:OU_nondim_voltage_vs_spikesonly}
\end{table}

\subsection{ML Control based on Spike-only estimates}
The final thing to do is control the ML model using the open-loop controller
based on the spikes only measurements. (It seems unnecessary to show the
performance of the closed-loop controller on spikes-only estimates - If you
can close-loop control, then you can voltage-estimate\ldots)

The performance is shown in
\cref{fig:ML_controled_simulation_ol_A10_spikesonly}.

Now Susanne suggested that the arbitrary, un-identifiable value of $C$ be fixed
to something external (from 'above'). Ok, let's do this, we set $C=20$ and show
the open-loop results in
\cref{fig:ML_controled_simulation_ol_A10_spikesonly_C20}

%%%%%%%%%%%% OL %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h] 
\begin{center}
\subfloat[example controlled trajectory] 
{
\label{fig:ML_example_controled_traj}
\includegraphics[width=\textwidth]
{Figs/MLTrainController/MLSim_example_ol_Amax10_spikesonly.pdf} 
}\\
\subfloat[raster plot for controlled simulations]
{
\label{fig:ML_raster_plot}
\includegraphics[width=\textwidth]
{Figs/MLTrainController/MLSim_raster_ol_Amax10_spikesonly.pdf} 
}
\caption[labelInTOC]{Same as \cref{fig:ML_controled_simulation_ol_A10}, but now
the parameters are estimated using spikes-only ($N=92 $spikes for estimation
$T=20$ secs.)}
\label{fig:ML_controled_simulation_ol_A10_spikesonly}
\end{center}
\end{figure}
\begin{figure}[h] 
\begin{center}
\subfloat[example controlled trajectory] 
{ 
\label{fig:ML_example_controled_traj}
\includegraphics[width=\textwidth]
{Figs/MLTrainController/MLSim_example_ol_Amax10_spikesonly_C20.pdf} 
}\\
\subfloat[raster plot for controlled simulations]
{
\label{fig:ML_raster_plot}
\includegraphics[width=\textwidth]
{Figs/MLTrainController/MLSim_raster_ol_Amax10_spikesonly_C20.pdf} 
}
\caption[labelInTOC]{Same as
\cref{fig:ML_controled_simulation_ol_A10_spikesonly}, but now the $C$ parameter
is fixed at $20.0$ (millisecs/ohms?) instead of using the voltage-based
estimation (when $C=39.46$ see \cref{tab:OU_from_ML1_ests}))}
\label{fig:ML_controled_simulation_ol_A10_spikesonly_C20}
\end{center}
\end{figure}

\bibliographystyle{plain}
\bibliography{library}
\end{document}

### 2014-02-18 abstract.tex
Adding ML Note

--- 2013-09-02 draft.tex
@@
 \begin{abstract}
+We have devised an algorithm for eliciting exact spike times from a noisy
+neuron. The strategy is especially designed to deal with a significant
+stochastic component in the impinging current and is applicable in both
+Supra- and Sub-threshold regimes. Depending on the measurements available to
+the controller, this involves solving a Hamilton-Jacobi-Bellman PDE or a pair of
+Kolmogorov PDEs (Forward and Backward). We provide a numerical implementation of
+the control strategy and test its performance in simulations for a variety of
+possible parameter regimes.
 
-{\em Objective.} External control of exact spike timings in single neurons is
-important for basic understanding of neuronal activity, and has potential in
-brain-machine interfaces, for improving neural prostheses, and in medical
-applications related to neuronal disorders, such as Parkinson's disease or
-epilepsy. The goal of this paper is the design of optimal controlled electrical
-stimulation patterns to a cell to achieve a target spike train under
-physiological constraints not to damage tissue and minimize possible side
-effects. {\em Approach.} We pose an optimal control problem to precisely specify
-the spike timing in a noisy leaky integrate-and-fire model of a neuron. In
-particular, we allow for the noise to be of arbitrary intensity. The optimal
-control problem is solved using Dynamic Programming if the controller has access
-to the voltage (closed-loop control) and using a Maximum Principle for the
-transition density if the controller only has access to the spike timings (open
-loop control). {\em Main results.} We have developed a stochastic optimal
-control scheme to obtain precise spike timings, which is not limited by the
-intensity of the noise in the system. Given this scope, the numerical demands
-are non-trivial, but simulations show that the algorithms are potentially
-feasible. It however remains to see if they can be successfully applied in a
-real-time experimental setting. Furthermore, we show how wrong the control
-strategy can perform if we assume that the noise is negligible. {\em
-Significance.} More precise targeting and control of neural activity to achieve
+\end{abstract}
+
+\maketitle
+
+\section{Novelty and Significance}
+External control of exact spike times in single neurons is important for
+understanding neuronal activity and has potential in brain-machine interfaces,
+for improving neural prostheses, and in medical applications to neuronal
+disorders, such as Parkinson's disease or epilepsy. The goal of this paper is
+the design of optimal controlled electrical stimulation patterns to a cell to
+achieve a target spike train under physiological constraints not to damage
+tissue. We pose a stochastic optimal control problem to precisely specify the
+spike times in a noisy leaky integrate-and-fire model of a neuron. In
+particular, we allow for noise of arbitrary intensity. The optimal control
+problem is solved using Dynamic Programming if the controller has access to the
+voltage (closed-loop control) and using a Maximum Principle for the transition
+density if the controller only has access to the spike times (open-loop
+control). The techniques are applicable in both the supra-threshold and
+sub-threshold regimes. Given this scope, our simulations show that our
+algorithms produce the desired results. They further reveal that the numerical
+demands are non-trivial, such that the algorithms may require efficiency
+refinements to achieve real-time control, in particular the open-loop context is
+more numerically demanding than the closed-loop one. Furthermore, we show how
+wrong the control strategy can perform if we assume that the noise is
+negligible.  More precise targeting and control of neural activity to achieve
 optimal performance in a noisy environment, in which most neurons are embedded,
@@
 
-\end{abstract}
-% \tableofcontents MSC KEYS: 49N90 90C39 93E20 93C20 93C95
-
-\maketitle
-
-\tableofcontents
-
-\section{Introduction}
-Manipulation of individual neurons through external electrical
-stimulation provides a means to
-controlling the spiking activity. Applications may arise, for
-example, in brain-machine interfaces or in artificial prosthetics.
-A natural goal is to make a cell produce a pre-specified spike train,
-under certain constraints on the modulation of the input.
-We consider this problem in the framework of Stochastic Optimal Control and
-give both a feedback solution (closed-loop), when the cell voltage is
-explicitly observable, as well as an open-loop solution when only the
-occurrence of spikes is observable. Importantly, we allow for an
-arbitrary noise intensity.
-
-Both theoretically and in practice related problems have been
-addressed in the literature. One objective has been to
-obtain either minimum or maximum interspike intervals lengths,
-when the input is constrained to be between some prespecified upper
-and lower bounds, see e.g. \cite{Lee1994,Lefebvre1987} for a
-mathematical treatment, or \cite{Danzl2009,Nabi2012,Wu2009} in a neuronal
-context. Another objective is to break a pathological synchronous
-firing pattern in clusters of neurons, highly relevant for neurological
-disorders such as epilepsy and Parkinson's disease,
-\cite{Nabi2013a,Nabi2011}, see also \cite{Feng2007b}.
-
-Our objective, namely targeting exact spike times in single neurons, has been
-considered mainly in the open-loop context, and in either absence of or for
-small noise. In \cite{Ahmadian2011} they use the Spike Response Model,
-\cite{Gerstner2002}, to control output target spike trains and implement their
-scheme on pyramidal cells in mouse cortical slices. Their method is numerically
-efficient and allows for the simultaneous control of many neurons. However the
-efficiency of the numerics strongly relies on the assumption that the diffusion
-noise, the value of $\b$ in \cref{eq:X_evolution_uo} below, is small enough, see
-the discussion above and below equation 12 in \cite{Ahmadian2011}. They only
-work with open-loop control. The difference between the objective in
-\cite{Ahmadian2011} and what we consider is that they maximize the probability
-of spiking at some given time $t^*$, whereas we minimize the mean squared
-difference between the realized spike time, $T_{sp}$ and the desired, $t^*$.
-Moehlis et al. \cite{Moehlis2006} work with a Phase Response Model to obtain
-spikes at exact times, while keeping the root mean square of the input to a
-minimum. A similar approach is taken in \cite{Dasanayake2011}. They do not
-consider noise, though, and only work with open-loop control. In
-\cite{Nabi2013}, the methods are implemented on brain slices of pyramidal
-neurons of rat hippocampus. While the
-Phase Response Curve Model is a parsimonious and effective way to describe a
-neuron's response to a stimulus, it is only valid in the supra-threshold regime,
-where the unstimulated neuron is periodically spiking. In \cite{Feng2003} they
-use a leaky integrate-and-fire model and open-loop control. To
-obtain a given spike time, they choose an objective of minimizing the
-variance of the membrane potential at the desired spike time, while forcing the mean of
-the membrane potential at this time point to equal the threshold. This provides
-exact solutions since it does not involve first-passage times, but has the
-drawback that there is a non-negligible probability that the obtained
-spike time will be far from the desired spike time.
-
-Our objective of imposing a certain timing sequence of the spike train using an
-externally applied control is obtained in both the closed- and
-open-loop setting, and we specifically include the noise in the
-calculations of the controls. We restrict the controlled input to stay
-within pre-specified bounds, and also include a cost function to
-minimize intervention.
-Our main contributions are that we specifically allow for a non-negligible
-noise component in the neural activity when calculating the control, and
-consider both open- and closed-loop controls. In particular we do not
-restrict our attention to the autonomously-spiking, supra-threshold regime.
-
-The paper is structured as follows: First we describe the neuron model and
-formalize the control objective. Then we describe a feedback-based solution,
-which assumes that the controller has detailed access to the voltage trajectory.
-Then we relax the observation assumption so that the controller only has access
-to the spike timings. Finally, we compare the two methods through
-simulations against a simple-minded control technique which ignores
-the stochastic input to the neuron.
-
-\section{Problem Formulation}
-A basic but useful model for the neural membrane potential evolution is the
-noisy leaky-integrate-and-fire model:
-\begin{equation}
-\begin{gathered}
-dX_s = \left(\Iext(t) - \frac{X_s}{\tc} \right) \intd{s} + \b \intd{W_s},
-\\
-X(0) = 0,
-\\
-X(\ts) = \xth \implies
-\begin{cases}
-X(\ts^+) = 0 &
-\end{cases}
-\end{gathered}
-\label{eq:X_evolution_uo}
-\end{equation}
-Here $X$ represents the membrane electric potential which decays to $0$
-with a time constant of $\tc$, $dW$ is a Brownian motion increment scaled by
-$\b$ and $\Iext(t)$ is the deterministic external input to the cell. Having last
-spiked at time $0$, the potential hits $\xth$ at some time $\ts$, the potential
-resets to $0$ and the process starts all over again. 
-We write $\ts^+$ for the voltage's limit from the right at $\ts$.
-Throughout the paper the threshold is set to one, $\xth = 1$.
-
-Suppose that we have some control over the external current such that it can be
-decomposed as
-\begin{equation}
-\Iext(t) = \m + \a(t)
-\label{eq:current_mu_alpha}
-\end{equation}
-where $\m$ is an uncontrollable, but constant part of the external current and
-$\a(t)$ is controllable, i.e.\ it can be chosen to achieve some goal.
-A natural goal is to attempt to control the spike time, $\ts$.
-That is how do we choose $\a(t)$ such that $\ts = \T$.
-A natural optimal control objective to achieve this is the least squares solution
-\begin{equation}
-\a(t) = \argmin_{\a(t)} \{\, \Exp[(\ts- \T)^2] \,\}
-\label{eq:OC_LS_variance}
-\end{equation}
-where expectation is taken with respect to the distribution of the trajectories
-of $X_t$.
-
-Often the control has certain constraints. The most common are simple box
-constraints:
-\begin{equation}
-\a(t) \in [\amin, \amax] \quad \forall t.
-\label{eq:bound_constraints_alpha}
-\end{equation}
-
-% \subsection{Optimal Control Formulation}
-In addition to \cref{eq:OC_LS_variance}, we will also add to our objective a
-running cost based on the control. This regularizes the problem eliminating the
-subtleties of singular-control situations and it serves to avoid excessive
-control as well as to avoid excessive charge building up on the cell.
-
-So we seek an optimal control, $\a^*$, that solves
-\begin{align}
-J[\a(\cdot)] =&
-\Exp\left[
-\e \int_0^\ts  \a^2(s) \intd{s}
-+
-(\ts - \T \big)^2 \right]
-\label{eq:OC_LS_variance_energy}
-\\ \notag
-\a^*(\cdot) =& \argmin_{\a(\cdot)} J[\a(\cdot)].
-\end{align}
-where $\e$ measures how much weight we put on minimizing the energy cost.
-If $\e = 0$, then we do not care at all about the expended cost.
-
-It will often be the case that $\a$ means either a function or a
-value of that function at a particular time. This function could be
-random, i.e.\ a stochastic process. Since $\a$ is used to control the behaviour
-of $X$, it will be at least measurable with respect to the filtration of $X$,
-see \cite{Oksendal2007} for the details. We will try to make clear below when we
-are refering to $\a$ as a stochastic process and when we are merely refering to its particular
-value at some time. For example, in \cref{eq:OC_LS_variance_energy}, $\a(\cdot)$
-refers to the function, while $\a(s)$ refers to that function's value, possibly
-random, at time $s$.
-
-We will consider two control contexts -- {\sl closed-loop} and  {\sl open-loop}
-control. In closed-loop control, the value of $X_t$ at time $t$ is observable
-and can be used in determining the control. In open-loop control only the spike
-times are observable. In the closed-loop context, we write the control as $$\a =
-\a(x,t)$$ to indicate its dependence on $X_t$ and to express that it will be
-updated based on the time-course of $X$. In the open-loop context, we write $$\a
-= \a(t)$$ to indicate that $\a(\cdot)$ is decided for all times at time 0. The
-techniques used to obtain the optimal controls in the two scenarios will be
-different. For the closed-loop scenario we use Dynamic Programing,
-see \cite{Fleming1975}, while for the open-loop scenario we use a form of the
-Maximum Principle applied on the transition density of the controlled process,
-see \cite{Borzi2012}.
-
-Crucially, we assume that the model parameters, $\m,\tc,\b$ in
-\cref{eq:X_evolution_uo} are known. This is a strong assumption and will be
-discussed later.
-
-\subsection{Parameter Regimes}
-Different  parameter regimes can be envisioned given \cref{eq:X_evolution_uo},
-depending on whether the noise intensity, $\b$, is
-relatively high or low, and whether the external, uncontrollable bias current,
-$\m$, induces spikes in the absence of noise or not. 
-Spikes will occur in the absence of noise, if and only
-if $ \m > \frac 1\tc,$ which is called the supra-threshold regime. When $\m
-\leq 1/\tc$, the regime is called the sub-threshold regime.
-In addition we will impose two values of $\b$, which we
-will call high-noise and low-noise, respectively.
-Example values for each parameter regime are given in \cref{tab:regimes}, and we
-visualize a single path for each regime in \cref{fig:regime_path_examples}.
-\begin{table}
-\begin{tabular}{|l||{c}|{c}|}
-\hline
-\backslashbox{$\m$}{$\b$}
-& $1.5$ & $0.3$ \\
-\hline
-$1.5 / \tc $ &Supra-threshold-High-noise & Supra-threshold-Low-noise \\
-\hline
-$0.1 / \tc$   &Sub-threshold-High-noise & Sub-threshold-Low-noise \\
-\hline
-\end{tabular}
-\caption{Regime labels and example values. Note that for the numerical
-experiments below, we use $\tc = 0.5$}
-\label{tab:regimes}
-\end{table}
-%\usepackage{graphics} is needed for \includegraphics
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=0.99\textwidth]{Figs/PathSimulator/path_T=14_combined.pdf}
-  \caption[labelInTOC]{Example Trajectories from \cref{eq:X_evolution_uo}
-  using the parameter values from \cref{tab:regimes}. 
-  A) Supra-Threshold-Low-Noise 
-  B)Supra-Threshold-High-Noise 
-  C) Sub-Threshold-Low-Noise
-  D) Sub-Threshold-High-Noise 
-  }
-  \label{fig:regime_path_examples}
-\end{center}
-\end{figure}
-
-\section{Closed-Loop Solution - Dynamic Programing}
-We now detail the Dynamic Programing approach to obtaining a feedback-optimal
-control. What this amounts to is that the controller can be continuously updated
-depending on the realization of the stochastic process, $X_t$.
-
-Given a time $t$ and a realized voltage value $X_t = x$, let $\trem$ be the
-(unknown) remaining time to spike, $\trem = \ts - t$. Starting from an arbitrary
-$t, x$, our remaining-cost objective, $J[\a(\cdot); x,t]$, will be
-\begin{equation}
-J[\a(\cdot); x,t]  =
-\Exp\left[
-\big(\trem - (\T-t) \big)^2
-+
-\e \int_t^\ts  \a^2(s) \intd{s}
-\,\Big|\, X_t = x
-\right]
-\label{eq:OC_LS_cost_to_go}
-\end{equation}
-i.e.\ if time $t$ has elapsed without a spike, we now want to minimize
-the difference between $\t = \ts - t$ and ($\T-t$), given the current state $x$.
-Recall that in the closed-loop scenario, we assume that the value of $X_t$
-is known to the controller. A similar problem is discussed analytically at
-length in the book on optimal control by Peter Whittle, \cite{Whittle1996},
-although there is no discussion there of the numerics required to solve it.
-
-\subsection{Hamilton-Jacobi-Bellman equation}
-The Hamilton-Jacobi-Bellman (HJB) equation,
-see \cite{Fleming1975,Whittle1996} or \cite{Danzl2009,Nabi2013a} for a neuroscience
-context, associated with
-\cref{eq:OC_LS_cost_to_go} is obtained as follows. We introduce the value function, $\v(x,t)$:
-\begin{align}
-\v(x,t) =&
- \min_{\a(\cdot)_{s \geq t}}
- \{J[\a(\cdot); x,t] \}
-\notag
-\\
-=&
-\min_{\a(\cdot)_{s \geq t}}
-\Exp\left[
-\big(\trem - (\T-t) \big)^2
-+
-\e \int_t^\ts  \a^2(s) \intd{s}
-\,\Big|\,X_t = x
-\right]
-\label{eq:OC_LS_mean_value}
-\end{align}
-
-Then $\v$ satisfies the following HJB partial differential equation (PDE):
-\begin{equation}
-\begin{gathered}
-\di_t \v(x,t) + \tfrac{\b^2}{2} \di_x^2 \v(x,t) + \\
-\min_{\a(x,t) \in [\amin,
-\amax]}\Big\{ \e \a^2 (x,t)+ \left(\m + \a(x,t)-\tfrac{x}{\tc}\right) \di_x
-\v(x,t) \Big\} = 0
-\label{eq:OC_LS_mean_HJB}
-\end{gathered}
-\end{equation}
-The special feature of \cref{eq:OC_LS_mean_HJB} in contrast to a generic
-parabolic PDE is that it contains an embedded optimisation that depends on the
-solution, $\v$. For each $x,t$ in the computational domain,
-$\a$ is chosen as to minimize $\{\e \a^2 + \left(\m + \a - \tfrac{x}{\tc}\right) \di_x \v \}$.
-Here we can solve for the optimal control, $\a^*(x,t)$ analytically as:
-\begin{align}
-\a^*(x,t)  =& \argmin_{\a \in [\amin, \amax]}\big\{\e \a^2 + (\m + \a-\tfrac{x}{\tc}) \di_x\v\big\}
-\notag
-\\
-=&
-\min \left(\amax, \max\left(\amin, -\frac{\di_x\v(x,t)}{2\e}\right)\right)
-\label{eq:OC_LS_variance_energy_quadratic_control}
-\end{align}
-With this the HJB PDE becomes
-\begin{equation}
-\di_t \v(x,t) + \tfrac{\b^2}{2} \di_x^2\v+
-\e (\astar(x,t))^2 + (\m + \astar(x,t)-\tfrac{x}{\tc}) \di_x\v
-= 0
-\label{eq:OC_LS_mean_HJB_bounded_control}
-\end{equation}
-
-We need to consider boundary conditions (BCs) for $\v$. If $X_t = \xth$ then we
-have a spike now and $\ts = t$. Thus $$ \v(\xth,t) = (t-\T)^2.$$ At the
-threshold, the value function equals the squared difference between the desired
-spike time and the realized one. For $x \ll 0$, we assume that $\v$ is not
-significantly affected by the change in $x$, i.e.\ that $$ \di_x \v(\xmin, t) =
-0 $$ for some lower boundary $\xmin$. Such a boundary condition will be
-justified if we impose artificial reflecting boundaries for the diffusion at
-$\xmin$. So we need to determine for which $\xmin$ is a reflecting boundary
-consistent with the dynamics. For example, we can take $\xmin$ to be two
-standard deviations below the mean of the stationary distribution of the
-maximally inhibited, $\a=\amin$, \cref{eq:current_mu_alpha}. That is we set
-$\xmin = \tc(\mu + \amin) - 2 \b / \sqrt{\tc/2}$. We further enforce
-that $\xmin \leq -0.5$.
-% Ooops, we need $\di_x\v\neq 0$ and here we are setting $\di_x\v= 0$ in one of
-% the BCs\ldots oh well.
-
-To determine the Terminal Conditions (TCs) for $\v$, the idea is
-simple: if we reach $\T$ without having spiked we apply maximum control in the
-positive direction, i.e.\ $$t>T \implies \a(t) = \amax$$ Thus:
-\begin{equation}\v(x,\T) =
-\Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big].
-\label{eq:valuefun_TCs}
-\end{equation}
-Note that we are making an approximation here
-- we are ignoring the energy term, $\eps u^2$, in the objective for
-$t>T$. Naturally, this approximation is ever more accurate for $\eps \ll 1$.
-We discuss in more detail the validity of the approximation in
-\cref{sec:effect_of_eps}.
-Alternatively, we could impose this approximate terminal condition at some $t^+
-> \T$.
-
-We will see the quantity on the right hand side of \cref{eq:valuefun_TCs}
-repeatedly so we will give it a special name.
-\begin{equation}
-\Ttwo(x) := \Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big].
-\label{eq:Trem_squared}
-\end{equation}
-$\Ttwo$ is the second moment of the remaining time to reach the
-threshold starting at $X_\t = x$ and applying $\amax$ throughout. This quantity can be found
-easily for all $x$ in the domain by solving a stationary backward Kolmogorov
-equation. We show the details in \ref{sec:valuefun_TCs}.
-
-Thus, we restate the HJB equation in its fully specified form:
-\begin{equation}
-\begin{gathered}
-\di_t \v(x,t) + \tfrac{\b^2}{2} \di_x^2\v(x,t)+
-\e \a^2(x,t) + \big(\m + \a(x,t) -\tfrac{x}{\tc}\big)\cdot \di_x\v(x,t)
-= 0
-\\
-\a (x,t) = \min \left(\amax, \max\left(\amin, -\frac{\di_x \v(x,t)}{2\e}\right)
-\right)
-\\
-\begin{cases}
-\v(\xth,t) = (t-\T)^2  \quad & \textrm{upper BC}
-\\
-\di_x \v(\xmin, t)  = 0  \quad &\textrm{lower BC}
-\\
-\v(x,\T)  =
-% \Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big]
-% \Exp[\t^2|x, \amax]  \quad
-\Ttwo(x)
-\,& \textrm{TC}
-\end{cases}
-\end{gathered}
-\label{eq:OC_LS_HJB_full}
-\end{equation}
-We are solving $\v(x,t)$ over the domain $[\xmin, \xth] \times [0,T]$.
-
-\subsection{The numerical method for the HJB equation}
-
-We now have a PDE for $\v$ and an algorithm for computing all the BCs and TCs of
-this PDE. It is time to discuss the numerical method for solving
-\cref{eq:OC_LS_HJB_full}. Since it is one-dimensional, it is straight forward to
-apply the standard centred finite difference using the Crank-Nicholson scheme to
-step in time. We have not found it necessary to upwind the first order
-derivatives for the regimes we have explored. To resolve the non-linearity,
-$(\di_x\v)^2$, in the PDE, we treat it as a mixed implicit-explicit term
-$$(\di_x \v(x,t_k))^2 = \di_x \v(x,t_k) \cdot \di_x \v(x,t_k) \approx
-\underbrace{\di_x \v(x,t_k)}_{\textrm{implicit}} \cdot
-\underbrace{\di_x\v(x,t_{k+1})}_{\textrm{explicit}}.$$ Note that the implicit
-term is in the previous time $t_k$ instead of, as is conventional, the next
-$t_{k+1}$, because we are solving for $\v$ {\sl backwards} in time.
-
-The numerical scheme is implemented in Python, using the Scipy/Numpy library,
-\cite{Scipy}. For the discretization in time and space, we choose $\Delta x$ and
-$\Delta t$ in relation to the parameter regimes. for example in the
-Supra-Threshold Low-Noise regime, we have $\Delta x = 0.0043$ and $\Delta t =
-0.0012$. We consider these to be a fairly fine discretization.
-
-We next demonstrate solutions for the value function and the associated
-control law for a parameter set from each regime in \cref{tab:regimes}.
-
-\subsection{Solutions of the HJB equation}
-We set $\T=1.5$ and $\eps = 0.001$, so that the primary focus is the accurate
-spiking and energy minimization considerations are secondary.
-
-In \cref{fig:HJB_4regimes_value_control_cuts}, we depict three snapshots
-in time of the value function and the control, $\v$ and $\a$, at $t = [0,\T/2,
-\T]$, i.e.\ at the beginning, middle and end of the relevant time interval. Then in
-\cref{fig:HJB_4regimes_value_surf} we show the entire value function, $\v(x,t)$,
-the solution to \cref{eq:OC_LS_HJB_full}, for each of the parameter regimes.
-% THE CUTS:
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/HJB/Regimes_vc_cuts.pdf}
-  \caption[labelInTOC]{Snapshots of the value function, $\v$, on the left and on
-  the optimal control, $\a$, on the right for $t$ fixed, at the start (blue),
-  mid-point (green) and end-point (red), for each of the four regimes. The desired spike time is set to $\T=1.5$, the energy penalty, $\eps
-  = 0.001$. The control bounds are $\a \in [-2,2]$. Note that the green and blue
-  curves in E) are lying on top of each other for all $x$.
-  \\
-  A,E) Supra-Threshold-Low-Noise
-  B,F) Supra-Threshold-High-Noise
-  C,G) Sub-Threshold-Low-Noise
-  D,H) Sub-Threshold-High-Noise
-  }
-\label{fig:HJB_4regimes_value_control_cuts}
-\end{center}
-\end{figure}
-% THE SURFS:
-\begin{figure}[h!]
-\begin{center}
-\includegraphics[width=0.99\textwidth]{Figs/HJB/Regimes_valuesurf.pdf}
-\caption{The numerical solution for the value function $\v(x,t)$ HJB PDE,
-\cref{eq:OC_LS_HJB_full}, for the four different parameter regimes.
-The desired spike time is set to $\T=1.5$.
-The control bounds are $\a \in [-2,2]$.
-The contours for fixed $t$ plotted in thick correspond to the graphs in \cref{fig:HJB_4regimes_value_control_cuts}.
-\\
-A) Supra-Threshold-Low-Noise
-B) Supra-Threshold-High-Noise
-C) Sub-Threshold-Low-Noise
-D) Sub-Threshold-High-Noise  }
-\label{fig:HJB_4regimes_value_surf}
-\end{center}
-\end{figure}
-
-Let us discuss the most salient features in
-\cref{fig:HJB_4regimes_value_surf,fig:HJB_4regimes_value_control_cuts}. Recall
-that lower values for the value function, $\v(x,t)$, are preferred, since we are
-minimizing. We see the same general shape in all four space-time surface plots
-of the value function in \cref{fig:HJB_4regimes_value_surf}. At the end of the
-interval, $\v(x,\T)$ is monotonically decreasing in $x$, which is to be expected
-given its terminal condition in \cref{eq:valuefun_TCs}. That is, the lower the
-value of $X_\T$ the more on average will we have to wait for the spike to occur.
-% This same basic shape represents the idea that we would like to be close to
-% the threshold, $x = 1$, near the end of the interval when $t \approx \T$, but
-% we would like to be far from threshold, for $t \ll \T$. Indeed, at the very
-% end of the interval, when $t = \T$, the value function, $v(x,\T)$, is
-% monotonically decreasing in $x$, because the lower the value of $X_\T$ the
-% more time on average will it take us to spike and the more delay will we have
-% from our desired spike time.
-As we go back in time $\v$ inverts, with a clear peak near the upper, threshold
-boundary. That is, for intermediate values of $t$ we have to consider the risk
-of spiking too early, represented by the high value of $\v$ near the upper
-boundary {\sl and} the risk of spiking too late, represented by the high values
-of $\v$ at the lower boundary. As we progress even further back in time to the
-beginning of the interval, the peak near the threshold rises further, since we are spiking
-{\sl earlier}, while the peak at the lower end flattens, since now there is
-enough time to reach threshold despite starting far from it. The full surface
-plots for $\v(x,t)$ in \cref{fig:HJB_4regimes_value_surf} can be thought of as
-the interpolation between these three basic phases.
-
-% Note, that exact shape is generated by Since we have put a very small weight
-% on the energy penalty, $\e = .001$, and we have a significantly high bound on
-% the maximum / minimum of the applied control, the value function becomes quite
-% flat for $x \ll 1$ and $t \ll \T$, since it is likely that we can apply enough
-% control to avoid an early spike and still drive the system close to threshold
-% at the desired time, $t = \T$
-Now focus on the controls, on the right in
-\cref{fig:HJB_4regimes_value_control_cuts}. Naturally, at the end of the
-interval, the control takes its maximum value, $\a(x,\T) = \amax$ for all $x$,
-i.e.\ it gives the maximum available push for the neuron to spike. This is
-implied \cref{eq:valuefun_TCs}. As we go back in time, however, the control $\a$
-decreases for $x \ll \xth$, and it becomes negative, i.e.\ inhibitory for $x
-\lesssim \xth$. That is intuitively consistent with the problem objective. 
-For $t<\T$, we want to bring $X_t$ closer to the threshold, but not too close, to
-avoid early spiking.
-
-Now consider the differences between the individual regimes, i.e.\ the
-effect of changing the bias and the noise intensity, $\m,\b$. All things being
-equal, the effect of increasing the noise intensity, $\b$, is to lift the value
-function, i.e.\ to make our objective worse, at the beginning of the interval
-and to decrease it at the end. This is to be expected since we are attempting to
-minimize the variance in the spike time and without noise there would be no
-variance at all. However, at the end of the interval, noise only helps, since
-now we are only interested in spiking as soon as possible, and a higher noise
-intensity will tend to decrease the spike time on average. Increasing $\b$ also
-has the effect of increasing the size of the boundary layer near $\xth$, where
-the value function rises steeply. I.e.\ for small $\b$ that layer is small since
-the risk of spiking early is only significant close to the threshold. For larger
-$\b$ this layer increases, see panels A,C, with a thin layer, vs.\ B,D, with a
-thicker layer, in \cref{fig:HJB_4regimes_value_control_cuts}. Similarly,
-increasing the bias, $\mu$, tends to decrease the value function, especially at
-the end since that has the unequivocal effect of preventing late spikes.
-
-\section{Open-loop Stochastic Control}
-When the value of $X_t$ is unobservable, the transition density can be used to
-perform the optimization. Since the transition density follows a deterministic
-PDE, we apply a Maximum Principle for PDEs as a method of obtaining the optimal
-control, see \cite{Borzi2012} for details on optimization with PDEs.
-
-\subsection{Fokker-Planck equation for the state density evolution}
-We write the transition density of $X$, conditional on no
-spikes having occurred as:
-\begin{align*}
-f(x,t) \intd{x} &= \Prob[X_t\in \intd{x} \,\big|\, X_0 = 0,\, X_{s} < \xth \quad
-\forall s < t]
-\end{align*}
-Then $f$ satisfies a Fokker-Planck equation with absorbing boundaries,
-see Ch.\ 7 in \cite{Jacobs},
-\begin{equation}
-\begin{gathered}
-\di_t f(x,t) =
-				\frac{\b^2 }{2}\cdot \di^2_x f -
-				\di_x \Big[ \left( \m + \a(t)- \frac{x}{\tc}\right)  \cdot f \Big]
-\\
-\\
-\begin{array}{lll}
-	&\textrm{st.\ }&
-	\left\{ \begin{array}{lcl}
-	 f(x,0) &=& \delta(x) \quad \textrm{delta function}
-	\\
-	(\m + \a(t)- \tfrac x \tc)f - \di_x \tfrac {\b^2}2 f] \big|_{x=\xmin} &\equiv&
-	0 \quad \textrm{reflecting BCs at some } \xmin
-	\\
-	f(x,t) |_{x=\xth} &\equiv& 0 \quad \textrm{absorbing BCs at } \xth
-\end{array} \right.
-\end{array}
-\label{eq:FP_pde_OU_PDF}
-\end{gathered}
-\end{equation}
-In theory, $\xmin = -\infty$, but in the numerics below we will need to
-truncate it to some finite value, exactly as in the HJB equation,
-\cref{eq:OC_LS_HJB_full}.
-Note that we write $\a(t)$ here instead of $\a(x,t)$ since we cannot use the
-value of $X_t$.
-
-The $f$ dynamics can also be written as
-$$
-\di_t f(x,t) = - \di_x \phi(x,t)
-$$
-for the probability flow
-$$
-\phi(x,t) = (\m + \a(t) - \tfrac x \tc)f - \di_x [\tfrac {\b^2}2 f]
-$$
-Then the lower BC is
-$$
-\phi(x,t) |_{x=\xmin} \equiv 0
-$$
-
-We will also need a short hand notation for the differential operator on the
-right side of \cref{eq:FP_pde_OU_PDF}. Let
-$$ \L_\a[\cdot] := \di^2_x \Big[\frac{\b^2 }{2} \cdot\Big] -
- \di_x \Big[ (\m + \a(t)- \frac{x}{\tc}) \cdot \Big] $$where
- $[\cdot]$ indicates the argument of the operator $\L_\a$. We will usually
- omit the subscript $\a$, but have written it now to emphasize that the
- differential operator is parametrized by the control, $\a$.
-
-\subsection{Restating the objective in terms of the transition density}
-Let us recall our objective, \cref{eq:OC_LS_variance}:
-$$
-J[\a(\cdot)] = \Exp\left[
-(\ts - \T \big)^2
-+
-\e \int_0^\ts  \a^2(s) \intd{s}
-\right].
-$$
-Rewriting this in terms of the transition density, $f$, reads:
-\begin{align}
-J[\a(\cdot)] =&
-\int_\xmin^\xth \Ttwo(x) \cdot f(x,\T) \intd{x}
-\notag
-\\
-&+ \int_0^\T \phi(\xth, t) (t-\T)^2 \intd{t}
-\label{eq:OC_LS_variance_density}
-\\
-&+  \e \int_0^\ts  \a^2(t)  \int_\xmin^\xth f(x,t) \intd{x} \intd{t}
-\notag
-\end{align}
-Let us explain in more detail each term on the right-hand side in
-\cref{eq:OC_LS_variance_density}.
-
-The first term $$ \int_\xmin^\xth \Ttwo(x) \cdot f(x,\T)\intd{x} $$ counts the cost of trajectories which spike too late. This cost is the expected
-squared time-to-hit starting at $x$, with  $\a = \amax$, weighted by the
-probability of $X_\T = x$, which is just $f(x,\T)$. Recall that $\Ttwo$ is
-defined in \cref{eq:Trem_squared}.
-
-The second term $$ \int_0^\T \phi(\xth,t) (t-\T)^2 \intd{t} $$ counts the cost
-of trajectories which spike too early, that is the squared difference between
-some realized spike time, $\ts = t$, and desired spike time $\T$, weighted by
-the probability of a spike at $t$ which is just the outflow of probability,
-$\phi(\xth,t)$. Recall that we assume that $\xth = 1$ throughout. Note further
-that due to the homogeneous Dirichlet BC at $\xth$, $f(\xth) = 0$, the outflow
-is simply: $$ \phi(\xth, t) = -\frac{\b^2 }{2} \di_x f(\xth, t).$$
-
-Finally, the third term
-$$
-\e \int_0^\ts  \a^2(t)  \left[  \int_\xmin^\xth f(x,t) \intd{x} \right] \intd{t}
-$$
-is the energy cost: the inner integral, $\int_\xmin^\xth f(x,t) \intd{x}$
-takes into account that we incur an energy cost only for those trajectories
-which have not yet spiked.
-
-With that our optimal control $\a^*(\cdot)$ will naturally be found via:
-$$
-\a^*(\cdot) = \argmin_{\a(\cdot)} J[\a(\cdot)]
-$$
-
-\subsection{Optimizing using a Maximum Principle}
-\label{sec:PDE_max_principle_for_pdf}
-By now our stochastic optimal control problem modelled by SDEs
-has become a deterministic optimal control problem modelled by PDEs. Our control
-$\a$ influences the evolution density $f$ and via $f$, the integrals in the
-objective, $J$.
-
-The Maximum Principle for PDEs, which is an extension to the famous Pontryagin
-Maximum Principle from finite dimensional systems, introduces an adjoint
-variable, $p$, which solves a PDE related to the PDE satisfied by the density
-$f$ and then calculates the optimal control, $\a(\cdot)$, as a functional of $f$
-and $p$.
-
-In short, the equation for the adjoint function, $p$, is
-\begin{equation}
-\begin{gathered}
-\begin{aligned}
-\di_t p =& - \Lstar[p]
-\\
-		=&
-			- \Big[ \frac{\b^2 }{2}\cdot \di^2_x p +
-			(\m + \a(t)- \frac{x}{\tc})  \cdot \di_x p \Big]
-\end{aligned}
-\\
-\st
-\begin{cases}
-	p(x,\T) &= \Ttwo(x)
-	\\
-	\di_x p  \big|_{x=\xmin} &\equiv 0
-	\\
-	p \big|_{x=\xth} &\equiv (s-T)^2
-\end{cases}
-\label{eq:adjoint_pde_OU}
-\end{gathered}
-\end{equation}
-
-and then $\a$ can be found via
-\begin{equation}
-\Big\{
- \tilde{\e}  2 \a(t)
-+ p f \Big|_\xmin
-- \int _\xmin^\xth p \cdot \di_x f \intd{x}
-\Big\} = 0
-\quad \forall t \in [0,T]
-\label{eq:J_necessary_condition}
-\end{equation}
-
-More practically, the quantity,
-$$\di_\a J =  \Big\{
- \tilde{\e}  2 \a(t)
-+ p(x,t) f(x,t) \Big|_\xmin
-- \int _\xmin^\xth p \cdot \di_x f(x,t) \intd{x}
-\Big\}
-$$
-gives the direction of increase of $J$ at $\a(t)$
-and we can use it as a gradient in a descent algorithm, given some initial
-guess, $\a_0(t)$, for the control.
-
-Finally, we are ready to calculate the open-loop stochastic optimal control for
-the four parameter regimes. The calculated optimal controls for each regime are
-in \cref{fig:FBK_Regimes_cs}. Recall that the optimal control, $\a^*(t)$, is
-open-loop and thus it is only a function of time.
-% \usepackage{graphics} is needed for \includegraphics
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/FP_Adjoint/Regimes_cs.pdf}
-  \caption[labelInTOC]{The deterministic optimal controls for each parameter
-  regime as functions of time, $t \in [0, \T]$.
-  The desired spike time is set to $\T=1.5$, the energy penalty, $\eps
-  = 0.001$ and the bounds are $\a \in [-2,2]$.
-  A) Supra-Threshold-Low-Noise B) Supra-Threshold-High-Noise
-C) Sub-Threshold-Low-Noise   D) Sub-Threshold-High-Noise  }
-  \label{fig:FBK_Regimes_cs}
-\end{center}
-\end{figure}
-
-Let us discuss the most salient features of the controls calculated in
-\cref{fig:FBK_Regimes_cs}. In general, the strategy is to inhibit the neuron at
-the beginning of the interval, $\a(t) = \amin$ for $t \ll \T$ and to excite it
-near the end, $\a(t) = \amax$ for $t \approx \T$. The smooth portion that connects these
-two segments in the middle of the interval is due to the energy penalty, $\e \int
-\a^2(s) \intd{s}$. The behaviour of the control in different regimes
-is relatively simple to explain and we see what we would expect. In a
-sense the bias current $\mu$ can be absorbed by the control and so increasing
-$\mu$ amounts to decreasing $\a$ modulo its bounds. That is exactly the
-difference between plots A) vs.\ C) and B) vs.\ D) in \cref{fig:FBK_Regimes_cs},
-where the reduction of $\mu$ in the lower plots, C) and D), results in an
-increase in $\a(t)$ compared to A) and B), respectively. The effect of the noise
-intensity, $\b$, is more subtle and it is not the same in the Supra vs.\ the
-Sub-Threshold regimes A,B) vs.\ C,D). In the Supra-Threshold regime,  panels
-A,B), reducing $\b$, reduces the need for excitatory  control towards the end,
-since the bias alone should put the neuron over the threshold. Thus the main
-part of the control with low noise in the Supra-Threshold regime is to stop the
-neuron from spiking too early. In the Sub-Threshold regime, panels C,D), the
-situation is somewhat reversed - reducing $\b$, obviates the need to apply an
-inhibitory control in the early part of the interval and also prompts the
-controller to apply excitatory input earlier, since it is the controller which
-is now the main drive for a spike.
-
-% An example for . The results are in
-% \cref{fig:FP_adjoint_objective_control_convergence}.
-% \begin{figure}[h]
-% \begin{center}
-% \subfloat[$\a_k(t)$ ]
-% {
-% \includegraphics[width=0.48\textwidth]
-% {Figs/FP_Adjoint/ExampleControlConvergence_control.png}
-% }
-% \subfloat[$J_k$ ]
-% {
-% \includegraphics[width=0.48\textwidth]
-% {Figs/FP_Adjoint/ExampleControlConvergence_objective.png}
-% }
-% \caption[ ]{The result of an entire optimization iteration, on the left the
-% iterates of the control, $\a_k(t)$, on the right the progress of the
-% objective with each iteration. The final control, in purple, is inhibitory at
-% the beginning of the time interval, $\a < 0$, unlike the the deterministic
-% control solution.
-% We see a significant
-% reduction in the objective value, $J$, and thus an improvement, on the
-% order of 50\%.}
-% \label{fig:FP_adjoint_objective_control_convergence}
-% \end{center}
-% \end{figure}
-%
-% \begin{table}[h]
-% \centering
-% \begin{tabular}{lc}
-% Control Law & Squared Error \\
-% \hline
-% Deterministic &  0.647 \\
-% Stochastic &  0.336\\
-% Theory (Value function) & 0.285
-% \end{tabular}
-% \caption{Realized performance of the different control laws and the theoretical
-% expected performance of the stochastic law (last row)}
-% \label{tab:realized_avg_errors_det_vs_stoch}
-% \end{table}
-%
-% \begin{figure}[h]
-% \begin{center}
-% \subfloat[A]
-% {
-% \label{fig:controlled_traj_ex1}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/example_controlled_trajectories_id1.png}
-% }
-% \subfloat[B]
-% {
-% \label{fig:controlled_traj_ex2}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/example_controlled_trajectories_id4.png}
-% }
-% \subfloat[C]
-% {
-% \label{fig:controlled_traj_ex3}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/example_controlled_trajectories_id3.png}
-% }
-% \caption[]{Examples for the controlled trajectories using both the deterministic
-% and the stochastic control approaches. The red vertical line in the plots
-% indicates the desired spike-time, $\T$. In A, B the performance of both
-% control laws is essentially the same, but in C we see the advantage of the
-% stochastic approach. $\tc, \b = [.75, 1.25]$.}
-% \label{fig:control_trajectories_examples}
-% \end{center}
-% \end{figure}
-% \begin{figure}[htp]
-% \begin{center}
-%   \includegraphics[width=.9\textwidth]{Figs/ControlSimulator/example_controlled_trajectories_hists.png}
-%   \caption[labelInTOC]{Histogram of the spike timing error for the
-%   deterministic (left) vs. the stochastic (right) control laws.}
-%   \label{fig:error_histograms_det_vs_stoch}
-% \end{center}
-% \end{figure}
-
-\section{Simulations} 
-\subsection{Single-Spike Control}
-\label{sec:probabilistic_numerical_test}
-Having obtained the optimal controllers, both closed- and open-loop, we evaluate
-their performance with simulated realizations of the voltage process,
-\cref{eq:X_evolution_uo}. In particular, while they both minimize the expected
-squared deviation of the spike-time from some desired spike time, we analyze
-the distribution of the squared deviation and the behaviour of the
-controls for different parameter regimes.
-
-In addition to the optimal controllers, we also show the behaviour of
-another control law - perhaps the most naive one - which is obtained by ignoring
-the noise and the energy penalty. That is we find a {\sl constant} value of $\a$
-which satisfies the desired boundary conditions, $x(0) = 0, x(\T) = \xth$. This naive
-controller will be called 'Deterministic', since it assumes deterministic
-dynamics in $X_t$, i.e.\ $\b = 0$.
-
-For the comparison, we sample $N$ realizations of the controlled system
-and apply in turn each of the three controls. Naturally, we reuse the same
-realization of the underlying stochastic process, $W_t$, for each of the
-three different controls.
-
-We set $\e = 0.001$ and $\amax = 2.0$. As such the energy cost is of secondary
-importance and the paramount effect on the objective is the difference $(\ts -
-\T)$, which we are trying to minimize.
-
-The performance of the three controllers for each parameter regime is given in
-\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch} and
-\cref{fig:error_histograms_det_vs_openloop_vs_stoch}. Naturally, the closed-loop
-achieves a lower error than the open-loop, and the naive controller fares worst.
-The difference in performance mostly depends on the strength of the noise. Thus,
-for low noise, the performance of the stochastic controllers, be they open- or
-closed-loop is not much superior to the naive deterministic controller. On the
-contrary - in the high-noise regime, using a stochastic controller gives a much
-lower error on average between the desired $\T$ and the realized $\ts$.
-
-For illustration sake, we also visualize some trajectories from the
-Sub-Threshold-High-Noise regime in \cref{fig:control_trajectories_examples},
-see panels A),C),E).
-
-\begin{table}[h]  
-\begin{center}  
-\subfloat[Supra-Threshold Low-Noise]{
-   \input{Figs/ControlSimulator/MeanSquaredErrors__SUPT_ln.txt}}\\ 
-\subfloat[Supra-Threshold High-Noise]{   
-\input{Figs/ControlSimulator/MeanSquaredErrors__SUPT_HN.txt}}\\
- \subfloat[Sub-Threshold Low-Noise]{
- \input{Figs/ControlSimulator/MeanSquaredErrors__subt_ln.txt}}\\ 
- \subfloat[Sub-Threshold High-Noise]{
- \input{Figs/ControlSimulator/MeanSquaredErrors__subt_HN.txt}}\\
-%  \begin{tabular}{lcc}
-% Control Law &  Squared Error &  Squared Error \\
-%  & Empirical & Theoretical\\% \hline 
-% Deterministic &  0.621 & - \\
-% Open-Loop Stochastic & 0.356 & 0.382\\
-% Feedback Stochastic &  0.288& 0.285\\
-% \hline % \end{tabular} - 
-\caption{
-Realized and theoretical performance of the different control laws. The
-empirical performance is obtained using $N=10000$ sample paths. The theoretical
-performance is found using the optimal value for $J$ for the open-loop
-stochastic control and the value function $\v(x=0, t =0)$ for the closed-loop
-stochastic control.}
-\label{tab:realized_avg_errors_det_vs_openloop_vs_stoch} 
-\end{center}
-\end{table}
-
-\begin{figure}[h]
-\begin{center}
-\subfloat[Supra-Threshold Low-Noise]{
-  \includegraphics[width=.99\textwidth]
-  {Figs/ControlSimulator/Regimes_SUPT_ln_errors_hist.pdf}
-}
-\\
-\subfloat[Supra-Threshold High-Noise]{
-  \includegraphics[width=.99\textwidth]
-  {Figs/ControlSimulator/Regimes_SUPT_HN_errors_hist.pdf}
-  }
-  \\
-\subfloat[Sub-Threshold Low-Noise]{
-  \includegraphics[width=.99\textwidth]
-  {Figs/ControlSimulator/Regimes_subt_ln_errors_hist.pdf}
-}
-\\
-\subfloat[Sub-Threshold High-Noise]{
-  \includegraphics[width=.99\textwidth]
-  {Figs/ControlSimulator/Regimes_subt_HN_errors_hist.pdf}
-}\\
-  \caption[labelInTOC]{Histogram of the spike timing error for the
-  deterministic (left) vs. open-loop stochastic(centre) vs. closed-loop
-  stochastic (right) control laws. This is for the same problem as in
-  \cref{fig:control_trajectories_examples}.
-  We have used $N=10000$ sample paths
-  to form the statistics}
-  \label{fig:error_histograms_det_vs_openloop_vs_stoch}
-\end{center}
-\end{figure}
-
-% \begin{figure}[h]
-% \begin{center}
-% \subfloat[]
-% {
-% \label{fig:controlled_traj_ex1}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/SubTHighNoise_Traj0.pdf}
-% }
-% \subfloat[ ]
-% {
-% \label{fig:controlled_traj_ex2}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/SubTHighNoise_Traj4.pdf}
-% }
-% \subfloat[ ]
-% {
-% \label{fig:controlled_traj_ex3}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/SubTHighNoise_Traj7.pdf}
-% }
-% \caption[]{Examples for the controlled trajectories using the
-% deterministic, open-loop stochastic and closed-loop stochastic control
-% approaches. The black vertical line in the plots indicates the desired spike-time, $\T$.
-% For the parameter values are $\m, \tc, \b = [.2, .5,  1.5]$ (Sub-Threshold
-% High-Noise regime). $\T = 1.5, \eps = 0.001$. The three panels from left to
-% right are three different realizations of the model dynamics. On the upper
-% plots, we show the voltage evolution, $X_t$, on the lower plots we show the
-% applied control, $\a(t)$. Note that $\a(t)$ obtained from the
-% deterministic and open-loop controls are the same for all three samples.}
-% \label{fig:control_trajectories_examples}
-% \end{center}
-% \end{figure}
-
-\subsection{Multi-Spike Control}
-We now turn to the ultimate goal of our analysis, the control of {\sl spike
-trains}. In particular, we will generate $M=10$ realizations of $N=20$ spikes
-each in an attempt to meet a prescribed spike train. We will focus on two
-regimes, the sub-threshold regime with either low or high noise.
-The results
-for parameters from the Sub-Threshold, Low-Noise regime are shown in
-results in \cref{fig:targettrain_cl_lownoise,fig:targettrain_ol_lownoise}.
-While, the results
-for parameters from the Sub-Threshold, High-Noise regime are shown in
-results in
-\cref{fig:targettrain_cl_SUPT_HN,fig:targettrain_cl_SUPT_ln,fig:targettrain_ol_SUPT_ln,fig:targettrain_ol_SUPT_HN}.
-%\usepackage{graphics} is needed for \includegraphics
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%% SUPER-THRESHOLD REGIME LOW-NOISe:
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_ln_cl_trains_sim_10.pdf}
-  \caption[ ]{How good the {\sl closed-loop} controller is at hitting a target
-  spike train. Uses the parameters from the Super-Threshold,Low-Noise regime.
-  The target spike train was generated using the model itself, with
-  parameters from the Super-Threshold, High-Noise regime, without an applied
-  control, $\a=0$.
-  There are $N=16$ spikes and $M=10$ realizations of the control strategy} 
-  \label{fig:targettrain_cl_SUPT_ln}
-\end{center}
-\end{figure}  
-\begin{figure}[htp] 
-\begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_ln_ol_trains_sim_10.pdf}
-  \caption[ ]{How good the {\sl open-loop} controller is at hitting a target
-  spike train. Uses the parameters from the Super-Threshold,Low-Noise regime.
-  The target spike train was generated using the model itself, with
-  parameters from the Super-Threshold, High-Noise regime, without an applied
-  control, $\a=0$.
-  There are $N=16$ spikes and $M=10$ realizations of the control strategy}
-  \label{fig:targettrain_ol_SUPT_ln}
-\end{center} 
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%% SUPER-THRESHOLD REGIME HIGH-NOISe:
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_HN_cl_trains_sim_10.pdf}
-  \caption[ ]{How good the {\sl closed-loop} controller is at hitting a target
-  spike train. Uses the parameters from the Super-Threshold,High-Noise regime.
-  The target spike train was generated using the model itself, with
-  parameters from the Super-Threshold, High-Noise regime, without an applied
-  control, $\a=0$.
-  There are $N=16$ spikes and $M=10$ realizations of the control strategy}
-  \label{fig:targettrain_cl_SUPT_HN}
-\end{center} 
-\end{figure}
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_HN_ol_trains_sim_10.pdf}
-  \caption[ ]{How good the {\sl open-loop} controller is at hitting a target
-  spike train. Uses the parameters from the Super-Threshold,High-Noise regime.
-  The target spike train was generated using the model itself, with
-  parameters from the Super-Threshold, High-Noise regime, without an applied
-  control, $\a=0$.
-  There are $N=16$ spikes and $M=10$ realizations of the control strategy}
-  \label{fig:targettrain_ol_SUPT_HN}    
-\end{center}
-\end{figure}
-
-\clearpage
-
-\section{Miscleaneous Analysis???}
-TODO: WE DON''T KNOW WHAT TO CALL THIS SECTION!
-\subsection{Effect of energy penalty}
-\label{sec:effect_of_eps}
-So far we have assumed a low value for the energy penalty parameter,
-$\eps$, in the objective defined in \cref{eq:OC_LS_variance_energy}. Our
-approach has been that we are most concerned with the accurate spiking and that
-penalizing energy is rather intended to regularize the control and avoid
-excessive chattering of the control between its extreme values, than as a goal
-in minimizing energy in its own right. Now, we take some time to explore the
-effect of a higher $\eps$ on the optimal controls. Intuitively, a
-higher value of $\eps$ will tend to bring the optimal control, $\a^*$, closer
-to zero. We make a comparison for the four regimes for both the
-closed-loop control and the open-loop control in
-\cref{fig:HJB_4regimes_control_different_eps,fig:FBK_Regimes_cs_different_es}.
-Indeed, for both the closed-loop and open-loop optimal controls, increasing
-$\eps$ tends to reduce the absolute value of $\a^*(t)$. In particular, for the
-closed-loop control it tends to broaden the area of transition for
-decreasing $x$ when the control swings from its minimal, i.e.\ inhibitory value,
-to its maximal, i.e.\ excitatory value. Similarly for the open-loop control,
-instead of banging from its minimal bound at the beginning of the interval to
-its maximal bound at the end, the optimal control for $\e = 0.1$ tends to have a
-more mild transition from a slightly inhibitory to slightly excitatory
-values.
-
-Note however that increasing $\eps$ has an important effect on the validity of
-the approximation, used to form the terminal conditions for
-the value function and the adjoint variable, $\v,p$. We assumed that applying
-$\amax$ is optimal for $t>\T$. This is indeed true in the
-limit $\eps \ra 0$. However, even for a finite value of $\eps$, this may still
-be the optimal thing to do. Indeed, in the original calculations,
-where $\eps=0.001$, see the panels on the left in
-\cref{fig:HJB_4regimes_control_different_eps,fig:FBK_Regimes_cs_different_es},
-the so-obtained value function {\sl implied} that $\a(x,\T) = \amax$.
-I.e.\ our
-guess that $\a(x, t) = \amax$ for $t > \t$, is self-consistent with the
-so-obtained value function. This is akin to confirming an ansatz.
-For a higher value of $\eps$, like $\eps = 0.1$, this is no longer the case. The red
-curves on the right-side of \cref{fig:HJB_4regimes_control_different_eps} are no
-longer pushed up at $\a = \amax$, and as such it is no longer valid to
-assume that the value function can be obtained by assuming $\a(t>\T) = \amax$
-and then just calculating the expected remaining time-to-spike.
-The exact same conclusion can be drawn from the open-loop control. There the
-optimal control always (almost) reaches its maximum, $\amax$ while choosing
-$\eps = 0.001$, but no longer so for the higher $\eps = 0.1$.
-In fact, as we mentioned after \cref{eq:valuefun_TCs}, the correct thing to do
-for higher values of $\eps$, is to push the
-calculation interval to some $t^+ > \T$, apply the same terminal condition at
-$t^+$ instead of $\T$ and then solve backwards (either for the closed- or
-open-loop control). One will be guaranteed to find some $t_f > \T$ that works
-since eventually the quadratic term $(t-\T)^2$ will dominate the energy term in
-the objective, which is linear in $t$. We do not explore this further here.
-% CLOSED_LOOP EPS:
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/HJB/Regimes_eps_comparison.pdf}
-  \caption[labelInTOC]{Effect of $\e$ on the closed-loop control. Snapshots of
-  $\a(x,t)$ for $t$ fixed, at the start (blue), mid-point (green) and end-point
-  (red), for each of the four regimes and different values of $\eps$.
-  On the left, we use $\eps = 0.001$ and on the right $\eps = 0.1$.
-  The desired spike time is set to
-  $\T=1.5$ and the bounds are $\a \in [-2,2]$. 
-  A,B)
-   Supra-Threshold-Low-Noise 
-  C,D) 
-   Supra-Threshold-High-Noise 
-  E,F)
-   Sub-Threshold-Low-Noise 
-  G,H)
-   Sub-Threshold-High-Noise}
-\label{fig:HJB_4regimes_control_different_eps} 
-\end{center}
-\end{figure}
-% OPEN_LOOP EPS:
-\begin{figure}[htp]
-\begin{center}
-  \includegraphics[width=.99\textwidth]{Figs/FP_Adjoint/Regimes_eps_comparison.pdf}
-  \caption[labelInTOC]{Effect of $\e$ on the open-loop control.
-  The deterministic time-trajectory of $\a(t)$ for each
-  parameter regime as function of time, $t \in [0, \T]$.
-  On the left, we use $\eps = 0.001$ and on the right $\eps =0.1$.
-  The desired spike time is set to $\T=1.5$ and the bounds are $\a \in [-2,2]$.
-    A,B)
-   Supra-Threshold-Low-Noise
-  C,D) 
-   Supra-Threshold-High-Noise
-  E,F)
-   Sub-Threshold-Low-Noise
-  G,H)
-   Sub-Threshold-High-Noise}
-  \label{fig:FBK_Regimes_cs_different_es} 
-\end{center}
-\end{figure}
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=0.99\textwidth]
-{Figs/ControlSimulator/Composite_Traj.pdf}
-\caption[]{Three examples for the controlled trajectories using the
-deterministic, open-loop stochastic and closed-loop stochastic control approaches.
-The plots on the left are obtained using a value of $\eps=0.001$, while the
-plots on the right use $\eps=0.1$.
-The black vertical line in the plots indicates the desired spike-time, $\T$.
-The parameter values are $\m, \tc, \b = [0.2, 0.5,  1.5]$ (Sub-Threshold
-High-Noise regime), with $\T = 1.5$. 
-On the upper plots, we show the voltage evolution, $X_t$, on the lower plots we
-show the applied control, $\a(t)$. Note that the optimal control, $\a(t)$, obtained from
-the deterministic and from the open-loop controls are the same across all three
-samples.}
-\label{fig:control_trajectories_examples}
-\end{center} 
-\end{figure}
-
-% 
-% \begin{figure}[h]
-% \begin{center}
-% \subfloat[]
-% {
-% \label{fig:controlled_traj_ex1}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/HighEps_Traj5.pdf}
-% }
-% \subfloat[ ]
-% {
-% \label{fig:controlled_traj_ex2}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/HighEps_Traj6.pdf}
-% }
-% \subfloat[ ]
-% {
-% \label{fig:controlled_traj_ex3}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/HighEps_Traj0.pdf}
-% }
-% \caption[]{The effect of higher energy penalty, $\eps = 0.1$.
-% Examples for the controlled trajectories using the deterministic,
-% open-loop stochastic and closed-loop stochastic control approaches. The black vertical line in the plots indicates the desired
-% spike-time, $\T$.
-% The parameter values are $\m, \tc, \b = [0.2, 0.5,  1.5]$ (Sub-Threshold
-% High-Noise regime). $\T = 1.5, \eps = 0.1$. The three panels from left to
-% right are three different realizations of the model dynamics. On the upper
-% plots, we show the voltage evolution, $X_t$, on the lower plots we show the
-% applied control, $\a(t)$. Note that the optimal control, $\a(t)$, obtained from
-% the deterministic and from the open-loop controls are the same across all three
-% samples.}
-% \label{fig:control_trajectories_examples_high_epsilon}
-% \end{center}
-% \end{figure}
-
-\subsection{Control Robustness}
-We have so far assumed that we know the parameters, $\m, \tc, \b$, determining
-the dynamics in \cref{eq:X_evolution_uo,eq:current_mu_alpha} exactly. That is we
-assume that the values for $\tc, \b$ used to obtain the controls are the same as the ones that drive the
-controlled dynamics. That is slightly naive, as in any realistic application
-these will have to be estimated and the estimates will have some error
-associated with them.
-Note that misestimating $\mu$ has the most immediate and clear effect on the
-control policy as that amounts to a direct shift of the control and so for now,
-let us assume that $\mu$ is estimated correctly.
-
-Let us then discuss one of the regimes, the Sub-Threshold,
-High-Noise regime, and see what is the impact of misspecifying the parameters.
-
-Assume that $\b$ and $\tc$ are underestimated, so that $\b_{est} = \b / 1.5$
-and $\tc_{est} = \tc / 1.5$.
-
-So we will have:
-$$\mu, \tc_{est}, \b_{est} = \{.2, .5, 1.5 \}$$ which is the control
-calculated already for the Sub-Threshold,
-High-Noise regime, while the actual simulated dynamics will be driven by
-$$\mu, \tc, \b = \{.2, .75, 2.25 \}$$
-
-We show the effect in
-\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch_misspecified}. Note that
-a misspecification of $\b$ should not have a direct influence on the behaviour
-of the 'deterministic' controller since it already assumes in any case that $\b=0$.
-\begin{table}[h]
-\begin{center}
- \subfloat[Sub-Threshold High-Noise]{
- \input{Figs/ControlSimulator/Misspec_MeanSquaredErrors__subt_HN.txt}}\\
- \caption{The effect of misspecifying the parameters. On the left, the
- system parameters and the parameters used to obtain the control are the
- same, i.e.\ accurate, on the right, they are misspecified.}
-\label{tab:realized_avg_errors_det_vs_openloop_vs_stoch_misspecified}
-\end{center}
-\end{table}
-
-\section{Discussion}
-We have analyzed and computed the optimal control of spikes in two scenarios -
-one where the underlying voltage of the neuron is observable to the controller
-(closed-loop control) and one where only the spikes are observable (open-loop
-control).
-
-Naturally, the ability to control the system is most clearly affected by the
-level of the noise. In a sense, the noise acts like an adversary - in our
-context it has no beneficial role, but to obstruct precise spiking.
-
-The bias current has a more helpful role at least in our examples, where a
-high value of the bias tends to help precise spiking. Of course, where it helps
-or hinders depends on the value of the desired spike time - a combination of a
-high positive bias and a 'distant' spike time, will tend to be difficult to
-control as the system will naturally tend to spike earlier than desired.
-
-In both cases finding the optimal control is computationally intensive as we
-numerically solve partial differential equations. If these algorithms were to be
-practical in an online setting, some more work would have to be done in order to
-ensure they can be computed very quickly (in the millisecond range) or that
-they can be pre-computed.
-
-Our work has strayed pretty close to the paradigm of \cite{Ahmadian2011} in
-trying to make the neuron spike at a particular time. Other goals have been
-formulated, most notably the idea of desynchronizing a population of neurons,
-\cite{Nabi2011}, where it is not so important when the neurons spike as long as
-they do not spike at the same time. We have also incorporated into our objective
-the minimization of total energy used to achieve our goal, which as discussed in
-\cite{Ahmadian2011} is sensible given the potential damage on the cell of
-accumulating charge. Furthermore, we have assumed that our control is
-constrained in magnitude, which is natural given physical limitations on
-equipment and safety considerations. Another possible constraint on the control
-is that it is charge-balanced, meaning that its time integral is zero, $\int u
-\intd{t} = 0$. Such constraints are most easily posed in the context of
-deterministic spiking models like the Phase-Response Curve, see e.g. Danzl et
-al. \cite{Danzl2010}. It is possible to add them to the open-loop control, since
-the control is still deterministic, but it is not-trivial. It is even less
-trivial when one uses Dynamic Programing. In
-particular, insisting on charge-balance in our schemes will make it much more
-difficult to apply the Terminal Conditions used in deriving both the closed- and
-open-loop controls.
-
-The novelty in our work is primarily in obtaining a stochastic control scheme,
-which is not limited by the intensity of the noise in the system. In fact, by
-comparison with the naive 'deterministic' controller, we have shown how wrong
-the control strategy can be if we assume that the noise is negligible. Given
-this scope, the numerical demands on our scheme are non-trivial, but our
-experience and experiments show that these demands are potentially feasible. It
-however remains to be seen whether they can be successfully applied in the lab
-and beyond.
-
-\appendix
-\section{Calculating the Terminal Conditions for the HJB equation}
-\label{sec:valuefun_TCs}
-Here we explain how to calculate the first two moments of the time to-spike,
-$\Exp \big[\trem^p \,\Big|\, X_\T = x, \a(t) = \amax \big].$, which is needed in setting the terminal conditions for both,
-$\v(x,\T)$ and $p(x,\T)$, see \cref{eq:OC_LS_HJB_full} and \cref{eq:adjoint_pde_OU}.
-
-Let
-\begin{equation}
-\begin{array}{lcl}
-\Tone(x_0) &=& \Exp \Big[\trem \,\Big|\, X_\T = x, \a(t) = \amax
-\Big]
-\\
-\Ttwo(x_0) &=&
-\Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big],
-\end{array}
-\end{equation}
-where as before, $\t$ is the remaining time-to-spike in
-\cref{eq:X_evolution_uo}, at the desired spike-time $t=\T$, given $X_\T = x_\T$
-and $\a(t) = \amax$.
-Naturally, the $\Ti$'s depend on $x_\T$ and are affected by the value of the
-various parameters, $\a,\b,\tc$.
-
-Actually, we only  need $\Ttwo$ for the TCs in \cref{eq:OC_LS_HJB_full} and \cref{eq:adjoint_pde_OU}.
-but $\Tone$ is needed as an auxiliary.
-
-Indeed, analytical expressions exist for $\Ti$, e.g.\ see \cite{Inoue1995}.
-However in our experience, they are awkward to work with numerically, especially
-as they involve infinite alternating sums which can easily overflow in numeric
-calculations. Moreover, they are not particularly convenient if you need $\Ti
-(x_0)$ for a wide range of $x_0$'s as we do here, since they calculate the value
-independently for each $x$. Instead we will follow and slightly extend Jacobs\footnote{Jacobs only discusses how to calculate the first moment
-$\Tone$, however it is quite clear from his discussion how to proceed
-recursively to obtain any higher moment $\Ti$, by using previously calculated
-lower moments. This is what we do here. Moreover the relation between mean exit
-times and boundary value problems is discussed in most
-introductory books on stochastic processes, see \cite{Oksendal2007} for
-example.}, \cite{Jacobs}, in deriving and solving a differential equation for
-$\Ti(x_0)$.
-
-Basically we are calculating the mean-time-squared to exit an interval $[\xmin,
-\xth]$. Theoretically, $\xmin = -\infty$, but for our purposes we will set it to
-some finite value and impose reflecting boundaries there - this is a very good
-approximation for $\xmin \ll 0$, as long as $\a = \amax >0$, which is what we
-have.
-
-Let $B(y,t|x) = \Prob[X_0 = y| X_t = x]$ for a fixed $x$. Then $B$ solves the
-following (backward Kolmogorov) PDE
-\begin{equation}
--\di_t B = (\a - \frac y\tc)\di_yB - \frac {\b^2} 2 \di_y^2 B
-\label{eq:backward_Xdensity}
-\end{equation}
-
-Let $\G(t,y)$ be the survival function for $T$ for $X_0 = y$.
-$$\G(t,y) = \Prob[T>t | X_0 = y] = \Prob[X_{s\leq t} \in [\xmin,
-\xth] | X_0 = y]$$.
-
-Because the dynamics do not depend explicitly on time, we have:
-$$
-\G(t,y) = \int_\xmin^\xth B(y,t|x) \intd{x}
-$$
-and so $\G$, being a definite integral of $B$ satisfies the same PDE:
-\begin{equation}
--\di_t \G = (\a - \frac y\tc) \di_y\G - \frac {\b^2} 2 \di_y^2 \G
-\label{eq:backward_SDF}
-\end{equation}
-
-With the BCs, ICs:
-\begin{equation}
-\begin{cases}
-\G(t,\xth) = .0 & \quad (T = t \implies \Prob[T>t] = 0)
-\\
-\di_x \G(t,\xmin) = .0  & \quad (\textrm{due to reflecting boundary for X})
-\\
-\G(0, y) = 1. & \quad (T \textrm{ is definitely} > 0 \textrm{ for } x_0 \in
-(\xmin, \xth) )
-\end{cases}
-\end{equation}
-
-If we solve the PDE for $\G$ over $t \in [0,\infty)$, then we can calculate
-$\Ttwo(x_0)$ as:
-\begin{eqnarray}
-\Ttwo(x_0) &=& \int_0^\infty t^2 \cdot (-\di_t\G(t,x_0)) \intd{t}
-\\
-		   &=& \int_0^\infty 2t \cdot \G(t,x_0) \intd{t}
-\end{eqnarray}
-assuming it exists.
-
-Since $\Ttwo$ is an integral over time, we can reduce the PDE in
-\cref{eq:backward_SDF} for the distribution to a BVP for the moment(s) by
-integrating over time. Here is how this is done: multiply both sides  of
-\cref{eq:backward_SDF} by $2t$, and integrate with respect to time:
-\begin{equation}
-\int_0^\infty 2t (-\di_t \G) \intd{t}
-=
-\int_0^\infty  2t \left[ (\a - \frac y\tc) \di_y\G - \frac {\b^2} 2
-\di_y^2 \G\right]
-\intd{t}
-\end{equation}
-The LHS turns out to be $2\Tone$, so
-\begin{equation}
-2\Tone
-=(\a - \frac y\tc)   \di_y \Ttwo
-- \frac {\b^2} 2
-\di_y^2 \Ttwo
-\label{eq:BVP_Ttwo}
-\end{equation}
-Similarly we can derive an equation for $\Tone$, namely:
-\begin{equation}
-1
-=(\a - \frac y\tc)   \di_y \Tone
-- \frac {\b^2} 2
-\di_y^2 \Tone
-\label{eq:BVP_Tone}
-\end{equation}
-In both cases, the BCs will be:
-\begin{equation}
-\begin{cases}
-\Ti(\xth) = .0 & \quad (\textrm{we are already spking})
-\\
-\di_x \Ti(\xmin) = .0  & \quad (\textrm{due to reflecting boundary for }X)
-\end{cases}
-\label{eq:BVP_Ti_BCs}
-\end{equation}
-Equations \ref{eq:BVP_Tone} and \ref{eq:BVP_Ttwo} can be reduced to first order
-differential equations and solved analytically, at least up to some definite integral,
-but the analytical solution is not particularly useful or necessary for
-numerical calculations. Instead, we opt to calculate the solutions to
-\ref{eq:BVP_Tone} and \ref{eq:BVP_Ttwo} directly via an ODE solver, starting
-from $T_i(\xth) = .0$ and integrating down for $x  \in [\xmin, \xth]$.
-That is how we have obtained the terminal conditions for both $\v(x,\T)$ and
-$p(x,\T)$ in our calculations above.
-
-\bibliographystyle{plain}
-\bibliographystyle{unsrt}
-\bibliographystyle{bmc_article}
-\section*{Acknowledgements}
-   \ifthenelse{\boolean{publ}}{\small}{}
-SD is supported by the Danish Council for Independent Research $\mid$ 
-Natural Sciences.
-\\ 
-AL acknowledges support from NSERC Canada.  
-\\
-AI is supported by a PGS-D scholarship from NSERC Canada.
-
-\section*{References}
-\bibliographystyle{iopartnum}
-\bibliography{library,local}
-
 \end{document}

### 2014-02-18 draft.tex
Adding ML Note

@@
 standard deviations below the mean of the stationary distribution of the
-maximally inhibited, $\a=\amin$, \cref{eq:current_mu_alpha}. That is we set
-$\xmin = \tc(\mu + \amin) - 2 \b / \sqrt{\tc/2}$. We further enforce
+maximally inhibited process, i.e.\ setting $\a=\amin$ in
+\cref{eq:current_mu_alpha}. That is we set $\xmin = \tc(\mu + \amin) - 2 \b / \sqrt{\tc/2}$. We further enforce
 that $\xmin \leq -0.5$.
@@
   Note that the green and blue
-  curves in F) are lying on top of each other for all $x$.
+  curves in D) are lying on top of each other for all $x$.
   \\
@@
   \includegraphics[width=\textwidth]{Figs/HJB/Regimes_eps_comparison.pdf}
-  \caption[labelInTOC]{Effect of $\e$ on the closed-loop control. Snapshots of
-  $\a(x,t)$ for $t$ fixed, at the start (blue), mid-point (green) and end-point
+  \caption[labelInTOC]{Effect of energy parameter $\e$ on the closed-loop
+  control. Snapshots of $\a(x,t)$ for $t$ fixed, at the start (blue), mid-point (green) and end-point
   (red), for each of the four regimes and different values of $\eps$.

### 2014-02-18 floats.tex
Adding ML Note

--- 2013-09-02 draft.tex
@@
-\documentclass[12pt]{iopart}
-%\documentclass{article}
-
-\usepackage{amsmath}
-\usepackage{amscd}
-\usepackage{amssymb} 
-\usepackage{amsfonts} 
-\usepackage{amsthm}
-\usepackage{amsfonts}
-\usepackage{amsthm}
-
-\usepackage{tikz}
-\usetikzlibrary{arrows,snakes,backgrounds}
-% \usepackage{circuitikz}
-% \usepackage{pgf}
-% \usetikz
-\usepackage{subfig}
-
-\usepackage{algpseudocode}
-\usepackage{algorithm}
-
-\usepackage[super]{nth}
-\usepackage{appendix}
-% \usepackage{listings}
-\usepackage{color}
-
-\usepackage{hyperref}
-%\usepackage{url}
-
-\usepackage{cleveref}
-\usepackage{cancel}
-\usepackage{slashbox}
-
-\usepackage{aviolov_style}
-\usepackage{local_style}
-
-\usepackage{iopams}
-
-% \newtheorem{thm}{Theorem}[section]
-% \newtheorem{lemma}{Theorem}[thm]
-% \theoremstyle{definition}
-% \newtheorem{ex}{Example}[thm]
-% \newtheorem{defn}{Definition}[thm]
-
-% \def\includegraphic{}
-% \def\includegraphics{}
-
-\begin{document}
-
-\title{Stochastic Optimal Control of Single Neuron Spike Trains}
-\author{Alexandre Iolov$^1$, Susanne Ditlevsen$^2$, Andr\'e Longtin$^1$
-		}	
-\address{%
-     (1) Department of Mathematics and Statistics, University of Ottawa, Ottawa,
-    Canada
-\\
-    (2) Department of Mathematical Sciences, University of Copenhagen,
-Copenhagen, Denmark
-}
-\ead{aiolo040@uottawa.ca}			
-% \address{Department of Mathematics and Statistics, University of Ottawa, Ottawa, Canada 
-% \\ Department of Mathematical Sciences, University of Copenhagen, Copenhagen,
-% Denmark }
-% \ead{$<$\href{mailto:aiolo040@uottawa.ca}
-% 		{aiolo040 at uottawa dot ca}$>$}
-
-% \date{\today}
-
-\begin{abstract}
-
-{\em Objective.} External control of exact spike timings in single neurons is
-important for basic understanding of neuronal activity, and has potential in
-brain-machine interfaces, for improving neural prostheses, and in medical
-applications related to neuronal disorders, such as Parkinson's disease or
-epilepsy. The goal of this paper is the design of optimal controlled electrical
-stimulation patterns to a cell to achieve a target spike train under
-physiological constraints not to damage tissue and minimize possible side
-effects. {\em Approach.} We pose an optimal control problem to precisely specify
-the spike timing in a noisy leaky integrate-and-fire model of a neuron. In
-particular, we allow for the noise to be of arbitrary intensity. The optimal
-control problem is solved using Dynamic Programming if the controller has access
-to the voltage (closed-loop control) and using a Maximum Principle for the
-transition density if the controller only has access to the spike timings (open
-loop control). {\em Main results.} We have developed a stochastic optimal
-control scheme to obtain precise spike timings, which is not limited by the
-intensity of the noise in the system. Given this scope, the numerical demands
-are non-trivial, but simulations show that the algorithms are potentially
-feasible. It however remains to see if they can be successfully applied in a
-real-time experimental setting. Furthermore, we show how wrong the control
-strategy can perform if we assume that the noise is negligible. {\em
-Significance.} More precise targeting and control of neural activity to achieve
-optimal performance in a noisy environment, in which most neurons are embedded,
-have great potential in health care. The main contribution is the online
-feedback control of a noisy system through modulation of the input, taking into
-account physiological constraints on the control, as well as incurred costs to
-diminish adverse effects.
-
-\end{abstract}
-% \tableofcontents MSC KEYS: 49N90 90C39 93E20 93C20 93C95
-
-\maketitle
-
-\tableofcontents
-
-\section{Introduction}
-Manipulation of individual neurons through external electrical
-stimulation provides a means to
-controlling the spiking activity. Applications may arise, for
-example, in brain-machine interfaces or in artificial prosthetics.
-A natural goal is to make a cell produce a pre-specified spike train,
-under certain constraints on the modulation of the input.
-We consider this problem in the framework of Stochastic Optimal Control and
-give both a feedback solution (closed-loop), when the cell voltage is
-explicitly observable, as well as an open-loop solution when only the
-occurrence of spikes is observable. Importantly, we allow for an
-arbitrary noise intensity.
-
-Both theoretically and in practice related problems have been
-addressed in the literature. One objective has been to
-obtain either minimum or maximum interspike intervals lengths,
-when the input is constrained to be between some prespecified upper
-and lower bounds, see e.g. \cite{Lee1994,Lefebvre1987} for a
-mathematical treatment, or \cite{Danzl2009,Nabi2012,Wu2009} in a neuronal
-context. Another objective is to break a pathological synchronous
-firing pattern in clusters of neurons, highly relevant for neurological
-disorders such as epilepsy and Parkinson's disease,
-\cite{Nabi2013a,Nabi2011}, see also \cite{Feng2007b}.
-
-Our objective, namely targeting exact spike times in single neurons, has been
-considered mainly in the open-loop context, and in either absence of or for
-small noise. In \cite{Ahmadian2011} they use the Spike Response Model,
-\cite{Gerstner2002}, to control output target spike trains and implement their
-scheme on pyramidal cells in mouse cortical slices. Their method is numerically
-efficient and allows for the simultaneous control of many neurons. However the
-efficiency of the numerics strongly relies on the assumption that the diffusion
-noise, the value of $\b$ in \cref{eq:X_evolution_uo} below, is small enough, see
-the discussion above and below equation 12 in \cite{Ahmadian2011}. They only
-work with open-loop control. The difference between the objective in
-\cite{Ahmadian2011} and what we consider is that they maximize the probability
-of spiking at some given time $t^*$, whereas we minimize the mean squared
-difference between the realized spike time, $T_{sp}$ and the desired, $t^*$.
-Moehlis et al. \cite{Moehlis2006} work with a Phase Response Model to obtain
-spikes at exact times, while keeping the root mean square of the input to a
-minimum. A similar approach is taken in \cite{Dasanayake2011}. They do not
-consider noise, though, and only work with open-loop control. In
-\cite{Nabi2013}, the methods are implemented on brain slices of pyramidal
-neurons of rat hippocampus. While the
-Phase Response Curve Model is a parsimonious and effective way to describe a
-neuron's response to a stimulus, it is only valid in the supra-threshold regime,
-where the unstimulated neuron is periodically spiking. In \cite{Feng2003} they
-use a leaky integrate-and-fire model and open-loop control. To
-obtain a given spike time, they choose an objective of minimizing the
-variance of the membrane potential at the desired spike time, while forcing the mean of
-the membrane potential at this time point to equal the threshold. This provides
-exact solutions since it does not involve first-passage times, but has the
-drawback that there is a non-negligible probability that the obtained
-spike time will be far from the desired spike time.
-
-Our objective of imposing a certain timing sequence of the spike train using an
-externally applied control is obtained in both the closed- and
-open-loop setting, and we specifically include the noise in the
-calculations of the controls. We restrict the controlled input to stay
-within pre-specified bounds, and also include a cost function to
-minimize intervention.
-Our main contributions are that we specifically allow for a non-negligible
-noise component in the neural activity when calculating the control, and
-consider both open- and closed-loop controls. In particular we do not
-restrict our attention to the autonomously-spiking, supra-threshold regime.
-
-The paper is structured as follows: First we describe the neuron model and
-formalize the control objective. Then we describe a feedback-based solution,
-which assumes that the controller has detailed access to the voltage trajectory.
-Then we relax the observation assumption so that the controller only has access
-to the spike timings. Finally, we compare the two methods through
-simulations against a simple-minded control technique which ignores
-the stochastic input to the neuron.
-
-\section{Problem Formulation}
-A basic but useful model for the neural membrane potential evolution is the
-noisy leaky-integrate-and-fire model:
-\begin{equation}
-\begin{gathered}
-dX_s = \left(\Iext(t) - \frac{X_s}{\tc} \right) \intd{s} + \b \intd{W_s},
-\\
-X(0) = 0,
-\\
-X(\ts) = \xth \implies
-\begin{cases}
-X(\ts^+) = 0 &
-\end{cases}
-\end{gathered}
-\label{eq:X_evolution_uo}
-\end{equation}
-Here $X$ represents the membrane electric potential which decays to $0$
-with a time constant of $\tc$, $dW$ is a Brownian motion increment scaled by
-$\b$ and $\Iext(t)$ is the deterministic external input to the cell. Having last
-spiked at time $0$, the potential hits $\xth$ at some time $\ts$, the potential
-resets to $0$ and the process starts all over again. 
-We write $\ts^+$ for the voltage's limit from the right at $\ts$.
-Throughout the paper the threshold is set to one, $\xth = 1$.
-
-Suppose that we have some control over the external current such that it can be
-decomposed as
-\begin{equation}
-\Iext(t) = \m + \a(t)
-\label{eq:current_mu_alpha}
-\end{equation}
-where $\m$ is an uncontrollable, but constant part of the external current and
-$\a(t)$ is controllable, i.e.\ it can be chosen to achieve some goal.
-A natural goal is to attempt to control the spike time, $\ts$.
-That is how do we choose $\a(t)$ such that $\ts = \T$.
-A natural optimal control objective to achieve this is the least squares solution
-\begin{equation}
-\a(t) = \argmin_{\a(t)} \{\, \Exp[(\ts- \T)^2] \,\}
-\label{eq:OC_LS_variance}
-\end{equation}
-where expectation is taken with respect to the distribution of the trajectories
-of $X_t$.
-
-Often the control has certain constraints. The most common are simple box
-constraints:
-\begin{equation}
-\a(t) \in [\amin, \amax] \quad \forall t.
-\label{eq:bound_constraints_alpha}
-\end{equation}
-
-% \subsection{Optimal Control Formulation}
-In addition to \cref{eq:OC_LS_variance}, we will also add to our objective a
-running cost based on the control. This regularizes the problem eliminating the
-subtleties of singular-control situations and it serves to avoid excessive
-control as well as to avoid excessive charge building up on the cell.
-
-So we seek an optimal control, $\a^*$, that solves
-\begin{align}
-J[\a(\cdot)] =&
-\Exp\left[
-\e \int_0^\ts  \a^2(s) \intd{s}
-+
-(\ts - \T \big)^2 \right]
-\label{eq:OC_LS_variance_energy}
-\\ \notag
-\a^*(\cdot) =& \argmin_{\a(\cdot)} J[\a(\cdot)].
-\end{align}
-where $\e$ measures how much weight we put on minimizing the energy cost.
-If $\e = 0$, then we do not care at all about the expended cost.
-
-It will often be the case that $\a$ means either a function or a
-value of that function at a particular time. This function could be
-random, i.e.\ a stochastic process. Since $\a$ is used to control the behaviour
-of $X$, it will be at least measurable with respect to the filtration of $X$,
-see \cite{Oksendal2007} for the details. We will try to make clear below when we
-are refering to $\a$ as a stochastic process and when we are merely refering to its particular
-value at some time. For example, in \cref{eq:OC_LS_variance_energy}, $\a(\cdot)$
-refers to the function, while $\a(s)$ refers to that function's value, possibly
-random, at time $s$.
-
-We will consider two control contexts -- {\sl closed-loop} and  {\sl open-loop}
-control. In closed-loop control, the value of $X_t$ at time $t$ is observable
-and can be used in determining the control. In open-loop control only the spike
-times are observable. In the closed-loop context, we write the control as $$\a =
-\a(x,t)$$ to indicate its dependence on $X_t$ and to express that it will be
-updated based on the time-course of $X$. In the open-loop context, we write $$\a
-= \a(t)$$ to indicate that $\a(\cdot)$ is decided for all times at time 0. The
-techniques used to obtain the optimal controls in the two scenarios will be
-different. For the closed-loop scenario we use Dynamic Programing,
-see \cite{Fleming1975}, while for the open-loop scenario we use a form of the
-Maximum Principle applied on the transition density of the controlled process,
-see \cite{Borzi2012}.
-
-Crucially, we assume that the model parameters, $\m,\tc,\b$ in
-\cref{eq:X_evolution_uo} are known. This is a strong assumption and will be
-discussed later.
-
-\subsection{Parameter Regimes}
-Different  parameter regimes can be envisioned given \cref{eq:X_evolution_uo},
-depending on whether the noise intensity, $\b$, is
-relatively high or low, and whether the external, uncontrollable bias current,
-$\m$, induces spikes in the absence of noise or not. 
-Spikes will occur in the absence of noise, if and only
-if $ \m > \frac 1\tc,$ which is called the supra-threshold regime. When $\m
-\leq 1/\tc$, the regime is called the sub-threshold regime.
-In addition we will impose two values of $\b$, which we
-will call high-noise and low-noise, respectively.
-Example values for each parameter regime are given in \cref{tab:regimes}, and we
-visualize a single path for each regime in \cref{fig:regime_path_examples}.
-\begin{table}
-\begin{tabular}{|l||{c}|{c}|}
-\hline
-\backslashbox{$\m$}{$\b$}
-& $1.5$ & $0.3$ \\
-\hline
-$1.5 / \tc $ &Supra-threshold-High-noise & Supra-threshold-Low-noise \\
-\hline
-$0.1 / \tc$   &Sub-threshold-High-noise & Sub-threshold-Low-noise \\
-\hline
-\end{tabular}
-\caption{Regime labels and example values. Note that for the numerical
-experiments below, we use $\tc = 0.5$}
-\label{tab:regimes}
-\end{table}
-%\usepackage{graphics} is needed for \includegraphics
 \begin{figure}[htp] 
@@
   A) Supra-Threshold-Low-Noise  
-  B)Supra-Threshold-High-Noise 
+  B) Supra-Threshold-High-Noise  
   C) Sub-Threshold-Low-Noise 
-  D) Sub-Threshold-High-Noise 
-  }
+  D) Sub-Threshold-High-Noise.  
+  Note the multiple crossings very close together in the high-noise regimes 
+  in B and D. } 
   \label{fig:regime_path_examples} 
@@
 \end{figure} 
-
-\section{Closed-Loop Solution - Dynamic Programing}
-We now detail the Dynamic Programing approach to obtaining a feedback-optimal
-control. What this amounts to is that the controller can be continuously updated
-depending on the realization of the stochastic process, $X_t$.
-
-Given a time $t$ and a realized voltage value $X_t = x$, let $\trem$ be the
-(unknown) remaining time to spike, $\trem = \ts - t$. Starting from an arbitrary
-$t, x$, our remaining-cost objective, $J[\a(\cdot); x,t]$, will be
-\begin{equation}
-J[\a(\cdot); x,t]  =
-\Exp\left[
-\big(\trem - (\T-t) \big)^2
-+
-\e \int_t^\ts  \a^2(s) \intd{s}
-\,\Big|\, X_t = x
-\right]
-\label{eq:OC_LS_cost_to_go}
-\end{equation}
-i.e.\ if time $t$ has elapsed without a spike, we now want to minimize
-the difference between $\t = \ts - t$ and ($\T-t$), given the current state $x$.
-Recall that in the closed-loop scenario, we assume that the value of $X_t$
-is known to the controller. A similar problem is discussed analytically at
-length in the book on optimal control by Peter Whittle, \cite{Whittle1996},
-although there is no discussion there of the numerics required to solve it.
-
-\subsection{Hamilton-Jacobi-Bellman equation}
-The Hamilton-Jacobi-Bellman (HJB) equation,
-see \cite{Fleming1975,Whittle1996} or \cite{Danzl2009,Nabi2013a} for a neuroscience
-context, associated with
-\cref{eq:OC_LS_cost_to_go} is obtained as follows. We introduce the value function, $\v(x,t)$:
-\begin{align}
-\v(x,t) =&
- \min_{\a(\cdot)_{s \geq t}}
- \{J[\a(\cdot); x,t] \}
-\notag
-\\
-=&
-\min_{\a(\cdot)_{s \geq t}}
-\Exp\left[
-\big(\trem - (\T-t) \big)^2
-+
-\e \int_t^\ts  \a^2(s) \intd{s}
-\,\Big|\,X_t = x
-\right]
-\label{eq:OC_LS_mean_value}
-\end{align}
-
-Then $\v$ satisfies the following HJB partial differential equation (PDE):
-\begin{equation}
-\begin{gathered}
-\di_t \v(x,t) + \tfrac{\b^2}{2} \di_x^2 \v(x,t) + \\
-\min_{\a(x,t) \in [\amin,
-\amax]}\Big\{ \e \a^2 (x,t)+ \left(\m + \a(x,t)-\tfrac{x}{\tc}\right) \di_x
-\v(x,t) \Big\} = 0
-\label{eq:OC_LS_mean_HJB}
-\end{gathered}
-\end{equation}
-The special feature of \cref{eq:OC_LS_mean_HJB} in contrast to a generic
-parabolic PDE is that it contains an embedded optimisation that depends on the
-solution, $\v$. For each $x,t$ in the computational domain,
-$\a$ is chosen as to minimize $\{\e \a^2 + \left(\m + \a - \tfrac{x}{\tc}\right) \di_x \v \}$.
-Here we can solve for the optimal control, $\a^*(x,t)$ analytically as:
-\begin{align}
-\a^*(x,t)  =& \argmin_{\a \in [\amin, \amax]}\big\{\e \a^2 + (\m + \a-\tfrac{x}{\tc}) \di_x\v\big\}
-\notag
-\\
-=&
-\min \left(\amax, \max\left(\amin, -\frac{\di_x\v(x,t)}{2\e}\right)\right)
-\label{eq:OC_LS_variance_energy_quadratic_control}
-\end{align}
-With this the HJB PDE becomes
-\begin{equation}
-\di_t \v(x,t) + \tfrac{\b^2}{2} \di_x^2\v+
-\e (\astar(x,t))^2 + (\m + \astar(x,t)-\tfrac{x}{\tc}) \di_x\v
-= 0
-\label{eq:OC_LS_mean_HJB_bounded_control}
-\end{equation}
-
-We need to consider boundary conditions (BCs) for $\v$. If $X_t = \xth$ then we
-have a spike now and $\ts = t$. Thus $$ \v(\xth,t) = (t-\T)^2.$$ At the
-threshold, the value function equals the squared difference between the desired
-spike time and the realized one. For $x \ll 0$, we assume that $\v$ is not
-significantly affected by the change in $x$, i.e.\ that $$ \di_x \v(\xmin, t) =
-0 $$ for some lower boundary $\xmin$. Such a boundary condition will be
-justified if we impose artificial reflecting boundaries for the diffusion at
-$\xmin$. So we need to determine for which $\xmin$ is a reflecting boundary
-consistent with the dynamics. For example, we can take $\xmin$ to be two
-standard deviations below the mean of the stationary distribution of the
-maximally inhibited, $\a=\amin$, \cref{eq:current_mu_alpha}. That is we set
-$\xmin = \tc(\mu + \amin) - 2 \b / \sqrt{\tc/2}$. We further enforce
-that $\xmin \leq -0.5$.
-% Ooops, we need $\di_x\v\neq 0$ and here we are setting $\di_x\v= 0$ in one of
-% the BCs\ldots oh well.
-
-To determine the Terminal Conditions (TCs) for $\v$, the idea is
-simple: if we reach $\T$ without having spiked we apply maximum control in the
-positive direction, i.e.\ $$t>T \implies \a(t) = \amax$$ Thus:
-\begin{equation}\v(x,\T) =
-\Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big].
-\label{eq:valuefun_TCs}
-\end{equation}
-Note that we are making an approximation here
-- we are ignoring the energy term, $\eps u^2$, in the objective for
-$t>T$. Naturally, this approximation is ever more accurate for $\eps \ll 1$.
-We discuss in more detail the validity of the approximation in
-\cref{sec:effect_of_eps}.
-Alternatively, we could impose this approximate terminal condition at some $t^+
-> \T$.
-
-We will see the quantity on the right hand side of \cref{eq:valuefun_TCs}
-repeatedly so we will give it a special name.
-\begin{equation}
-\Ttwo(x) := \Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big].
-\label{eq:Trem_squared}
-\end{equation}
-$\Ttwo$ is the second moment of the remaining time to reach the
-threshold starting at $X_\t = x$ and applying $\amax$ throughout. This quantity can be found
-easily for all $x$ in the domain by solving a stationary backward Kolmogorov
-equation. We show the details in \ref{sec:valuefun_TCs}.
-
-Thus, we restate the HJB equation in its fully specified form:
-\begin{equation}
-\begin{gathered}
-\di_t \v(x,t) + \tfrac{\b^2}{2} \di_x^2\v(x,t)+
-\e \a^2(x,t) + \big(\m + \a(x,t) -\tfrac{x}{\tc}\big)\cdot \di_x\v(x,t)
-= 0
-\\
-\a (x,t) = \min \left(\amax, \max\left(\amin, -\frac{\di_x \v(x,t)}{2\e}\right)
-\right)
-\\
-\begin{cases}
-\v(\xth,t) = (t-\T)^2  \quad & \textrm{upper BC}
-\\
-\di_x \v(\xmin, t)  = 0  \quad &\textrm{lower BC}
-\\
-\v(x,\T)  =
-% \Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big]
-% \Exp[\t^2|x, \amax]  \quad
-\Ttwo(x)
-\,& \textrm{TC}
-\end{cases}
-\end{gathered}
-\label{eq:OC_LS_HJB_full}
-\end{equation}
-We are solving $\v(x,t)$ over the domain $[\xmin, \xth] \times [0,T]$.
-
-\subsection{The numerical method for the HJB equation}
-
-We now have a PDE for $\v$ and an algorithm for computing all the BCs and TCs of
-this PDE. It is time to discuss the numerical method for solving
-\cref{eq:OC_LS_HJB_full}. Since it is one-dimensional, it is straight forward to
-apply the standard centred finite difference using the Crank-Nicholson scheme to
-step in time. We have not found it necessary to upwind the first order
-derivatives for the regimes we have explored. To resolve the non-linearity,
-$(\di_x\v)^2$, in the PDE, we treat it as a mixed implicit-explicit term
-$$(\di_x \v(x,t_k))^2 = \di_x \v(x,t_k) \cdot \di_x \v(x,t_k) \approx
-\underbrace{\di_x \v(x,t_k)}_{\textrm{implicit}} \cdot
-\underbrace{\di_x\v(x,t_{k+1})}_{\textrm{explicit}}.$$ Note that the implicit
-term is in the previous time $t_k$ instead of, as is conventional, the next
-$t_{k+1}$, because we are solving for $\v$ {\sl backwards} in time.
-
-The numerical scheme is implemented in Python, using the Scipy/Numpy library,
-\cite{Scipy}. For the discretization in time and space, we choose $\Delta x$ and
-$\Delta t$ in relation to the parameter regimes. for example in the
-Supra-Threshold Low-Noise regime, we have $\Delta x = 0.0043$ and $\Delta t =
-0.0012$. We consider these to be a fairly fine discretization.
-
-We next demonstrate solutions for the value function and the associated
-control law for a parameter set from each regime in \cref{tab:regimes}.
-
-\subsection{Solutions of the HJB equation}
-We set $\T=1.5$ and $\eps = 0.001$, so that the primary focus is the accurate
-spiking and energy minimization considerations are secondary.
-
-In \cref{fig:HJB_4regimes_value_control_cuts}, we depict three snapshots
-in time of the value function and the control, $\v$ and $\a$, at $t = [0,\T/2,
-\T]$, i.e.\ at the beginning, middle and end of the relevant time interval. Then in
-\cref{fig:HJB_4regimes_value_surf} we show the entire value function, $\v(x,t)$,
-the solution to \cref{eq:OC_LS_HJB_full}, for each of the parameter regimes.
-% THE CUTS:
 \begin{figure}[htp] 
 \begin{center} 
-  \includegraphics[width=.99\textwidth]{Figs/HJB/Regimes_vc_cuts.pdf}
+  \includegraphics[width=\textwidth]{Figs/HJB/Regimes_vc_cuts.pdf} 
   \caption[labelInTOC]{Snapshots of the value function, $\v$, on the left and on 
   the optimal control, $\a$, on the right for $t$ fixed, at the start (blue), 
-  mid-point (green) and end-point (red), for each of the four regimes. The desired spike time is set to $\T=1.5$, the energy penalty, $\eps
-  = 0.001$. The control bounds are $\a \in [-2,2]$. Note that the green and blue
-  curves in E) are lying on top of each other for all $x$.
+  mid-point (green) and end-point (red), for each of the four regimes. 
+  The desired spike time is set to $\T=1.5$ assuming the previous spike was at 
+  $t=0$, the energy penalty, $\eps = 0.001$. The control bounds are $\a \in 
+  [-2,2]$.  
+  Note that all red value function curves go to zero at the threshold to satisfy 
+  the upper BC in \cref{eq:OC_LS_HJB_full}. 
+  Note that the green and blue 
+  curves in F) are lying on top of each other for all $x$. 
   \\ 
-  A,E) Supra-Threshold-Low-Noise
-  B,F) Supra-Threshold-High-Noise
-  C,G) Sub-Threshold-Low-Noise
-  D,H) Sub-Threshold-High-Noise
+  A,B) Supra-Threshold-Low-Noise 
+  C,D) Supra-Threshold-High-Noise 
+  E,F) Sub-Threshold-Low-Noise 
+  G,H) Sub-Threshold-High-Noise 
   } 
@@
 \end{figure} 
-% THE SURFS:
 \begin{figure}[h!] 
@@
 The control bounds are $\a \in [-2,2]$.  
-The contours for fixed $t$ plotted in thick correspond to the graphs in \cref{fig:HJB_4regimes_value_control_cuts}.
+The coloured contours for fixed $t$ plotted in thick correspond to the 
+curves plotted in the left panels of \cref{fig:HJB_4regimes_value_control_cuts} 
+(blue: $t=0$, green: $t=t*/2$, red: $t=t*$). 
 \\ 
@@
 \end{figure} 
-
-Let us discuss the most salient features in
-\cref{fig:HJB_4regimes_value_surf,fig:HJB_4regimes_value_control_cuts}. Recall
-that lower values for the value function, $\v(x,t)$, are preferred, since we are
-minimizing. We see the same general shape in all four space-time surface plots
-of the value function in \cref{fig:HJB_4regimes_value_surf}. At the end of the
-interval, $\v(x,\T)$ is monotonically decreasing in $x$, which is to be expected
-given its terminal condition in \cref{eq:valuefun_TCs}. That is, the lower the
-value of $X_\T$ the more on average will we have to wait for the spike to occur.
-% This same basic shape represents the idea that we would like to be close to
-% the threshold, $x = 1$, near the end of the interval when $t \approx \T$, but
-% we would like to be far from threshold, for $t \ll \T$. Indeed, at the very
-% end of the interval, when $t = \T$, the value function, $v(x,\T)$, is
-% monotonically decreasing in $x$, because the lower the value of $X_\T$ the
-% more time on average will it take us to spike and the more delay will we have
-% from our desired spike time.
-As we go back in time $\v$ inverts, with a clear peak near the upper, threshold
-boundary. That is, for intermediate values of $t$ we have to consider the risk
-of spiking too early, represented by the high value of $\v$ near the upper
-boundary {\sl and} the risk of spiking too late, represented by the high values
-of $\v$ at the lower boundary. As we progress even further back in time to the
-beginning of the interval, the peak near the threshold rises further, since we are spiking
-{\sl earlier}, while the peak at the lower end flattens, since now there is
-enough time to reach threshold despite starting far from it. The full surface
-plots for $\v(x,t)$ in \cref{fig:HJB_4regimes_value_surf} can be thought of as
-the interpolation between these three basic phases.
-
-% Note, that exact shape is generated by Since we have put a very small weight
-% on the energy penalty, $\e = .001$, and we have a significantly high bound on
-% the maximum / minimum of the applied control, the value function becomes quite
-% flat for $x \ll 1$ and $t \ll \T$, since it is likely that we can apply enough
-% control to avoid an early spike and still drive the system close to threshold
-% at the desired time, $t = \T$
-Now focus on the controls, on the right in
-\cref{fig:HJB_4regimes_value_control_cuts}. Naturally, at the end of the
-interval, the control takes its maximum value, $\a(x,\T) = \amax$ for all $x$,
-i.e.\ it gives the maximum available push for the neuron to spike. This is
-implied \cref{eq:valuefun_TCs}. As we go back in time, however, the control $\a$
-decreases for $x \ll \xth$, and it becomes negative, i.e.\ inhibitory for $x
-\lesssim \xth$. That is intuitively consistent with the problem objective. 
-For $t<\T$, we want to bring $X_t$ closer to the threshold, but not too close, to
-avoid early spiking.
-
-Now consider the differences between the individual regimes, i.e.\ the
-effect of changing the bias and the noise intensity, $\m,\b$. All things being
-equal, the effect of increasing the noise intensity, $\b$, is to lift the value
-function, i.e.\ to make our objective worse, at the beginning of the interval
-and to decrease it at the end. This is to be expected since we are attempting to
-minimize the variance in the spike time and without noise there would be no
-variance at all. However, at the end of the interval, noise only helps, since
-now we are only interested in spiking as soon as possible, and a higher noise
-intensity will tend to decrease the spike time on average. Increasing $\b$ also
-has the effect of increasing the size of the boundary layer near $\xth$, where
-the value function rises steeply. I.e.\ for small $\b$ that layer is small since
-the risk of spiking early is only significant close to the threshold. For larger
-$\b$ this layer increases, see panels A,C, with a thin layer, vs.\ B,D, with a
-thicker layer, in \cref{fig:HJB_4regimes_value_control_cuts}. Similarly,
-increasing the bias, $\mu$, tends to decrease the value function, especially at
-the end since that has the unequivocal effect of preventing late spikes.
-
-\section{Open-loop Stochastic Control}
-When the value of $X_t$ is unobservable, the transition density can be used to
-perform the optimization. Since the transition density follows a deterministic
-PDE, we apply a Maximum Principle for PDEs as a method of obtaining the optimal
-control, see \cite{Borzi2012} for details on optimization with PDEs.
-
-\subsection{Fokker-Planck equation for the state density evolution}
-We write the transition density of $X$, conditional on no
-spikes having occurred as:
-\begin{align*}
-f(x,t) \intd{x} &= \Prob[X_t\in \intd{x} \,\big|\, X_0 = 0,\, X_{s} < \xth \quad
-\forall s < t]
-\end{align*}
-Then $f$ satisfies a Fokker-Planck equation with absorbing boundaries,
-see Ch.\ 7 in \cite{Jacobs},
-\begin{equation}
-\begin{gathered}
-\di_t f(x,t) =
-				\frac{\b^2 }{2}\cdot \di^2_x f -
-				\di_x \Big[ \left( \m + \a(t)- \frac{x}{\tc}\right)  \cdot f \Big]
-\\
-\\
-\begin{array}{lll}
-	&\textrm{st.\ }&
-	\left\{ \begin{array}{lcl}
-	 f(x,0) &=& \delta(x) \quad \textrm{delta function}
-	\\
-	(\m + \a(t)- \tfrac x \tc)f - \di_x \tfrac {\b^2}2 f] \big|_{x=\xmin} &\equiv&
-	0 \quad \textrm{reflecting BCs at some } \xmin
-	\\
-	f(x,t) |_{x=\xth} &\equiv& 0 \quad \textrm{absorbing BCs at } \xth
-\end{array} \right.
-\end{array}
-\label{eq:FP_pde_OU_PDF}
-\end{gathered}
-\end{equation}
-In theory, $\xmin = -\infty$, but in the numerics below we will need to
-truncate it to some finite value, exactly as in the HJB equation,
-\cref{eq:OC_LS_HJB_full}.
-Note that we write $\a(t)$ here instead of $\a(x,t)$ since we cannot use the
-value of $X_t$.
-
-The $f$ dynamics can also be written as
-$$
-\di_t f(x,t) = - \di_x \phi(x,t)
-$$
-for the probability flow
-$$
-\phi(x,t) = (\m + \a(t) - \tfrac x \tc)f - \di_x [\tfrac {\b^2}2 f]
-$$
-Then the lower BC is
-$$
-\phi(x,t) |_{x=\xmin} \equiv 0
-$$
-
-We will also need a short hand notation for the differential operator on the
-right side of \cref{eq:FP_pde_OU_PDF}. Let
-$$ \L_\a[\cdot] := \di^2_x \Big[\frac{\b^2 }{2} \cdot\Big] -
- \di_x \Big[ (\m + \a(t)- \frac{x}{\tc}) \cdot \Big] $$where
- $[\cdot]$ indicates the argument of the operator $\L_\a$. We will usually
- omit the subscript $\a$, but have written it now to emphasize that the
- differential operator is parametrized by the control, $\a$.
-
-\subsection{Restating the objective in terms of the transition density}
-Let us recall our objective, \cref{eq:OC_LS_variance}:
-$$
-J[\a(\cdot)] = \Exp\left[
-(\ts - \T \big)^2
-+
-\e \int_0^\ts  \a^2(s) \intd{s}
-\right].
-$$
-Rewriting this in terms of the transition density, $f$, reads:
-\begin{align}
-J[\a(\cdot)] =&
-\int_\xmin^\xth \Ttwo(x) \cdot f(x,\T) \intd{x}
-\notag
-\\
-&+ \int_0^\T \phi(\xth, t) (t-\T)^2 \intd{t}
-\label{eq:OC_LS_variance_density}
-\\
-&+  \e \int_0^\ts  \a^2(t)  \int_\xmin^\xth f(x,t) \intd{x} \intd{t}
-\notag
-\end{align}
-Let us explain in more detail each term on the right-hand side in
-\cref{eq:OC_LS_variance_density}.
-
-The first term $$ \int_\xmin^\xth \Ttwo(x) \cdot f(x,\T)\intd{x} $$ counts the cost of trajectories which spike too late. This cost is the expected
-squared time-to-hit starting at $x$, with  $\a = \amax$, weighted by the
-probability of $X_\T = x$, which is just $f(x,\T)$. Recall that $\Ttwo$ is
-defined in \cref{eq:Trem_squared}.
-
-The second term $$ \int_0^\T \phi(\xth,t) (t-\T)^2 \intd{t} $$ counts the cost
-of trajectories which spike too early, that is the squared difference between
-some realized spike time, $\ts = t$, and desired spike time $\T$, weighted by
-the probability of a spike at $t$ which is just the outflow of probability,
-$\phi(\xth,t)$. Recall that we assume that $\xth = 1$ throughout. Note further
-that due to the homogeneous Dirichlet BC at $\xth$, $f(\xth) = 0$, the outflow
-is simply: $$ \phi(\xth, t) = -\frac{\b^2 }{2} \di_x f(\xth, t).$$
-
-Finally, the third term
-$$
-\e \int_0^\ts  \a^2(t)  \left[  \int_\xmin^\xth f(x,t) \intd{x} \right] \intd{t}
-$$
-is the energy cost: the inner integral, $\int_\xmin^\xth f(x,t) \intd{x}$
-takes into account that we incur an energy cost only for those trajectories
-which have not yet spiked.
-
-With that our optimal control $\a^*(\cdot)$ will naturally be found via:
-$$
-\a^*(\cdot) = \argmin_{\a(\cdot)} J[\a(\cdot)]
-$$
-
-\subsection{Optimizing using a Maximum Principle}
-\label{sec:PDE_max_principle_for_pdf}
-By now our stochastic optimal control problem modelled by SDEs
-has become a deterministic optimal control problem modelled by PDEs. Our control
-$\a$ influences the evolution density $f$ and via $f$, the integrals in the
-objective, $J$.
-
-The Maximum Principle for PDEs, which is an extension to the famous Pontryagin
-Maximum Principle from finite dimensional systems, introduces an adjoint
-variable, $p$, which solves a PDE related to the PDE satisfied by the density
-$f$ and then calculates the optimal control, $\a(\cdot)$, as a functional of $f$
-and $p$.
-
-In short, the equation for the adjoint function, $p$, is
-\begin{equation}
-\begin{gathered}
-\begin{aligned}
-\di_t p =& - \Lstar[p]
-\\
-		=&
-			- \Big[ \frac{\b^2 }{2}\cdot \di^2_x p +
-			(\m + \a(t)- \frac{x}{\tc})  \cdot \di_x p \Big]
-\end{aligned}
-\\
-\st
-\begin{cases}
-	p(x,\T) &= \Ttwo(x)
-	\\
-	\di_x p  \big|_{x=\xmin} &\equiv 0
-	\\
-	p \big|_{x=\xth} &\equiv (s-T)^2
-\end{cases}
-\label{eq:adjoint_pde_OU}
-\end{gathered}
-\end{equation}
-
-and then $\a$ can be found via
-\begin{equation}
-\Big\{
- \tilde{\e}  2 \a(t)
-+ p f \Big|_\xmin
-- \int _\xmin^\xth p \cdot \di_x f \intd{x}
-\Big\} = 0
-\quad \forall t \in [0,T]
-\label{eq:J_necessary_condition}
-\end{equation}
-
-More practically, the quantity,
-$$\di_\a J =  \Big\{
- \tilde{\e}  2 \a(t)
-+ p(x,t) f(x,t) \Big|_\xmin
-- \int _\xmin^\xth p \cdot \di_x f(x,t) \intd{x}
-\Big\}
-$$
-gives the direction of increase of $J$ at $\a(t)$
-and we can use it as a gradient in a descent algorithm, given some initial
-guess, $\a_0(t)$, for the control.
-
-Finally, we are ready to calculate the open-loop stochastic optimal control for
-the four parameter regimes. The calculated optimal controls for each regime are
-in \cref{fig:FBK_Regimes_cs}. Recall that the optimal control, $\a^*(t)$, is
-open-loop and thus it is only a function of time.
-% \usepackage{graphics} is needed for \includegraphics
 \begin{figure}[htp] 
 \begin{center} 
-  \includegraphics[width=.99\textwidth]{Figs/FP_Adjoint/Regimes_cs.pdf}
+  \includegraphics[width=\textwidth]{Figs/FP_Adjoint/Regimes_cs_singleplot.pdf} 
   \caption[labelInTOC]{The deterministic optimal controls for each parameter 
@@
   = 0.001$ and the bounds are $\a \in [-2,2]$. 
-  A) Supra-Threshold-Low-Noise B) Supra-Threshold-High-Noise
-C) Sub-Threshold-Low-Noise   D) Sub-Threshold-High-Noise  }
+  From left-to-right, the control curves correspond to the 
+  Sub-Threshold-Low-Noise, Sub-Threshold-High-Noise, Super-Threshold-High-Noise, 
+  Sub-Threshold-Low-Noise regimes 
+  } 
   \label{fig:FBK_Regimes_cs} 
@@
 \end{figure}    
-
-Let us discuss the most salient features of the controls calculated in
-\cref{fig:FBK_Regimes_cs}. In general, the strategy is to inhibit the neuron at
-the beginning of the interval, $\a(t) = \amin$ for $t \ll \T$ and to excite it
-near the end, $\a(t) = \amax$ for $t \approx \T$. The smooth portion that connects these
-two segments in the middle of the interval is due to the energy penalty, $\e \int
-\a^2(s) \intd{s}$. The behaviour of the control in different regimes
-is relatively simple to explain and we see what we would expect. In a
-sense the bias current $\mu$ can be absorbed by the control and so increasing
-$\mu$ amounts to decreasing $\a$ modulo its bounds. That is exactly the
-difference between plots A) vs.\ C) and B) vs.\ D) in \cref{fig:FBK_Regimes_cs},
-where the reduction of $\mu$ in the lower plots, C) and D), results in an
-increase in $\a(t)$ compared to A) and B), respectively. The effect of the noise
-intensity, $\b$, is more subtle and it is not the same in the Supra vs.\ the
-Sub-Threshold regimes A,B) vs.\ C,D). In the Supra-Threshold regime,  panels
-A,B), reducing $\b$, reduces the need for excitatory  control towards the end,
-since the bias alone should put the neuron over the threshold. Thus the main
-part of the control with low noise in the Supra-Threshold regime is to stop the
-neuron from spiking too early. In the Sub-Threshold regime, panels C,D), the
-situation is somewhat reversed - reducing $\b$, obviates the need to apply an
-inhibitory control in the early part of the interval and also prompts the
-controller to apply excitatory input earlier, since it is the controller which
-is now the main drive for a spike.
-
-% An example for . The results are in
-% \cref{fig:FP_adjoint_objective_control_convergence}.
-% \begin{figure}[h]
-% \begin{center}
-% \subfloat[$\a_k(t)$ ]
-% {
-% \includegraphics[width=0.48\textwidth]
-% {Figs/FP_Adjoint/ExampleControlConvergence_control.png}
-% }
-% \subfloat[$J_k$ ]
-% {
-% \includegraphics[width=0.48\textwidth]
-% {Figs/FP_Adjoint/ExampleControlConvergence_objective.png}
-% }
-% \caption[ ]{The result of an entire optimization iteration, on the left the
-% iterates of the control, $\a_k(t)$, on the right the progress of the
-% objective with each iteration. The final control, in purple, is inhibitory at
-% the beginning of the time interval, $\a < 0$, unlike the the deterministic
-% control solution.
-% We see a significant
-% reduction in the objective value, $J$, and thus an improvement, on the
-% order of 50\%.}
-% \label{fig:FP_adjoint_objective_control_convergence}
-% \end{center}
-% \end{figure}
-%
-% \begin{table}[h]
-% \centering
-% \begin{tabular}{lc}
-% Control Law & Squared Error \\
-% \hline
-% Deterministic &  0.647 \\
-% Stochastic &  0.336\\
-% Theory (Value function) & 0.285
-% \end{tabular}
-% \caption{Realized performance of the different control laws and the theoretical
-% expected performance of the stochastic law (last row)}
-% \label{tab:realized_avg_errors_det_vs_stoch}
-% \end{table}
-%
-% \begin{figure}[h]
-% \begin{center}
-% \subfloat[A]
-% {
-% \label{fig:controlled_traj_ex1}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/example_controlled_trajectories_id1.png}
-% }
-% \subfloat[B]
-% {
-% \label{fig:controlled_traj_ex2}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/example_controlled_trajectories_id4.png}
-% }
-% \subfloat[C]
-% {
-% \label{fig:controlled_traj_ex3}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/example_controlled_trajectories_id3.png}
-% }
-% \caption[]{Examples for the controlled trajectories using both the deterministic
-% and the stochastic control approaches. The red vertical line in the plots
-% indicates the desired spike-time, $\T$. In A, B the performance of both
-% control laws is essentially the same, but in C we see the advantage of the
-% stochastic approach. $\tc, \b = [.75, 1.25]$.}
-% \label{fig:control_trajectories_examples}
-% \end{center}
-% \end{figure}
-% \begin{figure}[htp]
-% \begin{center}
-%   \includegraphics[width=.9\textwidth]{Figs/ControlSimulator/example_controlled_trajectories_hists.png}
-%   \caption[labelInTOC]{Histogram of the spike timing error for the
-%   deterministic (left) vs. the stochastic (right) control laws.}
-%   \label{fig:error_histograms_det_vs_stoch}
-% \end{center}
-% \end{figure}
-
-\section{Simulations} 
-\subsection{Single-Spike Control}
-\label{sec:probabilistic_numerical_test}
-Having obtained the optimal controllers, both closed- and open-loop, we evaluate
-their performance with simulated realizations of the voltage process,
-\cref{eq:X_evolution_uo}. In particular, while they both minimize the expected
-squared deviation of the spike-time from some desired spike time, we analyze
-the distribution of the squared deviation and the behaviour of the
-controls for different parameter regimes.
-
-In addition to the optimal controllers, we also show the behaviour of
-another control law - perhaps the most naive one - which is obtained by ignoring
-the noise and the energy penalty. That is we find a {\sl constant} value of $\a$
-which satisfies the desired boundary conditions, $x(0) = 0, x(\T) = \xth$. This naive
-controller will be called 'Deterministic', since it assumes deterministic
-dynamics in $X_t$, i.e.\ $\b = 0$.
-
-For the comparison, we sample $N$ realizations of the controlled system
-and apply in turn each of the three controls. Naturally, we reuse the same
-realization of the underlying stochastic process, $W_t$, for each of the
-three different controls.
-
-We set $\e = 0.001$ and $\amax = 2.0$. As such the energy cost is of secondary
-importance and the paramount effect on the objective is the difference $(\ts -
-\T)$, which we are trying to minimize.
-
-The performance of the three controllers for each parameter regime is given in
-\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch} and
-\cref{fig:error_histograms_det_vs_openloop_vs_stoch}. Naturally, the closed-loop
-achieves a lower error than the open-loop, and the naive controller fares worst.
-The difference in performance mostly depends on the strength of the noise. Thus,
-for low noise, the performance of the stochastic controllers, be they open- or
-closed-loop is not much superior to the naive deterministic controller. On the
-contrary - in the high-noise regime, using a stochastic controller gives a much
-lower error on average between the desired $\T$ and the realized $\ts$.
-
-For illustration sake, we also visualize some trajectories from the
-Sub-Threshold-High-Noise regime in \cref{fig:control_trajectories_examples},
-see panels A),C),E).
-
-\begin{table}[h]  
-\begin{center}  
-\subfloat[Supra-Threshold Low-Noise]{
-   \input{Figs/ControlSimulator/MeanSquaredErrors__SUPT_ln.txt}}\\ 
-\subfloat[Supra-Threshold High-Noise]{   
-\input{Figs/ControlSimulator/MeanSquaredErrors__SUPT_HN.txt}}\\
- \subfloat[Sub-Threshold Low-Noise]{
- \input{Figs/ControlSimulator/MeanSquaredErrors__subt_ln.txt}}\\ 
- \subfloat[Sub-Threshold High-Noise]{
- \input{Figs/ControlSimulator/MeanSquaredErrors__subt_HN.txt}}\\
-%  \begin{tabular}{lcc}
-% Control Law &  Squared Error &  Squared Error \\
-%  & Empirical & Theoretical\\% \hline 
-% Deterministic &  0.621 & - \\
-% Open-Loop Stochastic & 0.356 & 0.382\\
-% Feedback Stochastic &  0.288& 0.285\\
-% \hline % \end{tabular} - 
-\caption{
-Realized and theoretical performance of the different control laws. The
-empirical performance is obtained using $N=10000$ sample paths. The theoretical
-performance is found using the optimal value for $J$ for the open-loop
-stochastic control and the value function $\v(x=0, t =0)$ for the closed-loop
-stochastic control.}
-\label{tab:realized_avg_errors_det_vs_openloop_vs_stoch} 
-\end{center}
-\end{table}
-
 \begin{figure}[h]   
@@
 \subfloat[Supra-Threshold Low-Noise]{ 
-  \includegraphics[width=.99\textwidth]
+  \includegraphics[width=\textwidth] 
   {Figs/ControlSimulator/Regimes_SUPT_ln_errors_hist.pdf} 
@@
 \subfloat[Supra-Threshold High-Noise]{ 
-  \includegraphics[width=.99\textwidth]
+  \includegraphics[width=\textwidth] 
   {Figs/ControlSimulator/Regimes_SUPT_HN_errors_hist.pdf} 
@@
 \subfloat[Sub-Threshold Low-Noise]{ 
-  \includegraphics[width=.99\textwidth]
+  \includegraphics[width=\textwidth] 
   {Figs/ControlSimulator/Regimes_subt_ln_errors_hist.pdf} 
@@
 \subfloat[Sub-Threshold High-Noise]{ 
-  \includegraphics[width=.99\textwidth]
+  \includegraphics[width=\textwidth] 
   {Figs/ControlSimulator/Regimes_subt_HN_errors_hist.pdf} 
@@
   \caption[labelInTOC]{Histogram of the spike timing error for the 
-  deterministic (left) vs. open-loop stochastic(centre) vs. closed-loop
-  stochastic (right) control laws. This is for the same problem as in
+  deterministic (left) vs. open-loop stochastic (centre) vs. closed-loop 
+  stochastic (right) control laws. 
+  Recall that $\ts$ is the random realized spike time and $\T$ is the target 
+  spike time.   
+  This is for the same problem as in 
   \cref{fig:control_trajectories_examples}. 
   We have used $N=10000$ sample paths 
-  to form the statistics}
+  to form the statistics.} 
   \label{fig:error_histograms_det_vs_openloop_vs_stoch} 
@@
 \end{figure} 
-
-% \begin{figure}[h]
-% \begin{center}
-% \subfloat[]
-% {
-% \label{fig:controlled_traj_ex1}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/SubTHighNoise_Traj0.pdf}
-% }
-% \subfloat[ ]
-% {
-% \label{fig:controlled_traj_ex2}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/SubTHighNoise_Traj4.pdf}
-% }
-% \subfloat[ ]
-% {
-% \label{fig:controlled_traj_ex3}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/SubTHighNoise_Traj7.pdf}
-% }
-% \caption[]{Examples for the controlled trajectories using the
-% deterministic, open-loop stochastic and closed-loop stochastic control
-% approaches. The black vertical line in the plots indicates the desired spike-time, $\T$.
-% For the parameter values are $\m, \tc, \b = [.2, .5,  1.5]$ (Sub-Threshold
-% High-Noise regime). $\T = 1.5, \eps = 0.001$. The three panels from left to
-% right are three different realizations of the model dynamics. On the upper
-% plots, we show the voltage evolution, $X_t$, on the lower plots we show the
-% applied control, $\a(t)$. Note that $\a(t)$ obtained from the
-% deterministic and open-loop controls are the same for all three samples.}
-% \label{fig:control_trajectories_examples}
-% \end{center}
-% \end{figure}
-
-\subsection{Multi-Spike Control}
-We now turn to the ultimate goal of our analysis, the control of {\sl spike
-trains}. In particular, we will generate $M=10$ realizations of $N=20$ spikes
-each in an attempt to meet a prescribed spike train. We will focus on two
-regimes, the Supra-Threshold regime with either low or high noise.
-The results
-for parameters from the low-noise regime are shown in
-results in \cref{fig:targettrain_cl_lownoise,fig:targettrain_ol_lownoise}, 
-while, the results
-for the high-noise regime are shown in
-results in
-\cref{fig:targettrain_cl_highnoise,fig:targettrain_ol_highnoise}. In all cases,
-the target train is obtained by a simulation of the Supra-Threshold High Noise
-regime with no additional control, i.e.\ $\a(\cdot) = 0$.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%% SUPER-THRESHOLD REGIME LOW-NOISe:
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[h] 
+\begin{center} 
+\includegraphics[width=0.99\textwidth] 
+{Figs/ControlSimulator/Composite_Traj.pdf}  
+\caption[]{Three examples for the controlled trajectories using the 
+deterministic, open-loop stochastic and closed-loop stochastic control approaches. 
+The plots on the left, A,C,E, are obtained using a value of $\eps=0.001$, 
+while the plots on the right, B,D,F, use $\eps=0.1$ (for a discussion on the 
+effect of $\eps$ see \cref{sec:effect_of_eps}). The black vertical line in the plots indicates the 
+desired spike-time, $\T$. The parameter values are $\m, \tc, \b = [0.2, 0.5,  1.5]$ (Sub-Threshold 
+High-Noise regime), with $\T = 1.5$.  
+The upper plots show the voltage evolution, $X(t)$, while the lower 
+plots show the applied control, $\a(t)$. Note that the optimal control, $\a(t)$, obtained from 
+the deterministic and from the open-loop controls are the same across all three 
+samples.} 
+\label{fig:control_trajectories_examples}    
+\end{center}   
+\end{figure} 
 \begin{figure}[htp] 
 \begin{center} 
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_ln_cl_trains_sim_10.pdf}
+  \includegraphics[width=\textwidth]{Figs/TrainController/SUPT_ln_cl_trains_sim_50.pdf} 
   \caption[ ]{How good the {\sl closed-loop} controller is at hitting a target 
-  spike train. Uses the parameters from the Super-Threshold,Low-Noise regime.
-  The target spike train was generated using the model itself, with
-  parameters from the Super-Threshold, High-Noise regime, without an applied
+  spike train. In the panel above, the blue curve is the empirical firing 
+  rate of $M$ controlled trains.   strategy, the red curve is a 
+  smoothed-version of the target train. (See text for details of how they are calculated). In the panel below, the 
+  dashed, red lines indicate the target times, $\t_n$, while the blue dots are 
+  the realizations of the controlled trains.   
+  The parameters from the Super-Threshold, Low-Noise 
+  regime are used. The target spike train was generated using the model itself, 
+  with parameters from the Super-Threshold, High-Noise regime, without an applied 
   control, $\a=0$. 
-  There are $N=16$ spikes and $M=10$ realizations of the control strategy} 
+  There are $N=16$ target spikes and $M=50$ realizations of the control strategy}  
   \label{fig:targettrain_cl_lownoise} 
@@
 \begin{center} 
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_ln_ol_trains_sim_10.pdf}
+  \includegraphics[width=\textwidth]{Figs/TrainController/SUPT_ln_ol_trains_sim_50.pdf} 
   \caption[ ]{How good the {\sl open-loop} controller is at hitting a target 
-  spike train. Uses the parameters from the Super-Threshold,Low-Noise regime.
-  The target spike train was generated using the model itself, with
-  parameters from the Super-Threshold, High-Noise regime, without an applied
-  control, $\a=0$.
-  There are $N=16$ spikes and $M=10$ realizations of the control strategy}
+  spike train. 
+  The drawn curves have  the same meaning as in 
+  \cref{fig:targettrain_cl_lownoise}. 
+  The parameters from the Super-Threshold, Low-Noise regime are used.  
+  The target spike train was generated using the model itself, with parameters 
+  from the Super-Threshold, High-Noise regime, without an applied control, $\a=0$. 
+  There are $N=16$ target spikes and $M=50$ realizations of the control strategy} 
   \label{fig:targettrain_ol_lownoise} 
@@
 \end{figure} 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%% SUPER-THRESHOLD REGIME HIGH-NOISe:
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[htp] 
 \begin{center} 
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_HN_cl_trains_sim_10.pdf}
+  \includegraphics[width=\textwidth]{Figs/TrainController/SUPT_HN_cl_trains_sim_50.pdf} 
   \caption[ ]{How good the {\sl closed-loop} controller is at hitting a target 
-  spike train. Uses the parameters from the Super-Threshold,High-Noise regime.
-  The target spike train was generated using the model itself, with
+  spike train.  
+  The drawn curves have  the same meaning as in 
+  \cref{fig:targettrain_cl_lownoise}. 
+  The parameters from the Super-Threshold, High-Noise regime 
+  are used. The target spike train was generated using the model itself, with 
   parameters from the Super-Threshold, High-Noise regime, without an applied 
   control, $\a=0$. 
-  There are $N=16$ spikes and $M=10$ realizations of the control strategy}
+  There are $N=16$ target spikes and $M=50$ realizations of the control strategy} 
   \label{fig:targettrain_cl_highnoise} 
@@
 \begin{center} 
-  \includegraphics[width=.99\textwidth]{Figs/TrainController/SUPT_HN_ol_trains_sim_10.pdf}
-  \caption[ ]{How good the {\sl open-loop} controller is at hitting a target
-  spike train. Uses the parameters from the Super-Threshold,High-Noise regime.
-  The target spike train was generated using the model itself, with
+  \includegraphics[width=\textwidth]{Figs/TrainController/SUPT_HN_ol_trains_sim_50.pdf} 
+  \caption[]{How good the {\sl open-loop} controller is at hitting a target 
+  spike train.  
+  The drawn curves have  the same meaning as in 
+  \cref{fig:targettrain_cl_lownoise}. 
+  The parameters from the Super-Threshold, High-Noise regime 
+  are used. The target spike train was generated using the model itself, with 
   parameters from the Super-Threshold, High-Noise regime, without an applied 
   control, $\a=0$. 
-  There are $N=16$ spikes and $M=10$ realizations of the control strategy}
+  There are $N=16$ target spikes and $M=50$ realizations of the control strategy} 
   \label{fig:targettrain_ol_highnoise}     
@@
 \end{figure} 
-
-\clearpage
-
-\section{Miscleaneous Analysis???}
-TODO: WE DON''T KNOW WHAT TO CALL THIS SECTION!
-\subsection{Effect of energy penalty}
-\label{sec:effect_of_eps}
-So far we have assumed a low value for the energy penalty parameter,
-$\eps$, in the objective defined in \cref{eq:OC_LS_variance_energy}. Our
-approach has been that we are most concerned with the accurate spiking and that
-penalizing energy is rather intended to regularize the control and avoid
-excessive chattering of the control between its extreme values, than as a goal
-in minimizing energy in its own right. Now, we take some time to explore the
-effect of a higher $\eps$ on the optimal controls. Intuitively, a
-higher value of $\eps$ will tend to bring the optimal control, $\a^*$, closer
-to zero. We make a comparison for the four regimes for both the
-closed-loop control and the open-loop control in
-\cref{fig:HJB_4regimes_control_different_eps,fig:FBK_Regimes_cs_different_es}.
-Indeed, for both the closed-loop and open-loop optimal controls, increasing
-$\eps$ tends to reduce the absolute value of $\a^*(t)$. In particular, for the
-closed-loop control it tends to broaden the area of transition for
-decreasing $x$ when the control swings from its minimal, i.e.\ inhibitory value,
-to its maximal, i.e.\ excitatory value. Similarly for the open-loop control,
-instead of banging from its minimal bound at the beginning of the interval to
-its maximal bound at the end, the optimal control for $\e = 0.1$ tends to have a
-more mild transition from a slightly inhibitory to slightly excitatory
-values.
-
-Note however that increasing $\eps$ has an important effect on the validity of
-the approximation, used to form the terminal conditions for
-the value function and the adjoint variable, $\v,p$. We assumed that applying
-$\amax$ is optimal for $t>\T$. This is indeed true in the
-limit $\eps \ra 0$. However, even for a finite value of $\eps$, this may still
-be the optimal thing to do. Indeed, in the original calculations,
-where $\eps=0.001$, see the panels on the left in
-\cref{fig:HJB_4regimes_control_different_eps,fig:FBK_Regimes_cs_different_es},
-the so-obtained value function {\sl implied} that $\a(x,\T) = \amax$.
-I.e.\ our
-guess that $\a(x, t) = \amax$ for $t > \t$, is self-consistent with the
-so-obtained value function. This is akin to confirming an ansatz.
-For a higher value of $\eps$, like $\eps = 0.1$, this is no longer the case. The red
-curves on the right-side of \cref{fig:HJB_4regimes_control_different_eps} are no
-longer pushed up at $\a = \amax$, and as such it is no longer valid to
-assume that the value function can be obtained by assuming $\a(t>\T) = \amax$
-and then just calculating the expected remaining time-to-spike.
-The exact same conclusion can be drawn from the open-loop control. There the
-optimal control always (almost) reaches its maximum, $\amax$ while choosing
-$\eps = 0.001$, but no longer so for the higher $\eps = 0.1$.
-In fact, as we mentioned after \cref{eq:valuefun_TCs}, the correct thing to do
-for higher values of $\eps$, is to push the
-calculation interval to some $t^+ > \T$, apply the same terminal condition at
-$t^+$ instead of $\T$ and then solve backwards (either for the closed- or
-open-loop control). One will be guaranteed to find some $t_f > \T$ that works
-since eventually the quadratic term $(t-\T)^2$ will dominate the energy term in
-the objective, which is linear in $t$. We do not explore this further here.
-% CLOSED_LOOP EPS:
 \begin{figure}[htp] 
 \begin{center} 
-  \includegraphics[width=.99\textwidth]{Figs/HJB/Regimes_eps_comparison.pdf}
+  \includegraphics[width=\textwidth]{Figs/TrainController/SUPT_HN_cl_aplus_trains_sim_50.pdf} 
+  \caption[ ]{Same as \cref{fig:targettrain_cl_highnoise}, but with the control 
+  constraints, $[\amin, \amax]$ increased to $[-4,4]$ instead of $[-2,2]$.} 
+  \label{fig:targettrain_cl_highnoise_aplus}     
+\end{center} 
+\end{figure}  
+\begin{figure}[htp] 
+\begin{center} 
+  \includegraphics[width=\textwidth]{Figs/HJB/Regimes_eps_comparison.pdf} 
   \caption[labelInTOC]{Effect of $\e$ on the closed-loop control. Snapshots of 
@@
 \end{figure} 
-% OPEN_LOOP EPS:
 \begin{figure}[htp] 
 \begin{center} 
-  \includegraphics[width=.99\textwidth]{Figs/FP_Adjoint/Regimes_eps_comparison.pdf}
+  \includegraphics[width=\textwidth]{Figs/FP_Adjoint/Regimes_eps_comparison.pdf} 
   \caption[labelInTOC]{Effect of $\e$ on the open-loop control. 
@@
 \end{figure} 
+\clearpage
+\begin{table}
 
-\begin{figure}[h]
+\begin{tabular}{|l||{c}|{c}|}
+\hline
+\backslashbox{$\m$}{$\b$}
+& $1.5$ & $0.3$ \\
+\hline
+$1.5 / \tc $ &Supra-threshold-High-noise & Supra-threshold-Low-noise \\
+\hline
+$0.1 / \tc$   &Sub-threshold-High-noise & Sub-threshold-Low-noise \\
+\hline
+\end{tabular}
+\caption{Regime labels and example values. Note that for the numerical
+experiments below, we use $\tc = 0.5$}
+\label{tab:regimes}
+\end{table}
+\begin{table}[h]
+
 \begin{center}  
-\includegraphics[width=0.99\textwidth]
-{Figs/ControlSimulator/Composite_Traj.pdf}
-\caption[]{Three examples for the controlled trajectories using the
-deterministic, open-loop stochastic and closed-loop stochastic control approaches.
-The plots on the left are obtained using a value of $\eps=0.001$, while the
-plots on the right use $\eps=0.1$.
-The black vertical line in the plots indicates the desired spike-time, $\T$.
-The parameter values are $\m, \tc, \b = [0.2, 0.5,  1.5]$ (Sub-Threshold
-High-Noise regime), with $\T = 1.5$. 
-On the upper plots, we show the voltage evolution, $X_t$, on the lower plots we
-show the applied control, $\a(t)$. Note that the optimal control, $\a(t)$, obtained from
-the deterministic and from the open-loop controls are the same across all three
-samples.}
-\label{fig:control_trajectories_examples}
+\subfloat[Supra-Threshold Low-Noise]{
+   \input{Figs/ControlSimulator/MeanSquaredErrors__SUPT_ln.txt}}\\ 
+\subfloat[Supra-Threshold High-Noise]{   
+	\input{Figs/ControlSimulator/MeanSquaredErrors__SUPT_HN.txt}}\\
+ \subfloat[Sub-Threshold Low-Noise]{
+ 	\input{Figs/ControlSimulator/MeanSquaredErrors__subt_ln.txt}}\\ 
+ \subfloat[Sub-Threshold High-Noise]{
+ 	\input{Figs/ControlSimulator/MeanSquaredErrors__subt_HN.txt}}\\
+\caption{
+Realized and theoretical performance of the different control laws. The
+empirical performance is obtained using $N=10000$ sample paths. The theoretical
+performance is found using the optimal value for $J$ for the open-loop
+stochastic control and the value function $\v(x=0, t =0)$ for the closed-loop
+stochastic control.}
+\label{tab:realized_avg_errors_det_vs_openloop_vs_stoch}   
 \end{center} 
-\end{figure}
+\end{table} 
+\begin{table}[h]
 
-% 
-% \begin{figure}[h]
-% \begin{center}
-% \subfloat[]
-% {
-% \label{fig:controlled_traj_ex1}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/HighEps_Traj5.pdf}
-% }
-% \subfloat[ ]
-% {
-% \label{fig:controlled_traj_ex2}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/HighEps_Traj6.pdf}
-% }
-% \subfloat[ ]
-% {
-% \label{fig:controlled_traj_ex3}
-% \includegraphics[width=0.33\textwidth]
-% {Figs/ControlSimulator/HighEps_Traj0.pdf}
-% }
-% \caption[]{The effect of higher energy penalty, $\eps = 0.1$.
-% Examples for the controlled trajectories using the deterministic,
-% open-loop stochastic and closed-loop stochastic control approaches. The black vertical line in the plots indicates the desired
-% spike-time, $\T$.
-% The parameter values are $\m, \tc, \b = [0.2, 0.5,  1.5]$ (Sub-Threshold
-% High-Noise regime). $\T = 1.5, \eps = 0.1$. The three panels from left to
-% right are three different realizations of the model dynamics. On the upper
-% plots, we show the voltage evolution, $X_t$, on the lower plots we show the
-% applied control, $\a(t)$. Note that the optimal control, $\a(t)$, obtained from
-% the deterministic and from the open-loop controls are the same across all three
-% samples.}
-% \label{fig:control_trajectories_examples_high_epsilon}
-% \end{center}
-% \end{figure}
-
-\subsection{Control Robustness}
-We have so far assumed that we know the parameters, $\m, \tc, \b$, determining
-the dynamics in \cref{eq:X_evolution_uo,eq:current_mu_alpha} exactly. That is we
-assume that the values for $\tc, \b$ used to obtain the controls are the same as the ones that drive the
-controlled dynamics. That is slightly naive, as in any realistic application
-these will have to be estimated and the estimates will have some error
-associated with them.
-Note that misestimating $\mu$ has the most immediate and clear effect on the
-control policy as that amounts to a direct shift of the control and so for now,
-let us assume that $\mu$ is estimated correctly.
-
-Let us then discuss one of the regimes, the Sub-Threshold,
-High-Noise regime, and see what is the impact of misspecifying the parameters.
-
-Assume that $\b$ and $\tc$ are underestimated, so that $\b_{est} = \b / 1.5$
-and $\tc_{est} = \tc / 1.5$.
-
-So we will have:
-$$\mu, \tc_{est}, \b_{est} = \{.2, .5, 1.5 \}$$ which is the control
-calculated already for the Sub-Threshold,
-High-Noise regime, while the actual simulated dynamics will be driven by
-$$\mu, \tc, \b = \{.2, .75, 2.25 \}$$
-
-We show the effect in
-\cref{tab:realized_avg_errors_det_vs_openloop_vs_stoch_misspecified}. Note that
-a misspecification of $\b$ should not have a direct influence on the behaviour
-of the 'deterministic' controller since it already assumes in any case that $\b=0$.
-\begin{table}[h]
 \begin{center}
- \subfloat[Sub-Threshold High-Noise]{
- \input{Figs/ControlSimulator/Misspec_MeanSquaredErrors__subt_HN.txt}}\\
+ \input{Figs/ControlSimulator/Misspec_MeanSquaredErrors__subt_HN.txt}
  \caption{The effect of misspecifying the parameters. On the left, the
  system parameters and the parameters used to obtain the control are the
- same, i.e.\ accurate, on the right, they are misspecified.}
+ same, i.e.\ correct, on the right, they are misspecified.}
 \label{tab:realized_avg_errors_det_vs_openloop_vs_stoch_misspecified}
@@
 \end{table}
-
-\section{Discussion}
-We have analyzed and computed the optimal control of spikes in two scenarios -
-one where the underlying voltage of the neuron is observable to the controller
-(closed-loop control) and one where only the spikes are observable (open-loop
-control).
-
-Naturally, the ability to control the system is most clearly affected by the
-level of the noise. In a sense, the noise acts like an adversary - in our
-context it has no beneficial role, but to obstruct precise spiking.
-
-The bias current has a more helpful role at least in our examples, where a
-high value of the bias tends to help precise spiking. Of course, where it helps
-or hinders depends on the value of the desired spike time - a combination of a
-high positive bias and a 'distant' spike time, will tend to be difficult to
-control as the system will naturally tend to spike earlier than desired.
-
-In both cases finding the optimal control is computationally intensive as we
-numerically solve partial differential equations. If these algorithms were to be
-practical in an online setting, some more work would have to be done in order to
-ensure they can be computed very quickly (in the millisecond range) or that
-they can be pre-computed.
-
-Our work has strayed pretty close to the paradigm of \cite{Ahmadian2011} in
-trying to make the neuron spike at a particular time. Other goals have been
-formulated, most notably the idea of desynchronizing a population of neurons,
-\cite{Nabi2011}, where it is not so important when the neurons spike as long as
-they do not spike at the same time. We have also incorporated into our objective
-the minimization of total energy used to achieve our goal, which as discussed in
-\cite{Ahmadian2011} is sensible given the potential damage on the cell of
-accumulating charge. Furthermore, we have assumed that our control is
-constrained in magnitude, which is natural given physical limitations on
-equipment and safety considerations. Another possible constraint on the control
-is that it is charge-balanced, meaning that its time integral is zero, $\int u
-\intd{t} = 0$. Such constraints are most easily posed in the context of
-deterministic spiking models like the Phase-Response Curve, see e.g. Danzl et
-al. \cite{Danzl2010}. It is possible to add them to the open-loop control, since
-the control is still deterministic, but it is not-trivial. It is even less
-trivial when one uses Dynamic Programing. In
-particular, insisting on charge-balance in our schemes will make it much more
-difficult to apply the Terminal Conditions used in deriving both the closed- and
-open-loop controls.
-
-The novelty in our work is primarily in obtaining a stochastic control scheme,
-which is not limited by the intensity of the noise in the system. In fact, by
-comparison with the naive 'deterministic' controller, we have shown how wrong
-the control strategy can be if we assume that the noise is negligible. Given
-this scope, the numerical demands on our scheme are non-trivial, but our
-experience and experiments show that these demands are potentially feasible. It
-however remains to be seen whether they can be successfully applied in the lab
-and beyond.
-
-\appendix
-\section{Calculating the Terminal Conditions for the HJB equation}
-\label{sec:valuefun_TCs}
-Here we explain how to calculate the first two moments of the time to-spike,
-$\Exp \big[\trem^p \,\Big|\, X_\T = x, \a(t) = \amax \big].$, which is needed in setting the terminal conditions for both,
-$\v(x,\T)$ and $p(x,\T)$, see \cref{eq:OC_LS_HJB_full} and \cref{eq:adjoint_pde_OU}.
-
-Let
-\begin{equation}
-\begin{array}{lcl}
-\Tone(x_0) &=& \Exp \Big[\trem \,\Big|\, X_\T = x, \a(t) = \amax
-\Big]
-\\
-\Ttwo(x_0) &=&
-\Exp \Big[\trem^2 \,\Big|\, X_\T = x, \a(t) = \amax \Big],
-\end{array}
-\end{equation}
-where as before, $\t$ is the remaining time-to-spike in
-\cref{eq:X_evolution_uo}, at the desired spike-time $t=\T$, given $X_\T = x_\T$
-and $\a(t) = \amax$.
-Naturally, the $\Ti$'s depend on $x_\T$ and are affected by the value of the
-various parameters, $\a,\b,\tc$.
-
-Actually, we only  need $\Ttwo$ for the TCs in \cref{eq:OC_LS_HJB_full} and \cref{eq:adjoint_pde_OU}.
-but $\Tone$ is needed as an auxiliary.
-
-Indeed, analytical expressions exist for $\Ti$, e.g.\ see \cite{Inoue1995}.
-However in our experience, they are awkward to work with numerically, especially
-as they involve infinite alternating sums which can easily overflow in numeric
-calculations. Moreover, they are not particularly convenient if you need $\Ti
-(x_0)$ for a wide range of $x_0$'s as we do here, since they calculate the value
-independently for each $x$. Instead we will follow and slightly extend Jacobs\footnote{Jacobs only discusses how to calculate the first moment
-$\Tone$, however it is quite clear from his discussion how to proceed
-recursively to obtain any higher moment $\Ti$, by using previously calculated
-lower moments. This is what we do here. Moreover the relation between mean exit
-times and boundary value problems is discussed in most
-introductory books on stochastic processes, see \cite{Oksendal2007} for
-example.}, \cite{Jacobs}, in deriving and solving a differential equation for
-$\Ti(x_0)$.
-
-Basically we are calculating the mean-time-squared to exit an interval $[\xmin,
-\xth]$. Theoretically, $\xmin = -\infty$, but for our purposes we will set it to
-some finite value and impose reflecting boundaries there - this is a very good
-approximation for $\xmin \ll 0$, as long as $\a = \amax >0$, which is what we
-have.
-
-Let $B(y,t|x) = \Prob[X_0 = y| X_t = x]$ for a fixed $x$. Then $B$ solves the
-following (backward Kolmogorov) PDE
-\begin{equation}
--\di_t B = (\a - \frac y\tc)\di_yB - \frac {\b^2} 2 \di_y^2 B
-\label{eq:backward_Xdensity}
-\end{equation}
-
-Let $\G(t,y)$ be the survival function for $T$ for $X_0 = y$.
-$$\G(t,y) = \Prob[T>t | X_0 = y] = \Prob[X_{s\leq t} \in [\xmin,
-\xth] | X_0 = y]$$.
-
-Because the dynamics do not depend explicitly on time, we have:
-$$
-\G(t,y) = \int_\xmin^\xth B(y,t|x) \intd{x}
-$$
-and so $\G$, being a definite integral of $B$ satisfies the same PDE:
-\begin{equation}
--\di_t \G = (\a - \frac y\tc) \di_y\G - \frac {\b^2} 2 \di_y^2 \G
-\label{eq:backward_SDF}
-\end{equation}
-
-With the BCs, ICs:
-\begin{equation}
-\begin{cases}
-\G(t,\xth) = .0 & \quad (T = t \implies \Prob[T>t] = 0)
-\\
-\di_x \G(t,\xmin) = .0  & \quad (\textrm{due to reflecting boundary for X})
-\\
-\G(0, y) = 1. & \quad (T \textrm{ is definitely} > 0 \textrm{ for } x_0 \in
-(\xmin, \xth) )
-\end{cases}
-\end{equation}
-
-If we solve the PDE for $\G$ over $t \in [0,\infty)$, then we can calculate
-$\Ttwo(x_0)$ as:
-\begin{eqnarray}
-\Ttwo(x_0) &=& \int_0^\infty t^2 \cdot (-\di_t\G(t,x_0)) \intd{t}
-\\
-		   &=& \int_0^\infty 2t \cdot \G(t,x_0) \intd{t}
-\end{eqnarray}
-assuming it exists.
-
-Since $\Ttwo$ is an integral over time, we can reduce the PDE in
-\cref{eq:backward_SDF} for the distribution to a BVP for the moment(s) by
-integrating over time. Here is how this is done: multiply both sides  of
-\cref{eq:backward_SDF} by $2t$, and integrate with respect to time:
-\begin{equation}
-\int_0^\infty 2t (-\di_t \G) \intd{t}
-=
-\int_0^\infty  2t \left[ (\a - \frac y\tc) \di_y\G - \frac {\b^2} 2
-\di_y^2 \G\right]
-\intd{t}
-\end{equation}
-The LHS turns out to be $2\Tone$, so
-\begin{equation}
-2\Tone
-=(\a - \frac y\tc)   \di_y \Ttwo
-- \frac {\b^2} 2
-\di_y^2 \Ttwo
-\label{eq:BVP_Ttwo}
-\end{equation}
-Similarly we can derive an equation for $\Tone$, namely:
-\begin{equation}
-1
-=(\a - \frac y\tc)   \di_y \Tone
-- \frac {\b^2} 2
-\di_y^2 \Tone
-\label{eq:BVP_Tone}
-\end{equation}
-In both cases, the BCs will be:
-\begin{equation}
-\begin{cases}
-\Ti(\xth) = .0 & \quad (\textrm{we are already spking})
-\\
-\di_x \Ti(\xmin) = .0  & \quad (\textrm{due to reflecting boundary for }X)
-\end{cases}
-\label{eq:BVP_Ti_BCs}
-\end{equation}
-Equations \ref{eq:BVP_Tone} and \ref{eq:BVP_Ttwo} can be reduced to first order
-differential equations and solved analytically, at least up to some definite integral,
-but the analytical solution is not particularly useful or necessary for
-numerical calculations. Instead, we opt to calculate the solutions to
-\ref{eq:BVP_Tone} and \ref{eq:BVP_Ttwo} directly via an ODE solver, starting
-from $T_i(\xth) = .0$ and integrating down for $x  \in [\xmin, \xth]$.
-That is how we have obtained the terminal conditions for both $\v(x,\T)$ and
-$p(x,\T)$ in our calculations above.
-
-\bibliographystyle{plain}
-\bibliographystyle{unsrt}
-\bibliographystyle{bmc_article}
-\section*{Acknowledgements}
-   \ifthenelse{\boolean{publ}}{\small}{}
-SD is supported by the Danish Council for Independent Research $\mid$ 
-Natural Sciences.
-\\ 
-AL acknowledges support from NSERC Canada.  
-\\
-AI is supported by a PGS-D scholarship from NSERC Canada.
-
-\section*{References}
-\bibliographystyle{iopartnum}
-\bibliography{library,local}
-
-\end{document}

### 2014-02-18 ref_reply.tex
Adding ML Note

@@
 % \usetikzlibrary{arrows,snakes,backgrounds}
+% \usetikz
 % \usepackage{verbatim} 
-% \usetikz
-% \usepackage{subfig}
-% \usepackage{graphicx}
+\usepackage{subfig} 
+\usepackage{graphicx}
 \newtheorem*{rep}{Reply}
@@
 fashion. Below, we list the referees comments and then our replies immediately
-following each comment with '\begin{rep} TODO: \end{rep}' to indicate the reply.
-
+following each comment.
 \tableofcontents
@@
 \begin{rep}
-
+TODO
 \end{rep}
@@
 \begin{rep}
-
+Note that this is a function of $\alpha_{max}$. So this really needs to be a two
+dimensional plot with $\beta$ and $\alpha_{max}$ as independent axes.
+See figure \ref{fig:target_error_as_func_of_beta} for how this might look,
+where we investigate a Cartesian product of [Sub-Threshold, Supra-Threshold]
+Target Trains being tracked by [Sub-Threshold, Supra-Threshold] tracking
+trains, with showing three different error metrics as a function of $\beta$.
+%\usepackage{graphics} is needed for \includegraphics
+\begin{figure}[htp] 
+\begin{center}
+\subfloat[Sub-T Target / Sub-T Tracker]{
+  \includegraphics[width=.99\textwidth]{Figs/TrainController/ControlError_Stats_cl_subthn_6_16_subt}
+  }\\
+  \subfloat[Sub-T Target / Supra-T Tracker]{
+  \includegraphics[width=.99\textwidth]{Figs/TrainController/ControlError_Stats_cl_subthn_6_16_SUPT}
+  }
+  \\
+\subfloat[Supra-T Target / Sub-T Tracker]{
+  \includegraphics[width=.99\textwidth]{Figs/TrainController/ControlError_Stats_cl_supthn_8_16_subt}
+  }\\
+\subfloat[Supra-T Target / Supra-T Tracker]{
+  \includegraphics[width=.99\textwidth]{Figs/TrainController/ControlError_Stats_cl_supthn_8_16_SUPT}
+  }    
+  \caption[labelInTOC]{Errors as a function of noise intensity for the
+  CLOSED-LOOP controller. A 'correct' spike for the purposes of calculating
+  'Percent Correct' is one that is within .1$\cdot \hat I$ of the target spike, where $\hat I$ is the mean ISI of
+ the target train, here about .665}
+  \label{fig:target_error_as_func_of_beta}
+\end{center}
+\end{figure} 
 \end{rep}
@@
 
-In this manuscript, the problem of controlling the spike firing times in response to stimulating waveforms was considered in the closed- and open-loop control schemes.
-In particular, the authors have proposed a procedure for determining the stimulating waveform in the open-loop situation which is useful for clinical situations.
-The reviewer thinks that the results of the manuscript are interesting and the manuscript would be well organized.
-However, some issues need to be clarified before recommending the manuscript for publication in Journal of Neural Engineering.
+In this manuscript, the problem of controlling the spike firing times in
+response to stimulating waveforms was considered in the closed- and open-loop
+control schemes. In particular, the authors have proposed a procedure for
+determining the stimulating waveform in the open-loop situation which is useful
+for clinical situations. The reviewer thinks that the results of the manuscript
+are interesting and the manuscript would be well organized. However, some issues
+need to be clarified before recommending the manuscript for publication in
+Journal of Neural Engineering.
 
@@
 \begin{rep}
-
+TODO
+Put an explanation in beginning of section 5.2, and emphasize that quick
+succeeding ISIs are the main bane. There should be no accumulation of errors
+unless the controller is actually unable to catch up to an extremely high firing rate. 
+But otherwise the controller is adaptive to the realized spike times such that
+it is always targeting the original target train. 
 \end{rep}
@@
   \item 
---In l.6, Page 4 of 33, the authors stated “a non-neglegible noise component
+--In l.6, Page 4 of 33, the authors stated “a non-negligible noise component
 ...”. What kind of noise would the authors assume? Is that a weak or strong one?
@@
 \begin{rep}
-It is clear taht the referee is confused. By non-negligible noise component, we
-mean a value of $\b >> 0 $.
+By non-negligible noise component, we
+mean a value of $\beta >> 0 $, while the noise is always Gaussian in the sense
+that the SDE is driven by a Gaussian, white noise. 
 \end{rep}
@@
 the control instead?
+We make surface plots for $\alpha(x,t)$ IN ADDITION to fig. 3. An example is
+shown in figure \ref{fig:HJB_control_surface}. Note that the aspect
+of the new figure is rotated relative to the surface plots of $v(x,t)$, because
+the different surfaces are best seen from different angles respectively.
+
+TODO: Add to draft. 
+%\usepackage{graphics} is needed for \includegraphics
+\begin{figure}[htp]
+\begin{center}
+  \includegraphics[width=\textwidth]{Figs/HJB/Regimes_controlsurf.pdf}
+  \caption[labelInTOC]{Surface plots for $\alpha(x,t)$ analogous to the surface
+  plots for $v(x,t)$ in figure 3 in the draft}
+  \label{fig:HJB_control_surface}
+\end{center}
+\end{figure} 
 \end{rep}
@@
 \begin{rep}
-Both Susanne and Alex disagree. The point of the naive control as far as Alex is
-concerned is to provide a base-case against which to compare the 'optimal'
-stimulus. If the optimal stimulus improves things by 5,10\% then it is  hard to
-justify its computational cost. 
+Both Susanne and Alex disagree with the ref. The point of the naive control as
+far as Alex is concerned is to provide a base-case against which to compare the 'optimal'
+stimulus. Our main justification is that if the optimal stimulus improves things
+by 5,10\% then it is hard to justify its computational cost. 
 \end{rep}
@@
 \begin{rep}
-
+TODO
 \end{rep}
 
-\item In l. 7 in Page 23 of 33, “High Noise” should be deleted.
+\item In l. 7 in Page 23 of 33, 'High Noise' should be deleted.
 \begin{rep}
-TODO:
+TODO: Here is the offending line
+'In all cases,
+the target train is obtained by a simulation of the Supra-Threshold High Noise
+regime with no additional control, i.e.\ $\alpha(\cdot) = 0$.' It is actually on
+page 22 in draft (perhaps there is an extra first page in the referee's copy).
+
+I don't see why we should remove the High Noise qualifier. There is a
+significant difference between a target train generated by 'Supra-Threshold High
+Noise' vs.\ 'Supra-Threshold Low Noise'. 
 \end{rep}
@@
 \item In the fourth paragraph in Page 23 of 33, the authors mentioned “In fig. 11,
-we relax the bound constrains on the control $[\amin, \amax]$ from
+we relax the bound constrains on the control $[\alpha_{min}, \alpha_{max}]$ from
 [-2;2] to [-4;4].”. However they depicted the firing rates and dot raster plots only in
@@
 \begin{rep}
-TODO:
+TODO: Also do the relaxed bounds for open-loop???
+
+Unfortunately, averaging $\alpha(X_t,t)$ over the different tracking trains
+is not particularly sensible. In particular consider the case when one
+tracking train has just spiked, while the other is delayed. Then the average
+of the controls will be zero, even though one is max and the other is min. As
+showing an average seems uninformative of what is going on.
+
+Instead we show one closed-loop / open-loop full-train realization of the
+voltage / control signal, $X_t, \alpha(X_t, t)$ for a
+single train in figure \ref{fig:tracking_train_trajectories}. In a sense this is
+just 16 copies glued side-by-side of the realizations we see in section 5.1.
+
+%\usepackage{graphics} is needed for \includegraphics
+\begin{figure}[htp]
+\begin{center}
+\subfloat[Closed Loop]{
+  \includegraphics[width=\textwidth]{Figs/TrainController/SingleTrain_Trajectories_SUPT_supthn_8_16_cl.pdf}
+   }\\
+  \subfloat[Open Loop]{ 
+   \includegraphics[width=\textwidth]{Figs/TrainController/SingleTrain_Trajectories_SUPT_supthn_8_16_ol.pdf}
+  }
+  \caption[labelInTOC]{Full Tracking Train trajectories for $X_t$,
+  $\alpha(X_t,t)$}
+  \label{fig:tracking_train_trajectories}
+\end{center}
+\end{figure}
+
 \end{rep}
@@
 \begin{rep}
+
+Indeed, this is unprecise. Refractory period is clearly unreasonable, but more
+than that, we consider any period that is shorter than the time it takes to
+deterministically under maximum excitation.
+
+WE have added the following two sentences to the draft:
 TODO:
+
 \end{rep}
@@
 TODO:
+Are we going to make surface plots here again?
 \end{rep}
@@
 \begin{rep}
-The two traces are the same! They are added here to compare against
+The two traces are indeed the same! They are added here to compare against
 the traces on the right, so that people don't flip back and forth for figures to
-appreciate what has changed
+appreciate what has changed.
+We will put both epsilon, open-loop curves on the same plots (red vs.
+blue) see figure \ref{fig:epsilong_comparison_same_graph}.
+%\usepackage{graphics} is needed for \includegraphics
+\def \T {{ T}}
+\begin{figure}[htp]
+\begin{center}
+  \includegraphics[width=\textwidth]{Figs/FP_Adjoint/Regimes_eps_comparison_joined.pdf}
+  \caption[labelInTOC]{Effect of $\epsilon$ on the open-loop control.
+  The deterministic time-trajectory of $\alpha(t)$ for each
+  parameter regime as function of time, $t \in [0, \T]$.
+  On the left, we use $\epsilon = 0.001$ and on the right $\epsilon =0.1$.
+  The desired spike time is set to $\T=1.5$ and the bounds are $\alpha \in
+  [-2,2]$. A)
+   Supra-Threshold-Low-Noise
+    B) 
+   Supra-Threshold-High-Noise
+    C)
+   Sub-Threshold-Low-Noise 
+    D)
+   Sub-Threshold-High-Noise}
+  \label{fig:epsilong_comparison_same_graph}
+\end{center}
+\end{figure}
+
 \end{rep}
@@
 clarify that in all cases, the maximum value of the control puts the model in
-the Supra-Threshold regime. That is using the control one can always have
-Supra-Threshold behaviour. The Supra / Sub- distinction only applies to the
-intrinsic (uncontrolled dynamics)
+the Supra-Threshold regime. This means that by using the control one can
+always achieve Supra-Threshold behaviour. The Supra / Sub- distinction only
+applies to the intrinsic (uncontrolled dynamics) and they are defined in the
+absence of control.
+
+TODO: Add ANdre's notes on Stochastic Resonance from his letter here. 
 \end{rep}
@@
 TODO:
+This has to be intra-cellular injected current.
+
+We require to have access to the output of the neuron.
+
+We can use optogenetics\ldots
 \end{rep}
@@
 \begin{rep}
-TODO:
+TODO: Include more single-cell applications.
 \end{rep}
@@
 \begin{rep}
-TODO:
+It is our impression that (Pontryagin's) Maximum Principle and (Bellman's)
+Dynamic Programing approach are two distinct analytical methods for solving
+Optimal Control problems.
+
+The Maximum Principle is essentially a generalization of The Calculus of
+Variations wherein it finds a critical point for the objective, akin to the
+First Derivative Test in 1st year Calculus.
+(refer to sources here)
+
+Dynamic Programing is a sequential approach which builds up the optimal control
+from the end-time recursively, using a value function to efficiently represent
+the entire future payoff / cost associated with an incremental decision. (refer to
+sources here)
+
+There are connections between the two, most saliently, under certain conditions,
+the adjoint function in the Maximum Principle is the spatial gradient of the
+value function in Dynamic Programing (reference). But the two arise from
+different paradigms and in turn give rise to different numerical methods.
+
+The difference in numerical methods is apparent in particular in that when we
+apply Dynamic Programing, we can infer the optimal control right away, while for
+the Maximum Principle, we are forced to use an iterative, gradient descent
+procedure.
+
 \end{rep}
@@
 TODO:
+Running N parallel neurons is not a problem, naturally.
+
+However, controlling N coupled neurons is VERY challenging and certainly beyond
+the limits of the techniques presented as is. 
 \end{rep}
@@
 paper if the authors made a somewhat elaborate connection between their
-formulism to that of a bang-bang control formulism. And also explain why these
+formalism to that of a bang-bang control formalism. And also explain why these
 controls come out almost balanced in negative and positive values.
@@
 TODO:
+1. Indeed as $\epsilon$ goes to zero, the optimization becomes bang-bang.
+
+2. The occassional case where the controls are charge-balanced happens to be 
 \end{rep}
@@
 What do they mean by $x<<0$? Looking at Figure 1, min(x) is usually around zero,
-unless there is high noise, in which case it becomes in the order of -1 to -10 probably. 
-Please elaborate a little.
+unless there is high noise, in which case it becomes in the order of -1 to -10
+probably. Please elaborate a little.
 \begin{rep}
-TODO:
+TODO: Here is the offending paragraph redone:
+ \def \xmin {{ x_{-}}}
+  \def \amin {{ \alpha_{-}}}
+  \def \tc {{ \tau_c}}
+ 'For $x \ll 0$, we assume that $w$ is not
+significantly affected by the change in $x$, i.e.\ that $$ \partial_x
+w(x_{-}, t) = 0 $$ for some lower boundary $x_{-}$. Such a boundary condition will be
+justified if we impose artificial reflecting boundaries for the diffusion at
+$\xmin$. So we need to determine for which $\xmin$ is a reflecting boundary
+consistent with the dynamics. For example, we can take $\xmin$ to be two
+standard deviations below the mean of the stationary distribution of the
+maximally inhibited process, i.e.\ setting $\alpha=\amin$ in eq. [2].
+ That is we set $\xmin = \tc(\mu + \amin) - 2 \beta /
+\sqrt{\tc/2}$. We further enforce that $\xmin \leq -0.5$.'
 \end{rep}
-P10, Caption of Figure 2: "Note that the green and blue curves in F) are lying on top of each other for all x." This is not true.
-Did the authors mean subplot (D)?
-And can you explain why it is that the green and blue curves are on top of each other?
+
+P10, Caption of Figure 2: "Note that the green and blue curves in F) are lying
+on top of each other for all x." This is not true. Did the authors mean subplot
+(D)? And can you explain why it is that the green and blue curves are on top of
+each other?
 \begin{rep}
-TODO:
+ Indeed it is D) and thanks!
+
+TODO: Add explanation for why this is so.
 \end{rep}
-The initial condition for the computations seems to be the point of previous spike, x=0. Is that right? Please make note of this where ever appropriate for clarity, for example, when showing Figure 4.
+
+The initial condition for the computations seems to be the point of previous
+spike, x=0. Is that right? Please make note of this where ever appropriate for
+clarity, for example, when showing Figure 4.
 \begin{rep}
-TODO:
+TODO: Add a reminder in Fig 4 that x=0 at time t=0.
 \end{rep}
-Can the authors elaborate on the noise that they implemented in their simulations, and discuss the shape of the distributions in Figure 5, especially, for the deterministic case?
+
+Can the authors elaborate on the noise that they implemented in their
+simulations, and discuss the shape of the distributions in Figure 5, especially,
+for the deterministic case?
 \begin{rep}
-TODO:
+TODO: Hmm\ldots We use standard Wiener-type driving force for the dynamics in
+the stochastic. What is shown in Figure 5. is a function of the first passage
+time of such a noise. 
 \end{rep}
@@
 \begin{rep}
-TODO:
+TODO: The dots above the horizontal line indicate the target spike times. The
+dots below the horizontal line indicate the occurence of the tracking spike
+times. 
 \end{rep}
-"i.e." usually has commas before and after. Also, at the beginning of sentence, please consider using the words "That is" instead of "I.e." (P12,L19).
-For consistency, please consider using either Supra- or Super-Threshold throughout.
+
+"i.e." usually has commas before and after. Also, at the beginning of sentence,
+please consider using the words "That is" instead of "I.e." (P12,L19). For
+consistency, please consider using either Supra- or Super-Threshold throughout.
 \begin{rep}
+Definitely yes on consistent use of 'Supra-' vs. 'Super-'. Thanks. 
 TODO:  
@@
 
-\begin{rep} TODO: \end{rep}
+\begin{rep} TODO: 
+We agree.
+\end{rep}
 
@@
 
-\begin{rep} TODO: \end{rep}
+\begin{rep} TODO:
+
+Let's ignore the second part (on extracellular stimulation) and concentrate on
+the first part (more complicated models).
+
+TODO: Extracellular stimulation in the discussion only!
+\end{rep}
 
@@
 introduction. 
-
-\begin{rep} TODO: \end{rep}
+\begin{rep} 
+TODO: 
+\end{rep}
 
 \subsection{Specific points}
-
 Abstract: The authors state that their goal was to design a control scheme to
@@
 
-\begin{rep} TODO: \end{rep}
+\begin{rep} 
+TODO:
+ \end{rep}
 
@@
 
-p. 2, par. 3.  The sentence beginning “However the efficiency of the numerics …”
-is nearly impossible to parse and understand.
+p. 2, par. 3.  The sentence beginning ``However the efficiency of the
+numerics\ldots'' is nearly impossible to parse and understand.
 
